### 2017-09-12 Paper/disparity_review.tex
review INIT

\documentclass[12pt,letterpaper]{article}
\usepackage{natbib}

%Packages
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{dcolumn}
\newcolumntype{d}[1]{D{.}{.}{#1}}

\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
%\bibpunct{(}{)}{;}{a}{}{,}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\begin{document}

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: disparate views on disparity.

\bigskip
\medskip
\begin{center}

\noindent{\Large \bf Disparate views on disparity.} 
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Natalie Cooper$^2$, Philip Donoghue$^3$}\\
\noindent {\small \it 
$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\}
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

\newpage
\begin{abstract}

\begin{enumerate}
    blablalbalba
\end{enumerate}

\end{abstract}

\noindent (Keywords: disparity)\\

\vspace{1.5in}

\newpage

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------

\section{Introduction}

Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.
One solution to this problem is to analyse data in a multivariate framework.
Multivariate analyses aim to capture and incorporate this multidimensional complexity, while providing outputs that are interpretable in a physical world of only three dimensions.
Classical multidimensional species features that have been analysed through multivariate data ranges from their morphology \citep{raup1966geometric}, to their functional traits \citep{diaz2016global}.
However, such analysis are not limited at the species level and can also be applied to ecosystems \citep{DonohueDim}.

Key multidimensional features of species that have important roles in ecology and evolution include morphology, functional traits. etc. etc.
Many of these can be represented as matrices, these can be used in our package.
In the interests of clarity/brievety, we here focus on just one kind of data: morphological diversity.

Furthermore, we will also make a clear distinction between the multidimensional space which is a specific mathematical object and the disparity, which is a metric describing or summarising one or more aspects of this space.

Both can be defined in various ways depending on the authors, for example, disparity is defined as the weighted mean pairwise dissimilarity in \cite{Close2015} or the ellipsoid volume in \cite{DonohueDim} and the multidimensional space is defined as the morphospace in \cite{raup1966geometric} and the morpho-functional space in \cite{diaz2016global}.

The multidimensional space can be defined in many ways and arise from many mathematical transformations of the data such as the pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
Similarly, disparity metrics \citep[or indices;][]{Hopkins2017} can defined in many ways \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015,diaz2016global}.
Finally, difference between disparity metrics can also be measured in many ways: using NPMANOVA \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simple confidence interval overlap \citep[e.g.][]{halliday2016eutherian}.

This variety of definitions and analysis have been developed in an equal variety of softwares such as \texttt{GINGKO} in javascript \citep{bouxin2005ginkgo,de2007ginkgo} or \texttt{geomorph} \citep{adams2013geomorph,adams2017geometric}, \texttt{Claddis} \citep{Claddis}, or \texttt{vegan} \citep{oksanen2007vegan} in R \citep{R}.
This results in the need to learn different languages (or at least - when restricted to R - different packages with different standards) as well as making analysis sometimes idiosyncratic and often complex to repeat since they are based on a particular feature from a particular software.
For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function.
Unfortunately, however, the multidimensional space can only be defined as the ordination of the procrustes transform of geometric morphometric landmarks, the disparity can only be defined as the Procrustes variance and the difference between groups can only be measured through permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.

Disparity is an old concept blabalbal

\section{What \textit{is} disparity}

From a semantic view: morphospace or aspect of the morphospace?

From a biological view: what does an increase or a decrease in disparity represent?

\section{Why do we need disparity}

What are the fundamental questions we should use to answer disparity questions?

What are the questions we should not answer with disparity?

\section{What is missing}

Methods?
Data?

\section{Conclusion}

Yay!

% \section{Data availability and reproducibility}
% Data are available on Dryad or Figshare.
% Code for reproducing the analyses is available on GitHub (\url{github.com/TGuillerme/SpatioTemporal_Disparity}).

\section{Acknowledgments}
Royal Society..
I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.

\bibliographystyle{sysbio}
\bibliography{References}

\end{document}

### 2017-10-27 Functions/fuzzy.sort.data.R
Data handling functions

@@ -1,0 +1,45 @@
+#' @title Sorting the data
+#'
+#' @description Sorting the data for the disparity review by fuzzy matching
+#'
+#' @param data a column of the dataset
+#' @param what a vector of categories
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+fuzzy.sort.data <- function(data, what) {
+    ## Set all lower case for easy match
+    what_l <- tolower(what)
+    data_l <- tolower(data)
+
+    ## Setting up the matching list
+    total_matches <- data.frame()
+
+    ## Loop through the fuzzy matching
+    for(element in 1:length(what_l)) {
+        ## Check fuzzy match
+        fuzzy_match <- grep(what_l[element], data_l)
+        length_match <- length(fuzzy_match)
+        
+        if(length_match > 0) {
+            ## If match, record the matches
+            matches <- data.frame("data" = fuzzy_match, "what" = rep(what[element], length_match))
+            total_matches <- rbind(total_matches, matches)
+        }
+    }
+
+    ## Getting the non-matching
+    others <- which(!(1:length(data) %in% c(unique(total_matches$data))))
+    length_others <- length(others)
+    if(length_others > 0) {
+        others <- data.frame("data" = others, "what" = rep("other", length_others))
+        total_matches <- rbind(total_matches, others)
+    }
+
+    return(total_matches)
+}

### 2017-10-27 Functions/plot.data.R

@@ -1,0 +1,55 @@
+#' @title Plot data
+#'
+#' @description Plot the data sorting
+#'
+#' @param sorted.data the sorted data per categories
+#' @param proportion whether to use proportional data or not (default = FALSE)
+#' @param decreasing in which order to plot the categories (default, TRUE, right to left decreasing)
+#' @param las las parameter (default = 3)
+#' @param table whether to print the table in "R" format (default), "tex" or "md".
+#' @param ... optional arguments to be passed to \coe{barplot}
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+plot.data <- function(sorted.data, proportion = FALSE, decreasing = TRUE, las = 3, ...) {
+
+    ## Make the data table
+    plot_data <- apply(table(sorted.data), 2, sum)
+
+    ## Proportions
+    if(proportion) {
+        plot_data <- plot_data/sum(plot_data)*100
+    }
+
+    ## Plot the data
+    barplot(sort(plot_data, decreasing = decreasing), las = las, ...)
+
+}
+
+table.data <- function(sorted.data, proportion = FALSE, decreasing = TRUE, table = "R", ...) {
+    ## Make the data table
+    table_data <- apply(table(sorted.data), 2, sum)
+
+    ## Proportions
+    if(proportion) {
+        table_data <- table_data/sum(table_data)*100
+    }
+    
+    table_data <- t(t(table_data))
+
+    ## Table
+    if(table == "R") {
+        return(table_data)
+    }
+    if(table == "latex") {
+        return(xtable::xtable(table_data, ...))
+    }
+    if(table == "md") {
+        return(knitr::kable(table_data, ...))
+    }
+}

### 2017-10-27 Analysis/disparity_methods.Rmd
methods draft

---
title: "Disparate views of disparity"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

```{r}
source("../Functions/fuzzy.sort.data.R")
```

# Data

## Data collection from google scholars

I collected the data from Google scholar per year for the key words "morphological disparity".
I analysed all 100 <!--@@@ TG: check if I stick to this value or not --> papers for each year and collected the following data:

 * The paper details
 * Whether or not the paper contained a disparity analysis (for consistency)
 * The focal biological group
 * The type of data (3D landmarks, discrete data, etc...)
 * Whether the full organism was used or not.
 * How was the morphospace implicitly defined?
 * How was disparity implicitly defined?
 * Which test was applied for testing hypothesis?
 * Was phylogeny taken into account?

## Data wrangling

I then categorised the data as follwos.
<!-- @@@ -->

```{r}
## Reading the data
data_raw <- read.csv("../Data/disparity_methods.csv", stringsAsFactors = FALSE)

## Keeping only the studies with data
data <- data_raw[apply(data_raw[, 3:9], 1, function(X) ifelse(all(X == "-"), FALSE, TRUE)), ]
```

## Sorting the data to categorise it

The categories for each variable are:

```{r}
## Categories definition
data_type <- c("3D landmarks", "2D landmarks", "continuous", "discrete", "body", "metabolic")
full_organism <- c("yes", "no")
morphospace <- c("PCA", "MDS", "PCO", "PCOA", "Raup")
disparity <- c("sum of variance", "sum of range", "product of variance", "product of range", "distance", "observed") #TG: Area from observed to estimated (DTT)
test <- c("anova", "model fitting", "canonical")
phylo <- c("yes", "no", "mapped")

factors <- list("data type" = data_type, "full organism" = full_organism, "morphospace" = morphospace, "disparity" = disparity, "test" = test, "phylo correction" = phylo)
```

```{r, fig.width = 12, fig.height = 12}
## Plotting the results
par(mfrow = c(2, 3))
for(factor in 1:length(factors)) {
    plot.data(fuzzy.sort.data(data[,3+factor], factors[[factor]]), proportion = TRUE, ylim = c(0, 100), main = names(factors)[[factor]])    
}
```

```{r}
## Printing the tables
factor <- 1
table.data(fuzzy.sort.data(data[,3+factor], factors[[factor]]), table = "tex")

```

<!-- 
```{r, eval = FALSE, message = FALSE, fig.width = 12, fig.height = 12, echo = FALSE}
```
 -->

### 2017-11-01 Analysis/disparity_methods.Rmd
Updated pipeline

@@
 
-I then categorised the data as follwos.
+I then categorised the data as follows.
 <!-- @@@ -->
@@
 <!-- 
+How many studies use such and such definition: disparate views on disparity -> but maybe disparity is a vague concept?
+How many studies use different definition for such or such parameter recorded: on average, disparity is defined as a single thing...
+
+Parallel with diversity: it was a blurry thing - check out different diversity literature.
+ -->
+
+# Supplementary
+
+## Number of papers collected for each parameter
+
+```{r, fig.width = 12, fig.height = 12}
+## Get the data for cumulative curves
+total_cumulative <- data_raw[,c(1,3)]
+#data_type_cumulative <- data_raw[,c(1,4)]
+
+## Transform the cumulative data into binary
+results <- grep("-", total_cumulative[,2])
+results <- cbind(results, seq(1:length(results)))
+
+plot(add, type = "line", xlab = "Number of Google Scholar matches", ylab = "Number of new entries")
+
+```
+
+<!-- 
 ```{r, eval = FALSE, message = FALSE, fig.width = 12, fig.height = 12, echo = FALSE}

### 2017-11-01 Analysis/disparity_methods.Rmd
methods update

@@
 
+Additionally I collected the number of matches per year for the terms "morphological disparity" both in the text (fuzzy matching) or in the title (exact matching).
+
 ## Data wrangling
@@
 data_raw <- read.csv("../Data/disparity_methods.csv", stringsAsFactors = FALSE)
+search_results <- read.csv("../Data/GoogleSearch_occurences.csv", stringsAsFactors = FALSE)
 
@@
 
+# Occurence of morphological disparity papers
+
+```{r}
+disparity_fuzzy <- search_results[-c(1:2, 60:65),3]
+disparity_exact <- search_results[-c(1:2, 60:65),2]
+
+## Number of papers
+plot(seq(from = 1960, to = 2016), rev(disparity_fuzzy)/max(disparity_fuzzy), type = "l", xlab = "", ylab = "Proportion of papers")
+lines(seq(from = 1960, to = 2016), rev(disparity_exact)/max(disparity_exact), type = "l", col = "grey")
+```
+
 # Supplementary

### 2017-12-06 Writing/disparity_review.tex
paper INIT

--- 2017-09-12 Paper/disparity_review.tex
@@
 \usepackage{pdflscape}
-\usepackage{fixltx2e}
 \usepackage{textcomp}
@@
 
+Disparity, as a concept or set of toolkits is used a lot in biology and is more and more used every year.
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/GoogleScholarOccurences.pdf} 
+\caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year. In black, the match is in the paper and in grey, in the title.}
+\label{Fig:GoogleOccurences}
+\end{figure}
+
+In most of these papers, they trace back the disparity concept to seminal papers from the 90s: \cite{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution}.
+However, being such a broad and obvious concept, it can probably be stretched back to way earlier.
+
+\cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.
+
 Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.

### 2017-12-06 Analysis/data_cleaning.Rmd
data cleaning pipeline

---
title: "Cleaning the data"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Sorting the disparity methods dataset

```{r}
source("../Functions/fuzzy.sort.data.R")
```

First lets load the data:

```{r}
## Reading the data
data_raw <- read.csv("../Data/disparity_methods.csv", stringsAsFactors = FALSE)

## Keeping only the studies with data
data <- data_raw[apply(data_raw[, 3:9], 1, function(X) ifelse(all(X == "-"), FALSE, TRUE)), ]

## Removing rogue NA rows
data <- data[-231,]
```

Then, lets sort the following variables:

 * The focal biological group
 * The type of data (3D landmarks, discrete data, etc...)
 * Whether the full organism was used or not.
 * How was the morphospace explicitly defined?
 * How was disparity explicitly defined?
 * Which test was applied for testing hypothesis?
 * Was phylogeny taken into account?

## Biological group

Sorting the biological groups into broader categories

```{r}
## Biological group
data_taxa <- data[,3]

## Mammals synonyms
mammals <- c("Felids", "Carnivora", "Primates", "Rodents", "Whales", "Cetacean", "Cetaceans", "Metatheria", "Marsupials", "Humans", "Perissodactyla", "Lipotyphla", "Mammals")
data_taxa <- ifelse(data_taxa %in% mammals, "mammals", data_taxa)

## Aves synonyms
aves <- c("Meliphagidae", "Birds")
data_taxa <- ifelse(data_taxa %in% aves, "aves", data_taxa)

## Archosaurs (non-aves)
archosaurs <- c("Pterosaurs", "Crocodiles", "Sauropoda", "Dinosaurs", "Theropods", "Archosauria")
data_taxa <- ifelse(data_taxa %in% archosaurs, "archosaurs", data_taxa)

## Squamates
squamates <- c("Squamates", "Rhynchocephalia", "Anolis", "Squamates")
data_taxa <- ifelse(data_taxa %in% squamates, "squamates", data_taxa)

## Other tetrapods
amphib_other_tetrapods <- c("Plesiosaur", "Captorhinidae", "Anomodonts", "Ichtyosaurs", "Amphibians", "Frogs", "Caecilian")
data_taxa <- ifelse(data_taxa %in% amphib_other_tetrapods, "amphib_other_tetrapods", data_taxa)

## Arthropods
arthropods <- c("Odonata", "Trilobites", "Ostracodes", "Scorpions", "Insects", "Eurypterida", "Orthoptera", "Arthropods", "Crustacean", "Beetles", "Hymenoptera", "Amphipodes")
data_taxa <- ifelse(data_taxa %in% arthropods, "arthropods", data_taxa)

## Plants
plantae <- c("Angiosperms", "Rubiacea", "Marchantiophyta", "Ferns + angiosperms")
data_taxa <- ifelse(data_taxa %in% plantae, "plantae", data_taxa)

## Echinoderms
echinoderms <- c("Echinoderms", "Crinoids", "Echinoids")
data_taxa <- ifelse(data_taxa %in% echinoderms, "echinoderms", data_taxa)

## Nematods
cnidaria_other_invert <- c("Nematode", "Brachiopods", "Onychophora", "Palaeoscolecid worms", "Graptolites", "Cnidera", "Corals", "Cnidaria")
data_taxa <- ifelse(data_taxa %in% cnidaria_other_invert, "cnidaria_other_invert", data_taxa)

## Actinopterigygians
actino_chondry <- c("Cichlids", "Fish", "Fishes", "Shark")
data_taxa <- ifelse(data_taxa %in% actino_chondry, "actino_chondry", data_taxa)

## Molluscs
molluscs <- c("Ammonites", "Gastropod", "Ammonoids", "Cephalopoda", "Mollusca", "Bivalves")
data_taxa <- ifelse(data_taxa %in% molluscs, "molluscs", data_taxa)

## Unicellulars
unicellulars <- c("Foraminifera", "Desimidiales", "Acritarch")
data_taxa <- ifelse(data_taxa %in% unicellulars, "unicellulars", data_taxa)

## General
general <- c("Metazoans", "Ediacarian", "Blastozoa + Crinoids + Trilobite", "Echinoids + Trilobites", "Crinoids + Balstozoa", "Marine invertebrates", "Amniotes", "Vertebrates")
data_taxa <- ifelse(data_taxa %in% general, "general", data_taxa)

## Simulated
simualted <- c("Simulated", "-", "")
data_taxa <- ifelse(data_taxa %in% simualted, "simualted", data_taxa)

## Selecting the categories
data_taxa_categories <- unique(data_taxa)

data[,3] <- data_taxa
```

## Type of data

```{r}
## Types of data
data_type <- data[,4]

unique(data_type)

## continuous synonyms
continuous <- c("1D Continuous", "DTT pipeline", "Continuous")
data_type <- ifelse(data_type %in% continuous, "continuous", data_type)

## Math transfo
math_transfo <- c("B-spline", "Elliptic Fourier", "Outlines", "Ellipctic Fourier")
data_type <- ifelse(data_type %in% math_transfo, "math_transfo", data_type)

## Other_univar
other_univar <- c("Body mass", "Metabolic Rate", "Body size", "Glands for chemical communication", "Chemical signals", "Colours")
data_type <- ifelse(data_type %in% other_univar, "other_univar", data_type)

## Discrete
data_type <- ifelse(data_type == "Discrete", "discrete", data_type)

## Non
none <- c(NA, "-")
data_type <- ifelse(data_type %in% none, "none", data_type)

## 2d + discrete
d2_discrete <- c("Discrete + 2D landmarks")
data_type <- ifelse(data_type %in% d2_discrete, "discrete + 2D landmarks", data_type)

## 2d + continuous
d2_continu <- c("Continuous + 2D landmarks")
data_type <- ifelse(data_type %in% d2_continu, "continuous + 2D landmarks", data_type)

## continuous + other
continu_other <- c("Continuous + Body size")
data_type <- ifelse(data_type %in% continu_other, "continuous + other_univar", data_type)

## continuous + discrete
continu_discrete <- c("Continuous + discrete" )
data_type <- ifelse(data_type %in% continu_discrete, "continuous + discrete", data_type)

## Selecting the categories
data_type_categories <- unique(unlist(strsplit(unique(data_type), split = " \\+ ")))

data[,4] <- data_type
```

## Full organism

```{r}
## Full organism
full_organism <- data[,5]

## Selecting the categories
full_organism_categories <- c("Yes", "No", NA)
```

## Morphospaces

```{r}
## Morphospaces
morphospaces <- data[,6]

unique(morphospaces)

## PCA
PCA <- c("PCA", "Procrustes PCA", "pPCA", "Regression PCA")
morphospaces <- ifelse(morphospaces %in% PCA, "PCA", morphospaces)

## PCO PCoA
PCO <- c("PCO", "PCoA", "PcoA", "FMD (Foote 1993)")
morphospaces <- ifelse(morphospaces %in% PCO, "PCO", morphospaces)

## MDS NMDS
MDS <- c("MDS", "NMDS", "nMDS")
morphospaces <- ifelse(morphospaces %in% MDS, "MDS", morphospaces)

## RWA
RWA <- c("RWA", "RWA (Relative Warp Analysis)")
morphospaces <- ifelse(morphospaces %in% RWA, "RWA", morphospaces)

## Canonical Analysis
canonical <- c("Canonical variables", "canonical correlation analysis (CCA)")
morphospaces <- ifelse(morphospaces %in% canonical, "canonical", morphospaces)

## Correspondence Analysis
correspondence <- c("Correspondence analysis", "MCA (multiple correspondence analyses)")
morphospaces <- ifelse(morphospaces %in% correspondence, "correspondence", morphospaces)

## Other
other <- c("Eigen shape analysis", "Factorial planes","Biplot (two traits only)", "biplot (residuals of two traits)", "Factorial planes","Procrustes plane (2 D axis)", "Procrustes coordinates")
morphospaces <- ifelse(morphospaces %in% other, "other", morphospaces)

## None
none <- c("DTT pipeline")
morphospaces <- ifelse(morphospaces %in% none, "none", morphospaces)
morphospaces <- ifelse(is.na(morphospaces), "none", morphospaces)

## Combos
morphospaces <- ifelse(morphospaces %in% "PCA + canonical variables", "PCA + canonical", morphospaces)
morphospaces <- ifelse(morphospaces %in% "Dissimilarity matric + PCoA", "PCO + other", morphospaces)
morphospaces <- ifelse(morphospaces %in% "RWA (Relative Warp Analysis) + PCO", "PCO + RWA", morphospaces)
morphospaces <- ifelse(morphospaces %in% "CVA + PCA" , "PCA + canonical", morphospaces)

## Selecting the categories
morphospaces_categories <- unique(unlist(strsplit(unique(morphospaces), split = " \\+ ")))

```

```{r}
disparity <- c("sum of variance", "sum of range", "product of variance", "product of range", "distance", "observed") #TG: Area from observed to estimated (DTT)
test <- c("anova", "model fitting", "canonical", "confidence interval")
phylo <- c("yes", "no", "mapped")

factors <- list("data type" = data_type, "full organism" = full_organism, "morphospace" = morphospace, "disparity" = disparity, "test" = test, "phylo correction" = phylo)
```

### 2017-12-06 Analysis/disparity_methods.Rmd
data cleaning pipeline

--- 2017-11-01 Analysis/disparity_methods.Rmd
@@
 
-I collected the data from Google scholar per year for the key words "morphological disparity".
-I analysed all 100 <!--@@@ TG: check if I stick to this value or not --> papers for each year and collected the following data:
+I collected the data from the 500 first Google Scholar results for the key words "morphological disparity" per order of appearance (accessed 1st of Noveember 2017).
+This resulted in 232 relevant papers for which I collected:
 
- * The paper details
- * Whether or not the paper contained a disparity analysis (for consistency)
  * The focal biological group
@@
  * Whether the full organism was used or not.
- * How was the morphospace implicitly defined?
- * How was disparity implicitly defined?
+ * How was the morphospace explicitly defined?
+ * How was disparity explicitly defined?
  * Which test was applied for testing hypothesis?
  * Was phylogeny taken into account?
+
+In the other 268 papers disparity was mentioned but not measured.
+This happened for papers falling in the following categories:
+
+ * Out of topic
+ * Morphological descriptions
+ * Review paper
+ * Paper not accessible (broken link or paywall)
+ * Citation only
+
+Additionally I collected the number of matches per year from 1960 to 2016 for the terms "morphological disparity" both in the text (fuzzy matching) or in the title (exact matching).
 
@@
 
-## Sorting the data to categorise it
+# Occurrences of morphological disparity papers
 
-The categories for each variable are:
+This first analysis just shows that disparity analysis in morphology are steadily and rapidly increasing since the 2000s:
+
+```{r, fig.width = 8, fig.height = 8}
+disparity_fuzzy <- search_results[-c(1:2, 60:65),3]
+disparity_exact <- search_results[-c(1:2, 60:65),2]
+
+## Number of papers
+# pdf("../Paper/Figures/GoogleScholarOccurences.pdf", width = 8, height = 8)
+op <- par(bty = "n")
+plot(seq(from = 1960, to = 2016), rev(disparity_fuzzy)/search_results[1,3], type = "l", xlab = "", ylab = "Relative number of papers per year", ylim = c(0, 0.1))
+lines(seq(from = 1960, to = 2016), rev(disparity_exact)/search_results[1,2], type = "l", col = "grey")
+par(op)
+# dev.off()
+```
+
+Not much to say here but that it is a trendy topic!
+
+# Disparate views of disparity
+
+Disparity methods comes in all flavours!
+
+## Sorting the variables
+
+The variables can be sorted as follow:
 
 ```{r}
+
 ## Categories definition
@@
 disparity <- c("sum of variance", "sum of range", "product of variance", "product of range", "distance", "observed") #TG: Area from observed to estimated (DTT)
-test <- c("anova", "model fitting", "canonical")
+test <- c("anova", "model fitting", "canonical", "confidence interval")
 phylo <- c("yes", "no", "mapped")
@@
 
+# Occurrences of morphological disparity papers
+
+```{r}
+disparity_fuzzy <- search_results[-c(1:2, 60:65),3]
+disparity_exact <- search_results[-c(1:2, 60:65),2]
+
+## Number of papers
+pdf("../Writing/Figures/GoogleScholarOccurences.pdf", width = 8, height = 8)
+op <- par(bty = "n")
+plot(seq(from = 1960, to = 2016), rev(disparity_fuzzy)/search_results[1,3], type = "l", xlab = "", ylab = "Number of papers per year", ylim = c(0, 0.1), yaxt = "n")
+lines(seq(from = 1960, to = 2016), rev(disparity_exact)/search_results[1,2], type = "l", col = "grey")
+## Adding the axis
+axis(2, at = seq(from = 0, to = 0.08, by = 0.02), labels = round(seq(from = min(disparity_fuzzy), to = max(disparity_fuzzy), length.out = 5)) )
+axis(4, at = seq(from = 0, to = 0.1, by = 0.02), labels = round(seq(from = min(disparity_exact), to = max(disparity_exact), length.out = 6)), col = "grey", las = 2)
+
+par(op)
+dev.off()
+```
+
 # Supplementary

### 2017-12-07 Functions/fuzzy.sort.data.R
multi.gsub function

@@ -43,3 +43,33 @@ fuzzy.sort.data <- function(data, what) {
 
     return(total_matches)
 }
+
+#' @title Multiple gsub
+#'
+#' @description Using gsub with multiple pattern
+#'
+#' @param pattern a vector of greps to replace
+#' @param replacement the value to replace with
+#' @param x the data
+#' 
+#' @examples
+#'
+#' @seealso
+#' 
+#' @author Thomas Guillerme
+#' @export
+
+multi.gsub <- function(pattern, replacement, x, ignore.case = TRUE, ...) {
+
+    pattern <- sort(pattern, decreasing = TRUE)
+
+    for(pat in 1:length(pattern)) {
+        x <- gsub(pattern[pat], replacement, x, ignore.case = ignore.case, ...)
+    }
+
+    return(x)
+}
+
+
+
+

### 2017-12-07 Analysis/data_cleaning.Rmd
Data cleaning up to test

@@
 ## Mammals synonyms
-mammals <- c("Felids", "Carnivora", "Primates", "Rodents", "Whales", "Cetacean", "Cetaceans", "Metatheria", "Marsupials", "Humans", "Perissodactyla", "Lipotyphla", "Mammals")
-data_taxa <- ifelse(data_taxa %in% mammals, "mammals", data_taxa)
+mammals <- c("Felids", "Carnivora", "Primates", "Rodents", "Whales", "Cetaceans", "Metatheria", "Marsupials", "Humans", "Perissodactyla", "Lipotyphla", "Mammals", "mammals")
+data_taxa <- multi.gsub(mammals, "mammals", data_taxa)
 
@@
 aves <- c("Meliphagidae", "Birds")
-data_taxa <- ifelse(data_taxa %in% aves, "aves", data_taxa)
+data_taxa <- multi.gsub(aves, "aves", data_taxa)
 
@@
 archosaurs <- c("Pterosaurs", "Crocodiles", "Sauropoda", "Dinosaurs", "Theropods", "Archosauria")
-data_taxa <- ifelse(data_taxa %in% archosaurs, "archosaurs", data_taxa)
+data_taxa <- multi.gsub(archosaurs, "archosaurs", data_taxa)
 
@@
 squamates <- c("Squamates", "Rhynchocephalia", "Anolis", "Squamates")
-data_taxa <- ifelse(data_taxa %in% squamates, "squamates", data_taxa)
+data_taxa <- multi.gsub(squamates, "squamates", data_taxa)
 
@@
 amphib_other_tetrapods <- c("Plesiosaur", "Captorhinidae", "Anomodonts", "Ichtyosaurs", "Amphibians", "Frogs", "Caecilian")
-data_taxa <- ifelse(data_taxa %in% amphib_other_tetrapods, "amphib_other_tetrapods", data_taxa)
+data_taxa <- multi.gsub(amphib_other_tetrapods, "amphib_other_tetrapods", data_taxa)
 
 ## Arthropods
-arthropods <- c("Odonata", "Trilobites", "Ostracodes", "Scorpions", "Insects", "Eurypterida", "Orthoptera", "Arthropods", "Crustacean", "Beetles", "Hymenoptera", "Amphipodes")
-data_taxa <- ifelse(data_taxa %in% arthropods, "arthropods", data_taxa)
+arthropods <- c("Odonata", "Trilobites", "Trilobite", "Ostracodes", "Scorpions", "Insects", "Eurypterida", "Orthoptera", "Arthropods", "Crustacean", "Beetles", "Hymenoptera", "Amphipodes")
+data_taxa <- multi.gsub(arthropods, "arthropods", data_taxa)
 
 ## Plants
-plantae <- c("Angiosperms", "Rubiacea", "Marchantiophyta", "Ferns + angiosperms")
-data_taxa <- ifelse(data_taxa %in% plantae, "plantae", data_taxa)
+plantae <- c("Angiosperms", "Rubiacea", "Marchantiophyta", "angiosperms", "Ferns")
+data_taxa <- multi.gsub(plantae, "plantae", data_taxa)
 
 ## Echinoderms
-echinoderms <- c("Echinoderms", "Crinoids", "Echinoids")
-data_taxa <- ifelse(data_taxa %in% echinoderms, "echinoderms", data_taxa)
+echinoderms <- c("Echinoderms", "Crinoids", "Echinoids", "Blastozoa", "Balstozoa")
+data_taxa <- multi.gsub(echinoderms, "echinoderms", data_taxa)
 
 ## Nematods
-cnidaria_other_invert <- c("Nematode", "Brachiopods", "Onychophora", "Palaeoscolecid worms", "Graptolites", "Cnidera", "Corals", "Cnidaria")
-data_taxa <- ifelse(data_taxa %in% cnidaria_other_invert, "cnidaria_other_invert", data_taxa)
+other_invert <- c("Nematode", "Brachiopods", "Onychophora", "Palaeoscolecid worms", "Graptolites", "Cnidera", "Corals", "Cnidaria")
+data_taxa <- multi.gsub(other_invert, "other_invert", data_taxa)
 
@@
 actino_chondry <- c("Cichlids", "Fish", "Fishes", "Shark")
-data_taxa <- ifelse(data_taxa %in% actino_chondry, "actino_chondry", data_taxa)
+data_taxa <- multi.gsub(actino_chondry, "actino_chondry", data_taxa)
 
@@
 molluscs <- c("Ammonites", "Gastropod", "Ammonoids", "Cephalopoda", "Mollusca", "Bivalves")
-data_taxa <- ifelse(data_taxa %in% molluscs, "molluscs", data_taxa)
+data_taxa <- multi.gsub(molluscs, "molluscs", data_taxa)
 
@@
 unicellulars <- c("Foraminifera", "Desimidiales", "Acritarch")
-data_taxa <- ifelse(data_taxa %in% unicellulars, "unicellulars", data_taxa)
+data_taxa <- multi.gsub(unicellulars, "unicellulars", data_taxa)
 
 ## General
-general <- c("Metazoans", "Ediacarian", "Blastozoa + Crinoids + Trilobite", "Echinoids + Trilobites", "Crinoids + Balstozoa", "Marine invertebrates", "Amniotes", "Vertebrates")
-data_taxa <- ifelse(data_taxa %in% general, "general", data_taxa)
+general <- c("Metazoans", "Ediacarian", "Marine invertebrates", "Amniotes", "Vertebrates")
+data_taxa <- multi.gsub(general, "general", data_taxa)
 
 ## Simulated
-simualted <- c("Simulated", "-", "")
-data_taxa <- ifelse(data_taxa %in% simualted, "simualted", data_taxa)
+simulated <- c("Simulated", "-", "")
+data_taxa <- ifelse(data_taxa %in% simulated, "simulated", data_taxa)
 
 ## Selecting the categories
-data_taxa_categories <- unique(data_taxa)
+data_taxa_categories <- unique(unlist(strsplit(unique(data_taxa), split = " \\+ ")))
 
@@
 
-unique(data_type)
-
 ## continuous synonyms
 continuous <- c("1D Continuous", "DTT pipeline", "Continuous")
-data_type <- ifelse(data_type %in% continuous, "continuous", data_type)
+data_type <- multi.gsub(continuous, "continuous", data_type)
 
@@
 math_transfo <- c("B-spline", "Elliptic Fourier", "Outlines", "Ellipctic Fourier")
-data_type <- ifelse(data_type %in% math_transfo, "math_transfo", data_type)
+data_type <- multi.gsub(math_transfo, "math_transfo", data_type)
 
@@
 other_univar <- c("Body mass", "Metabolic Rate", "Body size", "Glands for chemical communication", "Chemical signals", "Colours")
-data_type <- ifelse(data_type %in% other_univar, "other_univar", data_type)
+data_type <- multi.gsub(other_univar, "other_univar", data_type)
 
 ## Discrete
+data_type <- multi.gsub("Discrete", "discrete", data_type)
+
 data_type <- ifelse(data_type == "Discrete", "discrete", data_type)
@@
 data_type <- ifelse(data_type %in% none, "none", data_type)
-
-## 2d + discrete
-d2_discrete <- c("Discrete + 2D landmarks")
-data_type <- ifelse(data_type %in% d2_discrete, "discrete + 2D landmarks", data_type)
-
-## 2d + continuous
-d2_continu <- c("Continuous + 2D landmarks")
-data_type <- ifelse(data_type %in% d2_continu, "continuous + 2D landmarks", data_type)
-
-## continuous + other
-continu_other <- c("Continuous + Body size")
-data_type <- ifelse(data_type %in% continu_other, "continuous + other_univar", data_type)
-
-## continuous + discrete
-continu_discrete <- c("Continuous + discrete" )
-data_type <- ifelse(data_type %in% continu_discrete, "continuous + discrete", data_type)
 
@@
 
-unique(morphospaces)
-
 ## PCA
-PCA <- c("PCA", "Procrustes PCA", "pPCA", "Regression PCA")
-morphospaces <- ifelse(morphospaces %in% PCA, "PCA", morphospaces)
+PCA <- c("Procrustes PCA", "pPCA", "Regression PCA")
+morphospaces <- multi.gsub(PCA, "PCA", morphospaces)
 
 ## PCO PCoA
-PCO <- c("PCO", "PCoA", "PcoA", "FMD (Foote 1993)")
-morphospaces <- ifelse(morphospaces %in% PCO, "PCO", morphospaces)
+PCO <- c("PCoA", "PcoA", "FMD (Foote 1993)")
+morphospaces <- multi.gsub(PCO, "PCO", morphospaces)
 
 ## MDS NMDS
-MDS <- c("MDS", "NMDS", "nMDS")
-morphospaces <- ifelse(morphospaces %in% MDS, "MDS", morphospaces)
+MDS <- c("NMDS", "nMDS")
+morphospaces <- multi.gsub(MDS, "MDS", morphospaces)
 
 ## RWA
-RWA <- c("RWA", "RWA (Relative Warp Analysis)")
-morphospaces <- ifelse(morphospaces %in% RWA, "RWA", morphospaces)
+RWA <- c("RWA\\ \\(Relative Warp Analysis\\)")
+morphospaces <- multi.gsub(RWA, "RWA", morphospaces)
 
 ## Canonical Analysis
-canonical <- c("Canonical variables", "canonical correlation analysis (CCA)")
-morphospaces <- ifelse(morphospaces %in% canonical, "canonical", morphospaces)
+canonical <- c("Canonical variables", "canonical correlation analysis\\ \\(CCA\\)", "canonical variables", "CVA")
+morphospaces <- multi.gsub(canonical, "canonical", morphospaces)
+
+## Raup spce
+morphospaces <- multi.gsub("Raup-space", "raup_space", morphospaces)
 
 ## Correspondence Analysis
-correspondence <- c("Correspondence analysis", "MCA (multiple correspondence analyses)")
-morphospaces <- ifelse(morphospaces %in% correspondence, "correspondence", morphospaces)
+correspondence <- c("Correspondence analysis", "MCA\\ \\(multiple correspondence analyses\\)")
+morphospaces <- multi.gsub(correspondence, "correspondence", morphospaces)
 
 ## Other
-other <- c("Eigen shape analysis", "Factorial planes","Biplot (two traits only)", "biplot (residuals of two traits)", "Factorial planes","Procrustes plane (2 D axis)", "Procrustes coordinates")
-morphospaces <- ifelse(morphospaces %in% other, "other", morphospaces)
+other <- c("Eigen shape analysis", "Factorial planes", "Biplot\\ \\(two traits only\\)", "biplot\\ \\(residuals of two traits\\)", "Procrustes plane \\(2 D axis\\)", "Procrustes coordinates", "FMD \\(Foote 1993\\)", "Dissimilarity matric")
+morphospaces <- multi.gsub(other, "other", morphospaces)
 
 ## None
-none <- c("DTT pipeline")
+none <- c("DTT pipeline", "-")
 morphospaces <- ifelse(morphospaces %in% none, "none", morphospaces)
 morphospaces <- ifelse(is.na(morphospaces), "none", morphospaces)
-
-## Combos
-morphospaces <- ifelse(morphospaces %in% "PCA + canonical variables", "PCA + canonical", morphospaces)
-morphospaces <- ifelse(morphospaces %in% "Dissimilarity matric + PCoA", "PCO + other", morphospaces)
-morphospaces <- ifelse(morphospaces %in% "RWA (Relative Warp Analysis) + PCO", "PCO + RWA", morphospaces)
-morphospaces <- ifelse(morphospaces %in% "CVA + PCA" , "PCA + canonical", morphospaces)
 
@@
 
+data[,6] <- morphospaces
+```
+
+## Disparity
+
+```{r}
+## Disparity
+disparity <- data[,7]
+
+number_unique_disparities <- length(unique(tolower(disparity)))
+
+## variances
+variance <- c("PCA variance", "sum of variance", "Sum of variance", "product of variances", "sum of variances", "Variance", "mean of variance", "sum of variances", "Total variance", "nth root of variance", "product of variance", "geometric mean of variance", "Procrustes variance", "Product of variance", "root product of variances", "total variance", "nth root of varaince")
+disparity <- multi.gsub(variance, "variance", disparity)
+
+## Sum of range
+range <- c("sum of ranges", "product of range", "product of ranges", "mean of range", "range", "total range", "sum of range", "nth root of range", "Range", "geometric mean of range", "root product of ranges", "mean or range")
+disparity <- multi.gsub(range, "range", disparity)
+
+## Distance
+distances <- c("Gower distance", "Euclidean distance", "mean square euclidean distance", "Procrustes distances \\(geomorph\\)", "other distances", "Procrustes distances", "Average square distance", "mean distance", "mean euclidean distance", "mean Manhattan distance", "Euclidean distance between groups", "average squared distance", "mean euclidean distance", "Procrustes distance", "MNND \\(Mean nearest neighbour distance\\)", "Euclidean Procrustes distances", "sum of Procrustes distances", "range of maximum euclidean distance", "average Procrustes distance", "sum of the squared Procrustes distances")
+disparity <- multi.gsub(distances, "distances", disparity)
+
+## Pairwise distance
+pair_distances <- c("DTT \\(morpho evolution in relation to diversification\\)", "DTT pairwise Euclidean distance", "weighted mean pairwise dissimilarity", "pairwise euclidean distance", "DTT pairwise average squared Euclidean distance", "mean pairwise dissimilarity", "sum of squared pairwise distance between coordinates", " pairwise distance", "average pairwise disparity", "Average pairwise squared distance", "average pairwise dissimilarity", "mean square distance between sp", "distance between species", "pairwise character dissimilarity", "mean pairwise distance", "pairwise distance", "mean squared pairwise difference in body size", "average morphological dissimilarity", "Average pairwise dissimilarity", "Pairwise Mahalanobis distance", "sum of the absolute distance for each variable")
+disparity <- multi.gsub(pair_distances, "pair_distances", disparity)
+
+## Distance from centroid
+centroids <- c("distance from centroids", "centroid", "Procrustes distances from centre", "distance from centroids", "distance from neighbour", "furthest distance", "Average square distance from centroid", "distance from centroid", "centroid size", "distance from centroid", "Group centroid", "mean distance from centroids", "mean distance to centroid", "mean Procrustes distance from centroid", "distances to centroids")
+disparity <- multi.gsub(centroids, "centroids", disparity)
+
+## Other
+other <- c("MDI \\(Morphological disparity index - observed disparity vs median of expected from BM\\)", "Area between observed and median", "Foote's disparity measurement \\(Foote 1993, Neige 2003\\)", "participation ratio", "Body mass", "Regression score", "lineage density", "function of distance to ancestor and character difference")
+disparity <- multi.gsub(other, "other", disparity)
+
+## Character based
+character <- c("Number of polymorphic characters", "number of unique pairwise characters", "unique pairwise combinations", "mean character difference")
+disparity <- multi.gsub(character, "character", disparity)
+
+## PC scores
+ord_scores <- c("PC1\\&2 values", "space coordinates", "PC contrast values", "sum of the diagonal of the variance covariance matrix of the PCA", "Probability of encountering neighbouring data points in 3D")
+disparity <- multi.gsub(ord_scores, "ord_scores", disparity)
+
+## Volume/surface
+volume <- c("Volume", "convex hull surface", "volume", "convex hull volume", "convex hull", "alpha shape volume", "Disparity Box 3D", "Number of cubes of theoretical morphospace", "PC scores", "PC axis scores")
+disparity <- multi.gsub(volume, "volume", disparity)
+
+## Missing
+disparity <- ifelse(disparity == "?", NA, disparity)
+disparity <- ifelse(disparity == "-", NA, disparity)
+
+## Clean the matchings
+disparity <- gsub("ss", "s", gsub("ss", "s", disparity))
+disparity <- gsub(" and ", " \\+ ", disparity)
+disparity <- gsub("DTT pair_distances", "pair_distances", disparity)
+disparity <- gsub("DTT pair_distances", "pair_distances", disparity)
+disparity <- gsub("range variance", "range \\+ variance", disparity)
+disparity <- gsub("distances from centre", "centroids", disparity)
+disparity <- gsub("distances from centroids", "centroids", disparity)
+disparity <- gsub("distances to centroids", "centroids", disparity)
+disparity <- gsub("mean ", "", disparity)
+disparity <- gsub("median ", "", disparity)
+disparity <- gsub("average ", "", disparity)
+disparity <- gsub("other _ character", "other + character", disparity)
+disparity <- gsub("range \\+variance", "range \\+ variance", disparity)
+disparity <- gsub("geometric ", "", disparity)
+
+disparity_categories <- unique(unlist(strsplit(unique(disparity), split = " \\+ ")))
+
+data[,7] <- disparity
+```
+
+## Tests
+
+```{r}
+# ## Disparity
+# tests <- data[,8]
+
+# unique(tests)
+
+# ## ANOVA
+# anova <- c("MANOVA", "PERMANOVA", "ANOVA")
+
+# ## LDA
+# LDA <- c("LDA")
+
+# ## Model test
+# models <- c("Model fitting")
+
+# ## Mean test
+# t_test <- c("t-test")
+
+# ## Correlation
+# correlation <- c("Correlation \\(Fisher's exact test \\+ Spearman's rank\\)")
+
+# ## linear models
+# linear <- c()
+
+# NA 
+# "MANOVA , LDA" 
+# "PERMANOVA" 
+# "Model fitting" 
+# "t-test" 
+# "ANOVA" 
+# "Correlation (Fisher's exact test + Spearman's rank)"
+# "Raup-space" 
+# "?" 
+# "PGLS (PC ~ trait)" 
+# "Mantel test" 
+# "Partial Least Squares" 
+# "Morphological Disparity Index (MDI) statistics" 
+# "Canonical analysis (PCM style)" 
+# "NPMANOVA" 
+# "Procrustes ANOVA" 
+# "phyloANOVA" 
+# "MANOVA" 
+# "lm + t-test" 
+# "Covariance ratio" 
+# "PERMANOVA + CI overlap" 
+# "Cluster dendrogram analysis" 
+# "LDA" 
+# "ANCOVA" 
+# "phyloANOVA + phyloLDA" 
+# "GLS" 
+# "distance PGLS" 
+# "neigbor joining of the centroids" 
+# "comparison to a null expectancy" 
+# "centre of gravity" 
+# "multivariate regression" 
+# "" 
+# "linear regressions" 
+# "confidence interval overlap" 
+# "correlation tests" 
+# "Mann-Whitney" 
+# "Kruskal-Wallis test + Walsh F-test" 
+# "Neares-neighbor test" 
+# "Spearman correlation" 
+# "coefficient of variance" 
+# "GLS multiple regression" 
+# "permutation test" 
+# "Z-test" 
+# "Anova + Kruskal-Wallis" 
+# "Canonical variates analysis" 
+# "F-statistics" 
+# "Wilcoxon" 
+# "multivariate t-test (Hotellings' T^2 test)" 
+# "Nearest neigbhor clustering" 
+# "Ripley's K" 
+# "Hierarchical clustering" 
+# "Canonical variance analysis" 
+# "bootstrap/rarefaction null hypothesis" 
+# "-" 
+# "MANCOVA + MANOVA" 
+# "Branching random walks" 
+# "PICs" 
+# "Kernel analysis" 
+# "Multidimensional Convergence Index" 
+# "Chi-squared" 
+# "Multivariate PGLS" 
+# "Kruskal-Wallis test" 
+# "Breush-Pagan + disparity test" 
+# "Non-linear regressions" 
+# "regression score"
+
+# test <- c("anova", "model fitting", "canonical", "confidence interval")
 ```
@@
 ```{r}
-disparity <- c("sum of variance", "sum of range", "product of variance", "product of range", "distance", "observed") #TG: Area from observed to estimated (DTT)
-test <- c("anova", "model fitting", "canonical", "confidence interval")
 phylo <- c("yes", "no", "mapped")

### 2017-12-07 Analysis/data_cleaning.Rmd
sorting testts

@@
 ```{r}
-# ## Disparity
-# tests <- data[,8]
+## Disparity
+tests <- data[,8]
 
-# unique(tests)
+## ANOVA
+anova <- c("MANOVA", "PERMANOVA", "ANOVA", "NPMANOVA", "Procrustes ANOVA", "phyloANOVA", "PERMANOVA", "ANCOVA", "phyloANOVA", "Kruskal-Wallis test", "coefficient of variance", "permutation test", "Anova", "Ripley's K", "MANCOVA", "Chi-squared")
+tests <- multi.gsub(anova, "anova", tests)
 
-# ## ANOVA
-# anova <- c("MANOVA", "PERMANOVA", "ANOVA")
+## LDA
+LDA <- c("LDA", "phyloLDA")
+tests <- multi.gsub(LDA, "LDA", tests)
 
-# ## LDA
-# LDA <- c("LDA")
+## Model test
+models <- c("Model fitting", "comparison to a null expectancy", "bootstrap/rarefaction null hypothesis")
+tests <- multi.gsub(models, "models", tests)
 
-# ## Model test
-# models <- c("Model fitting")
+## Mean test
+t_test <- c("t-test", "Mann-Whitney", "Walsh F-test", "Z-test", "F-statistics", "Wilcoxon", "multivariate t-test \\(Hotellings' T\\^2 test\\)")
+tests <- multi.gsub(t_test, "t_test", tests)
 
-# ## Mean test
-# t_test <- c("t-test")
+## Correlation
+correlation <- c("Correlation \\(Fisher's exact test \\+ Spearman's rank\\)", "Mantel test", "correlation tests", "Spearman correlation")
+tests <- multi.gsub(correlation, "correlation", tests)
 
-# ## Correlation
-# correlation <- c("Correlation \\(Fisher's exact test \\+ Spearman's rank\\)")
+## regressions
+linear <- c("PGLS \\(PC \\~ trait\\)", "Partial Least Squares", "Canonical analysis \\(PCM style\\)", "lm", "Covariance ratio", "GLS", "distance PGLS", "multivariate regression", "linear regressions", "GLS multiple regression", "PICs", "Multivariate PGLS", "Non-linear regressions", "regression score", "Breush-Pagan")
+tests <- multi.gsub(linear, "linear", tests)
 
-# ## linear models
-# linear <- c()
+## non stat
+non_stat <- c("CI overlap", "confidence interval overlap")
+tests <- multi.gsub(non_stat, "non_stat", tests)
 
-# NA 
-# "MANOVA , LDA" 
-# "PERMANOVA" 
-# "Model fitting" 
-# "t-test" 
-# "ANOVA" 
-# "Correlation (Fisher's exact test + Spearman's rank)"
-# "Raup-space" 
-# "?" 
-# "PGLS (PC ~ trait)" 
-# "Mantel test" 
-# "Partial Least Squares" 
-# "Morphological Disparity Index (MDI) statistics" 
-# "Canonical analysis (PCM style)" 
-# "NPMANOVA" 
-# "Procrustes ANOVA" 
-# "phyloANOVA" 
-# "MANOVA" 
-# "lm + t-test" 
-# "Covariance ratio" 
-# "PERMANOVA + CI overlap" 
-# "Cluster dendrogram analysis" 
-# "LDA" 
-# "ANCOVA" 
-# "phyloANOVA + phyloLDA" 
-# "GLS" 
-# "distance PGLS" 
-# "neigbor joining of the centroids" 
-# "comparison to a null expectancy" 
-# "centre of gravity" 
-# "multivariate regression" 
-# "" 
-# "linear regressions" 
-# "confidence interval overlap" 
-# "correlation tests" 
-# "Mann-Whitney" 
-# "Kruskal-Wallis test + Walsh F-test" 
-# "Neares-neighbor test" 
-# "Spearman correlation" 
-# "coefficient of variance" 
-# "GLS multiple regression" 
-# "permutation test" 
-# "Z-test" 
-# "Anova + Kruskal-Wallis" 
-# "Canonical variates analysis" 
-# "F-statistics" 
-# "Wilcoxon" 
-# "multivariate t-test (Hotellings' T^2 test)" 
-# "Nearest neigbhor clustering" 
-# "Ripley's K" 
-# "Hierarchical clustering" 
-# "Canonical variance analysis" 
-# "bootstrap/rarefaction null hypothesis" 
-# "-" 
-# "MANCOVA + MANOVA" 
-# "Branching random walks" 
-# "PICs" 
-# "Kernel analysis" 
-# "Multidimensional Convergence Index" 
-# "Chi-squared" 
-# "Multivariate PGLS" 
-# "Kruskal-Wallis test" 
-# "Breush-Pagan + disparity test" 
-# "Non-linear regressions" 
-# "regression score"
+## other
+other <- c("Morphological Disparity Index \\(MDI\\) statistics", "centre of gravity", "Canonical variates analysis", "Canonical variance analysis", "Multidimensional Convergence Index", "Kernel analysis", "disparity test")
+tests <- multi.gsub(other, "other", tests)
 
-# test <- c("anova", "model fitting", "canonical", "confidence interval")
+## clustering
+clustering <- c("Neares-neighbor test", "Cluster dendrogram analysis", "neigbor joining of the centroids", "Nearest neigbhor clustering", "Hierarchical clustering", "Branching random walks")
+tests <- multi.gsub(clustering, "clustering", tests)
+
+tests <- ifelse(tests == "", NA, tests)
+tests <- ifelse(tests == "?", NA, tests)
+tests <- ifelse(tests == "-", NA, tests)
+tests <- ifelse(tests == "Raup-space", NA, tests)
+
+## Clean the matchings
+tests <- gsub("distance Plinear", "linear", tests)
+tests <- gsub("multivariate t_test \\(Hotellings' T\\^2 test\\)", "t_test", tests)
+tests <- gsub("anova , LDA", "anova + LDA", tests)
+tests <- gsub("Kruskal-Wallis", "anova", tests)
+
+tests_categories <- unique(unlist(strsplit(unique(tests), split = " \\+ ")))
+
+data[,8] <- tests
 ```

### 2017-12-08 Analysis/data_cleaning.Rmd
Data wrangling finished

@@
 data <- data_raw[apply(data_raw[, 3:9], 1, function(X) ifelse(all(X == "-"), FALSE, TRUE)), ]
-
-## Removing rogue NA rows
-data <- data[-231,]
 ```
@@
 ## clustering
-clustering <- c("Neares-neighbor test", "Cluster dendrogram analysis", "neigbor joining of the centroids", "Nearest neigbhor clustering", "Hierarchical clustering", "Branching random walks")
+clustering <- c("Nearest-neighbour test", "Cluster dendrogram analysis", "neighbour joining of the centroids", "Nearest-neighbour clustering", "Hierarchical clustering", "Branching random walks")
 tests <- multi.gsub(clustering, "clustering", tests)
@@
 tests <- gsub("multivariate t_test \\(Hotellings' T\\^2 test\\)", "t_test", tests)
-tests <- gsub("anova , LDA", "anova + LDA", tests)
 tests <- gsub("Kruskal-Wallis", "anova", tests)
@@
 
+## Phylogeny
+
 ```{r}
-phylo <- c("yes", "no", "mapped")
+phylogeny <- data[,9]
 
-factors <- list("data type" = data_type, "full organism" = full_organism, "morphospace" = morphospace, "disparity" = disparity, "test" = test, "phylo correction" = phylo)
+phylogeny <- multi.gsub("Yes \\(pPC\\)", "pPC", phylogeny)
+phylogeny <- ifelse(phylogeny == "", NA, phylogeny)
+
+phylogeny_categories <- unique(unlist(strsplit(unique(phylogeny), split = " \\+ ")))
+
+data[,9] <- phylogeny
 ```
+
+## Wrapping up the data set
+
+```{r}
+## Cleaned data
+data <- data[,-c(1,2)]
+cleaned_data <- list("data" = data)
+cleaned_data$factors <- list(data_taxa = data_taxa_categories, data_type = data_type_categories, full_organism = full_organism_categories, morphospaces = morphospaces_categories, disparity = disparity_categories, tests = tests_categories, phylogeny = phylogeny_categories)
+
+## Save the file
+saveRDS(cleaned_data, file = paste0("../Data/Processed/cleaned_data.Rda"))
+
+```

### 2017-12-08 Analysis/data_cleaning.Rmd
cleaned data

@@
 ## General
-general <- c("Metazoans", "Ediacarian", "Marine invertebrates", "Amniotes", "Vertebrates")
+general <- c("Metazoans", "Ediacarian", "Marine invertebrates", "Amniotes", "Vertebrates", "Marine ingeneral")
 data_taxa <- multi.gsub(general, "general", data_taxa)
@@
 ## Pairwise distance
-pair_distances <- c("DTT \\(morpho evolution in relation to diversification\\)", "DTT pairwise Euclidean distance", "weighted mean pairwise dissimilarity", "pairwise euclidean distance", "DTT pairwise average squared Euclidean distance", "mean pairwise dissimilarity", "sum of squared pairwise distance between coordinates", " pairwise distance", "average pairwise disparity", "Average pairwise squared distance", "average pairwise dissimilarity", "mean square distance between sp", "distance between species", "pairwise character dissimilarity", "mean pairwise distance", "pairwise distance", "mean squared pairwise difference in body size", "average morphological dissimilarity", "Average pairwise dissimilarity", "Pairwise Mahalanobis distance", "sum of the absolute distance for each variable")
+pair_distances <- c("DTT \\(morpho evolution in relation to diversification\\)", "DTT pairwise Euclidean distance", "weighted mean pairwise dissimilarity", "pairwise euclidean distance", "DTT pairwise average squared Euclidean distance", "mean pairwise dissimilarity", "sum of squared pairwise distance between coordinates", " pairwise distance", "average pairwise disparity", "Average pairwise squared distance", "average pairwise dissimilarity", "mean square distance between sp", "distance between species", "pairwise character dissimilarity", "mean pairwise distance", "pairwise distance", "mean squared pairwise difference in body size", "average morphological dissimilarity", "Average pairwise dissimilarity", "Pairwise Mahalanobis distance", "sum of the absolute distance for each variable", "DTT pairwise squared distances")
 disparity <- multi.gsub(pair_distances, "pair_distances", disparity)
@@
 ## regressions
-linear <- c("PGLS \\(PC \\~ trait\\)", "Partial Least Squares", "Canonical analysis \\(PCM style\\)", "lm", "Covariance ratio", "GLS", "distance PGLS", "multivariate regression", "linear regressions", "GLS multiple regression", "PICs", "Multivariate PGLS", "Non-linear regressions", "regression score", "Breush-Pagan")
-tests <- multi.gsub(linear, "linear", tests)
+regressions <- c("PGLS \\(PC \\~ trait\\)", "Partial Least Squares", "Canonical analysis \\(PCM style\\)", "lm", "Covariance ratio", "GLS", "distance PGLS", "multivariate regression", "linear regressions", "GLS multiple regression", "PICs", "Multivariate PGLS", "Non-linear regressions", "regression score", "Breush-Pagan")
+tests <- multi.gsub(regressions, "regressions", tests)
 
@@
 tests <- gsub("Kruskal-Wallis", "anova", tests)
+tests <- gsub("distance Pregressions", "regressions", tests)
 
@@
 cleaned_data$factors <- list(data_taxa = data_taxa_categories, data_type = data_type_categories, full_organism = full_organism_categories, morphospaces = morphospaces_categories, disparity = disparity_categories, tests = tests_categories, phylogeny = phylogeny_categories)
+cleaned_data$unique_disparity <- number_unique_disparities
 

### 2017-12-08 Functions/fuzzy.sort.data.R
fuzzy sort with NAs

@@ -23,12 +23,18 @@ fuzzy.sort.data <- function(data, what) {
     ## Loop through the fuzzy matching
     for(element in 1:length(what_l)) {
         ## Check fuzzy match
-        fuzzy_match <- grep(what_l[element], data_l)
+        if(is.na(what_l[element])) {
+            fuzzy_match <- which(is.na(data_l))
+            element_name <- "NA"
+        } else {
+            fuzzy_match <- grep(what_l[element], data_l)
+            element_name <- what[element]
+        }
         length_match <- length(fuzzy_match)
         
         if(length_match > 0) {
             ## If match, record the matches
-            matches <- data.frame("data" = fuzzy_match, "what" = rep(what[element], length_match))
+            matches <- data.frame("data" = fuzzy_match, "what" = rep(element_name, length_match))
             total_matches <- rbind(total_matches, matches)
         }
     }

### 2017-12-08 Analysis/disparity_methods.Rmd
PCO plots

@@
 source("../Functions/fuzzy.sort.data.R")
+source("../Functions/plot.data.R")
 ```
@@
 ## Reading the data
-data_raw <- read.csv("../Data/disparity_methods.csv", stringsAsFactors = FALSE)
+cleaned_data <- readRDS("../Data/Processed/cleaned_data.Rda")
 search_results <- read.csv("../Data/GoogleSearch_occurences.csv", stringsAsFactors = FALSE)
 
-## Keeping only the studies with data
-data <- data_raw[apply(data_raw[, 3:9], 1, function(X) ifelse(all(X == "-"), FALSE, TRUE)), ]
+## Extracting the cleaned data
+factors <- cleaned_data$factors
+data <- cleaned_data$data
 ```
@@
 ```{r, fig.width = 8, fig.height = 8}
+## Selecting the right data
 disparity_fuzzy <- search_results[-c(1:2, 60:65),3]
@@
 ## Number of papers
-# pdf("../Paper/Figures/GoogleScholarOccurences.pdf", width = 8, height = 8)
+# pdf("../Writing/Figures/GoogleScholarOccurences.pdf", width = 8, height = 8)
 op <- par(bty = "n")
-plot(seq(from = 1960, to = 2016), rev(disparity_fuzzy)/search_results[1,3], type = "l", xlab = "", ylab = "Relative number of papers per year", ylim = c(0, 0.1))
+plot(seq(from = 1960, to = 2016), rev(disparity_fuzzy)/search_results[1,3], type = "l", xlab = "", ylab = "Number of papers per year", ylim = c(0, 0.1), yaxt = "n")
 lines(seq(from = 1960, to = 2016), rev(disparity_exact)/search_results[1,2], type = "l", col = "grey")
+## Adding the axis
+axis(2, at = seq(from = 0, to = 0.08, by = 0.02), labels = round(seq(from = min(disparity_fuzzy), to = max(disparity_fuzzy), length.out = 5)) )
+axis(4, at = seq(from = 0, to = 0.1, by = 0.02), labels = round(seq(from = min(disparity_exact), to = max(disparity_exact), length.out = 6)), col = "grey", las = 2)
 par(op)
@@
 
-## Sorting the variables
+## Variable distribution
 
-The variables can be sorted as follow:
+What is the proportion of use for each method?
+
+```{r, fig.width = 18, fig.height = 6}
+## Plotting the results
+pdf("../Writing/Figures/MethodsProportions.pdf", width = 18, height = 6)
+op <- par(mfrow = c(2, 3), bty = "n")
+for(factor in 2:length(factors)) {
+    plot.data(fuzzy.sort.data(data[,factor], factors[[factor]]), proportion = TRUE, ylim = c(0, 100), main = names(factors)[[factor]])
+}
+par(op)
+dev.off()
+```
+
+Methods seem to full in three main categories:
+
+  * The `Claddis` approach: Discrete data for the full organism ordinated with a PCO with a variance based disparity metric and no phylogenetic correction.
+  * The `geomorph` approach: landmark data for part of the organism ordinated with a PCA with a distance based disparity metric and no phylogenetic correction.
+  * The `dtt` approach: continuous data for the full organism, not ordinated, with a pairwise distance based disparity metric and phylogenetic correction.
 
 ```{r}
+## Methods types table (empty)
+methods_types <- matrix(NA, nrow = 3, ncol = 6, dimnames = list(c("Claddis", "geomorph", "dtt"), c(names(factors)[-1])))
 
-## Categories definition
-data_type <- c("3D landmarks", "2D landmarks", "continuous", "discrete", "body", "metabolic")
-full_organism <- c("yes", "no")
-morphospace <- c("PCA", "MDS", "PCO", "PCOA", "Raup")
-disparity <- c("sum of variance", "sum of range", "product of variance", "product of range", "distance", "observed") #TG: Area from observed to estimated (DTT)
-test <- c("anova", "model fitting", "canonical", "confidence interval")
-phylo <- c("yes", "no", "mapped")
+## Filling the table
+methods_types[,1] <- c("discrete", "landmarks", "continuous")
+methods_types[,2] <- c("yes", "no", "yes")
+methods_types[,3] <- c("PCO", "PCA", "\"pairwise matrix\"")
+methods_types[,4] <- c("variance", "distance", "pairwise distance")
+methods_types[,5] <- c("variable", "anova", "models")
+methods_types[,6] <- c("no", "sometimes", "yes")
 
-factors <- list("data type" = data_type, "full organism" = full_organism, "morphospace" = morphospace, "disparity" = disparity, "test" = test, "phylo correction" = phylo)
+knitr::kable(methods_types)
 ```
 
-```{r, fig.width = 12, fig.height = 12}
-## Plotting the results
-par(mfrow = c(2, 3))
-for(factor in 1:length(factors)) {
-    plot.data(fuzzy.sort.data(data[,3+factor], factors[[factor]]), proportion = TRUE, ylim = c(0, 100), main = names(factors)[[factor]])    
-}
+## A disparity-space?
+
+Multivariate data? Let's analysis it using disparity!
+
+But first, a bit of cleaning
+
+```{r}
+## Removing the taxa column
+multi_data <- apply(data[,-1], 2, as.character)
+
+## Replace NAs by "-"
+multi_data <- ifelse(is.na(multi_data), "-", multi_data)
+
+## Selecting only the first variable when multiples
+multi_data <- gsub(" \\+ .*", "", multi_data)
+
+## Transform the data as numeric
+multi_data <- as.list(as.data.frame(multi_data))
+multi_data <- lapply(multi_data, as.numeric)
+multi_data <- matrix(unlist(multi_data), ncol = ncol(data[,-1]), byrow = FALSE)
+
+## Make a pairwise matrix
+pairwise_data <- vegan::vegdist(multi_data, method = "bray")
+
+## Ordinating the distance matrix
+ordinated_data <- stats::cmdscale(pairwise_data, k = sum(unlist(lapply(factors, length))[-1]) - 2)
+#TG: Axis are selected as the number of factors - 2
 ```
 
-```{r}
+```{r, fig.width = 8, fig.height = 8}
+## Get the variance per axis
+scree_plot <- dispRity::variances(ordinated_data)
+percentages_scree <- round(scree_plot/sum(scree_plot)*100, digit = 2)
+
+## Variance per dimensions
+barplot(scree_plot, ylab = "Variance per dimension", xlab = "Dimensions")
+```
+
+```{r, fig.width = 8, fig.height = 8}
+
+factor <- 5
+# length(unique(multi_data[,1]))
+# length(factors[[2]])
+
+col_palette <- rainbow(length(unique(data[, factor])))
+colours <- col_palette[as.factor(multi_data[, factor-1])]
+
+plot(ordinated_data[,1], ordinated_data[,2], xlab = paste0("PC1 (",percentages_scree[1], "%)"), ylab = paste0("PC2 (",percentages_scree[2], "%)"), pch = 19, col = colours, main = paste0(names(factors)[factor]))
+
+legend("bottomright", legend = sort(factors[[factor]]), col = col_palette, pch = 19, cex = 0.5)
+
+```
+
+<!-- ```{r}
 ## Printing the tables
 factor <- 1
-table.data(fuzzy.sort.data(data[,3+factor], factors[[factor]]), table = "tex")
+table.data(fuzzy.sort.data(data[,factor], factors[[factor]]), table = "tex")
 
-```
+``` -->
 
@@
 
-# Occurrences of morphological disparity papers
+## Testing the three categories hypothesis
+
+Setting the data for the three categories hypothesis:
+
+  * The `Claddis` approach: Discrete data for the full organism ordinated with a PCO with a variance based disparity metric and no phylogenetic correction.
+  * The `geomorph` approach: landmark data for part of the organism ordinated with a PCA with a distance based disparity metric and no phylogenetic correction.
+  * The `dtt` approach: continuous data for the full organism, not ordinated, with a pairwise distance based disparity metric and phylogenetic correction.
 
 ```{r}
-disparity_fuzzy <- search_results[-c(1:2, 60:65),3]
-disparity_exact <- search_results[-c(1:2, 60:65),2]
+cleaned_data <- readRDS("../Data/Processed/cleaned_data.Rda")
 
-## Number of papers
-pdf("../Writing/Figures/GoogleScholarOccurences.pdf", width = 8, height = 8)
-op <- par(bty = "n")
-plot(seq(from = 1960, to = 2016), rev(disparity_fuzzy)/search_results[1,3], type = "l", xlab = "", ylab = "Number of papers per year", ylim = c(0, 0.1), yaxt = "n")
-lines(seq(from = 1960, to = 2016), rev(disparity_exact)/search_results[1,2], type = "l", col = "grey")
-## Adding the axis
-axis(2, at = seq(from = 0, to = 0.08, by = 0.02), labels = round(seq(from = min(disparity_fuzzy), to = max(disparity_fuzzy), length.out = 5)) )
-axis(4, at = seq(from = 0, to = 0.1, by = 0.02), labels = round(seq(from = min(disparity_exact), to = max(disparity_exact), length.out = 6)), col = "grey", las = 2)
+data <- cleaned_data$data
 
-par(op)
-dev.off()
 ```
@@
 plot(add, type = "line", xlab = "Number of Google Scholar matches", ylab = "Number of new entries")
+```
 
+## Groups studied
+
+Which groups are studied using disparity analysis?
+
+```{r}
+plot.data(fuzzy.sort.data(data[,1], factors[[1]]), proportion = TRUE, ylim = c(0, 100), main = names(factors)[[1]])
 ```
+
+It's pretty much even with no big majority!
 

### 2017-12-10 Analysis/data_cleaning.Rmd
Updated cleaning data

@@
 disparity <- gsub("geometric ", "", disparity)
+disparity <- gsub("ranges", "range", disparity)
+disparity <- gsub("DTT pairwise squared distances", "pair_distances", disparity)
 
@@
 tests <- gsub("distance Pregressions", "regressions", tests)
+tests <- gsub("Non-regressions", "regressions", tests)
 

### 2017-12-10 Analysis/disparity_methods.Rmd
Updated method proportions

@@
 ```{r}
+library(dispRity)
 source("../Functions/fuzzy.sort.data.R")
@@
 
-```{r, fig.width = 18, fig.height = 6}
+```{r, fig.width = 12, fig.height = 12}
 ## Plotting the results
-pdf("../Writing/Figures/MethodsProportions.pdf", width = 18, height = 6)
-op <- par(mfrow = c(2, 3), bty = "n")
-for(factor in 2:length(factors)) {
+pdf("../Writing/Figures/MethodsProportions.pdf", width = 12, height = 12)
+op <- par(mfrow = c(2, 2), bty = "n")
+for(factor in c(2,4,5,6)) {
     plot.data(fuzzy.sort.data(data[,factor], factors[[factor]]), proportion = TRUE, ylim = c(0, 100), main = names(factors)[[factor]])
@@
 ```{r}
-## Removing the taxa column
-multi_data <- apply(data[,-1], 2, as.character)
+## Removed columns
+removed_factors <- c(1)
+
+## Removing factors from the data
+multi_data <- apply(data[,-removed_factors], 2, as.character)
 
@@
 multi_data <- lapply(multi_data, as.numeric)
-multi_data <- matrix(unlist(multi_data), ncol = ncol(data[,-1]), byrow = FALSE)
+multi_data <- matrix(unlist(multi_data), ncol = ncol(data[,-removed_factors]), byrow = FALSE)
 
@@
 ## Ordinating the distance matrix
-ordinated_data <- stats::cmdscale(pairwise_data, k = sum(unlist(lapply(factors, length))[-1]) - 2)
+ordinated_data <- stats::cmdscale(pairwise_data, k = sum(unlist(lapply(factors[-removed_factors], length))) - 2)
+rownames(ordinated_data) <- seq(1:nrow(data))
 #TG: Axis are selected as the number of factors - 2
@@
 
-```{r, fig.width = 8, fig.height = 8}
+```{r, fig.width = 8, fig.height = 8, eval = FALSE, echo = FALSE}
 
@@
 
-data <- cleaned_data$data
+## Get the group variables
+Claddis_group_var <- list(factors$data_type[3], factors$morphospaces[2], factors$disparity[c(4,5)])
+geomorph_group_var <- list(factors$data_type[c(1,2)], factors$morphospaces[1], factors$disparity[c(2,9)])
+dtt_group_var <- list(factors$data_type[4], factors$morphospaces[4], factors$disparity[6])
+
+select.group <- function(data, group) {
+    ## Selecting the group variables
+    multi.grep <- function(x, pattern, ignore.case = TRUE, ...) {
+        ## Finding if the patten matches in x
+        is_grep <- unlist(sapply(pattern, grep, x, ignore.case = ignore.case, simplify = FALSE, USE.NAMES = FALSE))
+        ## Return true/false
+        return(ifelse(length(is_grep) > 0, TRUE, FALSE))
+    }
+
+    data_type <- sapply(data[,2], multi.grep, pattern = group[[1]], USE.NAMES = FALSE)
+    morphospace <- sapply(data[,4], multi.grep, pattern = group[[2]], USE.NAMES = FALSE)
+    disparity <- sapply(data[,5], multi.grep, pattern = group[[3]], USE.NAMES = FALSE)
+
+    all_match <- data_type & morphospace & disparity
+    two_match_1 <-  morphospace & disparity
+    two_match_2 <-  data_type & disparity
+    two_match_3 <-  data_type & morphospace
+
+    return(list("match3" = all_match, "match2" = list(two_match_1, two_match_2, two_match_3), "match1" = list(data_type, morphospace, disparity)))
+}
+
+## Select the groups
+Claddis_groups <- select.group(data, Claddis_group_var)
+geomorph_groups <- select.group(data, geomorph_group_var)
+dtt_groups <- select.group(data, dtt_group_var)
+
+## Select the full groups
+Claddis_group <- Claddis_groups$match3
+geomorph_group <- geomorph_groups$match3
+dtt_group <- dtt_groups$match3
+no_group <- !(Claddis_group | geomorph_group | dtt_group)
+
+## Create the group variable
+groups <- rep(NA, nrow(data))
+groups <- ifelse(Claddis_group, "Claddis", groups)
+groups <- ifelse(geomorph_group, "geomorph", groups)
+groups <- ifelse(dtt_group, "dtt", groups)
+groups <- ifelse(no_group, "none", groups)
+
+## Get the groups IDs
+Claddis_group_id <- which(Claddis_group)
+geomorph_group_id <- which(geomorph_group)
+dtt_group_id <- which(dtt_group)
+no_group_id <- which(no_group)
+
+```
+
+## Plotting the group's
+
+```{r, fig.width = 8, fig.height = 8}
+col_palette <- c("red", "blue", "green3", "grey")
+
+plot(ordinated_data[,1], ordinated_data[,2], xlab = paste0("PC1 (",percentages_scree[1], "%)"), ylab = paste0("PC2 (",percentages_scree[2], "%)"), pch = 19, col = col_palette[as.factor(groups)])
+
+legend("bottomright", legend = levels(as.factor(groups)), col = col_palette, pch = 19, cex = 0.5)
+```
+
+## Testing the effect of groups
+
+```{r}
+## Creating the custom groups list
+groups_list <- list("Claddis" = Claddis_group_id, "geomorph" = geomorph_group_id, "dtt" = dtt_group_id)
+
+## Creating the groups for the pairwise matrix and the ordinated one
+pair_matrix <- as.matrix(pairwise_data)
+rownames(pair_matrix) <- seq(1:nrow(data))
+groups_pair <- custom.subsets(pair_matrix, group = groups_list)
+groups_ordi <- custom.subsets(ordinated_data, group = groups_list)
+
+## Effect of groups
+vegan::adonis2(pairwise_data ~ group, data = data.frame(matrix(groups, ncol = 1, dimnames = list(seq(1:nrow(data)), c("group")))))
+
+## Effect of groups (individual)
+groups_data <- data.frame(cbind(as.factor(Claddis_group), as.factor(geomorph_group), as.factor(dtt_group), as.factor(no_group)))
+rownames(groups_data) <- seq(1:nrow(data))
+colnames(groups_data) <- c("Claddis", "geomorph", "dtt", "nope")
+vegan::adonis2(pairwise_data ~ Claddis + geomorph + dtt + nope, data = groups_data, permutations = 100)
 

### 2017-12-11 Analysis/disparity_methods.Rmd
typo

@@
 
-I collected the data from the 500 first Google Scholar results for the key words "morphological disparity" per order of appearance (accessed 1st of Noveember 2017).
+I collected the data from the 500 first Google Scholar results for the key words "morphological disparity" per order of appearance (accessed 1st of November 2017).
 This resulted in 232 relevant papers for which I collected:

### 2017-12-11 Writing/disparity_review.tex
Added methods to review

--- 2017-09-12 Paper/disparity_review.tex
@@
 
-Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.
-One solution to this problem is to analyse data in a multivariate framework.
-Multivariate analyses aim to capture and incorporate this multidimensional complexity, while providing outputs that are interpretable in a physical world of only three dimensions.
-Classical multidimensional species features that have been analysed through multivariate data ranges from their morphology \citep{raup1966geometric}, to their functional traits \citep{diaz2016global}.
-However, such analysis are not limited at the species level and can also be applied to ecosystems \citep{DonohueDim}.
+%§1: disparity is observed everytwhere
+Disparity is an old concept blabalbal coming from the observation that some groups of species are more similar/dissimilar than other.
 
-Key multidimensional features of species that have important roles in ecology and evolution include morphology, functional traits. etc. etc.
-Many of these can be represented as matrices, these can be used in our package.
-In the interests of clarity/brievety, we here focus on just one kind of data: morphological diversity.
+It can be also applied between groups as a concept of ecological niches: a group occupying many niches will have a stronger disparity than a groups with less niches.
 
-Furthermore, we will also make a clear distinction between the multidimensional space which is a specific mathematical object and the disparity, which is a metric describing or summarising one or more aspects of this space.
+It is decoupled from diversity.
 
-Both can be defined in various ways depending on the authors, for example, disparity is defined as the weighted mean pairwise dissimilarity in \cite{Close2015} or the ellipsoid volume in \cite{DonohueDim} and the multidimensional space is defined as the morphospace in \cite{raup1966geometric} and the morpho-functional space in \cite{diaz2016global}.
+\bigskip
+%§2: "phylogeny" of disparity
 
-The multidimensional space can be defined in many ways and arise from many mathematical transformations of the data such as the pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
-Similarly, disparity metrics \citep[or indices;][]{Hopkins2017} can defined in many ways \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015,diaz2016global}.
-Finally, difference between disparity metrics can also be measured in many ways: using NPMANOVA \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simple confidence interval overlap \citep[e.g.][]{halliday2016eutherian}.
+In palaeobiology, morphological disparity (the (dis)similarities between the morphologies of groups of organisms) concept as a tool for macroevolutionary studies stems to several seminal papers from the 90s: \cite{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution}.
 
-This variety of definitions and analysis have been developed in an equal variety of softwares such as \texttt{GINGKO} in javascript \citep{bouxin2005ginkgo,de2007ginkgo} or \texttt{geomorph} \citep{adams2013geomorph,adams2017geometric}, \texttt{Claddis} \citep{Claddis}, or \texttt{vegan} \citep{oksanen2007vegan} in R \citep{R}.
-This results in the need to learn different languages (or at least - when restricted to R - different packages with different standards) as well as making analysis sometimes idiosyncratic and often complex to repeat since they are based on a particular feature from a particular software.
-For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function.
-Unfortunately, however, the multidimensional space can only be defined as the ordination of the procrustes transform of geometric morphometric landmarks, the disparity can only be defined as the Procrustes variance and the difference between groups can only be measured through permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+This concept was then used in geometric morphometric and expanded to macroevolutionary studies in general (Harmon's dtt).
 
-Disparity is an old concept blabalbal
+\cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.
 
-\section{What \textit{is} disparity}
+\bigskip
 
-From a semantic view: morphospace or aspect of the morphospace?
+%§3: a very popular concept
 
-From a biological view: what does an increase or a decrease in disparity represent?
+Disparity is really popular (Fig. \ref{Fig:GoogleOccurences}):
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/GoogleScholarOccurences.pdf} 
+\caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year.
+In black, the match is in the paper and in grey, in the title.
+We collected the number of matches per year from 1960 to 2016 in Google Scholar for the terms "morphological disparity" both in the text (fuzzy matching) or in the title (exact matching).
+The data was collected on the 1st of November 2017.}
+\label{Fig:GoogleOccurences}
+\end{figure}
 
 \section{Why do we need disparity}
+Which macroevolutionary questions are tackled through disparity analysis?
 
@@
 
+Disparity can be 
+ -From a semantic view: morphospace or aspect of the morphospace?
+ -From a biological view: what does an increase or a decrease in disparity represent?
+
+But in the end, what \textit{is} disparity
+
+\section{But what \textit{is} disparity?}
+
+Disparity can describe either the metric (disparity index \citep{Hopkins2017} or disparity metric \citep{Wills2001}) or the whole pipeline \citep{Claddis,zelditch2012geometric}
+
+% §Methods for data:
+\subsection{Data collection methodology}
+To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
+For the 230 relevant papers among the 500 matches, we collected the following methodological data:
+
+\begin{itemize}
+    \item What was the focal biological group?
+    \item What kind of data was measured (e.g. landmarks, discrete data, etc.)?
+    \item Was data collected on the full organism or not?
+    \item How was the morphospace explicitly defined (e.g. PCA, PCO, MDS, etc.)?
+    \item How was the disparity metric(s) explicitly defined?
+    \item Which statistical test was applied to test the disparity related hypothesis?
+    \item Was phylogeny taken into account or not?
+\end{itemize}
+
+We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g.
+a disparity metric was measured but not described anywhere in the paper).
+
+The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).
+
+To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in the following categories:
+
+[DESCRIBE THE CATEGORIES HERE (from \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd})]
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
+\caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
+\label{Fig:MethodsProportions}
+\end{figure}
+
+[ADD Biological group, full organism and phylogeny in the supplementary]
+
+\subsection{Disparity data}
+Describing the many data inputs (from collected data to morphospace):
+
+\subsection{Disparity metrics}
+Describing the many metrics (what is actually measured):
+
+\subsection{Disparity hypothesis}
+Describing the many outputs (what and how is it tested):
+
+\subsection{The three main disparity analysis}
+From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.
+
+\begin{itemize}
+    \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data for the full organisms to build a PCO from the organism's pairwise distance as a morphospace.
+Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g.
+the sum of variance or/and the sum of ranges).
+Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.
+
+    \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA).
+Disparity is often measured as a distance metric (e.g.
+the distance between the species and a point in the morphospace such as its centroid).
+Hypothesis are then tested using ANOVA type tests with usually no phylogenetic correction (although phylogeny is sometimes used to correct the morphospace).
+
+    \item \textbf{The ``\texttt{dtt}'' approach:} continuous data for the full organism, not ordinated, with a pairwise distance based disparity metric and phylogenetic correction.
+
+\end{itemize}
+
+Of course some studies use a combination of these three methods or none of them at all!
+
+Also, among each category XX\% of studies use multiple approaches.
+
+[ADD Biological group, full organism and phylogeny in the supplementary]
+
 \section{What is missing}
 
-Methods?
-Data?
+How to compare disparity between groups?
+
+How to compare disparity between methods?
+
+Hot to compare disparity between disciplines?
+
+%TG: we can find solutions from other fields
+It bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.
 
@@
 
-Yay!
+A quick guideline for good disparity analysis:
 
-% \section{Data availability and reproducibility}
-% Data are available on Dryad or Figshare.
-% Code for reproducing the analyses is available on GitHub (\url{github.com/TGuillerme/SpatioTemporal_Disparity}).
+Maybe we need something like in Parham et al 2012 (Best Practices for Justifying Fossil Calibrations): an easy an identifiable description of the pipeline containing: 1) the type of data, 2) the morphospace and 3) the metric?
 
 \section{Acknowledgments}
-Royal Society..
+The Royal Society.
 I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
@@
 \end{document}
+
+% The multidimensional space can be defined in many ways and arise from many mathematical transformations of the data such as the pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
+% Similarly, disparity metrics \citep[or indices;][]{Hopkins2017} can defined in many ways \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015,diaz2016global}.
+% Finally, difference between disparity metrics can also be measured in many ways: using NPMANOVA \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simple confidence interval overlap \citep[e.g.][]{halliday2016eutherian}.
+
+% This variety of definitions and analysis have been developed in an equal variety of softwares such as \texttt{GINGKO} in javascript \citep{bouxin2005ginkgo,de2007ginkgo} or \texttt{geomorph} \citep{adams2013geomorph,adams2017geometric}, \texttt{Claddis} \citep{Claddis}, or \texttt{vegan} \citep{oksanen2007vegan} in R \citep{R}.
+% This results in the need to learn different languages (or at least - when restricted to R - different packages with different standards) as well as making analysis sometimes idiosyncratic and often complex to repeat since they are based on a particular feature from a particular software.
+% For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function.
+% Unfortunately, however, the multidimensional space can only be defined as the ordination of the procrustes transform of geometric morphometric landmarks, the disparity can only be defined as the Procrustes variance and the difference between groups can only be measured through permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+
+% In all these analysis, each set of multivariate traits forms a multidimensional space.
+% \hl{This space is represented as a matrix where rows are regarded as samples or observations (e.g.
+a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g.
+an embedding, scaling, ordination, etc.
+of the variables).}
+% \hl{These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (}
+% \citealt{lloyd2016estimating}
+% \hl{and references therein; e.g. in}
+% \citealt{Close2015}
+% \hl{), or as outputs from an ordination, whether it being a principal components analysis (PCA,}
+% \citealt{PCA}
+% \hl{; e.g. in}
+% \citealt{zelditch2012geometric}
+% \hl{), a metric scaling (PCO, PCoA,}
+% \citealt{PCO}
+% \hl{; e.g. in}
+% \citealt{Brusatte12092008}
+% \hl{) or a non-metric scaling (MDS, NMDS,}
+% \citealt{MDS}
+% \hl{; e.g. in}
+% \citealt{Liow2004,DonohueDim}
+% \hl{).}
+% The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
+% For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.
+
+% \hl{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
+% This requires the definition of a proxy for space occupancy: the disparity metric}
+% \citep[or index;][]{Hopkins2017}
+% \hl{which can be measured in a multitude of ways.}
+% \hl{For example, one could use a metric based on the variance or the range of each axis of space}
+% \citep{Wills2001, Ciampaglio2001}
+% \hl{, a distance (e.g. Euclidean) measured between observations}
+% \citep{foote1993contributions,Foote29111996}
+% \hl{, a more direct approximation of the hyper volume }
+% \citep{cornwell2006trait,DonohueDim}
+% \hl{, or many more }
+% \citep[e.g.][]{navarro2003MDA}.
+
+% Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (
+% \hl{NPMANOVA, }
+% \citealt{NPMANOVA}
+% \hl{; e.g. in }
+% \citealt{Brusatte12092008}
+% \hl{)}
+% multidimensional permutation tests
+% \hl{(}
+% \citealt{ManlyPermutations}
+% \hl{; e.g. in }
+% \citealt{diaz2016global}
+% \hl{)}
+% or even simply by looking at the confidence interval overlaps between disparity measurements
+% \hl{().}
+% In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.

### 2017-12-11 Analysis/data_cleaning.Rmd
updated method compiler

@@
 ## Distance
-distances <- c("Gower distance", "Euclidean distance", "mean square euclidean distance", "Procrustes distances \\(geomorph\\)", "other distances", "Procrustes distances", "Average square distance", "mean distance", "mean euclidean distance", "mean Manhattan distance", "Euclidean distance between groups", "average squared distance", "mean euclidean distance", "Procrustes distance", "MNND \\(Mean nearest neighbour distance\\)", "Euclidean Procrustes distances", "sum of Procrustes distances", "range of maximum euclidean distance", "average Procrustes distance", "sum of the squared Procrustes distances")
+distances <- c("Gower distance", "Euclidean distance", "mean square euclidean distance", "Procrustes distances \\(geomorph\\)", "other distances", "Procrustes distances", "Average square distance", "mean distance", "mean euclidean distance", "mean Manhattan distance", "Euclidean distance between groups", "average squared distance", "mean euclidean distance", "Procrustes distance", "MNND \\(Mean nearest neighbour distance\\)", "Euclidean Procrustes distances", "sum of Procrustes distances", "range of maximum euclidean distance", "average Procrustes distance", "sum of the squared Procrustes distances", "distance from centroids", "centroid", "Procrustes distances from centre", "distance from centroids", "distance from neighbour", "furthest distance", "Average square distance from centroid", "distance from centroid", "centroid size", "distance from centroid", "Group centroid", "mean distance from centroids", "mean distance to centroid", "mean Procrustes distance from centroid", "distances to centroids")
 disparity <- multi.gsub(distances, "distances", disparity)
@@
 
-## Distance from centroid
-centroids <- c("distance from centroids", "centroid", "Procrustes distances from centre", "distance from centroids", "distance from neighbour", "furthest distance", "Average square distance from centroid", "distance from centroid", "centroid size", "distance from centroid", "Group centroid", "mean distance from centroids", "mean distance to centroid", "mean Procrustes distance from centroid", "distances to centroids")
-disparity <- multi.gsub(centroids, "centroids", disparity)
+# ## Distance from centroid
+# centroids <- c("distance from centroids", "centroid", "Procrustes distances from centre", "distance from centroids", "distance from neighbour", "furthest distance", "Average square distance from centroid", "distance from centroid", "centroid size", "distance from centroid", "Group centroid", "mean distance from centroids", "mean distance to centroid", "mean Procrustes distance from centroid", "distances to centroids")
+# disparity <- multi.gsub(centroids, "centroids", disparity)
 

### 2017-12-11 Analysis/disparity_methods.Rmd
updated method compiler

@@
 
-```{r}
+```{r, eval = FALSE}
 ## Removed columns
@@
 
-```{r, fig.width = 8, fig.height = 8}
+```{r, fig.width = 8, fig.height = 8, eval = FALSE}
 ## Get the variance per axis
@@
 
-```{r}
+```{r, eval = FALSE}
 cleaned_data <- readRDS("../Data/Processed/cleaned_data.Rda")
@@
 
-```{r, fig.width = 8, fig.height = 8}
+```{r, fig.width = 8, fig.height = 8, eval = FALSE}
 col_palette <- c("red", "blue", "green3", "grey")
@@
 
-```{r}
+```{r, eval = FALSE}
 ## Creating the custom groups list
@@
 
-```{r, fig.width = 12, fig.height = 12}
+```{r, fig.width = 12, fig.height = 12, eval = FALSE}
 ## Get the data for cumulative curves
@@
 
-```{r}
+```{r, eval = FALSE}
 plot.data(fuzzy.sort.data(data[,1], factors[[1]]), proportion = TRUE, ylim = c(0, 100), main = names(factors)[[1]])

### 2017-12-11 Writing/disparity_review.tex
review reframed

@@
 %Packages
+\usepackage{xcolor}
+\usepackage{color,soul}
 \usepackage{pdflscape}
+\usepackage{fixltx2e}
 \usepackage{textcomp}
@@
 \usepackage{dcolumn}
+\usepackage{hyphenat}
+\usepackage[T1]{fontenc}
+\usepackage{textcomp}
 \newcolumntype{d}[1]{D{.}{.}{#1}}
@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Natalie Cooper$^2$, Philip Donoghue$^3$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Natalie Cooper$^2$, ..., Philip Donoghue$^3$}\\
 \noindent {\small \it 
@@
 
-%§1: disparity is observed everytwhere
-Disparity is an old concept blabalbal coming from the observation that some groups of species are more similar/dissimilar than other.
+\noindent \hl{\textit{1: disparity is observed everywhere}}
+Biodiversity is not a smooth gradient of forms, functions, ecologies and behaviours but it is rather variable and discontinuous.
+This is at the base of Biological Sciences and is linked to the concept of disparity with the observation that some groups of species are more similar or dissimilar than others.
+This diversity in morphologies is in fact proper in nature and decoupled from taxonomic diversity: some groups may exhibit low diversity of species by high diversity of shapes \citep[or the other way around]{ruta2013,hopkinsdecoupling2013}.
 
-It can be also applied between groups as a concept of ecological niches: a group occupying many niches will have a stronger disparity than a groups with less niches.
+\noindent \hl{\textit{2: historical evolution of disparity}}
+Morphological diversity (hereafter disparity) is based on the observed the (dis)similarities between the morphologies of groups of organisms.
+In palaeobiology, this concept, as a tool for macroevolutionary studies, stems from several seminal papers from the 90s describing and questioning how the diversity of life forms arose and evolved \citep{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution}.
+In parallel, disparity analysis have been extended with modifications to macroevolutionary analysis based on continuous traits or geometric morphometric.
+The former usage of data was popularised following \cite{Harmon961} and its implementation in \texttt{geiger} \citep{geiger2008}; while the latter was popularised following \citep{zelditch2012geometric} %Probably older
+and implementations in \texttt{geomorph} (\citealt{adams2013geomorph, adams2017geometric} and precursory work from \citealt{claude2008morphometrics}).
 
-It is decoupled from diversity.
-
-\bigskip
-%§2: "phylogeny" of disparity
-
-In palaeobiology, morphological disparity (the (dis)similarities between the morphologies of groups of organisms) concept as a tool for macroevolutionary studies stems to several seminal papers from the 90s: \cite{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution}.
-
-This concept was then used in geometric morphometric and expanded to macroevolutionary studies in general (Harmon's dtt).
+\noindent \hl{\textit{3: the great popularity leads it to be used in many methods}}
 
 \cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.
-
-\bigskip
-
-%§3: a very popular concept
-
 Disparity is really popular (Fig. \ref{Fig:GoogleOccurences}):
@@
 \section{Why do we need disparity}
+
+\noindent \hl{\textit{These points should be developed during the meeting}}
+
 Which macroevolutionary questions are tackled through disparity analysis?
+\begin{itemize}
+    \item Bodyplan evolution through time
+    \item Why does some morphologies exist and other not?
+    \item tempo and mode of innovation of genuine novelty
+    \item evolutionary competition
+    \item effects of mass extinctions
+    \item ecological niches
+\end{itemize}
 
-What are the fundamental questions we should use to answer disparity questions?
+To answer these question, disparity has been used as a proxy for the biological increase/decrease of distinguishable morphological variation.
+These proximal multidimensional spaces have been described as the morphospace, the eco-space, the functional space, the cladisto space, the developemental space, the allometric space, etc \citep[see][and references therein]{Hopkins2017}.
+This multitude of multidimensional spaces arising from the observation of variable and discontinuous morphology in Life arises the question of their equivalence and their pertinence.
+Are the description of these spaces all equivalent?
+If disparity is the method or the metric describing these spaces then what \textit{is} disparity?
 
-What are the questions we should not answer with disparity?
-
-Disparity can be 
- -From a semantic view: morphospace or aspect of the morphospace?
- -From a biological view: what does an increase or a decrease in disparity represent?
-
-But in the end, what \textit{is} disparity
-
-\section{But what \textit{is} disparity?}
+\section{What \textit{is} disparity (really)?}
 
@@
 
-To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in the following categories:
-
-[DESCRIBE THE CATEGORIES HERE (from \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd})]
+To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).
 
@@
 
-[ADD Biological group, full organism and phylogeny in the supplementary]
+\hl{[ADD Biological group, full organism and phylogeny in the supplementary]}
+
+\noindent \hl{\textit{These points should be developed during the meeting}}
 
 \subsection{Disparity data}
-Describing the many data inputs (from collected data to morphospace):
+The data used for disparity methods comes from three main sources:
+\begin{itemize}
+    \item Continuous data such as limb or body measurements \citep{slaterCetacean}.
+    \item Discrete morphological characters \citep[sometimes referred to as ``Cladistic'' characters; ][]{Brusatte12092008}.
+    \item Geometric morphometric data which is generally the Procrustes transformation of 2D or 3D landmarks \citep{cooney2017mega}.
+\end{itemize}
+
+However, data was by far not limited to these three categories (e.g. colours \citealt{maia2013key}; metabolic rate \citealt{nespolo2016studying} or chemicals signal \citealt{garcia2017heterogeneous}).
+
+\subsection{Morphospaces}
+\begin{itemize}
+    \item PCAs [CITE].
+    \item PCOs [CITE]
+    \item Pairwise distance matrices \citep[no ordination; ][]{Harmon961}.
+\end{itemize}
+
+But also RWA [CITE] or Raup-spaces [CITE].
 
 \subsection{Disparity metrics}
-Describing the many metrics (what is actually measured):
+Throughout the 230 analysed papers, we found 103 unique combinations of metrics!
+
+\begin{itemize}
+    \item Distances measurements between species or from a certain point in the morphospace [CITE].
+    \item Ranges and variances of each axis of the morphospace \cite{Wills2001,Ciampaglio2001}
+    \item Distance between species based on the pairwise distance matrix (not the ordinations) [CITE].
+\end{itemize}
+
+But also metrics based on the volume [CITE], the characters dissimilarity [CITE] or the coordinates of some axis of the ordination [CITE]
 
@@
 Describing the many outputs (what and how is it tested):
+
+\begin{itemize}
+    \item Variance based (ANOVA, etc.) [CITE].
+    \item Correlation based [CITE].
+    \item Regression based (PGLS, etc.) [CITE].
+\end{itemize}
+
+But also discriminant analysis [CITE] and clustering [CITE].
 
@@
 \begin{itemize}
-    \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data for the full organisms to build a PCO from the organism's pairwise distance as a morphospace.
+    \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace.
 Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g.
@@
 
-    \item \textbf{The ``\texttt{dtt}'' approach:} continuous data for the full organism, not ordinated, with a pairwise distance based disparity metric and phylogenetic correction.
+    \item \textbf{The ``\texttt{dtt}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance).
+Hypothesis of higher/lower disparity can then be measured using null evolutionary models.
 
@@
 
-\section{What is missing}
+\section{Expanding disparity}
 
-How to compare disparity between groups?
+How to compare disparity between groups? Is disparity relative?
 
@@
 
-Hot to compare disparity between disciplines?
+Can we really say things about competition when looking at disparity in a single group?
 
-%TG: we can find solutions from other fields
-It bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.
+Are all characters equal? Character contingency suggests otherwise
+
+What is the relationship between disparity and tree shape?
+
+Do we have a null model for investigating disparity?
+
+Can we really say things about competition when looking at disparity in a single group?
+
+\noindent \hl{\textit{Disparity in other fields}}
+
+In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.
 
@@
 The Royal Society.
-I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
+TG acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
 

### 2017-12-15 Writing/disparity_review.tex
rewritten intro (kindof)

--- 2017-12-06 Writing/disparity_review.tex
@@
 
-Disparity, as a concept or set of toolkits is used a lot in biology and is more and more used every year.
+Biodiversity is not a smooth gradient of forms, functions, ecologies, behaviours, etc., but it is rather variable and discontinuous.
+This is at the base of biological sciences and is linked to the concept of disparity \citep{Wills2001,Hopkins2017} with the observation that some groups of species are more similar or dissimilar than others.
+This diversity in morphologies is in fact proper in nature and decoupled from taxonomic diversity: some groups may exhibit low diversity of species by high diversity of shapes \citep[or the other way around]{ruta2013,hopkinsdecoupling2013}.
+In palaeobiology, this concept, has been studied since the 90s describing \citep{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution} and as been extended with modifications to macroevolutionary analysis in general since the 2000s (whether based on continuous traits; e.g.
+\citealt{Harmon961,geiger2008}; or on geometric morphometrics; e.g.
+\citealt{claude2008morphometrics,zelditch2012geometric,adams2013geomorph,adams2017geometric}).
+
+Since then, morphological disparity has been used for testing a vast array of biological hypotheses such as: how did body plans evolve through time?
+\citep{Wesley-Hunt2005}; why does some morphologies exist and not other?
+\cite{gerber2017geometry}; what is the tempo and mode of morphological innovations through time?
+\citep{Hughes20082013}; what is the effect of mass extinction on disparity?
+\citep{halliday2016eutherian}; how does competition between groups affect disparity?
+\citep{Brusatte12092008}; how did morpho-ecological niches get filled through time?
+\citep{price2014niche}; and many more.
+%
+This apparent versatility of morphological disparity analysis led to a phenomenal amount of papers based on disparity methods (Fig. \ref{Fig:GoogleOccurences})!
+%
+However, ``a lot of disparity studies are inductive fishing trips and are not often designed to test a specific hypothesis.''
+%
+Indeed, compared to the vast amount of publication using morphological disparity, only a handful highlight/solve problems and propose hypothesis testing.
 
@@
    \includegraphics[width=1\textwidth]{Figures/GoogleScholarOccurences.pdf} 
-\caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year. In black, the match is in the paper and in grey, in the title.}
+\caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year.
+In black, the match is in the paper and in grey, in the title.
+We collected the number of matches per year from 1960 to 2016 in Google Scholar for the terms "morphological disparity" both in the text (fuzzy matching) or in the title (exact matching).
+The data was collected on the 1st of November 2017.}
 \label{Fig:GoogleOccurences}
@@
 
-In most of these papers, they trace back the disparity concept to seminal papers from the 90s: \cite{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution}.
-However, being such a broad and obvious concept, it can probably be stretched back to way earlier.
+\section{What actually \textit{is} disparity?}
 
@@
 
-Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.
-One solution to this problem is to analyse data in a multivariate framework.
-Multivariate analyses aim to capture and incorporate this multidimensional complexity, while providing outputs that are interpretable in a physical world of only three dimensions.
-Classical multidimensional species features that have been analysed through multivariate data ranges from their morphology \citep{raup1966geometric}, to their functional traits \citep{diaz2016global}.
-However, such analysis are not limited at the species level and can also be applied to ecosystems \citep{DonohueDim}.
+Are all these methods the right tools to answer these questions (described in the intro) or are they meer “inductive fishing trips” and are actually not testing any of the hypothesis related to the disparity questions:
 
-Key multidimensional features of species that have important roles in ecology and evolution include morphology, functional traits. etc. etc.
-Many of these can be represented as matrices, these can be used in our package.
-In the interests of clarity/brievety, we here focus on just one kind of data: morphological diversity.
+Can the data used in disparity studies actually test the phenomenon researchers are trying to capture?
+E.g. for discrete morphological characters: disparity of fossilizable characters.
+E.g. for morphometrics: disparity of the skull procrustes
+E.g. for dtt: disparity of body/limb/beak length
+Are  these data types enough to test disparity hypothesis or do we need more data/tone down the hypothesis?
 
-Furthermore, we will also make a clear distinction between the multidimensional space which is a specific mathematical object and the disparity, which is a metric describing or summarising one or more aspects of this space.
+Does the transformation of this data reflect our hypothesis?
+E.g.
+a morphospace as a pairwise matrix reflects the dissimilarity between the data, a morphospace based on an ordination reflects the variability within the data, a morphospace on both reflects the variability within the dissimilarity, are each method reflecting that?
 
-Both can be defined in various ways depending on the authors, for example, disparity is defined as the weighted mean pairwise dissimilarity in \cite{Close2015} or the ellipsoid volume in \cite{DonohueDim} and the multidimensional space is defined as the morphospace in \cite{raup1966geometric} and the morpho-functional space in \cite{diaz2016global}.
+Do the disparity metrics really reflect the variation in the hypothesis that are tested?
+E.g. Are the variance/range based metrics a description of the ordination or the observed biological phenomenon (i.e. they are not including covariance and phylogenetic signal)
+E.g. What does the average distance between species or from a point really convey? Is the distribution not more interesting?
+Do we need more like metrics measuring explicitly what is tested in the hypothesis (rather than approximate it).
+For example, for competitive replacement, should we not measure whether group A occupies the same place as group B through time and whether that is due to competition or simply because any other place in the space cannot be occupied (and why are such spaces unoccupied)?
 
-The multidimensional space can be defined in many ways and arise from many mathematical transformations of the data such as the pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
-Similarly, disparity metrics \citep[or indices;][]{Hopkins2017} can defined in many ways \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015,diaz2016global}.
-Finally, difference between disparity metrics can also be measured in many ways: using NPMANOVA \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simple confidence interval overlap \citep[e.g.][]{halliday2016eutherian}.
+Are the methods used to test hypothesis in disparity actually testing what we think they’re testing (problem of multidimensional hypothesis)
+What does an anova or related variance based test real infirms/confirms regarding the disparity questions?
+How do we deal with bootstrapped data and pseudo-replications?
+Can we realistically use univariate gaussian models to test hypothesis based on multivariate, probably non-gaussian data?
 
-This variety of definitions and analysis have been developed in an equal variety of softwares such as \texttt{GINGKO} in javascript \citep{bouxin2005ginkgo,de2007ginkgo} or \texttt{geomorph} \citep{adams2013geomorph,adams2017geometric}, \texttt{Claddis} \citep{Claddis}, or \texttt{vegan} \citep{oksanen2007vegan} in R \citep{R}.
-This results in the need to learn different languages (or at least - when restricted to R - different packages with different standards) as well as making analysis sometimes idiosyncratic and often complex to repeat since they are based on a particular feature from a particular software.
-For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function.
-Unfortunately, however, the multidimensional space can only be defined as the ordination of the procrustes transform of geometric morphometric landmarks, the disparity can only be defined as the Procrustes variance and the difference between groups can only be measured through permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
+Conclusion:
+We need a more hypothesis driven way of analysing disparity. Disparity methods should be the tool for answering disparity hypothesis, not merely a description of the data.
 
-Disparity is an old concept blabalbal
+% \section{What \textit{is} disparity (really)?}
 
-\section{What \textit{is} disparity}
+% Disparity can describe either the metric (disparity index \citep{Hopkins2017} or disparity metric \citep{Wills2001}) or the whole pipeline \citep{Claddis,zelditch2012geometric}
 
-From a semantic view: morphospace or aspect of the morphospace?
+% % §Methods for data:
+% \subsection{Data collection methodology}
+% To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
+% For the 230 relevant papers among the 500 matches, we collected the following methodological data:
 
-From a biological view: what does an increase or a decrease in disparity represent?
+% \begin{itemize}
+%     \item What was the focal biological group?
+%     \item What kind of data was measured (e.g. landmarks, discrete data, etc.)?
+%     \item Was data collected on the full organism or not?
+%     \item How was the morphospace explicitly defined (e.g. PCA, PCO, MDS, etc.)?
+%     \item How was the disparity metric(s) explicitly defined?
+%     \item Which statistical test was applied to test the disparity related hypothesis?
+%     \item Was phylogeny taken into account or not?
+% \end{itemize}
 
-\section{Why do we need disparity}
+% We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g.
+a disparity metric was measured but not described anywhere in the paper).
 
-What are the fundamental questions we should use to answer disparity questions?
+% The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).
 
-What are the questions we should not answer with disparity?
+% To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).
 
-\section{What is missing}
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
+% \caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
+% \label{Fig:MethodsProportions}
+% \end{figure}
 
-Methods?
-Data?
+% \hl{[ADD Biological group, full organism and phylogeny in the supplementary]}
 
-\section{Conclusion}
+% \noindent \hl{\textit{These points should be developed during the meeting}}
 
-Yay!
+% \subsection{Disparity data}
+% The data used for disparity methods comes from three main sources:
+% \begin{itemize}
+%     \item Continuous data such as limb or body measurements \citep{slaterCetacean}.
+%     \item Discrete morphological characters \citep[sometimes referred to as ``Cladistic'' characters; ][]{Brusatte12092008}.
+%     \item Geometric morphometric data which is generally the Procrustes transformation of 2D or 3D landmarks \citep{cooney2017mega}.
+% \end{itemize}
 
-% \section{Data availability and reproducibility}
-% Data are available on Dryad or Figshare.
-% Code for reproducing the analyses is available on GitHub (\url{github.com/TGuillerme/SpatioTemporal_Disparity}).
+% However, data was by far not limited to these three categories (e.g.
+colours \citealt{maia2013key}; metabolic rate \citealt{nespolo2016studying} or chemicals signal \citealt{garcia2017heterogeneous}).
+
+% \subsection{Morphospaces}
+% \begin{itemize}
+%     \item PCAs [CITE].
+%     \item PCOs [CITE]
+%     \item Pairwise distance matrices \citep[no ordination; ][]{Harmon961}.
+% \end{itemize}
+
+% But also RWA [CITE] or Raup-spaces [CITE].
+
+% \subsection{Disparity metrics}
+% Throughout the 230 analysed papers, we found 103 unique combinations of metrics!
+
+% \begin{itemize}
+%     \item Distances measurements between species or from a certain point in the morphospace [CITE].
+%     \item Ranges and variances of each axis of the morphospace \cite{Wills2001,Ciampaglio2001}
+%     \item Distance between species based on the pairwise distance matrix (not the ordinations) [CITE].
+% \end{itemize}
+
+% But also metrics based on the volume [CITE], the characters dissimilarity [CITE] or the coordinates of some axis of the ordination [CITE]
+
+% \subsection{Disparity hypothesis}
+% Describing the many outputs (what and how is it tested):
+
+% \begin{itemize}
+%     \item Variance based (ANOVA, etc.) [CITE].
+%     \item Correlation based [CITE].
+%     \item Regression based (PGLS, etc.) [CITE].
+% \end{itemize}
+
+% But also discriminant analysis [CITE] and clustering [CITE].
+
+% \subsection{The three main disparity analysis}
+% From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.
+
+% \begin{itemize}
+%     \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace.
+Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g.
+the sum of variance or/and the sum of ranges).
+Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.
+
+%     \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA).
+Disparity is often measured as a distance metric (e.g.
+the distance between the species and a point in the morphospace such as its centroid).
+Hypothesis are then tested using ANOVA type tests with usually no phylogenetic correction (although phylogeny is sometimes used to correct the morphospace).
+
+%     \item \textbf{The ``\texttt{dtt}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance).
+Hypothesis of higher/lower disparity can then be measured using null evolutionary models.
+
+% \end{itemize}
+
+% Of course some studies use a combination of these three methods or none of them at all!
+
+% Also, among each category XX\% of studies use multiple approaches.
+
+% [ADD Biological group, full organism and phylogeny in the supplementary]
+
+% \section{Expanding disparity}
+
+% How to compare disparity between groups? Is disparity relative?
+
+% How to compare disparity between methods?
+
+% Can we really say things about competition when looking at disparity in a single group?
+
+% Are all characters equal? Character contingency suggests otherwise
+
+% What is the relationship between disparity and tree shape?
+
+% Do we have a null model for investigating disparity?
+
+% Can we really say things about competition when looking at disparity in a single group?
+
+% \noindent \hl{\textit{Disparity in other fields}}
+
+% In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.
+
+% \section{Conclusion}
+
+% A quick guideline for good disparity analysis:
+
+% Maybe we need something like in Parham et al 2012 (Best Practices for Justifying Fossil Calibrations): an easy an identifiable description of the pipeline containing: 1) the type of data, 2) the morphospace and 3) the metric?
 
 \section{Acknowledgments}
-Royal Society..
-I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
+The Royal Society.
+TG acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
 

### 2017-12-15 Writing/disparity_review.tex
first draft of the data part

@@
 \section{Introduction}
+\label{text:intro}
 
-Biodiversity is not a smooth gradient of forms, functions, ecologies, behaviours, etc., but it is rather variable and discontinuous.
+Biodiversity is not a smooth gradient of forms, functions, ecologies, behaviours, etc., but it is rather variable and discontinuous. %TG: Actually is it?
 This is at the base of biological sciences and is linked to the concept of disparity \citep{Wills2001,Hopkins2017} with the observation that some groups of species are more similar or dissimilar than others.
@@
 \cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.
+Disparity can describe either the metric (\citealt{Wills2001}; or index \citealt{Hopkins2017}) or the whole pipeline \citep{zelditch2012geometric,lloyd2016estimating}.
+To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
+For the 230 relevant papers among the 500 matches, we collected the following methodological data: (1) What was the focal biological group?
+; (2) What kind of data was measured (e.g.
+landmarks, discrete data, etc.)?
+; (3) Was data collected on the full organism or not?
+; (4) How was the morphospace explicitly defined (e.g.
+PCA, PCO, MDS, etc.)?
+; (5) How was the disparity metric(s) explicitly defined?
+; (6) Which statistical test was applied to test the disparity related hypothesis?
+; (7) Was phylogeny taken into account or not?
+We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g.
+a disparity metric was measured but not described anywhere in the paper).
+The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).
+To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).
 
-Are all these methods the right tools to answer these questions (described in the intro) or are they meer “inductive fishing trips” and are actually not testing any of the hypothesis related to the disparity questions:
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
+\caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
+\label{Fig:MethodsProportions}
+\end{figure}
 
+From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.
+
+\begin{itemize}
+    \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace.
+Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g.
+the sum of variance or/and the sum of ranges).
+Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.
+
+    \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA).
+Disparity is often measured as a distance metric (e.g.
+the distance between the species and a point in the morphospace such as its centroid).
+Hypothesis are then tested using ANOVA type tests with usually no phylogenetic correction (although phylogeny is sometimes used to correct the morphospace).
+
+    \item \textbf{The ``\texttt{approach}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance).
+Hypothesis of higher/lower disparity can then be measured using null evolutionary models.
+\end{itemize}
+
+%
+The advent of these three approaches coincides with the explosion of disparity analysis (Fig. \ref{Fig:GoogleOccurences}) and suggest a high popularity of disparity analysis.
+%
+Unfortunately, however, few actual tests have been performed on whether any of these approaches \textit{actually} allows to tackle the classic disparity questions (see \ref{text:intro}{Introduction}).
+Are all these methods the right tools to answer these questions are they mere ``inductive fishing trips'' used to describe the data rather than testing any of the hypothesis related to the disparity questions?
+
+\section{Does data used in disparity analysis allows to test hypothesis?}
+In fact, the question of which data to analyse \citep{hetherington2015cladistic} and which part of the data to use has been analysed \citep{hopkins2017well} has rarely been assessed.
 Can the data used in disparity studies actually test the phenomenon researchers are trying to capture?
-E.g. for discrete morphological characters: disparity of fossilizable characters.
-E.g. for morphometrics: disparity of the skull procrustes
-E.g. for dtt: disparity of body/limb/beak length
-Are  these data types enough to test disparity hypothesis or do we need more data/tone down the hypothesis?
 
-Does the transformation of this data reflect our hypothesis?
-E.g.
-a morphospace as a pairwise matrix reflects the dissimilarity between the data, a morphospace based on an ordination reflects the variability within the data, a morphospace on both reflects the variability within the dissimilarity, are each method reflecting that?
+For example, when studying discrete morphological characters (often following the ``\texttt{Claddis}'' approach), the data used is often recycled data used for phylogenetic analysis.
+These datasets are usually design to recover evolutionary history and differentiate clades and, often due to the preservations of the fossil record, are mainly based on hard tissue characters.
+In phylogenetics, this artefact have been shown to be leading to measurable topological recovery error (for soft \textit{vs} hard characters; \citealt{sansomfossilization2013}; or dental \textit{vs} cranial characters \citealt{sansom2017dental}).
+Therefore, is the morphospace composed of ``fossilisable'' characters enough for answering evolutionary hypothesis?
+For example, are niche replacement hypothesis equivalent to fossilisable-character-based-niches replacement hypothesis?
 
+Similarly, many studies in geometric morphometric are based solely on skull shape using landmark techniques combined with Procrustes sumperimposition \citep[the ``\texttt{geomorph}'' approach;][]{zelditch2012geometric,adams2017geometric}.
+However, it has been shown that the whole skeleton displays different integrated modules with different specific evolutionary constraints leading to variable rates and modes of evolution (\citealt{Goswami20130254}; and this even at the cranial level \citealt{goswami2010influence}).
+Is skull variation enough for measuring body plans variation through time for example?
+
+In the case where disparity is measured a single continuous trait (or a collection of traits; following the ``\texttt{geiger}'' approach), for example in \cite{price2014niche}, beak and limb length as well as body mass are used to differentiate groups of birds and different zones in the morphospace can then be interpreted as different niches.
+Is this set of traits studied in isolation sufficient to test niche occupancy hypothesis?
+
+Additionally, in many cases, the disparity metrics and hypothesis are often based on a transformation of the data (pairwise distances, PCA, PCO, MDS, RWA, etc.).
+Can these transformation readily reflect the raw data in our hypothesis?
+For examples a morphospace as a pairwise matrix reflects the dissimilarity between the data, a morphospace based on an ordination reflects the variability within the data, a morphospace based on both reflects the variability within the dissimilarity, are each method reflecting that?
+
+\section{Do the disparity metrics (or indices) really reflect variation in the studied phenomenon?}
 Do the disparity metrics really reflect the variation in the hypothesis that are tested?
@@
 
+\section{Is the current statistical toolkit sufficient for testing disparity hypothesis?}
 Are the methods used to test hypothesis in disparity actually testing what we think they’re testing (problem of multidimensional hypothesis)
@@
 
-Conclusion:
-We need a more hypothesis driven way of analysing disparity. Disparity methods should be the tool for answering disparity hypothesis, not merely a description of the data.
+\section{What are we missing?}
 
-% \section{What \textit{is} disparity (really)?}
+\subsection{Learning lessons from other fields}
+In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.
 
-% Disparity can describe either the metric (disparity index \citep{Hopkins2017} or disparity metric \citep{Wills2001}) or the whole pipeline \citep{Claddis,zelditch2012geometric}
-
-% % §Methods for data:
-% \subsection{Data collection methodology}
-% To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
-% For the 230 relevant papers among the 500 matches, we collected the following methodological data:
-
-% \begin{itemize}
-%     \item What was the focal biological group?
-%     \item What kind of data was measured (e.g. landmarks, discrete data, etc.)?
-%     \item Was data collected on the full organism or not?
-%     \item How was the morphospace explicitly defined (e.g. PCA, PCO, MDS, etc.)?
-%     \item How was the disparity metric(s) explicitly defined?
-%     \item Which statistical test was applied to test the disparity related hypothesis?
-%     \item Was phylogeny taken into account or not?
-% \end{itemize}
-
-% We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g.
-a disparity metric was measured but not described anywhere in the paper).
-
-% The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).
-
-% To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).
-
-% \begin{figure}[!htbp]
-% \centering
-%    \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
-% \caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
-% \label{Fig:MethodsProportions}
-% \end{figure}
+\section{Conclusion}
+We need a more hypothesis driven way of analysing disparity.
+Disparity methods should be the tool for answering disparity hypothesis, not merely a description of the data.
 
@@
 
-% \subsection{The three main disparity analysis}
-% From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.
-
-% \begin{itemize}
-%     \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace.
-Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g.
-the sum of variance or/and the sum of ranges).
-Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.
-
-%     \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA).
-Disparity is often measured as a distance metric (e.g.
-the distance between the species and a point in the morphospace such as its centroid).
-Hypothesis are then tested using ANOVA type tests with usually no phylogenetic correction (although phylogeny is sometimes used to correct the morphospace).
-
-%     \item \textbf{The ``\texttt{dtt}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance).
-Hypothesis of higher/lower disparity can then be measured using null evolutionary models.
-
-% \end{itemize}
-
 % Of course some studies use a combination of these three methods or none of them at all!

### 2017-12-15 Writing/disparity_review.tex
Added dimensions effect example

@@
 \section{Do the disparity metrics (or indices) really reflect variation in the studied phenomenon?}
+From the 230 papers analysed, we found 104 unique combination's of metrics to measure disparity (Fig. \ref{Fig:MethodsProportions}).
+Commonly, these metrics are used to approximate the volume occupancy within the morphospace \citep{Wills2001,DonohueDim,diaz2016global}.
+
+However, it has been shown that different metrics reacts differently depending on the dataset
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=1\textwidth]{Figures/DimensionsEffect.pdf} 
+\caption{Effect of the number of dimensions on different disparity metrics (note that the sum of variance is equal to the average squared pairwise distance in this case).}
+\label{Fig:DimensionsEffect}
+\end{figure}
+
+Product of variance
+
+Is disparity only relative?
+
 Do the disparity metrics really reflect the variation in the hypothesis that are tested?

### 2017-12-18 Analysis/effect_dimensionality.Rmd
Dimensionality test

---
title: "Effect of dimensionality on metrics"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Effect of dimensionality on disparity metrics

## Preparing the data

```{r}
set.seed(123)
library(dispRity)
data(BeckLee_mat50)

## Measuring for different metrics
sum_range <- sapply(2:48, function(X) dispRity(BeckLee_mat50, metric = c(sum, ranges), dimensions = X), simplify = FALSE)
prod_range <- sapply(2:48, function(X) dispRity(BeckLee_mat50, metric = c(prod, ranges), dimensions = X), simplify = FALSE)
sum_var <- sapply(2:48, function(X) dispRity(BeckLee_mat50, metric = c(sum, variances), dimensions = X), simplify = FALSE)
prod_var <- sapply(2:48, function(X) dispRity(BeckLee_mat50, metric = c(prod, variances), dimensions = X), simplify = FALSE)
med_centroid <- sapply(2:48, function(X) dispRity(BeckLee_mat50, metric = c(median, centroids), dimensions = X), simplify = FALSE)
avg_sq <- sapply(2:48, function(X) dispRity(BeckLee_mat50, metric = function(matrix){mean(pairwise.dist(matrix)^2)}, dimensions = X), simplify = FALSE)
ell_vol <- sapply(2:48, function(X) dispRity(BeckLee_mat50, metric = ellipse.volume, dimensions = X), simplify = FALSE)
n_ball <- sapply(2:48, function(X) dispRity(BeckLee_mat50, metric = n.ball.volume, dimensions = X), simplify = FALSE)

## Extracting the observed values
sum_range <- unlist(lapply(sum_range, function(X) return(summary(X)$obs)))
prod_range <- unlist(lapply(prod_range, function(X) return(summary(X)$obs)))
sum_var <- unlist(lapply(sum_var, function(X) return(summary(X)$obs)))
prod_var <- unlist(lapply(prod_var, function(X) return(summary(X)$obs)))
med_centroid <- unlist(lapply(med_centroid, function(X) return(summary(X)$obs)))
avg_sq <- unlist(lapply(avg_sq, function(X) return(summary(X)$obs)))
ell_vol <- unlist(lapply(ell_vol, function(X) return(summary(X)$obs)))
n_ball <- unlist(lapply(n_ball, function(X) return(summary(X)$obs)))

## Scale these scores
scale.max <- function(x) return(x/max(x))
sum_range_sca <- scale.max(sum_range)
prod_range_sca <- scale.max(prod_range)
sum_var_sca <- scale.max(sum_var)
prod_var_sca <- scale.max(prod_var)
med_centroid_sca <- scale.max(med_centroid)
avg_sq_sca <- scale.max(avg_sq)
ell_vol_sca <- scale.max(ell_vol)
n_ball_sca <- scale.max(n_ball)

metrics <- list(sum_range_sca, prod_range_sca, sum_var_sca, prod_var_sca, med_centroid_sca, avg_sq_sca, ell_vol_sca, n_ball_sca)
metrics_names <-  c("sum of ranges", "product of ranges", "sum of variance", "product of variance", "median distance from centroid", "average squared pairwise distance", "ellispoid volume", "minimum n-ball volume")
```

```{r, fig.width = 12, fig.height = 12}
## gg plot colours
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

cols <- gg_color_hue(8)
ylim <- c(0,1)
xlim <- c(2,48)

## Plotting the results
pdf("../Writing/Figures/DimensionsEffect.pdf", width = 12, height = 12)
op <- par(bty = "n")

plot(1,1, type = "n", ylim = ylim, xlim = xlim, xlab = "Dimensions", ylab = "Relative disparity")

for(metric in 1:8) {
    lines(x = c(2:48), metrics[[metric]], col = cols[metric])
}

legend("right", legend = metrics_names, col = cols, lty = 1)
par(op)
dev.off()
```

### 2017-12-18 Writing/disparity_review.tex
metric part up

@@
 
-However, it has been shown that different metrics reacts differently depending on the dataset
+However, it has been shown that different metrics reacts differently depending on the dataset \citep[][; and an anecdotal example in Fig. \ref{Fig:DimensionsEffect}]{Wills2001,Ciampaglio2001}.
+For example metrics involving a product of aspects of the matrix (e.g.
+the volume, the product of the dimensions) are likely to suffer from the \textit{Curse of dimensionality} \citep{bellman1957dynamic} and will quickly tend to null values.
+Some other metrics are not representing all the data captured in the morphospace (e.g. the sum or product of variance ignores the co-variance between dimensions).
+Are these metrics as a description of the ordination a good proxy for the observed biological phenomenon?
 
@@
    \includegraphics[width=1\textwidth]{Figures/DimensionsEffect.pdf} 
-\caption{Effect of the number of dimensions on different disparity metrics (note that the sum of variance is equal to the average squared pairwise distance in this case).}
+\caption{Effect of the number of dimensions on different disparity metrics (note that the sum of variance is equal to the average squared pairwise distance in this case).
+The data used is one of the example datasets from the \texttt{dispRity R} package \citep[\texttt{BeckLee\_mat50};][]{beckancient2014,dispRityv02} composed of 50 elements and 48 dimensions obtained from a ordination (MDS) of the distance matrix of a discrete morphological character matrix.}
 \label{Fig:DimensionsEffect}
@@
 
-Product of variance
+Despite these caveats, it is important to actually know what the metrics are measuring.
+There has been some attempts to test that through simulations \citep{Ciampaglio2001,  gerber2017geometry} but it is still unclear which metric reflects the best the hypothesis test.
+For example, the average squared pairwise distance (or the sum of ranges) are really popular \citep[e.g.][]{geiger2008} yet has been shown to be fairly senstive, at least in empirical data to the number of sampled taxa, the number of characters and the data completeness \citep[][; although effects were lesser in simulations]{Ciampaglio2001}.
+More fundamentally, the question arises whether one disparity metric could be sufficient to describe changes in disparity?
+Changes in multidimensions might not be readily grasped by unidimensional metrics!
+For example, a given taxonomic group might have the exact same volume as another one yet be in a complete different part of the morphospace \citep[or even overlap or not depending on the dimensions;][]{davis2012acanthodes,brusatte50}.
+In this case, it might more interesting to have multiple metrics (e.g.
+the position and the volume) [FABRE et.
+al in prep.] or to use distributions of metrics (e.g.
+what does the average distance between species or from a point really convey?
+Is the distribution not more interesting?).
 
-Is disparity only relative?
-
-Do the disparity metrics really reflect the variation in the hypothesis that are tested?
-E.g. Are the variance/range based metrics a description of the ordination or the observed biological phenomenon (i.e. they are not including covariance and phylogenetic signal)
-E.g. What does the average distance between species or from a point really convey? Is the distribution not more interesting?
-Do we need more like metrics measuring explicitly what is tested in the hypothesis (rather than approximate it).
+Maybe we need metrics that fit explicitly our hypothesis (or fit our hypothesis directly to the metric).
 For example, for competitive replacement, should we not measure whether group A occupies the same place as group B through time and whether that is due to competition or simply because any other place in the space cannot be occupied (and why are such spaces unoccupied)?
@@
 How do we deal with bootstrapped data and pseudo-replications?
-Can we realistically use univariate gaussian models to test hypothesis based on multivariate, probably non-gaussian data?
+Can we realistically use univariate gaussian models to test hypothesis based on multivariate, probably non-gaussian data? \citep{blomberg2017beyond}
+
+Should be take phylogeny into account? \citep{polly2013phylogenetic}
 

### 2018-01-04 Writing/disparity_review.tex
review up

@@
 
-Similarly, many studies in geometric morphometric are based solely on skull shape using landmark techniques combined with Procrustes sumperimposition \citep[the ``\texttt{geomorph}'' approach;][]{zelditch2012geometric,adams2017geometric}.
+Similarly, many studies in geometric morphometric are based solely on skull shape using landmark techniques combined with Procrustes superimposition \citep[the ``\texttt{geomorph}'' approach;][]{zelditch2012geometric,adams2017geometric}.
 However, it has been shown that the whole skeleton displays different integrated modules with different specific evolutionary constraints leading to variable rates and modes of evolution (\citealt{Goswami20130254}; and this even at the cranial level \citealt{goswami2010influence}).
@@
 Are the methods used to test hypothesis in disparity actually testing what we think they’re testing (problem of multidimensional hypothesis)
-What does an anova or related variance based test real infirms/confirms regarding the disparity questions?
+What does an anova or related variance based test real infirm/confirms regarding the disparity questions?
 How do we deal with bootstrapped data and pseudo-replications?
-Can we realistically use univariate gaussian models to test hypothesis based on multivariate, probably non-gaussian data? \citep{blomberg2017beyond}
+Can we realistically use univariate Gaussian models to test hypothesis based on multivariate, probably non-Gaussian data? \citep{blomberg2017beyond}
 

### 2018-01-16 Writing/disparity_review.tex
Rewriting the review (round 1!)

--- 2017-12-15 Writing/disparity_review.tex
@@
 
-\noindent{\Large \bf Disparate views on disparity.} 
+\noindent{\Large \bf Variations in multidimensional space.} 
 \bigskip
 
-\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Natalie Cooper$^2$, ..., Philip Donoghue$^3$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Natalie Cooper$^2$, Stephen Brusatte$^3$, Katie Davis$^4$, Andrew Jackson$^5$, Sylvain Gerber$^6$, Anjali Goswami$^2$, Kevin Healy$^7$, Melanie Hopkins$^8$, Graeme Lloyd$^9$, Joseph O'Reilly$^{10}$, Abi Pate$^{10}$, Emilie Rayfield$^{10}$, Erin Saupe$^{11}$, Emma Sherratt$^{12}$, Graham Slater$^{13}$, Gavin H Thomas$^{14}$ and Philip Donoghue$^{10}$}\\
+
 \noindent {\small \it 
-$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\}
+$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom; $^2$NHM; $^10$Bristol}
 \end{center}
@@
 \section{Introduction}
-\label{text:intro}
 
-Biodiversity is not a smooth gradient of forms, functions, ecologies, behaviours, etc., but it is rather variable and discontinuous. %TG: Actually is it?
-This is at the base of biological sciences and is linked to the concept of disparity \citep{Wills2001,Hopkins2017} with the observation that some groups of species are more similar or dissimilar than others.
-This diversity in morphologies is in fact proper in nature and decoupled from taxonomic diversity: some groups may exhibit low diversity of species by high diversity of shapes \citep[or the other way around]{ruta2013,hopkinsdecoupling2013}.
-In palaeobiology, this concept, has been studied since the 90s describing \citep{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution} and as been extended with modifications to macroevolutionary analysis in general since the 2000s (whether based on continuous traits; e.g.
+%1 -
+% History of disparity methods
+Since the advent of biology and probably before that, it has been observed that variation in morphologies follows discrete patterns rather than a continuous gradient of forms.
+Disparity has shifted from it's English meaning (i.e.
+dissimilarity) in the seventies to arrive to the ``standardised'' and popularised meaning we now today in biology [Runnegar 1987, Gould, Foote, Webb, etc.].
+In these seminal papers, disparity was used to describe a particular biological, phenotypic, and/or trait variation without strict distinction between the ontology and episthemology (see Semantic Box).
+Over the last 30 years, disparity has been usually used in the morphological disparity context.
+The numerous papers are mainly united through the study of phenotypes, using often multivariate data and methods focused on the species level or above in a macroevolutionary context.
+However, it is important to not that these sets of approaches are not restricted to areas.
+It has been used a microevolutionary levels or even in other fields such as ecology or linguistics.
+Our approach throughout this paper is to not constrain our definition of disparity to not build barriers between disciplines.
+Disparity provides data on the structure of biodiversity.
+In essence, disparity is a pattern of variation.
+
+%2§ what use: commonly used
+The description of these patterns (the disparity of patterns) has thus been used in evolutionary biology to tackle a vast array of questions:
+Developmental constraints [CITE]; 
+Niche width, competition  - ecological perspective / specialisation [CITE];
+Relationship between form and function and evolution functional traits [CITE];
+Disparity through time [CITE];
+Response to extinction - Extinction selectivity [CITE];
+Competition - ecologically relevant [CITE];
+Evolutionary mode [CITE];
+Evolutionary rate [CITE];
+Macroecology (disparity in space)[CITE];
+Tree Structure and the effect on disparity [CITE];
+Adaptive radiation ecological diversification [CITE];
+Scaling micro to macro [CITE];
+Competitive replacements [CITE];
+Tempo and mode of evolution [CITE].
+
+%3§ problems
+Disparity studies have a great deal promise in helping us to understand the evolution of biological diversity.
+However, we identify three main issues in current disparity studies.
+First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
+Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large cladistic datasets.
+Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
+Here we deal with some of these issues and present best practice guidelines/sets of things to think about before performing a disparity analysis.
+
+\subsubsection{Disparity patterns or processes}
+At the core of defining what \textit{is} disparity lies a divide between episthemology and ontology.
+Is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
+If disparity describes patterns of variation, the underlying scientific question ultimately always wants to use them as a proxy for explaining the biological processes.
+In the seminal papers from the late 80s, disparity was solely use to describe and understand the Cambrian Explosion were both the pattern (a high observed dissimilarity between groups) was directly linked to the process (and explosive radiation of multicellular life as championed by Gould).
+Restricting disparity to this initial informal definition (multidimensional morphological at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
+% That is, infer the mode of evolution (how does evolution occur and how does the structure of variation occur - both a constructive and destructive process) and the tempo (i.e., rate) at which it occurs.
+I think variation can also potentially tell us about ecology (as separate to evolution - for example, function, community structure, competition, etc.)...but...is ecology different from evolution??
+….Something I have been debating is whether it tells us about environment.
+Can we understand something about past environments from disparity??
+
+\section{Best practices for disparity analysis}
+
+\subsection{Find a question}
+Why to perform a disparity analysis?
+It might sound like an obvious question to start with but we would like to stress its crucial importance.
+The last decade have seen a vast array of excellent methodological implementation and documentations rendering disparity analysis easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
+However, this easiness might (conscientiously or not) push authors to perform disparity analysis ``just because''.
+In fact it is important to not rush into a disparity analysis to avoid some basic statistical caveats:
+For example, one major thing to consider is circularity: if your aim is to explore the variation in the data set, a multi-dimensional analysis is perfectly valid.
+However, the resulting groups are then no more appropriate for testing a hypothesis.
+Instead, rather do the data exploration on a subset of the data (preliminary analysis) and then use the full data set to test the resulting hypothesis.
+We believe that clearly framing an hypothesis and its null expectancy will help designing better disparity analysis.
+Once the question clearly framed, which data and which methods will be appropriate to answer it?
+
+\subsection{Collect the appropriate data}
+If disparity is used to describe pattern to explain some process, remember that the every possible data will always only be a proxy for these patterns.
+As a sub-note to this, more data is not necessarily better, especially when the data was originaly designed and collected for different questions (e.g. cladistic data).
+Again, framing the question helps to avoid this caveat.
+For example, if a study is focussing on the ecological adaptive radiation of a group, ecological traits must be crucial for such disparity analysis.
+Additionally different types of data can be collected to answer different questions: (1) discrete data that can be either describing feature (often referred to as ``cladistic'';e.g.
+the presence or absence of legs, the colour of a tail, etc.) or more simply counts (e.g.
+number of digits, etc.); (2) continuous data that can be either direct measurements (e.g.
+body length, etc.) or more sophisticated methods (e.g.
+geometric morphometrics, mathematical contour description, etc.); (3) or any mix and match of the two former (e.g.
+ratios, etc.).
+Again, in this case, the main question can help choosing the most appropriate data.
+For example, when looking at variation within bats species, landmarks (for geometric morphometrics analysis) or continuous measurements on bones can be approriate.
+On the other hand, when looking at convergence between bats and birds, these homology of these traits is more complex to define and the contour of the wings might more readily answer the question.
+
+Once the type of appropriate data is decided, sample size and sampling issues should be carefully considered next.
+Does one need to include intra-specific variation with multiple specimen per species when possible?
+For example, if you're studying variation of skull in canids, and you decide to include dogs, it is essential to capture their variation by including intra-specific one.
+On the other hand, if you're looking at variation of canids through time since the Eocene, intra-specific variation will not have major impact (up until the Holocene).
+http://www.journals.uchicago.edu/doi/abs/10.1086/650372
+
+Also, when comparing multiple groups, it is important to have fair comparative designs (same number of elements per group, etc.) or apply data corrections (e.g. rarefaction).
+-Taxonomic sampling...
+Depends on the question!
+E.g.
+in birds, if you're looking at the family level, if the variables measured do not distinguish any passerine from another family (i.e.
+all passerine are as dissimilar to a falcon), then taxonomic sampling can be reduced.
+On the other hand, if the variables distinguish within passerines, then it is important to sample more.
+-Sexual dimorphism.
+-Missing/overlapping/inapplicable/ambiguous/correlated/polymorphic data! Additional considerations: homology - cite a morphometrics paper, polymorphisms how to code,
+
+•   Scaling
+-For variables with huge ranges, do z or log transforms so that the variables with big ranges are not skewing the analysis towards these variables.
+
+ -    Biases in datasets
+If you're using other people's data (however, if you're collecting, try to not include biases yourself), you need an awareness of what the data was collected for.
+One major issue is in phylo data.
+-phylogenetic analysis vs disparity + if phylo with a particular aim: where the data collected to discriminate groups early in the tree or within a particular clade (i.e.
+number of synapomorphies vs number of autapomorphies).
+Recommend to select characters for analysis, not use all characters.
+E.g.
+if you;re looking at “feeding” niche, you don't need characters not related to the niche (e.g.
+sexual dimorphic ones).
+•   Talk about Foote's Crinoid non cladistic data example somewhere.
+Important distinction between discrete data collected for phylogeny building and that collected for disparity, i.e.
+one avoids homoplasy, one seeks homoplasies.
+
+Not a one-to-one mapping of form to function
+
+Congruence between data (different parts, different source, different results?).
+
+Finish with: again, think about your questions?
+No data type = best, it depends on Q.
+
+\subsection{Use methods appropriate} (avoid caveats)
+https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
+
+?controlling for phylogeny
+Methods are still immature
+Separate visualisation from analysis
+Size corrections? Allometry?
+
+\section{Where do we go from there?}
+4 - This a list of problems for the future/moving forward
+    Simulations 
+    Form/function
+Isospaces, glottospaces, acoustospace, ecospace, life-history-spaces, etc. All just a multidimensional analysis? Needs for terms?
+    
+``We asked whether we could but not whether we should'' (Jurassic Park)
+
+% \section{Introduction}
+% \label{text:intro}
+
+% Biodiversity is not a smooth gradient of forms, functions, ecologies, behaviours, etc., but it is rather variable and discontinuous. %TG: Actually is it?
+% This is at the base of biological sciences and is linked to the concept of disparity \citep{Wills2001,Hopkins2017} with the observation that some groups of species are more similar or dissimilar than others.
+% This diversity in morphologies is in fact proper in nature and decoupled from taxonomic diversity: some groups may exhibit low diversity of species by high diversity of shapes \citep[or the other way around]{ruta2013,hopkinsdecoupling2013}.
+% In palaeobiology, this concept, has been studied since the 90s describing \citep{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution} and as been extended with modifications to macroevolutionary analysis in general since the 2000s (whether based on continuous traits; e.g.
 \citealt{Harmon961,geiger2008}; or on geometric morphometrics; e.g.
@@
 
-Since then, morphological disparity has been used for testing a vast array of biological hypotheses such as: how did body plans evolve through time?
+% Since then, morphological disparity has been used for testing a vast array of biological hypotheses such as: how did body plans evolve through time?
 \citep{Wesley-Hunt2005}; why does some morphologies exist and not other?
@@
 \citep{price2014niche}; and many more.
-%
-This apparent versatility of morphological disparity analysis led to a phenomenal amount of papers based on disparity methods (Fig. \ref{Fig:GoogleOccurences})!
-%
-However, ``a lot of disparity studies are inductive fishing trips and are not often designed to test a specific hypothesis.''
-%
-Indeed, compared to the vast amount of publication using morphological disparity, only a handful highlight/solve problems and propose hypothesis testing.
+% %
+% This apparent versatility of morphological disparity analysis led to a phenomenal amount of papers based on disparity methods (Fig. \ref{Fig:GoogleOccurences})!
+% %
+% However, ``a lot of disparity studies are inductive fishing trips and are not often designed to test a specific hypothesis.''
+% %
+% Indeed, compared to the vast amount of publication using morphological disparity, only a handful highlight/solve problems and propose hypothesis testing.
 
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=1\textwidth]{Figures/GoogleScholarOccurences.pdf} 
-\caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year.
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=1\textwidth]{Figures/GoogleScholarOccurences.pdf} 
+% \caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year.
 In black, the match is in the paper and in grey, in the title.
@@
 The data was collected on the 1st of November 2017.}
-\label{Fig:GoogleOccurences}
-\end{figure}
+% \label{Fig:GoogleOccurences}
+% \end{figure}
 
-\section{What actually \textit{is} disparity?}
+% \section{What actually \textit{is} disparity?}
 
-\cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.
-Disparity can describe either the metric (\citealt{Wills2001}; or index \citealt{Hopkins2017}) or the whole pipeline \citep{zelditch2012geometric,lloyd2016estimating}.
-To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
-For the 230 relevant papers among the 500 matches, we collected the following methodological data: (1) What was the focal biological group?
+% \cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.
+% Disparity can describe either the metric (\citealt{Wills2001}; or index \citealt{Hopkins2017}) or the whole pipeline \citep{zelditch2012geometric,lloyd2016estimating}.
+% To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
+% For the 230 relevant papers among the 500 matches, we collected the following methodological data: (1) What was the focal biological group?
 ; (2) What kind of data was measured (e.g.
@@
 ; (7) Was phylogeny taken into account or not?
-We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g.
+% We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g.
 a disparity metric was measured but not described anywhere in the paper).
-The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).
-To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).
+% The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).
+% To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).
 
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
-\caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
-\label{Fig:MethodsProportions}
-\end{figure}
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
+% \caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
+% \label{Fig:MethodsProportions}
+% \end{figure}
 
-From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.
+% From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.
 
-\begin{itemize}
-    \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace.
+% \begin{itemize}
+%     \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace.
 Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g.
@@
 
-    \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA).
+%     \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA).
 Disparity is often measured as a distance metric (e.g.
@@
 
-    \item \textbf{The ``\texttt{approach}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance).
+%     \item \textbf{The ``\texttt{approach}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance).
 Hypothesis of higher/lower disparity can then be measured using null evolutionary models.
-\end{itemize}
+% \end{itemize}
 
-%
-The advent of these three approaches coincides with the explosion of disparity analysis (Fig. \ref{Fig:GoogleOccurences}) and suggest a high popularity of disparity analysis.
-%
-Unfortunately, however, few actual tests have been performed on whether any of these approaches \textit{actually} allows to tackle the classic disparity questions (see \ref{text:intro}{Introduction}).
-Are all these methods the right tools to answer these questions are they mere ``inductive fishing trips'' used to describe the data rather than testing any of the hypothesis related to the disparity questions?
+% %
+% The advent of these three approaches coincides with the explosion of disparity analysis (Fig. \ref{Fig:GoogleOccurences}) and suggest a high popularity of disparity analysis.
+% %
+% Unfortunately, however, few actual tests have been performed on whether any of these approaches \textit{actually} allows to tackle the classic disparity questions (see \ref{text:intro}{Introduction}).
+% Are all these methods the right tools to answer these questions are they mere ``inductive fishing trips'' used to describe the data rather than testing any of the hypothesis related to the disparity questions?
 
-\section{Does data used in disparity analysis allows to test hypothesis?}
-In fact, the question of which data to analyse \citep{hetherington2015cladistic} and which part of the data to use has been analysed \citep{hopkins2017well} has rarely been assessed.
-Can the data used in disparity studies actually test the phenomenon researchers are trying to capture?
+% \section{Does data used in disparity analysis allows to test hypothesis?}
+% In fact, the question of which data to analyse \citep{hetherington2015cladistic} and which part of the data to use has been analysed \citep{hopkins2017well} has rarely been assessed.
+% Can the data used in disparity studies actually test the phenomenon researchers are trying to capture?
 
-For example, when studying discrete morphological characters (often following the ``\texttt{Claddis}'' approach), the data used is often recycled data used for phylogenetic analysis.
-These datasets are usually design to recover evolutionary history and differentiate clades and, often due to the preservations of the fossil record, are mainly based on hard tissue characters.
-In phylogenetics, this artefact have been shown to be leading to measurable topological recovery error (for soft \textit{vs} hard characters; \citealt{sansomfossilization2013}; or dental \textit{vs} cranial characters \citealt{sansom2017dental}).
-Therefore, is the morphospace composed of ``fossilisable'' characters enough for answering evolutionary hypothesis?
-For example, are niche replacement hypothesis equivalent to fossilisable-character-based-niches replacement hypothesis?
+% For example, when studying discrete morphological characters (often following the ``\texttt{Claddis}'' approach), the data used is often recycled data used for phylogenetic analysis.
+% These datasets are usually design to recover evolutionary history and differentiate clades and, often due to the preservations of the fossil record, are mainly based on hard tissue characters.
+% In phylogenetics, this artefact have been shown to be leading to measurable topological recovery error (for soft \textit{vs} hard characters; \citealt{sansomfossilization2013}; or dental \textit{vs} cranial characters \citealt{sansom2017dental}).
+% Therefore, is the morphospace composed of ``fossilisable'' characters enough for answering evolutionary hypothesis?
+% For example, are niche replacement hypothesis equivalent to fossilisable-character-based-niches replacement hypothesis?
 
-Similarly, many studies in geometric morphometric are based solely on skull shape using landmark techniques combined with Procrustes sumperimposition \citep[the ``\texttt{geomorph}'' approach;][]{zelditch2012geometric,adams2017geometric}.
-However, it has been shown that the whole skeleton displays different integrated modules with different specific evolutionary constraints leading to variable rates and modes of evolution (\citealt{Goswami20130254}; and this even at the cranial level \citealt{goswami2010influence}).
-Is skull variation enough for measuring body plans variation through time for example?
+% Similarly, many studies in geometric morphometric are based solely on skull shape using landmark techniques combined with Procrustes superimposition \citep[the ``\texttt{geomorph}'' approach;][]{zelditch2012geometric,adams2017geometric}.
+% However, it has been shown that the whole skeleton displays different integrated modules with different specific evolutionary constraints leading to variable rates and modes of evolution (\citealt{Goswami20130254}; and this even at the cranial level \citealt{goswami2010influence}).
+% Is skull variation enough for measuring body plans variation through time for example?
 
-In the case where disparity is measured a single continuous trait (or a collection of traits; following the ``\texttt{geiger}'' approach), for example in \cite{price2014niche}, beak and limb length as well as body mass are used to differentiate groups of birds and different zones in the morphospace can then be interpreted as different niches.
-Is this set of traits studied in isolation sufficient to test niche occupancy hypothesis?
+% In the case where disparity is measured a single continuous trait (or a collection of traits; following the ``\texttt{geiger}'' approach), for example in \cite{price2014niche}, beak and limb length as well as body mass are used to differentiate groups of birds and different zones in the morphospace can then be interpreted as different niches.
+% Is this set of traits studied in isolation sufficient to test niche occupancy hypothesis?
 
-Additionally, in many cases, the disparity metrics and hypothesis are often based on a transformation of the data (pairwise distances, PCA, PCO, MDS, RWA, etc.).
-Can these transformation readily reflect the raw data in our hypothesis?
-For examples a morphospace as a pairwise matrix reflects the dissimilarity between the data, a morphospace based on an ordination reflects the variability within the data, a morphospace based on both reflects the variability within the dissimilarity, are each method reflecting that?
+% Additionally, in many cases, the disparity metrics and hypothesis are often based on a transformation of the data (pairwise distances, PCA, PCO, MDS, RWA, etc.).
+% Can these transformation readily reflect the raw data in our hypothesis?
+% For examples a morphospace as a pairwise matrix reflects the dissimilarity between the data, a morphospace based on an ordination reflects the variability within the data, a morphospace based on both reflects the variability within the dissimilarity, are each method reflecting that?
 
-\section{Do the disparity metrics (or indices) really reflect variation in the studied phenomenon?}
-From the 230 papers analysed, we found 104 unique combination's of metrics to measure disparity (Fig. \ref{Fig:MethodsProportions}).
-Commonly, these metrics are used to approximate the volume occupancy within the morphospace \citep{Wills2001,DonohueDim,diaz2016global}.
+% \section{Do the disparity metrics (or indices) really reflect variation in the studied phenomenon?}
+% From the 230 papers analysed, we found 104 unique combination's of metrics to measure disparity (Fig. \ref{Fig:MethodsProportions}).
+% Commonly, these metrics are used to approximate the volume occupancy within the morphospace \citep{Wills2001,DonohueDim,diaz2016global}.
 
-However, it has been shown that different metrics reacts differently depending on the dataset
+% However, it has been shown that different metrics reacts differently depending on the dataset \citep[][; and an anecdotal example in Fig. \ref{Fig:DimensionsEffect}]{Wills2001,Ciampaglio2001}.
+% For example metrics involving a product of aspects of the matrix (e.g.
+the volume, the product of the dimensions) are likely to suffer from the \textit{Curse of dimensionality} \citep{bellman1957dynamic} and will quickly tend to null values.
+% Some other metrics are not representing all the data captured in the morphospace (e.g. the sum or product of variance ignores the co-variance between dimensions).
+% Are these metrics as a description of the ordination a good proxy for the observed biological phenomenon?
 
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=1\textwidth]{Figures/DimensionsEffect.pdf} 
-\caption{Effect of the number of dimensions on different disparity metrics (note that the sum of variance is equal to the average squared pairwise distance in this case).}
-\label{Fig:DimensionsEffect}
-\end{figure}
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=1\textwidth]{Figures/DimensionsEffect.pdf} 
+% \caption{Effect of the number of dimensions on different disparity metrics (note that the sum of variance is equal to the average squared pairwise distance in this case).
+% The data used is one of the example datasets from the \texttt{dispRity R} package \citep[\texttt{BeckLee\_mat50};][]{beckancient2014,dispRityv02} composed of 50 elements and 48 dimensions obtained from a ordination (MDS) of the distance matrix of a discrete morphological character matrix.}
+% \label{Fig:DimensionsEffect}
+% \end{figure}
 
-Product of variance
+% Despite these caveats, it is important to actually know what the metrics are measuring.
+% There has been some attempts to test that through simulations \citep{Ciampaglio2001,  gerber2017geometry} but it is still unclear which metric reflects the best the hypothesis test.
+% For example, the average squared pairwise distance (or the sum of ranges) are really popular \citep[e.g.][]{geiger2008} yet has been shown to be fairly senstive, at least in empirical data to the number of sampled taxa, the number of characters and the data completeness \citep[][; although effects were lesser in simulations]{Ciampaglio2001}.
+% More fundamentally, the question arises whether one disparity metric could be sufficient to describe changes in disparity?
+% Changes in multidimensions might not be readily grasped by unidimensional metrics!
+% For example, a given taxonomic group might have the exact same volume as another one yet be in a complete different part of the morphospace \citep[or even overlap or not depending on the dimensions;][]{davis2012acanthodes,brusatte50}.
+% In this case, it might more interesting to have multiple metrics (e.g.
+the position and the volume) [FABRE et.
+al in prep.] or to use distributions of metrics (e.g.
+what does the average distance between species or from a point really convey?
+Is the distribution not more interesting?).
 
-Is disparity only relative?
+% Maybe we need metrics that fit explicitly our hypothesis (or fit our hypothesis directly to the metric).
+% For example, for competitive replacement, should we not measure whether group A occupies the same place as group B through time and whether that is due to competition or simply because any other place in the space cannot be occupied (and why are such spaces unoccupied)?
 
-Do the disparity metrics really reflect the variation in the hypothesis that are tested?
-E.g. Are the variance/range based metrics a description of the ordination or the observed biological phenomenon (i.e. they are not including covariance and phylogenetic signal)
-E.g. What does the average distance between species or from a point really convey? Is the distribution not more interesting?
-Do we need more like metrics measuring explicitly what is tested in the hypothesis (rather than approximate it).
-For example, for competitive replacement, should we not measure whether group A occupies the same place as group B through time and whether that is due to competition or simply because any other place in the space cannot be occupied (and why are such spaces unoccupied)?
+% \section{Is the current statistical toolkit sufficient for testing disparity hypothesis?}
+% Are the methods used to test hypothesis in disparity actually testing what we think they're testing (problem of multidimensional hypothesis)
+% What does an anova or related variance based test real infirm/confirms regarding the disparity questions?
+% How do we deal with bootstrapped data and pseudo-replications?
+% Can we realistically use univariate Gaussian models to test hypothesis based on multivariate, probably non-Gaussian data? \citep{blomberg2017beyond}
 
-\section{Is the current statistical toolkit sufficient for testing disparity hypothesis?}
-Are the methods used to test hypothesis in disparity actually testing what we think they’re testing (problem of multidimensional hypothesis)
-What does an anova or related variance based test real infirms/confirms regarding the disparity questions?
-How do we deal with bootstrapped data and pseudo-replications?
-Can we realistically use univariate gaussian models to test hypothesis based on multivariate, probably non-gaussian data?
+% Should be take phylogeny into account? \citep{polly2013phylogenetic}
 
-\section{What are we missing?}
+% \section{What are we missing?}
 
-\subsection{Learning lessons from other fields}
-In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.
+% \subsection{Learning lessons from other fields}
+% In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.
 
-\section{Conclusion}
-We need a more hypothesis driven way of analysing disparity.
-Disparity methods should be the tool for answering disparity hypothesis, not merely a description of the data.
+% \section{Conclusion}
+% We need a more hypothesis driven way of analysing disparity.
+% Disparity methods should be the tool for answering disparity hypothesis, not merely a description of the data.
 

### 2018-01-25 Writing/disparity_review.tex
Data part cleaned up

@@
 \noindent {\small \it 
-$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom; $^2$NHM; $^10$Bristol}
+$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.; $^2$NHM; $^10$Bristol}
 \end{center}
@@
 
-\subsubsection{Disparity patterns or processes}
+\subsubsection{[Box]Disparity patterns or processes}
 At the core of defining what \textit{is} disparity lies a divide between episthemology and ontology.
@@
 For example, if a study is focussing on the ecological adaptive radiation of a group, ecological traits must be crucial for such disparity analysis.
-Additionally different types of data can be collected to answer different questions: (1) discrete data that can be either describing feature (often referred to as ``cladistic'';e.g.
+
+\subsubsection{Which type of data}
+Additionally different types of data can be collected to answer different questions:
+(1) discrete data that can be either describing feature (often referred to as ``cladistic'';e.g.
 the presence or absence of legs, the colour of a tail, etc.) or more simply counts (e.g.
-number of digits, etc.); (2) continuous data that can be either direct measurements (e.g.
-body length, etc.) or more sophisticated methods (e.g.
-geometric morphometrics, mathematical contour description, etc.); (3) or any mix and match of the two former (e.g.
-ratios, etc.).
-Again, in this case, the main question can help choosing the most appropriate data.
+number of digits, etc.);
+(2) continuous data that can be either direct measurements (e.g. body length, etc.) or more sophisticated methods (e.g. geometric morphometrics, mathematical contour description, etc.);
+(3) or any mix and match of the two former (e.g. ratios, etc.).
+In this case, the main question can help choosing the most appropriate data.
 For example, when looking at variation within bats species, landmarks (for geometric morphometrics analysis) or continuous measurements on bones can be approriate.
@@
 On the other hand, if you're looking at variation of canids through time since the Eocene, intra-specific variation will not have major impact (up until the Holocene).
-http://www.journals.uchicago.edu/doi/abs/10.1086/650372
+%http://www.journals.uchicago.edu/doi/abs/10.1086/650372
 
-Also, when comparing multiple groups, it is important to have fair comparative designs (same number of elements per group, etc.) or apply data corrections (e.g. rarefaction).
--Taxonomic sampling...
-Depends on the question!
-E.g.
-in birds, if you're looking at the family level, if the variables measured do not distinguish any passerine from another family (i.e.
+Another aspect to consider regarding the amount of data to collect is to make sure it will allow to be used fairly for subsequent statistical analysis.
+For example, if the question is to compare disparity among groups, collect a similar number of elements per groups or, when not possible (e.g. palaeontological data), use rarefied data.
+Also, are the elements consistent and comparable them?
+Does the studied group present clear sexual dimorphism?
+Are they collected at a relevant taxonomic scope?
+For example in birds, if you're looking at the family level, if the variables measured do not distinguish any passerine from another family (i.e.
 all passerine are as dissimilar to a falcon), then taxonomic sampling can be reduced.
 On the other hand, if the variables distinguish within passerines, then it is important to sample more.
--Sexual dimorphism.
--Missing/overlapping/inapplicable/ambiguous/correlated/polymorphic data! Additional considerations: homology - cite a morphometrics paper, polymorphisms how to code,
 
-•   Scaling
--For variables with huge ranges, do z or log transforms so that the variables with big ranges are not skewing the analysis towards these variables.
+Finally, it is important to remember that morphological data can be patchy.
+Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
+Each of these might be a problem in specific cases.
+For example, hierarchical data can be problematic when comparing groups: if the question is about the disparity in locomotory modes among vertebrates, data can be collected from limbs but will then not be applicable to non-tetrapods or some specific tetrapods (e.g.
+snakes).
+This problem can be linked to whether only homologuous characters should be used or not.
 
- -    Biases in datasets
-If you're using other people's data (however, if you're collecting, try to not include biases yourself), you need an awareness of what the data was collected for.
-One major issue is in phylo data.
--phylogenetic analysis vs disparity + if phylo with a particular aim: where the data collected to discriminate groups early in the tree or within a particular clade (i.e.
-number of synapomorphies vs number of autapomorphies).
-Recommend to select characters for analysis, not use all characters.
-E.g.
-if you;re looking at “feeding” niche, you don't need characters not related to the niche (e.g.
-sexual dimorphic ones).
-•   Talk about Foote's Crinoid non cladistic data example somewhere.
-Important distinction between discrete data collected for phylogeny building and that collected for disparity, i.e.
-one avoids homoplasy, one seeks homoplasies.
+\subsubsection{Recycling data}
+Another important type of data is... other studies' data!
+In fact, it is a common (and good) practice to recycle data from previous studies whether published by the same author or not.
+It is then important to remember where this data comes from and why it was collected.
+For example, when using discrete morphological data used for phylogenetic analysis, it is important to be aware that disparity in such dataset can be maximised depending on the coded characters.
+It has actually been argued [by Foote - crinoids non-cladistic paper] that 
+Some discrete morphological datasets for phylogenetic uses can be designed to minimise homoplasy to obtain well resolved trees (thus increasing disparity among taxa).
+Conversely some discrete morphological datasets for the same group can be designed to maximise homoplasy to study the similarity among groups [cite Foote's crinoids paper].
 
-Not a one-to-one mapping of form to function
+One relatively simple way to control for this specific bias can be to not use all the available data but select within the dataset the most pertinent characters for the question.
+For example, in a study on disparity of dietary niches, removing non-dietary characters (e.g. sexually dimorphic ones) can be a good approach.
+However, it is important to keep in mind that there is no one-to-one mapping between form a function and that some characters can be linked to multiple functions (e.g.
+limb traits can be both linked to locomotion and diet!).
 
-Congruence between data (different parts, different source, different results?).
+\subsubsection{[Box]Data check list}
 
-Finish with: again, think about your questions?
-No data type = best, it depends on Q.
+When collecting data:
+\begin{itemize}
+    \item Which type of data?
+    \item Which traits?
+    \item Which sample size? (or correction)
+    \item Which taxonomic level?
+    \item Are their data caveats (missing, hierarchical, non-homologuous)?
+\end{itemize}
 
-\subsection{Use methods appropriate} (avoid caveats)
-https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
+When recycling data:
+\begin{itemize}
+    \item What was the data collected for?
+    \item Which traits are relevant to the question?
+    \item Are their data caveats (missing, hierarchical, non-homologuous)?
+\end{itemize}
 
-?controlling for phylogeny
-Methods are still immature
-Separate visualisation from analysis
-Size corrections? Allometry?
+\subsection{Use methods appropriate} 
+One of the major problem in disparity analysis is how to analysis data that is, in a vast majority of cases multidimensional.
+Multidimensional analysis can be a really powerful tool to analysis complex datests but also comes with many mathematical caveats and limitations.
+
+\subsubsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)}
+
+Can we go beyond ordination? Yes. By going before. (Unless data are landmarks?)
+We suggest we don’t automatically use ordinations, when the data and the question allow pre-ordination analysis.
+Why the pipeline 
+**data -> distance matrix -> ordination** and not
+**data -> zscore -> PCA**
+**data -> Procrustes superimposition -> projection into tangent space >- PCA [Melanie has image of pipeline; worthwhile getting Emma/Sylvain to check]
+Helps us plot data (e.g., summary stats could be similar for completely different spaces - see Wills 2001, disparity primer, his figure 18)
+For GM data, PCA reduces dimensionality to reflect number of degrees of freedom (which was reduced during superimposition); no issues with geometry because coordinates put into Euclidean space when projected into tangent space, and this is maintained in the PCA.
+But possible misstep is to rely only on first few axes when interpreting data.
+Note also that most subsequent analyses do not require coordinates to be in Euclidean space (e.g., linear modeling can be done on the Procrustes coordinates); the projection into tangent space is taken in order to facilitate the visualization.
+If you plan to use everything (e.g., in estimating sum of ranges or sum of variances), it is not necessary to do an ordination (calculate X on the raw data).
+Else do an ordination and decide which axis to use based on the metric properties (mathematical and computational) or the scree plot, Horns parallel analysis etc.
+In other methods (e.g.
+discrete data), ordination can be problematic (character hierarchy - i.e.
+inapplicable -, polymorphisms, missing data), particularly non-Euclidean distances have to be distorted into a Euclidean space, so think before using it.
+Best practices:
+What does an ideal morphospace represent?
+High variance on first two axes (relates to visualisation)
+Reduces dimensionality of data
+Axes that make morphological (or other if data not morphological) sense
+Points close in space are morphologically similar
+If you are going to include an ordination, make sure to do the following:
+Include \% variation associated with ordination axes, preferably on axis labels on figures, but at minimum include a scree plot in supplementary materials
+Make sure axes are plotted on same scale (to show variance in proportion)
+Never calculate disparity indices from discriminant analysis/canonical variate scores
+Never assume that because two (or more) groups overlap on a single bivariate plot they overlap in the whole space (or generally that the proximity between two points holds across the whole space; e.g., being close to the centroid on a single bivariate plot does not mean that a point is the closest to the centroid in the whole space)
+By contrast if two (or more) groups are separate on a single axis they are separate in the whole space
+
+\subsection{Phylogeny}
+\subsubsection{Ancestral states estimation}
+Do not use phylogeny just to increase sample size!
+Beware edge effects in time series if using phylogeny to "correct" data, especially if leading into an extinction event
+Disparity time series should be simulated along the sample phylogeny to give a sensible null expectation to assess how much of the recovered pattern is simply reflecting the phylogeny itself and not morphological change
+Pre/post ordination ancestral states debate -> see upcoming litterature.
+* if you filled the morphospace with ancestral estimation, the nearest neighbour distance would go to zero. 
+* ancestral estimation to balance sample sizes.. Is this appropriate?
+Ancestors -> estimate? Include?
+
+\subsubsection{Multivariate Phylogenetic Comparative Methods}
+This has been well covered here: %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
+
+\subsection{Metrics}
+Metrics can give the same results from very different morphospaces. Always plot your data (and see pros and cons of ordinations) and try to use multiple metrics.
+What should we be measuring?
+Volume?
+Position?
+Relative position?
+Density?
+Which metric?
+Using a few metrics in concert can be more informative than trying to make inferences from just one.
+For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution)
+Broad classifications of metrics could be those that relate to distance (mean pairwise, weighted mean pairwise, from centroid, from medioid, nearest neighbour, minimum spanning tree length, variance metrics), volume/occupancy (hypercube/hyperellipsoid/convex hull volume, N hyperquadrats occupied, range metrics) or position (shift of centroid/medioid, distance between medioids).
+Density?
+range: same problems as as max/min/hulls.. there are no unbiased estimators
+distance to centroid same as variance
+product of variances should be proportional to ellipsoid volume
+Should we use distributions or single metrics? (Distributions also relevant to tests and visualisations, see below)
+Can also classify as pre- or post-ordination metrics, although former can not be volume/occupancy(?)
+We are interested in separate questions relating to spread of group in space, or position of group in space (e.g., position can shift without altering volume, or vice versa)
+sums of variance (misses out correlations... this is ok if all PCA transformed, but not ok if subsets of data are then taken and calculated separately)
+
+Best practices:
+We suggest not calling metrics "disparity" (especially in figure labels) but rather "sum of variance", etc...
+
+\subsection{Tests}
+Just two sentences: check out the test assumptions.
+(and make sure your metrics are capturing what you are testing).
+Know if and how your dataset is violating assumptions.
+Check also if we need to talk about geometric morphometrics specific tests?
+Sample size is an issue (type I error inflation).
+Is bootstrapping valid?
+How can we test for differences statistically?
+Bayesian
+What stats are appropriate?
+permutation tests
+variance compared to range
+Mantel tests
+Including fossils is a good idea! (adds time’s arrow, constraints history/ancestral state estimates etc.)
+
+\subsection{Models}
+What are our null hypotheses?
+Joe’s simulations’ stuff - phylogeny can be used to build a null.One basic null is that disparity increases with diversity; how do we diverge from that (show these morpho/genome distance divergence/convergence).
+We can have a null model that is based on simple Brownian increase in disparity (gradual). 
+Null: is the null increase in disparity dependent on a random walk and on a tree shape?
+Maybe we don’t want to think about a null model but rather think about good models to fit. 
+We haven’t explored much about fitting models like explaining holes, effects of extinction etc…
+Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!
+
+\subsection{Visualisations}
+Hard for high-dimensional data (our brains can really only see three axes tops)
+What do our visualisations imply about the data (but we don’t necessarily want it to)?
+Straight lines in phylomorphospaces are particularly misleading (although no easy alternative)
+Spaghetti plots for time series more helpful than collapsing into a single point estimate (plus CI/HPD etc.)
+Can be hard to visualise uncertainty in an ordination space (be it phylogenetic or ancestral states), but this can be important for establishing robustness of our inferences
+Visualisations can be important for understanding what is going on with your data!
+Can we do an Anscombe’s Quartet for disparity?
+(see: %https://en.wikipedia.org/wiki/Anscombe\%27s_quartet; Revell has done this for PCMs: http://blog.phytools.org/2017/10/phylogenetic-anscombe-datasets.html; Wills 2001 has kinda already done this though, redraw his Figure 18?)
+If using ancestral estimates visually differentiate these from the empirical data
+Can be hard to, e.g., label tips in a phylomorphospace (or any ordination space) without cluttering plot - perhaps answer is to have interactive plots where hovering over a point with a cursor tells you what it is
+
+\subsection{Do we need to adapt our methods?}
+Are there lessons/methods from other fields we use?
+Sample sizes: we need to be careful about them! (e.g. for detecting holes etc.).
+Time series: same! Palaeontology data is not readily available (hence ancestral estimates) and might not be appropriate for doing time series analysis (across long time scales at least).
+Caveats of disparity
+Statistical issues e.g. covariance
+95\% or all PCs?
+Phylogenetic correlations
+
+Do a check list of suggestions for standardisation: put your \% variance on the axis (and scale them)!
+Andrew’s list of things
+* should we z-score our data always?
+* volume of a n-dimensional ellipsoid scales with number of dimensions, initially increasing and then decreasing with the shape contingent on the radius R.
+Need good understanding of your data
+Is space Euclidean?
+Visualisation vs. analysis
+Holes in morphospace? How to deal with these?
+What to do with phylogeny/about phylogenetic autocorrelation?
+What about other forms of autocorrelation
+Size correction?
+Avoid using shiny new methods unless you understand
+Avoid being too prescriptive in advice of paper
+Outline assumptions of different approaches--make clear caveats
 

### 2018-01-25 Writing/disparity_review.tex
ordination box

@@
 
-\subsubsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)}
+\subsection{Visualisations}
+\label{visualisation}
+Hard for high-dimensional data (our brains can really only see three axes tops)
+What do our visualisations imply about the data (but we don’t necessarily want it to)?
+Straight lines in phylomorphospaces are particularly misleading (although no easy alternative)
+Spaghetti plots for time series more helpful than collapsing into a single point estimate (plus CI/HPD etc.)
+Can be hard to visualise uncertainty in an ordination space (be it phylogenetic or ancestral states), but this can be important for establishing robustness of our inferences
+Visualisations can be important for understanding what is going on with your data!
+Can we do an Anscombe’s Quartet for disparity?
+(see: %https://en.wikipedia.org/wiki/Anscombe\%27s_quartet; Revell has done this for PCMs: http://blog.phytools.org/2017/10/phylogenetic-anscombe-datasets.html; Wills 2001 has kinda already done this though, redraw his Figure 18?)
+If using ancestral estimates visually differentiate these from the empirical data
+Can be hard to, e.g., label tips in a phylomorphospace (or any ordination space) without cluttering plot - perhaps answer is to have interactive plots where hovering over a point with a cursor tells you what it is
 
-Can we go beyond ordination? Yes. By going before. (Unless data are landmarks?)
-We suggest we don’t automatically use ordinations, when the data and the question allow pre-ordination analysis.
-Why the pipeline 
-**data -> distance matrix -> ordination** and not
-**data -> zscore -> PCA**
-**data -> Procrustes superimposition -> projection into tangent space >- PCA [Melanie has image of pipeline; worthwhile getting Emma/Sylvain to check]
-Helps us plot data (e.g., summary stats could be similar for completely different spaces - see Wills 2001, disparity primer, his figure 18)
-For GM data, PCA reduces dimensionality to reflect number of degrees of freedom (which was reduced during superimposition); no issues with geometry because coordinates put into Euclidean space when projected into tangent space, and this is maintained in the PCA.
-But possible misstep is to rely only on first few axes when interpreting data.
-Note also that most subsequent analyses do not require coordinates to be in Euclidean space (e.g., linear modeling can be done on the Procrustes coordinates); the projection into tangent space is taken in order to facilitate the visualization.
-If you plan to use everything (e.g., in estimating sum of ranges or sum of variances), it is not necessary to do an ordination (calculate X on the raw data).
-Else do an ordination and decide which axis to use based on the metric properties (mathematical and computational) or the scree plot, Horns parallel analysis etc.
-In other methods (e.g.
-discrete data), ordination can be problematic (character hierarchy - i.e.
-inapplicable -, polymorphisms, missing data), particularly non-Euclidean distances have to be distorted into a Euclidean space, so think before using it.
-Best practices:
-What does an ideal morphospace represent?
-High variance on first two axes (relates to visualisation)
-Reduces dimensionality of data
-Axes that make morphological (or other if data not morphological) sense
-Points close in space are morphologically similar
-If you are going to include an ordination, make sure to do the following:
-Include \% variation associated with ordination axes, preferably on axis labels on figures, but at minimum include a scree plot in supplementary materials
-Make sure axes are plotted on same scale (to show variance in proportion)
-Never calculate disparity indices from discriminant analysis/canonical variate scores
-Never assume that because two (or more) groups overlap on a single bivariate plot they overlap in the whole space (or generally that the proximity between two points holds across the whole space; e.g., being close to the centroid on a single bivariate plot does not mean that a point is the closest to the centroid in the whole space)
-By contrast if two (or more) groups are separate on a single axis they are separate in the whole space
+The percentage of variation associated with the ordination axes (preferably directly on the axis labels on a figure but at least in a scree plot).
+Make sure that the axes in the morphospace are plotted on the same scale (to who the proportion of variance)
+
+\subsubsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)?}
+One of the most common approach to analyse multidimensional data is to use ordination techniques.
+This consists in transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
+This can be done through three main analysis pipelines:
+\begin{enumerate}
+    \item data -> distance matrix -> ordination
+    \item data -> zscores -> PCA
+    \item data -> Procrustes superimposition -> projection into tangent space -> PCA
+\end{enumerate}
+%TG: Check with Melanie for the figure? - apparently she has one.
+
+%Why to ordinate:
+These pipelines allow to reduce the dimensionality of the data.
+This is advantageous for plotting and visualising the data (see Visualisation section \ref{visualisation}) which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
+Additionally, ordinating the data can allow to further reduce the number of dimensions by only selecting a number of axis based on the metric properties (mathematical - e.g.
+product base metric - or computational - e.g.
+hull based metric).
+This can be done through the scree plots (cuts off, majority axis, etc.) or other statistical analysis (Horns parallel, etc.).
+
+In the case of geometric morphometric data (point 3 above), ordination can be particularly relevant since it conserves mathematical properties of the data while efficiently reducing the dimensions: the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc.
+This has clear advantages for interpreting the results in a lower number of dimensions.
+For example, the axis will actually represent gradients of biological variation (e.g. elongation and flattening of the beak).
+This also allows easy interpretation of the proximity of the elements in space which reflects shared biological characters.
+
+However, some caveats can rise from such transformation of the data.
+For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g.
+only the few first) can lead to misinterpretation of the results erroneous statistical results.
+Furthermore, interpreting the biological variation along the axis is always a \textit{post-hoc} procedure and might bear little meaning regarding the disparity question (e.g.
+the elongation and flattening of the beak when the disparity question is about flight pattern disparity).
+
+In fact, in many cases, depending on the question, ordination might actually not be necessary.
+For example, if the disparity metric relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate in on ordinated data.
+Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
+For example, when the analysed data is non Euclidean (e.g.
+discrete morphological characters), the ordination can be problematic (character hierarchy - i.e.
+inapplicable -, polymorphisms, missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
+In such case, distances can be inappropriate in a mathematical (and colloquial) sense (e.g. the distance between A and B might differ from the distance between B and A).
+This can be problematic when comparing the position of groups in the multidimensional space since the measured distances might not be comparable among each other.
+Furthermore, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axis can be simply impossible or meaningless biologically (e.g.
+the ratio of data coded as 1 rather than 0).
+
+Finally, visualising ordinated data can be missleading to the human eye (or any other being living in three dimensions only).
+It is incorrect to assume that because some groups overlap on a bivariate (or 3D) plot, they overlap in the whole multidimensional space (especially if the space is not euclidean!).
+Note that conversely, when two groups don't overlap on a two or three dimensional plot, they \textit{do not} overlap in the multidimensional space.
+
+Regarding these advantages and inconvenients of ordination technique, we strongly suggest to not automatically ordinate data in multidimensional analysis and to carefully consider whether the question can be answered without ordination.
 
@@
 \subsection{Metrics}
+\label{metrics}
 Metrics can give the same results from very different morphospaces. Always plot your data (and see pros and cons of ordinations) and try to use multiple metrics.
@@
 Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!
-
-\subsection{Visualisations}
-Hard for high-dimensional data (our brains can really only see three axes tops)
-What do our visualisations imply about the data (but we don’t necessarily want it to)?
-Straight lines in phylomorphospaces are particularly misleading (although no easy alternative)
-Spaghetti plots for time series more helpful than collapsing into a single point estimate (plus CI/HPD etc.)
-Can be hard to visualise uncertainty in an ordination space (be it phylogenetic or ancestral states), but this can be important for establishing robustness of our inferences
-Visualisations can be important for understanding what is going on with your data!
-Can we do an Anscombe’s Quartet for disparity?
-(see: %https://en.wikipedia.org/wiki/Anscombe\%27s_quartet; Revell has done this for PCMs: http://blog.phytools.org/2017/10/phylogenetic-anscombe-datasets.html; Wills 2001 has kinda already done this though, redraw his Figure 18?)
-If using ancestral estimates visually differentiate these from the empirical data
-Can be hard to, e.g., label tips in a phylomorphospace (or any ordination space) without cluttering plot - perhaps answer is to have interactive plots where hovering over a point with a cursor tells you what it is
 

### 2018-02-04 Writing/disparity_review.tex
methods up past ordination

@@
 %2§ what use: commonly used
-The description of these patterns (the disparity of patterns) has thus been used in evolutionary biology to tackle a vast array of questions:
-Developmental constraints [CITE]; 
-Niche width, competition  - ecological perspective / specialisation [CITE];
-Relationship between form and function and evolution functional traits [CITE];
-Disparity through time [CITE];
-Response to extinction - Extinction selectivity [CITE];
-Competition - ecologically relevant [CITE];
-Evolutionary mode [CITE];
-Evolutionary rate [CITE];
+The description of these patterns (the disparity of patterns) has thus been used in evolutionary biology to tackle a vast array of questions.
+Disparity patterns have been used to tackle questions in Evo-devo through
+%evo-devo
+Developmental constraints [CITE],
+Relationship between form and function and evolution functional traits [CITE] or
+Scaling micro to macro [CITE];
+% Ecology
+in ecology with
+Competition - ecologically relevant [CITE],
+Niche width, competition  - ecological perspective / specialisation [CITE] or
 Macroecology (disparity in space)[CITE];
-Tree Structure and the effect on disparity [CITE];
-Adaptive radiation ecological diversification [CITE];
-Scaling micro to macro [CITE];
-Competitive replacements [CITE];
+% Evolution
+on in macroevolution:
+Response to extinction - Extinction selectivity [CITE],
+Competitive replacements [CITE] or
 Tempo and mode of evolution [CITE].
@@
 %3§ problems
-Disparity studies have a great deal promise in helping us to understand the evolution of biological diversity.
-However, we identify three main issues in current disparity studies.
+Disparity studies have a great deal of promise in helping us to understand the evolution of biological diversity.
+With the expansion of such methodological approaches in the last decade, however, we also identify three main issues that, if not tackled could reduce the power of such analysis.
 First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
-Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large cladistic datasets.
+Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets.
 Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
-Here we deal with some of these issues and present best practice guidelines/sets of things to think about before performing a disparity analysis.
+Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions before performing a disparity analysis.
 
@@
 Can we understand something about past environments from disparity??
+\subsubsection{[Box]End}
 
@@
 The last decade have seen a vast array of excellent methodological implementation and documentations rendering disparity analysis easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
-However, this easiness might (conscientiously or not) push authors to perform disparity analysis ``just because''.
-In fact it is important to not rush into a disparity analysis to avoid some basic statistical caveats:
+However, this facilitation might (conscientiously or not) push authors to perform disparity analysis ``just because''.
+It is important to not rush into a disparity analysis to avoid some basic statistical caveats.
 For example, one major thing to consider is circularity: if your aim is to explore the variation in the data set, a multi-dimensional analysis is perfectly valid.
@@
 Additionally different types of data can be collected to answer different questions:
-(1) discrete data that can be either describing feature (often referred to as ``cladistic'';e.g.
+(1) discrete data that can be either describing feature (often referred to as ``cladistic''; e.g.
 the presence or absence of legs, the colour of a tail, etc.) or more simply counts (e.g.
@@
 (2) continuous data that can be either direct measurements (e.g. body length, etc.) or more sophisticated methods (e.g. geometric morphometrics, mathematical contour description, etc.);
-(3) or any mix and match of the two former (e.g. ratios, etc.).
+(3) or any mix and match of the two former (e.g. ratios).
 In this case, the main question can help choosing the most appropriate data.
 For example, when looking at variation within bats species, landmarks (for geometric morphometrics analysis) or continuous measurements on bones can be approriate.
-On the other hand, when looking at convergence between bats and birds, these homology of these traits is more complex to define and the contour of the wings might more readily answer the question.
+On the other hand, when looking at convergence between bats and birds, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.
 
@@
 One of the major problem in disparity analysis is how to analysis data that is, in a vast majority of cases multidimensional.
-Multidimensional analysis can be a really powerful tool to analysis complex datests but also comes with many mathematical caveats and limitations.
+Multidimensional analysis can be a really powerful tool to analysis complex datasets but also comes with many mathematical caveats and limitations.
+As any great power, it comes with great responsibilities.
 
-\subsection{Visualisations}
+\subsection{Visualising the multidimensional problem}
 \label{visualisation}
-Hard for high-dimensional data (our brains can really only see three axes tops)
-What do our visualisations imply about the data (but we don’t necessarily want it to)?
-Straight lines in phylomorphospaces are particularly misleading (although no easy alternative)
-Spaghetti plots for time series more helpful than collapsing into a single point estimate (plus CI/HPD etc.)
-Can be hard to visualise uncertainty in an ordination space (be it phylogenetic or ancestral states), but this can be important for establishing robustness of our inferences
-Visualisations can be important for understanding what is going on with your data!
-Can we do an Anscombe’s Quartet for disparity?
+One really common first step in multidimensional analysis is to reduce the number of dimensions.
+Whether this entails an ordination or not (see box \ref{box_ordination}) it is often good practice to first visualise the data.
+However, the physical reality we're living in allows a maximum of three dimensions to be visualised.
+
+First, the interpretation of the data visualisation can be misleading:
+Classically, it is incorrect to assume that because some groups overlap on a bivariate (or 3D) plot, they overlap in the whole multidimensional space (especially if the space is not euclidean!).
+Note that conversely, when two groups don't overlap on a two or three dimensional plot, they \textit{do not} overlap in the multidimensional space.
+Similarly, straight lines in a 2/3D representation of a space can be misleading, depending on the geometrical properties of the space, they could not be interpreted in our intuitive Euclidean distance way.
+For example, the distance between two points A and B might not necessary be equal to the distance between B and A if the space is not Euclidean.
+
+Second, because of the complexity of multidimensional spaces, it is always important to be entirely clear about what is represented in their visualisation.
+In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a spatial pattern if not all the key properties of the space are present in the graph.
+We believe that these should always include the following:
+(1) the percentage of the variation of each axis of the plot (i.e. is the majority - or all - the variation of the data present in the plot or only a fraction?);
+(2) the scale of the axis (ideally, if the space is orthogonal, it is important to keep the same scale between the axis);
+and (3) the error associated with each point, whether this is in term of measurement error (on ordination plots for example) or in standard variation (in further analysis, e.g.
+disparity through time).
+
+Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worth, it won't be what it seems.
+
+% Can we do an Anscombe’s Quartet for disparity?
 (see: %https://en.wikipedia.org/wiki/Anscombe\%27s_quartet; Revell has done this for PCMs: http://blog.phytools.org/2017/10/phylogenetic-anscombe-datasets.html; Wills 2001 has kinda already done this though, redraw his Figure 18?)
-If using ancestral estimates visually differentiate these from the empirical data
-Can be hard to, e.g., label tips in a phylomorphospace (or any ordination space) without cluttering plot - perhaps answer is to have interactive plots where hovering over a point with a cursor tells you what it is
-
-The percentage of variation associated with the ordination axes (preferably directly on the axis labels on a figure but at least in a scree plot).
-Make sure that the axes in the morphospace are plotted on the same scale (to who the proportion of variance)
 
 \subsubsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)?}
+\label{box_ordination}
 One of the most common approach to analyse multidimensional data is to use ordination techniques.
@@
 
-Finally, visualising ordinated data can be missleading to the human eye (or any other being living in three dimensions only).
-It is incorrect to assume that because some groups overlap on a bivariate (or 3D) plot, they overlap in the whole multidimensional space (especially if the space is not euclidean!).
-Note that conversely, when two groups don't overlap on a two or three dimensional plot, they \textit{do not} overlap in the multidimensional space.
-
-Regarding these advantages and inconvenients of ordination technique, we strongly suggest to not automatically ordinate data in multidimensional analysis and to carefully consider whether the question can be answered without ordination.
+Regarding these advantages and inconvenient of ordination technique, we strongly suggest to not automatically ordinate data in multidimensional analysis and to carefully consider whether the question can be answered without ordination.
+\subsubsection{[Box] end}
 

### 2018-02-09 Writing/disparity_review.tex
slow updates

@@
 \subsection{Phylogeny}
+For many disparity analysis looking at macroevolutionary questions, phylogeny is often taken into account, whether for including estimates of ancestors or to add a phylogenetic structure to the data.
+For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
+and will thus not covered further in this review.
+
 \subsubsection{Ancestral states estimation}
-Do not use phylogeny just to increase sample size!
-Beware edge effects in time series if using phylogeny to "correct" data, especially if leading into an extinction event
-Disparity time series should be simulated along the sample phylogeny to give a sensible null expectation to assess how much of the recovered pattern is simply reflecting the phylogeny itself and not morphological change
-Pre/post ordination ancestral states debate -> see upcoming litterature.
-* if you filled the morphospace with ancestral estimation, the nearest neighbour distance would go to zero. 
-* ancestral estimation to balance sample sizes.. Is this appropriate?
-Ancestors -> estimate? Include?
+When looking at disparity in the past (e.g. disparity-through-time) ancestral states estimations are a common approach.
+Ancestral states estimations can be used in two ways:
+(1) either before transformation of the data (e.g. ordination, distance matrix, etc.) by simply estimating the ancestral states based on the original dataset and adding them to it [CITE].
+(2) else after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
+The first method is generally thought to be more thorough but can imply longer computation and will change the ordinated space configuration.
+The second is much more faster and does not change the ordinated space configuration (i.e.
+the ancestors can be seen as averages between tips in the space) but it will change the properties of the space (variance and degrees of freedom).
+MORE UPCOMING WORK ON THAT
 
-\subsubsection{Multivariate Phylogenetic Comparative Methods}
-This has been well covered here: %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
+More importantly however, one needs to keep in mind that using such methods can come with several caveats.
+First and foremost, although ancestral states estimations are valid methods in macroevolution, they results are highly dependent on the data and chosen methods.
+Secondly, more specific to disparity analysis, when using post-transformation ancestral states estimation, the morphospace can be saturated leading pairwise distance bases metrics to tend to zero.
+In general, using ancestral states estimation can bring much insights for recovering patterns of changes in disparity but should not be ``used'' simply to generate extra data points for statitstical reasons.
+In fact, these extra points are not independent and can also generate side effects which can be problematic, especially when testing the effect of extinctions events on disparity.
+
+Alternatively, ancestral states estimation can be used to provide realistic null expectations from the data.
+In disparity through time analysis, it is possible to sample many ancestral states estimates on the nodes and along to branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than the morphological changes.
+Furthermore, this method can be pertinent for fitting modes of evolution and see if the recovered disparity metric can be explained by the tested modes of evolution.
 
@@
 \label{metrics}
+
+[INCLUDE WILLS FIGURE HERE]
+
 Metrics can give the same results from very different morphospaces. Always plot your data (and see pros and cons of ordinations) and try to use multiple metrics.
@@
 Best practices:
-We suggest not calling metrics "disparity" (especially in figure labels) but rather "sum of variance", etc...
+Which ever metric is most suited to the disparity question and regarding the amount of different metrics available we would like to emphasise that it is crucial to define clearly the metric rather than calling it simply ``disparity'' but rather ``sum of variances'', etc.
+(especially in figure labels).
 

### 2018-02-15 Writing/disparity_review.tex
Metrics part worked on

@@
 
-Metrics can give the same results from very different morphospaces. Always plot your data (and see pros and cons of ordinations) and try to use multiple metrics.
-What should we be measuring?
-Volume?
-Position?
-Relative position?
-Density?
-Which metric?
-Using a few metrics in concert can be more informative than trying to make inferences from just one.
-For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution)
-Broad classifications of metrics could be those that relate to distance (mean pairwise, weighted mean pairwise, from centroid, from medioid, nearest neighbour, minimum spanning tree length, variance metrics), volume/occupancy (hypercube/hyperellipsoid/convex hull volume, N hyperquadrats occupied, range metrics) or position (shift of centroid/medioid, distance between medioids).
-Density?
-range: same problems as as max/min/hulls.. there are no unbiased estimators
-distance to centroid same as variance
-product of variances should be proportional to ellipsoid volume
-Should we use distributions or single metrics? (Distributions also relevant to tests and visualisations, see below)
-Can also classify as pre- or post-ordination metrics, although former can not be volume/occupancy(?)
-We are interested in separate questions relating to spread of group in space, or position of group in space (e.g., position can shift without altering volume, or vice versa)
-sums of variance (misses out correlations... this is ok if all PCA transformed, but not ok if subsets of data are then taken and calculated separately)
+Disparity metrics are essentially a one dimensional summary of a multidimensional object, hence, they are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
+With that in mind, the choice of the metric should be essential to summarise the aspect of the morphospace that is the most related to the question.
+In an unidimensional distributions (e.g. histograms) different metrics will readily reflect different aspect of the data.
+For example, metrics can be used to reflect the ``size'' of the distribution (e.g.
+the range, quantiles or variance) and some other will reflect the most likely value (the central tendency - e.g.
+the mean, median or mode).
+Among these metrics, some will have more preferable properties than other such as sensitivity to outliers (range, mean or mode \textit{vs.} quantiles, variance or median) and will thus be more or less appropriate to the different questions.
+
+In multidimensionality, this high availability of unidimensional descriptors of the space are really numerous and come with their equally numerous advantages and caveats.
+In this case, the disparity question should again drive the choice of the disparity metric.
+If the question refers to ``size'' of a group in the morphospace (e.g.
+``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriated (e.g.
+volume [CITE], variance [CITE], range [CITE], distance from the centroid [CITE]).
+Conversely, if the question refers to the ``position'' of a group in the morphospace (e.g.
+``does group A occupy the same space as group B?''), metrics relating to the distance of the elements to a point are most appropriated (e.g.
+distance from a fixed point - e.g.
+the opposite group).
+Finally, one might also be interested in the density of elements in some regions of the morphospace (e.g.
+``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriated (e.g.
+nearest neighbour distance, pairwise distances, etc.)
+
+When looking at this positively, any metric can be suitable to any question, when looking at it negatively, it is important to keep in mind several caveats that can apply in any case:
+First, it is important to relate the metrics to the space property.
+For example, measuring the sum of variances of the space's dimensions does not require the co-variance element between the axis in a PCA but only when the full space is considered: if only a subset is measured, co-variance should be added to the sum of variances.
+Second, some metrics automatically relate to the properties of the morphospace.
+For example, range or hull based metrics are really sensitive to outliers and thus biased based on the sample size.
+
+Of course some other aspects could be of interest (or more than one) and thus some other metrics would be more appropriated.
+For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution).
+Furthermore, one future avenue that could be explored is to go beyond single valued disparity metrics and rather express them in terms of distributions.
+For example, when the metric is expressed as the median distance from centroids [e.g. in CITE], it could be informative as well to express it as a full distribution.
+This would allow to both visualise the variation at a glance and use this variation in statistical tests for comparing these distributions.
 
 Best practices:
-Which ever metric is most suited to the disparity question and regarding the amount of different metrics available we would like to emphasise that it is crucial to define clearly the metric rather than calling it simply ``disparity'' but rather ``sum of variances'', etc.
-(especially in figure labels).
+\begin{itemize}
+\item Facing the amount of different ways to measure disparity and regardless which metric is the most suited to the question, it is crucial to clearly and mathematically define the metric in a disparity paper.
+The definition can be in plain English: e.g.
+``we measure disparity as the sum of the variance of each axis of the ordinated matrix''; or mathematically: e.g.
+``we measured disparity as $\sum\sigma_i$ where $\sigma_i$ is the variance of the each ordinated axis $i$.''.
+\item It is important to keep in mind that different metrics can lead to the same results even though the described spaces are different (the opposite is \textit{not} true).
+We therefore suggest to always plot some dimensions of the space (see Visualisation section \ref{visualisation}) and try multiple metrics.
+\end{itemize}
 
@@
 
-\subsection{Models}
+\subsubsection{Models}
 What are our null hypotheses?

### 2018-02-16 Writing/disparity_review.tex
added test part

@@
 \noindent {\small \it 
-$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.; $^2$NHM; $^10$Bristol}
+$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.; $^2$NHM; $^{10}$Bristol}
 \end{center}
@@
 \subsection{Tests}
-Just two sentences: check out the test assumptions.
-(and make sure your metrics are capturing what you are testing).
-Know if and how your dataset is violating assumptions.
-Check also if we need to talk about geometric morphometrics specific tests?
-Sample size is an issue (type I error inflation).
-Is bootstrapping valid?
-How can we test for differences statistically?
-Bayesian
-What stats are appropriate?
-permutation tests
-variance compared to range
-Mantel tests
-Including fossils is a good idea! (adds time’s arrow, constraints history/ancestral state estimates etc.)
+Similarly to all the points covered above, which statistical test to apply solely depends on which question is tackled with disparity analysis (i.e. which hypothesis to test).
+On the other hand, some tests (such as permanovas [CITE]) are actually \textit{not} taking into account the disparity measured but are simply based on the distribution of the taxa into the space (or simply their pairwise distances in the case of PERMANOVAS).
+
+Furthermore, bootstraps are often used in morphological disparity analysis as a mean to get a distribution from a single disparity value [CITE].
+While this method is valid in itself, it comes with several caveats.
+First, it is important to consider that bootstrapping the disparity values creates a distribution of them (e.g. a distribution of sum of variances from a bootstrapped morphospace).
+Therefore, the error associated with the bootstrap is reflecting the ``stability'' of the metric, not the extremes of the distribution of the data (\textit{c.f.} when bootstrapping a disparity distribution; e.g.
+the distribution of all the distances from centroids from a bootstrapped morphospace).
+In the first and most common case, the amount of bootstraps artificially increases the type I error.
+Furthermore, when comparing two groups, the comparison is one of the disparity of two groups using their ``stability'' as variance.
+
+Best practices:
+\begin{itemize}
+\item make sure your metrics are capturing what you are testing
+\item check out the test assumptions
+\item Know if and how your dataset is violating assumptions.
+\end{itemize}
 

### 2018-02-20 Writing/disparity_review.tex
Contributions?

@@
 
-\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Natalie Cooper$^2$, Stephen Brusatte$^3$, Katie Davis$^4$, Andrew Jackson$^5$, Sylvain Gerber$^6$, Anjali Goswami$^2$, Kevin Healy$^7$, Melanie Hopkins$^8$, Graeme Lloyd$^9$, Joseph O'Reilly$^{10}$, Abi Pate$^{10}$, Emilie Rayfield$^{10}$, Erin Saupe$^{11}$, Emma Sherratt$^{12}$, Graham Slater$^{13}$, Gavin H Thomas$^{14}$ and Philip Donoghue$^{10}$}\\
+\noindent {\normalsize \sc Thomas Guillerme$^{1,*,+}$, Natalie Cooper$^{2,+}$, Stephen Brusatte$^3$, Katie Davis$^4$, Andrew Jackson$^5$, Sylvain Gerber$^6$, Anjali Goswami$^2$, Kevin Healy$^7$, Melanie Hopkins$^8$, Graeme Lloyd$^9$, Joseph O'Reilly$^{10}$, Abi Pate$^{10}$, Emilie Rayfield$^{10}$, Erin Saupe$^{11}$, Emma Sherratt$^{12}$, Graham Slater$^{13}$, Gavin H Thomas$^{14}$ and Philip Donoghue$^{10,+}$}\\
 
@@
 \medskip
+\noindent{+ These authors contributed equally to the manuscript.}
 \noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
@@
 
+\section{Authors contribution}
+TG, NC and PD proposed the idea of this review. TG, NC, SB, KD, AJ, SG, AG, KH, MH, GL, JOR, AP, ER, ES, ES, GS, GHT and PD wrote the manuscript,
+
 \section{Acknowledgments}

### 2018-02-26 Analysis/disparity_methods.Rmd
Intro checked

@@
 for(factor in c(2,4,5,6)) {
-    plot.data(fuzzy.sort.data(data[,factor], factors[[factor]]), proportion = TRUE, ylim = c(0, 100), main = names(factors)[[factor]])
+    plot.data(fuzzy.sort.data(data[,factor], factors[[factor]]), proportion = TRUE, main = names(factors)[[factor]])
 }
@@
 ## Removed columns
-removed_factors <- c(1)
+removed_factors <- c(1,6)
 
@@
 
+Creating factors for the three approaches (roughly sorting according to `methods_type`).
+
+```{r}
+## Empty array of methods
+methods <- rep(NA, nrow(data))
+
+## Claddis methods
+Claddis_data <- grep("discrete", data[,2])
+Claddis_ordination <- grep("PCO", data[,4])
+Claddis_metric <- unique(grep("variance", data[,5]), grep("ranges", data[,5]))
+Claddis_methods <- unique(Claddis_data, Claddis_ordination, Claddis_metric)
+
+## Geomorph methods
+geomorph_data <- grep("landmarks", data[,2])
+geomorph_ordination <- grep("PCA", data[,4])
+geomorph_metric <- grep("distance", data[,5])
+geomorph_methods <- unique(geomorph_data, geomorph_ordination, geomorph_metric)
+
+## Geiger methods
+Geiger_data <- grep("continuous", data[,2])
+Geiger_ordination <- grep("none", data[,4])
+Geiger_metric <- grep("pair_distances", data[,5])
+Geiger_methods <- unique(Geiger_data, Geiger_ordination, Geiger_metric)
+
+## Multiple methods
+Claddis_methods2 <- Claddis_methods[-c(which(!is.na(match(Claddis_methods, geomorph_methods))), which(!is.na(match(Claddis_methods, Geiger_methods))))]
+geomorph_methods2 <- geomorph_methods[-c(which(!is.na(match(geomorph_methods, Claddis_methods))), which(!is.na(match(geomorph_methods, Geiger_methods))))]
+Geiger_methods2 <- Geiger_methods[-c(which(!is.na(match(Geiger_methods, geomorph_methods))), which(!is.na(match(Geiger_methods, Claddis_methods))))]
+
+## None classsified
+none_methods <- which(is.na(match(seq(1:nrow(data)), c(Claddis_methods2, geomorph_methods2, Geiger_methods2))))
+
+methods[none_methods] <- "other"
+methods[Claddis_methods2] <- "Claddis"
+methods[geomorph_methods2] <- "geomorph"
+methods[Geiger_methods2] <- "Geiger"
+```
+
 ```{r, fig.width = 8, fig.height = 8, eval = FALSE, echo = FALSE}
+## Colours
+cols <- presentation.palette()[c(1,4,9,8)]
+colours <- cols[as.factor(methods)]
+## Green = geomorph
+## Orange = Claddis
+## Blue = Geiger
+## Light = NA
 
-factor <- 5
-# length(unique(multi_data[,1]))
-# length(factors[[2]])
+plot(ordinated_data[,1], ordinated_data[,3], xlab = paste0("PC1 (",percentages_scree[1], "%)"), ylab = paste0("PC2 (",percentages_scree[2], "%)"), pch = 19, col = colours)
 
-col_palette <- rainbow(length(unique(data[, factor])))
-colours <- col_palette[as.factor(multi_data[, factor-1])]
+legend("bottomright", legend = levels(as.factor(methods)), col = cols, pch = 19, cex = 0.5)
+```
 
-plot(ordinated_data[,1], ordinated_data[,2], xlab = paste0("PC1 (",percentages_scree[1], "%)"), ylab = paste0("PC2 (",percentages_scree[2], "%)"), pch = 19, col = colours, main = paste0(names(factors)[factor]))
-
-legend("bottomright", legend = sort(factors[[factor]]), col = col_palette, pch = 19, cex = 0.5)
+```{r}
 

### 2018-02-26 Writing/disparity_review.tex
Intro checked

@@
 % History of disparity methods
-Since the advent of biology and probably before that, it has been observed that variation in morphologies follows discrete patterns rather than a continuous gradient of forms.
-Disparity has shifted from it's English meaning (i.e.
-dissimilarity) in the seventies to arrive to the ``standardised'' and popularised meaning we now today in biology [Runnegar 1987, Gould, Foote, Webb, etc.].
-In these seminal papers, disparity was used to describe a particular biological, phenotypic, and/or trait variation without strict distinction between the ontology and episthemology (see Semantic Box).
-Over the last 30 years, disparity has been usually used in the morphological disparity context.
-The numerous papers are mainly united through the study of phenotypes, using often multivariate data and methods focused on the species level or above in a macroevolutionary context.
-However, it is important to not that these sets of approaches are not restricted to areas.
-It has been used a microevolutionary levels or even in other fields such as ecology or linguistics.
-Our approach throughout this paper is to not constrain our definition of disparity to not build barriers between disciplines.
-Disparity provides data on the structure of biodiversity.
-In essence, disparity is a pattern of variation.
+Since the time of the ancient, it has been observed that variation in morphologies often follows discrete patterns rather than a continuous gradient of forms.
+These clusters of morphological groups, or, rather their differences (i.e.
+the disparity of shapes) was popularised in seminal palaeobiology papers in the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.].
+Disparity is since then used to describe a particular biological, phenotypic, and/or trait variation without strict distinction between the ontology and epistemology (see Semantic Box).
+Over the last 30 years, disparity has been commonly used in the morphological disparity context.
+The impressive amount of literature published since then tackles a vast array of biological questions, using different (sometimes contrasted) methods but often consists studies of the phenotype at a macroevolutionary level (i.e.
+species or above).
+To this date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
+Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
+In essence, we suggest that disparity is a pattern of variation.
 
 %2§ what use: commonly used
-The description of these patterns (the disparity of patterns) has thus been used in evolutionary biology to tackle a vast array of questions.
-Disparity patterns have been used to tackle questions in Evo-devo through
+In evolutionary biology, the description of these patterns has been used to tackle a vast array of questions.
+For example, it has been used (1) in Evo-devo to tackle questions of
 %evo-devo
-Developmental constraints [CITE],
+Developmental constraints [CITE];
 Relationship between form and function and evolution functional traits [CITE] or
@@
 % Ecology
-in ecology with
-Competition - ecologically relevant [CITE],
-Niche width, competition  - ecological perspective / specialisation [CITE] or
-Macroecology (disparity in space)[CITE];
+(2) in ecology or palaeoecology to study:
+intra-specific competition [CITE],
+niche occupants, speciation and competition[CITE] or
+macroecology (disparity in space)[CITE];
 % Evolution
-on in macroevolution:
-Response to extinction - Extinction selectivity [CITE],
-Competitive replacements [CITE] or
-Tempo and mode of evolution [CITE].
+(3) on in macroevolution to study:
+response to extinction and selectivity [CITE],
+intra-specific competitive replacements [CITE] or
+more globally, tempo and mode of evolution [CITE].
 
@@
 \subsubsection{[Box]Disparity patterns or processes}
-At the core of defining what \textit{is} disparity lies a divide between episthemology and ontology.
+At the core of defining what \textit{is} disparity lies a divide between epistemology and ontology.
 Is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
@@
 \subsubsection{Models}
+One 
 What are our null hypotheses?
@@
 Phylogenetic correlations
-
-Do a check list of suggestions for standardisation: put your \% variance on the axis (and scale them)!
-Andrew’s list of things
-* should we z-score our data always?
-* volume of a n-dimensional ellipsoid scales with number of dimensions, initially increasing and then decreasing with the shape contingent on the radius R.
-Need good understanding of your data
-Is space Euclidean?
-Visualisation vs. analysis
-Holes in morphospace? How to deal with these?
-What to do with phylogeny/about phylogenetic autocorrelation?
-What about other forms of autocorrelation
-Size correction?
 Avoid using shiny new methods unless you understand
-Avoid being too prescriptive in advice of paper
 Outline assumptions of different approaches--make clear caveats

### 2018-02-27 Writing/disparity_review.tex
re-written up to methods

@@
 \subsubsection{[Box]Disparity patterns or processes}
-At the core of defining what \textit{is} disparity lies a divide between epistemology and ontology.
-Is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
-If disparity describes patterns of variation, the underlying scientific question ultimately always wants to use them as a proxy for explaining the biological processes.
+\label{box_semantic}
+At the core of defining what \textit{is} disparity lies a potential divide between epistemology and ontology:
+is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
+If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
 In the seminal papers from the late 80s, disparity was solely use to describe and understand the Cambrian Explosion were both the pattern (a high observed dissimilarity between groups) was directly linked to the process (and explosive radiation of multicellular life as championed by Gould).
 Restricting disparity to this initial informal definition (multidimensional morphological at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
-% That is, infer the mode of evolution (how does evolution occur and how does the structure of variation occur - both a constructive and destructive process) and the tempo (i.e., rate) at which it occurs.
-I think variation can also potentially tell us about ecology (as separate to evolution - for example, function, community structure, competition, etc.)...but...is ecology different from evolution??
-….Something I have been debating is whether it tells us about environment.
-Can we understand something about past environments from disparity??
+In fact, in the last years, disparity was also used as a potential tool for investigating ecological processes (disconnected from evolution: e.g.
+function [CITE], community structure[CITE], competition[CITE]).
+In other words disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
 \subsubsection{[Box]End}
@@
 
-\subsection{Find a question}
+\section{Find a question}
 Why to perform a disparity analysis?
-It might sound like an obvious question to start with but we would like to stress its crucial importance.
+This question might sound like an obvious question to start with but we would like to stress its crucial importance.
 The last decade have seen a vast array of excellent methodological implementation and documentations rendering disparity analysis easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
-However, this facilitation might (conscientiously or not) push authors to perform disparity analysis ``just because''.
-It is important to not rush into a disparity analysis to avoid some basic statistical caveats.
-For example, one major thing to consider is circularity: if your aim is to explore the variation in the data set, a multi-dimensional analysis is perfectly valid.
-However, the resulting groups are then no more appropriate for testing a hypothesis.
-Instead, rather do the data exploration on a subset of the data (preliminary analysis) and then use the full data set to test the resulting hypothesis.
-We believe that clearly framing an hypothesis and its null expectancy will help designing better disparity analysis.
-Once the question clearly framed, which data and which methods will be appropriate to answer it?
+However, this facilitation might (conscientiously or not) pushes authors to perform disparity analysis ``just because'' (i.e.
+simply because of their facility rather than stemming from a conscious hypothetico-deductive approach).
+It is thus important not to rush in a disparity analysis to avoid some basic statistical caveats.
+For example, one major thing to consider is circularity: if ones aim is to explore the variation in the data set, a multi-dimensional analysis is perfectly valid.
+However, the groups resulting from this exploratory analysis are then no more appropriate for testing a hypothesis [CITE stats 1.0].
+Instead, it is more appropriate do the data exploration on a subset of the data (preliminary analysis) and then use the full data set to test the resulting hypothesis.
+Clearly framing such hypothesis and their null expectancy will help designing better disparity analysis: which data and methods will be appropriate to test it?
 
-\subsection{Collect the appropriate data}
-If disparity is used to describe pattern to explain some process, remember that the every possible data will always only be a proxy for these patterns.
-As a sub-note to this, more data is not necessarily better, especially when the data was originaly designed and collected for different questions (e.g. cladistic data).
-Again, framing the question helps to avoid this caveat.
-For example, if a study is focussing on the ecological adaptive radiation of a group, ecological traits must be crucial for such disparity analysis.
+\section{Collect the appropriate data}
+If disparity is used to describe patterns for explaining some process, every possible data will always only be a proxy for these patterns.
+Tangentially, more data is not necessarily better, especially when the data was originally designed and collected for different questions (e.g. cladistic data - see below).
+Therefore, clearly framing the question as indicated above will help avoiding some major caveats.
+For example, if a study is focussing on the ecological adaptive radiation of a group, ecological traits must be crucial an might be preferred to purely evolutionary traits.
 
-\subsubsection{Which type of data}
-Additionally different types of data can be collected to answer different questions:
-(1) discrete data that can be either describing feature (often referred to as ``cladistic''; e.g.
-the presence or absence of legs, the colour of a tail, etc.) or more simply counts (e.g.
-number of digits, etc.);
-(2) continuous data that can be either direct measurements (e.g. body length, etc.) or more sophisticated methods (e.g. geometric morphometrics, mathematical contour description, etc.);
-(3) or any mix and match of the two former (e.g. ratios).
-In this case, the main question can help choosing the most appropriate data.
-For example, when looking at variation within bats species, landmarks (for geometric morphometrics analysis) or continuous measurements on bones can be approriate.
-On the other hand, when looking at convergence between bats and birds, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.
+\subsection{Which type of data}
+Morphological disparity analysis can be based on three main types of data:
+\begin{enumerate}
+    \item discrete data that can be either describing feature (often referred to as ``cladistic''; e.g.
+the presence or absence of legs, the colour of a tail, etc.
+[CITE]) or more simply counts (e.g.
+number of digits, etc.
+[CITE]);
+    \item continuous data that can be either direct measurements (e.g.
+body length, etc.
+[CITE]) or more sophisticated methods (e.g.
+geometric morphometrics, mathematical contour description, etc.
+[CITE]);
+    \item or any mix and match of the two former (e.g. ratios [CITE]).
+\end{enumerate}
+None of these types of data are fundamentally better but rather each one will be more appropriate to a specific question/
+For example, when looking at variation within bats species wings shape, homologous landmarks (for geometric morphometrics analysis) or continuous measurements on bones can be appropriate.
+On the other hand, if the question focuses on the convergence between bats and birds wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.
 
-Once the type of appropriate data is decided, sample size and sampling issues should be carefully considered next.
-Does one need to include intra-specific variation with multiple specimen per species when possible?
-For example, if you're studying variation of skull in canids, and you decide to include dogs, it is essential to capture their variation by including intra-specific one.
-On the other hand, if you're looking at variation of canids through time since the Eocene, intra-specific variation will not have major impact (up until the Holocene).
+Regardless of the type of data, one really common follow up question is how much of it should be collected (both in terms of number of specimens and traits)?
+\begin{itemize}
+    \item \textit{Which taxonomic scope?}
+    Whether disparity is a process or merely a method to describe a pattern (see Box \ref{box_semantic}) some studies have successfully use morphological disparity at both a inter-specific [CITE] or intra-specific level [CITE].
+    The taxonomic scope will thus depends essentially on which type of variation ones want to analysis.
+    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation (specially domestic dogs) to capture the full scope of skull shapes.
+    On the other hand, if the study focuses on canids skull variation through time, a inter-specific variation will not have major impact (up until the Holocene).
 %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
+    \item \textit{Which sample size?}
+    The amount of data needed to test an hypothesis is by far not restricted to morphological disparity analysis.
+    There is always a trade-of between the amount of data to collect and the time allocated to a project.
+    In these conditions, it is mainly important to make sure that the data is collected fairly for subsequent statistical analysis.
+    For example, if the question is to compare disparity among groups, one should collect, when possible, a similar number of elements per groups or.
+    When this is not possible (e.g. for palaeontological data), one can always use statistical approaches to correct or asses biases due to data collection (e.g. rarefaction analysis [CITE]).
+    \item \textit{Are the data consistent?}
+    This is maybe the most important question related to the nature of the data.
+    It is essential that the data is consistent and homogeneous to make sure that the disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data set.
+    For example, are the observed differences between two groups not related to the group's specificities (e.g. sexual dimorphism)?
+    Are the elements fairly consistent and comparable them?
+    Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
+    In any of these cases, are the patterns a mere reflection of these inconsistencies?
+    For example, if ones use disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g.
+snakes) that could not reflect processes.
+\end{itemize}
+In practice, however, these considerations can sometimes only be caveats to keep in mind.
+The amount or type of data is often constrained by more mundane consideration like how much data can be collected in a particular space, time frame and financial set.
 
-Another aspect to consider regarding the amount of data to collect is to make sure it will allow to be used fairly for subsequent statistical analysis.
-For example, if the question is to compare disparity among groups, collect a similar number of elements per groups or, when not possible (e.g. palaeontological data), use rarefied data.
-Also, are the elements consistent and comparable them?
-Does the studied group present clear sexual dimorphism?
-Are they collected at a relevant taxonomic scope?
-For example in birds, if you're looking at the family level, if the variables measured do not distinguish any passerine from another family (i.e.
-all passerine are as dissimilar to a falcon), then taxonomic sampling can be reduced.
-On the other hand, if the variables distinguish within passerines, then it is important to sample more.
-
-Finally, it is important to remember that morphological data can be patchy.
-Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
-Each of these might be a problem in specific cases.
-For example, hierarchical data can be problematic when comparing groups: if the question is about the disparity in locomotory modes among vertebrates, data can be collected from limbs but will then not be applicable to non-tetrapods or some specific tetrapods (e.g.
-snakes).
-This problem can be linked to whether only homologuous characters should be used or not.
-
-\subsubsection{Recycling data}
-Another important type of data is... other studies' data!
-In fact, it is a common (and good) practice to recycle data from previous studies whether published by the same author or not.
-It is then important to remember where this data comes from and why it was collected.
-For example, when using discrete morphological data used for phylogenetic analysis, it is important to be aware that disparity in such dataset can be maximised depending on the coded characters.
-It has actually been argued [by Foote - crinoids non-cladistic paper] that 
-Some discrete morphological datasets for phylogenetic uses can be designed to minimise homoplasy to obtain well resolved trees (thus increasing disparity among taxa).
-Conversely some discrete morphological datasets for the same group can be designed to maximise homoplasy to study the similarity among groups [cite Foote's crinoids paper].
-
+\subsection{Recycling data}
+Following this more down to earth consideration, one common practice in disparity analysis is to recycle data from other studies.
+This leads to another set of consideration on the source and the hypothesis tested with the original data.
+One classic case is when using ``cladistic'' data in palaeobiology.
+This data is usually collected for phylogenetic analysis and are thus designed to maximise differences between taxa to recover a clear phylogenetic signal.
+This maximise difference can then artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
 One relatively simple way to control for this specific bias can be to not use all the available data but select within the dataset the most pertinent characters for the question.
@@
 However, it is important to keep in mind that there is no one-to-one mapping between form a function and that some characters can be linked to multiple functions (e.g.
-limb traits can be both linked to locomotion and diet!).
+limb traits can be both linked to locomotion and diet).
+Ultimately, whether the data has been originally sourced or is recycled and whether any of the caveats described above are taken into account, disparity patterns will eventually be measured and analysed in another multitude of ways a we suggest that, whenever possible, this should first and foremost be considered in the design of the data collection protocol.
 
-\subsubsection{[Box]Data check list}
-
-When collecting data:
-\begin{itemize}
-    \item Which type of data?
-    \item Which traits?
-    \item Which sample size? (or correction)
-    \item Which taxonomic level?
-    \item Are their data caveats (missing, hierarchical, non-homologuous)?
-\end{itemize}
-
-When recycling data:
-\begin{itemize}
-    \item What was the data collected for?
-    \item Which traits are relevant to the question?
-    \item Are their data caveats (missing, hierarchical, non-homologuous)?
-\end{itemize}
-
-\subsection{Use methods appropriate} 
+\section{Use appropriate methods} 
 One of the major problem in disparity analysis is how to analysis data that is, in a vast majority of cases multidimensional.
@@
 
-\subsubsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)?}
+\subsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)?}
 \label{box_ordination}
@@
 Regarding these advantages and inconvenient of ordination technique, we strongly suggest to not automatically ordinate data in multidimensional analysis and to carefully consider whether the question can be answered without ordination.
-\subsubsection{[Box] end}
+\subsection{[Box] end}
 

### 2018-02-27 Writing/Figures/Figures_in_R.Rmd
Overlap figure

---
title: "Figures"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Methods part

## Visualisation

A multidimensional space similar to Figure 3 in Davis et al 2012 Nature.

```{r}
set.seed(1)
## Sampling the four different groups
size <- 10            # x                      #y                       #z
group1 <- cbind(rnorm(size, -0.5, 0.25), rnorm(size,  0.5, 0.25), rnorm(size,  0.5, 0.25))
group2 <- cbind(rnorm(size, -0.5, 0.25), rnorm(size,  0.5, 0.25), rnorm(size, -0.5, 0.25))
group3 <- cbind(rnorm(size,  0.5, 0.25), rnorm(size,  0.5, 0.25), rnorm(size,  0.5, 0.25))
group4 <- cbind(rnorm(size,  0.5, 0.25), rnorm(size, -0.5, 0.25), rnorm(size,  0.5, 0.25))
groups <- list(group1, group2, group3, group4)

## Chull coordinates function
get.chull.coords <- function(x, axis) {
    chull_points <- chull(x[, c(axis)])
    chull_coords <- rbind(x[chull_points, axis], x[chull_points[1], axis])
}

## Getting the coordinates
chull_coords_xy <- lapply(groups, get.chull.coords, axis = c(1, 2))
chull_coords_xz <- lapply(groups, get.chull.coords, axis = c(1, 3))

## Empty plot function
empty.plot <- function(xlab = "Dimension 1", ylab) {
    par(bty = "n")
    plot(NULL, xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1), xlab = xlab, ylab = ylab, xaxt='n', yaxt='n')
    abline(v = 0)
    abline(h = 0)
}

## Setting the colour palette
col_palette <- as.list(viridis::viridis(4))

par(mfrow = c(2, 1))
par(mar = c(1, 4, 1, 3))
## Plotting the first coordinates sets
empty.plot(xlab = "", ylab = "Dimension 2")
text(-1,1, "1", cex = 2)
## Adding the points
plot.points <- function(x, col, axis) points(x[, axis], col = col, pch = 19, cex = 0.7)
silent <- mapply(plot.points, groups, col_palette, MoreArgs = list(axis = c(1, 2)))
## Adding the polygons
plot.polygons <- function(x, col) polygon(x[, 1], x[, 2], col = col, density = 10, border = col)
silent <- mapply(plot.polygons, chull_coords_xy, col_palette)
legend("bottomlef", legend = LETTERS[1:4], col = unlist(col_palette), bty = "n", pch = 19, lty = 1)

par(mar = c(4, 4, 1, 3))
## Plotting the second coordinates sets
empty.plot(ylab = "Dimension 3")
text(-1,1, "2", cex = 2)
## Adding the points
silent <- mapply(plot.points, groups, col_palette, MoreArgs = list(axis = c(1, 3)))
## Adding the polygons
silent <- mapply(plot.polygons, chull_coords_xz, col_palette)

```

```{r, eval = FALSE, message = FALSE, fig.width = 12, fig.height = 12, echo = FALSE}
```

### 2018-02-27 Writing/disparity_review.tex
Manuscript up to methods

@@
 \section{Use appropriate methods} 
-One of the major problem in disparity analysis is how to analysis data that is, in a vast majority of cases multidimensional.
+In the vast majority of morphological disparity analysis, the methodological approach ought to be multidimensional.
 Multidimensional analysis can be a really powerful tool to analysis complex datasets but also comes with many mathematical caveats and limitations.
@@
 
+One first step in any statistical analysis is to visualise the data to understand its distribution and limitations at a glance.
+Unfortunately, our brains and physical world only allows us to understand a maximum of three dimensions at the same time.
+Therefore one of the main challenges is multidimensional analysis is how to reduce the number of dimensions whether this is for analysis (see box \ref{box_ordination}) or visualisation.
+
 \subsection{Visualising the multidimensional problem}
 \label{visualisation}
-One really common first step in multidimensional analysis is to reduce the number of dimensions.
-Whether this entails an ordination or not (see box \ref{box_ordination}) it is often good practice to first visualise the data.
-However, the physical reality we're living in allows a maximum of three dimensions to be visualised.
-
-First, the interpretation of the data visualisation can be misleading:
-Classically, it is incorrect to assume that because some groups overlap on a bivariate (or 3D) plot, they overlap in the whole multidimensional space (especially if the space is not euclidean!).
+Because of the limitation of our physical world, it is important to remember first and foremost that visual interpretation of multidimensional data can easily be misleading.
+For example, it is incorrect to assume that because some groups overlap on a bivariate (or 3D) plot, they overlap in the whole multidimensional space (especially if the space is not euclidean!).
 Note that conversely, when two groups don't overlap on a two or three dimensional plot, they \textit{do not} overlap in the multidimensional space.
@@
 For example, the distance between two points A and B might not necessary be equal to the distance between B and A if the space is not Euclidean.
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
+\caption{\small{The overlap (or not) of groups can vary between dimensions.}}
+\label{Fig:RF_results_best}
+\end{figure}
 

### 2018-03-01 Writing/disparity_review.tex
updated up until metrics part

@@
 Note that conversely, when two groups don't overlap on a two or three dimensional plot, they \textit{do not} overlap in the multidimensional space.
-Similarly, straight lines in a 2/3D representation of a space can be misleading, depending on the geometrical properties of the space, they could not be interpreted in our intuitive Euclidean distance way.
-For example, the distance between two points A and B might not necessary be equal to the distance between B and A if the space is not Euclidean.
+Furthermore, multidimensional spaces are not necessary Euclidean.
+Such spaces can have some non-intuitive properties, for example straight lines in some dimensions are not straight in the space or, less intuitively, distance between two points A and B might not be equal to the distance between B and A [CITE Sylvain].
 
@@
    \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
-\caption{\small{The overlap (or not) of groups can vary between dimensions.}}
+\caption{\small{Two groups overlap in some dimensions do not necessary overlap in all the dimensions.
+Conversely, two groups not overlapping on any dimensions don't overlap in all dimensions.
+In this example, no group overlaps in the multidimensional space despite the appearances.}}
 \label{Fig:RF_results_best}
@@
 
-Second, because of the complexity of multidimensional spaces, it is always important to be entirely clear about what is represented in their visualisation.
+It is thus always important to be entirely clear about what is represented in multidimensional spaces visualisation.
 In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a spatial pattern if not all the key properties of the space are present in the graph.
@@
 disparity through time).
-
 Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worth, it won't be what it seems.
-
-% Can we do an Anscombe’s Quartet for disparity?
-(see: %https://en.wikipedia.org/wiki/Anscombe\%27s_quartet; Revell has done this for PCMs: http://blog.phytools.org/2017/10/phylogenetic-anscombe-datasets.html; Wills 2001 has kinda already done this though, redraw his Figure 18?)
 
@@
 This is advantageous for plotting and visualising the data (see Visualisation section \ref{visualisation}) which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
-Additionally, ordinating the data can allow to further reduce the number of dimensions by only selecting a number of axis based on the metric properties (mathematical - e.g.
-product base metric - or computational - e.g.
-hull based metric).
+Additionally, ordinating the data can allow to further reduce the number of dimensions by only selecting a number of axis based on the hypothesis (e.g.
+selecting only the axis that describe up to 95\% of the variation).
 This can be done through the scree plots (cuts off, majority axis, etc.) or other statistical analysis (Horns parallel, etc.).
-
 In the case of geometric morphometric data (point 3 above), ordination can be particularly relevant since it conserves mathematical properties of the data while efficiently reducing the dimensions: the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc.
@@
 Furthermore, interpreting the biological variation along the axis is always a \textit{post-hoc} procedure and might bear little meaning regarding the disparity question (e.g.
-the elongation and flattening of the beak when the disparity question is about flight pattern disparity).
+the elongation and flattening of the beak when the question is about flight pattern disparity).
 
 In fact, in many cases, depending on the question, ordination might actually not be necessary.
-For example, if the disparity metric relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate in on ordinated data.
+For example, if the disparity metric relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
 Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
 For example, when the analysed data is non Euclidean (e.g.
-discrete morphological characters), the ordination can be problematic (character hierarchy - i.e.
-inapplicable -, polymorphisms, missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
-In such case, distances can be inappropriate in a mathematical (and colloquial) sense (e.g. the distance between A and B might differ from the distance between B and A).
+discrete morphological characters), the ordination can be problematic (character hierarchy - inapplicable -, polymorphisms, missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
 This can be problematic when comparing the position of groups in the multidimensional space since the measured distances might not be comparable among each other.
@@
 For many disparity analysis looking at macroevolutionary questions, phylogeny is often taken into account, whether for including estimates of ancestors or to add a phylogenetic structure to the data.
-For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
+For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
 and will thus not covered further in this review.
@@
 Ancestral states estimations can be used in two ways:
-(1) either before transformation of the data (e.g. ordination, distance matrix, etc.) by simply estimating the ancestral states based on the original dataset and adding them to it [CITE].
-(2) else after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
-The first method is generally thought to be more thorough but can imply longer computation and will change the ordinated space configuration.
-The second is much more faster and does not change the ordinated space configuration (i.e.
-the ancestors can be seen as averages between tips in the space) but it will change the properties of the space (variance and degrees of freedom).
-MORE UPCOMING WORK ON THAT
+\begin{enumerate}
+\item pre-ordination: the estimations is done before transformation of the data (e.g. ordination, distance matrix, etc.) and is simply based on the original dataset[CITE].
+\item post-ordination: the estimation is done else after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
+\end{enumerate}
+Pre-ordination ancestral states estimation can change the ordinated space geometry (i.e.
+the relation between the points not estimated) and implies longer computational method but once the morphospace is defined, it will remain constant (i.e.
+its properties won't change).
+Post ordination ancestral states estimation will not change the ordinated space geometry and will be faster to compute, however, it will change the morphopspace's properties such as the number of parameters (i.e.
+elements in the space - this can be problematic for statistical tests down the line) or the variance per axis [CITE Graeme's paper to be?].
 
-More importantly however, one needs to keep in mind that using such methods can come with several caveats.
+More importantly, however, one needs to keep in mind that using such methods can come with several caveats.
 First and foremost, although ancestral states estimations are valid methods in macroevolution, they results are highly dependent on the data and chosen methods.

### 2018-03-02 Writing/disparity_review.tex
metrics

@@
 
-Disparity metrics are essentially a one dimensional summary of a multidimensional object, hence, they are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
+After building and visualising the morphospace, one can use disparity metrics (or indices) as a one dimensional summary of it to test hypotheses.
+Because of these reduction in dimensionality, these metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
 With that in mind, the choice of the metric should be essential to summarise the aspect of the morphospace that is the most related to the question.
@@
 
-In multidimensionality, this high availability of unidimensional descriptors of the space are really numerous and come with their equally numerous advantages and caveats.
-In this case, the disparity question should again drive the choice of the disparity metric.
+In multidimensionality, this high availability of unidimensional descriptors comes with equally numerous advantages and caveats.
 If the question refers to ``size'' of a group in the morphospace (e.g.

### 2018-03-08 Writing/disparity_review.tex
Minor edits to acknowledgemetns and addresses
Edits as far as methods

@@
 \bigskip
-\noindent RH: disparate views on disparity.
+\noindent RH: disparate views on disparity. % NC: Needs to be changed?
 
@@
 \noindent{\Large \bf Variations in multidimensional space.} 
+% NC: Not sure about this title, especially now we are just doing a mini series. I think it really needs disparity in the title somewhere.
 \bigskip
@@
 \noindent {\small \it 
-$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.; $^2$NHM; $^{10}$Bristol}
+$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.\\ 
+$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK.\\ 
+$^{10}$Bristol}
 \end{center}
@@
 
+%NC: OK I think this paragraph needs some rewriting.
+At present it's unclear to me what you're trying to tell me with this paragraph.
+Are you trying to tell me about the history of disparity?
+Are you trying to define it and tell me about the debates surrounding that?
+I think mostly you're aiming for the latter, but this needs to be much clearer.
+Try and be a bit more upfront about what you're doing, and don't be afraid of short, clear sentences without any fancy words.
+i.e.
+epistemiology and ontology helps no-one!
+Also note that you start by saying disparity = morphology, then you say it could be anything but people argue about it, then you say it can be anything.
+So perhaps you need to start broad and then say one type of disparity, and the most commonly analysed, is morphological, and that's what most people would think of when hearing the term.
+But it could apply to anything.
+
 %1 -
 % History of disparity methods
-Since the time of the ancient, it has been observed that variation in morphologies often follows discrete patterns rather than a continuous gradient of forms.
-These clusters of morphological groups, or, rather their differences (i.e.
-the disparity of shapes) was popularised in seminal palaeobiology papers in the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.].
-Disparity is since then used to describe a particular biological, phenotypic, and/or trait variation without strict distinction between the ontology and epistemology (see Semantic Box).
-Over the last 30 years, disparity has been commonly used in the morphological disparity context.
+%Since the time of the ancient, it has been observed that variation in morphologies often follows discrete patterns rather than a continuous gradient of forms.
+% NC: Since the time of the ancients is super clunky.
+I've tried something different
+Even classical scholars recognised that variation in the shape and form of organisms often follows discrete patterns rather than a continuous gradient.
+% NC: changed the next bit as you need to define disparity/introduce it. As previously written it was hiding in a bracket.
+Disparity analysis, the study of differences among these clusters of morphological groups, was popularised in seminal palaeobiology papers in the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.].
+Since then, disparity has been used to describe biological, phenotypic, and/or trait variation without strict distinction between the ontology and epistemology (see Semantics Box) 
+% NC: Or glossary rather than semantics?
+% NC: I have no idea what you mean with this ontology and epistemiology comment. 
+Over the last 30 years, disparity has been most commonly used in the context of morphological disparity.
 The impressive amount of literature published since then tackles a vast array of biological questions, using different (sometimes contrasted) methods but often consists studies of the phenotype at a macroevolutionary level (i.e.
 species or above).
-To this date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
+To date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
 Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
@@
 %2§ what use: commonly used
-In evolutionary biology, the description of these patterns has been used to tackle a vast array of questions.
+% NC: I feel like this perhaps need to fit in to the paragraph above, where you're saying it can be used for lots of things? 
+The description of these patterns % NC: what patterns?
+ has been used to tackle a vast array of questions.
 For example, it has been used (1) in Evo-devo to tackle questions of
@@
 
-%3§ problems
+%3§ problems %NC: Could maybe put the subheading "best practices" here?
 Disparity studies have a great deal of promise in helping us to understand the evolution of biological diversity.
-With the expansion of such methodological approaches in the last decade, however, we also identify three main issues that, if not tackled could reduce the power of such analysis.
+With the expansion of methodological approaches in the last decade, however, we also identify three main issues that, if not tackled could reduce the power of such analysis.
 First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
@@
 \label{box_semantic}
-At the core of defining what \textit{is} disparity lies a potential divide between epistemology and ontology:
+At the core of defining what \textit{is} disparity lies a potential divide between epistemology and ontology: % NC: Can you either just write out what these terms mean, or add the explanation in brackets?
+I hate un-necessary jargon.
+Most PhD students will not be familiar with these terms.
+Were you before you got into this?
+(maybe you were, maybe the French education system is better!)
 is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
 If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
-In the seminal papers from the late 80s, disparity was solely use to describe and understand the Cambrian Explosion were both the pattern (a high observed dissimilarity between groups) was directly linked to the process (and explosive radiation of multicellular life as championed by Gould).
-Restricting disparity to this initial informal definition (multidimensional morphological at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
-In fact, in the last years, disparity was also used as a potential tool for investigating ecological processes (disconnected from evolution: e.g.
-function [CITE], community structure[CITE], competition[CITE]).
-In other words disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
+In the seminal papers from the late 80s, disparity was used to describe and understand the Cambrian explosion where both the pattern (high dissimilarities among groups) was directly linked to the process (an explosive radiation of multicellular life as championed by Gould REF).
+Restricting disparity to this initial informal definition (multidimensional morphological dissimilarity at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
+In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
+Thus disparity % NC: I'm still unclear what you mean when you say disparity here. Do you mean the methods, the idea of discrete groups? I'm struggling a bit to parse how it can "inform us" per se.
+could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
 \subsubsection{[Box]End}
+% NC: I think I prefer the previous version where you actually had a subsection to discuss this within the text rather than a box.
+Seems particularly weird as you als odo some defining of disparity in the first paragraphs.
+Maybe move that into the box?
+Then just say here we define dispairty as "XXXX" in the text.
+Or shorten this and put it within the text?
+See what works.
+But the point of a box is to add details not needed in the text, so if it's in the text anyway then I'm dubious about the box...
 
@@
 
-\section{Find a question}
-Why to perform a disparity analysis?
-This question might sound like an obvious question to start with but we would like to stress its crucial importance.
-The last decade have seen a vast array of excellent methodological implementation and documentations rendering disparity analysis easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
-However, this facilitation might (conscientiously or not) pushes authors to perform disparity analysis ``just because'' (i.e.
-simply because of their facility rather than stemming from a conscious hypothetico-deductive approach).
-It is thus important not to rush in a disparity analysis to avoid some basic statistical caveats.
-For example, one major thing to consider is circularity: if ones aim is to explore the variation in the data set, a multi-dimensional analysis is perfectly valid.
-However, the groups resulting from this exploratory analysis are then no more appropriate for testing a hypothesis [CITE stats 1.0].
-Instead, it is more appropriate do the data exploration on a subset of the data (preliminary analysis) and then use the full data set to test the resulting hypothesis.
-Clearly framing such hypothesis and their null expectancy will help designing better disparity analysis: which data and methods will be appropriate to test it?
+\section{Find a question} %NC: I wonder if "find a question" is exactly what we mean here.
+This suggests you've decided to use disparity whatever happens, and just need the right question.
+Maybe change to What is your question?
+Why are you performing a disparity analysis?
+This seems like it should be obvious, but we stress its crucial importance.
+The last decade has seen a vast array of new methods being published and implemented in freely available software, making disparity analyses easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
+However, this (consciously or not) pushes authors towards performing disparity analyses simply because of their ease of use, rather than stemming from a question driven approach.
+Before rushing into a disparity analysis, one should consider some basic caveats.
+For example, one major thing to consider is circularity: if the aim is to explore the variation in a dataset, disparity analysis is perfectly valid.
+However, if this exploratory analysis reveals distinct groups, it would be circular to then test for differences among those groups as if this had been your hypothesis all along [CITE stats 1.0].
+This can easily be avoided by using a subset of the data for preliminary data exploration, and then using the full dataset to test for differences among any groups that emerge.
+Clearly framing your questions and their null expectations will help in designing better disparity analyses.
 
 \section{Collect the appropriate data}
-If disparity is used to describe patterns for explaining some process, every possible data will always only be a proxy for these patterns.
-Tangentially, more data is not necessarily better, especially when the data was originally designed and collected for different questions (e.g. cladistic data - see below).
-Therefore, clearly framing the question as indicated above will help avoiding some major caveats.
-For example, if a study is focussing on the ecological adaptive radiation of a group, ecological traits must be crucial an might be preferred to purely evolutionary traits.
+% If disparity is used to describe patterns that are explained by a certain process, every possible data will always only be a proxy for these patterns. % NC: Don't really understand your point here.
+Once you have a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
+For example, if a study focusses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over purely evolutionary traits.
+Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).
 
-\subsection{Which type of data}
+\subsection{Which types of data?} % NC: Not sure having these subheadings as Q looks best, maybe just Data types?
 Morphological disparity analysis can be based on three main types of data:
 \begin{enumerate}
-    \item discrete data that can be either describing feature (often referred to as ``cladistic''; e.g.
+    \item discrete data that can be either describing a feature (often referred to as ``cladistic''; e.g.
 the presence or absence of legs, the colour of a tail, etc.
-[CITE]) or more simply counts (e.g.
+[CITE]) or counts (e.g.
 number of digits, etc.
 [CITE]);
-    \item continuous data that can be either direct measurements (e.g.
+    \item continuous data that can be either simple measurements (e.g.
 body length, etc.
-[CITE]) or more sophisticated methods (e.g.
+[CITE]) or more abstracted methods (e.g.
 geometric morphometrics, mathematical contour description, etc.
 [CITE]);
-    \item or any mix and match of the two former (e.g. ratios [CITE]).
+    \item or any mixture of discrete and continuous data (e.g. ratios [CITE]).
 \end{enumerate}
-None of these types of data are fundamentally better but rather each one will be more appropriate to a specific question/
-For example, when looking at variation within bats species wings shape, homologous landmarks (for geometric morphometrics analysis) or continuous measurements on bones can be appropriate.
-On the other hand, if the question focuses on the convergence between bats and birds wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.
+None of these types of data are fundamentally better but each will be more appropriate for specific questions.
+For example, when investigating variation within bat wing shapes, homologous landmarks (for geometric morphometrics analysis) or continuous measurements of bones may be appropriate.
+On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.
+% NC: This section might need more explanation and/or some more examples
 
-Regardless of the type of data, one really common follow up question is how much of it should be collected (both in terms of number of specimens and traits)?
+\subsection{How much data to collect?}
+Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of both specimens and traits.
+There are several important considerations.
 \begin{itemize}
-    \item \textit{Which taxonomic scope?}
-    Whether disparity is a process or merely a method to describe a pattern (see Box \ref{box_semantic}) some studies have successfully use morphological disparity at both a inter-specific [CITE] or intra-specific level [CITE].
-    The taxonomic scope will thus depends essentially on which type of variation ones want to analysis.
-    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation (specially domestic dogs) to capture the full scope of skull shapes.
-    On the other hand, if the study focuses on canids skull variation through time, a inter-specific variation will not have major impact (up until the Holocene).
-%http://www.journals.uchicago.edu/doi/abs/10.1086/650372
-    \item \textit{Which sample size?}
-    The amount of data needed to test an hypothesis is by far not restricted to morphological disparity analysis.
-    There is always a trade-of between the amount of data to collect and the time allocated to a project.
-    In these conditions, it is mainly important to make sure that the data is collected fairly for subsequent statistical analysis.
-    For example, if the question is to compare disparity among groups, one should collect, when possible, a similar number of elements per groups or.
-    When this is not possible (e.g. for palaeontological data), one can always use statistical approaches to correct or asses biases due to data collection (e.g. rarefaction analysis [CITE]).
-    \item \textit{Are the data consistent?}
+    \item \textit{Taxonomic scope}
+    Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
+    Taxonomic scope will depend on the question being investigated, and what type of variation is being analysed.
+    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation to capture the full variety of skull shapes, particularly that seen in domestic dogs (CITE).
+    Conversely, if a study focuses on canid skull variation through time, interspecific data is most appropriate.
+    % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
+    \item \textit{Sample size}
+    There is always a trade-off between collecting enough data and the time allocated to a project.
+    It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
+    For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
+    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction analysis [CITE]).
+    Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
+    \item \textit{Are the data consistent?} %NC: Not suer I get this section. What do you mean by consistent???? Also does this fit into the "how much data" section or is it a seaprate consideration?
     This is maybe the most important question related to the nature of the data.
@@
     In any of these cases, are the patterns a mere reflection of these inconsistencies?
-    For example, if ones use disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g.
+    For example, if one uses disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g.
 snakes) that could not reflect processes.
 \end{itemize}
-In practice, however, these considerations can sometimes only be caveats to keep in mind.
-The amount or type of data is often constrained by more mundane consideration like how much data can be collected in a particular space, time frame and financial set.
+Of course in practice some of these caveats are unavoidable. 
+Data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
+% NC: I'd be tempted to move this to the end of the section to finish it off.
 
 \subsection{Recycling data}
-Following this more down to earth consideration, one common practice in disparity analysis is to recycle data from other studies.
-This leads to another set of consideration on the source and the hypothesis tested with the original data.
-One classic case is when using ``cladistic'' data in palaeobiology.
-This data is usually collected for phylogenetic analysis and are thus designed to maximise differences between taxa to recover a clear phylogenetic signal.
-This maximise difference can then artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
-One relatively simple way to control for this specific bias can be to not use all the available data but select within the dataset the most pertinent characters for the question.
-For example, in a study on disparity of dietary niches, removing non-dietary characters (e.g. sexually dimorphic ones) can be a good approach.
-However, it is important to keep in mind that there is no one-to-one mapping between form a function and that some characters can be linked to multiple functions (e.g.
-limb traits can be both linked to locomotion and diet).
-Ultimately, whether the data has been originally sourced or is recycled and whether any of the caveats described above are taken into account, disparity patterns will eventually be measured and analysed in another multitude of ways a we suggest that, whenever possible, this should first and foremost be considered in the design of the data collection protocol.
+A common practice in disparity analysis is to recycle data from other studies which can introduce additional problems.
+One classic case is using ``cladistic'' data in palaeobiology.
+These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
+This data can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
+A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all the data.
+For example, in a study of disparity of dietary niches, removing non-dietary characters is a good approach.
+However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g.
+limb traits can be used for both locomotion and feeding).
+%Ultimately, whether the data was collected for the purpose of the study or is recycled, and whether any of the caveats described above are taken into account, disparity patterns will eventually be measured and analysed in another multitude of ways a we suggest that, whenever possible, this should first and foremost be considered in the design of the data collection protocol.
+%NC: Not sure these linking statements are needed
 
@@
 ``We asked whether we could but not whether we should'' (Jurassic Park)
+
+% NC: I do like this quote.
+The actual quote is: "Your scientists were so preoccupied with whether or not they could that they didn't stop to think if they should." Dr Ian Malcolm.
+It would be magnificent ot get a jurassic park quote into a paper...
 
@@
 \section{Authors contribution}
-TG, NC and PD proposed the idea of this review. TG, NC, SB, KD, AJ, SG, AG, KH, MH, GL, JOR, AP, ER, ES, ES, GS, GHT and PD wrote the manuscript,
+TG, NC and PD proposed the idea of this review. All authors edited the manuscript.
 

### 2018-03-08 Writing/disparity_review.tex
Edits to methods

@@
 \section{Use appropriate methods} 
-In the vast majority of morphological disparity analysis, the methodological approach ought to be multidimensional.
-Multidimensional analysis can be a really powerful tool to analysis complex datasets but also comes with many mathematical caveats and limitations.
-As any great power, it comes with great responsibilities.
-
-One first step in any statistical analysis is to visualise the data to understand its distribution and limitations at a glance.
-Unfortunately, our brains and physical world only allows us to understand a maximum of three dimensions at the same time.
-Therefore one of the main challenges is multidimensional analysis is how to reduce the number of dimensions whether this is for analysis (see box \ref{box_ordination}) or visualisation.
+In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
+Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
+As with any great power, it comes with great responsibility. % NC: Ha are we citing spiderman too? Awesome
+The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
+Unfortunately, we can only perceive a maximum of three dimensions at the same time.
+Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).
 
@@
 \label{visualisation}
-Because of the limitation of our physical world, it is important to remember first and foremost that visual interpretation of multidimensional data can easily be misleading.
-For example, it is incorrect to assume that because some groups overlap on a bivariate (or 3D) plot, they overlap in the whole multidimensional space (especially if the space is not euclidean!).
-Note that conversely, when two groups don't overlap on a two or three dimensional plot, they \textit{do not} overlap in the multidimensional space.
+It is important to remember that visual interpretations of multidimensional data can easily be misleading.
+For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
+Conversely, when two groups do not overlap on a two or three dimensional plot, this does not mean they never overlap in the multidimensional space.
 Furthermore, multidimensional spaces are not necessary Euclidean.
-Such spaces can have some non-intuitive properties, for example straight lines in some dimensions are not straight in the space or, less intuitively, distance between two points A and B might not be equal to the distance between B and A [CITE Sylvain].
+Such non-Euclidean spaces often have non-intuitive properties, for example straight lines in some dimensions are not straight in the space and, even less intuitively, the distance between points A and B might not be equal to the distance between points B and A [CITE Sylvain].
 
@@
    \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
-\caption{\small{Two groups overlap in some dimensions do not necessary overlap in all the dimensions.
-Conversely, two groups not overlapping on any dimensions don't overlap in all dimensions.
+\caption{\small{Two groups that overlap in some dimensions do not necessary overlap in all dimensions. 
+Conversely, two groups that do not overlap in some dimensions can overlap in other dimensions. 
 In this example, no group overlaps in the multidimensional space despite the appearances.}} 
+% NC: Might need to explain more clearly, I'm not 100% sure I got your meaning
 \label{Fig:RF_results_best}
@@
 
-It is thus always important to be entirely clear about what is represented in multidimensional spaces visualisation.
-In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a spatial pattern if not all the key properties of the space are present in the graph.
-We believe that these should always include the following:
-(1) the percentage of the variation of each axis of the plot (i.e. is the majority - or all - the variation of the data present in the plot or only a fraction?);
-(2) the scale of the axis (ideally, if the space is orthogonal, it is important to keep the same scale between the axis);
-and (3) the error associated with each point, whether this is in term of measurement error (on ordination plots for example) or in standard variation (in further analysis, e.g.
+It is therefore vital to be clear about what is represented in multidimensional space visualisations.
+In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a pattern if all the key properties of the space are not present in the graph.
+We suggest visualisations should include the following:
+(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation of the data present in the plot or only a fraction?);
+(2) the scale of the axes (ideally, if the space is orthogonal, it is important to keep the same scale across the axes);
+and (3) the error associated with each point, whether this is in terms of measurement error (on ordination plots for example) or some other measure of standard variation (in further analysis, e.g.
 disparity through time).
-Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worth, it won't be what it seems.
+Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading.
 
@@
 \label{box_ordination}
-One of the most common approach to analyse multidimensional data is to use ordination techniques.
-This consists in transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
-This can be done through three main analysis pipelines:
+One of the most common approaches to analysing multidimensional data is to use ordination techniques.
+These consist of transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
+This is generally done via three analysis pipelines:
 \begin{enumerate}
     \item data -> distance matrix -> ordination
-    \item data -> zscores -> PCA
+    \item data -> zscores -> principal components analysis (PCA)
     \item data -> Procrustes superimposition -> projection into tangent space -> PCA
@@
 
-%Why to ordinate:
-These pipelines allow to reduce the dimensionality of the data.
+%Why to ordinate: % NC: This box seems quite long, can it be reduced?
+These pipelines reduce the dimensionality of the data.
 This is advantageous for plotting and visualising the data (see Visualisation section \ref{visualisation}) which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
-Additionally, ordinating the data can allow to further reduce the number of dimensions by only selecting a number of axis based on the hypothesis (e.g.
-selecting only the axis that describe up to 95\% of the variation).
-This can be done through the scree plots (cuts off, majority axis, etc.) or other statistical analysis (Horns parallel, etc.).
-In the case of geometric morphometric data (point 3 above), ordination can be particularly relevant since it conserves mathematical properties of the data while efficiently reducing the dimensions: the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc.
-This has clear advantages for interpreting the results in a lower number of dimensions.
-For example, the axis will actually represent gradients of biological variation (e.g. elongation and flattening of the beak).
-This also allows easy interpretation of the proximity of the elements in space which reflects shared biological characters.
+Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g.
+selecting only the axes that describe up to 95\% of the variation).
+In the case of geometric morphometric data (point 3 above), ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions: the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc.
+% NC: is the stuff after the : needed in this sentence?
+This has clear advantages for interpreting the results.
+For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).
+This also allows easy interpretation of the proximity of the elements in space which reflect shared biological characters.
+%%%%%%%%%%%
+However, some caveats can arise from such transformations of the data.
+For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g. only the first few) can lead to misinterpretation of the results.
+Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and might have little relation to the disparity question (e.g.
+if the first few axes represent the elongation and flattening of the beak, but the question is about wing disparity).
+In many cases, depending on the question, ordination might not be necessary.
+For example, if the disparity metric being used (see below) relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate it on ordinated data [e.g.
+CITE Close].
+Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
+For example, when the analysed data is non-Euclidean (e.g.
+discrete morphological characters), the ordination can be problematic (due to character hierarchy, inapplicable characters, polymorphisms, and missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
+This can be problematic when comparing the position of groups in the multidimensional space, as the measured distances might not be comparable.
+Furthermore, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axes can be impossible or biologically meaningless (e.g.
+the ratio of data coded as 1 rather than 0).
+% NC: Need to explain this point more clearly
 
-However, some caveats can rise from such transformation of the data.
-For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g.
-only the few first) can lead to misinterpretation of the results erroneous statistical results.
-Furthermore, interpreting the biological variation along the axis is always a \textit{post-hoc} procedure and might bear little meaning regarding the disparity question (e.g.
-the elongation and flattening of the beak when the question is about flight pattern disparity).
-
-In fact, in many cases, depending on the question, ordination might actually not be necessary.
-For example, if the disparity metric relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
-Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
-For example, when the analysed data is non Euclidean (e.g.
-discrete morphological characters), the ordination can be problematic (character hierarchy - inapplicable -, polymorphisms, missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
-This can be problematic when comparing the position of groups in the multidimensional space since the measured distances might not be comparable among each other.
-Furthermore, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axis can be simply impossible or meaningless biologically (e.g.
-the ratio of data coded as 1 rather than 0).
-
-Regarding these advantages and inconvenient of ordination technique, we strongly suggest to not automatically ordinate data in multidimensional analysis and to carefully consider whether the question can be answered without ordination.
+Because of these points, we strongly recommend that data for multidimensional analyses is not automatically ordinated, and that careful consideration is given to whether the question can be answered without ordination.
 \subsection{[Box] end}
@@
 \subsection{Phylogeny}
-For many disparity analysis looking at macroevolutionary questions, phylogeny is often taken into account, whether for including estimates of ancestors or to add a phylogenetic structure to the data.
+For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including estimates of ancestors or to add phylogenetic structure to the data.
 For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
-and will thus not covered further in this review.
+and therefore will not covered further in this review.
 
-\subsubsection{Ancestral states estimation}
-When looking at disparity in the past (e.g. disparity-through-time) ancestral states estimations are a common approach.
-Ancestral states estimations can be used in two ways:
+\subsubsection{Ancestral state estimation}
+When looking at disparity in the past (e.g. disparity-through-time) ancestral state estimations are a common approach.
+Ancestral state estimation can be performed in two ways:
 \begin{enumerate}
-\item pre-ordination: the estimations is done before transformation of the data (e.g. ordination, distance matrix, etc.) and is simply based on the original dataset[CITE].
-\item post-ordination: the estimation is done else after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
+\item pre-ordination: the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original dataset[CITE].
+\item post-ordination: the estimation is done after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
 \end{enumerate}
-Pre-ordination ancestral states estimation can change the ordinated space geometry (i.e.
-the relation between the points not estimated) and implies longer computational method but once the morphospace is defined, it will remain constant (i.e.
+Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e.
+the relationship between the points not estimated) and implies longer computational method but once the morphospace is defined, it will remain constant (i.e.
 its properties won't change).

### 2018-04-08 Writing/disparity_review.tex
rewritten intro

@@
 
-\noindent{\Large \bf Variations in multidimensional space.} 
+\noindent{\Large \bf Disparity in morphological disparity analysis} 
 % NC: Not sure about this title, especially now we are just doing a mini series. I think it really needs disparity in the title somewhere.
+%TG: back to the drawing board. Maybe something with "best practices" as well.
 \bigskip
@@
 
-%NC: OK I think this paragraph needs some rewriting.
-At present it's unclear to me what you're trying to tell me with this paragraph.
-Are you trying to tell me about the history of disparity?
-Are you trying to define it and tell me about the debates surrounding that?
-I think mostly you're aiming for the latter, but this needs to be much clearer.
-Try and be a bit more upfront about what you're doing, and don't be afraid of short, clear sentences without any fancy words.
-i.e.
-epistemiology and ontology helps no-one!
-Also note that you start by saying disparity = morphology, then you say it could be anything but people argue about it, then you say it can be anything.
-So perhaps you need to start broad and then say one type of disparity, and the most commonly analysed, is morphological, and that's what most people would think of when hearing the term.
-But it could apply to anything.
+Even classical scholars recognised that variation in the shape, form and functions of organisms often follows discrete patterns rather than a continuous gradients.
+These patterns can be pictured as multidimensional spaces encompassing all the possible life forms where each individual region are the result of a discrete combination of many variables.
+For example, the mammalian carnivores design space can be seen as a combination of phylogenetic variables and their associated shared traits (i.e.
+the synapomorphies of the order Carnivora) as well as a combination of ecological, life history and behavioural variables (i.e.
+the ecological guild of meat eating mammals).
 
-%1 -
-% History of disparity methods
-%Since the time of the ancient, it has been observed that variation in morphologies often follows discrete patterns rather than a continuous gradient of forms.
-% NC: Since the time of the ancients is super clunky.
-I've tried something different
-Even classical scholars recognised that variation in the shape and form of organisms often follows discrete patterns rather than a continuous gradient.
-% NC: changed the next bit as you need to define disparity/introduce it. As previously written it was hiding in a bracket.
-Disparity analysis, the study of differences among these clusters of morphological groups, was popularised in seminal palaeobiology papers in the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.].
-Since then, disparity has been used to describe biological, phenotypic, and/or trait variation without strict distinction between the ontology and epistemology (see Semantics Box) 
-% NC: Or glossary rather than semantics?
-% NC: I have no idea what you mean with this ontology and epistemiology comment. 
-Over the last 30 years, disparity has been most commonly used in the context of morphological disparity.
-The impressive amount of literature published since then tackles a vast array of biological questions, using different (sometimes contrasted) methods but often consists studies of the phenotype at a macroevolutionary level (i.e.
-species or above).
-To date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
-Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
-In essence, we suggest that disparity is a pattern of variation.
-
-%2§ what use: commonly used
-% NC: I feel like this perhaps need to fit in to the paragraph above, where you're saying it can be used for lots of things? 
-The description of these patterns % NC: what patterns?
- has been used to tackle a vast array of questions.
+One commonly used approach to analyse these discrete patterns and to understand the processes that led to it is to analyse the disparity of the morphological charactersitics of the group, often referred to as morphological disparity analysis.
+This type of analysis was popularised around the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] by studying the fossil record and focusing on broad questions such how does diversity evolve, are some body types more common that others or can shape evolve in all ``directions'' (of the multidimensional space).
+Since then, an impressive amount of literature and the equally vast amount of methods have been used and the term ``disparity'' has become a close synonym of ``analysis of patterns of morphological diversity'' and has been used in a range of sub-fields in biology.
+Disparity analysis has been used to tackle a vast array of questions.
 For example, it has been used (1) in Evo-devo to tackle questions of
@@
 
-%3§ problems %NC: Could maybe put the subheading "best practices" here?
-Disparity studies have a great deal of promise in helping us to understand the evolution of biological diversity.
-With the expansion of methodological approaches in the last decade, however, we also identify three main issues that, if not tackled could reduce the power of such analysis.
+To date, however, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
+Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
+In essence, we suggest that disparity is a pattern of variation.
+This diversity (and disparity) of methods to analysis disparity provides biologist with a powerful and comprehensive tool %TG: comprehensive in the sense "global"/"universal" not sure if that's the right word here.
+that has a great deal of promise in helping us to understand the evolution of biological diversity.
+However, as with any great power, it comes with great responsibility.
+
+With the expansion of methodological approaches in the last decade, we also identify three main issues that, if not tackled could reduce the power of such analysis.
 First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
@@
 Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
-As with any great power, it comes with great responsibility. % NC: Ha are we citing spiderman too? Awesome
 The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
@@
 \subsubsection{Models}
-One 
 What are our null hypotheses?

### 2018-04-08 Writing/disparity_review.tex
Removed definition box

@@
 Since then, an impressive amount of literature and the equally vast amount of methods have been used and the term ``disparity'' has become a close synonym of ``analysis of patterns of morphological diversity'' and has been used in a range of sub-fields in biology.
-Disparity analysis has been used to tackle a vast array of questions.
-For example, it has been used (1) in Evo-devo to tackle questions of
-%evo-devo
-Developmental constraints [CITE];
-Relationship between form and function and evolution functional traits [CITE] or
-Scaling micro to macro [CITE];
-% Ecology
-(2) in ecology or palaeoecology to study:
-intra-specific competition [CITE],
-niche occupants, speciation and competition[CITE] or
-macroecology (disparity in space)[CITE];
-% Evolution
-(3) on in macroevolution to study:
-response to extinction and selectivity [CITE],
-intra-specific competitive replacements [CITE] or
-more globally, tempo and mode of evolution [CITE].
+Disparity analysis has been used to tackle a vast array of questions in Evo-devo (Developmental constraints [CITE]; Relationship between form and function and evolution functional traits [CITE] or Scaling micro to macro [CITE];), in ecology or palaeoecology (intra-specific competition [CITE], niche occupants, speciation and competition[CITE] or macroecology (disparity in space)[CITE];)
+or in macroevolution (response to extinction and selectivity [CITE], intra-specific competitive replacements [CITE] or more globally, tempo and mode of evolution [CITE]).
 
-To date, however, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
+\subsection{What \textit{is} disparity?}
+
+To date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
+
+At the core of this debate a potential divide between whether disparity is a patterns or a processes: is there a difference between the disparity patterns (and how researchers characterise them) and how they were ultimately generated?
+If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
+In the seminal papers from the late 80s, disparity was used to describe and understand the Cambrian explosion where both the pattern (high dissimilarities among groups) was directly linked to the process (an explosive radiation of multicellular life as championed by Gould REF).
+
+Restricting disparity to this initial informal definition (multidimensional morphological dissimilarity at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
+In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
+Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
 Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
 In essence, we suggest that disparity is a pattern of variation.
+Thus, disparity methods or analysis can be seen as methods or analysis to study these patterns and eventually link them to macroevolutionary or macroecological processes
+
+\section{Best practices for disparity analysis}
+
 This diversity (and disparity) of methods to analysis disparity provides biologist with a powerful and comprehensive tool %TG: comprehensive in the sense "global"/"universal" not sure if that's the right word here.
@@
 However, as with any great power, it comes with great responsibility. 
-
 With the expansion of methodological approaches in the last decade, we also identify three main issues that, if not tackled could reduce the power of such analysis.
@@
 Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions before performing a disparity analysis.
-
-\subsubsection{[Box]Disparity patterns or processes}
-\label{box_semantic}
-At the core of defining what \textit{is} disparity lies a potential divide between epistemology and ontology: % NC: Can you either just write out what these terms mean, or add the explanation in brackets?
-I hate un-necessary jargon.
-Most PhD students will not be familiar with these terms.
-Were you before you got into this?
-(maybe you were, maybe the French education system is better!)
-is there a difference between the pattern (and how researchers characterise it) and how it was ultimately generated?
-If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
-In the seminal papers from the late 80s, disparity was used to describe and understand the Cambrian explosion where both the pattern (high dissimilarities among groups) was directly linked to the process (an explosive radiation of multicellular life as championed by Gould REF).
-Restricting disparity to this initial informal definition (multidimensional morphological dissimilarity at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
-In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
-Thus disparity % NC: I'm still unclear what you mean when you say disparity here. Do you mean the methods, the idea of discrete groups? I'm struggling a bit to parse how it can "inform us" per se.
-could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
-\subsubsection{[Box]End}
-% NC: I think I prefer the previous version where you actually had a subsection to discuss this within the text rather than a box.
-Seems particularly weird as you als odo some defining of disparity in the first paragraphs.
-Maybe move that into the box?
-Then just say here we define dispairty as "XXXX" in the text.
-Or shorten this and put it within the text?
-See what works.
-But the point of a box is to add details not needed in the text, so if it's in the text anyway then I'm dubious about the box...
-
-\section{Best practices for disparity analysis}
 

### 2018-04-08 Writing/disparity_review.tex
updates on the structure

@@
 
-\section{Find a question} %NC: I wonder if "find a question" is exactly what we mean here.
-This suggests you've decided to use disparity whatever happens, and just need the right question.
-Maybe change to What is your question?
+\section{What is your question}
 Why are you performing a disparity analysis?
@@
 \section{Collect the appropriate data}
-% If disparity is used to describe patterns that are explained by a certain process, every possible data will always only be a proxy for these patterns. % NC: Don't really understand your point here.
 Once you have a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
@@
 
-\subsection{Which types of data?} % NC: Not sure having these subheadings as Q looks best, maybe just Data types?
+\subsection{Data types}
 Morphological disparity analysis can be based on three main types of data:
@@
 % NC: This section might need more explanation and/or some more examples
+
+%TG: add the bat example for above
 

### 2018-04-10 Writing/disparity_review.tex
Minor edits to the data section

@@
 
-\section{What is your question}
-Why are you performing a disparity analysis?
+\section{What is the question?}
+Why perform a disparity analysis?
 This seems like it should be obvious, but we stress its crucial importance.
@@
 For example, one major thing to consider is circularity: if the aim is to explore the variation in a dataset, disparity analysis is perfectly valid.
-However, if this exploratory analysis reveals distinct groups, it would be circular to then test for differences among those groups as if this had been your hypothesis all along [CITE stats 1.0].
+However, if this exploratory analysis reveals distinct groups, it would be circular to then test for differences among those groups as if this had been the hypothesis all along [CITE stats 1.0].
 This can easily be avoided by using a subset of the data for preliminary data exploration, and then using the full dataset to test for differences among any groups that emerge.
-Clearly framing your questions and their null expectations will help in designing better disparity analyses.
+Clearly framing questions and their null expectations will help in designing better disparity analyses.
 
 \section{Collect the appropriate data}
-Once you have a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
+Once you 
+%TG: not sure about using "you" all the time. I feel it does add some patronising tone the whole yoke.
+have a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
 For example, if a study focusses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over purely evolutionary traits.
@@
 None of these types of data are fundamentally better but each will be more appropriate for specific questions.
-For example, when investigating variation within bat wing shapes, homologous landmarks (for geometric morphometrics analysis) or continuous measurements of bones may be appropriate.
+For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometrics analysis) or continuous measurements of bones may be appropriate.
 On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.
-% NC: This section might need more explanation and/or some more examples
-
-%TG: add the bat example for above
+% NC: This section might need more explanation and/or some more examples - %TG: here's a try.
+This illustrates the importance of framing the question.
+In this above example, if the question is whether the aerodynamic properties of wings varies within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
+wingspan, aspect ratio, etc.).
+Conversely if the question focuses on phylogenetic convergence aspects between different types of active flights, it might be more pertinent to use evolutionary traits that facilitate active flights within and between both groups (e.g.
+digits length, head shape, etc.).
 
@@
     \item \textit{Sample size}
-    There is always a trade-off between collecting enough data and the time allocated to a project.
+    There is always a trade-off between collecting enough data and the time and resources allocated to a project.
     It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
@@
     Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
-    \item \textit{Are the data consistent?} %NC: Not suer I get this section. What do you mean by consistent???? Also does this fit into the "how much data" section or is it a seaprate consideration?
-    This is maybe the most important question related to the nature of the data.
-    It is essential that the data is consistent and homogeneous to make sure that the disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data set.
-    For example, are the observed differences between two groups not related to the group's specificities (e.g. sexual dimorphism)?
-    Are the elements fairly consistent and comparable them?
-    Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
-    In any of these cases, are the patterns a mere reflection of these inconsistencies?
-    For example, if one uses disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g.
+    % \item \textit{Are the data consistent?}
+        %NC: Not suer I get this section. What do you mean by consistent???? Also does this fit into the "how much data" section or is it a seaprate consideration?
+        %TG: fair enough. This was kind of the miscellaneous section steming from the discussion on data (bullet points about the importance of sexual dimorphism, etc...)
+    % This is maybe the most important question related to the nature of the data.
+    % It is essential that the data is consistent and homogeneous to make sure that the disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data set.
+    % For example, are the observed differences between two groups not related to the group's specificities (e.g. sexual dimorphism)?
+    % Are the elements fairly consistent and comparable them?
+    % Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
+    % In any of these cases, are the patterns a mere reflection of these inconsistencies?
+    % For example, if one uses disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g.
 snakes) that could not reflect processes.
 \end{itemize}
-Of course in practice some of these caveats are unavoidable. 
-Data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
-% NC: I'd be tempted to move this to the end of the section to finish it off.
+Of course in practice some of these caveats are unavoidable, especially when dealing with the vertebrate fossil record.
 
@@
 limb traits can be used for both locomotion and feeding).
-%Ultimately, whether the data was collected for the purpose of the study or is recycled, and whether any of the caveats described above are taken into account, disparity patterns will eventually be measured and analysed in another multitude of ways a we suggest that, whenever possible, this should first and foremost be considered in the design of the data collection protocol.
-%NC: Not sure these linking statements are needed
+
+Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
 

### 2018-04-11 Writing/disparity_review.tex
Suggestion for title
New reordered version of the introduction

@@
 %TG: back to the drawing board. Maybe something with "best practices" as well.
+% NC: how about something like "Analysing disparity in morphology data: best practice and future directions"
 \bigskip
@@
 
+% NC: OK I've chopped this up a lot as it felt like you were repeating yourself quite a bit (e.g.
+you said here is what kinds of analyses people did in the 90s, and what they do now, then you had a section on here is what we defined disparity as in the 1980s and here is what we do now - so I've combined it all) and could say the same thing in half the space.
+Feel free to revert back.
 \section{Introduction}
 
-Even classical scholars recognised that variation in the shape, form and functions of organisms often follows discrete patterns rather than a continuous gradients.
-These patterns can be pictured as multidimensional spaces encompassing all the possible life forms where each individual region are the result of a discrete combination of many variables.
-For example, the mammalian carnivores design space can be seen as a combination of phylogenetic variables and their associated shared traits (i.e.
-the synapomorphies of the order Carnivora) as well as a combination of ecological, life history and behavioural variables (i.e.
-the ecological guild of meat eating mammals).
+% NC: New version
+Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
+These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
+One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
 
-One commonly used approach to analyse these discrete patterns and to understand the processes that led to it is to analyse the disparity of the morphological charactersitics of the group, often referred to as morphological disparity analysis.
-This type of analysis was popularised around the 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] by studying the fossil record and focusing on broad questions such how does diversity evolve, are some body types more common that others or can shape evolve in all ``directions'' (of the multidimensional space).
-Since then, an impressive amount of literature and the equally vast amount of methods have been used and the term ``disparity'' has become a close synonym of ``analysis of patterns of morphological diversity'' and has been used in a range of sub-fields in biology.
-Disparity analysis has been used to tackle a vast array of questions in Evo-devo (Developmental constraints [CITE]; Relationship between form and function and evolution functional traits [CITE] or Scaling micro to macro [CITE];), in ecology or palaeoecology (intra-specific competition [CITE], niche occupants, speciation and competition[CITE] or macroecology (disparity in space)[CITE];)
-or in macroevolution (response to extinction and selectivity [CITE], intra-specific competitive replacements [CITE] or more globally, tempo and mode of evolution [CITE]).
+In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF).
+Early disparity analyses focused on broad questions such as How does diversity evolve?
+Are some body types more common than others?
+Can shape evolve in all ``directions'' (of the multidimensional space)?
+Since the 1980s, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity''.
+Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e.
+developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e.
+intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e.
+responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).
 
-\subsection{What \textit{is} disparity?}
+The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
+Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation (REF Katie et al's paper somewhere here).
+Here we use a broader definition and suggest that disparity is a pattern of variation.
+Throughout this review, we aim to highlight the diversity of ``disparity'' analyses and how they can aid our understanding of the structure of biodiversity.
 
-To date, there is no consensus on whether disparity should be restricted and specific to these kind of studies or whether it should be used as a broader umbrella term designating studies of multivariate variation, not restricted in its questions, methods and disciplines.
+% OLD VERSION
+\subsection{old version of intro}
+%Even classical scholars recognised that NC: I'm unsure we need this qualifier, just get right to it!
+Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
+These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
+The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
+Each individual carnivore species occupies a discrete region of this space related to its values for these traits.
+% NC: I've tried to make the example clearer, but I'm unsure if it actually helps. Could cut the examples and go straight into the next sentence? Then the intro is just one paragraph.
+One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as morphological disparity analysis.
 
-At the core of this debate a potential divide between whether disparity is a patterns or a processes: is there a difference between the disparity patterns (and how researchers characterise them) and how they were ultimately generated?
-If disparity describes patterns of variation, the underlying scientific question ultimately always needs to use them as a proxy for explaining the biological processes.
-In the seminal papers from the late 80s, disparity was used to describe and understand the Cambrian explosion where both the pattern (high dissimilarities among groups) was directly linked to the process (an explosive radiation of multicellular life as championed by Gould REF).
+This type of analysis was popularised in the 1980s and 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] through studies of the fossil record, focused on broad questions such as how does diversity evolve?
+Are some body types more common than others?
+Can shape evolve in all ``directions'' (of the multidimensional space)?
 
-Restricting disparity to this initial informal definition (multidimensional morphological dissimilarity at a macroevolutionary scale) can be advantageous for advances in this field but broadening it will allow more interdisciplinary discussion.
+In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
+The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
+Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
+There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
+Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
+The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).
+
+Since then, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity'' as a synonym for ``analysis of patterns of morphological diversity''.
+Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e.
+developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e.
+intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e.
+responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).
+
+What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
+The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
+Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
+There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
+Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
+The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).
+
+Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
 In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
 Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
+
 Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.

### 2018-04-12 Writing/disparity_review.tex
Updated data bit

@@
 
-\noindent{\Large \bf Disparity in morphological disparity analysis} 
-% NC: Not sure about this title, especially now we are just doing a mini series. I think it really needs disparity in the title somewhere.
-%TG: back to the drawing board. Maybe something with "best practices" as well.
-% NC: how about something like "Analysing disparity in morphology data: best practice and future directions"
+\noindent{\Large \bf Analysing disparity in morphology data: best practice and future directions}
+
 \bigskip
@@
 %---------------------------------------------
-
-% NC: OK I've chopped this up a lot as it felt like you were repeating yourself quite a bit (e.g.
-you said here is what kinds of analyses people did in the 90s, and what they do now, then you had a section on here is what we defined disparity as in the 1980s and here is what we do now - so I've combined it all) and could say the same thing in half the space.
-Feel free to revert back.
 \section{Introduction}
 
-% NC: New version
 Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
@@
 One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
+% The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
+% Each individual carnivore species occupies a discrete region of this space related to its values for these traits.
 
-In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF).
-Early disparity analyses focused on broad questions such as How does diversity evolve?
+In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF)[Runnegar 1987, Gould, Foote, Webb, etc.].
+Early disparity analyses focused on broad questions such as: How does diversity evolve?
 Are some body types more common than others?
@@
 
-% OLD VERSION
-\subsection{old version of intro}
-%Even classical scholars recognised that NC: I'm unsure we need this qualifier, just get right to it!
-Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
-These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
-The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
-Each individual carnivore species occupies a discrete region of this space related to its values for these traits.
-% NC: I've tried to make the example clearer, but I'm unsure if it actually helps. Could cut the examples and go straight into the next sentence? Then the intro is just one paragraph.
-One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as morphological disparity analysis.
+% \subsection{What \textit{is} disparity?} %TG: maybe useful for box or paraphrapsing summaries.
+% What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
+% The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
+% Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
+% There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
+% Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
+% The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).
 
-This type of analysis was popularised in the 1980s and 1990s [Runnegar 1987, Gould, Foote, Webb, etc.] through studies of the fossil record, focused on broad questions such as how does diversity evolve?
-Are some body types more common than others?
-Can shape evolve in all ``directions'' (of the multidimensional space)?
-
-In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
-The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
-Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
-There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
-Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
-The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).
-
-Since then, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity'' as a synonym for ``analysis of patterns of morphological diversity''.
-Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e.
-developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e.
-intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e.
-responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).
-
-What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
-The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
-Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
-There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
-Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
-The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).
-
-Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
-In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
-Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
-
-Throughout this review, we aim to highlight the diversity of disparity analyses and how they can help understanding or provide a structure to biodiversity.
-In essence, we suggest that disparity is a pattern of variation.
-Thus, disparity methods or analysis can be seen as methods or analysis to study these patterns and eventually link them to macroevolutionary or macroecological processes
+% Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
+% In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
+% Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
 
@@
 \section{Collect the appropriate data}
-Once you 
-%TG: not sure about using "you" all the time. I feel it does add some patronising tone the whole yoke.
-have a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
-For example, if a study focusses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over purely evolutionary traits.
+Once one has a clear question, it is important to collect the appropriate data to avoid some major caveats.
+For example, if a study focusses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over solely evolutionary traits.
 Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).
@@
 On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.
-% NC: This section might need more explanation and/or some more examples - %TG: here's a try.
 This illustrates the importance of framing the question.
@@
 wingspan, aspect ratio, etc.).
-Conversely if the question focuses on phylogenetic convergence aspects between different types of active flights, it might be more pertinent to use evolutionary traits that facilitate active flights within and between both groups (e.g.
-digits length, head shape, etc.).
+Conversely if the question focuses on phylogenetic convergence aspects between different types of active flights, it might be more pertinent to use evolutionary traits that would have facilitated active flights within and between both groups (e.g.
+digits length, integumentary system, etc.).
 
 \subsection{How much data to collect?}
-Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of both specimens and traits.
-There are several important considerations.
+Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of specimens and traits.
 \begin{itemize}
@@
     \item \textit{Sample size}
-    There is always a trade-off between collecting enough data and the time and resources allocated to a project.
     It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
@@
     % \item \textit{Are the data consistent?}
-        %NC: Not suer I get this section. What do you mean by consistent???? Also does this fit into the "how much data" section or is it a seaprate consideration?
+        %NC: Not sure I get this section. What do you mean by consistent???? Also does this fit into the "how much data" section or is it a seaprate consideration?
         %TG: fair enough. This was kind of the miscellaneous section steming from the discussion on data (bullet points about the importance of sexual dimorphism, etc...)
@@
 \subsection{Recycling data}
-A common practice in disparity analysis is to recycle data from other studies which can introduce additional problems.
+A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
 One classic case is using ``cladistic'' data in palaeobiology.
 These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
-This data can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
+It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
 A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all the data.
@@
 
-\section{Use appropriate methods} 
+\section{Use the appropriate methods} 
 In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
@@
 Conversely, two groups that do not overlap in some dimensions can overlap in other dimensions.
-In this example, no group overlaps in the multidimensional space despite the appearances.}} 
-% NC: Might need to explain more clearly, I'm not 100% sure I got your meaning
+In this example for the groups A and B to truly overlap, they must overlap in all three dimensions displayed here.
+Although these groups overlap in the two first dimensions (panel 1), they do not overlap between the dimension 1 and 3 thus do no overlap in the multidimensional space at all.
+The inverse is of course true as well: if two groups don't overlap between two dimensions (D and C in panel 1), they do not overlap in the multidimensional space, even if they overlap between any other combination of dimensions (dimensions 1 and 3 in panel 2).}}
 \label{Fig:RF_results_best}
@@
 disparity through time).
-Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading.
+Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading (see box \ref{box_ordination}).
 
@@
 \label{box_ordination}
-One of the most common approaches to analysing multidimensional data is to use ordination techniques.
-These consist of transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
-This is generally done via three analysis pipelines:
-\begin{enumerate}
-    \item data -> distance matrix -> ordination
-    \item data -> zscores -> principal components analysis (PCA)
-    \item data -> Procrustes superimposition -> projection into tangent space -> PCA
-\end{enumerate}
-%TG: Check with Melanie for the figure? - apparently she has one.
+% One of the most common approaches to analysing multidimensional data is to use ordination techniques.
+% These consist of transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
+% This is generally done via three analysis pipelines:
 
-%Why to ordinate: % NC: This box seems quite long, can it be reduced?
-These pipelines reduce the dimensionality of the data.
-This is advantageous for plotting and visualising the data (see Visualisation section \ref{visualisation}) which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
+% \begin{enumerate}
+%     \item data -> distance matrix -> ordination
+%     \item data -> zscores -> principal components analysis (PCA)
+%     \item data -> Procrustes superimposition -> projection into tangent space -> PCA
+% \end{enumerate}
+% %TG: Check with Melanie for the figure? - apparently she has one.
+
+Reducing the dimensionality through ordination of the data can be advantageous for plotting and visualising the data which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
 Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g.
 selecting only the axes that describe up to 95\% of the variation).
-In the case of geometric morphometric data (point 3 above), ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions: the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc.
-% NC: is the stuff after the : needed in this sentence?
+In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions.
+%the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc.%TG: removed below for now.
 This has clear advantages for interpreting the results.
 For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).
-This also allows easy interpretation of the proximity of the elements in space which reflect shared biological characters.
-%%%%%%%%%%%
+
 However, some caveats can arise from such transformations of the data.
@@
 [INCLUDE WILLS FIGURE HERE]
+
+%TG: add the curse of dimensionality
 

### 2018-04-12 Writing/disparity_review.tex
rewriting up until ancestral states

@@
 Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and might have little relation to the disparity question (e.g.
-if the first few axes represent the elongation and flattening of the beak, but the question is about wing disparity).
-In many cases, depending on the question, ordination might not be necessary.
-For example, if the disparity metric being used (see below) relies on all the data (for example the sum of variance or the sum of ranges) it is not necessary to calculate it on ordinated data [e.g.
-CITE Close].
+if the first few axes represent the elongation of the beak, but the question is about wing disparity).
+
+In many cases, ordination might not be necessary.
+For example, if the disparity metric being used (see below) relies on all the data (see Metric session) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
 Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
 For example, when the analysed data is non-Euclidean (e.g.
-discrete morphological characters), the ordination can be problematic (due to character hierarchy, inapplicable characters, polymorphisms, and missing data), and the resulting non-Euclidean ordinated space can be difficult to interpret.
+discrete morphological characters), the ordination can be problematic ([cite Melanie's future paper here]), and the resulting non-Euclidean ordinated space can be difficult to interpret ([probably cite Sylvain's here]).
 This can be problematic when comparing the position of groups in the multidimensional space, as the measured distances might not be comparable.
-Furthermore, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axes can be impossible or biologically meaningless (e.g.
-the ratio of data coded as 1 rather than 0).
-% NC: Need to explain this point more clearly
+Finally, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axes can be impossible or biologically meaningless.
+Although some gradients might be easy to detect or interpret (e.g.
+the elongation of the beak along the first axis), they can often be hard to interpret in morphological gradients terms or simply non-interpretable (e.g.
+with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more states 0).
 

### 2018-04-16 Writing/disparity_review.tex
Edits to question and data sections

@@
 
-This diversity (and disparity) of methods to analysis disparity provides biologist with a powerful and comprehensive tool %TG: comprehensive in the sense "global"/"universal" not sure if that's the right word here.
-that has a great deal of promise in helping us to understand the evolution of biological diversity.
+This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
 However, as with any great power, it comes with great responsibility. 
-With the expansion of methodological approaches in the last decade, we also identify three main issues that, if not tackled could reduce the power of such analysis.
+With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
 First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
-Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets.
+Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
 Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
-Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions before performing a disparity analysis.
+Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions % NC: Do we mean questions here or analyses or something else?
+ before performing a disparity analysis.
 
@@
 The last decade has seen a vast array of new methods being published and implemented in freely available software, making disparity analyses easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
-However, this (consciously or not) pushes authors towards performing disparity analyses simply because of their ease of use, rather than stemming from a question driven approach.
+However, this (consciously or not) pushes authors towards performing disparity analyses simply because of their ease of implementation, rather than stemming from a question driven approach.
 Before rushing into a disparity analysis, one should consider some basic caveats.
@@
 \section{Collect the appropriate data}
-Once one has a clear question, it is important to collect the appropriate data to avoid some major caveats.
-For example, if a study focusses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over solely evolutionary traits.
+Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
+For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for species delimitation.
+% NC: I changed it from evolutionary here to make it clear what an evolutionary trait would be
 Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).
@@
 \begin{enumerate}
-    \item discrete data that can be either describing a feature (often referred to as ``cladistic''; e.g.
+    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g.
 the presence or absence of legs, the colour of a tail, etc.
-[CITE]) or counts (e.g.
+[CITE]) or count the number of features (e.g.
 number of digits, etc.
@@
 body length, etc.
-[CITE]) or more abstracted methods (e.g.
+[CITE]) or more abstracted measures (e.g.
 geometric morphometrics, mathematical contour description, etc.
 [CITE]);
-    \item or any mixture of discrete and continuous data (e.g. ratios [CITE]).
+    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
 \end{enumerate}
 None of these types of data are fundamentally better but each will be more appropriate for specific questions.
-For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometrics analysis) or continuous measurements of bones may be appropriate.
-On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the contour of the wings might more readily answer the question.
-This illustrates the importance of framing the question.
-In this above example, if the question is whether the aerodynamic properties of wings varies within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
+For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
+On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
+This again illustrates the importance of framing the question.
+In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
 wingspan, aspect ratio, etc.).
-Conversely if the question focuses on phylogenetic convergence aspects between different types of active flights, it might be more pertinent to use evolutionary traits that would have facilitated active flights within and between both groups (e.g.
-digits length, integumentary system, etc.).
+Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g.
+digit length, integumentary system, etc.).
 
-\subsection{How much data to collect?}
+\subsection{How much data should be collected?}
 Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of specimens and traits.
 \begin{itemize}
-    \item \textit{Taxonomic scope}
+    \item \textit{Taxonomic scope.}
     Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
@@
     % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
-    \item \textit{Sample size}
+    \item \textit{Sample size.}
     It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
     For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
-    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction analysis [CITE]).
+    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction [CITE]).
     Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
+    \end{itemize}
+
+Of course in practice some of these caveats are unavoidable, especially when dealing with the vertebrate fossil record.
+
     % \item \textit{Are the data consistent?}
@@
         %TG: fair enough. This was kind of the miscellaneous section steming from the discussion on data (bullet points about the importance of sexual dimorphism, etc...) 
-    % This is maybe the most important question related to the nature of the data.
-    % It is essential that the data is consistent and homogeneous to make sure that the disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data set.
-    % For example, are the observed differences between two groups not related to the group's specificities (e.g. sexual dimorphism)?
-    % Are the elements fairly consistent and comparable them?
-    % Data can be missing, non overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
-    % In any of these cases, are the patterns a mere reflection of these inconsistencies?
-    % For example, if one uses disparity analysis of locomotory modes among vertebrates and the collected data is hierarchical, limb data could create an artificial grouping of non-tetrapods and limbless tetrapods (e.g.
-snakes) that could not reflect processes.
-\end{itemize}
-Of course in practice some of these caveats are unavoidable, especially when dealing with the vertebrate fossil record.
+        % NC: How about adding an "other considerations" subsection to mention all this stuff? Draft below, you may want to edit.
 
@@
 It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
-A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all the data.
-For example, in a study of disparity of dietary niches, removing non-dietary characters is a good approach.
+A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all of the data.
+For example, in a study of the disparity of dietary niches, removing non-dietary characters is a good approach.
 However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g.
 limb traits can be used for both locomotion and feeding).
+
+\subsection{Other data considerations}
+There are many other considerations when choosing the appropriate data to use. 
+It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
+For example, are the observed differences between two groups related to factors unrelated to group designation, e.g. sexual dimorphism?
+Are the taxa/groups being compared fairly consistent and truly comparable?
+In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
+It is important to ensure that any patterns detected are not merely reflections of these features of the data.
 

### 2018-04-16 Writing/disparity_review.tex
Edits to ace section

@@
 In this example for the groups A and B to truly overlap, they must overlap in all three dimensions displayed here.
-Although these groups overlap in the two first dimensions (panel 1), they do not overlap between the dimension 1 and 3 thus do no overlap in the multidimensional space at all.
-The inverse is of course true as well: if two groups don't overlap between two dimensions (D and C in panel 1), they do not overlap in the multidimensional space, even if they overlap between any other combination of dimensions (dimensions 1 and 3 in panel 2).}}
+Although these groups overlap in the two first dimensions (panel 1), they do not overlap in dimensions 1 and 3 thus do not overlap in the multidimensional space at all.
+The inverse is of course true: if two groups don't overlap in two dimensions (D and C in panel 1), they do not overlap in the multidimensional space, even if they overlap in any other combination of dimensions (dimensions 1 and 3 in panel 2).}}
 \label{Fig:RF_results_best}
@@
 We suggest visualisations should include the following:
-(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation of the data present in the plot or only a fraction?);
+(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation in the data present in the plot or only a small fraction?);
 (2) the scale of the axes (ideally, if the space is orthogonal, it is important to keep the same scale across the axes);
 and (3) the error associated with each point, whether this is in terms of measurement error (on ordination plots for example) or some other measure of standard variation (in further analysis, e.g.
-disparity through time).
+disparity-through-time plots).
 Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading (see box \ref{box_ordination}).
 
-\subsection{[Box]To ordinate or not to ordinate (that is the multidimensional question)?}
+\subsection{[Box]To ordinate or not to ordinate; that is the (multidimensional) question?}
 \label{box_ordination}
@@
 
-Reducing the dimensionality through ordination of the data can be advantageous for plotting and visualising the data which can reveal properties of the morphospace not captured by the metrics (see Metrics section \ref{metrics}).
+Reducing the dimensionality of data through ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
 Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g.
@@
 In many cases, ordination might not be necessary.
-For example, if the disparity metric being used (see below) relies on all the data (see Metric session) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
+For example, if the disparity metric being used relies on all the data (see Metrics section \ref{metrics}) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
 Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
@@
 Although some gradients might be easy to detect or interpret (e.g.
-the elongation of the beak along the first axis), they can often be hard to interpret in morphological gradients terms or simply non-interpretable (e.g.
-with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more states 0).
+the elongation of the beak along the first axis), they are often hard to interpret in terms of morphological gradients or simply non-interpretable (e.g.
+with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0 has no biological meaning).
 

### 2018-04-18 Writing/disparity_review.tex
Added abstract

@@
 
-\begin{enumerate}
-    blablalbalba
-\end{enumerate}
+Variation in organisms' shape, ecology or behaviour is often more discrete rather than continuous.
+One way to study this variation was pioneered in the 1990s by looking at the disparity of morphologies in the Cambrian.
+Morphological disparity analysis - a term encompassing both the patterns and process of morphological evolution through time - became rapidly popular and lead to numerous publications in the last two decades.
+
+Here, we suggest that this diversification of disparity analysis lead to two contrasted effects:
+(1) the great diversity of methodological approaches to study disparity makes it sometimes hard to understand all caveats and pitfalls of such analysis - a problem that can lead to major misinterpretation of disparity patterns - and;
+(2) in essence, disparity is a pattern of variation meaning that disparity analysis can be expanded beyond the study of morphology.
+
+In this review, we propose a set of non-prescriptive guidelines that will help researchers to think about specific disparity analysis designs, with their caveats and their advantages.
+Furthermore, we highlight potential avenues of expansion of disparity analysis both in terms of improving the existing methods for studying morphological disparity and in terms of expanding disparity analysis beyond morphology.
 
@@
 Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
-Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their questions % NC: Do we mean questions here or analyses or something else?
- before performing a disparity analysis.
+Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.
 
@@
 For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for species delimitation.
-% NC: I changed it from evolutionary here to make it clear what an evolutionary trait would be
 Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).
@@
 
-    % \item \textit{Are the data consistent?}
-        %NC: Not sure I get this section. What do you mean by consistent???? Also does this fit into the "how much data" section or is it a seaprate consideration?
-        %TG: fair enough. This was kind of the miscellaneous section steming from the discussion on data (bullet points about the importance of sexual dimorphism, etc...) 
-        % NC: How about adding an "other considerations" subsection to mention all this stuff? Draft below, you may want to edit.
-
 \subsection{Recycling data}
@@
 It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
-For example, are the observed differences between two groups related to factors unrelated to group designation, e.g. sexual dimorphism?
+For example, are the observed differences between two groups related to factors unrelated to group designation such as sexual dimorphism?
 Are the taxa/groups being compared fairly consistent and truly comparable?
-In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated.
+In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated [String of morpho data problems cite].
 It is important to ensure that any patterns detected are not merely reflections of these features of the data.
@@
 In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions.
-%the PCA reflects the number of degrees of freedom (resulting from the superimposition), its geometry is conserved (coordinates remain Euclidean when projected into tangent space and ordinated), etc.%TG: removed below for now.
 This has clear advantages for interpreting the results.

### 2018-04-18 Writing/disparity_review.tex
updated the metrics section

@@
 
-[INCLUDE WILLS FIGURE HERE]
+[INCLUDE WILLS FIGURE SOMWHERE AROUND]
 
@@
 
-After building and visualising the morphospace, one can use disparity metrics (or indices) as a one dimensional summary of it to test hypotheses.
-Because of these reduction in dimensionality, these metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
-With that in mind, the choice of the metric should be essential to summarise the aspect of the morphospace that is the most related to the question.
-In an unidimensional distributions (e.g. histograms) different metrics will readily reflect different aspect of the data.
-For example, metrics can be used to reflect the ``size'' of the distribution (e.g.
+After building and visualising the morphospace, one can use disparity metrics (or indices) as a one dimensional summary of the space and to test hypotheses.
+As any summary metric and because of this reduction in dimensionality, disparity metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
+It is often beneficial to use more than one such a metric to summarise the different aspects of interest of the morphospace (e.g. it's width, it's spread, etc.).
+
+The choice of the disparity metric is thus essential to the disparity analysis and should always be driven by the question.
+When considering only one dimension, disparity metrics can be used to reflect the ``size'' of the distribution (e.g.
 the range, quantiles or variance) and some other will reflect the most likely value (the central tendency - e.g.
 the mean, median or mode).
-Among these metrics, some will have more preferable properties than other such as sensitivity to outliers (range, mean or mode \textit{vs.} quantiles, variance or median) and will thus be more or less appropriate to the different questions.
+Among these metrics, some will have more preferable properties than other such as sensitivity to outliers (range, mean or mode \textit{vs.} quantiles, variance or median) and will thus be more or less appropriate to the different questions and to the actual properties distribution.
+For example, if the data collected is likely to contain outliers, but is fairly normally distributed, one should prefer the median to the mean.
 
-In multidimensionality, this high availability of unidimensional descriptors comes with equally numerous advantages and caveats.
-If the question refers to ``size'' of a group in the morphospace (e.g.
+With multidimensionality, this high availability of unidimensional descriptors comes with an equally numerous advantages and caveats and again, the question can help determining which metric would be the most appropriate.
+For example, if the question refers to ``size'' of a group in the morphospace (e.g.
 ``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriated (e.g.
@@
 Conversely, if the question refers to the ``position'' of a group in the morphospace (e.g.
-``does group A occupy the same space as group B?''), metrics relating to the distance of the elements to a point are most appropriated (e.g.
-distance from a fixed point - e.g.
-the opposite group).
+``does group A occupy the same space as group B?''), metrics relating to the distance between the elements within a group and a fixed point in the morphospace are most appropriated.
 Finally, one might also be interested in the density of elements in some regions of the morphospace (e.g.
 ``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriated (e.g.
-nearest neighbour distance, pairwise distances, etc.)
+nearest neighbour distance, pairwise distances, etc.).
 
-When looking at this positively, any metric can be suitable to any question, when looking at it negatively, it is important to keep in mind several caveats that can apply in any case:
-First, it is important to relate the metrics to the space property.
-For example, measuring the sum of variances of the space's dimensions does not require the co-variance element between the axis in a PCA but only when the full space is considered: if only a subset is measured, co-variance should be added to the sum of variances.
-Second, some metrics automatically relate to the properties of the morphospace.
-For example, range or hull based metrics are really sensitive to outliers and thus biased based on the sample size.
+Additionally to the desired properties of what the metrics should capture, it is important to also keep in mind the mathematical properties of the metrics and their associated caveats.
 
-Of course some other aspects could be of interest (or more than one) and thus some other metrics would be more appropriated.
-For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution).
-Furthermore, one future avenue that could be explored is to go beyond single valued disparity metrics and rather express them in terms of distributions.
-For example, when the metric is expressed as the median distance from centroids [e.g. in CITE], it could be informative as well to express it as a full distribution.
-This would allow to both visualise the variation at a glance and use this variation in statistical tests for comparing these distributions.
+\begin{itemize} %TG: not sure how to introduce this list. Do we even need a list? I can shorten these examples and discuss them more in my paper-project on disparity metrics?
+    \item For example, measuring the full sum of the variance of each dimensions of the space does not require to add the co-variance between the axis in a ordinated space using a PCA.
+    This is not true however in other mathematical spaces or (even in a PCA) when not all dimensions or all the elements are considered (e.g.
+when only looking at a group within the morphospace, the sum of variances is not mathematically equal to the sum of the variance of each dimensions in the space).
 
-Best practices:
-\begin{itemize}
-\item Facing the amount of different ways to measure disparity and regardless which metric is the most suited to the question, it is crucial to clearly and mathematically define the metric in a disparity paper.
+    \item Furthermore, multidimensional space have some counter-intuitive properties that needs to be taken into account such as the curse of dimensionality [CITE].
+    For example, product based metrics (e.g. product of ranges, hypervolume, hypercube, etc.) will tend towards 0 fairly quickly for spaces with more than 6 to 10 dimensions.
+
+    \item Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size such as range or convex hull based metrics.
+
+\end{itemize}
+
+%TG: again, I feel this section is already fairly long, I might include this bit in another paper
+% Of course some other aspects could be of interest (or more than one) and thus some other metrics would be more appropriated.
+% For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution).
+% Furthermore, one future avenue that could be explored is to go beyond single valued disparity metrics and rather express them in terms of distributions.
+% For example, when the metric is expressed as the median distance from centroids [e.g. in CITE], it could be informative as well to express it as a full distribution.
+% This would allow to both visualise the variation at a glance and use this variation in statistical tests for comparing these distributions.
+
+% Best practices:
+% \begin{itemize}
+% \item Facing the amount of different ways to measure disparity and regardless which metric is the most suited to the question, it is crucial to clearly and mathematically define the metric in a disparity paper.
 The definition can be in plain English: e.g.
@@
 ``we measured disparity as $\sum\sigma_i$ where $\sigma_i$ is the variance of the each ordinated axis $i$.''.
-\item It is important to keep in mind that different metrics can lead to the same results even though the described spaces are different (the opposite is \textit{not} true).
+% \item It is important to keep in mind that different metrics can lead to the same results even though the described spaces are different (the opposite is \textit{not} true).
 We therefore suggest to always plot some dimensions of the space (see Visualisation section \ref{visualisation}) and try multiple metrics.
-\end{itemize}
+% \end{itemize}
 

### 2018-05-07 Writing/disparity_review.tex
Removed old ideas
add curse of dimensionality example

@@
 \bigskip
-\noindent RH: disparate views on disparity. % NC: Needs to be changed?
+\noindent RH: pest practices for disparity analysis.
 
@@
 \label{box_ordination}
-% One of the most common approaches to analysing multidimensional data is to use ordination techniques.
-% These consist of transforming the dataset along its major axis of variation through various mathematical methods [Legender \& Legendre].
-% This is generally done via three analysis pipelines:
-
-% \begin{enumerate}
-%     \item data -> distance matrix -> ordination
-%     \item data -> zscores -> principal components analysis (PCA)
-%     \item data -> Procrustes superimposition -> projection into tangent space -> PCA
-% \end{enumerate}
-% %TG: Check with Melanie for the figure? - apparently she has one.
-
 Reducing the dimensionality of data through ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
@@
 Additionally to the desired properties of what the metrics should capture, it is important to also keep in mind the mathematical properties of the metrics and their associated caveats.
-
-\begin{itemize} %TG: not sure how to introduce this list. Do we even need a list? I can shorten these examples and discuss them more in my paper-project on disparity metrics?
-    \item For example, measuring the full sum of the variance of each dimensions of the space does not require to add the co-variance between the axis in a ordinated space using a PCA.
+For example, measuring the full sum of the variance of each dimensions of the space does not require to add the co-variance between the axis in a ordinated space using a PCA.
 This is not true however in other mathematical spaces or (even in a PCA) when not all dimensions or all the elements are considered (e.g.
 when only looking at a group within the morphospace, the sum of variances is not mathematically equal to the sum of the variance of each dimensions in the space).
-
-    \item Furthermore, multidimensional space have some counter-intuitive properties that needs to be taken into account such as the curse of dimensionality [CITE].
-    For example, product based metrics (e.g. product of ranges, hypervolume, hypercube, etc.) will tend towards 0 fairly quickly for spaces with more than 6 to 10 dimensions.
-
-    \item Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size such as range or convex hull based metrics.
-
-\end{itemize}
+Furthermore, multidimensional space have some counter-intuitive properties that needs to be taken into account such as the curse of dimensionality [CITE].
+For example, product based metrics as proxies of volumes (e.g.
+product of ranges, hypervolume, hypercube, etc.) will tend towards 0 fairly quickly for spaces even with a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem].
+Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size such as range or convex hull based metrics.
 
@@
 Maybe we don’t want to think about a null model but rather think about good models to fit. 
-We haven’t explored much about fitting models like explaining holes, effects of extinction etc…
+We haven’t explored much about fitting models like explaining holes, effects of extinction etc...
 Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!
@@
 
-% \section{Introduction}
-% \label{text:intro}
-
-% Biodiversity is not a smooth gradient of forms, functions, ecologies, behaviours, etc., but it is rather variable and discontinuous. %TG: Actually is it?
-% This is at the base of biological sciences and is linked to the concept of disparity \citep{Wills2001,Hopkins2017} with the observation that some groups of species are more similar or dissimilar than others.
-% This diversity in morphologies is in fact proper in nature and decoupled from taxonomic diversity: some groups may exhibit low diversity of species by high diversity of shapes \citep[or the other way around]{ruta2013,hopkinsdecoupling2013}.
-% In palaeobiology, this concept, has been studied since the 90s describing \citep{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution} and as been extended with modifications to macroevolutionary analysis in general since the 2000s (whether based on continuous traits; e.g.
-\citealt{Harmon961,geiger2008}; or on geometric morphometrics; e.g.
-\citealt{claude2008morphometrics,zelditch2012geometric,adams2013geomorph,adams2017geometric}).
-
-% Since then, morphological disparity has been used for testing a vast array of biological hypotheses such as: how did body plans evolve through time?
-\citep{Wesley-Hunt2005}; why does some morphologies exist and not other?
-\cite{gerber2017geometry}; what is the tempo and mode of morphological innovations through time?
-\citep{Hughes20082013}; what is the effect of mass extinction on disparity?
-\citep{halliday2016eutherian}; how does competition between groups affect disparity?
-\citep{Brusatte12092008}; how did morpho-ecological niches get filled through time?
-\citep{price2014niche}; and many more.
-% %
-% This apparent versatility of morphological disparity analysis led to a phenomenal amount of papers based on disparity methods (Fig. \ref{Fig:GoogleOccurences})!
-% %
-% However, ``a lot of disparity studies are inductive fishing trips and are not often designed to test a specific hypothesis.''
-% %
-% Indeed, compared to the vast amount of publication using morphological disparity, only a handful highlight/solve problems and propose hypothesis testing.
-
-% \begin{figure}[!htbp]
-% \centering
-%    \includegraphics[width=1\textwidth]{Figures/GoogleScholarOccurences.pdf} 
-% \caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year.
-In black, the match is in the paper and in grey, in the title.
-We collected the number of matches per year from 1960 to 2016 in Google Scholar for the terms "morphological disparity" both in the text (fuzzy matching) or in the title (exact matching).
-The data was collected on the 1st of November 2017.}
-% \label{Fig:GoogleOccurences}
-% \end{figure}
-
-% \section{What actually \textit{is} disparity?}
-
-% \cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.
-% Disparity can describe either the metric (\citealt{Wills2001}; or index \citealt{Hopkins2017}) or the whole pipeline \citep{zelditch2012geometric,lloyd2016estimating}.
-% To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
-% For the 230 relevant papers among the 500 matches, we collected the following methodological data: (1) What was the focal biological group?
-; (2) What kind of data was measured (e.g.
-landmarks, discrete data, etc.)?
-; (3) Was data collected on the full organism or not?
-; (4) How was the morphospace explicitly defined (e.g.
-PCA, PCO, MDS, etc.)?
-; (5) How was the disparity metric(s) explicitly defined?
-; (6) Which statistical test was applied to test the disparity related hypothesis?
-; (7) Was phylogeny taken into account or not?
-% We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g.
-a disparity metric was measured but not described anywhere in the paper).
-% The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).
-% To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).
-
-% \begin{figure}[!htbp]
-% \centering
-%    \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
-% \caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
-% \label{Fig:MethodsProportions}
-% \end{figure}
-
-% From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.
-
-% \begin{itemize}
-%     \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace.
-Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g.
-the sum of variance or/and the sum of ranges).
-Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.
-
-%     \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA).
-Disparity is often measured as a distance metric (e.g.
-the distance between the species and a point in the morphospace such as its centroid).
-Hypothesis are then tested using ANOVA type tests with usually no phylogenetic correction (although phylogeny is sometimes used to correct the morphospace).
-
-%     \item \textbf{The ``\texttt{approach}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance).
-Hypothesis of higher/lower disparity can then be measured using null evolutionary models.
-% \end{itemize}
-
-% %
-% The advent of these three approaches coincides with the explosion of disparity analysis (Fig. \ref{Fig:GoogleOccurences}) and suggest a high popularity of disparity analysis.
-% %
-% Unfortunately, however, few actual tests have been performed on whether any of these approaches \textit{actually} allows to tackle the classic disparity questions (see \ref{text:intro}{Introduction}).
-% Are all these methods the right tools to answer these questions are they mere ``inductive fishing trips'' used to describe the data rather than testing any of the hypothesis related to the disparity questions?
-
-% \section{Does data used in disparity analysis allows to test hypothesis?}
-% In fact, the question of which data to analyse \citep{hetherington2015cladistic} and which part of the data to use has been analysed \citep{hopkins2017well} has rarely been assessed.
-% Can the data used in disparity studies actually test the phenomenon researchers are trying to capture?
-
-% For example, when studying discrete morphological characters (often following the ``\texttt{Claddis}'' approach), the data used is often recycled data used for phylogenetic analysis.
-% These datasets are usually design to recover evolutionary history and differentiate clades and, often due to the preservations of the fossil record, are mainly based on hard tissue characters.
-% In phylogenetics, this artefact have been shown to be leading to measurable topological recovery error (for soft \textit{vs} hard characters; \citealt{sansomfossilization2013}; or dental \textit{vs} cranial characters \citealt{sansom2017dental}).
-% Therefore, is the morphospace composed of ``fossilisable'' characters enough for answering evolutionary hypothesis?
-% For example, are niche replacement hypothesis equivalent to fossilisable-character-based-niches replacement hypothesis?
-
-% Similarly, many studies in geometric morphometric are based solely on skull shape using landmark techniques combined with Procrustes superimposition \citep[the ``\texttt{geomorph}'' approach;][]{zelditch2012geometric,adams2017geometric}.
-% However, it has been shown that the whole skeleton displays different integrated modules with different specific evolutionary constraints leading to variable rates and modes of evolution (\citealt{Goswami20130254}; and this even at the cranial level \citealt{goswami2010influence}).
-% Is skull variation enough for measuring body plans variation through time for example?
-
-% In the case where disparity is measured a single continuous trait (or a collection of traits; following the ``\texttt{geiger}'' approach), for example in \cite{price2014niche}, beak and limb length as well as body mass are used to differentiate groups of birds and different zones in the morphospace can then be interpreted as different niches.
-% Is this set of traits studied in isolation sufficient to test niche occupancy hypothesis?
-
-% Additionally, in many cases, the disparity metrics and hypothesis are often based on a transformation of the data (pairwise distances, PCA, PCO, MDS, RWA, etc.).
-% Can these transformation readily reflect the raw data in our hypothesis?
-% For examples a morphospace as a pairwise matrix reflects the dissimilarity between the data, a morphospace based on an ordination reflects the variability within the data, a morphospace based on both reflects the variability within the dissimilarity, are each method reflecting that?
-
-% \section{Do the disparity metrics (or indices) really reflect variation in the studied phenomenon?}
-% From the 230 papers analysed, we found 104 unique combination's of metrics to measure disparity (Fig. \ref{Fig:MethodsProportions}).
-% Commonly, these metrics are used to approximate the volume occupancy within the morphospace \citep{Wills2001,DonohueDim,diaz2016global}.
-
-% However, it has been shown that different metrics reacts differently depending on the dataset \citep[][; and an anecdotal example in Fig. \ref{Fig:DimensionsEffect}]{Wills2001,Ciampaglio2001}.
-% For example metrics involving a product of aspects of the matrix (e.g.
-the volume, the product of the dimensions) are likely to suffer from the \textit{Curse of dimensionality} \citep{bellman1957dynamic} and will quickly tend to null values.
-% Some other metrics are not representing all the data captured in the morphospace (e.g. the sum or product of variance ignores the co-variance between dimensions).
-% Are these metrics as a description of the ordination a good proxy for the observed biological phenomenon?
-
-% \begin{figure}[!htbp]
-% \centering
-%    \includegraphics[width=1\textwidth]{Figures/DimensionsEffect.pdf} 
-% \caption{Effect of the number of dimensions on different disparity metrics (note that the sum of variance is equal to the average squared pairwise distance in this case).
-% The data used is one of the example datasets from the \texttt{dispRity R} package \citep[\texttt{BeckLee\_mat50};][]{beckancient2014,dispRityv02} composed of 50 elements and 48 dimensions obtained from a ordination (MDS) of the distance matrix of a discrete morphological character matrix.}
-% \label{Fig:DimensionsEffect}
-% \end{figure}
-
-% Despite these caveats, it is important to actually know what the metrics are measuring.
-% There has been some attempts to test that through simulations \citep{Ciampaglio2001,  gerber2017geometry} but it is still unclear which metric reflects the best the hypothesis test.
-% For example, the average squared pairwise distance (or the sum of ranges) are really popular \citep[e.g.][]{geiger2008} yet has been shown to be fairly senstive, at least in empirical data to the number of sampled taxa, the number of characters and the data completeness \citep[][; although effects were lesser in simulations]{Ciampaglio2001}.
-% More fundamentally, the question arises whether one disparity metric could be sufficient to describe changes in disparity?
-% Changes in multidimensions might not be readily grasped by unidimensional metrics!
-% For example, a given taxonomic group might have the exact same volume as another one yet be in a complete different part of the morphospace \citep[or even overlap or not depending on the dimensions;][]{davis2012acanthodes,brusatte50}.
-% In this case, it might more interesting to have multiple metrics (e.g.
-the position and the volume) [FABRE et.
-al in prep.] or to use distributions of metrics (e.g.
-what does the average distance between species or from a point really convey?
-Is the distribution not more interesting?).
-
-% Maybe we need metrics that fit explicitly our hypothesis (or fit our hypothesis directly to the metric).
-% For example, for competitive replacement, should we not measure whether group A occupies the same place as group B through time and whether that is due to competition or simply because any other place in the space cannot be occupied (and why are such spaces unoccupied)?
-
-% \section{Is the current statistical toolkit sufficient for testing disparity hypothesis?}
-% Are the methods used to test hypothesis in disparity actually testing what we think they're testing (problem of multidimensional hypothesis)
-% What does an anova or related variance based test real infirm/confirms regarding the disparity questions?
-% How do we deal with bootstrapped data and pseudo-replications?
-% Can we realistically use univariate Gaussian models to test hypothesis based on multivariate, probably non-Gaussian data? \citep{blomberg2017beyond}
-
-% Should be take phylogeny into account? \citep{polly2013phylogenetic}
-
-% \section{What are we missing?}
-
-% \subsection{Learning lessons from other fields}
-% In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.
-
-% \section{Conclusion}
-% We need a more hypothesis driven way of analysing disparity.
-% Disparity methods should be the tool for answering disparity hypothesis, not merely a description of the data.
-
-% \hl{[ADD Biological group, full organism and phylogeny in the supplementary]}
-
-% \noindent \hl{\textit{These points should be developed during the meeting}}
-
-% \subsection{Disparity data}
-% The data used for disparity methods comes from three main sources:
-% \begin{itemize}
-%     \item Continuous data such as limb or body measurements \citep{slaterCetacean}.
-%     \item Discrete morphological characters \citep[sometimes referred to as ``Cladistic'' characters; ][]{Brusatte12092008}.
-%     \item Geometric morphometric data which is generally the Procrustes transformation of 2D or 3D landmarks \citep{cooney2017mega}.
-% \end{itemize}
-
-% However, data was by far not limited to these three categories (e.g.
-colours \citealt{maia2013key}; metabolic rate \citealt{nespolo2016studying} or chemicals signal \citealt{garcia2017heterogeneous}).
-
-% \subsection{Morphospaces}
-% \begin{itemize}
-%     \item PCAs [CITE].
-%     \item PCOs [CITE]
-%     \item Pairwise distance matrices \citep[no ordination; ][]{Harmon961}.
-% \end{itemize}
-
-% But also RWA [CITE] or Raup-spaces [CITE].
-
-% \subsection{Disparity metrics}
-% Throughout the 230 analysed papers, we found 103 unique combinations of metrics!
-
-% \begin{itemize}
-%     \item Distances measurements between species or from a certain point in the morphospace [CITE].
-%     \item Ranges and variances of each axis of the morphospace \cite{Wills2001,Ciampaglio2001}
-%     \item Distance between species based on the pairwise distance matrix (not the ordinations) [CITE].
-% \end{itemize}
-
-% But also metrics based on the volume [CITE], the characters dissimilarity [CITE] or the coordinates of some axis of the ordination [CITE]
-
-% \subsection{Disparity hypothesis}
-% Describing the many outputs (what and how is it tested):
-
-% \begin{itemize}
-%     \item Variance based (ANOVA, etc.) [CITE].
-%     \item Correlation based [CITE].
-%     \item Regression based (PGLS, etc.) [CITE].
-% \end{itemize}
-
-% But also discriminant analysis [CITE] and clustering [CITE].
-
-% Of course some studies use a combination of these three methods or none of them at all!
-
-% Also, among each category XX\% of studies use multiple approaches.
-
-% [ADD Biological group, full organism and phylogeny in the supplementary]
-
-% \section{Expanding disparity}
-
-% How to compare disparity between groups? Is disparity relative?
-
-% How to compare disparity between methods?
-
-% Can we really say things about competition when looking at disparity in a single group?
-
-% Are all characters equal? Character contingency suggests otherwise
-
-% What is the relationship between disparity and tree shape?
-
-% Do we have a null model for investigating disparity?
-
-% Can we really say things about competition when looking at disparity in a single group?
-
-% \noindent \hl{\textit{Disparity in other fields}}
-
-% In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.
-
-% \section{Conclusion}
-
-% A quick guideline for good disparity analysis:
-
-% Maybe we need something like in Parham et al 2012 (Best Practices for Justifying Fossil Calibrations): an easy an identifiable description of the pipeline containing: 1) the type of data, 2) the morphospace and 3) the metric?
-
 \section{Authors contribution}

### 2018-05-07 Writing/disparity_review.tex
tests section

@@
 \subsection{Tests}
-Similarly to all the points covered above, which statistical test to apply solely depends on which question is tackled with disparity analysis (i.e. which hypothesis to test).
-On the other hand, some tests (such as permanovas [CITE]) are actually \textit{not} taking into account the disparity measured but are simply based on the distribution of the taxa into the space (or simply their pairwise distances in the case of PERMANOVAS).
+Similarly to all the points covered above, the statistical test to apply solely depends on which question is tackled with disparity analysis (i.e. which hypothesis to test).
+Considering the question, the data, the ordination method used (or the absence thereof) and the metric used, the statistical test should be based on a clearly state hypothesis.
+This might seem obvious but we believe doing so greatly enhances the framing of the analysis and its comprehension by future readers.
 
-Furthermore, bootstraps are often used in morphological disparity analysis as a mean to get a distribution from a single disparity value [CITE].
-While this method is valid in itself, it comes with several caveats.
-First, it is important to consider that bootstrapping the disparity values creates a distribution of them (e.g. a distribution of sum of variances from a bootstrapped morphospace).
-Therefore, the error associated with the bootstrap is reflecting the ``stability'' of the metric, not the extremes of the distribution of the data (\textit{c.f.} when bootstrapping a disparity distribution; e.g.
-the distribution of all the distances from centroids from a bootstrapped morphospace).
-In the first and most common case, the amount of bootstraps artificially increases the type I error.
-Furthermore, when comparing two groups, the comparison is one of the disparity of two groups using their ``stability'' as variance.
+For example, one commonly used test in morphological analysis, is the non parametric permutation anovas [NPANOVA or PERMANOVA; CITE Anderson + examples].
+This test and its implementations is perfectly valid to test whether two groups have a different variance/covariance but is based on the pairwise distances of the group elements.
+This test will ignore any eventual ordination (if the data was ordinated) and is not based on a uni-dimensional disparity metric.
+In other words, if one observes no overlap between two groups on the two first axis of a PCA, the hypothesis H0 that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance based in a ``distance-space''.
+Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the disparity analysis question.
 
-Best practices:
-\begin{itemize}
-\item make sure your metrics are capturing what you are testing
-\item check out the test assumptions
-\item Know if and how your dataset is violating assumptions.
-\end{itemize}
+Another caveat for applying statistical tests on disparity data is on which data to actually apply the test.
+In fact, in morphological disparity analysis (especially for palaebiological questions), data is often bootstrapped.
+This has two useful advantages: first, when the disparity metric is unidimensional (e.g.
+the sum of variances or the average squared pairwise distance), bootstrapping the data will generate a distribution of this unidimensional metric that can then be analysed using the vast statistical toolkit to compare distributions; second, when the data is scarce (e.g.
+for palaeontological data), bootstrapping the data will allow to introduce statistical variance and will render the test less sensitive to outliers.
+However, bootstrapping the data also comes with an important caveat: the data analysed is pseudo-replicate and thus non-independant.
+This can violate most of the parametric statistical tests assumptions.
+Furthermore, it is important to remember that the number of bootstrap pseudo-replicates will inevitably increase the type I error.
 
-\subsubsection{Models}
-What are our null hypotheses?
-Joe’s simulations’ stuff - phylogeny can be used to build a null.One basic null is that disparity increases with diversity; how do we diverge from that (show these morpho/genome distance divergence/convergence).
-We can have a null model that is based on simple Brownian increase in disparity (gradual). 
-Null: is the null increase in disparity dependent on a random walk and on a tree shape?
-Maybe we don’t want to think about a null model but rather think about good models to fit. 
-We haven’t explored much about fitting models like explaining holes, effects of extinction etc...
-Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!
+\section{Future directions}
 
@@
 
-\section{Where do we go from there?}
+\subsubsection{Disparity models}
+What are our null hypotheses?
+Joe’s simulations’ stuff - phylogeny can be used to build a null.One basic null is that disparity increases with diversity; how do we diverge from that (show these morpho/genome distance divergence/convergence).
+We can have a null model that is based on simple Brownian increase in disparity (gradual). 
+Null: is the null increase in disparity dependent on a random walk and on a tree shape?
+Maybe we don’t want to think about a null model but rather think about good models to fit. 
+We haven’t explored much about fitting models like explaining holes, effects of extinction etc...
+Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!
+
+\subsection{Where do we go from there?}
 4 - This a list of problems for the future/moving forward
@@
 
-``We asked whether we could but not whether we should'' (Jurassic Park)
+%Concluding words:
+Although disparity analysis have been rendered easy to perform from the user end (CITE packages tools), it is crucial to remember that they are first and foremost multidimensional analysis; and multidimensional analysis are complex.
+In Jurassic Park, Dr Ian Malcolm summarises this problem in an elegant way: ``[We] scientists were so preoccupied with whether or not [we] could that [we] didn't stop to think if [we] should.''
+We believe that morphological analysis in general will greatly benefit in emphasising the methodological decisions (\textit{whether we should}) rather than simply using disparity analysis because \textit{we can}.
 

### 2018-05-07 Writing/disparity_review.tex
adapting our methods

@@
 For example, product based metrics as proxies of volumes (e.g.
-product of ranges, hypervolume, hypercube, etc.) will tend towards 0 fairly quickly for spaces even with a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem].
+product of ranges, hypervolume, hypercube, etc.) will tend towards 0 fairly quickly for spaces even with a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem and Donoho 2000 Aide-Memoire].
 Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size such as range or convex hull based metrics.
@@
 for palaeontological data), bootstrapping the data will allow to introduce statistical variance and will render the test less sensitive to outliers.
-However, bootstrapping the data also comes with an important caveat: the data analysed is pseudo-replicate and thus non-independant.
+However, bootstrapping the data also comes with an important caveat: the data analysed is pseudo-replicate and thus non-independent.
 This can violate most of the parametric statistical tests assumptions.
@@
 \section{Future directions}
+As mentioned above, there is no ``one-size-fit-all'' morphological disparity analysis pipeline.
+As in any multidimensional analysis, there are a lot of nuances in which data to use and how to analyse it, all stemming from the explicit formulation of the disparity hypothesis.
+To understand these nuances, it is important to remember that: (1) morphological disparity analysis are not new in essence and much effort has been made in developing the our methods and (2) multidimensional analysis are by far not restricted to morphological data analysis and solutions to specific problems can be found in the literature not focused on morphology (especially ecology and mathematics).
 
-\subsection{Do we need to adapt our methods?}
-Are there lessons/methods from other fields we use?
-Sample sizes: we need to be careful about them! (e.g. for detecting holes etc.).
-Time series: same! Palaeontology data is not readily available (hence ancestral estimates) and might not be appropriate for doing time series analysis (across long time scales at least).
-Caveats of disparity
-Statistical issues e.g. covariance
-95\% or all PCs?
-Phylogenetic correlations
-Avoid using shiny new methods unless you understand
-Outline assumptions of different approaches--make clear caveats
+We believe that many of the caveats in morphological disparity analysis can arise simply by applying methodological pipelines (often new and powerful ones) but at the expense of sometimes loosing sight of the actual biological question tested in the analysis.
+We therefore advice to keep the methodology simple and tractable, at least in the preliminary analysis.
+Also, many of the caveats and technical problems in disparity analysis have been tackled in other fields namely ecology.
+In fact, ecological questions are often closely related to evolutionary ones and ecological data is also often multidimensional [CITE DONOHUE's papers].
 
-\subsubsection{Disparity models}
-What are our null hypotheses?
-Joe’s simulations’ stuff - phylogeny can be used to build a null.One basic null is that disparity increases with diversity; how do we diverge from that (show these morpho/genome distance divergence/convergence).
-We can have a null model that is based on simple Brownian increase in disparity (gradual). 
-Null: is the null increase in disparity dependent on a random walk and on a tree shape?
-Maybe we don’t want to think about a null model but rather think about good models to fit. 
-We haven’t explored much about fitting models like explaining holes, effects of extinction etc...
-Dealing with uncertainty? Bootstrap Bayesian G matrices \& P matrices!
+\subsection{How can we adapt our methods?}
+Each of the points outlined above can be the start of an entire new methodological question (and often already are).
+Although it is impossible to cover the state-of-the-art exhaustively for each of these points here are several solutions and/or directions for adapting our methodological choices:
+
+\begin{itemize}
+    \item{Which data?}
+    [One line abstract of Melanie's paper on inapplicable data].
+    Furthermore, some simple methods already exist to correct for caveats in the datasets. 
+    For example, to correct for the scarcity of the fossil record, on can use methods as simple as rarefaction analysis to take sample size into account [CITE].
+    \item{How to visualise it?}
+    [One line abstract of Sylvain's paper on morphospaces plottings].
+    One major problem of such multidimensional analyses is that treating with high dimensional spaces is hard (or even impossible) to conceptualise.
+    One solution to that could be to use more simple 2D/3D ordinations.
+    For examples using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation [e.g Raup-spaces CITE].
+    \item{What to measure?}
+    This question, although crucial might actually be the easier to tackle by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combinations of properties of interest in the morphospace (for example, focusing on both the size and a position of a group in the morphospace).
+    \item{How to test hypothesis?}
+    Finally, it is possible to use non-parametric methods to test specific disparity questions.
+    For example, using the extant of available software and computer power available, it could be possible to propose specific null models on the expectancy of what disparity results would be expected to look like under such or such model.
+    This has been already successfully used in ecology (DIAZ paper) and is currently implemented in palaeobiology (Mark stuff).
+\end{itemize}
 

### 2018-05-16 Writing/disparity_review.tex
Updated conclusion

@@
 As in any multidimensional analysis, there are a lot of nuances in which data to use and how to analyse it, all stemming from the explicit formulation of the disparity hypothesis.
-To understand these nuances, it is important to remember that: (1) morphological disparity analysis are not new in essence and much effort has been made in developing the our methods and (2) multidimensional analysis are by far not restricted to morphological data analysis and solutions to specific problems can be found in the literature not focused on morphology (especially ecology and mathematics).
+To understand these nuances, it is important to remember that: (1) morphological disparity analysis are not new and much effort has been made in developing the our methods and (2) multidimensional analysis are by far not restricted to morphological data analysis and solutions to specific problems can be found in the literature that is not necessary focused on morphology (especially ecology and mathematics).
 
-We believe that many of the caveats in morphological disparity analysis can arise simply by applying methodological pipelines (often new and powerful ones) but at the expense of sometimes loosing sight of the actual biological question tested in the analysis.
+We believe that many of the caveats in morphological disparity analysis can arise when ``blindly'' applying methodological pipelines that can come at the expense of loosing sight of the actual biological question tested.
 We therefore advice to keep the methodology simple and tractable, at least in the preliminary analysis.
@@
 \subsection{How can we adapt our methods?}
-Each of the points outlined above can be the start of an entire new methodological question (and often already are).
+Each of the points outlined above can be the start of new methodological questions (and often already are).
 Although it is impossible to cover the state-of-the-art exhaustively for each of these points here are several solutions and/or directions for adapting our methodological choices:
@@
     \item{What to measure?}
-    This question, although crucial might actually be the easier to tackle by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combinations of properties of interest in the morphospace (for example, focusing on both the size and a position of a group in the morphospace).
+    This question, although crucial might actually be the easier to tackle by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combinations of properties of interest in the morphospace.
+    For example, when reducing a multidimensional variation to one metric, it could be judicious to indicate what this metric represents the results (e.g.
+in a plot or a table, if the metric used was the sum of variances, labelling it as ``size of the group'' rather than ``disparity'').
+    This will be especially advantageous when using multiple disparity metrics (e.g. ``size of the group'' and ``position of the group'').
     \item{How to test hypothesis?}
@@
     For example, using the extant of available software and computer power available, it could be possible to propose specific null models on the expectancy of what disparity results would be expected to look like under such or such model.
-    This has been already successfully used in ecology (DIAZ paper) and is currently implemented in palaeobiology (Mark stuff).
+    This has been already successfully used in ecology [CITE Diaz paper] and is currently implemented in palaeobiology (Mark stuff).
 \end{itemize}
@@
 \subsection{Where do we go from there?}
-4 - This a list of problems for the future/moving forward
-    Simulations 
-    Form/function
-Isospaces, glottospaces, acoustospace, ecospace, life-history-spaces, etc. All just a multidimensional analysis? Needs for terms?
+Morphological disparity analysis are a major tool in evolutionary biology proposing a good and more versatile alternative to studying species richness (especially in macroevolutionary studies).
+We believe that morphological disparity analyses, as the analysis of patterns of morphological differences can be of great use for linking evolutionary and ecological processes (e.g.
+when using morphological disparity as a proxy for ecological niches to test competition through time hypotheses).
+
+However, as demonstrated above, such analyses would greatly benefit both more development and (mainly) that more care is put into the study design so that the analysis actually reflects the biological question at hand.
+This can not only be done through the points highlighted above but also through linking the observed patterns to both explicit process (for example, the link between organism's form and function) or by putting care into developing null models to potentially explain the observed patterns.
+
+Finally, we encourage to a more cross-disciplinary approach to morphological disparity analyses.
+In fact, as mentioned above, morphological disparity could greatly benefit from advances in multidimensional analysis in different fields (namely ecology) but similarly, the concept of a morphospace could easily be exported to other fields.
+For example, the multidimensional analysis in Diaz [CITE] looking at the patterns of form and functions in plants can be thought as a morphospace or a life-history-space (an ecospace?);
+isotopic analysis in [CITE Andrew's stuff] can be represented as a isotope-space, etc.
+And these generalisation could of course also be exported for any set of traits (e.g.
+acoustospaces for accoustic traits) and that even beyond evolutionary biology (e.g.
+glottospaces for linguistic traits) [CITE Katie's review].
 
@@
 The Royal Society.
-TG acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
 

### 2018-05-18 Writing/disparity_review.tex
Edits to metrics

@@
 \subsection{Phylogeny}
-For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including estimates of ancestors or to add phylogenetic structure to the data.
+For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
 For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
@@
 Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e.
-the relationship between the points not estimated) and implies longer computational method but once the morphospace is defined, it will remain constant (i.e.
-its properties won't change).
-Post ordination ancestral states estimation will not change the ordinated space geometry and will be faster to compute, however, it will change the morphopspace's properties such as the number of parameters (i.e.
-elements in the space - this can be problematic for statistical tests down the line) or the variance per axis [CITE Graeme's paper to be?].
+the relationship between the points not estimated) and implies longer computational times but, once the morphospace is defined, it will remain constant (i.e.
+its properties will not change).
+Post ordination ancestral state estimation will not change the ordinated space's geometry and will be faster to compute, however, it will change the morphospace's properties such as the number of parameters (i.e.
+elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper to be?].
 
-More importantly, however, one needs to keep in mind that using such methods can come with several caveats.
-First and foremost, although ancestral states estimations are valid methods in macroevolution, they results are highly dependent on the data and chosen methods.
-Secondly, more specific to disparity analysis, when using post-transformation ancestral states estimation, the morphospace can be saturated leading pairwise distance bases metrics to tend to zero.
-In general, using ancestral states estimation can bring much insights for recovering patterns of changes in disparity but should not be ``used'' simply to generate extra data points for statitstical reasons.
-In fact, these extra points are not independent and can also generate side effects which can be problematic, especially when testing the effect of extinctions events on disparity.
+More importantly, however, one needs to remember that using such ancestral state estimation methods comes with several caveats.
+First and foremost, although ancestral state estimations are valid methods in macroevolution, their results are highly dependent on the data and chosen methods.
+Secondly, more specific to disparity analysis, when using post-transformation ancestral state estimates, the morphospace can be saturated leading pairwise distance metrics to tend to zero.
+In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points for statistical reasons.
+In fact, these extra points are not independent and can also generate side effects that can be problematic, especially when testing the effect of extinction events on disparity.
 
-Alternatively, ancestral states estimation can be used to provide realistic null expectations from the data.
-In disparity through time analysis, it is possible to sample many ancestral states estimates on the nodes and along to branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than the morphological changes.
-Furthermore, this method can be pertinent for fitting modes of evolution and see if the recovered disparity metric can be explained by the tested modes of evolution.
+Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
+In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes.
+Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution.
 
@@
 We believe that morphological disparity analyses, as the analysis of patterns of morphological differences can be of great use for linking evolutionary and ecological processes (e.g.
-when using morphological disparity as a proxy for ecological niches to test competition through time hypotheses).
+when using morphological disparity as a proxy for ecological niches to test competition-through-time hypotheses).
 

### 2018-05-18 Writing/disparity_review.tex
Edits to Tests

@@
 After building and visualising the morphospace, one can use disparity metrics (or indices) as a one dimensional summary of the space and to test hypotheses.
-As any summary metric and because of this reduction in dimensionality, disparity metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
-It is often beneficial to use more than one such a metric to summarise the different aspects of interest of the morphospace (e.g. it's width, it's spread, etc.).
+As with any summary metric, and because of this reduction in dimensionality, disparity metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
+It is often beneficial to use more than one metric to summarise the different aspects of interest of the morphospace (e.g. its width or spread).
 
-The choice of the disparity metric is thus essential to the disparity analysis and should always be driven by the question.
+The choice of disparity metric is thus essential to the disparity analysis and should always be driven by the question.
 When considering only one dimension, disparity metrics can be used to reflect the ``size'' of the distribution (e.g.
-the range, quantiles or variance) and some other will reflect the most likely value (the central tendency - e.g.
+the range, quantiles or variance), or reflect the most likely value (the central tendency, i.e.
 the mean, median or mode).
-Among these metrics, some will have more preferable properties than other such as sensitivity to outliers (range, mean or mode \textit{vs.} quantiles, variance or median) and will thus be more or less appropriate to the different questions and to the actual properties distribution.
+Among these metrics, some will have more preferable properties than others, such as sensitivity to outliers (range, mean or mode are highly sensitive \textit{vs.} quantiles, variance or median are less sensitive) and will thus be more or less appropriate to the different questions and the properties of the distribution.
 For example, if the data collected is likely to contain outliers, but is fairly normally distributed, one should prefer the median to the mean.
 
-With multidimensionality, this high availability of unidimensional descriptors comes with an equally numerous advantages and caveats and again, the question can help determining which metric would be the most appropriate.
+With multidimensionality, this high availability of unidimensional descriptors comes with equally numerous advantages and caveats and again, the question can help in determining which metric is the most appropriate.
 For example, if the question refers to ``size'' of a group in the morphospace (e.g.
-``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriated (e.g.
+``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriate (e.g.
 volume [CITE], variance [CITE], range [CITE], distance from the centroid [CITE]).
 Conversely, if the question refers to the ``position'' of a group in the morphospace (e.g.
-``does group A occupy the same space as group B?''), metrics relating to the distance between the elements within a group and a fixed point in the morphospace are most appropriated.
+``does group A occupy the same space as group B?''), metrics relating to the distance between the elements within a group and a fixed point in the morphospace are most appropriate.
 Finally, one might also be interested in the density of elements in some regions of the morphospace (e.g.
-``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriated (e.g.
+``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriate (e.g.
 nearest neighbour distance, pairwise distances, etc.).
@@
 Additionally to the desired properties of what the metrics should capture, it is important to also keep in mind the mathematical properties of the metrics and their associated caveats.
-For example, measuring the full sum of the variance of each dimensions of the space does not require to add the co-variance between the axis in a ordinated space using a PCA.
-This is not true however in other mathematical spaces or (even in a PCA) when not all dimensions or all the elements are considered (e.g.
-when only looking at a group within the morphospace, the sum of variances is not mathematically equal to the sum of the variance of each dimensions in the space).
-Furthermore, multidimensional space have some counter-intuitive properties that needs to be taken into account such as the curse of dimensionality [CITE].
-For example, product based metrics as proxies of volumes (e.g.
-product of ranges, hypervolume, hypercube, etc.) will tend towards 0 fairly quickly for spaces even with a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem and Donoho 2000 Aide-Memoire].
-Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size such as range or convex hull based metrics.
+For example, measuring the full sum of the variance of each dimension of the space does not require one to add the co-variance between the axes in a ordinated space using a PCA.
+This is not true however in other mathematical spaces or (even in a PCA) when not all dimensions or elements are considered (e.g.
+when only looking at a group within the morphospace, the sum of variances is not mathematically equal to the sum of the variance of each dimension in the space).
+Furthermore, multidimensional space has some counter-intuitive properties that need to be taken into account such as the curse of dimensionality [CITE].
+For example, product-based metrics as proxies of volumes (e.g.
+product of ranges, hypervolume, hypercube, etc.) will tend towards zero fairly quickly for spaces with even a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem and Donoho 2000 Aide-Memoire].
+Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size, for example range or convex hull based metrics.
 
 %TG: again, I feel this section is already fairly long, I might include this bit in another paper
+
+% NC: Sounds reasonable. This could be cut down more, but for now I've just edited the grammar and spelling so we can look at the thing as a whole and also get Phil's input.
+
 % Of course some other aspects could be of interest (or more than one) and thus some other metrics would be more appropriated.

### 2018-05-18 Writing/disparity_review.tex
Edits down to Future

@@
 \subsection{Tests}
-Similarly to all the points covered above, the statistical test to apply solely depends on which question is tackled with disparity analysis (i.e. which hypothesis to test).
-Considering the question, the data, the ordination method used (or the absence thereof) and the metric used, the statistical test should be based on a clearly state hypothesis.
-This might seem obvious but we believe doing so greatly enhances the framing of the analysis and its comprehension by future readers.
+The final stage of most disparity analyses is statistical testing of the disparity metrics, to determine whether changes thought time or between groups are statistically significant. 
+Similar to the points above, the statistical test to apply depends on which question is being tackled with disparity analysis, i.e. which hypothesis is being tested.
+It also requires consideration of the data, the ordination method used (or the absence thereof) and the metric used.
 
-For example, one commonly used test in morphological analysis, is the non parametric permutation anovas [NPANOVA or PERMANOVA; CITE Anderson + examples].
-This test and its implementations is perfectly valid to test whether two groups have a different variance/covariance but is based on the pairwise distances of the group elements.
-This test will ignore any eventual ordination (if the data was ordinated) and is not based on a uni-dimensional disparity metric.
-In other words, if one observes no overlap between two groups on the two first axis of a PCA, the hypothesis H0 that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance based in a ``distance-space''.
-Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the disparity analysis question.
+For example, one commonly used test in morphological analysis, is the non parametric permutation analysis of variance [NPANOVA or PERMANOVA; CITE Anderson + examples].
+This test and its implementation is perfectly valid to test whether two groups have a different variance/covariance, but is based on the pairwise distances of the group elements.
+This test will ignore any ordination (if the data was ordinated) and is not based on a unidimensional disparity metric.
+In other words, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
+Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the question.
 
-Another caveat for applying statistical tests on disparity data is on which data to actually apply the test.
-In fact, in morphological disparity analysis (especially for palaebiological questions), data is often bootstrapped.
+Another caveat for applying statistical tests on disparity data is that it is important to consider which data to apply the test on.
+For example, in morphological disparity analysis (especially for palaeobiological questions), data is often bootstrapped.
 This has two useful advantages: first, when the disparity metric is unidimensional (e.g.
-the sum of variances or the average squared pairwise distance), bootstrapping the data will generate a distribution of this unidimensional metric that can then be analysed using the vast statistical toolkit to compare distributions; second, when the data is scarce (e.g.
-for palaeontological data), bootstrapping the data will allow to introduce statistical variance and will render the test less sensitive to outliers.
-However, bootstrapping the data also comes with an important caveat: the data analysed is pseudo-replicate and thus non-independent.
-This can violate most of the parametric statistical tests assumptions.
-Furthermore, it is important to remember that the number of bootstrap pseudo-replicates will inevitably increase the type I error.
+the sum of variances or the average squared pairwise distance), bootstrapping the data generates a distribution of this unidimensional metric that can then be analysed using the vast statistical toolkit available for comparing distributions; second, when the data is scarce (e.g.
+for palaeontological data), bootstrapping the data allows users to introduce statistical variance, rendering the test less sensitive to outliers.
+However, bootstrapping the data also comes with an important caveat: the data analysed are pseudo-replicated and thus non-independent.
+This can violate the assumptions of parametric statistical tests.
+Furthermore, it is important to remember that the number of bootstrap pseudo-replicates will inevitably increase the Type I error.
 

### 2018-05-18 Writing/disparity_review.tex
Edits to adpt

@@
 \section{Future directions}
-As mentioned above, there is no ``one-size-fit-all'' morphological disparity analysis pipeline.
-As in any multidimensional analysis, there are a lot of nuances in which data to use and how to analyse it, all stemming from the explicit formulation of the disparity hypothesis.
-To understand these nuances, it is important to remember that: (1) morphological disparity analysis are not new and much effort has been made in developing the our methods and (2) multidimensional analysis are by far not restricted to morphological data analysis and solutions to specific problems can be found in the literature that is not necessary focused on morphology (especially ecology and mathematics).
+As mentioned above, there is no ``one-size-fits-all'' morphological disparity analysis pipeline.
+As with any multidimensional analysis, there are many nuances in deciding which data to use and how to analyse it, stemming from the explicit formulation of the disparity hypothesis.
+To understand these nuances, it is important to remember that: (1) morphological disparity analyses are not new and much effort has been made in developing methods; and (2) multidimensional analyses are not restricted to morphological data analysis and solutions to specific problems can be found in literature that is not necessary focused on morphology (especially ecology and mathematics).
 
-We believe that many of the caveats in morphological disparity analysis can arise when ``blindly'' applying methodological pipelines that can come at the expense of loosing sight of the actual biological question tested.
-We therefore advice to keep the methodology simple and tractable, at least in the preliminary analysis.
-Also, many of the caveats and technical problems in disparity analysis have been tackled in other fields namely ecology.
-In fact, ecological questions are often closely related to evolutionary ones and ecological data is also often multidimensional [CITE DONOHUE's papers].
+We believe that many of the caveats in morphological disparity analysis can arise when ``blindly'' applying methodological pipelines at the expense of losing sight of the  biological question being tested.
+We therefore advise keeping the methodology simple and tractable, at least in the preliminary analysis.
+Also, many of the caveats and technical problems in disparity analyses have been tackled in other fields, mainly ecology.
+This is perhaps unsurprisingly as ecological questions are often closely related to evolutionary questions, and ecological data is also often multidimensional [CITE DONOHUE's papers].
 

### 2018-05-18 Writing/disparity_review.tex
Edits to end

@@
 \subsection{How can we adapt our methods?}
-Each of the points outlined above can be the start of new methodological questions (and often already are).
-Although it is impossible to cover the state-of-the-art exhaustively for each of these points here are several solutions and/or directions for adapting our methodological choices:
+Each of the points outlined above may drive new methodological questions (and often already are).
+Although it is impossible to cover the state-of-the-art exhaustively for each of these points here are several solutions and/or directions for adapting existing methods:
 
 \begin{itemize}
-    \item{Which data?}
-    [One line abstract of Melanie's paper on inapplicable data].
-    Furthermore, some simple methods already exist to correct for caveats in the datasets. 
-    For example, to correct for the scarcity of the fossil record, on can use methods as simple as rarefaction analysis to take sample size into account [CITE].
-    \item{How to visualise it?}
+    \item{Which data should we use in disparity analyses?}
+    %[One line abstract of Melanie's paper on inapplicable data].
+    Furthermore, some simple methods already exist to account for caveats in data. 
+    For example, to correct for the paucity of the fossil record, one can use methods like rarefaction analysis to take sample size into account [CITE].
+    % NC: This item currently doesn't seem to actually give advice about which data to use...
+    \item{How should we visualise disparity?} % NC: Tried to make the bullets a bit clearer, might need to edit for correctness and consistency with the rest of the document.
     [One line abstract of Sylvain's paper on morphospaces plottings].
-    One major problem of such multidimensional analyses is that treating with high dimensional spaces is hard (or even impossible) to conceptualise.
-    One solution to that could be to use more simple 2D/3D ordinations.
-    For examples using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation [e.g Raup-spaces CITE].
-    \item{What to measure?}
-    This question, although crucial might actually be the easier to tackle by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combinations of properties of interest in the morphospace.
-    For example, when reducing a multidimensional variation to one metric, it could be judicious to indicate what this metric represents the results (e.g.
+    One major problem of multidimensional analyses is that highly dimensional spaces are hard (or even impossible) to conceptualise.
+    One solution to this is to use more simple 2D/3D ordinations.
+    For example, using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation [e.g Raup-spaces CITE].
+    \item{Which disparity metrics should we use?}
+    This can be tackled by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combination of properties of interest in the morphospace.
+    For example, when reducing a multidimensional space to one metric, it would be judicious to indicate what this metric represents in the results (e.g.
 in a plot or a table, if the metric used was the sum of variances, labelling it as ``size of the group'' rather than ``disparity'').
-    This will be especially advantageous when using multiple disparity metrics (e.g. ``size of the group'' and ``position of the group'').
-    \item{How to test hypothesis?}
+    This is especially advantageous when using multiple disparity metrics (e.g. ``size of the group'' and ``position of the group'').
+    \item{How should we test disparity hypotheses?}
     Finally, it is possible to use non-parametric methods to test specific disparity questions.
-    For example, using the extant of available software and computer power available, it could be possible to propose specific null models on the expectancy of what disparity results would be expected to look like under such or such model.
-    This has been already successfully used in ecology [CITE Diaz paper] and is currently implemented in palaeobiology (Mark stuff).
+    For example, using available software and computer power, it is possible to propose specific null models for the expectation of what disparity would be expected to look like under various null models.
+    This has been already successfully used in ecology [CITE Diaz paper] and is currently being implemented in palaeobiology (Mark stuff).
 \end{itemize}
 
-\subsection{Where do we go from there?}
-Morphological disparity analysis are a major tool in evolutionary biology proposing a good and more versatile alternative to studying species richness (especially in macroevolutionary studies).
-We believe that morphological disparity analyses, as the analysis of patterns of morphological differences can be of great use for linking evolutionary and ecological processes (e.g.
+\subsection{Where do we go from here?}
+Morphological disparity analyses are a major tool in evolutionary biology, posing a more versatile alternative to studying species richness, especially in macroevolutionary studies.
+We believe that morphological disparity analyses, as the analysis of patterns of morphological differences, can be of great use for linking evolutionary and ecological processes (e.g.
 when using morphological disparity as a proxy for ecological niches to test competition-through-time hypotheses).
 
-However, as demonstrated above, such analyses would greatly benefit both more development and (mainly) that more care is put into the study design so that the analysis actually reflects the biological question at hand.
-This can not only be done through the points highlighted above but also through linking the observed patterns to both explicit process (for example, the link between organism's form and function) or by putting care into developing null models to potentially explain the observed patterns.
+However, as demonstrated above, such analyses would greatly benefit from more development, and  more care is needed in study design so that the analysis actually reflects the biological question at hand.
+This can not only be done through the points highlighted above but also through linking the observed patterns to an explicit process (for example, the link between an organism's form and function) or by developing null models to potentially explain the observed patterns.
 
@@
 In fact, as mentioned above, morphological disparity could greatly benefit from advances in multidimensional analysis in different fields (namely ecology) but similarly, the concept of a morphospace could easily be exported to other fields.
-For example, the multidimensional analysis in Diaz [CITE] looking at the patterns of form and functions in plants can be thought as a morphospace or a life-history-space (an ecospace?);
-isotopic analysis in [CITE Andrew's stuff] can be represented as a isotope-space, etc.
-And these generalisation could of course also be exported for any set of traits (e.g.
-acoustospaces for accoustic traits) and that even beyond evolutionary biology (e.g.
+For example, the multidimensional analysis in Diaz [CITE] looking at the patterns of form and function in plants can be thought as a morphospace or a life-history-space (an ecospace?); isotopic analyses in [CITE Andrew's stuff] can be represented as a isotope-space, etc.
+These generalisations could also be exported for any set of traits (e.g.
+acoustospaces for acoustic traits) and even beyond evolutionary biology (e.g.
 glottospaces for linguistic traits) [CITE Katie's review].
@@
 %Concluding words:
-Although disparity analysis have been rendered easy to perform from the user end (CITE packages tools), it is crucial to remember that they are first and foremost multidimensional analysis; and multidimensional analysis are complex.
+Although disparity analyses are now simple to implement (CITE packages tools), it is crucial to remember that they are multidimensional analyses; and multidimensional analyses are complex.
 In Jurassic Park, Dr Ian Malcolm summarises this problem in an elegant way: ``[We] scientists were so preoccupied with whether or not [we] could that [we] didn't stop to think if [we] should.''
-We believe that morphological analysis in general will greatly benefit in emphasising the methodological decisions (\textit{whether we should}) rather than simply using disparity analysis because \textit{we can}.
+We believe that morphological analysis in general will greatly benefit from emphasising the methodological decisions made (\textit{whether we should}), rather than simply using disparity analysis because \textit{we can}.
 

### 2018-08-02 Writing/best_practices_version.tex
Backup best practices version

--- 2018-05-18 Writing/disparity_review.tex
@@
 \bigskip
-\noindent RH: pest practices for disparity analysis.
+\noindent RH: best practices for disparity analysis.
 
@@
 
-\noindent (Keywords: disparity)\\
+\noindent (Keywords: disparity, multidimensionality, palaeobiology, ecology)\\
 
@@
 %
+%       NEW PLAN
+%
+%---------------------------------------------
+
+% Intro
+% IN -> history
+%     -Disparity or diversity?
+%     Is disparity just diversity? No.
+% OU -> Disparity can be many things
+
+% -Diversity of disparities
+% IN -> There are many different disparity
+%     -Ecological
+%     -Functional
+%     -Phylogenetic
+
+%     However, do these disparities represent anything more than the disparity of those traits rather than the disparity of the thing?
+%     Maybe.
+Disparity traits need to be carefully selected in order to properly reflect the thing.
+Disparity is only a proxy to describe the pattern and does not necesserly make conclusions on the process (but see disparity pattern vs disparity process)
+
+% OU -> But in general, disparity is a disparity of traits (Best practice: by the way, it's good to have a clear selection of the traits)
+
+% - Trait disparity
+% IN -> If it's just traits, can all of the traits be used?
+%     Which traits should be collected?
+%     Do discrete and continuous trait reflect the same thing? If yes, what does that imply?
+%     Can disparity analysis work with any kind of trait?
+%     Does the part represent the whole?
+% OU -> And how do these traits need to be analysed?
+
+% - Disparity is multidimensional
+% IN -> Disparity is multidimensional by essence
+%     Do we need trait reduction (ordination) or can we do without?
+%     What do we measure from these multidimensional spaces?
+% OU -> And how do we test these things?
+
+% - Disparity hypothesis
+% IN -> So how do we go for testing hypotheses in disparity?
+%     What are our null expectations?
+%     How can we analyse it through time?
+
+% - A disparity future
+% IN -> The definition of disparity and its uses have broadened since the early disparity start.
+%     What can we get from other fields to make disparity better
+%     But also what can we bring to other fields (discussions with Abigail)
+
+%---------------------------------------------
+%
 %       INTRODUCTION
@@
 %---------------------------------------------
+
+\section{Introduction}
+
+%Historical blurb
+Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
+These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
+One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
+
+In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF)[Runnegar 1987, Gould, Foote, Webb, etc.].
+Early disparity analyses focused on broad questions such as: How does diversity evolve?
+Are some body types more common than others?
+Can shape evolve in all ``directions'' (of the multidimensional space)?
+Since the 1980s, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity''.
+Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e.
+developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e.
+intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e.
+responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).
+
+The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
+Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation (REF Katie et al's paper somewhere here).
+Here we use a broader definition and suggest that disparity is a pattern of variation.
+Throughout this review, we aim to highlight the diversity of ``disparity'' analyses and how they can aid our understanding of the structure of biodiversity.
+
+\section{Introduction}
+% IN -> history
+\subsection{Is disparity equivalent to diversity?}
+% OU -> Disparity can be many things
+
+\section{A diversity of disparities}
+% IN -> There are many different disparity
+
+\subsubsection{Morphological disparity}
+\subsubsection{Functional disparity}
+\subsubsection{Ecological disparity}
+\subsubsection{Phylogenetic disparity}
+
+%However, do these disparities represent anything more than the disparity of those traits rather than the disparity of the thing?
+
+%Maybe.
+Disparity traits need to be carefully selected in order to properly reflect the thing.
+Disparity is only a proxy to describe the pattern and does not necesserly make conclusions on the process (but see disparity pattern vs disparity process)
+
+% OU -> But in general, disparity is a disparity of traits (Best practice: by the way, it's good to have a clear selection of the traits)
+
+\section{Trait disparity}
+% IN -> If it's just traits, can all of the traits be used?
+%     Which traits should be collected?
+
+\subsection{Are different traits equivalent?}
+%     Can disparity analysis work with any kind of trait?
+
+\subsection{Are different dataset equivalent?}
+%     Do discrete and continuous trait reflect the same thing? If yes, what does that imply?
+
+\subsection{Does the part represent the whole?}
+%     Does the part represent the whole?
+
+% OU -> And how do these traits need to be analysed?
+
+\section{Disparity is multidimensional}
+% IN -> Disparity is multidimensional by essence
+%     Do we need trait reduction (ordination) or can we do without?
+%     What do we measure from these multidimensional spaces?
+% OU -> And how do we test these things?
+
+\section{Disparity hypotheses}
+% IN -> So how do we go for testing hypotheses in disparity?
+%     What are our null expectations?
+%     How can we analyse it through time?
+
+\section{A disparate future?}
+% IN -> The definition of disparity and its uses have broadened since the early disparity start.
+%     What can we get from other fields to make disparity better
+%     But also what can we bring to other fields (discussions with Abigail)
+
+%i) the relationship between diversity and disparity
+1 - What is the relation between disparity and diversity? They are decoupled
+Disparity can be used as a parallel metric for biodiversity: species diversity (or alpha or beta diversity) aim to describe the number of population as reproductive isolates.
+On the other hand, disparity aims to describe the number of traits combination (e.g. shapes, ecologies, etc.).
+They can sometimes approximate the same thing (i.e. really low diversity can sometimes mean really low disparity) but have often been showed as decouple [CITES].
+A classic example is among mammals, really specious groups like Rodents display low morphological disparity (at least at the superficial scale) but high diversity.
+Conversely, carnivorans display a relatively lower diversity but a much higher morphological disparity.
+
+%ii) the ensuing ideas on evolvability
+2 - Disparity can be linked to evolvability?
+A morphological disparate group is likely to occupy more niches (but see the debate for the link between traits and ecological niches [CITES]) and more likely to be the result from character displacement and/or competition.
+Also, more disparate groups are more likely to lead to disparate radiations as different ecological strategies get explored.
+
+%iii) parallels with the evolution of gene regulatory evolution, and the extension of this approach to studying the evolution of ecological and functional disparity
+3.1- Ecological disparity: promising tool at a macro-evolutionary scales (but maybe less at micro-evolutionary scales).
+E.g.
+the concept of ecological guilds can be distinguished in the fossil record (carnivoran guild etc.) but it might be more difficult to grasp at a micro-scale level (CITE plant trait/niche papers).
+3.2- Functional disparity: seems to work at any scale and can be used to test convergence (species with similar trait combination can have similar functions) or the link between form and function (Diaz paper).
+However, the link between trait and function can sometimes be hard to grasp since traits can be linked to multiple functions and multiple functions can be linked to a single trait.
+
+%do continuous and categorical characters capture the same disparity?
+%Does a part represent the whole?
+%Do paleontological/ ecological/ functional disparities represent anything more than a disparity of those characters, rather than disparity of morphology/ ecology/ function?
+%Time slicing?
+%Character equivalence?
+%Is ordination meaningful?
+
+% It is inevitable that this manuscript is going to be a shit sandwich but I think it is currently an open sandwich and needs an introduction that reviews not only what disparity is, from whence it came and whither it goes, we also need to review what we have learned so far: (, (, ( of (iv) ecological and (v) functional disparity.
+
+% I would follow this with a discussion of the problems that have emerged in study: do continuous and categorical characters capture the same disparity?
+Does a part represent the whole?
+Do paleontological/ ecological/ functional disparities represent anything more than a disparity of those characters, rather than disparity of morphology/ ecology/ function?
+Time slicing?
+Character equivalence?
+Is ordination meaningful?
+
+% I think that I would make this discussion less abstract and use examples.
+We should also write this as though the prospective articles do not exist – because they don’t.
+What are the solutions?
+I think they usually involve forethought in place of playing with data and hoping some interesting but untestable phenomenon falls out.
+And, of course, follow with a prospect on what could be done in future….
+
+% Most of it is there – it just needs jiggling about and editing a bit – as well as adding to.
+
+% We were asked originally to write a review on this topic and so we’re pushing the envelope in writing about best practices.
+I’m comfortable with that, but we still need to achieve the basic expectation of a review – though I always prefer an insightful (more than the sum of its parts) review, rather than a piece of journalism.
+
+% So in a round about way, yes, I think we do need to have a ‘what we have learned’ aspect to the manuscript.
+
 \section{Introduction}
@@
     \item{Which data should we use in disparity analyses?}
-    %[One line abstract of Melanie's paper on inapplicable data].
+    [One line abstract of Melanie's paper on inapplicable data - Needs to be developed].
     Furthermore, some simple methods already exist to account for caveats in data. 

### 2018-08-02 Writing/disparity_review.tex
added placeholder for Natalie's comment
For exporting the manuscript to google docs
Typo in title
Restructured paper INIT

@@
 %---------------------------------------------
+
+\section{Introduction}
+
+%Historical blurb
+Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
+These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
+One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
+
+In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF)[Runnegar 1987, Gould, Foote, Webb, etc.].
+Early disparity analyses focused on broad questions such as: How does diversity evolve?
+Are some body types more common than others?
+Can shape evolve in all ``directions'' (of the multidimensional space)?
+Since the 1980s, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity''.
+Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e.
+developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e.
+intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e.
+responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).
+
+The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
+Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
+Here we use a broader definition and suggest that disparity is a pattern of variation.
+Throughout this review, we aim to highlight the diversity of ``disparity'' analyses and how they can aid our understanding of the structure of biodiversity.
+
+\subsection{Is disparity equivalent to diversity?}
+One obvious aspect when measuring biodiversity is the measurement of raw species diversity (at any scale; $\alpha$, $\beta$, $\gamma$).
+Such biodiversity metric intuitively describes the amount of reproductive isolated populations in a system.
+On the other hand, disparity aims at describing the amount of measured traits in a system (see below), in such a way, a high morphological disparity value is a measurement of a high diversity of morphologies.
+The link between both disparity and diversity has been scrutinised under different scenarios with most conclusions supporting an effective decoupling between both [CITE the disparity/diversity papers].
+This is intuitively understandable when comparing the both metrics for two contrasted groups of organisms, for example, rodents would display a high diversity (@@@ sp) with a relatively low disparity (a majority of rodents have the typical rat shape) whereas carnivores would display a relatively lower diversity (@@@ sp) with a higher disparity (with body shapes ranging from a weasel to an elephant seal).
+However, it is important to note that in more nuanced groups, diversity and disparity can be link where a really low diversity would tend to yield a low disparity and vice versa.
+
+%ii) the ensuing ideas on evolvability
+2 - Disparity can be linked to evolvability?
+A morphological disparate group is likely to occupy more niches (but see the debate for the link between traits and ecological niches [CITES]) and more likely to be the result from character displacement and/or competition.
+Also, more disparate groups are more likely to lead to disparate radiations as different ecological strategies get explored.
+
+% OU -> Disparity can be many things
+
+\section{A diversity of disparities}
+% IN -> There are many different disparity
+This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
+However, as with any great power, it comes with great responsibility. 
+With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
+First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
+Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
+Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
+Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.
+
+\subsubsection{Morphological disparity}
+
+\subsubsection{Functional disparity}
+3.2- Functional disparity: seems to work at any scale and can be used to test convergence (species with similar trait combination can have similar functions) or the link between form and function (Diaz paper).
+However, the link between trait and function can sometimes be hard to grasp since traits can be linked to multiple functions and multiple functions can be linked to a single trait.
+\subsubsection{Ecological disparity}
+3.1- Ecological disparity: promising tool at a macro-evolutionary scales (but maybe less at micro-evolutionary scales).
+E.g.
+the concept of ecological guilds can be distinguished in the fossil record (carnivoran guild etc.) but it might be more difficult to grasp at a micro-scale level (CITE plant trait/niche papers).
+\subsubsection{Phylogenetic disparity}
+
+%However, do these disparities represent anything more than the disparity of those traits rather than the disparity of the thing?
+
+%Maybe.
+Disparity traits need to be carefully selected in order to properly reflect the thing.
+Disparity is only a proxy to describe the pattern and does not necesserly make conclusions on the process (but see disparity pattern vs disparity process)
+
+% OU -> But in general, disparity is a disparity of traits (Best practice: by the way, it's good to have a clear selection of the traits)
+
+\section{Trait disparity}
+% IN -> If it's just traits, can all of the traits be used?
+Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
+For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for spec).ies delimitation.
+Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below
+
+\subsection{Are different traits equivalent?}
+%     Can disparity analysis work with any kind of trait?
+A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
+One classic case is using ``cladistic'' data in palaeobiology.
+These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
+It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
+A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all of the data.
+For example, in a study of the disparity of dietary niches, removing non-dietary characters is a good approach.
+However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g.
+limb traits can be used for both locomotion and feeding).
+
+\subsection{Are different dataset equivalent?}
+%     Do discrete and continuous trait reflect the same thing? If yes, what does that imply?
+Morphological disparity analysis can be based on three main types of data:
+\begin{enumerate}
+    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g.
+the presence or absence of legs, the colour of a tail, etc.
+[CITE]) or count the number of features (e.g.
+number of digits, etc.
+[CITE]);
+    \item continuous data that can be either simple measurements (e.g.
+body length, etc.
+[CITE]) or more abstracted measures (e.g.
+geometric morphometrics, mathematical contour description, etc.
+[CITE]);
+    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
+\end{enumerate}
+None of these types of data are fundamentally better but each will be more appropriate for specific questions.
+For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
+On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
+This again illustrates the importance of framing the question.
+In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
+wingspan, aspect ratio, etc.).
+Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g.
+digit length, integumentary system, etc.).
+
+\subsection{Does the part represent the whole?}
+%     Does the part represent the whole?
+Melanie's work.
+
+%Bonus
+There are many other considerations when choosing the appropriate data to use. 
+It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
+For example, are the observed differences between two groups related to factors unrelated to group designation such as sexual dimorphism?
+Are the taxa/groups being compared fairly consistent and truly comparable?
+In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated [String of morpho data problems cite].
+It is important to ensure that any patterns detected are not merely reflections of these features of the data.
+
+Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
+% OU -> And how do these traits need to be analysed?
+
+\section{Disparity is multidimensional}
+% IN -> Disparity is multidimensional by essence
+% Visualisation problem
+\subsection{Visualising the multidimensional problem}
+\label{visualisation}
+It is important to remember that visual interpretations of multidimensional data can easily be misleading.
+For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
+Conversely, when two groups do not overlap on a two or three dimensional plot, this does not mean they never overlap in the multidimensional space.
+Furthermore, multidimensional spaces are not necessary Euclidean.
+Such non-Euclidean spaces often have non-intuitive properties, for example straight lines in some dimensions are not straight in the space and, even less intuitively, the distance between points A and B might not be equal to the distance between points B and A [CITE Sylvain].
+
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
+\caption{\small{Two groups that overlap in some dimensions do not necessary overlap in all dimensions. 
+Conversely, two groups that do not overlap in some dimensions can overlap in other dimensions.
+In this example for the groups A and B to truly overlap, they must overlap in all three dimensions displayed here.
+Although these groups overlap in the two first dimensions (panel 1), they do not overlap in dimensions 1 and 3 thus do not overlap in the multidimensional space at all.
+The inverse is of course true: if two groups don't overlap in two dimensions (D and C in panel 1), they do not overlap in the multidimensional space, even if they overlap in any other combination of dimensions (dimensions 1 and 3 in panel 2).}}
+\label{Fig:RF_results_best}
+\end{figure}
+
+%     Do we need trait reduction (ordination) or can we do without?
+
+\subsection{[Box]To ordinate or not to ordinate; that is the (multidimensional) question?}
+\label{box_ordination}
+Reducing the dimensionality of data through ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
+Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g.
+selecting only the axes that describe up to 95\% of the variation).
+In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions.
+This has clear advantages for interpreting the results.
+For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).
+
+However, some caveats can arise from such transformations of the data.
+For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g. only the first few) can lead to misinterpretation of the results.
+Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and might have little relation to the disparity question (e.g.
+if the first few axes represent the elongation of the beak, but the question is about wing disparity).
+
+In many cases, ordination might not be necessary.
+For example, if the disparity metric being used relies on all the data (see Metrics section \ref{metrics}) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
+Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
+For example, when the analysed data is non-Euclidean (e.g.
+discrete morphological characters), the ordination can be problematic ([cite Melanie's future paper here]), and the resulting non-Euclidean ordinated space can be difficult to interpret ([probably cite Sylvain's here]).
+This can be problematic when comparing the position of groups in the multidimensional space, as the measured distances might not be comparable.
+Finally, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axes can be impossible or biologically meaningless.
+Although some gradients might be easy to detect or interpret (e.g.
+the elongation of the beak along the first axis), they are often hard to interpret in terms of morphological gradients or simply non-interpretable (e.g.
+with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0 has no biological meaning).
+
+Because of these points, we strongly recommend that data for multidimensional analyses is not automatically ordinated, and that careful consideration is given to whether the question can be answered without ordination.
+\subsection{[Box] end}
+
+%     What do we measure from these multidimensional spaces?
+% OU -> And how do we test these things?
+
+\section{Disparity hypotheses}
+% IN -> So how do we go for testing hypotheses in disparity?
+%     What are our null expectations?
+The final stage of most disparity analyses is statistical testing of the disparity metrics, to determine whether changes thought time or between groups are statistically significant. 
+Similar to the points above, the statistical test to apply depends on which question is being tackled with disparity analysis, i.e. which hypothesis is being tested.
+It also requires consideration of the data, the ordination method used (or the absence thereof) and the metric used.
+
+For example, one commonly used test in morphological analysis, is the non parametric permutation analysis of variance [NPANOVA or PERMANOVA; CITE Anderson + examples].
+This test and its implementation is perfectly valid to test whether two groups have a different variance/covariance, but is based on the pairwise distances of the group elements.
+This test will ignore any ordination (if the data was ordinated) and is not based on a unidimensional disparity metric.
+In other words, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
+Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the question.
+
+Another caveat for applying statistical tests on disparity data is that it is important to consider which data to apply the test on.
+For example, in morphological disparity analysis (especially for palaeobiological questions), data is often bootstrapped.
+This has two useful advantages: first, when the disparity metric is unidimensional (e.g.
+the sum of variances or the average squared pairwise distance), bootstrapping the data generates a distribution of this unidimensional metric that can then be analysed using the vast statistical toolkit available for comparing distributions; second, when the data is scarce (e.g.
+for palaeontological data), bootstrapping the data allows users to introduce statistical variance, rendering the test less sensitive to outliers.
+However, bootstrapping the data also comes with an important caveat: the data analysed are pseudo-replicated and thus non-independent.
+This can violate the assumptions of parametric statistical tests.
+Furthermore, it is important to remember that the number of bootstrap pseudo-replicates will inevitably increase the Type I error.
+
+%     How do we take phylogeny into account?
+
+\subsection{Phylogeny}
+For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
+For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
+and therefore will not covered further in this review.
+
+\subsubsection{Ancestral state estimation}
+When looking at disparity in the past (e.g. disparity-through-time) ancestral state estimations are a common approach.
+Ancestral state estimation can be performed in two ways:
+\begin{enumerate}
+\item pre-ordination: the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original dataset[CITE].
+\item post-ordination: the estimation is done after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
+\end{enumerate}
+Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e.
+the relationship between the points not estimated) and implies longer computational times but, once the morphospace is defined, it will remain constant (i.e.
+its properties will not change).
+Post ordination ancestral state estimation will not change the ordinated space's geometry and will be faster to compute, however, it will change the morphospace's properties such as the number of parameters (i.e.
+elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper to be?].
+
+More importantly, however, one needs to remember that using such ancestral state estimation methods comes with several caveats.
+First and foremost, although ancestral state estimations are valid methods in macroevolution, their results are highly dependent on the data and chosen methods.
+Secondly, more specific to disparity analysis, when using post-transformation ancestral state estimates, the morphospace can be saturated leading pairwise distance metrics to tend to zero.
+In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points for statistical reasons.
+In fact, these extra points are not independent and can also generate side effects that can be problematic, especially when testing the effect of extinction events on disparity.
+
+Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
+In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes.
+Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution.
+%     How can we analyse it through time?
+
+\section{A disparate future?}
+% IN -> The definition of disparity and its uses have broadened since the early disparity start.
+
+As mentioned above, there is no ``one-size-fits-all'' morphological disparity analysis pipeline.
+As with any multidimensional analysis, there are many nuances in deciding which data to use and how to analyse it, stemming from the explicit formulation of the disparity hypothesis.
+To understand these nuances, it is important to remember that: (1) morphological disparity analyses are not new and much effort has been made in developing methods; and (2) multidimensional analyses are not restricted to morphological data analysis and solutions to specific problems can be found in literature that is not necessary focused on morphology (especially ecology and mathematics).
+
+We believe that many of the caveats in morphological disparity analysis can arise when ``blindly'' applying methodological pipelines at the expense of losing sight of the  biological question being tested.
+We therefore advise keeping the methodology simple and tractable, at least in the preliminary analysis.
+Also, many of the caveats and technical problems in disparity analyses have been tackled in other fields, mainly ecology.
+This is perhaps unsurprisingly as ecological questions are often closely related to evolutionary questions, and ecological data is also often multidimensional [CITE DONOHUE's papers].
+
+%     What can we get from other fields to make disparity better
+%     But also what can we bring to other fields (discussions with Abigail)
+
+\subsection{How can we adapt our methods?}
+Each of the points outlined above may drive new methodological questions (and often already are).
+Although it is impossible to cover the state-of-the-art exhaustively for each of these points here are several solutions and/or directions for adapting existing methods:
+
+\begin{itemize}
+    \item{Which data should we use in disparity analyses?}
+    [One line abstract of Melanie's paper on inapplicable data - Needs to be developed].
+    Furthermore, some simple methods already exist to account for caveats in data. 
+    For example, to correct for the paucity of the fossil record, one can use methods like rarefaction analysis to take sample size into account [CITE].
+    % NC: This item currently doesn't seem to actually give advice about which data to use...
+    \item{How should we visualise disparity?} % NC: Tried to make the bullets a bit clearer, might need to edit for correctness and consistency with the rest of the document.
+    [One line abstract of Sylvain's paper on morphospaces plottings].
+    One major problem of multidimensional analyses is that highly dimensional spaces are hard (or even impossible) to conceptualise.
+    One solution to this is to use more simple 2D/3D ordinations.
+    For example, using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation [e.g Raup-spaces CITE].
+    \item{Which disparity metrics should we use?}
+    This can be tackled by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combination of properties of interest in the morphospace.
+    For example, when reducing a multidimensional space to one metric, it would be judicious to indicate what this metric represents in the results (e.g.
+in a plot or a table, if the metric used was the sum of variances, labelling it as ``size of the group'' rather than ``disparity'').
+    This is especially advantageous when using multiple disparity metrics (e.g. ``size of the group'' and ``position of the group'').
+    \item{How should we test disparity hypotheses?}
+    Finally, it is possible to use non-parametric methods to test specific disparity questions.
+    For example, using available software and computer power, it is possible to propose specific null models for the expectation of what disparity would be expected to look like under various null models.
+    This has been already successfully used in ecology [CITE Diaz paper] and is currently being implemented in palaeobiology (Mark stuff).
+\end{itemize}
+
+\subsection{Where do we go from here?}
+Morphological disparity analyses are a major tool in evolutionary biology, posing a more versatile alternative to studying species richness, especially in macroevolutionary studies.
+We believe that morphological disparity analyses, as the analysis of patterns of morphological differences, can be of great use for linking evolutionary and ecological processes (e.g.
+when using morphological disparity as a proxy for ecological niches to test competition-through-time hypotheses).
+
+However, as demonstrated above, such analyses would greatly benefit from more development, and  more care is needed in study design so that the analysis actually reflects the biological question at hand.
+This can not only be done through the points highlighted above but also through linking the observed patterns to an explicit process (for example, the link between an organism's form and function) or by developing null models to potentially explain the observed patterns.
+
+Finally, we encourage to a more cross-disciplinary approach to morphological disparity analyses.
+In fact, as mentioned above, morphological disparity could greatly benefit from advances in multidimensional analysis in different fields (namely ecology) but similarly, the concept of a morphospace could easily be exported to other fields.
+For example, the multidimensional analysis in Diaz [CITE] looking at the patterns of form and function in plants can be thought as a morphospace or a life-history-space (an ecospace?); isotopic analyses in [CITE Andrew's stuff] can be represented as a isotope-space, etc.
+These generalisations could also be exported for any set of traits (e.g.
+acoustospaces for acoustic traits) and even beyond evolutionary biology (e.g.
+glottospaces for linguistic traits) [CITE Katie's review].
+
+%Concluding words:
+Although disparity analyses are now simple to implement (CITE packages tools), it is crucial to remember that they are multidimensional analyses; and multidimensional analyses are complex.
+In Jurassic Park, Dr Ian Malcolm summarises this problem in an elegant way: ``[We] scientists were so preoccupied with whether or not [we] could that [we] didn't stop to think if [we] should.''
+We believe that morphological analysis in general will greatly benefit from emphasising the methodological decisions made (\textit{whether we should}), rather than simply using disparity analysis because \textit{we can}.
+
 \section{Introduction}

### 2018-09-11 Writing/disparity_review.tex
Updated shit sandwich

@@
 \bigskip
-\noindent RH: best practices for disparity analysis.
+\noindent RH: disparate disparity analysis.
 
@@
 
-\noindent{\Large \bf Analysing disparity in morphology data: best practice and future directions}
+\noindent{\Large \bf The disparite ways to analysis morphological disparity} %TG: oooooh the ugly title!
 
@@
 Such biodiversity metric intuitively describes the amount of reproductive isolated populations in a system.
-On the other hand, disparity aims at describing the amount of measured traits in a system (see below), in such a way, a high morphological disparity value is a measurement of a high diversity of morphologies.
+On the other hand, disparity aims at describing the amount of measured traits in a system (see below), in such a way, a high morphological disparity value is a measurement of a high diversity of morphologies (i.e.
+shapes or body plans).
 The link between both disparity and diversity has been scrutinised under different scenarios with most conclusions supporting an effective decoupling between both [CITE the disparity/diversity papers].
-This is intuitively understandable when comparing the both metrics for two contrasted groups of organisms, for example, rodents would display a high diversity (@@@ sp) with a relatively low disparity (a majority of rodents have the typical rat shape) whereas carnivores would display a relatively lower diversity (@@@ sp) with a higher disparity (with body shapes ranging from a weasel to an elephant seal).
+This is intuitively understandable when comparing both metrics for two contrasted groups of organisms, for example, living rodents would display a high diversity ($>$2200 sp; CITE Wilson\&Reeder) with a relatively low disparity (a majority of rodents have the typical rat shape) whereas carnivores would display a relatively lower diversity ($>$280 sp; CITE Wilson\&Reeder) with a higher disparity (with body shapes ranging from a weasel to an elephant seal).
 However, it is important to note that in more nuanced groups, diversity and disparity can be link where a really low diversity would tend to yield a low disparity and vice versa.
 
-%ii) the ensuing ideas on evolvability
-2 - Disparity can be linked to evolvability?
-A morphological disparate group is likely to occupy more niches (but see the debate for the link between traits and ecological niches [CITES]) and more likely to be the result from character displacement and/or competition.
-Also, more disparate groups are more likely to lead to disparate radiations as different ecological strategies get explored.
-
-% OU -> Disparity can be many things
+This decoupling has led researchers to work on the relation between disparity and evolvability: a more disparate group (population, group, etc...) will have more genetic diversity to adapt to biotic and abiotic changes [CITES].
+This led to the use of disparity as a tool for investigating morphological modularity [CITE].
+However, this is based on the assumption that there is a link between diversity in traits (disparity) and subsequent diversity in genes and/or in ecological niches ([Trait plant paper CITE] see below for more details).
+Ultimately, both links between disparity and diversity and disparity and evolvability are dependent on the multiple definitions of disparity and the multiple datasets used.
 
@@
 This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
-However, as with any great power, it comes with great responsibility. 
-With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
-First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
-Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
-Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
-Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.
+From a same analysis comparing morphological traits among different organisms, one can study four main types of disparity:
 
-\subsubsection{Morphological disparity}
+\begin{enumerate}
 
-\subsubsection{Functional disparity}
-3.2- Functional disparity: seems to work at any scale and can be used to test convergence (species with similar trait combination can have similar functions) or the link between form and function (Diaz paper).
+\item{Morphological disparity.}
+The disparity of morphological trait configurations has been at the genesis of disparity analysis either through Foote's pioneering work [Foote Crinoids] or Raup's morphospaces [Raup].
+This type of disparity analysis often consists in a more descriptive multidimensionality analysis of the patterns in the diversity of morphological traits and has been successfully used to answer questions such as why are some morphological traits combinations more common than other [Raup] and what are the properties of the resulting morphospace [Gerber].
+It is important to highlight that the morphological data used in this context was collected specifically for these analyses and not ``recycled'' from phylogenetic analysis.
+
+\item{Phylogenetic disparity.}
+Another disparity approach is to use the same morphological disparity approach as mentioned above but using discrete morphological data commonly used for phylogenetic analysis.
+This approach is widely used in palaeobiology to answer a range of macroevolutionary questions [CITEs string].
+Although very useful for describing macroevolutionary patterns, this approach can tend to artificially increase disparity between phylogenetically distinct groups due to the nature of so phylogenetic characters use to discriminate groups [CITE Foote].
+
+\item{Functional disparity.}
+Functional disparity has since then been assumed to work at a large scale and can be used to study convergence (species with similar trait combination can assume similar functions).
+Furthermore there as been work on linking form and function (Diaz paper).
 However, the link between trait and function can sometimes be hard to grasp since traits can be linked to multiple functions and multiple functions can be linked to a single trait.
-\subsubsection{Ecological disparity}
-3.1- Ecological disparity: promising tool at a macro-evolutionary scales (but maybe less at micro-evolutionary scales).
-E.g.
-the concept of ecological guilds can be distinguished in the fossil record (carnivoran guild etc.) but it might be more difficult to grasp at a micro-scale level (CITE plant trait/niche papers).
-\subsubsection{Phylogenetic disparity}
 
-%However, do these disparities represent anything more than the disparity of those traits rather than the disparity of the thing?
+\item{Ecological disparity.}
+Linking to functional disparity, people have proposed to use morphological trait disparity as a proxy for ecological niches [CITE].
+This seems to work at a macro-evolutionary scale but is still debated at a micro-evolutionary scale [CITE plant trait paper].
+For example, the concept of different ecological guilds can easily be distinguished through the fossil record (apex predator have share morphological characteristics through time; e.g.
+large body mass, pointy teeth, etc.) but can be more difficult at a finer scales and when defining the actual ecological niche (i.e.
+does a T-Rex occupy the same top predator niche as an Orca?).
 
-%Maybe.
-Disparity traits need to be carefully selected in order to properly reflect the thing.
-Disparity is only a proxy to describe the pattern and does not necesserly make conclusions on the process (but see disparity pattern vs disparity process)
+\end{enumerate}
 
-% OU -> But in general, disparity is a disparity of traits (Best practice: by the way, it's good to have a clear selection of the traits)
+This diversity of patterns that can be studied using disparity analyses links to the ambiguous distinction between disparity as a pattern or a process.
+In other words, do these disparities represent anything more than the variance of those traits or is it link to specific biological properties of these traits?
+For example, does the distribution of traits in apex predators allow them to be distinguished in disparity analysis (i.e.
+disparity is the pattern) or is there something about being an apex predator that results in this combination of traits (i.e.
+disparity is the process).
+As highlighted above, for the rest of this review, we will essentially use disparity as a term for describing morphological traits patterns but we acknowledge that disparity can sometimes also be the process leading to the observed patterns (e.g.
+in the case of trait convergence).
+Furthermore, as a best practice, we do advice researchers to select the traits that will be fit there question \textit{a priori} and not measure all the traits available.
+For example, when studying disparity in locomotion in a group of animals, postcranial characters might more readily answer the question than cranial ones.
 
@@
 % IN -> If it's just traits, can all of the traits be used?
-Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
-For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for spec).ies delimitation.
-Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below
+In essence, disparity studies are multidimensional analyses of morphological traits.
+Such traits can be collected using three main traits: 1) discrete ``cladistic'' characters (e.g.
+coded as the absence or presence of features [CITE any paleo disp paper]); 2) continuous measurements (e.g.
+length of features [CITE Anderson fish paper]) or 3) more sophisticated mathematical measurements from in geometric morphometric data (e.g.
+Pcrocrustes superimpositions of entire features [CITE a geomorph paper], Fourier contour analyses [CITE], etc.).
+
+None of these types of data are fundamentally better and each have a range of materialistic pros and cons (e.g.
+cost, time, range, etc.), however, some specific questions or methodological pipelines can be more appropriated to some types of data.
+For example, when investigating variation within bat wing shapes, both homologous landmarks (e.g.
+used in geometric morphometric analysis) or continuous measurements of bones may be appropriate to capture patterns of variations.
+Conversely, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question from a mechanical perspective.
+In the above example, if the question is about functional disparity, say, whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
+wingspan, aspect ratio, etc.).
+On the opposite, if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g.
+digit length, integumentary system, etc.).
+It is therefore important to consider disparity patterns as traits disparity patterns approximating biologically complex patterns.
 
 \subsection{Are different traits equivalent?}
-%     Can disparity analysis work with any kind of trait?
-A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
-One classic case is using ``cladistic'' data in palaeobiology.
-These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
-It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
-A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all of the data.
-For example, in a study of the disparity of dietary niches, removing non-dietary characters is a good approach.
-However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g.
-limb traits can be used for both locomotion and feeding).
 
-\subsection{Are different dataset equivalent?}
-%     Do discrete and continuous trait reflect the same thing? If yes, what does that imply?
-Morphological disparity analysis can be based on three main types of data:
-\begin{enumerate}
-    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g.
-the presence or absence of legs, the colour of a tail, etc.
-[CITE]) or count the number of features (e.g.
-number of digits, etc.
-[CITE]);
-    \item continuous data that can be either simple measurements (e.g.
-body length, etc.
-[CITE]) or more abstracted measures (e.g.
-geometric morphometrics, mathematical contour description, etc.
-[CITE]);
-    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
-\end{enumerate}
-None of these types of data are fundamentally better but each will be more appropriate for specific questions.
-For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
-On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
-This again illustrates the importance of framing the question.
-In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
-wingspan, aspect ratio, etc.).
-Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g.
-digit length, integumentary system, etc.).
+From the example above, the question can be asked whether using different methods to approximate the same traits on the same features would lead to similar results.
+In other words, does discrete, continuous or geometric morphometric data collected from a same feature capture the same disparity signal?
+This question has been assessed several times by comparing the ``disparity signal'' between cladistic and geometric morphometric data in vertebrates leading to the conclusions that data originating from the same source leads to the same results [CITE].
+Although the advantages and caveats of each type of traits are well known, there is no clear boundaries to which extent the equivalence between dataset holds.
+These equivalence might be more primordial at the trait level (i.e. which traits do we measure) rather than at the measurement method level (i.e. how do we measure them) [Foote Crinoids].
 
 \subsection{Does the part represent the whole?}
-%     Does the part represent the whole?
-Melanie's work.
 
-%Bonus
-There are many other considerations when choosing the appropriate data to use. 
-It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
-For example, are the observed differences between two groups related to factors unrelated to group designation such as sexual dimorphism?
-Are the taxa/groups being compared fairly consistent and truly comparable?
-In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated [String of morpho data problems cite].
-It is important to ensure that any patterns detected are not merely reflections of these features of the data.
+In this context, it can be important to understand the relations between the analysis of one specific set of traits and the macroevolutionary questions asked at a organismal level.
+For example, when studying mammalian morphological evolution through time based on cladistic characters, do the disparity patterns reflect the changes through time of mammalian morphological niches or do they merely approximate variation in one specific set of traits?
+In mammalian discrete datasets, many characters are dental and may not actually represent one organsims' niche (i.e.
+if the dataset is missing other crucial characters like locomotory ones).
+%TG: maybe add an example with stuff with the same dentition but living in a different environement, e.g.
+trees/ground).
+However, studies have shown that
+-[mammalian teeth are not that bad - Zou and Zhang] 
+-[the head can be enough to represent disparity - Melanie's work]
 
+Furthermore, trait data suffers from the same short comings as most data in biology (i.e.
+allometry, homology, sexual dimorphism, etc...) or to short comings from data in general (missing, non-overlapping, hierarchical - e.g.
+inapplicable -, ambiguous, polymorphic or correlated, sample size [String of morpho data problems cite]).
 Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
@@
 \section{Disparity is multidimensional}
+
+Additionally to pitfalls and caveats stemming from the biological questions and the physical datasets analysed through the lens of morphological disparity, the multidimensional aspect of these analysis can also add a layer of technical complexity.
+In fact, disparity analysis are inherently multidimensional and our human brains and the physical word is bound to three spatial dimensions.
+This problem is sometime apparent in over-interpretation of disparity analysis results.
+For example, in a classic morphological disparity pipelined analysis using geometric morphometric data, the collected data's dimensionality is the number of collected landmarks and their 3 or 2 dimensional set of coordinates.
+However, using dimensionality reduction techniques can lead to misinterpretation of the results if not careful.
+In the case of ordination, it is not uncommon to then use only the first axis of ordination (e.g. PC1) and draw conclusions based on this axis.
+The changes in values along this dimension can display some disparity pattern or gradient at a glance but don't often reflect the actual changes of the set of all the landmarks' dimensions [CITE Vera and Thomas paper].
+Therefore, whether for visualising or analysing disparity data, it is crucial to remember it's multidimensionality aspect.
+
 % IN -> Disparity is multidimensional by essence
@@
 \label{visualisation}
-It is important to remember that visual interpretations of multidimensional data can easily be misleading.
+Visual interpretations of multidimensional data can easily be misleading.
 For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
@@
 
-\subsection{[Box]To ordinate or not to ordinate; that is the (multidimensional) question?}
-\label{box_ordination}
-Reducing the dimensionality of data through ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
+\subsection{To ordinate or not to ordinate; that is the (multidimensional) question?}
+Similarly as for visualisation, reducing the dimensionality of a disparity dataset can be tricky.
+Ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
 Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g.
@@
 Because of these points, we strongly recommend that data for multidimensional analyses is not automatically ordinated, and that careful consideration is given to whether the question can be answered without ordination.
-\subsection{[Box] end}
 
 %     What do we measure from these multidimensional spaces?
+\subsection{Measuring disparity}
+
+A second important point stemming from the multidimensional nature of disparity analyses is considering how to extract some meaningful interpretable summary information: the disparity metric (or index).
+As with any summary metric, and because of this reduction in dimensionality, disparity metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
+It is often beneficial to use more than one metric to summarise the different aspects of interest of the morphospace (e.g. its width or spread).
+
+The choice of disparity metric is thus essential to the disparity analysis and should always be driven by the question.
+When considering only one dimension, disparity metrics can be used to reflect the ``size'' of the distribution (e.g.
+the range, quantiles or variance), or reflect the most likely value (the central tendency, i.e.
+the mean, median or mode).
+Among these metrics, some will have more preferable properties than others, such as sensitivity to outliers (range, mean or mode are highly sensitive \textit{vs.} quantiles, variance or median are less sensitive) and will thus be more or less appropriate to the different questions and the properties of the distribution.
+For example, if the question refers to ``size'' of a group in the morphospace (e.g.
+``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriate (e.g.
+volume [CITE], variance [CITE], range [CITE], distance from the centroid [CITE]).
+Conversely, if the question refers to the ``position'' of a group in the morphospace (e.g.
+``does group A occupy the same space as group B?''), metrics relating to the distance between the elements within a group and a fixed point in the morphospace are most appropriate.
+Finally, one might also be interested in the density of elements in some regions of the morphospace (e.g.
+``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriate (e.g.
+nearest neighbour distance, pairwise distances, etc.).
+
+Additionally to the desired properties of what the metrics should capture, it is important to also keep in mind the mathematical properties of the metrics and their associated caveats.
+For example, measuring the full sum of the variance of each dimension of the space does not require one to add the co-variance between the axes in a ordinated space using a PCA.
+This is not true however in other mathematical spaces or (even in a PCA) when not all dimensions or elements are considered (e.g.
+when only looking at a group within the morphospace, the sum of variances is not mathematically equal to the sum of the variance of each dimension in the space).
+Furthermore, multidimensional space has some counter-intuitive properties that need to be taken into account such as the curse of dimensionality [CITE].
+For example, product-based metrics as proxies of volumes (e.g.
+product of ranges, hypervolume, hypercube, etc.) will tend towards zero fairly quickly for spaces with even a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem and Donoho 2000 Aide-Memoire].
+Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size, for example range [CITE] or convex hull based metrics [CITE].
+
 % OU -> And how do we test these things?
+The technical properties of multidimensional disparity analyses themselves allow for the vast amount of methods and metrics described above to be used in specific context.
+However, as mentioned above equally, these tools should always be used in the frame of a specific stated scientific question.
+This leads to the importance of clearly defining and rigorously testing on or more hypotheses.
 
@@
 %     What are our null expectations?
-The final stage of most disparity analyses is statistical testing of the disparity metrics, to determine whether changes thought time or between groups are statistically significant. 
-Similar to the points above, the statistical test to apply depends on which question is being tackled with disparity analysis, i.e. which hypothesis is being tested.
-It also requires consideration of the data, the ordination method used (or the absence thereof) and the metric used.
-
+The multidimensional statistical toolkit has been greatly developed in the last years [CITE Dean Adams, Simone, etc.] but oddly enough not routinely applied to disparity analyses.
+Instead, disparity hypotheses testing seemed to be confined to a small set of (well established) methods.
 For example, one commonly used test in morphological analysis, is the non parametric permutation analysis of variance [NPANOVA or PERMANOVA; CITE Anderson + examples].
-This test and its implementation is perfectly valid to test whether two groups have a different variance/covariance, but is based on the pairwise distances of the group elements.
-This test will ignore any ordination (if the data was ordinated) and is not based on a unidimensional disparity metric.
-In other words, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
+This test is an analysis of variance in the pairwise distances between different groups and is often completely independent of the type of morphospace and the disparity metric used.
+Although perfectly valid, this test can sometimes not be directly related to the hypotheses stated.
+For example, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
 Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the question.
@@
 
-%     How do we take phylogeny into account?
-
 \subsection{Phylogeny}
-For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
+More specifically within the realm of testing hypothesis, it is important to remember that when analysing biological data, the data is not independent [CITE Felsenstein or so].
+For many disparity analyses of macroevolutionary questions, phylogeny must thus be taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
 For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
@@
 Post ordination ancestral state estimation will not change the ordinated space's geometry and will be faster to compute, however, it will change the morphospace's properties such as the number of parameters (i.e.
-elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper to be?].
+elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper].
 
@@
 Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
-In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes.
-Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution.
+In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes [CITE time slice].
+Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution [CITE STD paper?].
 %     How can we analyse it through time?
@@
 
-\section{Introduction}
-
-Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
-These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
-One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
-% The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
-% Each individual carnivore species occupies a discrete region of this space related to its values for these traits.
-
-In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF)[Runnegar 1987, Gould, Foote, Webb, etc.].
-Early disparity analyses focused on broad questions such as: How does diversity evolve?
-Are some body types more common than others?
-Can shape evolve in all ``directions'' (of the multidimensional space)?
-Since the 1980s, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity''.
-Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e.
-developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e.
-intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e.
-responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).
-
-The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
-Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation (REF Katie et al's paper somewhere here).
-Here we use a broader definition and suggest that disparity is a pattern of variation.
-Throughout this review, we aim to highlight the diversity of ``disparity'' analyses and how they can aid our understanding of the structure of biodiversity.
-
-% \subsection{What \textit{is} disparity?} %TG: maybe useful for box or paraphrapsing summaries.
-% What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
-% The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
-% Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
-% There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
-% Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
-% The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).
-
-% Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
-% In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
-% Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
-
-\section{Best practices for disparity analysis}
-
-This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
+\subsection{Conclusion}
 However, as with any great power, it comes with great responsibility. 
@@
 Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.
-
-\section{What is the question?}
-Why perform a disparity analysis?
-This seems like it should be obvious, but we stress its crucial importance.
-The last decade has seen a vast array of new methods being published and implemented in freely available software, making disparity analyses easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
-However, this (consciously or not) pushes authors towards performing disparity analyses simply because of their ease of implementation, rather than stemming from a question driven approach.
-Before rushing into a disparity analysis, one should consider some basic caveats.
-For example, one major thing to consider is circularity: if the aim is to explore the variation in a dataset, disparity analysis is perfectly valid.
-However, if this exploratory analysis reveals distinct groups, it would be circular to then test for differences among those groups as if this had been the hypothesis all along [CITE stats 1.0].
-This can easily be avoided by using a subset of the data for preliminary data exploration, and then using the full dataset to test for differences among any groups that emerge.
-Clearly framing questions and their null expectations will help in designing better disparity analyses.
-
-\section{Collect the appropriate data}
-Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
-For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for species delimitation.
-Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).
-
-\subsection{Data types}
-Morphological disparity analysis can be based on three main types of data:
-\begin{enumerate}
-    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g.
-the presence or absence of legs, the colour of a tail, etc.
-[CITE]) or count the number of features (e.g.
-number of digits, etc.
-[CITE]);
-    \item continuous data that can be either simple measurements (e.g.
-body length, etc.
-[CITE]) or more abstracted measures (e.g.
-geometric morphometrics, mathematical contour description, etc.
-[CITE]);
-    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
-\end{enumerate}
-None of these types of data are fundamentally better but each will be more appropriate for specific questions.
-For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
-On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
-This again illustrates the importance of framing the question.
-In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
-wingspan, aspect ratio, etc.).
-Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g.
-digit length, integumentary system, etc.).
-
-\subsection{How much data should be collected?}
-Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of specimens and traits.
-\begin{itemize}
-    \item \textit{Taxonomic scope.}
-    Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
-    Taxonomic scope will depend on the question being investigated, and what type of variation is being analysed.
-    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation to capture the full variety of skull shapes, particularly that seen in domestic dogs (CITE).
-    Conversely, if a study focuses on canid skull variation through time, interspecific data is most appropriate.
-    % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
-    \item \textit{Sample size.}
-    It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
-    For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
-    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction [CITE]).
-    Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
-    \end{itemize}
-
-Of course in practice some of these caveats are unavoidable, especially when dealing with the vertebrate fossil record.
-
-\subsection{Recycling data}
-A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
-One classic case is using ``cladistic'' data in palaeobiology.
-These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
-It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
-A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all of the data.
-For example, in a study of the disparity of dietary niches, removing non-dietary characters is a good approach.
-However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g.
-limb traits can be used for both locomotion and feeding).
-
-\subsection{Other data considerations}
-There are many other considerations when choosing the appropriate data to use. 
-It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
-For example, are the observed differences between two groups related to factors unrelated to group designation such as sexual dimorphism?
-Are the taxa/groups being compared fairly consistent and truly comparable?
-In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated [String of morpho data problems cite].
-It is important to ensure that any patterns detected are not merely reflections of these features of the data.
-
-Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
-
-\section{Use the appropriate methods} 
-In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
-Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
-The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
-Unfortunately, we can only perceive a maximum of three dimensions at the same time.
-Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).
-
-\subsection{Visualising the multidimensional problem}
-\label{visualisation}
-It is important to remember that visual interpretations of multidimensional data can easily be misleading.
-For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
-Conversely, when two groups do not overlap on a two or three dimensional plot, this does not mean they never overlap in the multidimensional space.
-Furthermore, multidimensional spaces are not necessary Euclidean.
-Such non-Euclidean spaces often have non-intuitive properties, for example straight lines in some dimensions are not straight in the space and, even less intuitively, the distance between points A and B might not be equal to the distance between points B and A [CITE Sylvain].
-
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
-\caption{\small{Two groups that overlap in some dimensions do not necessary overlap in all dimensions. 
-Conversely, two groups that do not overlap in some dimensions can overlap in other dimensions.
-In this example for the groups A and B to truly overlap, they must overlap in all three dimensions displayed here.
-Although these groups overlap in the two first dimensions (panel 1), they do not overlap in dimensions 1 and 3 thus do not overlap in the multidimensional space at all.
-The inverse is of course true: if two groups don't overlap in two dimensions (D and C in panel 1), they do not overlap in the multidimensional space, even if they overlap in any other combination of dimensions (dimensions 1 and 3 in panel 2).}}
-\label{Fig:RF_results_best}
-\end{figure}
-
-It is therefore vital to be clear about what is represented in multidimensional space visualisations.
-In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a pattern if all the key properties of the space are not present in the graph.
-We suggest visualisations should include the following:
-(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation in the data present in the plot or only a small fraction?);
-(2) the scale of the axes (ideally, if the space is orthogonal, it is important to keep the same scale across the axes);
-and (3) the error associated with each point, whether this is in terms of measurement error (on ordination plots for example) or some other measure of standard variation (in further analysis, e.g.
-disparity-through-time plots).
-Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading (see box \ref{box_ordination}).
-
-\subsection{[Box]To ordinate or not to ordinate; that is the (multidimensional) question?}
-\label{box_ordination}
-Reducing the dimensionality of data through ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
-Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g.
-selecting only the axes that describe up to 95\% of the variation).
-In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions.
-This has clear advantages for interpreting the results.
-For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).
-
-However, some caveats can arise from such transformations of the data.
-For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g. only the first few) can lead to misinterpretation of the results.
-Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and might have little relation to the disparity question (e.g.
-if the first few axes represent the elongation of the beak, but the question is about wing disparity).
-
-In many cases, ordination might not be necessary.
-For example, if the disparity metric being used relies on all the data (see Metrics section \ref{metrics}) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
-Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
-For example, when the analysed data is non-Euclidean (e.g.
-discrete morphological characters), the ordination can be problematic ([cite Melanie's future paper here]), and the resulting non-Euclidean ordinated space can be difficult to interpret ([probably cite Sylvain's here]).
-This can be problematic when comparing the position of groups in the multidimensional space, as the measured distances might not be comparable.
-Finally, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axes can be impossible or biologically meaningless.
-Although some gradients might be easy to detect or interpret (e.g.
-the elongation of the beak along the first axis), they are often hard to interpret in terms of morphological gradients or simply non-interpretable (e.g.
-with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0 has no biological meaning).
-
-Because of these points, we strongly recommend that data for multidimensional analyses is not automatically ordinated, and that careful consideration is given to whether the question can be answered without ordination.
-\subsection{[Box] end}
-
-\subsection{Phylogeny}
-For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
-For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
-and therefore will not covered further in this review.
-
-\subsubsection{Ancestral state estimation}
-When looking at disparity in the past (e.g. disparity-through-time) ancestral state estimations are a common approach.
-Ancestral state estimation can be performed in two ways:
-\begin{enumerate}
-\item pre-ordination: the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original dataset[CITE].
-\item post-ordination: the estimation is done after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
-\end{enumerate}
-Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e.
-the relationship between the points not estimated) and implies longer computational times but, once the morphospace is defined, it will remain constant (i.e.
-its properties will not change).
-Post ordination ancestral state estimation will not change the ordinated space's geometry and will be faster to compute, however, it will change the morphospace's properties such as the number of parameters (i.e.
-elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper to be?].
-
-More importantly, however, one needs to remember that using such ancestral state estimation methods comes with several caveats.
-First and foremost, although ancestral state estimations are valid methods in macroevolution, their results are highly dependent on the data and chosen methods.
-Secondly, more specific to disparity analysis, when using post-transformation ancestral state estimates, the morphospace can be saturated leading pairwise distance metrics to tend to zero.
-In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points for statistical reasons.
-In fact, these extra points are not independent and can also generate side effects that can be problematic, especially when testing the effect of extinction events on disparity.
-
-Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
-In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes.
-Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution.
-
-\subsection{Metrics}
-\label{metrics}
-
-[INCLUDE WILLS FIGURE SOMWHERE AROUND]
-
-%TG: add the curse of dimensionality
-
-After building and visualising the morphospace, one can use disparity metrics (or indices) as a one dimensional summary of the space and to test hypotheses.
-As with any summary metric, and because of this reduction in dimensionality, disparity metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
-It is often beneficial to use more than one metric to summarise the different aspects of interest of the morphospace (e.g. its width or spread).
-
-The choice of disparity metric is thus essential to the disparity analysis and should always be driven by the question.
-When considering only one dimension, disparity metrics can be used to reflect the ``size'' of the distribution (e.g.
-the range, quantiles or variance), or reflect the most likely value (the central tendency, i.e.
-the mean, median or mode).
-Among these metrics, some will have more preferable properties than others, such as sensitivity to outliers (range, mean or mode are highly sensitive \textit{vs.} quantiles, variance or median are less sensitive) and will thus be more or less appropriate to the different questions and the properties of the distribution.
-For example, if the data collected is likely to contain outliers, but is fairly normally distributed, one should prefer the median to the mean.
-
-With multidimensionality, this high availability of unidimensional descriptors comes with equally numerous advantages and caveats and again, the question can help in determining which metric is the most appropriate.
-For example, if the question refers to ``size'' of a group in the morphospace (e.g.
-``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriate (e.g.
-volume [CITE], variance [CITE], range [CITE], distance from the centroid [CITE]).
-Conversely, if the question refers to the ``position'' of a group in the morphospace (e.g.
-``does group A occupy the same space as group B?''), metrics relating to the distance between the elements within a group and a fixed point in the morphospace are most appropriate.
-Finally, one might also be interested in the density of elements in some regions of the morphospace (e.g.
-``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriate (e.g.
-nearest neighbour distance, pairwise distances, etc.).
-
-Additionally to the desired properties of what the metrics should capture, it is important to also keep in mind the mathematical properties of the metrics and their associated caveats.
-For example, measuring the full sum of the variance of each dimension of the space does not require one to add the co-variance between the axes in a ordinated space using a PCA.
-This is not true however in other mathematical spaces or (even in a PCA) when not all dimensions or elements are considered (e.g.
-when only looking at a group within the morphospace, the sum of variances is not mathematically equal to the sum of the variance of each dimension in the space).
-Furthermore, multidimensional space has some counter-intuitive properties that need to be taken into account such as the curse of dimensionality [CITE].
-For example, product-based metrics as proxies of volumes (e.g.
-product of ranges, hypervolume, hypercube, etc.) will tend towards zero fairly quickly for spaces with even a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem and Donoho 2000 Aide-Memoire].
-Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size, for example range or convex hull based metrics.
-
-%TG: again, I feel this section is already fairly long, I might include this bit in another paper
-
-% NC: Sounds reasonable. This could be cut down more, but for now I've just edited the grammar and spelling so we can look at the thing as a whole and also get Phil's input.
-
-% Of course some other aspects could be of interest (or more than one) and thus some other metrics would be more appropriated.
-% For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution).
-% Furthermore, one future avenue that could be explored is to go beyond single valued disparity metrics and rather express them in terms of distributions.
-% For example, when the metric is expressed as the median distance from centroids [e.g. in CITE], it could be informative as well to express it as a full distribution.
-% This would allow to both visualise the variation at a glance and use this variation in statistical tests for comparing these distributions.
-
-% Best practices:
-% \begin{itemize}
-% \item Facing the amount of different ways to measure disparity and regardless which metric is the most suited to the question, it is crucial to clearly and mathematically define the metric in a disparity paper.
-The definition can be in plain English: e.g.
-``we measure disparity as the sum of the variance of each axis of the ordinated matrix''; or mathematically: e.g.
-``we measured disparity as $\sum\sigma_i$ where $\sigma_i$ is the variance of the each ordinated axis $i$.''.
-% \item It is important to keep in mind that different metrics can lead to the same results even though the described spaces are different (the opposite is \textit{not} true).
-We therefore suggest to always plot some dimensions of the space (see Visualisation section \ref{visualisation}) and try multiple metrics.
-% \end{itemize}
-
-\subsection{Tests}
-The final stage of most disparity analyses is statistical testing of the disparity metrics, to determine whether changes thought time or between groups are statistically significant. 
-Similar to the points above, the statistical test to apply depends on which question is being tackled with disparity analysis, i.e. which hypothesis is being tested.
-It also requires consideration of the data, the ordination method used (or the absence thereof) and the metric used.
-
-For example, one commonly used test in morphological analysis, is the non parametric permutation analysis of variance [NPANOVA or PERMANOVA; CITE Anderson + examples].
-This test and its implementation is perfectly valid to test whether two groups have a different variance/covariance, but is based on the pairwise distances of the group elements.
-This test will ignore any ordination (if the data was ordinated) and is not based on a unidimensional disparity metric.
-In other words, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
-Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the question.
-
-Another caveat for applying statistical tests on disparity data is that it is important to consider which data to apply the test on.
-For example, in morphological disparity analysis (especially for palaeobiological questions), data is often bootstrapped.
-This has two useful advantages: first, when the disparity metric is unidimensional (e.g.
-the sum of variances or the average squared pairwise distance), bootstrapping the data generates a distribution of this unidimensional metric that can then be analysed using the vast statistical toolkit available for comparing distributions; second, when the data is scarce (e.g.
-for palaeontological data), bootstrapping the data allows users to introduce statistical variance, rendering the test less sensitive to outliers.
-However, bootstrapping the data also comes with an important caveat: the data analysed are pseudo-replicated and thus non-independent.
-This can violate the assumptions of parametric statistical tests.
-Furthermore, it is important to remember that the number of bootstrap pseudo-replicates will inevitably increase the Type I error.
-
-\section{Future directions}
-As mentioned above, there is no ``one-size-fits-all'' morphological disparity analysis pipeline.
-As with any multidimensional analysis, there are many nuances in deciding which data to use and how to analyse it, stemming from the explicit formulation of the disparity hypothesis.
-To understand these nuances, it is important to remember that: (1) morphological disparity analyses are not new and much effort has been made in developing methods; and (2) multidimensional analyses are not restricted to morphological data analysis and solutions to specific problems can be found in literature that is not necessary focused on morphology (especially ecology and mathematics).
-
-We believe that many of the caveats in morphological disparity analysis can arise when ``blindly'' applying methodological pipelines at the expense of losing sight of the  biological question being tested.
-We therefore advise keeping the methodology simple and tractable, at least in the preliminary analysis.
-Also, many of the caveats and technical problems in disparity analyses have been tackled in other fields, mainly ecology.
-This is perhaps unsurprisingly as ecological questions are often closely related to evolutionary questions, and ecological data is also often multidimensional [CITE DONOHUE's papers].
-
-\subsection{How can we adapt our methods?}
-Each of the points outlined above may drive new methodological questions (and often already are).
-Although it is impossible to cover the state-of-the-art exhaustively for each of these points here are several solutions and/or directions for adapting existing methods:
-
-\begin{itemize}
-    \item{Which data should we use in disparity analyses?}
-    [One line abstract of Melanie's paper on inapplicable data - Needs to be developed].
-    Furthermore, some simple methods already exist to account for caveats in data. 
-    For example, to correct for the paucity of the fossil record, one can use methods like rarefaction analysis to take sample size into account [CITE].
-    % NC: This item currently doesn't seem to actually give advice about which data to use...
-    \item{How should we visualise disparity?} % NC: Tried to make the bullets a bit clearer, might need to edit for correctness and consistency with the rest of the document.
-    [One line abstract of Sylvain's paper on morphospaces plottings].
-    One major problem of multidimensional analyses is that highly dimensional spaces are hard (or even impossible) to conceptualise.
-    One solution to this is to use more simple 2D/3D ordinations.
-    For example, using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation [e.g Raup-spaces CITE].
-    \item{Which disparity metrics should we use?}
-    This can be tackled by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combination of properties of interest in the morphospace.
-    For example, when reducing a multidimensional space to one metric, it would be judicious to indicate what this metric represents in the results (e.g.
-in a plot or a table, if the metric used was the sum of variances, labelling it as ``size of the group'' rather than ``disparity'').
-    This is especially advantageous when using multiple disparity metrics (e.g. ``size of the group'' and ``position of the group'').
-    \item{How should we test disparity hypotheses?}
-    Finally, it is possible to use non-parametric methods to test specific disparity questions.
-    For example, using available software and computer power, it is possible to propose specific null models for the expectation of what disparity would be expected to look like under various null models.
-    This has been already successfully used in ecology [CITE Diaz paper] and is currently being implemented in palaeobiology (Mark stuff).
-\end{itemize}
-
-\subsection{Where do we go from here?}
-Morphological disparity analyses are a major tool in evolutionary biology, posing a more versatile alternative to studying species richness, especially in macroevolutionary studies.
-We believe that morphological disparity analyses, as the analysis of patterns of morphological differences, can be of great use for linking evolutionary and ecological processes (e.g.
-when using morphological disparity as a proxy for ecological niches to test competition-through-time hypotheses).
-
-However, as demonstrated above, such analyses would greatly benefit from more development, and  more care is needed in study design so that the analysis actually reflects the biological question at hand.
-This can not only be done through the points highlighted above but also through linking the observed patterns to an explicit process (for example, the link between an organism's form and function) or by developing null models to potentially explain the observed patterns.
-
-Finally, we encourage to a more cross-disciplinary approach to morphological disparity analyses.
-In fact, as mentioned above, morphological disparity could greatly benefit from advances in multidimensional analysis in different fields (namely ecology) but similarly, the concept of a morphospace could easily be exported to other fields.
-For example, the multidimensional analysis in Diaz [CITE] looking at the patterns of form and function in plants can be thought as a morphospace or a life-history-space (an ecospace?); isotopic analyses in [CITE Andrew's stuff] can be represented as a isotope-space, etc.
-These generalisations could also be exported for any set of traits (e.g.
-acoustospaces for acoustic traits) and even beyond evolutionary biology (e.g.
-glottospaces for linguistic traits) [CITE Katie's review].
 

### 2018-11-02 Analysis/Figures_in_R.Rmd
Moved figures Rmd to analysis

---
title: "Figures"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---

# Methods part

```{r}
library(dispRity)
```

## Visualisation

A multidimensional space similar to Figure 3 in Davis et al 2012 Nature.

```{r, fig.width = 5, fig.height = 8.33333}
set.seed(42)
## Sampling the four different groups
size <- 10            # x                      #y                       #z
group1 <- cbind(rnorm(size, -0.5, 0.25), rnorm(size,  0.5, 0.25), rnorm(size,  0.5, 0.25))
group2 <- cbind(rnorm(size, -0.5, 0.25), rnorm(size,  0.5, 0.25), rnorm(size, -0.5, 0.25))
group3 <- cbind(rnorm(size,  0.5, 0.25), rnorm(size,  0.5, 0.25), rnorm(size,  0.5, 0.25))
group4 <- cbind(rnorm(size,  0.5, 0.25), rnorm(size, -0.5, 0.25), rnorm(size,  0.5, 0.25))
groups <- list(group1, group2, group3, group4)

## Chull coordinates function
get.chull.coords <- function(x, axis) {
    chull_points <- chull(x[, c(axis)])
    chull_coords <- rbind(x[chull_points, axis], x[chull_points[1], axis])
}

## Getting the coordinates
chull_coords_xy <- lapply(groups, get.chull.coords, axis = c(1, 2))
chull_coords_xz <- lapply(groups, get.chull.coords, axis = c(1, 3))

## Empty plot function
empty.plot <- function(xlab = "Dimension 1", ylab) {
    par(bty = "n")
    plot(NULL, xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1), xlab = xlab, ylab = ylab, xaxt='n', yaxt='n')
    abline(v = 0)
    abline(h = 0)
}

## Setting the colour palette
col_palette <- as.list(viridis::viridis(4))

plot.figure.dimensionality <- function(){
    par(mfrow = c(2, 1))
    par(mar = c(1, 4, 1, 3))
    ## Plotting the first coordinates sets
    empty.plot(xlab = "", ylab = "Dimension 2")
    text(-1,1, "1", cex = 2)
    ## Adding the points
    plot.points <- function(x, col, axis) points(x[, axis], col = col, pch = 19, cex = 0.7)
    silent <- mapply(plot.points, groups, col_palette, MoreArgs = list(axis = c(1, 2)))
    ## Adding the polygons
    plot.polygons <- function(x, col) polygon(x[, 1], x[, 2], col = col, density = 10, border = col)
    silent <- mapply(plot.polygons, chull_coords_xy, col_palette)
    legend("bottomlef", legend = LETTERS[1:4], col = unlist(col_palette), bty = "n", pch = 19, lty = 1)

    par(mar = c(4, 4, 1, 3))
    ## Plotting the second coordinates sets
    empty.plot(ylab = "Dimension 3")
    text(-1,1, "2", cex = 2)
    ## Adding the points
    silent <- mapply(plot.points, groups, col_palette, MoreArgs = list(axis = c(1, 3)))
    ## Adding the polygons
    silent <- mapply(plot.polygons, chull_coords_xz, col_palette)
}

plot.figure.dimensionality()

## Figure for the paper
pdf(file = "../Writing/Figures/figure_dimensionality.pdf", width = 5, height = 8.33333)
plot.figure.dimensionality()
dev.off()
```

## Metrics

```{r, fig.width = 10, fig.height = 10}
set.seed(42)

## Scale the results to be between -1 and 1
scale.matrix <- function(matrix) {
    ## Get the range for scaling
    range <- apply(matrix, 2, range)

    ## Shift to positive
    shift_matrix <- matrix
    shift_matrix[,1] <- matrix[,1] + abs(range[1,1])
    shift_matrix[,2] <- matrix[,2] + abs(range[1,2])

    ## Scale the matrix (to + 2)
    shift_matrix <- apply(shift_matrix, 2, function(x) x/(max(x)/2))

    ## Reshift the matrix
    return(shift_matrix - 1)
}

## Plotting the two matrices
plot.both.matrices <- function(matrix1, matrix2, col = c("blue", "orange"), ...) {
    plot(matrix1, pch = 21, bg = col[1], ..., xlab = "", ylab = "")
    points(matrix2, pch = 21, bg = col[2])
}

## Two similar correlation matrices
cor_mat_positive <- matrix(c(1, 0.8, 0.8, 1), 2, 2)
cor_mat_negative <- matrix(c(1, -0.8, -0.8, 1), 2, 2)

## Correlation differences
correlated_positive <- scale.matrix(space.maker(100, 2, rnorm, cor.matrix = cor_mat_positive))
correlated_negative <- scale.matrix(space.maker(100, 2, rnorm, cor.matrix = cor_mat_negative))

## Doughnut differences
outer <- space.maker(100, 2, random.circle, arguments = list(list(runif, inner = 0.5)))
inner <- space.maker(100, 2, random.circle, arguments = list(list(runif, outer = 0.5)))

## Density differences
base_matrix <- space.maker(100, 2, runif, arguments = list(list(max = 1, min = -1)))
modif_matrix <- apply(base_matrix, 2, function(x) x+rnorm(length(x), sd = 0.035))

## Position differences
upper_right <- space.maker(100, 2, rnorm, arguments = list(list(mean = 0.5, sd = 0.2)))
lower_left <- space.maker(100, 2, rnorm, arguments = list(list(mean = -0.5, sd = 0.2)))

plot.figure.metrics <- function(){
    par(mfrow = c(2,2), mar = c(2.5,2.5,2.5,2.5))
    plot.both.matrices(correlated_positive, correlated_negative, main = "Correlation")
    plot.both.matrices(outer, inner, main = "Filling")
    plot.both.matrices(base_matrix, modif_matrix, main = "Density")
    plot.both.matrices(upper_right, lower_left, xlim = c(-1, 1), ylim = c(-1, 1), main = "Position")
}

plot.figure.metrics()

## Figure for the paper
pdf(file = "../Writing/Figures/figure_metrics.pdf", width = 10, height = 10)
plot.figure.metrics()
dev.off()

```

### 2018-11-19 Analysis/Figures_in_R.Rmd
updated figures plots

@@
 ```
+
+## Figure methods
+
+```{r}
+set.seed(42)
+## Creating a 5 trait space
+iris_rand <- cbind(iris[seq(from = 1, to = 150, by = 5), 1:4], rnorm(30))
+
+## Plotting the 5 traits
+pdf(file = "../Writing/Figures/figure_methods_biplot.pdf", width = 5, height = 5)
+plot(iris_rand, pch = 19, xaxt = "n", yaxt = "n")
+dev.off()
+
+## Making a distance matrix
+iris_dist <- dist(iris_rand)
+
+## Making it 2D
+pdf(file = "../Writing/Figures/figure_methods_2dPCO.pdf", width = 5, height = 5)
+plot(cmdscale(iris_dist, k = 2), pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
+dev.off()
+
+## Making a PCA
+pca <- princomp(iris_rand)$scores
+pdf(file = "../Writing/Figures/figure_methods_PCA1.pdf", width = 5, height = 5)
+plot(pca[,1:2], pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
+dev.off()
+pdf(file = "../Writing/Figures/figure_methods_PCA2.pdf", width = 5, height = 5)
+plot(pca[,3:4], pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
+dev.off()
+pdf(file = "../Writing/Figures/figure_methods_PCA3.pdf", width = 5, height = 5)
+plot(pca[,c(5,1)], pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
+dev.off()
+```

### 2019-10-30 Analysis/Figures_in_R.Rmd
Updated figures

@@
 ```
+
+## Iris example
+
+```{r}
+## Preparing the dataset
+colnames(iris)[1:4] <- gsub("\\.", "\n", colnames(iris)[1:4])
+rownames(iris) <- c(paste0("setosa", which(iris[,5] == "setosa")),
+                    paste0("versicolor", which(iris[,5] == "versicolor")-50),
+                    paste0("virginica", which(iris[,5] == "virginica")-100))
+
+## Morphospaces:
+## No transformations
+write.csv(head(iris[, -5]), file = "../Writing/Figures/trait.space.no_trans.csv")
+
+## Distance matrix
+iris_dist <- dist(iris[, -5])
+write.csv(round(head(as.matrix(iris_dist)[1:5, 1:5]), digits = 2), file = "../Writing/Figures/trait.space.distance.csv")
+
+## Ordinations
+iris_pca <- princomp(iris[, -5])$scores
+write.csv(round(head(iris_pca), digits = 2), file = "../Writing/Figures/trait.space.pca.csv")
+iris_pco <-cmdscale(iris_dist, k = 2)
+
+## Visualisations
+## biplots
+col_var <- c("orange", "blue", "green")[as.factor(iris[,5])]
+pdf(file = "../Writing/Figures/figure_methods_biplots.pdf", width = 5, height = 5)
+plot(iris[, -5],
+    pch = 19,
+    col = col_var)
+dev.off()
+
+## PCO
+vars <- round(apply(iris_pco, 2, var)/sum(apply(iris_pco, 2, var))*100, 2)
+plot(iris_pco,
+    pch = 19,
+    col = col_var,
+    xlab = paste0("PCO 1 (", vars[1], "%)"),
+    ylab = paste0("PCO 21 (", vars[2], "%)"))
+
+## PCA
+pdf(file = "../Writing/Figures/figure_methods_PCA.pdf", width = 5, height = 5)
+vars <- round(apply(iris_pca, 2, var)/sum(apply(iris_pca, 2, var))*100, 2) 
+ranges <- apply(iris_pca, 2, range)
+
+par(mfrow = c(4,4), mar = c(1,1,1,1), oma = c(2,2,2,2))
+## First row
+# 1
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PCA 1\n(", vars[1], "%)"), cex = 1)
+# 2
+plot(iris_pca[, c(1,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(3)
+# 3
+plot(iris_pca[, c(1,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 4
+plot(iris_pca[, c(1,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(3)
+axis(4)
+
+## Second row
+# 1
+plot(iris_pca[, c(2,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(2)
+# 2
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PCA 2\n(", vars[2], "%)"), cex = 1)
+# 3
+plot(iris_pca[, c(2,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 4
+plot(iris_pca[, c(2,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(4)
+
+## Third row
+# 1
+plot(iris_pca[, c(3,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(2)
+# 3
+plot(iris_pca[, c(3,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 3
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PCA 3\n(", vars[3], "%)"), cex = 1)
+# 4
+plot(iris_pca[, c(3,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(4)
+
+## Fourth row
+# 1
+plot(iris_pca[, c(4,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(1)
+axis(2)
+# 3
+plot(iris_pca[, c(4,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 3
+plot(iris_pca[, c(4,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(1)
+# 4
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PCA 4\n(", vars[4], "%)"), cex = 1)
+dev.off()

### 2019-10-30 Writing/disparity_review.tex
Updated figures

--- 2018-05-18 Writing/disparity_review.tex
@@
-\documentclass[12pt,letterpaper]{article}
-\usepackage{natbib}
+RH: disparate disparity analyses.
 
-%Packages
-\usepackage{xcolor}
-\usepackage{color,soul}
-\usepackage{pdflscape}
-\usepackage{fixltx2e}
-\usepackage{textcomp}
-\usepackage{fullpage}
-\usepackage{float}
-\usepackage{latexsym}
-\usepackage{url}
-\usepackage{epsfig}
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{array}
-\usepackage[version=3]{mhchem}
-\usepackage{ifthen}
-\usepackage{caption}
-\usepackage{hyperref}
-\usepackage{amsthm}
-\usepackage{amstext}
-\usepackage{enumerate}
-\usepackage[osf]{mathpazo}
-\usepackage{dcolumn}
-\usepackage{lineno}
-\usepackage{dcolumn}
-\usepackage{hyphenat}
-\usepackage[T1]{fontenc}
-\usepackage{textcomp}
-\newcolumntype{d}[1]{D{.}{.}{#1}}
+\textbf{The disparate ways to analyse morphological disparity}
 
-\pagenumbering{arabic}
+Thomas
+Guillerme\includegraphics[width=0.4902in,height=0.1807in]{./ObjectReplacements/Object 2},
+Natalie
+Cooper\includegraphics[width=0.3984in,height=0.1807in]{./ObjectReplacements/Object 4},
+Stephen
+Brusatte\includegraphics[width=0.1937in,height=0.1807in]{./ObjectReplacements/Object 6},
+Katie
+Davis\includegraphics[width=0.1992in,height=0.1807in]{./ObjectReplacements/Object 8},
+Andrew L
+Jackson\includegraphics[width=0.1937in,height=0.1807in]{./ObjectReplacements/Object 10},
+Sylvain
+Gerber\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 12},
+Anjali
+Goswami\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 14},
+Kevin
+Healy\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 16},
+Melanie
+Hopkins\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 18},
+Graeme
+Lloyd\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 20},
+Joseph
+O'Reilly\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 22},
+Abi
+Pate\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 24},
+Emilie
+Rayfield\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 26},
+Erin
+Saupe\includegraphics[width=0.239in,height=0.1807in]{./ObjectReplacements/Object 28},
+Emma
+Sherratt\includegraphics[width=0.2425in,height=0.1807in]{./ObjectReplacements/Object 30},
+Graham
+Slater\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 32},
+Gavin H
+Thomas\includegraphics[width=0.2465in,height=0.1807in]{./ObjectReplacements/Object 34}
+and Philip
+Donoghue\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 36}\\
+\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 38}\emph{School
+of Biological Sciences, University of Queensland, St. Lucia, Queensland,
+Australia.}\\
+\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 40}\emph{Department
+of Life Sciences, Natural History Museum, London, Cromwell Road, London,
+SW7 5BD, UK.}\\
+\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 42}\emph{Bristol}
 
-%Pagination style and stuff
-\linespread{2}
-\raggedright
-\setlength{\parindent}{0.5in}
-\setcounter{secnumdepth}{0} 
-\renewcommand{\section}[1]{%
-\bigskip
-\begin{center}
-\begin{Large}
-\normalfont\scshape #1
-\medskip
-\end{Large}
-\end{center}}
-\renewcommand{\subsection}[1]{%
-\bigskip
-\begin{center}
-\begin{large}
-\normalfont\itshape #1
-\end{large}
-\end{center}}
-\renewcommand{\subsubsection}[1]{%
-\vspace{2ex}
-\noindent
-\textit{#1.}---}
-\renewcommand{\tableofcontents}{}
-%\bibpunct{(}{)}{;}{a}{}{,}
++ These authors contributed equally to the manuscript.
 
-%---------------------------------------------
-%
-%       START
-%
-%---------------------------------------------
+*\textbf{Corresponding author: }\emph{guillert@tcd.ie}\\
 
-\begin{document}
+\textbf{NOTE TO ALL THE CO-AUTHORS:}
 
-%Running head
-\begin{flushright}
-Version dated: \today
-\end{flushright}
-\bigskip
-\noindent RH: pest practices for disparity analysis.
+\textbf{-Please write in Google docs using }\emph{\textbf{Suggesting
+mode}}\textbf{ only }(upper right corner of your screen)
 
-\bigskip
-\medskip
-\begin{center}
+\textbf{-The word limit for this is 5000 words so we can't add any large
+sections.}
 
-\noindent{\Large \bf Analysing disparity in morphology data: best practice and future directions}
+\textbf{-There is no reference limit per se but please don't add too
+many. I will sort out the references later so don't correct the
+formatting of them here.}
 
-\bigskip
+-I suggest you add your references using the \textbf{Paperpile} addon
+(go in Add-ons menu \textgreater{} Get Add-ons\ldots{} \textgreater{}
+Search for Paperpile \textgreater{} Add/Install; then go in Add-ons
+menus \textgreater{} Paperpile \textgreater{} Manage citation). If you
+don't use Paperpile, please don't add references as just (``Someone et
+al. 2000'' but add also a link to the correct Google Scholar entry in
+comments).
 
-\noindent {\normalsize \sc Thomas Guillerme$^{1,*,+}$, Natalie Cooper$^{2,+}$, Stephen Brusatte$^3$, Katie Davis$^4$, Andrew Jackson$^5$, Sylvain Gerber$^6$, Anjali Goswami$^2$, Kevin Healy$^7$, Melanie Hopkins$^8$, Graeme Lloyd$^9$, Joseph O'Reilly$^{10}$, Abi Pate$^{10}$, Emilie Rayfield$^{10}$, Erin Saupe$^{11}$, Emma Sherratt$^{12}$, Graham Slater$^{13}$, Gavin H Thomas$^{14}$ and Philip Donoghue$^{10,+}$}\\
+\textbf{-Add your affiliation in the author list above}. The
+affiliations count in the word count so please 1) keep it short (max 1
+per person!) and 2) try to use one that's already in there. Also add any
+initials I've missed in your names.
 
-\noindent {\small \it 
-$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia.\\ 
-$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK.\\ 
-$^{10}$Bristol}
-\end{center}
-\medskip
-\noindent{+ These authors contributed equally to the manuscript.}
-\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
-\vspace{1in}
+- If in doubt, check formatting here:
+\href{https://royalsociety.org/journals/authors/author-guidelines/}{\emph{https://royalsociety.org/journals/authors/author-guidelines/}}
 
-%Line numbering
-\modulolinenumbers[1]
-\linenumbers
+Thanks!
 
-%---------------------------------------------
-%
-%       ABSTRACT
-%
-%---------------------------------------------
+\textbf{Abstract (200 words max)}
 
-\newpage
-\begin{abstract}
+Morphological disparity analysis is used to characterize and investigate
+discrete patterns of variation in the shape, form, function, and the
+ecology of organisms. Since their conception in the 1980s, these methods
+have been used to ask a variety of questions about how and why
+morphological diversity evolves, and the ecological consequences of
+these changes. There are many methodological approaches for studying
+disparity, however, all of these come with caveats and pitfalls that
+can, if ignored, lead to misinterpretations of disparity patterns. Here
+we review morphological disparity methods and their caveats, and provide
+best practice guidelines for their application and interpretation. We
+discuss the data required for disparity studies, as well as
+methodological challenges, including multidimensionality, ordination,
+metric choice, hypothesis testing, and ancestral state estimations.
+There is no ``one-size-fits-all'' approach to characterizing and
+investigating the evolution of morphological disparity, and the
+available tools should always be used in the context of a specific
+biological question that will determine data and method selection at
+every stage of the analysis.
 
-Variation in organisms' shape, ecology or behaviour is often more discrete rather than continuous.
-One way to study this variation was pioneered in the 1990s by looking at the disparity of morphologies in the Cambrian.
-Morphological disparity analysis - a term encompassing both the patterns and process of morphological evolution through time - became rapidly popular and lead to numerous publications in the last two decades.
+\textbf{Keywords}: multidimensionality, palaeobiology, ecology\\
 
-Here, we suggest that this diversification of disparity analysis lead to two contrasted effects:
-(1) the great diversity of methodological approaches to study disparity makes it sometimes hard to understand all caveats and pitfalls of such analysis - a problem that can lead to major misinterpretation of disparity patterns - and;
-(2) in essence, disparity is a pattern of variation meaning that disparity analysis can be expanded beyond the study of morphology.
+1. Introduction
 
-In this review, we propose a set of non-prescriptive guidelines that will help researchers to think about specific disparity analysis designs, with their caveats and their advantages.
-Furthermore, we highlight potential avenues of expansion of disparity analysis both in terms of improving the existing methods for studying morphological disparity and in terms of expanding disparity analysis beyond morphology.
+Variation in the shape, form and function of organisms often exhibits a
+pattern of discrete clusters rather than continuous gradients. These
+patterns can be pictured as multidimensional spaces encompassing all the
+possible forms an organism can take, and where each individual region of
+this space is the result of a unique combination of many variables. One
+commonly used approach to analyse these patterns and to understand the
+processes that lead to them, is to analyse the morphological diversity
+of a group, often referred to as a ``morphological disparity analysis''
+\href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline et al. 2018)}.
 
-\end{abstract}
+The initial informal definition of morphological disparity was
+``multidimensional morphological dissimilarity at a macroevolutionary
+scale'' (Runnegar 1987,
+\href{https://paperpile.com/c/sTGYvp/Uns3}{(Gould 2000)}). In the
+seminal disparity papers of the late 1980s, disparity was used to
+investigate body plan evolution after the Cambrian explosion
+\href{https://paperpile.com/c/sTGYvp/CiPy}{(Briggs, Fortey, and Wills
+1992)}\href{https://paperpile.com/c/sTGYvp/CidX+6tNm+oenu+TtGs}{(Gould
+1991; Valentine 1986; M. Foote and Gould 1992; Morris 1989)};
+\href{https://paperpile.com/c/sTGYvp/CiPy}{(Briggs, Fortey, and Wills
+1992)};. Early disparity analyses focused on broad questions such as:
+How does diversity evolve? Are some phenotypes more common than others?
+and Can anatomical shapes evolve in all ``directions'' of the
+multidimensional space
+\href{https://paperpile.com/c/sTGYvp/yqPw+fTJ3}{(Mike Foote 1997,
+1995)}? Since the 1990s, a great deal of literature from a range of
+sub-fields in biology and palaeontology, and an equally large number of
+methods, have used ``disparity'' as a broader umbrella term for studies
+of multivariate variation. Although most influential reviews have
+treated disparity as a pattern
+\href{https://paperpile.com/c/sTGYvp/yqPw+nFf7+vTHS}{(Mike Foote 1997;
+Wills 2001; Melanie J. Hopkins and Gerber 2017)}, there is currently no
+consensus on whether to restrict use of the term disparity to its
+original macroevolutionary-scale definition, or whether it can refer to
+any study of multivariate variation. Here we use a broader definition
+and suggest that morphological disparity is a \emph{pattern of variation
+}in phenotype. Disparity analysis has been used to tackle a vast array
+of questions. We identify four main classes of disparity analysis as
+follows.
 
-\noindent (Keywords: disparity)\\
+(1) \emph{Descriptive disparity} \emph{analyses. }The seminal studies of
+disparity described the shapes of organisms and how they differed among
+groups \href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Mike Foote 1995;
+Wills, Briggs, and Fortey 1994)}. These consist of descriptive
+multidimensionality analyses of the patterns in the diversity of
+morphological traits, addressing questions such as: why are some
+morphological traits combinations more common than others and what are
+the properties of the resulting morphospace?
+\href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Mike Foote 1995;
+Raup 1961; Gerber 2017)}.
 
-\vspace{1.5in}
+(2) \emph{Disparity-through-time} \emph{analyses. }This approach
+investigates how the shapes of organisms have changed through time,
+focussing on the disparity of taxa in particular time bins or slices.
+This approach has been widely used in palaeobiology to answer a range of
+macroevolutionary questions
+\href{https://paperpile.com/c/sTGYvp/PbSx+khc9+xxh5}{(Close et al. 2015;
+Cisneros and Ruta 2010; Hughes, Gerber, and Wills 2013)}, such as: how
+does disparity accumulate over the history of a clade
+\href{https://paperpile.com/c/sTGYvp/ekU4+s33b}{(Guillerme and Cooper
+2018; Wright 2017)}, and how does disparity change across mass
+extinction events \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman
+2010)}?
 
-\newpage
+(3) \emph{Disparity and diversity. }Morphological disparity provides
+another perspective on biodiversity; high morphological disparity
+represents a high diversity of morphologies (i.e. shapes or body plans)
+and is, presumably, associated with high levels of ecological and
+functional diversity (see below). This makes disparity a more useful
+measure of diversity than merely species richness. Indeed, most studies
+that have investigated disparity and taxonomic diversity support an
+effective decoupling
+\href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(Fortey,
+Briggs, and Wills 1996; Ruta et al. 2013; M. J. Hopkins 2013; Moyne and
+Neige 2007)} - though this result breaks down at extremes of diversity;
+extremely low diversity tends to yield low disparity and \emph{vice
+versa}. This approach has been used to investigate whether some groups
+are more successful than other in their exploration of new evolutionary
+strategies @@@.
 
-%---------------------------------------------
-%
-%       INTRODUCTION
-%
-%---------------------------------------------
-\section{Introduction}
+(4) \emph{Disparity and ecology. }The disparity of a group can be used
+as a proxy for either the functional role it plays within an ecosystem
+or its ecological niche. These analyses assume that groups with high
+disparity are also likely to be functionally and ecologically diverse,
+and that groups found in similar regions of shape space will have
+similar functional and ecological roles
+\href{https://paperpile.com/c/sTGYvp/EETc+tSIy+qjj9}{(Friedman 2010;
+Pierce, Angielczyk, and Rayfield 2008; P. S. L. Anderson et al. 2011)}.
+Note however, that the links between form and function are not always
+clear cut; traits can be linked to multiple functions and multiple
+functions can be linked to a single trait
+\href{https://paperpile.com/c/sTGYvp/Ejzr}{\emph{(Wainwright et al.
+2005)}}. This approach is used to investigate competitive replacements
+\href{https://paperpile.com/c/sTGYvp/EeC8}{(Stephen L. Brusatte et al.
+2008)} and changes in ecosystem function during and after mass
+extinctions \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}.
+It is particularly common in paleobiology where it is not possible to
+observe directly the ecological or functional characteristics of extinct
+species. \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.
+2005)}
 
-Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
-These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
-One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
-% The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
-% Each individual carnivore species occupies a discrete region of this space related to its values for these traits.
+Of course, these four broad categories are not independent and many
+studies use a combination of approaches
+\href{https://paperpile.com/c/sTGYvp/EeC8}{(Stephen L. Brusatte et al.
+2008)}. Furthermore, disparity analysis is continuously expanding into
+areas. While this is exciting for the field, there are several issues
+that may reduce the utility of disparity analyses if they are not
+tackled. Here we review these issues and present best practice
+guidelines to allow researchers to optimise their analyses. We first
+discuss the appropriate data required for disparity analyses, then
+review various challenging methodological aspects of disparity analyses.
+Throughout it is important to remember that these tools should always be
+used in the context of a specific scientific question, as this will
+drive data and methodological choices at every stage of the process.
 
-In the seminal disparity papers from the late 1980s, disparity was used to investigate body plan evolution after the Cambrian explosion (REF)[Runnegar 1987, Gould, Foote, Webb, etc.].
-Early disparity analyses focused on broad questions such as: How does diversity evolve?
-Are some body types more common than others?
-Can shape evolve in all ``directions'' (of the multidimensional space)?
-Since the 1980s, a great deal of literature from a range of sub-fields in biology, and an equally large number of methods, have used the term ``disparity''.
-Disparity analysis has been used to tackle a vast array of questions, for example in evo-devo (i.e.
-developmental constraints [CITE]; relationships between form and function [CITE]; evolution functional traits [CITE]; and scaling from micro- to macro-level processes [CITE]), in ecology and palaeoecology (i.e.
-intra-specific competition [CITE], niche occupancy[CITE]; speciation[CITE]; competition [CITE]; and macroecology/disparity in space[CITE]), and macroevolution (i.e.
-responses to extinction and extinction selectivity [CITE]; intra-specific competitive replacements [CITE]; and tempo and mode of evolution [CITE]).
+\includegraphics[width=6.5in,height=4.9445in]{Pictures/10000201000002EE0000023BE5B78D4C4B79F709.png}
 
-The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
-Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation (REF Katie et al's paper somewhere here).
-Here we use a broader definition and suggest that disparity is a pattern of variation.
-Throughout this review, we aim to highlight the diversity of ``disparity'' analyses and how they can aid our understanding of the structure of biodiversity.
+\emph{Figure disparities {[}figure\_disparities.pdf{]}: The four main
+types of disparity analysis. Descriptive disparity analyses focus on
+describing the properties of the morphospace ; Disparity-through-time
+analyses investigate the evolution of the morphospace through time
+including the effect of extinction events; Disparity and diversity
+analysis use disparity as a proxy for species diversity; Disparity and
+ecology analyses use disparity as a proxy for the ecological or
+functional role of a group. These categories are not independent and
+many analyses will cover more than one.}
 
-% \subsection{What \textit{is} disparity?} %TG: maybe useful for box or paraphrapsing summaries.
-% What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
-% The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
-% Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
-% There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
-% Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
-% The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).
+2. Data in Disparity Analyses
 
-% Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
-% In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
-% Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
+Regardless of the precise approach being used, disparity analyses are
+based on morphological traits. It is important, therefore, to consider
+which traits to use in the context of the specific question under
+investigation. Traits for disparity analyses come from three main types
+of data: (1) discrete ``cladistic'' characters, i.e coding the
+absence-or-presence of features or a discrete characteristic of a trait
+\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}); (2)
+continuous measurements of features
+\href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al.
+2011)}; or (3) more sophisticated mathematical measurements from
+geometric morphometric data (e.g. Procrustes superimpositions
+\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, or
+Fourier contours \href{https://paperpile.com/c/sTGYvp/ZEDR}{(Spriggs et
+al. 2018)}). None of these are fundamentally better, however, some
+questions and methodological pipelines are more appropriate with
+specific types of data.
 
-\section{Best practices for disparity analysis}
+When investigating variation within bat wing shapes, for example, both
+homologous landmarks from geometric morphometric analysis and continuous
+measurements of bones may be appropriate to capture patterns of wing
+variation. However, if the question focuses on the convergence between
+bat and bird wings, the homology of the traits to measure is more
+complex to define and the outline shape of the wings might more readily
+answer the question from a mechanical perspective. Similarly, if the
+focus is on wing function, i.e. whether the aerodynamic properties of
+wings vary within bats or between bats and birds, the traits collected
+should reflect these aerodynamic properties (e.g. wingspan, aspect
+ratio, etc.). Whereas if the focus is on convergence between different
+bats and birds, it would be preferable to use traits that have
+facilitated flight in both groups (e.g. digit length, integumentary
+system, etc.). Where there is any doubt about the appropriate traits to
+choose, we advise using different kinds of data for the same feature to
+determine whether it captures the same disparity.
 
-This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
-However, as with any great power, it comes with great responsibility. 
-With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
-First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
-Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
-Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
-Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.
+This dicussion assumes that researchers are collecting their own data
+for disparity analyses, but often this is not the case. It is common,
+especially in disparity-through-time analyses, to recycle discrete
+character data collected for use in phylogenetic inference (e.g. REFs
+that have done this). This has the advantage of being quick, cheap and
+efficient, but tends to artificially increase disparity between
+phylogenetically distinct groups because phylogenetic characters are
+collected specifically for the purpose of discriminating groups
+\href{https://paperpile.com/c/sTGYvp/fTJ3}{(Mike Foote 1995)}. This
+needs to be considered when interpreting results, especially as when new
+clades appear, the synapomorphies of that clade will naturally lead to
+an apparent shift or increase in disparity. Similarly, data used for
+phylogenetic reconstruction is likely to be highly skewed towards easily
+available and taxonomically relevant traits, for example teeth in
+mammals. If your disparity question is related to feeding ecology or
+similar then this may be appropriate, but if the focus is on dispersal
+ability then it would be sensible to select only postcranial traits. It
+is often tempting to use all of a large dataset to increase the power of
+your analyses, but if the data are not suitable for the question at hand
+then the results may be meaningless (although ssome studies have suggest
+this may not be a major issue;
+\href{https://paperpile.com/c/sTGYvp/ZqML+Qrba}{(Zou and Zhang 2016;
+Dávalos et al. 2014)}
+\href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}).
 
-\section{What is the question?}
-Why perform a disparity analysis?
-This seems like it should be obvious, but we stress its crucial importance.
-The last decade has seen a vast array of new methods being published and implemented in freely available software, making disparity analyses easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
-However, this (consciously or not) pushes authors towards performing disparity analyses simply because of their ease of implementation, rather than stemming from a question driven approach.
-Before rushing into a disparity analysis, one should consider some basic caveats.
-For example, one major thing to consider is circularity: if the aim is to explore the variation in a dataset, disparity analysis is perfectly valid.
-However, if this exploratory analysis reveals distinct groups, it would be circular to then test for differences among those groups as if this had been the hypothesis all along [CITE stats 1.0].
-This can easily be avoided by using a subset of the data for preliminary data exploration, and then using the full dataset to test for differences among any groups that emerge.
-Clearly framing questions and their null expectations will help in designing better disparity analyses.
+Finally, trait data suffers from the same shortcomings as most data --
+data can be missing, non-overlapping, hierarchical, inapplicable,
+ambiguous, polymorphic, correlated, or have insufficient sample size
+\href{https://paperpile.com/c/sTGYvp/yO2t}{(Palci and Lee 2018)},
+\href{https://paperpile.com/c/sTGYvp/Yrbg}{(Brazeau, Guillerme, and
+Smith 2017)} {[}String of morpho data problems cite{]}. Biological
+phenomena such as allometry and sexual dimorphism may also influence
+trait data. More practically, data collection is constrained by the time
+and money available, making collating a ``perfect'' dataset difficult.
+We suggest that trait data are collected with the question in mind and
+considering all of the caveats above, but where issues are unavoidable
+at the very least these caveats must be considered during the analyses
+and subsequent interpretation of the results.
 
-\section{Collect the appropriate data}
-Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
-For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for species delimitation.
-Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).
+\emph{Figure data {[}figure\_data.pdf{]}: major routes to obtain
+disparity data. Data can be collected as discrete trait observations
+(e.g. presence or absence data) or as continuous data. Continuous data
+can be collected by different methods: linear measurements, landmark
+coordinates or contours (curves). These measurement can then in turn be
+mathematically transformed (logarithm, scaling, Procrustes
+Superimposition, Elliptic Fourier, etc.). Regardless of the method, this
+data collection results in a trait matrix where the observed traits
+constitute columns and the studied elements (e.g. taxa) the rows.}
 
-\subsection{Data types}
-Morphological disparity analysis can be based on three main types of data:
-\begin{enumerate}
-    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g.
-the presence or absence of legs, the colour of a tail, etc.
-[CITE]) or count the number of features (e.g.
-number of digits, etc.
-[CITE]);
-    \item continuous data that can be either simple measurements (e.g.
-body length, etc.
-[CITE]) or more abstracted measures (e.g.
-geometric morphometrics, mathematical contour description, etc.
-[CITE]);
-    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
-\end{enumerate}
-None of these types of data are fundamentally better but each will be more appropriate for specific questions.
-For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
-On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
-This again illustrates the importance of framing the question.
-In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
-wingspan, aspect ratio, etc.).
-Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g.
-digit length, integumentary system, etc.).
+\includegraphics[width=6.5in,height=6.5in]{Pictures/10000201000007D0000007D0F8A72A6590C667D3.png}
 
-\subsection{How much data should be collected?}
-Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of specimens and traits.
-\begin{itemize}
-    \item \textit{Taxonomic scope.}
-    Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
-    Taxonomic scope will depend on the question being investigated, and what type of variation is being analysed.
-    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation to capture the full variety of skull shapes, particularly that seen in domestic dogs (CITE).
-    Conversely, if a study focuses on canid skull variation through time, interspecific data is most appropriate.
-    % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
-    \item \textit{Sample size.}
-    It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
-    For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
-    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction [CITE]).
-    Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
-    \end{itemize}
+3. Disparity Methods
 
-Of course in practice some of these caveats are unavoidable, especially when dealing with the vertebrate fossil record.
+Once suitable trait data have been collected, we need to consider the
+disparity analysis. Below we address several key aspects including (i)
+the difficulty of dealing with multidimensional data; (ii) caveats
+associated with using ordination techniques to reduce the dimensionality
+of the data; (iii) various metrics used to summarise the disparity of a
+group; (iv) caveats related to methods that exist for hypothesis testing
+within a disparity analysis framework; and (v) the specific case of
+ancestral state estimations in disparity-through-time analyses. As noted
+above, these tools should always be used in the context of a specific
+question, as this will determine which methods are most appropriate.
 
-\subsection{Recycling data}
-A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
-One classic case is using ``cladistic'' data in palaeobiology.
-These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
-It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
-A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all of the data.
-For example, in a study of the disparity of dietary niches, removing non-dietary characters is a good approach.
-However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g.
-limb traits can be used for both locomotion and feeding).
+\protect\hypertarget{anchor}{}{}3.1. Multidimensional mayhem
 
-\subsection{Other data considerations}
-There are many other considerations when choosing the appropriate data to use. 
-It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
-For example, are the observed differences between two groups related to factors unrelated to group designation such as sexual dimorphism?
-Are the taxa/groups being compared fairly consistent and truly comparable?
-In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated [String of morpho data problems cite].
-It is important to ensure that any patterns detected are not merely reflections of these features of the data.
+The multidimensional aspect of disparity analyses adds a layer of
+technical complexity because the human brain naturally struggles to
+comprehend more than three spatial dimensions. This problem is often
+apparent in how people interpret the results of disparity analyses. For
+example, it is standard for disparity results to use ordination
+techniques to reduce their dimensionality, and for only the first or
+second axes of ordination (e.g. principal components (PCs) one and two)
+to be interpreted and used to draw conclusions. Changes in disparity
+along these axes may show a pattern or gradient, but often do not
+reflect the true changes across all the dimensions of the data
+\href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(Bookstein
+1997, 2015, {[}b{]} 2017, {[}a{]}
+20}\href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1}\href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{7)}{[}CITE
+Vera and Thomas paper{]}. Visual interpretations of multidimensional
+data can also easily be misleading. For example, an obvious
+discontinuity (or overlap) between two groups in a morphospace of PC1
+and PC2, may not be apparent when looking at the same groups in a
+morphospace of PC2 and PC3 etc. or across the whole multidimensional
+space. Furthermore, multidimensional spaces are not necessary Euclidean.
+Such non-Euclidean spaces often have non-intuitive properties, for
+example straight lines in some dimensions are not straight in
+non-Euclidean space and, even less intuitively, the distance between
+points A and B might not be equal to the distance between points B and A
+\href{https://paperpile.com/c/sTGYvp/SJbC}{(Gerber 2014)}.
 
-Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
+3.2. To ordinate or not to ordinate; that is the (multidimensional)
+question?
 
-\section{Use the appropriate methods} 
-In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
-Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
-The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
-Unfortunately, we can only perceive a maximum of three dimensions at the same time.
-Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).
+Reducing the dimensionality of a disparity dataset can be tricky.
+Ordination is advantageous for plotting and visualising the data, and
+can reveal properties of the morphospace not captured by disparity
+metrics (see Metrics section {[}metrics{]}). Additionally, after
+ordinating the data it is possible to further reduce the number of
+dimensions by only selecting a certain number of axes, for example by
+selecting only the axes that describe up to 95\% of the variation. In
+the case of geometric morphometric data, ordination is particularly
+useful as it conserves the mathematical properties of the data while
+efficiently reducing the dimensions
+\href{https://paperpile.com/c/sTGYvp/oFiP}{(Legendre and Legendre
+2012)}. This has clear advantages for interpreting the results. For
+example, the axes will represent gradients of biological variation (e.g.
+elongation and flattening of the beak
+\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}).
 
-\subsection{Visualising the multidimensional problem}
-\label{visualisation}
-It is important to remember that visual interpretations of multidimensional data can easily be misleading.
-For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
-Conversely, when two groups do not overlap on a two or three dimensional plot, this does not mean they never overlap in the multidimensional space.
-Furthermore, multidimensional spaces are not necessary Euclidean.
-Such non-Euclidean spaces often have non-intuitive properties, for example straight lines in some dimensions are not straight in the space and, even less intuitively, the distance between points A and B might not be equal to the distance between points B and A [CITE Sylvain].
+However, various caveats can arise from such transformations of the
+data. For example, in the case of an ordination in a geometric
+morphometric context, not using all the axes from the ordination can
+lead to misinterpretation of the results. Furthermore, interpreting
+biological variation along the axes is always a \emph{post-hoc}
+procedure and may have little relation to the disparity question, for
+example, if the first few ordination axes represent the elongation of
+the beak in birds, but the question is about wing disparity.
 
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.5\textwidth]{Figures/dimensionsOverlap.pdf}
-\caption{\small{Two groups that overlap in some dimensions do not necessary overlap in all dimensions. 
-Conversely, two groups that do not overlap in some dimensions can overlap in other dimensions.
-In this example for the groups A and B to truly overlap, they must overlap in all three dimensions displayed here.
-Although these groups overlap in the two first dimensions (panel 1), they do not overlap in dimensions 1 and 3 thus do not overlap in the multidimensional space at all.
-The inverse is of course true: if two groups don't overlap in two dimensions (D and C in panel 1), they do not overlap in the multidimensional space, even if they overlap in any other combination of dimensions (dimensions 1 and 3 in panel 2).}}
-\label{Fig:RF_results_best}
-\end{figure}
+In many cases, ordination might not be necessary. For example, if the
+disparity metric being used relies on all the data (see Metrics section
+{[}metrics{]}) it is not necessary to calculate it on ordinated data
+{[}e.g.\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
+2015)}{]}. Additionally, in some cases, reducing the dimensionality of a
+dataset can render its interpretation more problematic. For example,
+when the analysed data is non-Euclidean (e.g. discrete morphological
+characters), ordination can be difficult, and the resulting
+non-Euclidean ordinated space can be hard to interpret
+\href{https://paperpile.com/c/sTGYvp/SJbC}{(Gerber 2014)}. This is
+problematic when comparing the position of groups in the
+multidimensional space, as the measured distances might not be
+comparable. Finally, the \emph{post-hoc} interpretation of the gradient
+of variation on the ordination axes can be impossible or biologically
+meaningless. Although some gradients are easy to detect or interpret
+(e.g. the elongation of the beak along the first axis; cooney paper),
+some are not. For example, with discrete morphological data, a gradient
+between the species that have many characters in state 1 and the ones
+that have more in state 0 has no biological meaning. Because of these
+caveats, we strongly recommend that data for multidimensional analyses
+is not automatically ordinated, and that careful consideration is given
+to whether the question can be answered without ordination.
 
-It is therefore vital to be clear about what is represented in multidimensional space visualisations.
-In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a pattern if all the key properties of the space are not present in the graph.
-We suggest visualisations should include the following:
-(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation in the data present in the plot or only a small fraction?);
-(2) the scale of the axes (ideally, if the space is orthogonal, it is important to keep the same scale across the axes);
-and (3) the error associated with each point, whether this is in terms of measurement error (on ordination plots for example) or some other measure of standard variation (in further analysis, e.g.
-disparity-through-time plots).
-Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading (see box \ref{box_ordination}).
+\includegraphics[width=6.5in,height=5.7638in]{Pictures/10000201000002730000022C18503E8C16A21886.png}
 
-\subsection{[Box]To ordinate or not to ordinate; that is the (multidimensional) question?}
-\label{box_ordination}
-Reducing the dimensionality of data through ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
-Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g.
-selecting only the axes that describe up to 95\% of the variation).
-In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions.
-This has clear advantages for interpreting the results.
-For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak REF).
+\emph{Figure methods: Different ways to define and visualise a
+morphospace: A.0: The morphospace is the collected trait space (with
+some potential modification - see data section above): this can be
+visualised using bi-plots (or correlation plots - B.3) between all the
+traits but becomes hard to interpret after a few traits. Another
+approach is to directly ordinate the trait matrix (with a PCA - A.1 -
+e.g. in }\href{https://paperpile.com/c/sTGYvp/ZQS9}{\emph{(Zelditch,
+Swiderski, and David Sheets 2012)}}\emph{- or a PCO - A.2; e.g. in
+Brusatte et al. 2008). This results in a k (\textless{}m) space that can
+also be hard to visualise even though the first axis hold most of the
+variation (B.1). Another approach is to create a distance matrix (A.3;
+e.g. in Close et al., 2015) this matrix can then be ordinated as well
+for further analysis (A.2 and used as the ordinated morphospaces
+described above) or ``flattened'' to a manageable number of dimensions
+(2 or 3 - B.2).}
 
-However, some caveats can arise from such transformations of the data.
-For example, in the case of an ordination in a geometric morphometric data context, not using all the axes from the ordination (e.g. only the first few) can lead to misinterpretation of the results.
-Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and might have little relation to the disparity question (e.g.
-if the first few axes represent the elongation of the beak, but the question is about wing disparity).
+\protect\hypertarget{anchor-1}{}{}3.3. Disparity metrics
 
-In many cases, ordination might not be necessary.
-For example, if the disparity metric being used relies on all the data (see Metrics section \ref{metrics}) it is not necessary to calculate it on ordinated data [e.g. CITE Close].
-Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
-For example, when the analysed data is non-Euclidean (e.g.
-discrete morphological characters), the ordination can be problematic ([cite Melanie's future paper here]), and the resulting non-Euclidean ordinated space can be difficult to interpret ([probably cite Sylvain's here]).
-This can be problematic when comparing the position of groups in the multidimensional space, as the measured distances might not be comparable.
-Finally, the \textit{post-hoc} interpretation of the gradient of variation on the ordination axes can be impossible or biologically meaningless.
-Although some gradients might be easy to detect or interpret (e.g.
-the elongation of the beak along the first axis), they are often hard to interpret in terms of morphological gradients or simply non-interpretable (e.g.
-with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0 has no biological meaning).
+Because disparity is multidimensional, a large component of any
+disparity analysis involves considering how to extract a meaningful (and
+interpretable) summary of disparity, i.e. choosing which disparity
+metric (or index)\href{https://paperpile.com/c/sTGYvp/vTHS}{(Melanie J.
+Hopkins and Gerber 2017)} to use. As with any summary metric, and
+because of this reduction in dimensionality, disparity metrics will only
+reflect some aspects of the morphospace, never its whole complexity. It
+is therefore often beneficial to use more than one metric to summarise
+different aspects of interest. The choice of disparity metric is
+essential to the disparity analysis and should always be driven by the
+question at hand.
 
-Because of these points, we strongly recommend that data for multidimensional analyses is not automatically ordinated, and that careful consideration is given to whether the question can be answered without ordination.
-\subsection{[Box] end}
+When considering only one dimension, disparity metrics can be used to
+reflect the ``size'' of the distribution (e.g. the range, quantiles or
+variance), or the most likely value (i.e. the central tendency, mean,
+median or mode). Among these metrics, some will have more attractive
+properties than others, such as sensitivity to outliers -- range, mean
+and mode are highly sensitive whereas quantiles, variance and median are
+less so -- and will thus be more or less appropriate for different
+questions. For example, if the question refers to ``size'' of a group in
+the morphospace (e.g. does group A occupy as much space as group B?),
+metrics related to the spread of the group in the morphospace are the
+most appropriate (e.g. volume
+\href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, distance
+from the centroid
+\href{https://paperpile.com/c/sTGYvp/vTHS+yyNa}{(Melanie J. Hopkins and
+Gerber 2017; Finlay and Cooper 2015)}, variance and range
+\href{https://paperpile.com/c/sTGYvp/tGyd}{(S. L. Brusatte et al.
+2008)}). Conversely, if the question refers to the ``position'' of a
+group in the morphospace (e.g. does group A occupy the same space as
+group B?), metrics related to the distance between the elements within a
+group and a fixed point in the morphospace are most appropriate (e.g.
+????). Finally, if the question relates to the density of elements in
+some regions of the morphospace (e.g. is group A more dense than group
+B?), metrics related to the pairwise distances between elements will be
+most appropriate (e.g. nearest neighbour distance, pairwise distances,
+etc. \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}).
 
-\subsection{Phylogeny}
-For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
-For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
-and therefore will not covered further in this review.
+In addition to considering what properties of disparity the metrics
+should capture, it is important to also consider the mathematical
+properties of the metrics and their associated caveats
+\href{https://paperpile.com/c/sTGYvp/nFf7+ROH8}{(Wills 2001; Ciampaglio,
+Kemp, and McShea 2001)}. For example, measuring the full sum of the
+variance of each dimension of the space does not require one to add the
+covariance between the axes in a ordinated space using a principal
+components analysis (PCA). This is not true however, in other
+mathematical spaces or when not all dimensions or elements are
+considered, even in a PCA. More simply, when looking at a group within
+morphospace, the sum of variances is not mathematically equal to the sum
+of the variance of each dimension in the space. Further,
+multidimensional space has some counter-intuitive properties that need
+to be considered such as the ``curse of dimensionality''
+\href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}.
+Product-based metrics used as proxies of volumes (e.g. product of
+ranges, hypervolume, hypercube, etc.) will tend towards zero fairly
+quickly for spaces with even a modest number of dimensions
+{[}\href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}; Donoho
+2000{]}. Some other types of metrics are also extremely sensitive to
+outliers and can be easily biased by sample size, for example range
+\href{https://paperpile.com/c/sTGYvp/aSSL}{(Butler et al. 2012)} or
+convex hull based metrics
+\href{https://paperpile.com/c/sTGYvp/aSSL+PwyQ}{(Butler et al. 2012;
+Jackson et al. 2011)}.
 
-\subsubsection{Ancestral state estimation}
-When looking at disparity in the past (e.g. disparity-through-time) ancestral state estimations are a common approach.
-Ancestral state estimation can be performed in two ways:
-\begin{enumerate}
-\item pre-ordination: the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original dataset[CITE].
-\item post-ordination: the estimation is done after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
-\end{enumerate}
-Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e.
-the relationship between the points not estimated) and implies longer computational times but, once the morphospace is defined, it will remain constant (i.e.
-its properties will not change).
-Post ordination ancestral state estimation will not change the ordinated space's geometry and will be faster to compute, however, it will change the morphospace's properties such as the number of parameters (i.e.
-elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper to be?].
+\protect\hypertarget{anchor-2}{}{}3.4. Testing disparity hypotheses
 
-More importantly, however, one needs to remember that using such ancestral state estimation methods comes with several caveats.
-First and foremost, although ancestral state estimations are valid methods in macroevolution, their results are highly dependent on the data and chosen methods.
-Secondly, more specific to disparity analysis, when using post-transformation ancestral state estimates, the morphospace can be saturated leading pairwise distance metrics to tend to zero.
-In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points for statistical reasons.
-In fact, these extra points are not independent and can also generate side effects that can be problematic, especially when testing the effect of extinction events on disparity.
+As mentioned above, disparity analysis should always be used in the
+context of a specific question. Once disparity metrics have been
+calculated, the associated question must be tested in an appropriate
+statistical framework. The multidimensional statistical toolkit for
+ecology and evolution has been greatly expanded in recent years
+\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)} but
+is not routinely applied to disparity analyses. Instead, disparity
+hypothesis testing is mostly confined to a small set of well-established
+methods. One commonly used test is the non-parametric permutation
+analysis of variance \href{https://paperpile.com/c/sTGYvp/3hy2+SC6L}{(M.
+J. Anderson and Walsh 2013; M. J. Anderson 2001)}. This test is an
+analysis of variance (ANOVA) of the pairwise distances between different
+groups. Although statistically valid, this test is sometimes not
+directly related to the hypothesis under test. For example, PERMANOVA
+tests whether two groups share the same variance/covariance in a
+``distance-space''. This is not the same as testing whether the two
+groups overlap in morphospace. Explicitly stating the hypothesis being
+tested can help with understanding which statistical test to apply for a
+specific question.
 
-Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
-In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes.
-Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution.
+It is also important to consider which data to apply a statistical test
+on. For example, in morphological disparity analysis, especially for
+palaeobiological questions, data are often bootstrapped. This has two
+advantages: first, when the disparity metric is unidimensional, e.g. the
+sum of variances, bootstrapping the data generates a distribution of the
+metric that can then be analysed using the vast statistical toolkit
+available for comparing distributions; second, when data are scarce,
+bootstrapping the data allows users to introduce variance, rendering the
+test less sensitive to outliers. However, bootstrapped data are
+pseudoreplicates and thus non-independent. This violates the assumptions
+of most parametric statistical tests. Furthermore, the number of
+bootstrap pseudoreplicates will inevitably increase the Type I error
+rate. These factors are often ignored in disparity analyses.
 
-\subsection{Metrics}
-\label{metrics}
+3.5. Phylogenetic autocorrelation
 
-[INCLUDE WILLS FIGURE SOMWHERE AROUND]
+As with all comparative datasets, the data used in disparity analyses
+are not independent because close relatives will tend to have more
+similar morphologies than more distant relatives
+\href{https://paperpile.com/c/sTGYvp/WXik}{(Harvey and Pagel 1998)}.
+Thus for many disparity analyses phylogenetic relationships should be
+taken into account. Multivariate phylogenetic comparative methods have
+been reviewed recently by
+\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)} and
+therefore will not covered further in this review. However, we do
+briefly draw attention to the fact that phylogenetic corrections of PCA
+are inappropriate (cite Pennell Uyeda paper on phylo PCA).
 
-%TG: add the curse of dimensionality
+\protect\hypertarget{anchor-3}{}{}3.6. Ancestral state estimation in
+disparity-through-time analyses
 
-After building and visualising the morphospace, one can use disparity metrics (or indices) as a one dimensional summary of the space and to test hypotheses.
-As with any summary metric, and because of this reduction in dimensionality, disparity metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
-It is often beneficial to use more than one metric to summarise the different aspects of interest of the morphospace (e.g. its width or spread).
+One final methodological detail to consider applies only to
+disparity-through-time analyses. Often these analyses use ancestral
+state estimation (often mistakenly referred to as ancestral state
+reconstruction) to extract disparity estimates for non-sampled taxa
+and/or nodes of a phylogeny. Ancestral state estimation can be performed
+at two points in the disparity analysis pipeline, either (1)
+pre-ordination, i.e. the estimation is done before transformation of the
+data (e.g. ordination, or distance matrix construction) and is simply
+based on the original data; or (2) post-ordination, i.e. the estimation
+is done after transformation of the data by estimating the ancestral
+states using the transformed matrix (e.g. the ordination scores)
+\href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
 
-The choice of disparity metric is thus essential to the disparity analysis and should always be driven by the question.
-When considering only one dimension, disparity metrics can be used to reflect the ``size'' of the distribution (e.g.
-the range, quantiles or variance), or reflect the most likely value (the central tendency, i.e.
-the mean, median or mode).
-Among these metrics, some will have more preferable properties than others, such as sensitivity to outliers (range, mean or mode are highly sensitive \textit{vs.} quantiles, variance or median are less sensitive) and will thus be more or less appropriate to the different questions and the properties of the distribution.
-For example, if the data collected is likely to contain outliers, but is fairly normally distributed, one should prefer the median to the mean.
+Pre-ordination ancestral state estimation can change the ordinated
+space's geometry -- i.e. the relationship between the points are not yet
+estimated -- and implies longer computational times, but once the
+morphospace is defined its properties will not change. Post-ordination
+ancestral state estimation will not change the ordinated space's
+geometry and is faster to compute, however, it will change the
+morphospace properties such as the number of parameters, i.e. elements
+in the space, or the variance on each axis, which can be problematic for
+statistical tests down the line
+\href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
 
-With multidimensionality, this high availability of unidimensional descriptors comes with equally numerous advantages and caveats and again, the question can help in determining which metric is the most appropriate.
-For example, if the question refers to ``size'' of a group in the morphospace (e.g.
-``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriate (e.g.
-volume [CITE], variance [CITE], range [CITE], distance from the centroid [CITE]).
-Conversely, if the question refers to the ``position'' of a group in the morphospace (e.g.
-``does group A occupy the same space as group B?''), metrics relating to the distance between the elements within a group and a fixed point in the morphospace are most appropriate.
-Finally, one might also be interested in the density of elements in some regions of the morphospace (e.g.
-``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriate (e.g.
-nearest neighbour distance, pairwise distances, etc.).
+More importantly, however, using any ancestral state estimation method
+has several caveats. First and foremost, ancestral state estimations are
+highly dependent on the data and chosen methods. Second, when using
+post-ordination ancestral state estimates, the morphospace can be
+saturated leading pairwise-distance metrics to tend to zero. In general,
+using ancestral state estimation can help with recovering patterns of
+changes in disparity but should not be used simply to generate extra
+data points to increase statistical power. In fact, these extra points
+are not independent and can also generate problematic side effects,
+especially when testing for the effects of mass extinctions on
+disparity.
 
-Additionally to the desired properties of what the metrics should capture, it is important to also keep in mind the mathematical properties of the metrics and their associated caveats.
-For example, measuring the full sum of the variance of each dimension of the space does not require one to add the co-variance between the axes in a ordinated space using a PCA.
-This is not true however in other mathematical spaces or (even in a PCA) when not all dimensions or elements are considered (e.g.
-when only looking at a group within the morphospace, the sum of variances is not mathematically equal to the sum of the variance of each dimension in the space).
-Furthermore, multidimensional space has some counter-intuitive properties that need to be taken into account such as the curse of dimensionality [CITE].
-For example, product-based metrics as proxies of volumes (e.g.
-product of ranges, hypervolume, hypercube, etc.) will tend towards zero fairly quickly for spaces with even a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem and Donoho 2000 Aide-Memoire].
-Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size, for example range or convex hull based metrics.
+4. Disparity analyses for the future
 
-%TG: again, I feel this section is already fairly long, I might include this bit in another paper
+Morphological disparity analyses are a major tool in evolutionary
+biology, but they have many caveats. There, there is no
+``one-size-fits-all'' morphological disparity analysis pipeline. As with
+any multidimensional analysis, there are many nuances in deciding which
+data to use and how to analyse it, stemming from the explicit hypothesis
+being tested. We believe that many of the problems in morphological
+disparity analysis can arise when ``blindly'' applying methodological
+pipelines while losing sight of the biological question being tested. We
+therefore advise keeping the methodology simple and tractable, at least
+in the preliminary analysis. Also, many of the caveats and technical
+problems in disparity analyses have been tackled in other fields, mainly
+ecology. This is unsurprising, as ecological data are also often
+multidimensional
+\href{https://paperpile.com/c/sTGYvp/krNU+60H0}{(Donohue et al. 2013;
+Canter et al. 2018)}. Seeking solutions using these literature is a
+sensible next step for disparity analysis.
 
-% NC: Sounds reasonable. This could be cut down more, but for now I've just edited the grammar and spelling so we can look at the thing as a whole and also get Phil's input.
+While morphological disparity would greatly benefit from advances in
+multidimensional analysis in different fields similarly, the concept of
+a morphospace could benefit other fields. For example, the
+multidimensional analysis in Diaz
+\href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)} looking
+at the patterns of form and function in plants can be thought as a
+morphospace or a life-history-space (an eco-space?); isotopic analyses
+in \href{https://paperpile.com/c/sTGYvp/PwyQ}{(Jackson et al. 2011)} can
+be represented as a isotope-space; ecosystem functioning in (Donohue et
+al. 2013) as an ecosystem-space, etc. These generalisations could also
+be exported for any set of traits (e.g. acousto-spaces for acoustic
+traits?) and even beyond evolutionary biology (e.g. glotto-spaces for
+linguistic traits?).
 
-% Of course some other aspects could be of interest (or more than one) and thus some other metrics would be more appropriated.
-% For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution).
-% Furthermore, one future avenue that could be explored is to go beyond single valued disparity metrics and rather express them in terms of distributions.
-% For example, when the metric is expressed as the median distance from centroids [e.g. in CITE], it could be informative as well to express it as a full distribution.
-% This would allow to both visualise the variation at a glance and use this variation in statistical tests for comparing these distributions.
+Although disparity analyses are now simple to implement
+\href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{(Guillerme
+2018; Adams and Otárola-Castillo 2013; Bouxin 2005; Harmon et al. 2008;
+Lloyd 2016; Navarro 2003; Dixon 2003)}, it is crucial to remember that
+they are multidimensional analyses; and multidimensional analyses are
+complex. In Jurassic Park, Dr Ian Malcolm summarises this problem in an
+elegant way: ``{[}We{]} scientists were so preoccupied with whether or
+not {[}we{]} could that {[}we{]} didn't stop to think if {[}we{]}
+should.'' We believe that morphological analysis in general will greatly
+benefit from emphasising the methodological decisions made
+(\emph{whether we should}), rather than simply using disparity analysis
+because \emph{we can}.
 
-% Best practices:
-% \begin{itemize}
-% \item Facing the amount of different ways to measure disparity and regardless which metric is the most suited to the question, it is crucial to clearly and mathematically define the metric in a disparity paper.
-The definition can be in plain English: e.g.
-``we measure disparity as the sum of the variance of each axis of the ordinated matrix''; or mathematically: e.g.
-``we measured disparity as $\sum\sigma_i$ where $\sigma_i$ is the variance of the each ordinated axis $i$.''.
-% \item It is important to keep in mind that different metrics can lead to the same results even though the described spaces are different (the opposite is \textit{not} true).
-We therefore suggest to always plot some dimensions of the space (see Visualisation section \ref{visualisation}) and try multiple metrics.
-% \end{itemize}
+5. Author contributions
 
-\subsection{Tests}
-The final stage of most disparity analyses is statistical testing of the disparity metrics, to determine whether changes thought time or between groups are statistically significant. 
-Similar to the points above, the statistical test to apply depends on which question is being tackled with disparity analysis, i.e. which hypothesis is being tested.
-It also requires consideration of the data, the ordination method used (or the absence thereof) and the metric used.
+TG, NC and PD proposed this review. All authors edited the manuscript.
 
-For example, one commonly used test in morphological analysis, is the non parametric permutation analysis of variance [NPANOVA or PERMANOVA; CITE Anderson + examples].
-This test and its implementation is perfectly valid to test whether two groups have a different variance/covariance, but is based on the pairwise distances of the group elements.
-This test will ignore any ordination (if the data was ordinated) and is not based on a unidimensional disparity metric.
-In other words, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
-Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the question.
+6. Acknowledgments
 
-Another caveat for applying statistical tests on disparity data is that it is important to consider which data to apply the test on.
-For example, in morphological disparity analysis (especially for palaeobiological questions), data is often bootstrapped.
-This has two useful advantages: first, when the disparity metric is unidimensional (e.g.
-the sum of variances or the average squared pairwise distance), bootstrapping the data generates a distribution of this unidimensional metric that can then be analysed using the vast statistical toolkit available for comparing distributions; second, when the data is scarce (e.g.
-for palaeontological data), bootstrapping the data allows users to introduce statistical variance, rendering the test less sensitive to outliers.
-However, bootstrapping the data also comes with an important caveat: the data analysed are pseudo-replicated and thus non-independent.
-This can violate the assumptions of parametric statistical tests.
-Furthermore, it is important to remember that the number of bootstrap pseudo-replicates will inevitably increase the Type I error.
+The Royal Society. TG funding.
 
-\section{Future directions}
-As mentioned above, there is no ``one-size-fits-all'' morphological disparity analysis pipeline.
-As with any multidimensional analysis, there are many nuances in deciding which data to use and how to analyse it, stemming from the explicit formulation of the disparity hypothesis.
-To understand these nuances, it is important to remember that: (1) morphological disparity analyses are not new and much effort has been made in developing methods; and (2) multidimensional analyses are not restricted to morphological data analysis and solutions to specific problems can be found in literature that is not necessary focused on morphology (especially ecology and mathematics).
+Donoho, D. L. (2000). High-dimensional data analysis: The curses and
+blessings of dimensionality. \emph{AMS math challenges lecture},
+\emph{1}(2000), 32.
 
-We believe that many of the caveats in morphological disparity analysis can arise when ``blindly'' applying methodological pipelines at the expense of losing sight of the  biological question being tested.
-We therefore advise keeping the methodology simple and tractable, at least in the preliminary analysis.
-Also, many of the caveats and technical problems in disparity analyses have been tackled in other fields, mainly ecology.
-This is perhaps unsurprisingly as ecological questions are often closely related to evolutionary questions, and ecological data is also often multidimensional [CITE DONOHUE's papers].
+References
 
-\subsection{How can we adapt our methods?}
-Each of the points outlined above may drive new methodological questions (and often already are).
-Although it is impossible to cover the state-of-the-art exhaustively for each of these points here are several solutions and/or directions for adapting existing methods:
+\href{http://paperpile.com/b/sTGYvp/ZnDd}{Adams, Dean C., and Michael L.
+Collyer. 2018. ``Multivariate Phylogenetic Comparative Methods:
+Evaluations, Comparisons, and Recommendations.''
+}\href{http://paperpile.com/b/sTGYvp/ZnDd}{\emph{Systematic
+Biology}}\href{http://paperpile.com/b/sTGYvp/ZnDd}{ 67 (1): 14--31.}
 
-\begin{itemize}
-    \item{Which data should we use in disparity analyses?}
-    %[One line abstract of Melanie's paper on inapplicable data].
-    Furthermore, some simple methods already exist to account for caveats in data. 
-    For example, to correct for the paucity of the fossil record, one can use methods like rarefaction analysis to take sample size into account [CITE].
-    % NC: This item currently doesn't seem to actually give advice about which data to use...
-    \item{How should we visualise disparity?} % NC: Tried to make the bullets a bit clearer, might need to edit for correctness and consistency with the rest of the document.
-    [One line abstract of Sylvain's paper on morphospaces plottings].
-    One major problem of multidimensional analyses is that highly dimensional spaces are hard (or even impossible) to conceptualise.
-    One solution to this is to use more simple 2D/3D ordinations.
-    For example, using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation [e.g Raup-spaces CITE].
-    \item{Which disparity metrics should we use?}
-    This can be tackled by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combination of properties of interest in the morphospace.
-    For example, when reducing a multidimensional space to one metric, it would be judicious to indicate what this metric represents in the results (e.g.
-in a plot or a table, if the metric used was the sum of variances, labelling it as ``size of the group'' rather than ``disparity'').
-    This is especially advantageous when using multiple disparity metrics (e.g. ``size of the group'' and ``position of the group'').
-    \item{How should we test disparity hypotheses?}
-    Finally, it is possible to use non-parametric methods to test specific disparity questions.
-    For example, using available software and computer power, it is possible to propose specific null models for the expectation of what disparity would be expected to look like under various null models.
-    This has been already successfully used in ecology [CITE Diaz paper] and is currently being implemented in palaeobiology (Mark stuff).
-\end{itemize}
+\href{http://paperpile.com/b/sTGYvp/J2G1}{Adams, Dean C., and Erik
+Otárola-Castillo. 2013. ``Geomorph: Anrpackage for the Collection and
+Analysis of Geometric Morphometric Shape Data.''
+}\href{http://paperpile.com/b/sTGYvp/J2G1}{\emph{Methods in Ecology and
+Evolution / British Ecological
+Society}}\href{http://paperpile.com/b/sTGYvp/J2G1}{ 4 (4): 393--99.}
 
-\subsection{Where do we go from here?}
-Morphological disparity analyses are a major tool in evolutionary biology, posing a more versatile alternative to studying species richness, especially in macroevolutionary studies.
-We believe that morphological disparity analyses, as the analysis of patterns of morphological differences, can be of great use for linking evolutionary and ecological processes (e.g.
-when using morphological disparity as a proxy for ecological niches to test competition-through-time hypotheses).
+\href{http://paperpile.com/b/sTGYvp/SC6L}{Anderson, Marti J. 2001. ``A
+New Method for Non-Parametric Multivariate Analysis of Variance.''
+}\href{http://paperpile.com/b/sTGYvp/SC6L}{\emph{Austral
+Ecology}}\href{http://paperpile.com/b/sTGYvp/SC6L}{ 26 (1): 32--46.}
 
-However, as demonstrated above, such analyses would greatly benefit from more development, and  more care is needed in study design so that the analysis actually reflects the biological question at hand.
-This can not only be done through the points highlighted above but also through linking the observed patterns to an explicit process (for example, the link between an organism's form and function) or by developing null models to potentially explain the observed patterns.
+\href{http://paperpile.com/b/sTGYvp/3hy2}{Anderson, Marti J., and Daniel
+C. I. Walsh. 2013. ``PERMANOVA, ANOSIM, and the Mantel Test in the Face
+of Heterogeneous Dispersions: What Null Hypothesis Are You Testing?''
+}\href{http://paperpile.com/b/sTGYvp/3hy2}{\emph{Ecological
+Monographs}}\href{http://paperpile.com/b/sTGYvp/3hy2}{ 83 (4): 557--74.}
 
-Finally, we encourage to a more cross-disciplinary approach to morphological disparity analyses.
-In fact, as mentioned above, morphological disparity could greatly benefit from advances in multidimensional analysis in different fields (namely ecology) but similarly, the concept of a morphospace could easily be exported to other fields.
-For example, the multidimensional analysis in Diaz [CITE] looking at the patterns of form and function in plants can be thought as a morphospace or a life-history-space (an ecospace?); isotopic analyses in [CITE Andrew's stuff] can be represented as a isotope-space, etc.
-These generalisations could also be exported for any set of traits (e.g.
-acoustospaces for acoustic traits) and even beyond evolutionary biology (e.g.
-glottospaces for linguistic traits) [CITE Katie's review].
+\href{http://paperpile.com/b/sTGYvp/qjj9}{Anderson, Philip S. L., Matt
+Friedman, Martin D. Brazeau, and Emily J. Rayfield. 2011. ``Initial
+Radiation of Jaws Demonstrated Stability despite Faunal and
+Environmental Change.''
+}\href{http://paperpile.com/b/sTGYvp/qjj9}{\emph{Nature}}\href{http://paperpile.com/b/sTGYvp/qjj9}{
+476 (7359): 206--9.}
 
-%Concluding words:
-Although disparity analyses are now simple to implement (CITE packages tools), it is crucial to remember that they are multidimensional analyses; and multidimensional analyses are complex.
-In Jurassic Park, Dr Ian Malcolm summarises this problem in an elegant way: ``[We] scientists were so preoccupied with whether or not [we] could that [we] didn't stop to think if [we] should.''
-We believe that morphological analysis in general will greatly benefit from emphasising the methodological decisions made (\textit{whether we should}), rather than simply using disparity analysis because \textit{we can}.
+\href{http://paperpile.com/b/sTGYvp/Qsl3}{Bellman, R. 1966. ``Dynamic
+Programming.''
+}\href{http://paperpile.com/b/sTGYvp/Qsl3}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/Qsl3}{
+153 (3731): 34--37.}
 
-% NC: I do like this quote.
-The actual quote is: "Your scientists were so preoccupied with whether or not they could that they didn't stop to think if they should." Dr Ian Malcolm.
-It would be magnificent ot get a jurassic park quote into a paper...
+\href{http://paperpile.com/b/sTGYvp/1SD2}{Bookstein, Fred L. 1997.
+}\href{http://paperpile.com/b/sTGYvp/1SD2}{\emph{Morphometric Tools for
+Landmark Data: Geometry and
+Biology}}\href{http://paperpile.com/b/sTGYvp/1SD2}{. Cambridge
+University Press.}
 
-\section{Authors contribution}
-TG, NC and PD proposed the idea of this review. All authors edited the manuscript.
+\href{http://paperpile.com/b/sTGYvp/sN5d}{---------. 2015. ``The
+Relation between Geometric Morphometrics and Functional Morphology, as
+Explored by Procrustes Interpretation of Individual Shape Measures
+Pertinent to Function.''
+}\href{http://paperpile.com/b/sTGYvp/sN5d}{\emph{Anatomical Record
+}}\href{http://paperpile.com/b/sTGYvp/sN5d}{ 298 (1): 314--27.}
 
-\section{Acknowledgments}
-The Royal Society.
+\href{http://paperpile.com/b/sTGYvp/o4w7}{---------. 2017a. ``A Newly
+Noticed Formula Enforces Fundamental Limits on Geometric Morphometric
+Analyses.''
+}\href{http://paperpile.com/b/sTGYvp/o4w7}{\emph{Evolutionary
+Biology}}\href{http://paperpile.com/b/sTGYvp/o4w7}{ 44 (4): 522--41.}
 
-\bibliographystyle{sysbio}
-\bibliography{References}
+\href{http://paperpile.com/b/sTGYvp/xaUx}{---------. 2017b. ``A Method
+of Factor Analysis for Shape Coordinates.''
+}\href{http://paperpile.com/b/sTGYvp/xaUx}{\emph{American Journal of
+Physical Anthropology}}\href{http://paperpile.com/b/sTGYvp/xaUx}{ 164
+(2): 221--45.}
 
-\end{document}
+\href{http://paperpile.com/b/sTGYvp/9JdS}{Bouxin, Guy. 2005. ``Ginkgo, a
+Multivariate Analysis Package.''
+}\href{http://paperpile.com/b/sTGYvp/9JdS}{\emph{Journal of Vegetation
+Science: Official Organ of the International Association for Vegetation
+Science}}\href{http://paperpile.com/b/sTGYvp/9JdS}{ 16 (3): 355--59.}
+
+\href{http://paperpile.com/b/sTGYvp/Yrbg}{Brazeau, Martin D., Thomas
+Guillerme, and Martin R. Smith. 2017. ``Morphological Phylogenetic
+Analysis with Inapplicable Data.''
+https://doi.org/}\href{http://dx.doi.org/10.1101/209775}{10.1101/209775}\href{http://paperpile.com/b/sTGYvp/Yrbg}{.}
+
+\href{http://paperpile.com/b/sTGYvp/CiPy}{Briggs, D. E., R. A. Fortey,
+and M. A. Wills. 1992. ``Morphological Disparity in the Cambrian.''
+}\href{http://paperpile.com/b/sTGYvp/CiPy}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/CiPy}{
+256 (5064): 1670--73.}
+
+\href{http://paperpile.com/b/sTGYvp/tGyd}{Brusatte, S. L., M. J. Benton,
+M. Ruta, and G. T. Lloyd. 2008. ``The First 50 Myr of Dinosaur
+Evolution: Macroevolutionary Pattern and Morphological Disparity.''
+}\href{http://paperpile.com/b/sTGYvp/tGyd}{\emph{Biology
+Letters}}\href{http://paperpile.com/b/sTGYvp/tGyd}{ 4 (6): 733--36.}
+
+\href{http://paperpile.com/b/sTGYvp/EeC8}{Brusatte, Stephen L., Michael
+J. Benton, Marcello Ruta, and Graeme T. Lloyd. 2008. ``Superiority,
+Competition, and Opportunism in the Evolutionary Radiation of
+Dinosaurs.''
+}\href{http://paperpile.com/b/sTGYvp/EeC8}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/EeC8}{
+321 (5895): 1485--88.}
+
+\href{http://paperpile.com/b/sTGYvp/aSSL}{Butler, Richard J., Stephen L.
+Brusatte, Brian Andres, and Roger B. J. Benson. 2012. ``How Do
+Geological Sampling Biases Affect Studies of Morphological Evolution in
+Deep Time? A Case Study of Pterosaur (Reptilia: Archosauria)
+Disparity.'' }\href{http://paperpile.com/b/sTGYvp/aSSL}{\emph{Evolution;
+International Journal of Organic
+Evolution}}\href{http://paperpile.com/b/sTGYvp/aSSL}{ 66 (1): 147--62.}
+
+\href{http://paperpile.com/b/sTGYvp/60H0}{Canter, Erin J., Catalina
+Cuellar-Gempeler, Abigail I. Pastore, Thomas E. Miller, and Olivia U.
+Mason. 2018. ``Predator Identity More than Predator Richness Structures
+Aquatic Microbial Assemblages in Sarracenia Purpurea Leaves.''
+}\href{http://paperpile.com/b/sTGYvp/60H0}{\emph{Ecology}}\href{http://paperpile.com/b/sTGYvp/60H0}{
+99 (3): 652--60.}
+
+\href{http://paperpile.com/b/sTGYvp/ROH8}{Ciampaglio, Charles N.,
+Matthieu Kemp, and Daniel W. McShea. 2001. ``Detecting Changes in
+Morphospace Occupation Patterns in the Fossil Record: Characterization
+and Analysis of Measures of Disparity.''
+}\href{http://paperpile.com/b/sTGYvp/ROH8}{\emph{Paleobiology}}\href{http://paperpile.com/b/sTGYvp/ROH8}{
+27 (4): 695--715.}
+
+\href{http://paperpile.com/b/sTGYvp/khc9}{Cisneros, Juan Carlos, and
+Marcello Ruta. 2010. ``Morphological Diversity and Biogeography of
+Procolophonids (Amniota: Parareptilia).''
+}\href{http://paperpile.com/b/sTGYvp/khc9}{\emph{Journal of Systematic
+Palaeontology}}\href{http://paperpile.com/b/sTGYvp/khc9}{ 8 (4):
+607--25.}
+
+\href{http://paperpile.com/b/sTGYvp/PbSx}{Close, Roger A., Matt
+Friedman, Graeme T. Lloyd, and Roger B. J. Benson. 2015. ``Evidence for
+a Mid-Jurassic Adaptive Radiation in Mammals.''
+}\href{http://paperpile.com/b/sTGYvp/PbSx}{\emph{Current Biology:
+CB}}\href{http://paperpile.com/b/sTGYvp/PbSx}{ 25 (16): 2137--42.}
+
+\href{http://paperpile.com/b/sTGYvp/RjqE}{Cooney, Christopher R., Jen A.
+Bright, Elliot J. R. Capp, Angela M. Chira, Emma C. Hughes, Christopher
+J. A. Moody, Lara O. Nouri, Zoë K. Varley, and Gavin H. Thomas. 2017.
+``Mega-Evolutionary Dynamics of the Adaptive Radiation of Birds.''
+}\href{http://paperpile.com/b/sTGYvp/RjqE}{\emph{Nature}}\href{http://paperpile.com/b/sTGYvp/RjqE}{
+542 (7641): 344--47.}
+
+\href{http://paperpile.com/b/sTGYvp/Qrba}{Dávalos, Liliana M., Paúl M.
+Velazco, Omar M. Warsi, Peter D. Smits, and Nancy B. Simmons. 2014.
+``Integrating Incomplete Fossils by Isolating Conflicting Signal in
+Saturated and Non-Independent Morphological Characters.''
+}\href{http://paperpile.com/b/sTGYvp/Qrba}{\emph{Systematic
+Biology}}\href{http://paperpile.com/b/sTGYvp/Qrba}{ 63 (4): 582--600.}
+
+\href{http://paperpile.com/b/sTGYvp/0y4V}{Deline, Bradley, Jennifer M.
+Greenwood, James W. Clark, Mark N. Puttick, Kevin J. Peterson, and
+Philip C. J. Donoghue. 2018. ``Evolution of Metazoan Morphological
+Disparity.''
+}\href{http://paperpile.com/b/sTGYvp/0y4V}{\emph{Proceedings of the
+National Academy of Sciences of the United States of
+America}}\href{http://paperpile.com/b/sTGYvp/0y4V}{ 115 (38):
+E8909--18.}
+
+\href{http://paperpile.com/b/sTGYvp/47fI}{Díaz, Sandra, Jens Kattge,
+Johannes H. C. Cornelissen, Ian J. Wright, Sandra Lavorel, Stéphane
+Dray, Björn Reu, et al. 2016. ``The Global Spectrum of Plant Form and
+Function.''
+}\href{http://paperpile.com/b/sTGYvp/47fI}{\emph{Nature}}\href{http://paperpile.com/b/sTGYvp/47fI}{
+529 (7585): 167--71.}
+
+\href{http://paperpile.com/b/sTGYvp/2KmX}{Dixon, Philip. 2003. ``VEGAN,
+a Package of R Functions for Community Ecology.''
+}\href{http://paperpile.com/b/sTGYvp/2KmX}{\emph{Journal of Vegetation
+Science: Official Organ of the International Association for Vegetation
+Science}}\href{http://paperpile.com/b/sTGYvp/2KmX}{ 14 (6): 927.}
+
+\href{http://paperpile.com/b/sTGYvp/krNU}{Donohue, Ian, Owen L. Petchey,
+José M. Montoya, Andrew L. Jackson, Luke McNally, Mafalda Viana, Kevin
+Healy, Miguel Lurgi, Nessa E. O'Connor, and Mark C. Emmerson. 2013. ``On
+the Dimensionality of Ecological Stability.''
+}\href{http://paperpile.com/b/sTGYvp/krNU}{\emph{Ecology
+Letters}}\href{http://paperpile.com/b/sTGYvp/krNU}{ 16 (4): 421--29.}
+
+\href{http://paperpile.com/b/sTGYvp/yyNa}{Finlay, Sive, and Natalie
+Cooper. 2015. ``Morphological Diversity in Tenrecs (Afrosoricida,
+Tenrecidae): Comparing Tenrec Skull Diversity to Their Closest
+Relatives.''
+}\href{http://paperpile.com/b/sTGYvp/yyNa}{\emph{PeerJ}}\href{http://paperpile.com/b/sTGYvp/yyNa}{
+3 (April): e927.}
+
+\href{http://paperpile.com/b/sTGYvp/oenu}{Foote, M., and S. J. Gould.
+1992. ``Cambrian and Recent Morphological Disparity.''
+}\href{http://paperpile.com/b/sTGYvp/oenu}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/oenu}{
+258 (5089): 1816.}
+
+\href{http://paperpile.com/b/sTGYvp/fTJ3}{Foote, Mike. 1995.
+``Morphological Diversification of Paleozoic Crinoids.''
+}\href{http://paperpile.com/b/sTGYvp/fTJ3}{\emph{Paleobiology}}\href{http://paperpile.com/b/sTGYvp/fTJ3}{
+21 (03): 273--99.}
+
+\href{http://paperpile.com/b/sTGYvp/yqPw}{---------. 1997. ``THE
+EVOLUTION OF MORPHOLOGICAL DIVERSITY.''
+}\href{http://paperpile.com/b/sTGYvp/yqPw}{\emph{Annual Review of
+Ecology and Systematics}}\href{http://paperpile.com/b/sTGYvp/yqPw}{ 28
+(1): 129--52.}
+
+\href{http://paperpile.com/b/sTGYvp/2tbJ}{Fortey, R. A., D. E. G.
+Briggs, and M. A. Wills. 1996. ``The Cambrian Evolutionary `explosion':
+Decoupling Cladogenesis from Morphological Disparity.''
+}\href{http://paperpile.com/b/sTGYvp/2tbJ}{\emph{Biological Journal of
+the Linnean Society. Linnean Society of
+London}}\href{http://paperpile.com/b/sTGYvp/2tbJ}{ 57 (1): 13--33.}
+
+\href{http://paperpile.com/b/sTGYvp/EETc}{Friedman, Matt. 2010.
+``Explosive Morphological Diversification of Spiny-Finned Teleost Fishes
+in the Aftermath of the End-Cretaceous Extinction.''
+}\href{http://paperpile.com/b/sTGYvp/EETc}{\emph{Proceedings. Biological
+Sciences / The Royal Society}}\href{http://paperpile.com/b/sTGYvp/EETc}{
+277 (1688): 1675--83.}
+
+\href{http://paperpile.com/b/sTGYvp/SJbC}{Gerber, Sylvain. 2014. ``Not
+All Roads Can Be Taken: Development Induces Anisotropic Accessibility in
+Morphospace.''
+}\href{http://paperpile.com/b/sTGYvp/SJbC}{\emph{Evolution \&
+Development}}\href{http://paperpile.com/b/sTGYvp/SJbC}{ 16 (6):
+373--81.}
+
+\href{http://paperpile.com/b/sTGYvp/QVvv}{---------. 2017. ``The
+Geometry of Morphospaces: Lessons from the Classic Raup Shell Coiling
+Model.'' }\href{http://paperpile.com/b/sTGYvp/QVvv}{\emph{Biological
+Reviews of the Cambridge Philosophical
+Society}}\href{http://paperpile.com/b/sTGYvp/QVvv}{ 92 (2): 1142--55.}
+
+\href{http://paperpile.com/b/sTGYvp/CidX}{Gould, Stephen Jay. 1991.
+``The Disparity of the Burgess Shale Arthropod Fauna and the Limits of
+Cladistic Analysis: Why We Must Strive to Quantify Morphospace.''
+}\href{http://paperpile.com/b/sTGYvp/CidX}{\emph{Paleobiology}}\href{http://paperpile.com/b/sTGYvp/CidX}{
+17 (04): 411--23.}
+
+\href{http://paperpile.com/b/sTGYvp/Uns3}{---------. 2000.
+}\href{http://paperpile.com/b/sTGYvp/Uns3}{\emph{Wonderful Life: The
+Burgess Shale and the Nature of
+History}}\href{http://paperpile.com/b/sTGYvp/Uns3}{. Random House.}
+
+\href{http://paperpile.com/b/sTGYvp/xDqf}{Guillerme, Thomas. 2018.
+``dispRity : A Modular R Package for Measuring Disparity.''
+}\href{http://paperpile.com/b/sTGYvp/xDqf}{\emph{Methods in Ecology and
+Evolution / British Ecological
+Society}}\href{http://paperpile.com/b/sTGYvp/xDqf}{ 9 (7): 1755--63.}
+
+\href{http://paperpile.com/b/sTGYvp/ekU4}{Guillerme, Thomas, and Natalie
+Cooper. 2018. ``Time for a Rethink: Time Sub-Sampling Methods in
+Disparity-through-Time Analyses.''
+}\href{http://paperpile.com/b/sTGYvp/ekU4}{\emph{Palaeontology}}\href{http://paperpile.com/b/sTGYvp/ekU4}{
+61 (4): 481--93.}
+
+\href{http://paperpile.com/b/sTGYvp/9Zoi}{Harmon, Luke J., Jason T.
+Weir, Chad D. Brock, Richard E. Glor, and Wendell Challenger. 2008.
+``GEIGER: Investigating Evolutionary Radiations.''
+}\href{http://paperpile.com/b/sTGYvp/9Zoi}{\emph{Bioinformatics
+}}\href{http://paperpile.com/b/sTGYvp/9Zoi}{ 24 (1): 129--31.}
+
+\href{http://paperpile.com/b/sTGYvp/WXik}{Harvey, Paul H., and Mark D.
+Pagel. 1998. }\href{http://paperpile.com/b/sTGYvp/WXik}{\emph{The
+Comparative Method in Evolutionary
+Biology}}\href{http://paperpile.com/b/sTGYvp/WXik}{. Oxford University
+Press, USA.}
+
+\href{http://paperpile.com/b/sTGYvp/xLdm}{Hopkins, Melanie J. 2017.
+``How Well Does a Part Represent the Whole? A Comparison of Cranidial
+Shape Evolution with Exoskeletal Character Evolution in the Trilobite
+Family Pterocephaliidae.''
+}\href{http://paperpile.com/b/sTGYvp/xLdm}{\emph{Palaeontology}}\href{http://paperpile.com/b/sTGYvp/xLdm}{
+60 (3): 309--18.}
+
+\href{http://paperpile.com/b/sTGYvp/vTHS}{Hopkins, Melanie J., and
+Sylvain Gerber. 2017. ``Morphological Disparity.'' In
+}\href{http://paperpile.com/b/sTGYvp/vTHS}{\emph{Evolutionary
+Developmental Biology}}\href{http://paperpile.com/b/sTGYvp/vTHS}{,
+1--12.}
+
+\href{http://paperpile.com/b/sTGYvp/hea5}{Hopkins, M. J. 2013.
+``Decoupling of Taxonomic Diversity and Morphological Disparity during
+Decline of the Cambrian Trilobite Family Pterocephaliidae.''
+}\href{http://paperpile.com/b/sTGYvp/hea5}{\emph{Journal of Evolutionary
+Biology}}\href{http://paperpile.com/b/sTGYvp/hea5}{ 26 (8): 1665--76.}
+
+\href{http://paperpile.com/b/sTGYvp/xxh5}{Hughes, Martin, Sylvain
+Gerber, and Matthew Albion Wills. 2013. ``Clades Reach Highest
+Morphological Disparity Early in Their Evolution.''
+}\href{http://paperpile.com/b/sTGYvp/xxh5}{\emph{Proceedings of the
+National Academy of Sciences of the United States of
+America}}\href{http://paperpile.com/b/sTGYvp/xxh5}{ 110 (34):
+13875--79.}
+
+\href{http://paperpile.com/b/sTGYvp/PwyQ}{Jackson, Andrew L., Richard
+Inger, Andrew C. Parnell, and Stuart Bearhop. 2011. ``Comparing Isotopic
+Niche Widths among and within Communities: SIBER - Stable Isotope
+Bayesian Ellipses in R.''
+}\href{http://paperpile.com/b/sTGYvp/PwyQ}{\emph{The Journal of Animal
+Ecology}}\href{http://paperpile.com/b/sTGYvp/PwyQ}{ 80 (3): 595--602.}
+
+\href{http://paperpile.com/b/sTGYvp/oFiP}{Legendre, P., and Loic F. J.
+Legendre. 2012.
+}\href{http://paperpile.com/b/sTGYvp/oFiP}{\emph{Numerical
+Ecology}}\href{http://paperpile.com/b/sTGYvp/oFiP}{. Elsevier.}
+
+\href{http://paperpile.com/b/sTGYvp/bCsU}{Lloyd, Graeme T. 2016.
+``Estimating Morphological Diversity and Tempo with Discrete
+Character-Taxon Matrices: Implementation, Challenges, Progress, and
+Future Directions.''
+}\href{http://paperpile.com/b/sTGYvp/bCsU}{\emph{Biological Journal of
+the Linnean Society. Linnean Society of
+London}}\href{http://paperpile.com/b/sTGYvp/bCsU}{ 118 (1): 131--51.}
+
+\href{http://paperpile.com/b/sTGYvp/53SJ}{---------. 2018. ``Journeys
+through Discrete-Character Morphospace: Synthesizing Phylogeny, Tempo,
+and Disparity.''
+}\href{http://paperpile.com/b/sTGYvp/53SJ}{\emph{Palaeontology}}\href{http://paperpile.com/b/sTGYvp/53SJ}{
+61 (5): 637--45.}
+
+\href{http://paperpile.com/b/sTGYvp/TtGs}{Morris, S. C. 1989. ``Burgess
+Shale Faunas and the Cambrian Explosion.''
+}\href{http://paperpile.com/b/sTGYvp/TtGs}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/TtGs}{
+246 (4928): 339--46.}
+
+\href{http://paperpile.com/b/sTGYvp/aVVj}{Moyne, Sébastien, and Pascal
+Neige. 2007. ``The Space-Time Relationship of Taxonomic Diversity and
+Morphological Disparity in the Middle Jurassic Ammonite Radiation.''
+}\href{http://paperpile.com/b/sTGYvp/aVVj}{\emph{Palaeogeography,
+Palaeoclimatology,
+Palaeoecology}}\href{http://paperpile.com/b/sTGYvp/aVVj}{ 248 (1-2):
+82--95.}
+
+\href{http://paperpile.com/b/sTGYvp/EmTR}{Navarro, Nicolas. 2003. ``MDA:
+A MATLAB-Based Program for Morphospace-Disparity Analysis.''
+}\href{http://paperpile.com/b/sTGYvp/EmTR}{\emph{Computers \&
+Geosciences}}\href{http://paperpile.com/b/sTGYvp/EmTR}{ 29 (5):
+655--64.}
+
+\href{http://paperpile.com/b/sTGYvp/yO2t}{Palci, Alessandro, and Michael
+S. Y. Lee. 2018. ``Geometric Morphometrics, Homology and Cladistics:
+Review and Recommendations.''
+}\href{http://paperpile.com/b/sTGYvp/yO2t}{\emph{Cladistics: The
+International Journal of the Willi Hennig
+Society}}\href{http://paperpile.com/b/sTGYvp/yO2t}{.
+https://doi.org/}\href{http://dx.doi.org/10.1111/cla.12340}{10.1111/cla.12340}\href{http://paperpile.com/b/sTGYvp/yO2t}{.}
+
+\href{http://paperpile.com/b/sTGYvp/tSIy}{Pierce, Stephanie E., Kenneth
+D. Angielczyk, and Emily J. Rayfield. 2008. ``Patterns of Morphospace
+Occupation and Mechanical Performance in Extant Crocodilian Skulls: A
+Combined Geometric Morphometric and Finite Element Modeling Approach.''
+}\href{http://paperpile.com/b/sTGYvp/tSIy}{\emph{Journal of
+Morphology}}\href{http://paperpile.com/b/sTGYvp/tSIy}{ 269 (7):
+840--64.}
+
+\href{http://paperpile.com/b/sTGYvp/I0Ic}{Raup, D. M. 1961. ``THE
+GEOMETRY OF COILING IN GASTROPODS.''
+}\href{http://paperpile.com/b/sTGYvp/I0Ic}{\emph{Proceedings of the
+National Academy of Sciences of the United States of
+America}}\href{http://paperpile.com/b/sTGYvp/I0Ic}{ 47 (4): 602--9.}
+
+\href{http://paperpile.com/b/sTGYvp/geAO}{Ruta, Marcello, Kenneth D.
+Angielczyk, Jörg Fröbisch, and Michael J. Benton. 2013. ``Decoupling of
+Morphological Disparity and Taxic Diversity during the Adaptive
+Radiation of Anomodont Therapsids.''
+}\href{http://paperpile.com/b/sTGYvp/geAO}{\emph{Proceedings. Biological
+Sciences / The Royal Society}}\href{http://paperpile.com/b/sTGYvp/geAO}{
+280 (1768): 20131071.}
+
+\href{http://paperpile.com/b/sTGYvp/ZEDR}{Spriggs, Elizabeth L., Samuel
+B. Schmerler, Erika J. Edwards, and Michael J. Donoghue. 2018. ``Leaf
+Form Evolution in Viburnum Parallels Variation within Individual
+Plants.'' }\href{http://paperpile.com/b/sTGYvp/ZEDR}{\emph{The American
+Naturalist}}\href{http://paperpile.com/b/sTGYvp/ZEDR}{ 191 (2):
+235--49.}
+
+\href{http://paperpile.com/b/sTGYvp/6tNm}{Valentine, J. W. 1986.
+``Fossil Record of the Origin of Baupläne and Its Implications.'' In
+}\href{http://paperpile.com/b/sTGYvp/6tNm}{\emph{Patterns and Processes
+in the History of Life}}\href{http://paperpile.com/b/sTGYvp/6tNm}{,
+209--22.}
+
+\href{http://paperpile.com/b/sTGYvp/Ejzr}{Wainwright, Peter C., Michael
+E. Alfaro, Daniel I. Bolnick, and C. Darrin Hulsey. 2005. ``Many-to-One
+Mapping of Form to Function: A General Principle in Organismal Design?''
+}\href{http://paperpile.com/b/sTGYvp/Ejzr}{\emph{Integrative and
+Comparative Biology}}\href{http://paperpile.com/b/sTGYvp/Ejzr}{ 45 (2):
+256--62.}
+
+\href{http://paperpile.com/b/sTGYvp/nFf7}{Wills, Matthew A. 2001.
+``Morphological Disparity: A Primer.'' In
+}\href{http://paperpile.com/b/sTGYvp/nFf7}{\emph{Topics in
+Geobiology}}\href{http://paperpile.com/b/sTGYvp/nFf7}{, 55--144.}
+
+\href{http://paperpile.com/b/sTGYvp/eZ3F}{Wills, Matthew A., Derek E. G.
+Briggs, and Richard A. Fortey. 1994. ``Disparity as an Evolutionary
+Index: A Comparison of Cambrian and Recent Arthropods.''
+}\href{http://paperpile.com/b/sTGYvp/eZ3F}{\emph{Paleobiology}}\href{http://paperpile.com/b/sTGYvp/eZ3F}{
+20 (02): 93--130.}
+
+\href{http://paperpile.com/b/sTGYvp/s33b}{Wright, David F. 2017.
+``Phenotypic Innovation and Adaptive Constraints in the Evolutionary
+Radiation of Palaeozoic Crinoids.''
+}\href{http://paperpile.com/b/sTGYvp/s33b}{\emph{Scientific
+Reports}}\href{http://paperpile.com/b/sTGYvp/s33b}{ 7 (1): 13745.}
+
+\href{http://paperpile.com/b/sTGYvp/ZQS9}{Zelditch, Miriam Leah, Donald
+L. Swiderski, and H. David Sheets. 2012.
+}\href{http://paperpile.com/b/sTGYvp/ZQS9}{\emph{Geometric Morphometrics
+for Biologists: A Primer}}\href{http://paperpile.com/b/sTGYvp/ZQS9}{.
+Academic Press.}
+
+\href{http://paperpile.com/b/sTGYvp/ZqML}{Zou, Zhengting, and Jianzhi
+Zhang. 2016. ``Morphological and Molecular Convergences in Mammalian
+Phylogenetics.'' }\href{http://paperpile.com/b/sTGYvp/ZqML}{\emph{Nature
+Communications}}\href{http://paperpile.com/b/sTGYvp/ZqML}{ 7
+(September): 12758.}

### 2019-10-30 Writing/disparity_review_old.tex
Updated figures

--- 2018-05-18 Writing/disparity_review.tex
@@
 \bigskip
-\noindent RH: pest practices for disparity analysis.
+\noindent RH: disparate disparity analysis.
 
@@
 %---------------------------------------------
+
 \section{Introduction}
 
+%Historical blurb
 Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
@@
 One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.
-% The multidimensional space for mammalian carnivores, for example, encompasses all combinations of the traits of the order Carnivora, along with the ecological, life history and behavioural variables characteristic of meat eating mammals.
-% Each individual carnivore species occupies a discrete region of this space related to its values for these traits.
 
@@
 The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
-Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation (REF Katie et al's paper somewhere here).
+Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
 Here we use a broader definition and suggest that disparity is a pattern of variation.
@@
 
-% \subsection{What \textit{is} disparity?} %TG: maybe useful for box or paraphrapsing summaries.
-% What \textit{is} disparity? In the seminal disparity papers from the late 80s, disparity was used in analyses of body plan evolution after the Cambrian explosion (REF).
-% The initial informal definition of disparity was, therefore, ``multidimensional morphological dissimilarity at a macroevolutionary scale''.
-% Currently, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation.
-% There is also debate about whether disparity refers to a pattern, a process that generates the pattern, or both. 
-% Even if disparity only describes patterns of variation, the scientific questions of disparity analyses generally use these patterns as a proxy for explaining the biological processes underlying them.
-% The original Cambrian explosion papers, for example, use disparity to describe both the pattern of high dissimilarity in body plans among groups, and the process of explosive radiation that generated them (as championed by Gould; REF).
+\subsection{Is disparity equivalent to diversity?}
+One obvious aspect when measuring biodiversity is the measurement of raw species diversity (at any scale; $\alpha$, $\beta$, $\gamma$).
+Such biodiversity metric intuitively describes the amount of reproductive isolated populations in a system.
+On the other hand, disparity aims at describing the amount of measured traits in a system (see below), in such a way, a high morphological disparity value is a measurement of a high diversity of morphologies (i.e.
+shapes or body plans).
+The link between both disparity and diversity has been scrutinised under different scenarios with most conclusions supporting an effective decoupling between both [CITE the disparity/diversity papers].
+This is intuitively understandable when comparing both metrics for two contrasted groups of organisms, for example, living rodents would display a high diversity ($>$2200 sp; CITE Wilson\&Reeder) with a relatively low disparity (a majority of rodents have the typical rat shape) whereas carnivores would display a relatively lower diversity ($>$280 sp; CITE Wilson\&Reeder) with a higher disparity (with body shapes ranging from a weasel to an elephant seal).
+However, it is important to note that in more nuanced groups, diversity and disparity can be link where a really low diversity would tend to yield a low disparity and vice versa.
 
-% Restricting disparity to its original definition can be advantageous for advances in this field but broadening it allows more interdisciplinary discussion.
-% In recent years, disparity has been used to investigate ecological processes, disconnected from evolution (e.g. function [CITE], community structure[CITE], competition[CITE]).
-% Thus studying disparity could not only inform us on macroevolutionary patterns and processes but also something about environments and ecosystem functioning (with the advantages of easily allowing palaeocological and palaeoenvironemental perspectives).
+This decoupling has led researchers to work on the relation between disparity and evolvability: a more disparate group (population, group, etc...) will have more genetic diversity to adapt to biotic and abiotic changes [CITES].
+This led to the use of disparity as a tool for investigating morphological modularity [CITE].
+However, this is based on the assumption that there is a link between diversity in traits (disparity) and subsequent diversity in genes and/or in ecological niches ([Trait plant paper CITE] see below for more details).
+Ultimately, both links between disparity and diversity and disparity and evolvability are dependent on the multiple definitions of disparity and the multiple datasets used.
 
-\section{Best practices for disparity analysis}
+\section{A diversity of disparities}
+% IN -> There are many different disparity
+This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
+From a same analysis comparing morphological traits among different organisms, one can study four main types of disparity:
 
-This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
-However, as with any great power, it comes with great responsibility. 
-With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
-First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
-Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
-Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
-Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.
+\begin{enumerate}
 
-\section{What is the question?}
-Why perform a disparity analysis?
-This seems like it should be obvious, but we stress its crucial importance.
-The last decade has seen a vast array of new methods being published and implemented in freely available software, making disparity analyses easy to perform \citep{bouxin2005ginkgo,oksanen2007vegan,geiger2008,zelditch2012geometric,adams2013geomorph,Claddis,dispRityv02,adams2017geometric}.
-However, this (consciously or not) pushes authors towards performing disparity analyses simply because of their ease of implementation, rather than stemming from a question driven approach.
-Before rushing into a disparity analysis, one should consider some basic caveats.
-For example, one major thing to consider is circularity: if the aim is to explore the variation in a dataset, disparity analysis is perfectly valid.
-However, if this exploratory analysis reveals distinct groups, it would be circular to then test for differences among those groups as if this had been the hypothesis all along [CITE stats 1.0].
-This can easily be avoided by using a subset of the data for preliminary data exploration, and then using the full dataset to test for differences among any groups that emerge.
-Clearly framing questions and their null expectations will help in designing better disparity analyses.
+\item{Morphological disparity.}
+The disparity of morphological trait configurations has been at the genesis of disparity analysis either through Foote's pioneering work [Foote Crinoids] or Raup's morphospaces [Raup].
+This type of disparity analysis often consists in a more descriptive multidimensionality analysis of the patterns in the diversity of morphological traits and has been successfully used to answer questions such as why are some morphological traits combinations more common than other [Raup] and what are the properties of the resulting morphospace [Gerber].
+It is important to highlight that the morphological data used in this context was collected specifically for these analyses and not ``recycled'' from phylogenetic analysis.
 
-\section{Collect the appropriate data}
-Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
-For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for species delimitation.
-Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below).
+\item{Phylogenetic disparity.}
+Another disparity approach is to use the same morphological disparity approach as mentioned above but using discrete morphological data commonly used for phylogenetic analysis.
+This approach is widely used in palaeobiology to answer a range of macroevolutionary questions [CITEs string].
+Although very useful for describing macroevolutionary patterns, this approach can tend to artificially increase disparity between phylogenetically distinct groups due to the nature of so phylogenetic characters use to discriminate groups [CITE Foote].
 
-\subsection{Data types}
-Morphological disparity analysis can be based on three main types of data:
-\begin{enumerate}
-    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g.
-the presence or absence of legs, the colour of a tail, etc.
-[CITE]) or count the number of features (e.g.
-number of digits, etc.
-[CITE]);
-    \item continuous data that can be either simple measurements (e.g.
-body length, etc.
-[CITE]) or more abstracted measures (e.g.
-geometric morphometrics, mathematical contour description, etc.
-[CITE]);
-    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
+\item{Functional disparity.}
+Functional disparity has since then been assumed to work at a large scale and can be used to study convergence (species with similar trait combination can assume similar functions).
+Furthermore there as been work on linking form and function (Diaz paper).
+However, the link between trait and function can sometimes be hard to grasp since traits can be linked to multiple functions and multiple functions can be linked to a single trait.
+
+\item{Ecological disparity.}
+Linking to functional disparity, people have proposed to use morphological trait disparity as a proxy for ecological niches [CITE].
+This seems to work at a macro-evolutionary scale but is still debated at a micro-evolutionary scale [CITE plant trait paper].
+For example, the concept of different ecological guilds can easily be distinguished through the fossil record (apex predator have share morphological characteristics through time; e.g.
+large body mass, pointy teeth, etc.) but can be more difficult at a finer scales and when defining the actual ecological niche (i.e.
+does a T-Rex occupy the same top predator niche as an Orca?).
+
 \end{enumerate}
-None of these types of data are fundamentally better but each will be more appropriate for specific questions.
-For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
-On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
-This again illustrates the importance of framing the question.
-In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
+
+This diversity of patterns that can be studied using disparity analyses links to the ambiguous distinction between disparity as a pattern or a process.
+In other words, do these disparities represent anything more than the variance of those traits or is it link to specific biological properties of these traits?
+For example, does the distribution of traits in apex predators allow them to be distinguished in disparity analysis (i.e.
+disparity is the pattern) or is there something about being an apex predator that results in this combination of traits (i.e.
+disparity is the process).
+As highlighted above, for the rest of this review, we will essentially use disparity as a term for describing morphological traits patterns but we acknowledge that disparity can sometimes also be the process leading to the observed patterns (e.g.
+in the case of trait convergence).
+Furthermore, as a best practice, we do advice researchers to select the traits that will be fit there question \textit{a priori} and not measure all the traits available.
+For example, when studying disparity in locomotion in a group of animals, postcranial characters might more readily answer the question than cranial ones.
+
+\section{Trait disparity}
+% IN -> If it's just traits, can all of the traits be used?
+In essence, disparity studies are multidimensional analyses of morphological traits.
+Such traits can be collected using three main traits: 1) discrete ``cladistic'' characters (e.g.
+coded as the absence or presence of features [CITE any paleo disp paper]); 2) continuous measurements (e.g.
+length of features [CITE Anderson fish paper]) or 3) more sophisticated mathematical measurements from in geometric morphometric data (e.g.
+Pcrocrustes superimpositions of entire features [CITE a geomorph paper], Fourier contour analyses [CITE], etc.).
+
+None of these types of data are fundamentally better and each have a range of materialistic pros and cons (e.g.
+cost, time, range, etc.), however, some specific questions or methodological pipelines can be more appropriated to some types of data.
+For example, when investigating variation within bat wing shapes, both homologous landmarks (e.g.
+used in geometric morphometric analysis) or continuous measurements of bones may be appropriate to capture patterns of variations.
+Conversely, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question from a mechanical perspective.
+In the above example, if the question is about functional disparity, say, whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
 wingspan, aspect ratio, etc.).
-Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g.
+On the opposite, if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g.
 digit length, integumentary system, etc.).
+It is therefore important to consider disparity patterns as traits disparity patterns approximating biologically complex patterns.
 
-\subsection{How much data should be collected?}
-Regardless of the type of data, another important consideration is how much data to collect, in terms of numbers of specimens and traits.
-\begin{itemize}
-    \item \textit{Taxonomic scope.}
-    Morphological disparity has been studied at both inter-[CITE] and intra-specific levels [CITE].
-    Taxonomic scope will depend on the question being investigated, and what type of variation is being analysed.
-    For example, when studying skull shape plasticity in canids, it is essential to include intra-specific variation to capture the full variety of skull shapes, particularly that seen in domestic dogs (CITE).
-    Conversely, if a study focuses on canid skull variation through time, interspecific data is most appropriate.
-    % intra-specific variation will not have major impact (up until the Holocene). %http://www.journals.uchicago.edu/doi/abs/10.1086/650372
-    \item \textit{Sample size.}
-    It is important to ensure that the data is collected in a way that will not bias subsequent statistical analysis.
-    For example, if the aim is to compare disparity among groups, one should collect, where possible, a similar amount of data for each group.
-    When this is not possible (e.g. for palaeontological data), one should use statistical approaches to correct or assess biases due to data collection (e.g. rarefaction [CITE]).
-    Equally, one should ensure that enough data is available so that the methods have power to detect the patterns being tested.
-    \end{itemize}
+\subsection{Are different traits equivalent?}
 
-Of course in practice some of these caveats are unavoidable, especially when dealing with the vertebrate fossil record.
+From the example above, the question can be asked whether using different methods to approximate the same traits on the same features would lead to similar results.
+In other words, does discrete, continuous or geometric morphometric data collected from a same feature capture the same disparity signal?
+This question has been assessed several times by comparing the ``disparity signal'' between cladistic and geometric morphometric data in vertebrates leading to the conclusions that data originating from the same source leads to the same results [CITE].
+Although the advantages and caveats of each type of traits are well known, there is no clear boundaries to which extent the equivalence between dataset holds.
+These equivalence might be more primordial at the trait level (i.e. which traits do we measure) rather than at the measurement method level (i.e. how do we measure them) [Foote Crinoids].
 
-\subsection{Recycling data}
-A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
-One classic case is using ``cladistic'' data in palaeobiology.
-These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
-It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
-A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all of the data.
-For example, in a study of the disparity of dietary niches, removing non-dietary characters is a good approach.
-However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g.
-limb traits can be used for both locomotion and feeding).
+\subsection{Does the part represent the whole?}
 
-\subsection{Other data considerations}
-There are many other considerations when choosing the appropriate data to use. 
-It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
-For example, are the observed differences between two groups related to factors unrelated to group designation such as sexual dimorphism?
-Are the taxa/groups being compared fairly consistent and truly comparable?
-In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated [String of morpho data problems cite].
-It is important to ensure that any patterns detected are not merely reflections of these features of the data.
+In this context, it can be important to understand the relations between the analysis of one specific set of traits and the macroevolutionary questions asked at a organismal level.
+For example, when studying mammalian morphological evolution through time based on cladistic characters, do the disparity patterns reflect the changes through time of mammalian morphological niches or do they merely approximate variation in one specific set of traits?
+In mammalian discrete datasets, many characters are dental and may not actually represent one organsims' niche (i.e.
+if the dataset is missing other crucial characters like locomotory ones).
+%TG: maybe add an example with stuff with the same dentition but living in a different environement, e.g.
+trees/ground).
+However, studies have shown that
+-[mammalian teeth are not that bad - Zou and Zhang] 
+-[the head can be enough to represent disparity - Melanie's work]
 
+Furthermore, trait data suffers from the same short comings as most data in biology (i.e.
+allometry, homology, sexual dimorphism, etc...) or to short comings from data in general (missing, non-overlapping, hierarchical - e.g.
+inapplicable -, ambiguous, polymorphic or correlated, sample size [String of morpho data problems cite]).
 Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
+% OU -> And how do these traits need to be analysed?
 
-\section{Use the appropriate methods} 
-In the vast majority of morphological disparity analyses, the methodological approach will be multidimensional.
-Multidimensional analysis can be a powerful tool for analysing complex datasets but also comes with many mathematical caveats and limitations.
-The first step in any statistical analysis should be to visualise the data to understand its distribution and limitations.
-Unfortunately, we can only perceive a maximum of three dimensions at the same time.
-Therefore one of the main challenges in multidimensional analysis is how to reduce the number of dimensions, both for visualisation and analysis (see box \ref{box_ordination}).
+\section{Disparity is multidimensional}
 
+Additionally to pitfalls and caveats stemming from the biological questions and the physical datasets analysed through the lens of morphological disparity, the multidimensional aspect of these analysis can also add a layer of technical complexity.
+In fact, disparity analysis are inherently multidimensional and our human brains and the physical word is bound to three spatial dimensions.
+This problem is sometime apparent in over-interpretation of disparity analysis results.
+For example, in a classic morphological disparity pipelined analysis using geometric morphometric data, the collected data's dimensionality is the number of collected landmarks and their 3 or 2 dimensional set of coordinates.
+However, using dimensionality reduction techniques can lead to misinterpretation of the results if not careful.
+In the case of ordination, it is not uncommon to then use only the first axis of ordination (e.g. PC1) and draw conclusions based on this axis.
+The changes in values along this dimension can display some disparity pattern or gradient at a glance but don't often reflect the actual changes of the set of all the landmarks' dimensions [CITE Vera and Thomas paper].
+Therefore, whether for visualising or analysing disparity data, it is crucial to remember it's multidimensionality aspect.
+
+% IN -> Disparity is multidimensional by essence
+% Visualisation problem
 \subsection{Visualising the multidimensional problem}
 \label{visualisation}
-It is important to remember that visual interpretations of multidimensional data can easily be misleading.
+Visual interpretations of multidimensional data can easily be misleading.
 For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
@@
 
-It is therefore vital to be clear about what is represented in multidimensional space visualisations.
-In fact, it can be easy (intentionally or not) to mislead a reader into over/under interpreting a pattern if all the key properties of the space are not present in the graph.
-We suggest visualisations should include the following:
-(1) the percentage of the variation explained by each axis of the plot (i.e. is the majority - or all - the variation in the data present in the plot or only a small fraction?);
-(2) the scale of the axes (ideally, if the space is orthogonal, it is important to keep the same scale across the axes);
-and (3) the error associated with each point, whether this is in terms of measurement error (on ordination plots for example) or some other measure of standard variation (in further analysis, e.g.
-disparity-through-time plots).
-Finally, it is important to remember that for any dimensionality reduction of a multidimensional problem, the visual result will be at best a distortion of the measured reality, at worst, it will be misleading (see box \ref{box_ordination}).
+%     Do we need trait reduction (ordination) or can we do without?
 
-\subsection{[Box]To ordinate or not to ordinate; that is the (multidimensional) question?}
-\label{box_ordination}
-Reducing the dimensionality of data through ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
+\subsection{To ordinate or not to ordinate; that is the (multidimensional) question?}
+Similarly as for visualisation, reducing the dimensionality of a disparity dataset can be tricky.
+Ordination can be advantageous for plotting and visualising the data, and can reveal properties of the morphospace not captured by disparity metrics (see Metrics section \ref{metrics}).
 Additionally, after ordinating the data it is possible to further reduce the number of dimensions by only selecting a certain number of axes (e.g.
@@
 Because of these points, we strongly recommend that data for multidimensional analyses is not automatically ordinated, and that careful consideration is given to whether the question can be answered without ordination.
-\subsection{[Box] end}
 
-\subsection{Phylogeny}
-For many disparity analyses of macroevolutionary questions, phylogeny is often taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
-For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
-and therefore will not covered further in this review.
+%     What do we measure from these multidimensional spaces?
+\subsection{Measuring disparity}
 
-\subsubsection{Ancestral state estimation}
-When looking at disparity in the past (e.g. disparity-through-time) ancestral state estimations are a common approach.
-Ancestral state estimation can be performed in two ways:
-\begin{enumerate}
-\item pre-ordination: the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original dataset[CITE].
-\item post-ordination: the estimation is done after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
-\end{enumerate}
-Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e.
-the relationship between the points not estimated) and implies longer computational times but, once the morphospace is defined, it will remain constant (i.e.
-its properties will not change).
-Post ordination ancestral state estimation will not change the ordinated space's geometry and will be faster to compute, however, it will change the morphospace's properties such as the number of parameters (i.e.
-elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper to be?].
-
-More importantly, however, one needs to remember that using such ancestral state estimation methods comes with several caveats.
-First and foremost, although ancestral state estimations are valid methods in macroevolution, their results are highly dependent on the data and chosen methods.
-Secondly, more specific to disparity analysis, when using post-transformation ancestral state estimates, the morphospace can be saturated leading pairwise distance metrics to tend to zero.
-In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points for statistical reasons.
-In fact, these extra points are not independent and can also generate side effects that can be problematic, especially when testing the effect of extinction events on disparity.
-
-Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
-In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes.
-Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution.
-
-\subsection{Metrics}
-\label{metrics}
-
-[INCLUDE WILLS FIGURE SOMWHERE AROUND]
-
-%TG: add the curse of dimensionality
-
-After building and visualising the morphospace, one can use disparity metrics (or indices) as a one dimensional summary of the space and to test hypotheses.
+A second important point stemming from the multidimensional nature of disparity analyses is considering how to extract some meaningful interpretable summary information: the disparity metric (or index).
 As with any summary metric, and because of this reduction in dimensionality, disparity metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
@@
 Among these metrics, some will have more preferable properties than others, such as sensitivity to outliers (range, mean or mode are highly sensitive \textit{vs.} quantiles, variance or median are less sensitive) and will thus be more or less appropriate to the different questions and the properties of the distribution.
-For example, if the data collected is likely to contain outliers, but is fairly normally distributed, one should prefer the median to the mean.
-
-With multidimensionality, this high availability of unidimensional descriptors comes with equally numerous advantages and caveats and again, the question can help in determining which metric is the most appropriate.
 For example, if the question refers to ``size'' of a group in the morphospace (e.g.
@@
 product of ranges, hypervolume, hypercube, etc.) will tend towards zero fairly quickly for spaces with even a modest number of dimensions [CITING Bellman 1957; also see \url{https://beta.observablehq.com/@tophtucker/theres-plenty-of-room-in-the-corners} for an excellent illustration of the problem and Donoho 2000 Aide-Memoire].
-Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size, for example range or convex hull based metrics.
+Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size, for example range [CITE] or convex hull based metrics [CITE].
 
-%TG: again, I feel this section is already fairly long, I might include this bit in another paper
+% OU -> And how do we test these things?
+The technical properties of multidimensional disparity analyses themselves allow for the vast amount of methods and metrics described above to be used in specific context.
+However, as mentioned above equally, these tools should always be used in the frame of a specific stated scientific question.
+This leads to the importance of clearly defining and rigorously testing on or more hypotheses.
 
-% NC: Sounds reasonable. This could be cut down more, but for now I've just edited the grammar and spelling so we can look at the thing as a whole and also get Phil's input.
-
-% Of course some other aspects could be of interest (or more than one) and thus some other metrics would be more appropriated.
-% For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution).
-% Furthermore, one future avenue that could be explored is to go beyond single valued disparity metrics and rather express them in terms of distributions.
-% For example, when the metric is expressed as the median distance from centroids [e.g. in CITE], it could be informative as well to express it as a full distribution.
-% This would allow to both visualise the variation at a glance and use this variation in statistical tests for comparing these distributions.
-
-% Best practices:
-% \begin{itemize}
-% \item Facing the amount of different ways to measure disparity and regardless which metric is the most suited to the question, it is crucial to clearly and mathematically define the metric in a disparity paper.
-The definition can be in plain English: e.g.
-``we measure disparity as the sum of the variance of each axis of the ordinated matrix''; or mathematically: e.g.
-``we measured disparity as $\sum\sigma_i$ where $\sigma_i$ is the variance of the each ordinated axis $i$.''.
-% \item It is important to keep in mind that different metrics can lead to the same results even though the described spaces are different (the opposite is \textit{not} true).
-We therefore suggest to always plot some dimensions of the space (see Visualisation section \ref{visualisation}) and try multiple metrics.
-% \end{itemize}
-
-\subsection{Tests}
-The final stage of most disparity analyses is statistical testing of the disparity metrics, to determine whether changes thought time or between groups are statistically significant. 
-Similar to the points above, the statistical test to apply depends on which question is being tackled with disparity analysis, i.e. which hypothesis is being tested.
-It also requires consideration of the data, the ordination method used (or the absence thereof) and the metric used.
-
+\section{Disparity hypotheses}
+% IN -> So how do we go for testing hypotheses in disparity?
+%     What are our null expectations?
+The multidimensional statistical toolkit has been greatly developed in the last years [CITE Dean Adams, Simone, etc.] but oddly enough not routinely applied to disparity analyses.
+Instead, disparity hypotheses testing seemed to be confined to a small set of (well established) methods.
 For example, one commonly used test in morphological analysis, is the non parametric permutation analysis of variance [NPANOVA or PERMANOVA; CITE Anderson + examples].
-This test and its implementation is perfectly valid to test whether two groups have a different variance/covariance, but is based on the pairwise distances of the group elements.
-This test will ignore any ordination (if the data was ordinated) and is not based on a unidimensional disparity metric.
-In other words, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
+This test is an analysis of variance in the pairwise distances between different groups and is often completely independent of the type of morphospace and the disparity metric used.
+Although perfectly valid, this test can sometimes not be directly related to the hypotheses stated.
+For example, if one observes no overlap between two groups on the two first axes of a PCA, the null hypothesis that the two groups overlap in the morphospace is not directly tested by a PERMANOVA which tests rather whether the two groups share the same variance/covariance in a ``distance-space''.
 Therefore, explicitly stating the hypothesis might help understanding which test to apply depending on the question.
@@
 
-\section{Future directions}
+\subsection{Phylogeny}
+More specifically within the realm of testing hypothesis, it is important to remember that when analysing biological data, the data is not independent [CITE Felsenstein or so].
+For many disparity analyses of macroevolutionary questions, phylogeny must thus be taken into account, whether for including ancestral state estimates or to incorporate phylogenetic structure in the data.
+For the latter, multivariate phylogenetic comparative methods have been reviewed recently by [CITE Adams] %https://academic.oup.com/sysbio/article-abstract/67/1/14/3867043
+and therefore will not covered further in this review.
+
+\subsubsection{Ancestral state estimation}
+When looking at disparity in the past (e.g. disparity-through-time) ancestral state estimations are a common approach.
+Ancestral state estimation can be performed in two ways:
+\begin{enumerate}
+\item pre-ordination: the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original dataset[CITE].
+\item post-ordination: the estimation is done after transformation of the data by estimating the ancestral states based on the transformed matrix (e.g. the ordination scores) [CITE].
+\end{enumerate}
+Pre-ordination ancestral state estimation can change the ordinated space's geometry (i.e.
+the relationship between the points not estimated) and implies longer computational times but, once the morphospace is defined, it will remain constant (i.e.
+its properties will not change).
+Post ordination ancestral state estimation will not change the ordinated space's geometry and will be faster to compute, however, it will change the morphospace's properties such as the number of parameters (i.e.
+elements in the space - this can be problematic for statistical tests down the line) or the variance on each axis [CITE Graeme's paper].
+
+More importantly, however, one needs to remember that using such ancestral state estimation methods comes with several caveats.
+First and foremost, although ancestral state estimations are valid methods in macroevolution, their results are highly dependent on the data and chosen methods.
+Secondly, more specific to disparity analysis, when using post-transformation ancestral state estimates, the morphospace can be saturated leading pairwise distance metrics to tend to zero.
+In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points for statistical reasons.
+In fact, these extra points are not independent and can also generate side effects that can be problematic, especially when testing the effect of extinction events on disparity.
+
+Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
+In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes [CITE time slice].
+Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution [CITE STD paper?].
+%     How can we analyse it through time?
+
+\section{A disparate future?}
+% IN -> The definition of disparity and its uses have broadened since the early disparity start.
+
 As mentioned above, there is no ``one-size-fits-all'' morphological disparity analysis pipeline.
@@
 
+%     What can we get from other fields to make disparity better
+%     But also what can we bring to other fields (discussions with Abigail)
+
 \subsection{How can we adapt our methods?}
@@
 
+\subsection{Conclusion}
+However, as with any great power, it comes with great responsibility. 
+With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
+First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
+Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
+Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
+Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.
+
+%Concluding words:
+Although disparity analyses are now simple to implement (CITE packages tools), it is crucial to remember that they are multidimensional analyses; and multidimensional analyses are complex.
+In Jurassic Park, Dr Ian Malcolm summarises this problem in an elegant way: ``[We] scientists were so preoccupied with whether or not [we] could that [we] didn't stop to think if [we] should.''
+We believe that morphological analysis in general will greatly benefit from emphasising the methodological decisions made (\textit{whether we should}), rather than simply using disparity analysis because \textit{we can}.
+
 % NC: I do like this quote.
@@
 \end{document}
+
+% Old bits from Google doc (2019-01-16)
+% OLD section 4
+% TAs mentioned above, there is no “one-size-fits-all” morphological disparity analysis pipeline.
+As with any multidimensional analysis, there are many nuances in deciding which data to use and how to analyse it, stemming from the explicit formulation of the disparity hypothesis being tested.
+However, it is worth remembering that To understand these nuances, it is important to remember that: (1) morphological disparity analyses are not new and much effort has been made in developing methods; and (2) multidimensional analyses are not restricted to morphological data, analysis thusand thus solutions to specific problems are oftencan be availablefound in literature that is not necessary focused on morphology,  (especially within ecology and mathematics).
+% We believe that many of the problemscaveats in morphological disparity analysis can arise when “blindly” applying methodological pipelines whileat the expense of losing sight of the biological question being tested.
+We therefore advise keeping the methodology simple and tractable, at least in the preliminary analysis.
+Also, many of the caveats and technical problems in disparity analyses have been tackled in other fields, mainly ecology.
+This is perhaps unsurprising,ly as ecological questions are often closely related to evolutionary questions, and ecological data are also often multidimensional (Donohue et al.
+2013; Canter et al.
+2018).
+Seeking solutions using these literature is a sensible next step for disparity analysis.
+% 4.1. How can we adapt our methods?
+% Each of the caveatspoints outlined above may drive new methodological questions (and often already are).
+Although it is impossible to cover the state-of-the-art exhaustively for each of these points here are several solutions and/or directions for adapting existing methods:
+% Which data should we use in disparity analyses?
+Some simple methods already exist to account for caveats in data.
+For example, to correct for the paucity of the fossil record, one can use methods like rarefaction analysis to take sample size into account (Butler et al.
+2012).
+% How should we visualise disparity?
+One major problem of multidimensional analyses is that highly dimensional spaces are hard (or even impossible) to conceptualise.
+One solution to this is to use more simple 2D/3D ordinations.
+For example, using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation (Raup 1961; Gerber 2017).
+% Which disparity metrics should we use?
+This can be tackled by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combination of properties of interest in the morphospace.
+For example, when reducing a multidimensional space to one metric, it would be judicious to indicate what this metric represents in the results (e.g.
+in a plot or a table, if the metric used was the sum of variances, labelling it as “size of the group” rather than “disparity”).
+This is especially advantageous when using multiple disparity metrics (e.g.
+“size of the group” and “position of the group”).
+% How should we test disparity hypotheses?
+Using available software and computer power, it is possible to propose specific null models for the expectation of what disparity would be expected to look like under various null models.
+This has been already successfully used in ecology (Díaz et al.
+2016) and is currently being implemented in palaeobiology (Murrell 2017; Slater and Pennell 2013).
+% 4.2. Where do we go from here?
+% Morphological disparity analyses are a major tool in evolutionary biology, posing a more versatile alternative to studying species richness, especially in macroevolutionary studies.
+We believe that morphological disparity analyses, as the analysis of patterns of morphological differences, can be of great use for linking evolutionary and ecological processes, for example when using morphological disparity as a proxy for ecological niches to test competition-through-time hypotheses.
+% However, as demonstrated above, such analyses would greatly benefit from more development, and more care is needed in study design so that the analysis actually reflects the biological question at hand.
+This can not only be done through the points highlighted above but also through linking the observed patterns to an explicit process (for example, the link between an organism’s form and function) or by developing null models to potentially explain the observed patterns.
+% Finally, we encourage to a more cross-disciplinary approach to morphological disparity analyses.
+In fact, as mentioned above, morphological disparity would greatly benefit from advances in multidimensional analysis in different fields but similarly, the concept of a morphospace could easily be exported to other fields.
+For example, the multidimensional analysis in Diaz (Díaz et al.
+2016) looking at the patterns of form and function in plants can be thought as a morphospace or a life-history-space (an eco-space?); isotopic analyses in (Jackson et al.
+2011) can be represented as a isotope-space; ecosystem functioning in (Donohue et al.
+2013) as an ecosystem-space, etc.
+These generalisations could also be exported for any set of traits (e.g.
+acousto-spaces for acoustic traits?) and even beyond evolutionary biology (e.g.
+glotto-spaces for linguistic traits?).
+% Although disparity analyses are now simple to implement (Guillerme 2018; Adams and Otárola-Castillo 2013; Bouxin 2005; Harmon et al.
+2008; Lloyd 2016; Navarro 2003; Dixon 2003), it is crucial to remember that they are multidimensional analyses; and multidimensional analyses are complex.
+In Jurassic Park, Dr Ian Malcolm summarises this problem in an elegant way: “[We] scientists were so preoccupied with whether or not [we] could that [we] didn’t stop to think if [we] should.” We believe that morphological analysis in general will greatly benefit from emphasising the methodological decisions made (whether we should), rather than simply using disparity analysis because we can.

### 2019-10-31 Analysis/Figures_in_R.Rmd
Fig script up

--- 2018-11-19 Analysis/Figures_in_R.Rmd
@@
 ```
+
+## Iris example
+
+```{r}
+## Preparing the dataset
+colnames(iris)[1:4] <- gsub("\\.", "\n", colnames(iris)[1:4])
+rownames(iris) <- c(paste0("setosa", which(iris[,5] == "setosa")),
+                    paste0("versicolor", which(iris[,5] == "versicolor")-50),
+                    paste0("virginica", which(iris[,5] == "virginica")-100))
+
+## Morphospaces:
+## No transformations
+write.csv(head(iris[, -5]), file = "../Writing/Figures/trait.space.no_trans.csv")
+
+## Distance matrix
+iris_dist <- dist(iris[, -5])
+write.csv(round(head(as.matrix(iris_dist)[1:5, 1:5]), digits = 2), file = "../Writing/Figures/trait.space.distance.csv")
+
+## Ordinations
+iris_pca <- princomp(iris[, -5])$scores
+write.csv(round(head(iris_pca), digits = 2), file = "../Writing/Figures/trait.space.pca.csv")
+iris_pco <-cmdscale(iris_dist, k = 2)
+
+## Visualisations
+## biplots
+col_var <- c("orange", "blue", "green")[as.factor(iris[,5])]
+pdf(file = "../Writing/Figures/figure_methods_biplots.pdf", width = 5, height = 5)
+plot(iris[, -5],
+    pch = 19,
+    col = col_var)
+dev.off()
+
+## PCO
+pdf(file = "../Writing/Figures/figure_methods_PCO.pdf", width = 5, height = 5)
+
+vars <- round(apply(iris_pco, 2, var)/sum(apply(iris_pco, 2, var))*100, 2)
+plot(iris_pco,
+    pch = 19,
+    col = col_var,
+    xlab = paste0("PCO 1 (", vars[1], "%)"),
+    ylab = paste0("PCO 21 (", vars[2], "%)"),
+    xlim = range(iris_pco),
+    ylim = range(iris_pco))
+dev.off()
+
+## PCA
+pdf(file = "../Writing/Figures/figure_methods_PCA.pdf", width = 5, height = 5)
+vars <- round(apply(iris_pca, 2, var)/sum(apply(iris_pca, 2, var))*100, 2) 
+ranges <- apply(iris_pca, 2, range)
+
+par(mfrow = c(4,4), mar = c(1,1,1,1), oma = c(2,2,2,2))
+## First row
+# 1
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PC 1\n(", vars[1], "%)"), cex = 1.9)
+# 2
+plot(iris_pca[, c(1,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(3)
+# 3
+plot(iris_pca[, c(1,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 4
+plot(iris_pca[, c(1,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(3)
+axis(4)
+
+## Second row
+# 1
+plot(iris_pca[, c(2,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(2)
+# 2
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PC 2\n(", vars[2], "%)"), cex = 1.9)
+# 3
+plot(iris_pca[, c(2,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 4
+plot(iris_pca[, c(2,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(4)
+
+## Third row
+# 1
+plot(iris_pca[, c(3,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(2)
+# 3
+plot(iris_pca[, c(3,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 3
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PC 3\n(", vars[3], "%)"), cex = 1.9)
+# 4
+plot(iris_pca[, c(3,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(4)
+
+## Fourth row
+# 1
+plot(iris_pca[, c(4,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(1)
+axis(2)
+# 3
+plot(iris_pca[, c(4,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 3
+plot(iris_pca[, c(4,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(1)
+# 4
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PC 4\n(", vars[4], "%)"), cex = 1.9)
+dev.off()

### 2020-01-24 Writing/disparity_review.tex
Updated latex

@@
 
-\textbf{The disparate ways to analyse morphological disparity}
-
-Thomas
-Guillerme\includegraphics[width=0.4902in,height=0.1807in]{./ObjectReplacements/Object 2},
-Natalie
-Cooper\includegraphics[width=0.3984in,height=0.1807in]{./ObjectReplacements/Object 4},
-Stephen
-Brusatte\includegraphics[width=0.1937in,height=0.1807in]{./ObjectReplacements/Object 6},
-Katie
-Davis\includegraphics[width=0.1992in,height=0.1807in]{./ObjectReplacements/Object 8},
-Andrew L
-Jackson\includegraphics[width=0.1937in,height=0.1807in]{./ObjectReplacements/Object 10},
-Sylvain
-Gerber\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 12},
-Anjali
-Goswami\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 14},
-Kevin
-Healy\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 16},
-Melanie
-Hopkins\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 18},
-Graeme
-Lloyd\includegraphics[width=0.1953in,height=0.1807in]{./ObjectReplacements/Object 20},
-Joseph
-O'Reilly\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 22},
-Abi
-Pate\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 24},
-Emilie
-Rayfield\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 26},
-Erin
-Saupe\includegraphics[width=0.239in,height=0.1807in]{./ObjectReplacements/Object 28},
-Emma
-Sherratt\includegraphics[width=0.2425in,height=0.1807in]{./ObjectReplacements/Object 30},
-Graham
-Slater\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 32},
-Gavin H
-Thomas\includegraphics[width=0.2465in,height=0.1807in]{./ObjectReplacements/Object 34}
-and Philip
-Donoghue\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 36}\\
-\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 38}\emph{School
-of Biological Sciences, University of Queensland, St. Lucia, Queensland,
-Australia.}\\
-\includegraphics[width=0.1929in,height=0.1807in]{./ObjectReplacements/Object 40}\emph{Department
-of Life Sciences, Natural History Museum, London, Cromwell Road, London,
-SW7 5BD, UK.}\\
-\includegraphics[width=0.2437in,height=0.1807in]{./ObjectReplacements/Object 42}\emph{Bristol}
+\textbf{Disparities in the analysis of morphological disparity}
+
+Thomas Guillerme\({}^{1,*, +}\), Natalie Cooper\({}^{2, +}\), Stephen L.
+Brusatte\({}^{3}\), Katie E. Davis\({}^{4}\), Andrew L
+Jackson\({}^{5}\), Sylvain Gerber\({}^{6}\), Anjali Goswami\({}^{2}\),
+Kevin Healy\({}^{7}\), Melanie Hopkins\({}^{8}\), Marc EH
+Jones\({}^{9}\), Graeme T. Lloyd\({}^{10}\), Joseph E.
+O'Reilly\({}^{11}\), Abi Pate\({}^{11}\), Mark N Puttick\({}^{12}\),
+Emiliy Rayfield\({}^{11}\), Erin E. Saupe\({}^{13}\), Emma
+Sherratt\({}^{14}\), Graham Slater\({}^{15}\), Vera
+Weisbecker\({}^{1}\), Gavin H Thomas\({}^{16}\) and Philip
+Donoghue\({}^{11}\)\\
+\({}^{1}\)\emph{School of Biological Sciences, University of Queensland,
+St. Lucia, Queensland, Australia.}\\
+\({}^{2}\)\emph{Department of Life Sciences, Natural History Museum,
+London, Cromwell Road, London, SW7 5BD, UK.}
+
+\emph{\textsuperscript{3}School of GeoSciences, University of Edinburgh,
+Grant Institute, Edinburgh EH9 3FE, UK}
+
+\textsuperscript{4} Department of Biology, University of York, YO10 5DD
+
+\textsuperscript{5} Department of Zoology, School of Natural Sciences,
+Trinity College.
+
+\textsuperscript{6} Institut de Systématique, Évolution, Biodiversité
+(ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne
+Université, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005
+Paris, France.
+
+\textsuperscript{7} Ryan Institute, School of Natural Sciences, National
+University of Ireland Galway, Ireland.
+
+\textsuperscript{10} \emph{School of Earth and Environment, University
+of Leeds, Leeds LS2 9JT, UK}\\
+\textsuperscript{11}\emph{School of Earth Sciences, University of
+Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK}
+
+\textsuperscript{12} \emph{Milner Centre for Evolution, University of
+Bath, BA2 7AY}
+
+\textsuperscript{13} \emph{Department of Earth Sciences, University of
+Oxford, S Parks Road, Oxford OX1 3AN, UK}
+
+\textsuperscript{14} \emph{School of Biological Sciences, The University
+of Adelaide, Adelaide, South Australia 5005, Australia}
 
@@
 
-*\textbf{Corresponding author: }\emph{guillert@tcd.ie}\\
+*\textbf{Corresponding author:} \emph{guillert@tcd.ie}
 
 \textbf{NOTE TO ALL THE CO-AUTHORS:}
-
-\textbf{-Please write in Google docs using }\emph{\textbf{Suggesting
-mode}}\textbf{ only }(upper right corner of your screen)
 
@@
 
--I suggest you add your references using the \textbf{Paperpile} addon
-(go in Add-ons menu \textgreater{} Get Add-ons\ldots{} \textgreater{}
-Search for Paperpile \textgreater{} Add/Install; then go in Add-ons
-menus \textgreater{} Paperpile \textgreater{} Manage citation). If you
-don't use Paperpile, please don't add references as just (``Someone et
-al. 2000'' but add also a link to the correct Google Scholar entry in
-comments).
-
 \textbf{-Add your affiliation in the author list above}. The
 affiliations count in the word count so please 1) keep it short (max 1
-per person!) and 2) try to use one that's already in there. Also add any
+per person!) and 2) try to use one that's already there. Also add any
 initials I've missed in your names.
@@
 - If in doubt, check formatting here:
-\href{https://royalsociety.org/journals/authors/author-guidelines/}{\emph{https://royalsociety.org/journals/authors/author-guidelines/}}
+\href{https://royalsociety.org/journals/authors/author-guidelines/}{{https://royalsociety.org/journals/authors/author-guidelines/}}
 
@@
 
-Morphological disparity analysis is used to characterize and investigate
-discrete patterns of variation in the shape, form, function, and the
-ecology of organisms. Since their conception in the 1980s, these methods
-have been used to ask a variety of questions about how and why
-morphological diversity evolves, and the ecological consequences of
-these changes. There are many methodological approaches for studying
-disparity, however, all of these come with caveats and pitfalls that
-can, if ignored, lead to misinterpretations of disparity patterns. Here
-we review morphological disparity methods and their caveats, and provide
-best practice guidelines for their application and interpretation. We
-discuss the data required for disparity studies, as well as
-methodological challenges, including multidimensionality, ordination,
-metric choice, hypothesis testing, and ancestral state estimations.
-There is no ``one-size-fits-all'' approach to characterizing and
-investigating the evolution of morphological disparity, and the
-available tools should always be used in the context of a specific
-biological question that will determine data and method selection at
-every stage of the analysis.
-
-\textbf{Keywords}: multidimensionality, palaeobiology, ecology\\
-
-1. Introduction
-
-Variation in the shape, form and function of organisms often exhibits a
-pattern of discrete clusters rather than continuous gradients. These
-patterns can be pictured as multidimensional spaces encompassing all the
-possible forms an organism can take, and where each individual region of
-this space is the result of a unique combination of many variables. One
-commonly used approach to analyse these patterns and to understand the
-processes that lead to them, is to analyse the morphological diversity
-of a group, often referred to as a ``morphological disparity analysis''
-\href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline et al. 2018)}.
-
-The initial informal definition of morphological disparity was
-``multidimensional morphological dissimilarity at a macroevolutionary
-scale'' (Runnegar 1987,
-\href{https://paperpile.com/c/sTGYvp/Uns3}{(Gould 2000)}). In the
-seminal disparity papers of the late 1980s, disparity was used to
-investigate body plan evolution after the Cambrian explosion
-\href{https://paperpile.com/c/sTGYvp/CiPy}{(Briggs, Fortey, and Wills
-1992)}\href{https://paperpile.com/c/sTGYvp/CidX+6tNm+oenu+TtGs}{(Gould
-1991; Valentine 1986; M. Foote and Gould 1992; Morris 1989)};
-\href{https://paperpile.com/c/sTGYvp/CiPy}{(Briggs, Fortey, and Wills
-1992)};. Early disparity analyses focused on broad questions such as:
-How does diversity evolve? Are some phenotypes more common than others?
-and Can anatomical shapes evolve in all ``directions'' of the
-multidimensional space
-\href{https://paperpile.com/c/sTGYvp/yqPw+fTJ3}{(Mike Foote 1997,
-1995)}? Since the 1990s, a great deal of literature from a range of
-sub-fields in biology and palaeontology, and an equally large number of
-methods, have used ``disparity'' as a broader umbrella term for studies
-of multivariate variation. Although most influential reviews have
-treated disparity as a pattern
-\href{https://paperpile.com/c/sTGYvp/yqPw+nFf7+vTHS}{(Mike Foote 1997;
-Wills 2001; Melanie J. Hopkins and Gerber 2017)}, there is currently no
-consensus on whether to restrict use of the term disparity to its
-original macroevolutionary-scale definition, or whether it can refer to
-any study of multivariate variation. Here we use a broader definition
-and suggest that morphological disparity is a \emph{pattern of variation
-}in phenotype. Disparity analysis has been used to tackle a vast array
-of questions. We identify four main classes of disparity analysis as
-follows.
-
-(1) \emph{Descriptive disparity} \emph{analyses. }The seminal studies of
-disparity described the shapes of organisms and how they differed among
-groups \href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Mike Foote 1995;
-Wills, Briggs, and Fortey 1994)}. These consist of descriptive
-multidimensionality analyses of the patterns in the diversity of
-morphological traits, addressing questions such as: why are some
-morphological traits combinations more common than others and what are
-the properties of the resulting morphospace?
-\href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Mike Foote 1995;
-Raup 1961; Gerber 2017)}.
-
-(2) \emph{Disparity-through-time} \emph{analyses. }This approach
-investigates how the shapes of organisms have changed through time,
-focussing on the disparity of taxa in particular time bins or slices.
-This approach has been widely used in palaeobiology to answer a range of
-macroevolutionary questions
-\href{https://paperpile.com/c/sTGYvp/PbSx+khc9+xxh5}{(Close et al. 2015;
-Cisneros and Ruta 2010; Hughes, Gerber, and Wills 2013)}, such as: how
-does disparity accumulate over the history of a clade
+Statistical nalyses of morphological disparity have been used to
+characterise and investigate the evolution of variation in the anatomy,
+function, and ecology of organisms since the 1980s. While a diversity of
+methods have been employed, it is unclear whether they provide
+equivalent insights. Here we review the most commonly used approaches
+for characterising and analysing morphological disparity, all of which
+have associated limitations that, if ignored, can lead to
+misinterpretation. We provide best practice guidelines for disparity
+analyses, while noting that there can be no ``one-size-fits-all''
+approach. The available tools should always be used in the context of a
+specific biological question that will determine data and method
+selection at every stage of the analysis.
+
+\textbf{Keywords}: multidimensionality, palaeobiology, ecology,
+morphology, disparity, variance/variation
+
+\hypertarget{introduction}{%
+\section{1. Introduction}\label{introduction}}
+
+Clades of organisms are characterised by variation in both numbers of
+species and range of phenotypes through time. At the extremes, clades
+may be exceptionally species-rich and phenotypically diverse (e.g.
+cichlids and molluscs), species-rich but phenotypically conservative
+(e.g. bacteria and nematodes), species-poor but phenotypically diverse
+(e.g. Afrotheria), or depauperate in both species and phenotypic
+diversity (e.g. lungfish). These phenomena suggest that taxonomic and
+phenotypic diversity are not inextricably linked, raising important
+questions about how phenotypic diversity arises, such as: How does
+diversity evolve? Are some morphologies more common than others? Can
+anatomy evolve in all ``directions'' or are some anatomies impossible to
+achieve? (Mike Foote 1997, 1995) What role does ecology play in
+structuring morphological diversity?. Analyses of species diversity have
+a venerable history, but those of phenotypic diversity (hereafter
+\emph{disparity}) are a comparatively more recent phenomenon. Originally
+defined as ``multidimensional morphological dissimilarity at a
+macroevolutionary scale'' (Runnegar 1987, Gould
+\href{https://paperpile.com/c/sTGYvp/Uns3}{2000}), this concept of
+disparity emerged from attempts by palaeobiologists to characterise the
+evolutionary origin of animal bodyplans and from attempts by comparative
+developmental biologists to provide causal explanations for their
+emergence. However, analyses of ``disparity'' have since expanded into
+comparative biology as a means of capturing the effect of intrinsic and
+extrinsic causal agents in morphological evolution. Typically, methods
+to capture disparity are based on multidimensional spaces where each
+dimension represents an aspect of morphological variation (a trait) and
+biological observations (taxa) can be placed in this space based on
+their trait values. Such multidimensional spaces (or morphospaces) can
+then be used to tackle a diverse array of questions that can be grouped
+into four main (non-mutually exclusive) classes (Fig. 1):
+
+(1) \emph{Descriptive disparity.} The pioneering studies of disparity
+characterised the shapes of organisms and how they differed among groups
+\href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Foote 1995; Wills,
+Briggs, and Fortey 1994)}. These studies consist of describing
+multidimensional patterns in the diversity of morphological traits,
+addressing questions such as: why are some morphological trait
+combinations more common than others and what are the biological (or
+mathematical) properties of the resulting morphospace?
+\href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Foote 1995; Raup
+1961; Gerber 2017)}.
+
+(2) \emph{Disparity-through-time.} This approach investigates how the
+morphologies of organisms have changed on a temporal scale, focussing on
+the disparity of taxa in particular time intervals or slices. This
+approach has been used widely in palaeobiology to answer a range of
+macroevolutionary questions, such as: how does disparity accumulate over
+the history of a clade
 \href{https://paperpile.com/c/sTGYvp/ekU4+s33b}{(Guillerme and Cooper
-2018; Wright 2017)}, and how does disparity change across mass
-extinction events \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman
-2010)}?
-
-(3) \emph{Disparity and diversity. }Morphological disparity provides
-another perspective on biodiversity; high morphological disparity
-represents a high diversity of morphologies (i.e. shapes or body plans)
-and is, presumably, associated with high levels of ecological and
-functional diversity (see below). This makes disparity a more useful
-measure of diversity than merely species richness. Indeed, most studies
-that have investigated disparity and taxonomic diversity support an
-effective decoupling
-\href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(Fortey,
+2018; Wright 2017)}, or how does disparity change leading up to and
+across mass extinction events
+\href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}?
+
+(3) \emph{Disparity and taxonomic diversity.} Morphological disparity
+provides another perspective on biodiversity; high morphological
+disparity represents a high diversity of morphologies (i.e. shapes or
+body plans) and is, presumably, associated with high levels of
+ecological and functional diversity (though see below). This makes
+disparity an informative complement to diversity measures based on
+species richness alone. Indeed, most studies that have investigated
+disparity and taxonomic diversity support an effective decoupling
+\href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(e.g., Fortey,
 Briggs, and Wills 1996; Ruta et al. 2013; M. J. Hopkins 2013; Moyne and
-Neige 2007)} - though this result breaks down at extremes of diversity;
-extremely low diversity tends to yield low disparity and \emph{vice
-versa}. This approach has been used to investigate whether some groups
-are more successful than other in their exploration of new evolutionary
-strategies @@@.
-
-(4) \emph{Disparity and ecology. }The disparity of a group can be used
-as a proxy for either the functional role it plays within an ecosystem
-or its ecological niche. These analyses assume that groups with high
-disparity are also likely to be functionally and ecologically diverse,
-and that groups found in similar regions of shape space will have
-similar functional and ecological roles
+Neige 2007)}. This approach has been used to investigate whether some
+groups are more successful than others in their exploration of new
+evolutionary strategies
+\href{https://paperpile.com/c/sTGYvp/dJHu}{(Losos 2011)}.
+
+(4) \emph{Disparity as a proxy for ecology.} The disparity of a group
+can be used as a proxy for either the functional role it plays within an
+ecosystem or its ecological niche. This approach assumes that groups
+with high disparity are also likely to be functionally and ecologically
+diverse, and that groups found in similar regions of shape space will
+have similar functional and ecological roles
 \href{https://paperpile.com/c/sTGYvp/EETc+tSIy+qjj9}{(Friedman 2010;
 Pierce, Angielczyk, and Rayfield 2008; P. S. L. Anderson et al. 2011)}.
-Note however, that the links between form and function are not always
-clear cut; traits can be linked to multiple functions and multiple
-functions can be linked to a single trait
-\href{https://paperpile.com/c/sTGYvp/Ejzr}{\emph{(Wainwright et al.
-2005)}}. This approach is used to investigate competitive replacements
-\href{https://paperpile.com/c/sTGYvp/EeC8}{(Stephen L. Brusatte et al.
-2008)} and changes in ecosystem function during and after mass
-extinctions \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}.
-It is particularly common in paleobiology where it is not possible to
-observe directly the ecological or functional characteristics of extinct
-species. \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.
-2005)}
-
-Of course, these four broad categories are not independent and many
-studies use a combination of approaches
-\href{https://paperpile.com/c/sTGYvp/EeC8}{(Stephen L. Brusatte et al.
-2008)}. Furthermore, disparity analysis is continuously expanding into
-areas. While this is exciting for the field, there are several issues
-that may reduce the utility of disparity analyses if they are not
-tackled. Here we review these issues and present best practice
-guidelines to allow researchers to optimise their analyses. We first
-discuss the appropriate data required for disparity analyses, then
-review various challenging methodological aspects of disparity analyses.
-Throughout it is important to remember that these tools should always be
-used in the context of a specific scientific question, as this will
-drive data and methodological choices at every stage of the process.
-
-\includegraphics[width=6.5in,height=4.9445in]{Pictures/10000201000002EE0000023BE5B78D4C4B79F709.png}
+The links between form and function, however, are not always clear cut.
+Traits can be linked to multiple functions and multiple functions can be
+linked to a single trait
+\href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al. 2005)}.
+This approach has been used to investigate hypotheses of competitive
+replacement (Stephen L. Brusatte et al. 2008) and changes in ecosystem
+function during and after mass extinctions
+\href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}. It is
+particularly common in palaeobiology where it is not possible to
+directly observe the ecological or functional characteristics of extinct
+species \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.}
+2005\href{https://paperpile.com/c/sTGYvp/Ejzr}{)}.
+
+(5) Disparity and evo-devo. Morphological disparity blalbalbal
+
+Fundamental insights into evolutionary biology have been elicited from
+these four types of analysis. One of the most important insights is the
+discovery that morphological disparity is often greatest early in the
+evolutionary history of clades
+\href{https://paperpile.com/c/sTGYvp/xxh5+yqPw}{(Hughes, Gerber, and
+Wills 2013; Foote 1997;}
+\href{https://paperpile.com/c/sTGYvp/EPJ2}{Erwin 2007)}, indicating that
+the capacity for evolutionary innovation wanes with clade age, which
+some have argued reflects the evolutionary assembly of gene regulatory
+networks that constrain later fundamental change
+\href{https://paperpile.com/c/sTGYvp/xxh5+Z6l6}{(Hughes, Gerber, and
+Wills 2013; Erwin 2011)}. However, this example highlights one of the
+most challenging problems confronting researchers who are attempting,
+increasingly, to obtain general insights from multiple independent
+studies: can the insights gained from studies using the diversity of
+methods, approaches and data types employed be considered equivalent?
+
+In attempting to answer this question, we review current methods and
+highlight their limitations, as part of a more general attempt to
+propose best practice guidelines for studies of disparity . We first
+discuss the appropriate data required for characterising disparity, then
+review various challenging aspects of these approaches. Throughout, it
+is important to remember these tools should always be used in the
+context of a specific scientific question, as this will drive data and
+methodological choices at every stage of the process.
+
+\includegraphics[width=6.5in,height=4.94444in]{media/image1.png}
 
 \emph{Figure disparities {[}figure\_disparities.pdf{]}: The four main
-types of disparity analysis. Descriptive disparity analyses focus on
-describing the properties of the morphospace ; Disparity-through-time
-analyses investigate the evolution of the morphospace through time
-including the effect of extinction events; Disparity and diversity
-analysis use disparity as a proxy for species diversity; Disparity and
-ecology analyses use disparity as a proxy for the ecological or
-functional role of a group. These categories are not independent and
-many analyses will cover more than one.}
-
-2. Data in Disparity Analyses
-
-Regardless of the precise approach being used, disparity analyses are
-based on morphological traits. It is important, therefore, to consider
-which traits to use in the context of the specific question under
-investigation. Traits for disparity analyses come from three main types
-of data: (1) discrete ``cladistic'' characters, i.e coding the
-absence-or-presence of features or a discrete characteristic of a trait
-\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}); (2)
-continuous measurements of features
+types of disparity analysis. \textbf{Descriptive disparity} focuses on
+describing the features of morphospace occupation;
+\textbf{disparity-through-time} investigates the evolution of the
+morphospace through time including the effect of extinction events;
+\textbf{disparity and taxonomic diversity} compares different measures
+of biodiversity ; \textbf{disparity as a proxy for ecology} uses
+disparity as a proxy for the ecological or functional role of a group.
+These categories are not independent and many disparity studies will
+cover more than one.}
+
+\hypertarget{data-and-disparity}{%
+\section{2. Data and disparity }\label{data-and-disparity}}
+
+Disparity analyses are based on traits, but traits can be characterised
+in a number of ways: (1) discrete ``cladistic'' characters, e.g. coding
+the absence or presence of features or a discrete characteristic of a
+trait \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)};
+(2) continuous measurements of features (e.g., lengths)
 \href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al.
-2011)}; or (3) more sophisticated mathematical measurements from
-geometric morphometric data (e.g. Procrustes superimpositions
-\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, or
-Fourier contours \href{https://paperpile.com/c/sTGYvp/ZEDR}{(Spriggs et
-al. 2018)}). None of these are fundamentally better, however, some
-questions and methodological pipelines are more appropriate with
-specific types of data.
-
-When investigating variation within bat wing shapes, for example, both
-homologous landmarks from geometric morphometric analysis and continuous
-measurements of bones may be appropriate to capture patterns of wing
-variation. However, if the question focuses on the convergence between
-bat and bird wings, the homology of the traits to measure is more
-complex to define and the outline shape of the wings might more readily
-answer the question from a mechanical perspective. Similarly, if the
-focus is on wing function, i.e. whether the aerodynamic properties of
-wings vary within bats or between bats and birds, the traits collected
-should reflect these aerodynamic properties (e.g. wingspan, aspect
-ratio, etc.). Whereas if the focus is on convergence between different
-bats and birds, it would be preferable to use traits that have
-facilitated flight in both groups (e.g. digit length, integumentary
-system, etc.). Where there is any doubt about the appropriate traits to
-choose, we advise using different kinds of data for the same feature to
-determine whether it captures the same disparity.
-
-This dicussion assumes that researchers are collecting their own data
-for disparity analyses, but often this is not the case. It is common,
-especially in disparity-through-time analyses, to recycle discrete
-character data collected for use in phylogenetic inference (e.g. REFs
-that have done this). This has the advantage of being quick, cheap and
-efficient, but tends to artificially increase disparity between
-phylogenetically distinct groups because phylogenetic characters are
-collected specifically for the purpose of discriminating groups
-\href{https://paperpile.com/c/sTGYvp/fTJ3}{(Mike Foote 1995)}. This
-needs to be considered when interpreting results, especially as when new
-clades appear, the synapomorphies of that clade will naturally lead to
-an apparent shift or increase in disparity. Similarly, data used for
-phylogenetic reconstruction is likely to be highly skewed towards easily
-available and taxonomically relevant traits, for example teeth in
-mammals. If your disparity question is related to feeding ecology or
-similar then this may be appropriate, but if the focus is on dispersal
-ability then it would be sensible to select only postcranial traits. It
-is often tempting to use all of a large dataset to increase the power of
-your analyses, but if the data are not suitable for the question at hand
-then the results may be meaningless (although ssome studies have suggest
-this may not be a major issue;
-\href{https://paperpile.com/c/sTGYvp/ZqML+Qrba}{(Zou and Zhang 2016;
-Dávalos et al. 2014)}
-\href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}).
-
-Finally, trait data suffers from the same shortcomings as most data --
-data can be missing, non-overlapping, hierarchical, inapplicable,
-ambiguous, polymorphic, correlated, or have insufficient sample size
-\href{https://paperpile.com/c/sTGYvp/yO2t}{(Palci and Lee 2018)},
-\href{https://paperpile.com/c/sTGYvp/Yrbg}{(Brazeau, Guillerme, and
-Smith 2017)} {[}String of morpho data problems cite{]}. Biological
-phenomena such as allometry and sexual dimorphism may also influence
-trait data. More practically, data collection is constrained by the time
-and money available, making collating a ``perfect'' dataset difficult.
-We suggest that trait data are collected with the question in mind and
-considering all of the caveats above, but where issues are unavoidable
-at the very least these caveats must be considered during the analyses
-and subsequent interpretation of the results.
-
-\emph{Figure data {[}figure\_data.pdf{]}: major routes to obtain
-disparity data. Data can be collected as discrete trait observations
-(e.g. presence or absence data) or as continuous data. Continuous data
-can be collected by different methods: linear measurements, landmark
-coordinates or contours (curves). These measurement can then in turn be
-mathematically transformed (logarithm, scaling, Procrustes
-Superimposition, Elliptic Fourier, etc.). Regardless of the method, this
-data collection results in a trait matrix where the observed traits
-constitute columns and the studied elements (e.g. taxa) the rows.}
-
-\includegraphics[width=6.5in,height=6.5in]{Pictures/10000201000007D0000007D0F8A72A6590C667D3.png}
-
-3. Disparity Methods
-
-Once suitable trait data have been collected, we need to consider the
-disparity analysis. Below we address several key aspects including (i)
-the difficulty of dealing with multidimensional data; (ii) caveats
-associated with using ordination techniques to reduce the dimensionality
-of the data; (iii) various metrics used to summarise the disparity of a
-group; (iv) caveats related to methods that exist for hypothesis testing
-within a disparity analysis framework; and (v) the specific case of
-ancestral state estimations in disparity-through-time analyses. As noted
-above, these tools should always be used in the context of a specific
-question, as this will determine which methods are most appropriate.
-
-\protect\hypertarget{anchor}{}{}3.1. Multidimensional mayhem
-
-The multidimensional aspect of disparity analyses adds a layer of
-technical complexity because the human brain naturally struggles to
-comprehend more than three spatial dimensions. This problem is often
-apparent in how people interpret the results of disparity analyses. For
-example, it is standard for disparity results to use ordination
-techniques to reduce their dimensionality, and for only the first or
-second axes of ordination (e.g. principal components (PCs) one and two)
-to be interpreted and used to draw conclusions. Changes in disparity
-along these axes may show a pattern or gradient, but often do not
-reflect the true changes across all the dimensions of the data
-\href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(Bookstein
-1997, 2015, {[}b{]} 2017, {[}a{]}
-20}\href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1}\href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{7)}{[}CITE
-Vera and Thomas paper{]}. Visual interpretations of multidimensional
-data can also easily be misleading. For example, an obvious
-discontinuity (or overlap) between two groups in a morphospace of PC1
-and PC2, may not be apparent when looking at the same groups in a
-morphospace of PC2 and PC3 etc. or across the whole multidimensional
-space. Furthermore, multidimensional spaces are not necessary Euclidean.
-Such non-Euclidean spaces often have non-intuitive properties, for
-example straight lines in some dimensions are not straight in
-non-Euclidean space and, even less intuitively, the distance between
-points A and B might not be equal to the distance between points B and A
-\href{https://paperpile.com/c/sTGYvp/SJbC}{(Gerber 2014)}.
-
-3.2. To ordinate or not to ordinate; that is the (multidimensional)
-question?
-
-Reducing the dimensionality of a disparity dataset can be tricky.
-Ordination is advantageous for plotting and visualising the data, and
-can reveal properties of the morphospace not captured by disparity
-metrics (see Metrics section {[}metrics{]}). Additionally, after
-ordinating the data it is possible to further reduce the number of
-dimensions by only selecting a certain number of axes, for example by
-selecting only the axes that describe up to 95\% of the variation. In
-the case of geometric morphometric data, ordination is particularly
-useful as it conserves the mathematical properties of the data while
-efficiently reducing the dimensions
-\href{https://paperpile.com/c/sTGYvp/oFiP}{(Legendre and Legendre
-2012)}. This has clear advantages for interpreting the results. For
-example, the axes will represent gradients of biological variation (e.g.
-elongation and flattening of the beak
-\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}).
-
-However, various caveats can arise from such transformations of the
-data. For example, in the case of an ordination in a geometric
-morphometric context, not using all the axes from the ordination can
-lead to misinterpretation of the results. Furthermore, interpreting
-biological variation along the axes is always a \emph{post-hoc}
-procedure and may have little relation to the disparity question, for
-example, if the first few ordination axes represent the elongation of
-the beak in birds, but the question is about wing disparity.
-
-In many cases, ordination might not be necessary. For example, if the
-disparity metric being used relies on all the data (see Metrics section
-{[}metrics{]}) it is not necessary to calculate it on ordinated data
-{[}e.g.\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
-2015)}{]}. Additionally, in some cases, reducing the dimensionality of a
-dataset can render its interpretation more problematic. For example,
-when the analysed data is non-Euclidean (e.g. discrete morphological
-characters), ordination can be difficult, and the resulting
-non-Euclidean ordinated space can be hard to interpret
-\href{https://paperpile.com/c/sTGYvp/SJbC}{(Gerber 2014)}. This is
-problematic when comparing the position of groups in the
-multidimensional space, as the measured distances might not be
-comparable. Finally, the \emph{post-hoc} interpretation of the gradient
-of variation on the ordination axes can be impossible or biologically
-meaningless. Although some gradients are easy to detect or interpret
-(e.g. the elongation of the beak along the first axis; cooney paper),
-some are not. For example, with discrete morphological data, a gradient
-between the species that have many characters in state 1 and the ones
-that have more in state 0 has no biological meaning. Because of these
-caveats, we strongly recommend that data for multidimensional analyses
-is not automatically ordinated, and that careful consideration is given
-to whether the question can be answered without ordination.
-
-\includegraphics[width=6.5in,height=5.7638in]{Pictures/10000201000002730000022C18503E8C16A21886.png}
-
-\emph{Figure methods: Different ways to define and visualise a
-morphospace: A.0: The morphospace is the collected trait space (with
-some potential modification - see data section above): this can be
-visualised using bi-plots (or correlation plots - B.3) between all the
-traits but becomes hard to interpret after a few traits. Another
-approach is to directly ordinate the trait matrix (with a PCA - A.1 -
-e.g. in }\href{https://paperpile.com/c/sTGYvp/ZQS9}{\emph{(Zelditch,
-Swiderski, and David Sheets 2012)}}\emph{- or a PCO - A.2; e.g. in
-Brusatte et al. 2008). This results in a k (\textless{}m) space that can
-also be hard to visualise even though the first axis hold most of the
-variation (B.1). Another approach is to create a distance matrix (A.3;
-e.g. in Close et al., 2015) this matrix can then be ordinated as well
-for further analysis (A.2 and used as the ordinated morphospaces
-described above) or ``flattened'' to a manageable number of dimensions
-(2 or 3 - B.2).}
-
-\protect\hypertarget{anchor-1}{}{}3.3. Disparity metrics
-
-Because disparity is multidimensional, a large component of any
-disparity analysis involves considering how to extract a meaningful (and
-interpretable) summary of disparity, i.e. choosing which disparity
-metric (or index)\href{https://paperpile.com/c/sTGYvp/vTHS}{(Melanie J.
-Hopkins and Gerber 2017)} to use. As with any summary metric, and
-because of this reduction in dimensionality, disparity metrics will only
-reflect some aspects of the morphospace, never its whole complexity. It
-is therefore often beneficial to use more than one metric to summarise
-different aspects of interest. The choice of disparity metric is
-essential to the disparity analysis and should always be driven by the
-question at hand.
+2011)}; or (3) more mathematical descriptors from geometric morphometric
+landmark data (e.g. Procrustes coordinates)
+\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, and
+Fourier coefficients \href{https://paperpile.com/c/sTGYvp/2Neu}{(Foote
+1989;} \href{https://paperpile.com/c/sTGYvp/ZEDR}{Spriggs et al. 2018)})
+(Fig. 2). None of these approaches are superior, but they may be more or
+less well-suited to characterising traits under comparison and to the
+question being asked of those traits (Hetherington et al 2015; Hopkins
+2017).
+
+For example, if investigating variation of bat wing shapes, both
+homologous landmarks and continuous measurements of bones may be
+appropriate to capture patterns of wing variation. However, if the
+question focuses on comparing wings between bats and birds, different
+measurements might be more appropriate depending on the specific
+question: for example if the focus is on wing function, i.e. whether the
+aerodynamic properties of wings vary within bats or between bats and
+birds, the traits collected should reflect these aerodynamic properties
+(e.g. wingspan, aspect ratio, etc.). However, if the focus is on
+convergence between different bats and birds, it would be preferable to
+use traits that have facilitated flight in both groups (e.g. digit
+length, integumentary system, etc.). Where there is any doubt about the
+appropriate traits to choose, it may be preferable to use different
+kinds of data for the same feature to determine whether they capture the
+same pattern of disparity.
+
+The points above assume that researchers are collecting their own data
+for disparity analyses, but often this is not the case. Discrete
+character data are commonly recycled from phylogenetic studies
+(e.g.Foote 1989\href{https://paperpile.com/c/sTGYvp/2Neu+0y4V}{; Deline
+et al. 2018}). This is an efficient approach to character sampling, but
+it may artifactually increase disparity between phylogenetically
+distinct groups because phylogenetic characters are often collected to
+discriminate among groups
+\href{https://paperpile.com/c/sTGYvp/fTJ3}{(Foote 1995)}. This needs to
+be considered when interpreting results, especially as synapomorphies
+will naturally lead to an apparent shift or increase in disparity when
+new clades appear. Furthermore, many datasets are limited to subsets of
+anatomy that are at least implicit samples of overall anatomy, but
+explicit tests of this assumption have shown that different aspects of
+morphology can exhibit different patterns of disparity
+\href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}.
+This effect of anatomical part on disparity patterns can be especially
+challenging when working on datasets where the available data has
+non-random missing anatomical parts, such as the absence of soft tissue
+in the fossil record \href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline
+et al. 2018)}.
+
+Trait data suffer from the same shortcomings as most biological datasets
+-- data can be missing, non-overlapping, hierarchical, inapplicable,
+ambiguous, polymorphic, correlated, or there may be an insufficient
+sample size \href{https://paperpile.com/c/sTGYvp/Yrbg}{(Brazeau,
+Guillerme, and Smith 2017};
+\href{https://paperpile.com/c/sTGYvp/yO2t}{Palci and Lee 2018)}.
+Biological phenomena such as allometry and sexual dimorphism may also
+influence trait data. More mundanely, data collection is constrained by
+the time and money available, making collating a ``perfect'' dataset
+difficult. Ultimately, disparity analyses characterise the data, and
+subsamples of the universe of possible data may not have the power to
+uncover holistic patterns of disparity. Therefore, trait data should be
+collected with the question in mind, or the question asked should be
+tailored to the limits of the data available.
+
+\includegraphics[width=5.63542in,height=3.875in]{media/image2.png}3.
+
+\emph{Figure data {[}figure\_data.pdf{]}: Major routes to obtain
+morphological data for disparity analyses. Data can be collected as
+discrete trait observations (e.g. presence or absence data) or as
+continuous data. Continuous data can be collected by various methods
+including linear measurements and landmark coordinates or contours
+(curves). These measurements can then be mathematically transformed
+(logarithm transforms, scaling, Procrustes superimposition, elliptic
+Fourier transforms, etc.). Regardless of the method, data collection
+produces a trait matrix where the observed traits constitute columns and
+the studied elements (generally taxa or OTUs) the rows.}
+
+\hypertarget{disparity-methods}{%
+\section{Disparity methods}\label{disparity-methods}}
+
+Once suitable trait data have been collected, the design of the
+disparity analysis itself needs to be considered. Study design
+encompasses several key aspects including (a) the difficulty of dealing
+with multidimensional data; (b) the variety of ordination (dimension
+reduction) techniques available and their limitations; (c) the metrics
+used to summarise the relative disparity of groups; (d) the methods used
+for hypothesis testing within the disparity analysis framework; and (e)
+ancestral state estimations in disparity-through-time analyses. We
+consider these aspects in turn below.
+
+\hypertarget{a-multidimensionality}{%
+\subsection{(a) Multidimensionality}\label{a-multidimensionality}}
+
+Disparity analyses often use ordination techniques for dimension reduction.
+These are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g.
+principal component analysis -- PCA; principal coordinates analysis -- PCO/PCoA; non-metric multidimensional scaling -- NMDS; etc.).
+Using ordinations in disparity analysis makes it easier for researchers to comprehend patterns in just two or three spatial dimensions at a time - or, indeed, just the first few axes of a PCA or a PCO.
+Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(}Bookstein \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1997, 2015, {[}b{]} 2017, {[}a{]} 2017)}{[}CITE Vera and Thomas paper{]}.
+%TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
+Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \href{https://paperpile.com/c/sTGYvp/0y4V+QVvv}{(Deline et al.
+2018;
+Gerber 2017)}.
+Categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
+contrary to a PCA, the first few axis do not encompass most variance of the dataset).
+Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric (\href{https://paperpile.com/c/sTGYvp/SJbC}{i.e.
+the distance between A and B is not equal to the distance between B and A; Gerber 2014)}.
+
+\hypertarget{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}{%
+\subsection{(b) To ordinate or not to ordinate - that is the
+(multidimensional)
+question}\label{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}}
+
+Dimension reduction (or ordination) is a powerful tool for visualizing data distribution in disparity analyses, helping to elucidate the underlying structure within datasets.
+However, like most other aspects of disparity analyses, reducing dimensionality can be tricky.
+Ordination techniques come in many flavours depending on the data and the desired morphospace properties.
+Quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCoA, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) (see Legendre and Legendre 2012,chapter 9, for a detailed explanation of ordination methods and properties).
+
+Ordination is advantageous for plotting and visualising data, and can reveal properties of the morphospace not intuitively captured by disparity metrics (see Metrics section below {[}metrics{]}).
+Additionally, after ordinating the data, it is possible to reduce further the number of dimensions by focussing on just a subset of axes (e.g.
+selecting only those axes that describe the majority of the variation) or by fitting the variation to a prescribed number of axes.
+In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions (Legendre and Legendre 2012).
+%TG: cite also: Dryden and Mardia "Statistical Shape Analysis book"
+This has clear advantages for interpreting the results.
+For example, the axes will represent gradients of biological variation (e.g.
+elongation and flattening of the beak in birds \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al.
+2017}).
+However, there are a number of potential problems with these transformations of the data.
+%TG: comment from Sylvain: difference between a morphospace and an ordination space
+%TG: comment from Graeme: the ordination space is decided by the data present (e.g. shoebill)
+In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results.
+%TG: from Anjali: add that people should use 95% or 99% percent of the data (why not 100% though?)
+Furthermore, interpreting biological variation along the axes is always a \emph{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
+
+However, In many cases, ordination might not be necessary.
+For example, if a metric (e.g.
+sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below {[}metrics{]}), it is not necessary to calculate it on ordinated data {[}e.g.\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
+2015)}{]}.
+Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
+For example, when the analysed data is non-Euclidean (e.g.
+some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \href{https://paperpile.com/c/sTGYvp/SJbC}{(}Gerber 2014\href{https://paperpile.com/c/sTGYvp/SJbC}{)}.
+%TG: should be Gerber 2019
+This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear.
+Finally, \emph{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible. %TG: Cite Gerber 2019 10.1111/pala.12407
+Although some gradients are easy to detect or interpret (e.g.
+the elongation and depth of mandibles in fishes on first and second PC axes, respectively; {[}\href{https://paperpile.com/c/sTGYvp/3JPy}{Hill et al.
+2018{]}}), some are not (e.g.
+\href{https://paperpile.com/c/sTGYvp/TZzO}{Weisbecker et al.
+2019)}.
+For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0, has no biological meaning if these are binary alternate states.
+For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination (Lloyd 2016, 2018).
+
+\includegraphics[width=6.5in,height=5.26389in]{media/image3.png}
+
+\emph{Figure methods: Morphospaces: different mathematical
+representations of a} \emph{morphospace. A trait matrix can be
+transformed into a distance matrix (e.g. in Close et al., 2015) or an
+ordinated matrix (e.g. in Brusatte et al.} \emph{2008). In each case,
+all three matrices represent the morphospace for the observations at
+hand. Visualisation: different ways to represent the morphospace in 2D.
+Visualisations can use either trait plots (directly from the trait
+matrix); ``flattening'' ordination of the data (e.g. using
+multidimensional scaling} \emph{forcing all the variance to be contained
+in two dimensions); or ordination axis plots (directly from the
+ordinated matrix).}
+
+\hypertarget{c-disparity-metrics}{%
+\subsection{(c) Disparity metrics}\label{c-disparity-metrics}}
+
+Most disparity datasets are multidimensional and, consequently, a large
+component of any disparity analysis involves considering how to extract
+a meaningful (i.e. interpretable) summary of disparity. This is usually
+achieved with a disparity metric or index
+\href{https://paperpile.com/c/sTGYvp/vTHS}{(Melanie J. Hopkins and
+Gerber 2017)}. As with any summary of multidimensional data, disparity
+metrics will reflect only some aspects of the morphological variation,
+never its whole complexity {[}metrics paper{]}. It is therefore often
+beneficial to use more than one metric to summarise different aspects of
+variation, guided by the aim of the study.
 
 When considering only one dimension, disparity metrics can be used to
-reflect the ``size'' of the distribution (e.g. the range, quantiles or
-variance), or the most likely value (i.e. the central tendency, mean,
-median or mode). Among these metrics, some will have more attractive
-properties than others, such as sensitivity to outliers -- range, mean
-and mode are highly sensitive whereas quantiles, variance and median are
-less so -- and will thus be more or less appropriate for different
-questions. For example, if the question refers to ``size'' of a group in
-the morphospace (e.g. does group A occupy as much space as group B?),
-metrics related to the spread of the group in the morphospace are the
-most appropriate (e.g. volume
+reflect the spread of the distribution (e.g. the range, quantiles or
+variance) or its central tendency (i.e., mean, median or mode). Among
+these metrics, some will have more attractive properties than others,
+such as sensitivity to outliers. Range and mean are highly sensitive,
+whereas quantiles, variance and median are less so, making them more or
+less appropriate for different questions. For example, if we want to
+characterise the extent of morphospace occupied by a group (e.g. does
+group A occupy as much space as group B?), metrics related to the spread
+of the group in the morphospace are most appropriate (e.g. volume
 \href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, distance
@@
 \href{https://paperpile.com/c/sTGYvp/tGyd}{(S. L. Brusatte et al.
-2008)}). Conversely, if the question refers to the ``position'' of a
-group in the morphospace (e.g. does group A occupy the same space as
-group B?), metrics related to the distance between the elements within a
-group and a fixed point in the morphospace are most appropriate (e.g.
-????). Finally, if the question relates to the density of elements in
-some regions of the morphospace (e.g. is group A more dense than group
-B?), metrics related to the pairwise distances between elements will be
-most appropriate (e.g. nearest neighbour distance, pairwise distances,
-etc. \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}).
-
-In addition to considering what properties of disparity the metrics
-should capture, it is important to also consider the mathematical
-properties of the metrics and their associated caveats
+2008)}). Conversely, if we wish to describe the ``position'' of a group
+in a morphospace (e.g. does group A occupy the same space as group B?),
+metrics related to the distance between the elements within a group and
+a fixed point in the morphospace are most appropriate
+(e.g.\href{https://paperpile.com/c/sTGYvp/u1KE}{(Guillerme et al.,
+n.d.)}). Finally, if we aim to characterise the density of morphospace
+occupation (e.g. is group A more closely packed than group B?) metrics
+related to the pairwise distances between elements will be most
+appropriate (e.g. nearest neighbour distance, pairwise distances, etc.
+\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}).
+
+In addition to considering what properties of disparity metrics should
+capture, it is important to also consider the mathematical properties of
+the metrics and their associated caveats
 \href{https://paperpile.com/c/sTGYvp/nFf7+ROH8}{(Wills 2001; Ciampaglio,
 Kemp, and McShea 2001)}. For example, measuring the full sum of the
-variance of each dimension of the space does not require one to add the
-covariance between the axes in a ordinated space using a principal
-components analysis (PCA). This is not true however, in other
-mathematical spaces or when not all dimensions or elements are
-considered, even in a PCA. More simply, when looking at a group within
-morphospace, the sum of variances is not mathematically equal to the sum
-of the variance of each dimension in the space. Further,
-multidimensional space has some counter-intuitive properties that need
-to be considered such as the ``curse of dimensionality''
-\href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}.
-Product-based metrics used as proxies of volumes (e.g. product of
-ranges, hypervolume, hypercube, etc.) will tend towards zero fairly
-quickly for spaces with even a modest number of dimensions
-{[}\href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}; Donoho
-2000{]}. Some other types of metrics are also extremely sensitive to
-outliers and can be easily biased by sample size, for example range
-\href{https://paperpile.com/c/sTGYvp/aSSL}{(Butler et al. 2012)} or
-convex hull based metrics
-\href{https://paperpile.com/c/sTGYvp/aSSL+PwyQ}{(Butler et al. 2012;
-Jackson et al. 2011)}.
-
-\protect\hypertarget{anchor-2}{}{}3.4. Testing disparity hypotheses
-
-As mentioned above, disparity analysis should always be used in the
-context of a specific question. Once disparity metrics have been
-calculated, the associated question must be tested in an appropriate
-statistical framework. The multidimensional statistical toolkit for
-ecology and evolution has been greatly expanded in recent years
-\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)} but
-is not routinely applied to disparity analyses. Instead, disparity
-hypothesis testing is mostly confined to a small set of well-established
-methods. One commonly used test is the non-parametric permutation
-analysis of variance \href{https://paperpile.com/c/sTGYvp/3hy2+SC6L}{(M.
-J. Anderson and Walsh 2013; M. J. Anderson 2001)}. This test is an
-analysis of variance (ANOVA) of the pairwise distances between different
-groups. Although statistically valid, this test is sometimes not
-directly related to the hypothesis under test. For example, PERMANOVA
-tests whether two groups share the same variance/covariance in a
-``distance-space''. This is not the same as testing whether the two
-groups overlap in morphospace. Explicitly stating the hypothesis being
-tested can help with understanding which statistical test to apply for a
-specific question.
-
-It is also important to consider which data to apply a statistical test
-on. For example, in morphological disparity analysis, especially for
-palaeobiological questions, data are often bootstrapped. This has two
-advantages: first, when the disparity metric is unidimensional, e.g. the
-sum of variances, bootstrapping the data generates a distribution of the
-metric that can then be analysed using the vast statistical toolkit
-available for comparing distributions; second, when data are scarce,
-bootstrapping the data allows users to introduce variance, rendering the
-test less sensitive to outliers. However, bootstrapped data are
-pseudoreplicates and thus non-independent. This violates the assumptions
-of most parametric statistical tests. Furthermore, the number of
-bootstrap pseudoreplicates will inevitably increase the Type I error
-rate. These factors are often ignored in disparity analyses.
-
-3.5. Phylogenetic autocorrelation
+variance in each dimension of the space does not require we add the
+covariance between the axes in an ordinated space using a PCA. However,
+this is not true of other mathematical spaces or when not all dimensions
+or elements are considered, even in a PCA {[}Legendre \& Legendre
+2012{]}.
+
+Furthermore, multidimensional spaces also have some counter-intuitive
+properties that need to be considered such as the ``curse of
+dimensionality'' \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman
+1966)}. In spaces with some axis of variance lower than 1, product-based
+metrics used as proxies of volumes (e.g. product of ranges, hypervolume,
+hypercube, etc.) can tend towards zero fairly quickly for spaces with
+even a modest number of dimensions
+\href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}; Donoho
+2000). Some other types of metrics are also extremely sensitive to
+outliers and can be biased easily by sample size, for example range
+\href{https://paperpile.com/c/sTGYvp/aSSL}{(}Butler et al.
+2012\href{https://paperpile.com/c/sTGYvp/aSSL}{)} or convex hull based
+metrics \href{https://paperpile.com/c/sTGYvp/aSSL+PwyQ}{(Butler et al.
+2012; Jackson et al. 2011)}.
+
+\hypertarget{d-testing-hypotheses-on-the-evolution-of-disparity}{%
+\subsection{(d) Testing hypotheses on the evolution of
+disparity}\label{d-testing-hypotheses-on-the-evolution-of-disparity}}
+
+No matter which disparity metrics have been calculated, the research
+question must be framed in an appropriate statistical context. The
+multidimensional statistical toolkit for ecology and evolution has been
+greatly expanded in recent years
+\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer}
+2018\href{https://paperpile.com/c/sTGYvp/ZnDd}{)}, but some of these
+advances have yet to be implemented in disparity analyses. Instead,
+hypothesis testing has mostly been confined to a small set of
+well-established methods. One commonly used test is the non-parametric
+permutation analysis of variance
+\href{https://paperpile.com/c/sTGYvp/3hy2+SC6L}{(M. J. Anderson and
+Walsh 2013; M. J. Anderson 2001)}, an analysis of variance (ANOVA) of
+the pairwise distances between different groups. Although statistically
+valid, this test is not always directly related to the hypothesis under
+question. For example, PERMANOVA tests whether two groups share the same
+variance/covariance in a ``distance-space''. This is not the same as
+testing whether the two groups overlap in morphospace. Statistical tests
+should be employed that are tailored to the question at hand, rather
+than simply following common practices.
+
+It is also important to consider which data should be subjected to a
+statistical test. For example, in morphological disparity analysis,
+especially for palaeobiological questions, data are often bootstrapped.
+This has two advantages: (i) when the disparity metric is unidimensional
+(e.g. the sum of variances), bootstrapping the data generates a
+distribution of the metric that can be analysed using the vast
+statistical toolkit available for comparing distributions; (ii) when
+data are scarce, bootstrapping the data allows users to introduce
+variance, rendering the test less sensitive to outliers. However,
+bootstrapped data are pseudoreplicates and thus non-independent. This
+violates the assumptions of most parametric statistical tests.
+Furthermore, the number of bootstrap pseudoreplicates will inevitably
+increase the Type I error rate. These factors are often ignored in
+disparity analyses.
+
+\hypertarget{e-phylogenetic-autocorrelation}{%
+\subsection{(e) Phylogenetic
+autocorrelation}\label{e-phylogenetic-autocorrelation}}
 
@@
 \href{https://paperpile.com/c/sTGYvp/WXik}{(Harvey and Pagel 1998)}.
-Thus for many disparity analyses phylogenetic relationships should be
-taken into account. Multivariate phylogenetic comparative methods have
-been reviewed recently by
-\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)} and
-therefore will not covered further in this review. However, we do
-briefly draw attention to the fact that phylogenetic corrections of PCA
-are inappropriate (cite Pennell Uyeda paper on phylo PCA).
-
-\protect\hypertarget{anchor-3}{}{}3.6. Ancestral state estimation in
-disparity-through-time analyses
-
-One final methodological detail to consider applies only to
-disparity-through-time analyses. Often these analyses use ancestral
-state estimation (often mistakenly referred to as ancestral state
-reconstruction) to extract disparity estimates for non-sampled taxa
-and/or nodes of a phylogeny. Ancestral state estimation can be performed
-at two points in the disparity analysis pipeline, either (1)
-pre-ordination, i.e. the estimation is done before transformation of the
-data (e.g. ordination, or distance matrix construction) and is simply
-based on the original data; or (2) post-ordination, i.e. the estimation
-is done after transformation of the data by estimating the ancestral
-states using the transformed matrix (e.g. the ordination scores)
+Thus, for many disparity analyses, phylogenetic relationships should be
+taken into account. However, it has been noted that some popular
+phylogenetic correction methods (like pPCA) can be inappropriate if
+incorrect assumptions are made about the data (Uyeda et al).
+\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)}
+provide a thorough review of multivariate phylogenetic comparative
+methods, and so we do not consider them further here.
+
+\hypertarget{f-ancestral-state-estimation-in-disparity-through-time-analyses}{%
+\subsection{(f) Ancestral state estimation in disparity-through-time
+analyses}\label{f-ancestral-state-estimation-in-disparity-through-time-analyses}}
+
+Disparity-through-time analyses often use ancestral state estimation to
+extract disparity estimates for non-sampled taxa and/or nodes of a
+phylogeny. Ancestral state estimation can be performed at two points in
+the disparity analysis pipeline: either (1) pre-ordination, i.e. the
+estimation is done before transformation of the data (e.g. ordination,
+or distance matrix construction) and is simply based on the original
+data; or (2) post-ordination, i.e. the estimation is done after
+transformation of the data by estimating the ancestral states using the
+transformed matrix (e.g. the ordination scores)
 \href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
 
-Pre-ordination ancestral state estimation can change the ordinated
-space's geometry -- i.e. the relationship between the points are not yet
-estimated -- and implies longer computational times, but once the
-morphospace is defined its properties will not change. Post-ordination
-ancestral state estimation will not change the ordinated space's
-geometry and is faster to compute, however, it will change the
-morphospace properties such as the number of parameters, i.e. elements
-in the space, or the variance on each axis, which can be problematic for
-statistical tests down the line
+Pre-ordination ancestral state estimation will change the way the
+ordination space is defined -- i.e. the relationship between the points
+are not yet estimated -- and requires longer computational times.
+However, once the morphospace is defined its properties will not change.
+Post-ordination ancestral state estimation will not change the empirical
+ordination space and is faster to compute, but it will add elements in
+the space, whose estimated positions can be problematic for statistical
+tests and evolutionary inferences down the line
 \href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
 
-More importantly, however, using any ancestral state estimation method
-has several caveats. First and foremost, ancestral state estimations are
-highly dependent on the data and chosen methods. Second, when using
-post-ordination ancestral state estimates, the morphospace can be
-saturated leading pairwise-distance metrics to tend to zero. In general,
-using ancestral state estimation can help with recovering patterns of
-changes in disparity but should not be used simply to generate extra
-data points to increase statistical power. In fact, these extra points
-are not independent and can also generate problematic side effects,
-especially when testing for the effects of mass extinctions on
-disparity.
-
-4. Disparity analyses for the future
-
-Morphological disparity analyses are a major tool in evolutionary
-biology, but they have many caveats. There, there is no
-``one-size-fits-all'' morphological disparity analysis pipeline. As with
-any multidimensional analysis, there are many nuances in deciding which
-data to use and how to analyse it, stemming from the explicit hypothesis
-being tested. We believe that many of the problems in morphological
-disparity analysis can arise when ``blindly'' applying methodological
-pipelines while losing sight of the biological question being tested. We
-therefore advise keeping the methodology simple and tractable, at least
-in the preliminary analysis. Also, many of the caveats and technical
-problems in disparity analyses have been tackled in other fields, mainly
-ecology. This is unsurprising, as ecological data are also often
-multidimensional
-\href{https://paperpile.com/c/sTGYvp/krNU+60H0}{(Donohue et al. 2013;
-Canter et al. 2018)}. Seeking solutions using these literature is a
-sensible next step for disparity analysis.
-
-While morphological disparity would greatly benefit from advances in
-multidimensional analysis in different fields similarly, the concept of
-a morphospace could benefit other fields. For example, the
-multidimensional analysis in Diaz
-\href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)} looking
-at the patterns of form and function in plants can be thought as a
-morphospace or a life-history-space (an eco-space?); isotopic analyses
-in \href{https://paperpile.com/c/sTGYvp/PwyQ}{(Jackson et al. 2011)} can
-be represented as a isotope-space; ecosystem functioning in (Donohue et
-al. 2013) as an ecosystem-space, etc. These generalisations could also
-be exported for any set of traits (e.g. acousto-spaces for acoustic
-traits?) and even beyond evolutionary biology (e.g. glotto-spaces for
-linguistic traits?).
-
-Although disparity analyses are now simple to implement
-\href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{(Guillerme
-2018; Adams and Otárola-Castillo 2013; Bouxin 2005; Harmon et al. 2008;
-Lloyd 2016; Navarro 2003; Dixon 2003)}, it is crucial to remember that
-they are multidimensional analyses; and multidimensional analyses are
-complex. In Jurassic Park, Dr Ian Malcolm summarises this problem in an
-elegant way: ``{[}We{]} scientists were so preoccupied with whether or
-not {[}we{]} could that {[}we{]} didn't stop to think if {[}we{]}
-should.'' We believe that morphological analysis in general will greatly
-benefit from emphasising the methodological decisions made
-(\emph{whether we should}), rather than simply using disparity analysis
-because \emph{we can}.
-
-5. Author contributions
-
-TG, NC and PD proposed this review. All authors edited the manuscript.
-
-6. Acknowledgments
-
-The Royal Society. TG funding.
+No ancestral state estimation method is without drawbacks and above all
+else are highly dependent on the data and method used. In general, using
+ancestral state estimation can help with recovering patterns of changes
+in disparity but should not be used simply to generate extra data points
+to increase statistical power. In fact, these extra points are not
+independent and can also have problematic side effects, especially when
+testing for the influence of mass extinctions on disparity.
+
+\hypertarget{disparity-analyses-for-the-future}{%
+\section{4. Disparity analyses for the
+future}\label{disparity-analyses-for-the-future}}
+
+Morphological disparity analyses are widely employed in evolutionary
+palaeobiology, and they are based on a diversity of methods and data.
+There is no ``one-size-fits-all'' pipeline for morphological disparity
+analyses. As with any multidimensional analysis, there are many
+variables that have to be considered when deciding which data to use and
+how to analyse it, stemming from the explicit hypotheses being tested.
+Many of the problems in morphological disparity analysis arise from
+``blind'' application of established methodological pipelines without
+consideration of the biological question being addressed. We advocate
+the bespoke assembly of analytic approaches and, given the computational
+efficiency of these methods, an experimental approach that explores the
+impact of competing approaches, such as choice of distance measure,
+ordination method and ancestral state estimation method on disparity
+analysis results. Many of the methods employed in disparity analysis are
+used more widely in other fields, including genomics and ecology, which
+also encompass analyses of multidimensional datasets
+\href{https://paperpile.com/c/sTGYvp/krNU+60H0+cV3v}{(Donohue et al.
+2013; Canter et al. 2018; Saupe et al. 2015)}. Innovations in
+morphological disparity analyses likely await discovery in their
+respective literatures.
+
+While studies of morphological disparity would benefit from advances in
+multidimensional analysis in other fields, the concept of a morphospace
+could reciprocally benefit other disciplines. For example, the
+multidimensional analysis of Diaz
+\href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, which
+analysed patterns of form and function in plants, is essentially an
+ecomorphospace; isotopic analyses of organisms
+\href{https://paperpile.com/c/sTGYvp/PwyQ}{(Jackson et al. 2011)} can be
+represented as an isotope-space; ecosystem functioning in Donohue et al.
+(2013) as an ecosystem-space, etc. These generalisations could also be
+exported for any set of traits (e.g. acousto-spaces for acoustic traits
+or glotto-spaces for linguistic traits). Cognate approaches have been
+adopted recently in the analysis of single cell comparative
+transcriptome data
+\href{https://paperpile.com/c/sTGYvp/856K}{(Sebé-Pedrós et al. 2018)}
+where interpretation of the resulting transcriptome-spaces would be
+improved by heeding the concerns we highlight concerning morphospaces.
+
+Although disparity analyses are now simple to implement in freely
+available softwares
+\href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{(e.g.,
+Guillerme 2018;} Adams and Otárola-Castillo
+2013\href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{;
+Bouxin 2005; Harmon et al. 2008; Lloyd 2016; Navarro 2003; Dixon 2003)},
+it is crucial to remember that they are multidimensional analyses and
+multidimensional analyses are complex. We assert that future
+morphological analyses will benefit by emphasising the methodological
+decisions made, rather than simply using disparity analysis because
+\emph{we can}.
+
+\hypertarget{author-contributions}{%
+\section{5. Author contributions}\label{author-contributions}}
+
+TG, NC and PD proposed this review; TG and NC led the writing supported
+by PD and GT. All authors edited drafts and approved the final version.
+
+\hypertarget{acknowledgments}{%
+\section{6. Acknowledgments}\label{acknowledgments}}
+
+This article results from discussion at the Royal Society International
+Science Seminar on `Reconciling disparate views on disparity' held at
+Chicheley Hall, 9-10th January, 2018.
+
+-AG was funded by European Research Council Starting grant 637171
+ADaPTiVE.
+
+-ALJ was funded by an Irish Research Council Laureate Award
+IRCLA/2017/186. ES was funded by a University of Adelaide Research
+Fellowship;
+
+-EES was funded by a Leverhulme Trust Research Project Grant (DGR01020).
+
+-PD was funded by NERC (NE/P013678/1; NE/N002067/1) and BBSRC
+(BB/N000919/1);
+
+-SB was funded by European Research Council (ERC) under the European
+Union's Horizon 2020 research and innovation program (grant agreement No
+756226, ERC Starting Grant: PalM) and a Leverhulme Trust Research
+Project Grant (RPG-2017-167);
+
+-TG was funded by ARC DP170103227 and FT180100634 awarded to VW;
 
@@
 Collyer. 2018. ``Multivariate Phylogenetic Comparative Methods:
-Evaluations, Comparisons, and Recommendations.''
-}\href{http://paperpile.com/b/sTGYvp/ZnDd}{\emph{Systematic
-Biology}}\href{http://paperpile.com/b/sTGYvp/ZnDd}{ 67 (1): 14--31.}
+Evaluations, Comparisons, and Recommendations.'' \emph{Systematic
+Biology} 67 (1): 14--31.}
 
@@
 Otárola-Castillo. 2013. ``Geomorph: Anrpackage for the Collection and
-Analysis of Geometric Morphometric Shape Data.''
-}\href{http://paperpile.com/b/sTGYvp/J2G1}{\emph{Methods in Ecology and
-Evolution / British Ecological
-Society}}\href{http://paperpile.com/b/sTGYvp/J2G1}{ 4 (4): 393--99.}
+Analysis of Geometric Morphometric Shape Data.'' \emph{Methods in
+Ecology and Evolution / British Ecological Society} 4 (4): 393--99.}
 
@@
 New Method for Non-Parametric Multivariate Analysis of Variance.''
-}\href{http://paperpile.com/b/sTGYvp/SC6L}{\emph{Austral
-Ecology}}\href{http://paperpile.com/b/sTGYvp/SC6L}{ 26 (1): 32--46.}
+\emph{Austral Ecology} 26 (1): 32--46.}
 
@@
 of Heterogeneous Dispersions: What Null Hypothesis Are You Testing?''
-}\href{http://paperpile.com/b/sTGYvp/3hy2}{\emph{Ecological
-Monographs}}\href{http://paperpile.com/b/sTGYvp/3hy2}{ 83 (4): 557--74.}
+\emph{Ecological Monographs} 83 (4): 557--74.}
 
@@
 Radiation of Jaws Demonstrated Stability despite Faunal and
-Environmental Change.''
-}\href{http://paperpile.com/b/sTGYvp/qjj9}{\emph{Nature}}\href{http://paperpile.com/b/sTGYvp/qjj9}{
-476 (7359): 206--9.}
+Environmental Change.'' \emph{Nature} 476 (7359): 206--9.}
 
 \href{http://paperpile.com/b/sTGYvp/Qsl3}{Bellman, R. 1966. ``Dynamic
-Programming.''
-}\href{http://paperpile.com/b/sTGYvp/Qsl3}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/Qsl3}{
-153 (3731): 34--37.}
+Programming.'' \emph{Science} 153 (3731): 34--37.}
 
 \href{http://paperpile.com/b/sTGYvp/1SD2}{Bookstein, Fred L. 1997.
-}\href{http://paperpile.com/b/sTGYvp/1SD2}{\emph{Morphometric Tools for
-Landmark Data: Geometry and
-Biology}}\href{http://paperpile.com/b/sTGYvp/1SD2}{. Cambridge
-University Press.}
+\emph{Morphometric Tools for Landmark Data: Geometry and Biology}.
+Cambridge University Press.}
 
@@
 Explored by Procrustes Interpretation of Individual Shape Measures
-Pertinent to Function.''
-}\href{http://paperpile.com/b/sTGYvp/sN5d}{\emph{Anatomical Record
-}}\href{http://paperpile.com/b/sTGYvp/sN5d}{ 298 (1): 314--27.}
+Pertinent to Function.'' \emph{Anatomical Record} 298 (1): 314--27.}
 
@@
 Noticed Formula Enforces Fundamental Limits on Geometric Morphometric
-Analyses.''
-}\href{http://paperpile.com/b/sTGYvp/o4w7}{\emph{Evolutionary
-Biology}}\href{http://paperpile.com/b/sTGYvp/o4w7}{ 44 (4): 522--41.}
+Analyses.'' \emph{Evolutionary Biology} 44 (4): 522--41.}
 
 \href{http://paperpile.com/b/sTGYvp/xaUx}{---------. 2017b. ``A Method
-of Factor Analysis for Shape Coordinates.''
-}\href{http://paperpile.com/b/sTGYvp/xaUx}{\emph{American Journal of
-Physical Anthropology}}\href{http://paperpile.com/b/sTGYvp/xaUx}{ 164
-(2): 221--45.}
+of Factor Analysis for Shape Coordinates.'' \emph{American Journal of
+Physical Anthropology} 164 (2): 221--45.}
 
 \href{http://paperpile.com/b/sTGYvp/9JdS}{Bouxin, Guy. 2005. ``Ginkgo, a
-Multivariate Analysis Package.''
-}\href{http://paperpile.com/b/sTGYvp/9JdS}{\emph{Journal of Vegetation
-Science: Official Organ of the International Association for Vegetation
-Science}}\href{http://paperpile.com/b/sTGYvp/9JdS}{ 16 (3): 355--59.}
+Multivariate Analysis Package.'' \emph{Journal of Vegetation Science:
+Official Organ of the International Association for Vegetation Science}
+16 (3): 355--59.}
 
@@
 
-\href{http://paperpile.com/b/sTGYvp/CiPy}{Briggs, D. E., R. A. Fortey,
-and M. A. Wills. 1992. ``Morphological Disparity in the Cambrian.''
-}\href{http://paperpile.com/b/sTGYvp/CiPy}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/CiPy}{
-256 (5064): 1670--73.}
-
 \href{http://paperpile.com/b/sTGYvp/tGyd}{Brusatte, S. L., M. J. Benton,
@@
 Evolution: Macroevolutionary Pattern and Morphological Disparity.''
-}\href{http://paperpile.com/b/sTGYvp/tGyd}{\emph{Biology
-Letters}}\href{http://paperpile.com/b/sTGYvp/tGyd}{ 4 (6): 733--36.}
+\emph{Biology Letters} 4 (6): 733--36.}
 
@@
 Competition, and Opportunism in the Evolutionary Radiation of
-Dinosaurs.''
-}\href{http://paperpile.com/b/sTGYvp/EeC8}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/EeC8}{
-321 (5895): 1485--88.}
+Dinosaurs.'' \emph{Science} 321 (5895): 1485--88.}
 
@@
 Deep Time? A Case Study of Pterosaur (Reptilia: Archosauria)
-Disparity.'' }\href{http://paperpile.com/b/sTGYvp/aSSL}{\emph{Evolution;
-International Journal of Organic
-Evolution}}\href{http://paperpile.com/b/sTGYvp/aSSL}{ 66 (1): 147--62.}
+Disparity.'' \emph{Evolution; International Journal of Organic
+Evolution} 66 (1): 147--62.}
 
@@
 Aquatic Microbial Assemblages in Sarracenia Purpurea Leaves.''
-}\href{http://paperpile.com/b/sTGYvp/60H0}{\emph{Ecology}}\href{http://paperpile.com/b/sTGYvp/60H0}{
-99 (3): 652--60.}
+\emph{Ecology} 99 (3): 652--60.}
 
@@
 Morphospace Occupation Patterns in the Fossil Record: Characterization
-and Analysis of Measures of Disparity.''
-}\href{http://paperpile.com/b/sTGYvp/ROH8}{\emph{Paleobiology}}\href{http://paperpile.com/b/sTGYvp/ROH8}{
-27 (4): 695--715.}
-
-\href{http://paperpile.com/b/sTGYvp/khc9}{Cisneros, Juan Carlos, and
-Marcello Ruta. 2010. ``Morphological Diversity and Biogeography of
-Procolophonids (Amniota: Parareptilia).''
-}\href{http://paperpile.com/b/sTGYvp/khc9}{\emph{Journal of Systematic
-Palaeontology}}\href{http://paperpile.com/b/sTGYvp/khc9}{ 8 (4):
-607--25.}
+and Analysis of Measures of Disparity.'' \emph{Paleobiology} 27 (4):
+695--715.}
 
@@
 Friedman, Graeme T. Lloyd, and Roger B. J. Benson. 2015. ``Evidence for
-a Mid-Jurassic Adaptive Radiation in Mammals.''
-}\href{http://paperpile.com/b/sTGYvp/PbSx}{\emph{Current Biology:
-CB}}\href{http://paperpile.com/b/sTGYvp/PbSx}{ 25 (16): 2137--42.}
+a Mid-Jurassic Adaptive Radiation in Mammals.'' \emph{Current Biology:
+CB} 25 (16): 2137--42.}
 
@@
 ``Mega-Evolutionary Dynamics of the Adaptive Radiation of Birds.''
-}\href{http://paperpile.com/b/sTGYvp/RjqE}{\emph{Nature}}\href{http://paperpile.com/b/sTGYvp/RjqE}{
-542 (7641): 344--47.}
-
-\href{http://paperpile.com/b/sTGYvp/Qrba}{Dávalos, Liliana M., Paúl M.
-Velazco, Omar M. Warsi, Peter D. Smits, and Nancy B. Simmons. 2014.
-``Integrating Incomplete Fossils by Isolating Conflicting Signal in
-Saturated and Non-Independent Morphological Characters.''
-}\href{http://paperpile.com/b/sTGYvp/Qrba}{\emph{Systematic
-Biology}}\href{http://paperpile.com/b/sTGYvp/Qrba}{ 63 (4): 582--600.}
+\emph{Nature} 542 (7641): 344--47.}
 
@@
 Philip C. J. Donoghue. 2018. ``Evolution of Metazoan Morphological
-Disparity.''
-}\href{http://paperpile.com/b/sTGYvp/0y4V}{\emph{Proceedings of the
-National Academy of Sciences of the United States of
-America}}\href{http://paperpile.com/b/sTGYvp/0y4V}{ 115 (38):
-E8909--18.}
+Disparity.'' \emph{Proceedings of the National Academy of Sciences of
+the United States of America} 115 (38): E8909--18.}
 
@@
 Dray, Björn Reu, et al. 2016. ``The Global Spectrum of Plant Form and
-Function.''
-}\href{http://paperpile.com/b/sTGYvp/47fI}{\emph{Nature}}\href{http://paperpile.com/b/sTGYvp/47fI}{
-529 (7585): 167--71.}
+Function.'' \emph{Nature} 529 (7585): 167--71.}
 
 \href{http://paperpile.com/b/sTGYvp/2KmX}{Dixon, Philip. 2003. ``VEGAN,
-a Package of R Functions for Community Ecology.''
-}\href{http://paperpile.com/b/sTGYvp/2KmX}{\emph{Journal of Vegetation
-Science: Official Organ of the International Association for Vegetation
-Science}}\href{http://paperpile.com/b/sTGYvp/2KmX}{ 14 (6): 927.}
+a Package of R Functions for Community Ecology.'' \emph{Journal of
+Vegetation Science: Official Organ of the International Association for
+Vegetation Science} 14 (6): 927.}
 
@@
 Healy, Miguel Lurgi, Nessa E. O'Connor, and Mark C. Emmerson. 2013. ``On
-the Dimensionality of Ecological Stability.''
-}\href{http://paperpile.com/b/sTGYvp/krNU}{\emph{Ecology
-Letters}}\href{http://paperpile.com/b/sTGYvp/krNU}{ 16 (4): 421--29.}
+the Dimensionality of Ecological Stability.'' \emph{Ecology Letters} 16
+(4): 421--29.}
+
+\href{http://paperpile.com/b/sTGYvp/EPJ2}{Erwin, Douglas H. 2007.
+``DISPARITY: MORPHOLOGICAL PATTERN AND DEVELOPMENTAL CONTEXT.''
+\emph{Palaeontology}.
+https://doi.org/}\href{http://dx.doi.org/10.1111/j.1475-4983.2006.00614.x}{10.1111/j.1475-4983.2006.00614.x}\href{http://paperpile.com/b/sTGYvp/EPJ2}{.}
+
+\href{http://paperpile.com/b/sTGYvp/Z6l6}{---------. 2011.
+``Evolutionary Uniformitarianism.'' \emph{Developmental Biology} 357
+(1): 27--34.}
 
@@
 Tenrecidae): Comparing Tenrec Skull Diversity to Their Closest
-Relatives.''
-}\href{http://paperpile.com/b/sTGYvp/yyNa}{\emph{PeerJ}}\href{http://paperpile.com/b/sTGYvp/yyNa}{
-3 (April): e927.}
-
-\href{http://paperpile.com/b/sTGYvp/oenu}{Foote, M., and S. J. Gould.
-1992. ``Cambrian and Recent Morphological Disparity.''
-}\href{http://paperpile.com/b/sTGYvp/oenu}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/oenu}{
-258 (5089): 1816.}
-
-\href{http://paperpile.com/b/sTGYvp/fTJ3}{Foote, Mike. 1995.
+Relatives.'' \emph{PeerJ} 3 (April): e927.}
+
+\href{http://paperpile.com/b/sTGYvp/2Neu}{Foote, Mike. 1989.
+``Perimeter-Based Fourier Analysis: A New Morphometric Method Applied to
+the Trilobite Cranidium.'' \emph{Journal of Paleontology}.
+https://doi.org/}\href{http://dx.doi.org/10.1017/s0022336000036556}{10.1017/s0022336000036556}\href{http://paperpile.com/b/sTGYvp/2Neu}{.}
+
+\href{http://paperpile.com/b/sTGYvp/fTJ3}{---------. 1995.
 ``Morphological Diversification of Paleozoic Crinoids.''
-}\href{http://paperpile.com/b/sTGYvp/fTJ3}{\emph{Paleobiology}}\href{http://paperpile.com/b/sTGYvp/fTJ3}{
-21 (03): 273--99.}
+\emph{Paleobiology} 21 (03): 273--99.}
 
 \href{http://paperpile.com/b/sTGYvp/yqPw}{---------. 1997. ``THE
-EVOLUTION OF MORPHOLOGICAL DIVERSITY.''
-}\href{http://paperpile.com/b/sTGYvp/yqPw}{\emph{Annual Review of
-Ecology and Systematics}}\href{http://paperpile.com/b/sTGYvp/yqPw}{ 28
-(1): 129--52.}
+EVOLUTION OF MORPHOLOGICAL DIVERSITY.'' \emph{Annual Review of Ecology
+and Systematics} 28 (1): 129--52.}
 
@@
 Briggs, and M. A. Wills. 1996. ``The Cambrian Evolutionary `explosion':
-Decoupling Cladogenesis from Morphological Disparity.''
-}\href{http://paperpile.com/b/sTGYvp/2tbJ}{\emph{Biological Journal of
-the Linnean Society. Linnean Society of
-London}}\href{http://paperpile.com/b/sTGYvp/2tbJ}{ 57 (1): 13--33.}
+Decoupling Cladogenesis from Morphological Disparity.'' \emph{Biological
+Journal of the Linnean Society. Linnean Society of London} 57 (1):
+13--33.}
 
@@
 ``Explosive Morphological Diversification of Spiny-Finned Teleost Fishes
-in the Aftermath of the End-Cretaceous Extinction.''
-}\href{http://paperpile.com/b/sTGYvp/EETc}{\emph{Proceedings. Biological
-Sciences / The Royal Society}}\href{http://paperpile.com/b/sTGYvp/EETc}{
-277 (1688): 1675--83.}
+in the Aftermath of the End-Cretaceous Extinction.'' \emph{Proceedings.
+Biological Sciences / The Royal Society} 277 (1688): 1675--83.}
 
@@
 All Roads Can Be Taken: Development Induces Anisotropic Accessibility in
-Morphospace.''
-}\href{http://paperpile.com/b/sTGYvp/SJbC}{\emph{Evolution \&
-Development}}\href{http://paperpile.com/b/sTGYvp/SJbC}{ 16 (6):
-373--81.}
+Morphospace.'' \emph{Evolution \& Development} 16 (6): 373--81.}
 
@@
 Geometry of Morphospaces: Lessons from the Classic Raup Shell Coiling
-Model.'' }\href{http://paperpile.com/b/sTGYvp/QVvv}{\emph{Biological
-Reviews of the Cambridge Philosophical
-Society}}\href{http://paperpile.com/b/sTGYvp/QVvv}{ 92 (2): 1142--55.}
-
-\href{http://paperpile.com/b/sTGYvp/CidX}{Gould, Stephen Jay. 1991.
-``The Disparity of the Burgess Shale Arthropod Fauna and the Limits of
-Cladistic Analysis: Why We Must Strive to Quantify Morphospace.''
-}\href{http://paperpile.com/b/sTGYvp/CidX}{\emph{Paleobiology}}\href{http://paperpile.com/b/sTGYvp/CidX}{
-17 (04): 411--23.}
-
-\href{http://paperpile.com/b/sTGYvp/Uns3}{---------. 2000.
-}\href{http://paperpile.com/b/sTGYvp/Uns3}{\emph{Wonderful Life: The
-Burgess Shale and the Nature of
-History}}\href{http://paperpile.com/b/sTGYvp/Uns3}{. Random House.}
+Model.'' \emph{Biological Reviews of the Cambridge Philosophical
+Society} 92 (2): 1142--55.}
+
+\href{http://paperpile.com/b/sTGYvp/Uns3}{Gould, Stephen Jay. 2000.
+\emph{Wonderful Life: The Burgess Shale and the Nature of History}.
+Random House.}
 
@@
 ``dispRity : A Modular R Package for Measuring Disparity.''
-}\href{http://paperpile.com/b/sTGYvp/xDqf}{\emph{Methods in Ecology and
-Evolution / British Ecological
-Society}}\href{http://paperpile.com/b/sTGYvp/xDqf}{ 9 (7): 1755--63.}
+\emph{Methods in Ecology and Evolution / British Ecological Society} 9
+(7): 1755--63.}
 
@@
 Cooper. 2018. ``Time for a Rethink: Time Sub-Sampling Methods in
-Disparity-through-Time Analyses.''
-}\href{http://paperpile.com/b/sTGYvp/ekU4}{\emph{Palaeontology}}\href{http://paperpile.com/b/sTGYvp/ekU4}{
-61 (4): 481--93.}
+Disparity-through-Time Analyses.'' \emph{Palaeontology} 61 (4):
+481--93.}
+
+\href{http://paperpile.com/b/sTGYvp/u1KE}{Guillerme, Thomas, Mark N.
+Puttick, Ariel E. Marcy, and Vera Weisbecker. n.d. ``Shifting Spaces:
+Which Disparity or Dissimilarity Metrics Best Summarise Occupancy in
+Multidimensional Spaces?''
+https://doi.org/}\href{http://dx.doi.org/10.1101/801571}{10.1101/801571}\href{http://paperpile.com/b/sTGYvp/u1KE}{.}
 
@@
 Weir, Chad D. Brock, Richard E. Glor, and Wendell Challenger. 2008.
-``GEIGER: Investigating Evolutionary Radiations.''
-}\href{http://paperpile.com/b/sTGYvp/9Zoi}{\emph{Bioinformatics
-}}\href{http://paperpile.com/b/sTGYvp/9Zoi}{ 24 (1): 129--31.}
+``GEIGER: Investigating Evolutionary Radiations.'' \emph{Bioinformatics}
+24 (1): 129--31.}
 
 \href{http://paperpile.com/b/sTGYvp/WXik}{Harvey, Paul H., and Mark D.
-Pagel. 1998. }\href{http://paperpile.com/b/sTGYvp/WXik}{\emph{The
-Comparative Method in Evolutionary
-Biology}}\href{http://paperpile.com/b/sTGYvp/WXik}{. Oxford University
-Press, USA.}
+Pagel. 1998. \emph{The Comparative Method in Evolutionary Biology}.
+Oxford University Press, USA.}
+
+\href{http://paperpile.com/b/sTGYvp/3JPy}{Hill, Jennifer J., Mark N.
+Puttick, Thomas L. Stubbs, Emily J. Rayfield, and Philip C. J. Donoghue.
+2018. ``Evolution of Jaw Disparity in Fishes.'' \emph{Palaeontology}.
+https://doi.org/}\href{http://dx.doi.org/10.1111/pala.12371}{10.1111/pala.12371}\href{http://paperpile.com/b/sTGYvp/3JPy}{.}
 
@@
 Shape Evolution with Exoskeletal Character Evolution in the Trilobite
-Family Pterocephaliidae.''
-}\href{http://paperpile.com/b/sTGYvp/xLdm}{\emph{Palaeontology}}\href{http://paperpile.com/b/sTGYvp/xLdm}{
-60 (3): 309--18.}
+Family Pterocephaliidae.'' \emph{Palaeontology} 60 (3): 309--18.}
 
 \href{http://paperpile.com/b/sTGYvp/vTHS}{Hopkins, Melanie J., and
-Sylvain Gerber. 2017. ``Morphological Disparity.'' In
-}\href{http://paperpile.com/b/sTGYvp/vTHS}{\emph{Evolutionary
-Developmental Biology}}\href{http://paperpile.com/b/sTGYvp/vTHS}{,
-1--12.}
+Sylvain Gerber. 2017. ``Morphological Disparity.'' In \emph{Evolutionary
+Developmental Biology}, 1--12.}
 
@@
 Decline of the Cambrian Trilobite Family Pterocephaliidae.''
-}\href{http://paperpile.com/b/sTGYvp/hea5}{\emph{Journal of Evolutionary
-Biology}}\href{http://paperpile.com/b/sTGYvp/hea5}{ 26 (8): 1665--76.}
+\emph{Journal of Evolutionary Biology} 26 (8): 1665--76.}
 
@@
 Gerber, and Matthew Albion Wills. 2013. ``Clades Reach Highest
-Morphological Disparity Early in Their Evolution.''
-}\href{http://paperpile.com/b/sTGYvp/xxh5}{\emph{Proceedings of the
-National Academy of Sciences of the United States of
-America}}\href{http://paperpile.com/b/sTGYvp/xxh5}{ 110 (34):
-13875--79.}
+Morphological Disparity Early in Their Evolution.'' \emph{Proceedings of
+the National Academy of Sciences of the United States of America} 110
+(34): 13875--79.}
 
@@
 Niche Widths among and within Communities: SIBER - Stable Isotope
-Bayesian Ellipses in R.''
-}\href{http://paperpile.com/b/sTGYvp/PwyQ}{\emph{The Journal of Animal
-Ecology}}\href{http://paperpile.com/b/sTGYvp/PwyQ}{ 80 (3): 595--602.}
+Bayesian Ellipses in R.'' \emph{The Journal of Animal Ecology} 80 (3):
+595--602.}
 
 \href{http://paperpile.com/b/sTGYvp/oFiP}{Legendre, P., and Loic F. J.
-Legendre. 2012.
-}\href{http://paperpile.com/b/sTGYvp/oFiP}{\emph{Numerical
-Ecology}}\href{http://paperpile.com/b/sTGYvp/oFiP}{. Elsevier.}
+Legendre. 2012. \emph{Numerical Ecology}. Elsevier.}
 
@@
 Character-Taxon Matrices: Implementation, Challenges, Progress, and
-Future Directions.''
-}\href{http://paperpile.com/b/sTGYvp/bCsU}{\emph{Biological Journal of
-the Linnean Society. Linnean Society of
-London}}\href{http://paperpile.com/b/sTGYvp/bCsU}{ 118 (1): 131--51.}
+Future Directions.'' \emph{Biological Journal of the Linnean Society.
+Linnean Society of London} 118 (1): 131--51.}
 
@@
 through Discrete-Character Morphospace: Synthesizing Phylogeny, Tempo,
-and Disparity.''
-}\href{http://paperpile.com/b/sTGYvp/53SJ}{\emph{Palaeontology}}\href{http://paperpile.com/b/sTGYvp/53SJ}{
-61 (5): 637--45.}
-
-\href{http://paperpile.com/b/sTGYvp/TtGs}{Morris, S. C. 1989. ``Burgess
-Shale Faunas and the Cambrian Explosion.''
-}\href{http://paperpile.com/b/sTGYvp/TtGs}{\emph{Science}}\href{http://paperpile.com/b/sTGYvp/TtGs}{
-246 (4928): 339--46.}
+and Disparity.'' \emph{Palaeontology} 61 (5): 637--45.}
+
+\href{http://paperpile.com/b/sTGYvp/dJHu}{Losos, Jonathan B. 2011.
+\emph{Lizards in an Evolutionary Tree: Ecology and Adaptive Radiation of
+Anoles}. Univ of California Press.}
 
@@
 Morphological Disparity in the Middle Jurassic Ammonite Radiation.''
-}\href{http://paperpile.com/b/sTGYvp/aVVj}{\emph{Palaeogeography,
-Palaeoclimatology,
-Palaeoecology}}\href{http://paperpile.com/b/sTGYvp/aVVj}{ 248 (1-2):
+\emph{Palaeogeography, Palaeoclimatology, Palaeoecology} 248 (1-2):
 82--95.}
@@
 A MATLAB-Based Program for Morphospace-Disparity Analysis.''
-}\href{http://paperpile.com/b/sTGYvp/EmTR}{\emph{Computers \&
-Geosciences}}\href{http://paperpile.com/b/sTGYvp/EmTR}{ 29 (5):
-655--64.}
+\emph{Computers \& Geosciences} 29 (5): 655--64.}
 
@@
 S. Y. Lee. 2018. ``Geometric Morphometrics, Homology and Cladistics:
-Review and Recommendations.''
-}\href{http://paperpile.com/b/sTGYvp/yO2t}{\emph{Cladistics: The
-International Journal of the Willi Hennig
-Society}}\href{http://paperpile.com/b/sTGYvp/yO2t}{.
+Review and Recommendations.'' \emph{Cladistics: The International
+Journal of the Willi Hennig Society}.
 https://doi.org/}\href{http://dx.doi.org/10.1111/cla.12340}{10.1111/cla.12340}\href{http://paperpile.com/b/sTGYvp/yO2t}{.}
@@
 Combined Geometric Morphometric and Finite Element Modeling Approach.''
-}\href{http://paperpile.com/b/sTGYvp/tSIy}{\emph{Journal of
-Morphology}}\href{http://paperpile.com/b/sTGYvp/tSIy}{ 269 (7):
-840--64.}
+\emph{Journal of Morphology} 269 (7): 840--64.}
 
 \href{http://paperpile.com/b/sTGYvp/I0Ic}{Raup, D. M. 1961. ``THE
-GEOMETRY OF COILING IN GASTROPODS.''
-}\href{http://paperpile.com/b/sTGYvp/I0Ic}{\emph{Proceedings of the
-National Academy of Sciences of the United States of
-America}}\href{http://paperpile.com/b/sTGYvp/I0Ic}{ 47 (4): 602--9.}
+GEOMETRY OF COILING IN GASTROPODS.'' \emph{Proceedings of the National
+Academy of Sciences of the United States of America} 47 (4): 602--9.}
 
@@
 Morphological Disparity and Taxic Diversity during the Adaptive
-Radiation of Anomodont Therapsids.''
-}\href{http://paperpile.com/b/sTGYvp/geAO}{\emph{Proceedings. Biological
-Sciences / The Royal Society}}\href{http://paperpile.com/b/sTGYvp/geAO}{
-280 (1768): 20131071.}
+Radiation of Anomodont Therapsids.'' \emph{Proceedings. Biological
+Sciences / The Royal Society} 280 (1768): 20131071.}
+
+\href{http://paperpile.com/b/sTGYvp/cV3v}{Saupe, Erin E., Huijie Qiao,
+Jonathan R. Hendricks, Roger W. Portell, Stephen J. Hunter, Jorge
+Soberón, and Bruce S. Lieberman. 2015. ``Niche Breadth and Geographic
+Range Size as Determinants of Species Survival on Geological Time
+Scales.'' \emph{Global Ecology and Biogeography}.
+https://doi.org/}\href{http://dx.doi.org/10.1111/geb.12333}{10.1111/geb.12333}\href{http://paperpile.com/b/sTGYvp/cV3v}{.}
+
+\href{http://paperpile.com/b/sTGYvp/856K}{Sebé-Pedrós, Arnau, Elad
+Chomsky, Kevin Pang, David Lara-Astiaso, Federico Gaiti, Zohar Mukamel,
+Ido Amit, Andreas Hejnol, Bernard M. Degnan, and Amos Tanay. 2018.
+``Early Metazoan Cell Type Diversity and the Evolution of Multicellular
+Gene Regulation.'' \emph{Nature Ecology \& Evolution} 2 (7): 1176--88.}
 
@@
 Form Evolution in Viburnum Parallels Variation within Individual
-Plants.'' }\href{http://paperpile.com/b/sTGYvp/ZEDR}{\emph{The American
-Naturalist}}\href{http://paperpile.com/b/sTGYvp/ZEDR}{ 191 (2):
-235--49.}
-
-\href{http://paperpile.com/b/sTGYvp/6tNm}{Valentine, J. W. 1986.
-``Fossil Record of the Origin of Baupläne and Its Implications.'' In
-}\href{http://paperpile.com/b/sTGYvp/6tNm}{\emph{Patterns and Processes
-in the History of Life}}\href{http://paperpile.com/b/sTGYvp/6tNm}{,
-209--22.}
+Plants.'' \emph{The American Naturalist} 191 (2): 235--49.}
 
@@
 Mapping of Form to Function: A General Principle in Organismal Design?''
-}\href{http://paperpile.com/b/sTGYvp/Ejzr}{\emph{Integrative and
-Comparative Biology}}\href{http://paperpile.com/b/sTGYvp/Ejzr}{ 45 (2):
-256--62.}
+\emph{Integrative and Comparative Biology} 45 (2): 256--62.}
+
+\href{http://paperpile.com/b/sTGYvp/TZzO}{Weisbecker, Vera, Thomas
+Guillerme, Cruise Speck, Emma Sherratt, Hyab Mehari Abraha, Alana C.
+Sharp, Claire E. Terhune, Simon Collins, Stephen Johnston, and Olga
+Panagiotopoulou. 2019. ``Individual Variation of the Masticatory System
+Dominates 3D Skull Shape in the Herbivory-Adapted Marsupial Wombats.''
+\emph{Frontiers in Zoology} 16 (November): 41.}
 
 \href{http://paperpile.com/b/sTGYvp/nFf7}{Wills, Matthew A. 2001.
-``Morphological Disparity: A Primer.'' In
-}\href{http://paperpile.com/b/sTGYvp/nFf7}{\emph{Topics in
-Geobiology}}\href{http://paperpile.com/b/sTGYvp/nFf7}{, 55--144.}
+``Morphological Disparity: A Primer.'' In \emph{Topics in Geobiology},
+55--144.}
 
@@
 Index: A Comparison of Cambrian and Recent Arthropods.''
-}\href{http://paperpile.com/b/sTGYvp/eZ3F}{\emph{Paleobiology}}\href{http://paperpile.com/b/sTGYvp/eZ3F}{
-20 (02): 93--130.}
+\emph{Paleobiology} 20 (02): 93--130.}
 
@@
 ``Phenotypic Innovation and Adaptive Constraints in the Evolutionary
-Radiation of Palaeozoic Crinoids.''
-}\href{http://paperpile.com/b/sTGYvp/s33b}{\emph{Scientific
-Reports}}\href{http://paperpile.com/b/sTGYvp/s33b}{ 7 (1): 13745.}
-
-\href{http://paperpile.com/b/sTGYvp/ZQS9}{Zelditch, Miriam Leah, Donald
-L. Swiderski, and H. David Sheets. 2012.
-}\href{http://paperpile.com/b/sTGYvp/ZQS9}{\emph{Geometric Morphometrics
-for Biologists: A Primer}}\href{http://paperpile.com/b/sTGYvp/ZQS9}{.
-Academic Press.}
-
-\href{http://paperpile.com/b/sTGYvp/ZqML}{Zou, Zhengting, and Jianzhi
-Zhang. 2016. ``Morphological and Molecular Convergences in Mammalian
-Phylogenetics.'' }\href{http://paperpile.com/b/sTGYvp/ZqML}{\emph{Nature
-Communications}}\href{http://paperpile.com/b/sTGYvp/ZqML}{ 7
-(September): 12758.}
+Radiation of Palaeozoic Crinoids.'' \emph{Scientific Reports} 7 (1):
+13745.}

### 2020-01-24 Writing/disparity_review.tex
Rearranged ordination part (to write properly)

@@
 
-\hypertarget{a-multidimensionality}{%
-\subsection{(a) Multidimensionality}\label{a-multidimensionality}}
+\subsection{a) To ordinate or not to ordinate - that is the (multidimensional) question}
 
+% -Disparity analysis have a history of using ordination techniques
 Disparity analyses often use ordination techniques for dimension reduction.
+Dimension reduction (or ordination) is a powerful tool for visualizing data distribution in disparity analyses, helping to elucidate the underlying structure within datasets.
+%     -Ordination techniques are blablabla
 These are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g.
 principal component analysis -- PCA; principal coordinates analysis -- PCO/PCoA; non-metric multidimensional scaling -- NMDS; etc.).
-Using ordinations in disparity analysis makes it easier for researchers to comprehend patterns in just two or three spatial dimensions at a time - or, indeed, just the first few axes of a PCA or a PCO.
-Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(}Bookstein \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1997, 2015, {[}b{]} 2017, {[}a{]} 2017)}{[}CITE Vera and Thomas paper{]}.
-%TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
-Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \href{https://paperpile.com/c/sTGYvp/0y4V+QVvv}{(Deline et al.
-2018;
-Gerber 2017)}.
-Categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
-contrary to a PCA, the first few axis do not encompass most variance of the dataset).
-Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric (\href{https://paperpile.com/c/sTGYvp/SJbC}{i.e.
-the distance between A and B is not equal to the distance between B and A; Gerber 2014)}.
-
-\hypertarget{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}{%
-\subsection{(b) To ordinate or not to ordinate - that is the
-(multidimensional)
-question}\label{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}}
-
-Dimension reduction (or ordination) is a powerful tool for visualizing data distribution in disparity analyses, helping to elucidate the underlying structure within datasets.
-However, like most other aspects of disparity analyses, reducing dimensionality can be tricky.
 Ordination techniques come in many flavours depending on the data and the desired morphospace properties.
 Quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCoA, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) (see Legendre and Legendre 2012,chapter 9, for a detailed explanation of ordination methods and properties).
-
+% -But mainly ordination is good for visualising.
 Ordination is advantageous for plotting and visualising data, and can reveal properties of the morphospace not intuitively captured by disparity metrics (see Metrics section below {[}metrics{]}).
+% -Ordination can be good for reducing dimensions if the data has a lot of measurements for a few number of observations (e.g. landmark data, discrete morphological characters, etc.)
+Using ordinations in disparity analysis makes it easier for researchers to comprehend patterns in just two or three spatial dimensions at a time - or, indeed, just the first few axes of a PCA or a PCO.
 Additionally, after ordinating the data, it is possible to reduce further the number of dimensions by focussing on just a subset of axes (e.g.
@@
 This has clear advantages for interpreting the results.
-For example, the axes will represent gradients of biological variation (e.g.
-elongation and flattening of the beak in birds \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al.
-2017}).
+For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak in birds \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017}).
+
+% -However ordination can be problematic
+However, like most other aspects of disparity analyses, reducing dimensionality can be tricky.
 However, there are a number of potential problems with these transformations of the data.
@@
 %TG: comment from Graeme: the ordination space is decided by the data present (e.g. shoebill)
+%     -It should contain all axis
 In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results.
 %TG: from Anjali: add that people should use 95% or 99% percent of the data (why not 100% though?)
+%     -It is hard to interpret
+Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(}Bookstein \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1997, 2015, {[}b{]} 2017, {[}a{]} 2017)}{[}CITE Vera and Thomas paper{]}.
+%TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
+Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \href{https://paperpile.com/c/sTGYvp/0y4V+QVvv}{(Deline et al.
+2018; Gerber 2017)}.
 Furthermore, interpreting biological variation along the axes is always a \emph{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
-
-However, In many cases, ordination might not be necessary.
-For example, if a metric (e.g.
-sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below {[}metrics{]}), it is not necessary to calculate it on ordinated data {[}e.g.\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
-2015)}{]}.
 Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
@@
 For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0, has no biological meaning if these are binary alternate states.
+%     -It can make non-metric spaces
+Categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
+contrary to a PCA, the first few axis do not encompass most variance of the dataset).
+Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric (\href{https://paperpile.com/c/sTGYvp/SJbC}{i.e.
+the distance between A and B is not equal to the distance between B and A; Gerber 2014)}.
+%     -It is often not necessary
+And last but not least reason, In many cases, ordination might not be necessary.
+For example, if a metric (e.g.
+sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below {[}metrics{]}), it is not necessary to calculate it on ordinated data {[}e.g.\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
+2015)}{]}.
+% -Disparity can be measured directly from the non-ordinated dataset (e.g. via distance matrices)
 For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination (Lloyd 2016, 2018).
+
+% Stopped p.13
+
+% \hypertarget{a-multidimensionality}{%
+% \subsection{(a) Multidimensionality}\label{a-multidimensionality}}
+
+% Disparity analyses often use ordination techniques for dimension reduction.
+% These are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g.
+principal component analysis -- PCA; principal coordinates analysis -- PCO/PCoA; non-metric multidimensional scaling -- NMDS; etc.).
+% Using ordinations in disparity analysis makes it easier for researchers to comprehend patterns in just two or three spatial dimensions at a time - or, indeed, just the first few axes of a PCA or a PCO.
+% Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(}Bookstein \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1997, 2015, {[}b{]} 2017, {[}a{]} 2017)}{[}CITE Vera and Thomas paper{]}.
+%TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
+% Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \href{https://paperpile.com/c/sTGYvp/0y4V+QVvv}{(Deline et al.
+2018; Gerber 2017)}.
+% Categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
+contrary to a PCA, the first few axis do not encompass most variance of the dataset).
+% Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric (\href{https://paperpile.com/c/sTGYvp/SJbC}{i.e.
+the distance between A and B is not equal to the distance between B and A; Gerber 2014)}.
+
+% \hypertarget{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}{%
+% \subsection{(b) To ordinate or not to ordinate - that is the
+% (multidimensional)
+% question}\label{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}}
+
+% Dimension reduction (or ordination) is a powerful tool for visualizing data distribution in disparity analyses, helping to elucidate the underlying structure within datasets.
+% However, like most other aspects of disparity analyses, reducing dimensionality can be tricky.
+% Ordination techniques come in many flavours depending on the data and the desired morphospace properties.
+% Quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCoA, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) (see Legendre and Legendre 2012,chapter 9, for a detailed explanation of ordination methods and properties).
+
+% Ordination is advantageous for plotting and visualising data, and can reveal properties of the morphospace not intuitively captured by disparity metrics (see Metrics section below {[}metrics{]}).
+% Additionally, after ordinating the data, it is possible to reduce further the number of dimensions by focussing on just a subset of axes (e.g.
+selecting only those axes that describe the majority of the variation) or by fitting the variation to a prescribed number of axes.
+% In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions (Legendre and Legendre 2012).
+%TG: cite also: Dryden and Mardia "Statistical Shape Analysis book"
+% This has clear advantages for interpreting the results.
+% For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak in birds \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017}).
+% However, there are a number of potential problems with these transformations of the data.
+% %TG: comment from Sylvain: difference between a morphospace and an ordination space
+% %TG: comment from Graeme: the ordination space is decided by the data present (e.g. shoebill)
+% In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results.
+%TG: from Anjali: add that people should use 95% or 99% percent of the data (why not 100% though?)
+% Furthermore, interpreting biological variation along the axes is always a \emph{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
+
+% However, In many cases, ordination might not be necessary.
+For example, if a metric (e.g.
+sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below {[}metrics{]}), it is not necessary to calculate it on ordinated data {[}e.g.\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
+2015)}{]}.
+% Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
+% For example, when the analysed data is non-Euclidean (e.g.
+some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \href{https://paperpile.com/c/sTGYvp/SJbC}{(}Gerber 2014\href{https://paperpile.com/c/sTGYvp/SJbC}{)}.
+%TG: should be Gerber 2019
+% This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear.
+% Finally, \emph{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible. %TG: Cite Gerber 2019 10.1111/pala.12407
+% Although some gradients are easy to detect or interpret (e.g.
+the elongation and depth of mandibles in fishes on first and second PC axes, respectively; {[}\href{https://paperpile.com/c/sTGYvp/3JPy}{Hill et al.
+2018{]}}), some are not (e.g.
+\href{https://paperpile.com/c/sTGYvp/TZzO}{Weisbecker et al.
+2019)}.
+% For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0, has no biological meaning if these are binary alternate states.
+% For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination (Lloyd 2016, 2018).
 
@@
 
-\hypertarget{c-disparity-metrics}{%
-\subsection{(c) Disparity metrics}\label{c-disparity-metrics}}
+\hypertarget{b-disparity-indices}{%
+\subsection{(b) Measuring disparity}\label{c-disparity-indices}}
 
@@
 
-\hypertarget{d-testing-hypotheses-on-the-evolution-of-disparity}{%
-\subsection{(d) Testing hypotheses on the evolution of
-disparity}\label{d-testing-hypotheses-on-the-evolution-of-disparity}}
+\hypertarget{c-testing-hypotheses-on-the-evolution-of-disparity}{%
+\subsection{(c) Testing hypotheses on the evolution of
+disparity}\label{c-testing-hypotheses-on-the-evolution-of-disparity}}
 
@@
 
-\hypertarget{e-phylogenetic-autocorrelation}{%
-\subsection{(e) Phylogenetic
-autocorrelation}\label{e-phylogenetic-autocorrelation}}
+%TG: Add section d) within c)
+\hypertarget{d-phylogenetic-autocorrelation}{%
+\subsection{(d) Phylogenetic
+autocorrelation}\label{d-phylogenetic-autocorrelation}}
 
@@
 
-\hypertarget{f-ancestral-state-estimation-in-disparity-through-time-analyses}{%
-\subsection{(f) Ancestral state estimation in disparity-through-time
-analyses}\label{f-ancestral-state-estimation-in-disparity-through-time-analyses}}
+%TG: Add section e) within c)???
+\hypertarget{e-ancestral-state-estimation-in-disparity-through-time-analyses}{%
+\subsection{(e) Ancestral state estimation in disparity-through-time
+analyses}\label{e-ancestral-state-estimation-in-disparity-through-time-analyses}}
 

### 2020-02-18 Writing/disparity_review.tex
Reformated until p.13

@@
 
-Statistical nalyses of morphological disparity have been used to
-characterise and investigate the evolution of variation in the anatomy,
-function, and ecology of organisms since the 1980s. While a diversity of
-methods have been employed, it is unclear whether they provide
-equivalent insights. Here we review the most commonly used approaches
-for characterising and analysing morphological disparity, all of which
-have associated limitations that, if ignored, can lead to
-misinterpretation. We provide best practice guidelines for disparity
-analyses, while noting that there can be no ``one-size-fits-all''
-approach. The available tools should always be used in the context of a
-specific biological question that will determine data and method
-selection at every stage of the analysis.
+Analyses of morphological disparity have been used to characterise and investigate the evolution of variation in the anatomy, function, and ecology of organisms since the 1980s.
+While a diversity of methods have been employed, it is unclear whether they provide
+equivalent insights.
+Here we review the most commonly used approaches for characterising and analysing morphological disparity, all of which have associated limitations that, if ignored, can lead to misinterpretation.
+We provide best practice guidelines for disparity analyses, while noting that there can be no ``one-size-fits-all'' approach.
+The available tools should always be used in the context of a specific biological question that will determine data and method selection at every stage of the analysis.
 
@@
 
-Clades of organisms are characterised by variation in both numbers of
-species and range of phenotypes through time. At the extremes, clades
-may be exceptionally species-rich and phenotypically diverse (e.g.
-cichlids and molluscs), species-rich but phenotypically conservative
-(e.g. bacteria and nematodes), species-poor but phenotypically diverse
-(e.g. Afrotheria), or depauperate in both species and phenotypic
-diversity (e.g. lungfish). These phenomena suggest that taxonomic and
-phenotypic diversity are not inextricably linked, raising important
-questions about how phenotypic diversity arises, such as: How does
-diversity evolve? Are some morphologies more common than others? Can
-anatomy evolve in all ``directions'' or are some anatomies impossible to
-achieve? (Mike Foote 1997, 1995) What role does ecology play in
-structuring morphological diversity?. Analyses of species diversity have
-a venerable history, but those of phenotypic diversity (hereafter
-\emph{disparity}) are a comparatively more recent phenomenon. Originally
-defined as ``multidimensional morphological dissimilarity at a
-macroevolutionary scale'' (Runnegar 1987, Gould
-\href{https://paperpile.com/c/sTGYvp/Uns3}{2000}), this concept of
-disparity emerged from attempts by palaeobiologists to characterise the
-evolutionary origin of animal bodyplans and from attempts by comparative
-developmental biologists to provide causal explanations for their
-emergence. However, analyses of ``disparity'' have since expanded into
-comparative biology as a means of capturing the effect of intrinsic and
-extrinsic causal agents in morphological evolution. Typically, methods
-to capture disparity are based on multidimensional spaces where each
-dimension represents an aspect of morphological variation (a trait) and
-biological observations (taxa) can be placed in this space based on
-their trait values. Such multidimensional spaces (or morphospaces) can
-then be used to tackle a diverse array of questions that can be grouped
-into four main (non-mutually exclusive) classes (Fig. 1):
+Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
+At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g.
+cichlids and molluscs), species-rich but phenotypically conservative (e.g.
+bacteria and nematodes), species-poor but phenotypically diverse (e.g.
+Afrotheria), or depauperate in both species and phenotypic diversity (e.g.
+lungfish).
+These phenomena suggest that taxonomic and phenotypic diversity are not inextricably linked, raising important questions about how phenotypic diversity arises, such as:
+How does diversity evolve?
+Are some morphologies more common than others?
+Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve? (Mike Foote 1997, 1995)
+What role does ecology play in structuring morphological diversity?
+Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are a comparatively more recent phenomenon.
+Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (Runnegar 1987, Gould \href{https://paperpile.com/c/sTGYvp/Uns3}{@@@}),
+% Runnegar, Bruce "Rates and modes of evolution in the Mollusca." Rates of evolution. Allen and Unwin, London (1987): 39-60.
+ this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
+However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
+Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (taxa) can be placed in this space based on their trait values.
+Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):
 
-(1) \emph{Descriptive disparity.} The pioneering studies of disparity
-characterised the shapes of organisms and how they differed among groups
-\href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Foote 1995; Wills,
-Briggs, and Fortey 1994)}. These studies consist of describing
-multidimensional patterns in the diversity of morphological traits,
-addressing questions such as: why are some morphological trait
-combinations more common than others and what are the biological (or
-mathematical) properties of the resulting morphospace?
-\href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Foote 1995; Raup
-1961; Gerber 2017)}.
+(1) \emph{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Foote 1995; Wills, Briggs, and Fortey 1994)}.
+These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace?
+\href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Foote 1995; Raup 1961; Gerber 2017)}.
 
-(2) \emph{Disparity-through-time.} This approach investigates how the
-morphologies of organisms have changed on a temporal scale, focussing on
-the disparity of taxa in particular time intervals or slices. This
-approach has been used widely in palaeobiology to answer a range of
-macroevolutionary questions, such as: how does disparity accumulate over
-the history of a clade
-\href{https://paperpile.com/c/sTGYvp/ekU4+s33b}{(Guillerme and Cooper
-2018; Wright 2017)}, or how does disparity change leading up to and
-across mass extinction events
-\href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}?
+(2) \emph{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
+This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \href{https://paperpile.com/c/sTGYvp/ekU4+s33b}{(Guillerme and Cooper 2018; Wright 2017)}, or how does disparity change leading up to and across mass extinction events \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}?
 
-(3) \emph{Disparity and taxonomic diversity.} Morphological disparity
-provides another perspective on biodiversity; high morphological
-disparity represents a high diversity of morphologies (i.e. shapes or
-body plans) and is, presumably, associated with high levels of
-ecological and functional diversity (though see below). This makes
-disparity an informative complement to diversity measures based on
-species richness alone. Indeed, most studies that have investigated
-disparity and taxonomic diversity support an effective decoupling
-\href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(e.g., Fortey,
-Briggs, and Wills 1996; Ruta et al. 2013; M. J. Hopkins 2013; Moyne and
-Neige 2007)}. This approach has been used to investigate whether some
-groups are more successful than others in their exploration of new
-evolutionary strategies
-\href{https://paperpile.com/c/sTGYvp/dJHu}{(Losos 2011)}.
+(3) \emph{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e.
+shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
+This makes disparity an informative complement to diversity measures based on species richness alone.
+Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling \href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(e.g., Fortey, Briggs, and Wills 1996; Ruta et al.
+2013; M.
+J.
+Hopkins 2013; Moyne and Neige 2007)}.
+This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \href{https://paperpile.com/c/sTGYvp/dJHu}{(Losos 2011)}.
 
-(4) \emph{Disparity as a proxy for ecology.} The disparity of a group
-can be used as a proxy for either the functional role it plays within an
-ecosystem or its ecological niche. This approach assumes that groups
-with high disparity are also likely to be functionally and ecologically
-diverse, and that groups found in similar regions of shape space will
-have similar functional and ecological roles
-\href{https://paperpile.com/c/sTGYvp/EETc+tSIy+qjj9}{(Friedman 2010;
-Pierce, Angielczyk, and Rayfield 2008; P. S. L. Anderson et al. 2011)}.
+(4) \emph{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
+This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles \href{https://paperpile.com/c/sTGYvp/EETc+tSIy+qjj9}{(Friedman 2010; Pierce, Angielczyk, and Rayfield 2008; P.
+S.
+L.
+Anderson et al.
+2011)}.
 The links between form and function, however, are not always clear cut.
-Traits can be linked to multiple functions and multiple functions can be
-linked to a single trait
-\href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al. 2005)}.
-This approach has been used to investigate hypotheses of competitive
-replacement (Stephen L. Brusatte et al. 2008) and changes in ecosystem
-function during and after mass extinctions
-\href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}. It is
-particularly common in palaeobiology where it is not possible to
-directly observe the ecological or functional characteristics of extinct
-species \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.}
-2005\href{https://paperpile.com/c/sTGYvp/Ejzr}{)}.
+Traits can be linked to multiple functions and multiple functions can be linked to a single trait \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al. 2005)}.
+This approach has been used to investigate hypotheses of competitive replacement (Stephen L.
+Brusatte et al.
+2008) and changes in ecosystem function during and after mass extinctions \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}.
+It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.} 2005\href{https://paperpile.com/c/sTGYvp/Ejzr}{)}.
 
 (5) Disparity and evo-devo. Morphological disparity blalbalbal
+% AG: Although intrinsic constraints are mentioned earlier, it doesn’t enter into any of these topics explicitly, but it is certainly one of the most interesting things one can use disparity for – to bridge development and evolution, looking at developmental biases/constraints.
+This also includes integration and modularity, which are also central to the topic of disparity and evolvability.
+Although you touch on these in places, I think it's really important to incorporate it into one of these topics, as there probably isn't room to give it one of its own (although i would argue it warrants equal footing with ecology, given that these are the processes that generate variation in the first place).
+There is a massive literature on using disparity to test hypotheses of developmental constraints, and also a huge one on disparity and integration/modularity.
+Let me know if you want me to send you refs.
 
-Fundamental insights into evolutionary biology have been elicited from
-these four types of analysis. One of the most important insights is the
-discovery that morphological disparity is often greatest early in the
-evolutionary history of clades
-\href{https://paperpile.com/c/sTGYvp/xxh5+yqPw}{(Hughes, Gerber, and
-Wills 2013; Foote 1997;}
-\href{https://paperpile.com/c/sTGYvp/EPJ2}{Erwin 2007)}, indicating that
-the capacity for evolutionary innovation wanes with clade age, which
-some have argued reflects the evolutionary assembly of gene regulatory
-networks that constrain later fundamental change
-\href{https://paperpile.com/c/sTGYvp/xxh5+Z6l6}{(Hughes, Gerber, and
-Wills 2013; Erwin 2011)}. However, this example highlights one of the
-most challenging problems confronting researchers who are attempting,
-increasingly, to obtain general insights from multiple independent
-studies: can the insights gained from studies using the diversity of
-methods, approaches and data types employed be considered equivalent?
+Fundamental insights into evolutionary biology have been elicited from these five types of analysis.
+One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \href{https://paperpile.com/c/sTGYvp/xxh5+yqPw}{(Hughes, Gerber, and Wills 2013; Foote 1997;} \href{https://paperpile.com/c/sTGYvp/EPJ2}{Erwin 2007)}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change \href{https://paperpile.com/c/sTGYvp/xxh5+Z6l6}{(Hughes, Gerber, and Wills 2013; Erwin 2011)}.
+However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?
 
-In attempting to answer this question, we review current methods and
-highlight their limitations, as part of a more general attempt to
-propose best practice guidelines for studies of disparity . We first
-discuss the appropriate data required for characterising disparity, then
-review various challenging aspects of these approaches. Throughout, it
-is important to remember these tools should always be used in the
-context of a specific scientific question, as this will drive data and
-methodological choices at every stage of the process.
+In attempting to answer this question, we review current methods and highlight their limitations, as part of a more general attempt to propose best practice guidelines for studies of disparity.
+We first discuss the appropriate data required for characterising disparity, then review various challenging aspects of these approaches.
+Throughout, it is important to remember these tools should always be used in the context of a specific scientific question, as this will drive data and methodological choices at every stage of the process.
 
@@
 
-\emph{Figure disparities {[}figure\_disparities.pdf{]}: The four main
-types of disparity analysis. \textbf{Descriptive disparity} focuses on
-describing the features of morphospace occupation;
-\textbf{disparity-through-time} investigates the evolution of the
-morphospace through time including the effect of extinction events;
-\textbf{disparity and taxonomic diversity} compares different measures
-of biodiversity ; \textbf{disparity as a proxy for ecology} uses
-disparity as a proxy for the ecological or functional role of a group.
-These categories are not independent and many disparity studies will
-cover more than one.}
+\emph{Figure disparities {[}figure\_disparities.pdf{]}: The four main types of disparity analysis.
+\textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
+These categories are not independent and many disparity studies will cover more than one.}
+
+%@@@ Add disparity and evo-devo
 
@@
 
-Disparity analyses are based on traits, but traits can be characterised
-in a number of ways: (1) discrete ``cladistic'' characters, e.g. coding
-the absence or presence of features or a discrete characteristic of a
-trait \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)};
-(2) continuous measurements of features (e.g., lengths)
-\href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al.
-2011)}; or (3) more mathematical descriptors from geometric morphometric
-landmark data (e.g. Procrustes coordinates)
-\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, and
-Fourier coefficients \href{https://paperpile.com/c/sTGYvp/2Neu}{(Foote
-1989;} \href{https://paperpile.com/c/sTGYvp/ZEDR}{Spriggs et al. 2018)})
-(Fig. 2). None of these approaches are superior, but they may be more or
-less well-suited to characterising traits under comparison and to the
-question being asked of those traits (Hetherington et al 2015; Hopkins
-2017).
+Disparity analyses are based on traits, but traits can be characterised in a number of ways:
+% MH: We do not mention model-based descriptors at all, or the spaces defined by them (e.g.
+Raup's shell coiling models).
+I'm not convinced that we must, but it may come across as an notable omission to reviewers
+% SG: I agree. There's for instance Saunders et al. (2004) disparity study of Paleozoic ammonoids using Raup's shell coiling parameters
+(1) discrete ``cladistic'' characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)};
+(2) continuous measurements of features (e.g., lengths) \href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al. 2011)}; or
+(3) more mathematical descriptors from geometric morphometric landmark data (e.g.
+Procrustes coordinates) \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al.
+2017)}, and Fourier coefficients \href{https://paperpile.com/c/sTGYvp/2Neu}{(Foote 1989;} \href{https://paperpile.com/c/sTGYvp/ZEDR}{Spriggs et al.
+2018)}) (Fig.
+2).
+None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits (Hetherington et al 2015; Hopkins 2017).
 
-For example, if investigating variation of bat wing shapes, both
-homologous landmarks and continuous measurements of bones may be
-appropriate to capture patterns of wing variation. However, if the
-question focuses on comparing wings between bats and birds, different
-measurements might be more appropriate depending on the specific
-question: for example if the focus is on wing function, i.e. whether the
-aerodynamic properties of wings vary within bats or between bats and
-birds, the traits collected should reflect these aerodynamic properties
-(e.g. wingspan, aspect ratio, etc.). However, if the focus is on
-convergence between different bats and birds, it would be preferable to
-use traits that have facilitated flight in both groups (e.g. digit
-length, integumentary system, etc.). Where there is any doubt about the
-appropriate traits to choose, it may be preferable to use different
-kinds of data for the same feature to determine whether they capture the
-same pattern of disparity.
+For example, if investigating variation of bat wing shapes, both homologous landmarks and continuous measurements of bones may be appropriate to capture patterns of wing variation.
+However, if the question focuses on comparing wings between bats and birds, different measurements might be more appropriate depending on the specific question: for example if the focus is on wing function, i.e.
+whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
+wingspan, aspect ratio, etc.).
+However, if the focus is on convergence between different bats and birds, it would be preferable to use traits that have facilitated flight in both groups (e.g.
+digit length, integumentary system, etc.).
+Where there is any doubt about the appropriate traits to choose, it may be preferable to use different kinds of data for the same feature to determine whether they capture the same pattern of disparity.
 
-The points above assume that researchers are collecting their own data
-for disparity analyses, but often this is not the case. Discrete
-character data are commonly recycled from phylogenetic studies
-(e.g.Foote 1989\href{https://paperpile.com/c/sTGYvp/2Neu+0y4V}{; Deline
-et al. 2018}). This is an efficient approach to character sampling, but
-it may artifactually increase disparity between phylogenetically
-distinct groups because phylogenetic characters are often collected to
-discriminate among groups
-\href{https://paperpile.com/c/sTGYvp/fTJ3}{(Foote 1995)}. This needs to
-be considered when interpreting results, especially as synapomorphies
-will naturally lead to an apparent shift or increase in disparity when
-new clades appear. Furthermore, many datasets are limited to subsets of
-anatomy that are at least implicit samples of overall anatomy, but
-explicit tests of this assumption have shown that different aspects of
-morphology can exhibit different patterns of disparity
-\href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}.
-This effect of anatomical part on disparity patterns can be especially
-challenging when working on datasets where the available data has
-non-random missing anatomical parts, such as the absence of soft tissue
-in the fossil record \href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline
-et al. 2018)}.
+The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
+Discrete character data are commonly recycled from phylogenetic studies (e.g.Foote 1989\href{https://paperpile.com/c/sTGYvp/2Neu+0y4V}{; Deline et al. 2018}).
+% SG: Foote (1989) is a study of trilobite disparity from Fourier analysis of cranidia. I don't think it has anything to do with discrete data or phylogenetic studies
+% MH: Also the Foote papers that used discrete data weren't taken from phylogenetic studies but collected for the specific purpose of the project -- I think this might be true for the Deline paper as well.
+But there are many studies out there that used matrices initially intended for phylogenetics.
+I can look for some if needed.
+This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups \href{https://paperpile.com/c/sTGYvp/fTJ3}{(Foote 1995)}.
+This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
+Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J.
+Hopkins 2017)}.
+This effect of anatomical part on disparity patterns can be especially challenging when working on datasets where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline et al.
+2018)}.
 
-Trait data suffer from the same shortcomings as most biological datasets
--- data can be missing, non-overlapping, hierarchical, inapplicable,
-ambiguous, polymorphic, correlated, or there may be an insufficient
-sample size \href{https://paperpile.com/c/sTGYvp/Yrbg}{(Brazeau,
-Guillerme, and Smith 2017};
-\href{https://paperpile.com/c/sTGYvp/yO2t}{Palci and Lee 2018)}.
-Biological phenomena such as allometry and sexual dimorphism may also
-influence trait data. More mundanely, data collection is constrained by
-the time and money available, making collating a ``perfect'' dataset
-difficult. Ultimately, disparity analyses characterise the data, and
-subsamples of the universe of possible data may not have the power to
-uncover holistic patterns of disparity. Therefore, trait data should be
-collected with the question in mind, or the question asked should be
-tailored to the limits of the data available.
+Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \href{https://paperpile.com/c/sTGYvp/Yrbg}{(Brazeau, Guillerme, and Smith 2017}; \href{https://paperpile.com/c/sTGYvp/yO2t}{Palci and Lee 2018)}.
+Biological phenomena such as allometry and sexual dimorphism may also influence trait data.
+More mundanely, data collection is constrained by the time and money available, making collating a ``perfect'' dataset difficult.
+Ultimately, disparity analyses characterise the data, and subsamples of the universe of possible data may not have the power to uncover holistic patterns of disparity.
+Therefore, trait data should be collected with the question in mind, or the question asked should be tailored to the limits of the data available.
 
@@
 
-\emph{Figure data {[}figure\_data.pdf{]}: Major routes to obtain
-morphological data for disparity analyses. Data can be collected as
-discrete trait observations (e.g. presence or absence data) or as
-continuous data. Continuous data can be collected by various methods
-including linear measurements and landmark coordinates or contours
-(curves). These measurements can then be mathematically transformed
-(logarithm transforms, scaling, Procrustes superimposition, elliptic
-Fourier transforms, etc.). Regardless of the method, data collection
-produces a trait matrix where the observed traits constitute columns and
-the studied elements (generally taxa or OTUs) the rows.}
+\emph{Figure data {[}figure\_data.pdf{]}: Major routes to obtain morphological data for disparity analyses.
+Data can be collected as discrete trait observations (e.g.
+presence or absence data) or as continuous data.
+Continuous data can be collected by various methods including linear measurements and landmark coordinates or contours (curves).
+These measurements can then be mathematically transformed (logarithm transforms, scaling, Procrustes superimposition, elliptic Fourier transforms, etc.).
+Regardless of the method, data collection produces a trait matrix where the observed traits constitute columns and the studied elements (generally taxa or OTUs) the rows.}
 
@@
 
-Once suitable trait data have been collected, the design of the
-disparity analysis itself needs to be considered. Study design
-encompasses several key aspects including (a) the difficulty of dealing
-with multidimensional data; (b) the variety of ordination (dimension
-reduction) techniques available and their limitations; (c) the metrics
-used to summarise the relative disparity of groups; (d) the methods used
-for hypothesis testing within the disparity analysis framework; and (e)
-ancestral state estimations in disparity-through-time analyses. We
-consider these aspects in turn below.
+Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
+Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the variety of ordination (dimension reduction) techniques available and their limitations; (c) the metrics used to summarise the relative disparity of groups; (d) the methods used for hypothesis testing within the disparity analysis framework; and (e) ancestral state estimations in disparity-through-time analyses.
+% @@@ Check the numbering here
+We consider these aspects in turn below.
 

### 2020-02-18 Writing/disparity_review.tex
reformated for tex compatible

@@
-RH: disparate disparity analyses.
+\documentclass[12pt,letterpaper]{article}
+\usepackage{natbib}
 
-\textbf{Disparities in the analysis of morphological disparity}
+%Packages
+\usepackage{fixltx2e}
+\usepackage{textcomp}
+\usepackage{fullpage}
+\usepackage{float}
+\usepackage{latexsym}
+\usepackage{url}
+\usepackage{epsfig}
+\usepackage{graphicx}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{mathtools}
+\usepackage{bm}
+\usepackage{array}
+\usepackage[version=3]{mhchem}
+\usepackage{ifthen}
+\usepackage{caption}
+\usepackage{hyperref}
+\usepackage{amsthm}
+\usepackage{amstext}
+\usepackage{enumerate}
+\usepackage[osf]{mathpazo}
+\usepackage{dcolumn}
+\usepackage{lineno}
+\usepackage{pdflscape}
 
-Thomas Guillerme\({}^{1,*, +}\), Natalie Cooper\({}^{2, +}\), Stephen L.
-Brusatte\({}^{3}\), Katie E. Davis\({}^{4}\), Andrew L
-Jackson\({}^{5}\), Sylvain Gerber\({}^{6}\), Anjali Goswami\({}^{2}\),
-Kevin Healy\({}^{7}\), Melanie Hopkins\({}^{8}\), Marc EH
-Jones\({}^{9}\), Graeme T. Lloyd\({}^{10}\), Joseph E.
-O'Reilly\({}^{11}\), Abi Pate\({}^{11}\), Mark N Puttick\({}^{12}\),
-Emiliy Rayfield\({}^{11}\), Erin E. Saupe\({}^{13}\), Emma
-Sherratt\({}^{14}\), Graham Slater\({}^{15}\), Vera
-Weisbecker\({}^{1}\), Gavin H Thomas\({}^{16}\) and Philip
-Donoghue\({}^{11}\)\\
-\({}^{1}\)\emph{School of Biological Sciences, University of Queensland,
-St. Lucia, Queensland, Australia.}\\
-\({}^{2}\)\emph{Department of Life Sciences, Natural History Museum,
-London, Cromwell Road, London, SW7 5BD, UK.}
+\usepackage{color,soul}
 
-\emph{\textsuperscript{3}School of GeoSciences, University of Edinburgh,
-Grant Institute, Edinburgh EH9 3FE, UK}
+\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
+\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
+\newcolumntype{d}[1]{D{.}{.}{#1}}
 
-\textsuperscript{4} Department of Biology, University of York, YO10 5DD
+\pagenumbering{arabic}
 
-\textsuperscript{5} Department of Zoology, School of Natural Sciences,
-Trinity College.
+%Pagination style and stuff
+\linespread{2}
+\raggedright
+\setlength{\parindent}{0.5in}
+\setcounter{secnumdepth}{0} 
+\renewcommand{\section}[1]{%
+\bigskip
+\begin{center}
+\begin{Large}
+\normalfont\scshape #1
+\medskip
+\end{Large}
+\end{center}}
+\renewcommand{\subsection}[1]{%
+\bigskip
+\begin{center}
+\begin{large}
+\normalfont\itshape #1
+\end{large}
+\end{center}}
+\renewcommand{\subsubsection}[1]{%
+\vspace{2ex}
+\noindent
+\textit{#1.}---}
+\renewcommand{\tableofcontents}{}
+%\bibpunct{(}{)}{;}{a}{}{,}
 
-\textsuperscript{6} Institut de Systématique, Évolution, Biodiversité
-(ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne
-Université, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005
-Paris, France.
+%---------------------------------------------
+%
+%       START
+%
+%---------------------------------------------
 
-\textsuperscript{7} Ryan Institute, School of Natural Sciences, National
-University of Ireland Galway, Ireland.
+\begin{document}
+%Running head
+\begin{flushright}
+Version dated: \today
+\end{flushright}
 
-\textsuperscript{10} \emph{School of Earth and Environment, University
-of Leeds, Leeds LS2 9JT, UK}\\
-\textsuperscript{11}\emph{School of Earth Sciences, University of
-Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK}
+\bigskip
+\noindent RH: disparate disparity analyses
+\bigskip
+\medskip
+\begin{center}
+\noindent{\Large \bf Disparities in the analysis of morphological disparity}
+\bigskip
 
-\textsuperscript{12} \emph{Milner Centre for Evolution, University of
-Bath, BA2 7AY}
+Thomas Guillerme$^{1,16,*,+}$,
+Natalie Cooper$^{2,+}$,
+Stephen L. Brusatte$^{3}$,
+Katie E. Davis$^{4}$,
+Andrew L Jackson$^{5}$,
+Sylvain Gerber$^{6}$,
+Anjali Goswami$^{2}$,
+Kevin Healy$^{7}$,
+Melanie Hopkins$^{8}$,
+Marc EH Jones$^{9}$,
+Graeme T. Lloyd$^{10}$,
+Joseph E. O'Reilly$^{11}$,
+Abi Pate$^{11}$,
+Mark N Puttick$^{12}$,
+Emiliy Rayfield$^{11}$,
+Erin E. Saupe$^{13}$,
+Emma Sherratt$^{14}$,
+Graham Slater$^{15}$,
+Vera Weisbecker$^{1}$,
+Gavin H Thomas$^{16}$
+and Philip Donoghue$^{11}$\\
 
-\textsuperscript{13} \emph{Department of Earth Sciences, University of
-Oxford, S Parks Road, Oxford OX1 3AN, UK}
+\noindent {\small \it 
+$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
+$^2$IDepartment of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
+$^{3}$School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
+$^{4}$Department of Biology, University of York, YO10 5DD, UK;
+$^{5}$Department of Zoology, School of Natural Sciences, Trinity College, Ireland;
+$^{6}$Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
+$^{7}$Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
+$^{8}$ @@@;
+$^{10}$School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
+$^{11}$School of Earth Sciences, University of Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
+$^{12}$Milner Centre for Evolution, University of Bath, BA2 7AY UK;
+$^{13}$Department of Earth Sciences, University of Oxford, S Parks Road, Oxford OX1 3AN, UK;
+$^{14}$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
+$^{15}$ @@@;
+$^{16}$ @@@;\\
+$^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding author: guillert@tcd.ie\\}
 
-\textsuperscript{14} \emph{School of Biological Sciences, The University
-of Adelaide, Adelaide, South Australia 5005, Australia}
-
-+ These authors contributed equally to the manuscript.
-
-*\textbf{Corresponding author:} \emph{guillert@tcd.ie}
-
-\textbf{NOTE TO ALL THE CO-AUTHORS:}
-
-\textbf{-The word limit for this is 5000 words so we can't add any large
-sections.}
-
-\textbf{-There is no reference limit per se but please don't add too
-many. I will sort out the references later so don't correct the
-formatting of them here.}
-
-\textbf{-Add your affiliation in the author list above}. The
-affiliations count in the word count so please 1) keep it short (max 1
-per person!) and 2) try to use one that's already there. Also add any
-initials I've missed in your names.
-
-- If in doubt, check formatting here:
-\href{https://royalsociety.org/journals/authors/author-guidelines/}{{https://royalsociety.org/journals/authors/author-guidelines/}}
-
-Thanks!
+\end{center}
 
@@
 
-\includegraphics[width=6.5in,height=4.94444in]{media/image1.png}
+%\includegraphics[width=6.5in,height=4.94444in]{media/image1.png}
 
@@
 
-\includegraphics[width=5.63542in,height=3.875in]{media/image2.png}3.
+%\includegraphics[width=5.63542in,height=3.875in]{media/image2.png}3.
 
@@
 
-% Stopped p.13
-
 % \hypertarget{a-multidimensionality}{%
@@
 
-\includegraphics[width=6.5in,height=5.26389in]{media/image3.png}
+%\includegraphics[width=6.5in,height=5.26389in]{media/image3.png}
 
-\emph{Figure methods: Morphospaces: different mathematical
-representations of a} \emph{morphospace. A trait matrix can be
-transformed into a distance matrix (e.g. in Close et al., 2015) or an
-ordinated matrix (e.g. in Brusatte et al.} \emph{2008). In each case,
-all three matrices represent the morphospace for the observations at
-hand. Visualisation: different ways to represent the morphospace in 2D.
-Visualisations can use either trait plots (directly from the trait
-matrix); ``flattening'' ordination of the data (e.g. using
-multidimensional scaling} \emph{forcing all the variance to be contained
-in two dimensions); or ordination axis plots (directly from the
-ordinated matrix).}
+\emph{Figure methods: Morphospaces: different mathematical representations of a} \emph{morphospace.
+% SG: The morphospace is the original trait space within which lie the specimens.
+Then one performs an ordination to visualize the space.
+So I would distinguish the morphospace from the ordination space.
+See also Mitteroecker and Huttegger (2009)
+A trait matrix can be transformed into a distance matrix (e.g. in Close et al., 2015) or an ordinated matrix (e.g. in Brusatte et al.} \emph{2008).
+In each case, all three matrices represent the morphospace for the observations at hand.
+Visualisation: different ways to represent the morphospace in 2D.
+Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g.
+using multidimensional scaling} \emph{forcing all the variance to be contained in two dimensions)
+% SG: No, the flattening will not force all the variance to be contained in 2D.
+First because NMDS is non metric and variance is therefore not preserved as such.
+Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation.
+Some information (variance) is thus necessarily lost.
+; or ordination axis plots (directly from the ordinated matrix).}
+% ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
+% SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..).
+But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
+% @@@ Change references
 
 \hypertarget{b-disparity-indices}{%
+% @@@ Change to indices throughout
 \subsection{(b) Measuring disparity}\label{c-disparity-indices}}
 
-Most disparity datasets are multidimensional and, consequently, a large
-component of any disparity analysis involves considering how to extract
-a meaningful (i.e. interpretable) summary of disparity. This is usually
-achieved with a disparity metric or index
-\href{https://paperpile.com/c/sTGYvp/vTHS}{(Melanie J. Hopkins and
-Gerber 2017)}. As with any summary of multidimensional data, disparity
-metrics will reflect only some aspects of the morphological variation,
-never its whole complexity {[}metrics paper{]}. It is therefore often
-beneficial to use more than one metric to summarise different aspects of
-variation, guided by the aim of the study.
+Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
+This is usually achieved with a disparity metric or index \href{https://paperpile.com/c/sTGYvp/vTHS}{(Melanie J. Hopkins and Gerber 2017)}.
+As with any summary of multidimensional data, disparity metrics will reflect only some aspects of the morphological variation, never its whole complexity {[}metrics paper{]}.
+It is therefore often beneficial to use more than one metric to summarise different aspects of variation, guided by the aim of the study.
 
-When considering only one dimension, disparity metrics can be used to
-reflect the spread of the distribution (e.g. the range, quantiles or
-variance) or its central tendency (i.e., mean, median or mode). Among
-these metrics, some will have more attractive properties than others,
-such as sensitivity to outliers. Range and mean are highly sensitive,
-whereas quantiles, variance and median are less so, making them more or
-less appropriate for different questions. For example, if we want to
-characterise the extent of morphospace occupied by a group (e.g. does
-group A occupy as much space as group B?), metrics related to the spread
-of the group in the morphospace are most appropriate (e.g. volume
-\href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, distance
-from the centroid
-\href{https://paperpile.com/c/sTGYvp/vTHS+yyNa}{(Melanie J. Hopkins and
-Gerber 2017; Finlay and Cooper 2015)}, variance and range
-\href{https://paperpile.com/c/sTGYvp/tGyd}{(S. L. Brusatte et al.
-2008)}). Conversely, if we wish to describe the ``position'' of a group
-in a morphospace (e.g. does group A occupy the same space as group B?),
-metrics related to the distance between the elements within a group and
-a fixed point in the morphospace are most appropriate
-(e.g.\href{https://paperpile.com/c/sTGYvp/u1KE}{(Guillerme et al.,
-n.d.)}). Finally, if we aim to characterise the density of morphospace
-occupation (e.g. is group A more closely packed than group B?) metrics
-related to the pairwise distances between elements will be most
-appropriate (e.g. nearest neighbour distance, pairwise distances, etc.
-\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}).
+When considering only one dimension, disparity metrics can be used to reflect the spread of the distribution (e.g.
+the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
+Among these metrics, some will have more attractive properties than others, such as sensitivity to outliers.
+Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
+For example, if we want to characterise the extent of morphospace occupied by a group (e.g.
+does group A occupy as much space as group B?), metrics related to the spread of the group in the morphospace are most appropriate (e.g.
+volume \href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al.
+2016)}, distance from the centroid \href{https://paperpile.com/c/sTGYvp/vTHS+yyNa}{(Melanie J.
+Hopkins and Gerber 2017; Finlay and Cooper 2015)}, variance and range \href{https://paperpile.com/c/sTGYvp/tGyd}{(S.
+L.
+Brusatte et al.
+2008)}).
+Conversely, if we wish to describe the ``position''
+% SG: Note that conceptually and technically,  disparity is about variation.
+So information on position is another aspect of morphospace occupation.
+but it is distinct from the meaning of  disparity as originally intended.
+ of a group in a morphospace (e.g.
+does group A occupy the same space as group B?), metrics related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate (e.g.\href{https://paperpile.com/c/sTGYvp/u1KE}{(Guillerme et al., n.d.)}).
+Finally, if we aim to characterise the density of morphospace occupation (e.g.
+is group A more closely packed than group B?) metrics related to the pairwise distances between elements will be most appropriate (e.g.
+nearest neighbour distance, pairwise distances, etc.
+\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
+2015)}).
+%ES: this para is probably most useful to refer to for the ordinate or not part
 
-In addition to considering what properties of disparity metrics should
-capture, it is important to also consider the mathematical properties of
-the metrics and their associated caveats
-\href{https://paperpile.com/c/sTGYvp/nFf7+ROH8}{(Wills 2001; Ciampaglio,
-Kemp, and McShea 2001)}. For example, measuring the full sum of the
-variance in each dimension of the space does not require we add the
-covariance between the axes in an ordinated space using a PCA. However,
-this is not true of other mathematical spaces or when not all dimensions
-or elements are considered, even in a PCA {[}Legendre \& Legendre
-2012{]}.
+In addition to considering what properties of disparity metrics should capture, it is important to also consider the mathematical properties of the metrics and their associated caveats \href{https://paperpile.com/c/sTGYvp/nFf7+ROH8}{(Wills 2001; Ciampaglio, Kemp, and McShea 2001)}.
+For example, measuring the full sum of the variance in each dimension of the space does not require we add the covariance between the axes in an ordinated space using a PCA.
+% SG: Not very clear to me
+However, this is not true of other mathematical spaces or when not all dimensions or elements are considered, even in a PCA {[}Legendre \& Legendre 2012{]}.
 
-Furthermore, multidimensional spaces also have some counter-intuitive
-properties that need to be considered such as the ``curse of
-dimensionality'' \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman
-1966)}. In spaces with some axis of variance lower than 1, product-based
-metrics used as proxies of volumes (e.g. product of ranges, hypervolume,
-hypercube, etc.) can tend towards zero fairly quickly for spaces with
-even a modest number of dimensions
-\href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}; Donoho
-2000). Some other types of metrics are also extremely sensitive to
-outliers and can be biased easily by sample size, for example range
-\href{https://paperpile.com/c/sTGYvp/aSSL}{(}Butler et al.
-2012\href{https://paperpile.com/c/sTGYvp/aSSL}{)} or convex hull based
-metrics \href{https://paperpile.com/c/sTGYvp/aSSL+PwyQ}{(Butler et al.
-2012; Jackson et al. 2011)}.
+Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}.
+In spaces with some axis of variance lower than 1, product-based metrics used as proxies of volumes (e.g.
+product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}; Donoho 2000).
+Some other types of metrics are also extremely sensitive to outliers and can be biased easily by sample size, for example range \href{https://paperpile.com/c/sTGYvp/aSSL}{(}Butler et al.
+2012\href{https://paperpile.com/c/sTGYvp/aSSL}{)}
+%SG: Or even better, twenty years earlier, Foote (1992), who orginally described the problem in the context of disparity and suggested a rarefaction approach.
+or convex hull based metrics \href{https://paperpile.com/c/sTGYvp/aSSL+PwyQ}{(Butler et al. 2012; Jackson et al. 2011)}.
 
@@
 
-No matter which disparity metrics have been calculated, the research
-question must be framed in an appropriate statistical context. The
-multidimensional statistical toolkit for ecology and evolution has been
-greatly expanded in recent years
-\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer}
-2018\href{https://paperpile.com/c/sTGYvp/ZnDd}{)}, but some of these
-advances have yet to be implemented in disparity analyses. Instead,
-hypothesis testing has mostly been confined to a small set of
-well-established methods. One commonly used test is the non-parametric
-permutation analysis of variance
-\href{https://paperpile.com/c/sTGYvp/3hy2+SC6L}{(M. J. Anderson and
-Walsh 2013; M. J. Anderson 2001)}, an analysis of variance (ANOVA) of
-the pairwise distances between different groups. Although statistically
-valid, this test is not always directly related to the hypothesis under
-question. For example, PERMANOVA tests whether two groups share the same
-variance/covariance in a ``distance-space''. This is not the same as
-testing whether the two groups overlap in morphospace. Statistical tests
-should be employed that are tailored to the question at hand, rather
-than simply following common practices.
+No matter which disparity metrics have been calculated, the research question must be framed in an appropriate statistical context.
+The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer} 2018\href{https://paperpile.com/c/sTGYvp/ZnDd}{)}
+%AG: there are others that should be mentioned - mvmorph for example, which is especially important given that there are extremely different approaches to multivariate data in these different packages
+, but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
+One commonly used test is the non-parametric permutation analysis of variance \href{https://paperpile.com/c/sTGYvp/3hy2+SC6L}{(M.
+J.
+Anderson and Walsh 2013; M.
+J.
+Anderson 2001)}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
+Although statistically valid, this test is not always directly related to the hypothesis under question.
+For example, PERMANOVA tests whether two groups share the same variance/covariance in a ``distance-space''.
+% SG: What is meant by distance-sapce and in what way is it distinguished from the spaces discussed above?
+This is not the same as testing whether the two groups overlap in morphospace.
+%SG: You're sure of this?
+Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.
 
-It is also important to consider which data should be subjected to a
-statistical test. For example, in morphological disparity analysis,
-especially for palaeobiological questions, data are often bootstrapped.
-This has two advantages: (i) when the disparity metric is unidimensional
-(e.g. the sum of variances), bootstrapping the data generates a
-distribution of the metric that can be analysed using the vast
-statistical toolkit available for comparing distributions; (ii) when
-data are scarce, bootstrapping the data allows users to introduce
-variance, rendering the test less sensitive to outliers. However,
-bootstrapped data are pseudoreplicates and thus non-independent. This
-violates the assumptions of most parametric statistical tests.
-Furthermore, the number of bootstrap pseudoreplicates will inevitably
-increase the Type I error rate. These factors are often ignored in
-disparity analyses.
+It is also important to consider which data should be subjected to a statistical test.
+% AG: It would be good to also note here that data should be comparable in terms of having a common reference frame - e.g.
+comparing disparities across elements with different numbers of landmarks or different landmark configurations that haven't been/can't be superimposed together is problematic but is popping up in the disparity literature
+For example, in morphological disparity analysis, especially for palaeobiological questions, data are often bootstrapped.
+This has two advantages: (i) when the disparity metric is unidimensional (e.g.
+the sum of variances), bootstrapping the data generates a distribution of the metric that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
+However, bootstrapped data are pseudoreplicates and thus non-independent.
+This violates the assumptions of most parametric statistical tests.
+Furthermore, the number of bootstrap pseudoreplicates will inevitably increase the false positive rate (Type I error).
+% @@@ check positive/negative type
+These factors are often ignored in disparity analyses.
+% VW: Surely there's a reference for this?
+% MH: I might be wrong about this, but my impression is that CIs generated from bootstrapping are compared visually (like to assess significance in changes in disparity over time), but that statistical tests are not commonly applied.
 
@@
 
-As with all comparative datasets, the data used in disparity analyses
-are not independent because close relatives will tend to have more
-similar morphologies than more distant relatives
-\href{https://paperpile.com/c/sTGYvp/WXik}{(Harvey and Pagel 1998)}.
+As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \href{https://paperpile.com/c/sTGYvp/WXik}{(Harvey and Pagel 1998)}.
 Thus, for many disparity analyses, phylogenetic relationships should be
-taken into account. However, it has been noted that some popular
-phylogenetic correction methods (like pPCA) can be inappropriate if
-incorrect assumptions are made about the data (Uyeda et al).
-\href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)}
-provide a thorough review of multivariate phylogenetic comparative
-methods, and so we do not consider them further here.
+taken into account.
+% SG: Why?
+However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate
+% SG: In what way?
+if incorrect assumptions are made about the data (Uyeda et al). \href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)}
+% @@@ref: https://academic.oup.com/sysbio/article/64/4/677/1649888
+% SG: Which assumption? This part (e) seems a bit vague to me
+% AG: as long as you are using most or ideally all of the variance in a dataset, the rotation of pPCA should have little to no impact on disparity analyses. polly et al 2013 i think
+provide a thorough review of multivariate phylogenetic comparative methods, and so we do not consider them further here.
+% AG: Given this is given quite short treatment, i think if you need room, you could cut it altogether or just have a sentence in another section stating this even more briefly
 
@@
 
-Disparity-through-time analyses often use ancestral state estimation to
-extract disparity estimates for non-sampled taxa and/or nodes of a
-phylogeny. Ancestral state estimation can be performed at two points in
-the disparity analysis pipeline: either (1) pre-ordination, i.e. the
-estimation is done before transformation of the data (e.g. ordination,
-or distance matrix construction) and is simply based on the original
-data; or (2) post-ordination, i.e. the estimation is done after
-transformation of the data by estimating the ancestral states using the
-transformed matrix (e.g. the ordination scores)
-\href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
+Disparity-through-time analyses often use ancestral state estimation to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
+Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-ordination, i.e.
+the estimation is done before transformation of the data (e.g.
+ordination, or distance matrix construction) and is simply based on the original data; or (2) post-ordination, i.e.
+the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g.
+the ordination scores) \href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
 
-Pre-ordination ancestral state estimation will change the way the
-ordination space is defined -- i.e. the relationship between the points
-are not yet estimated -- and requires longer computational times.
+Pre-ordination ancestral state estimation will change the way the ordination space is defined -- i.e.
+the relationship between the points are not yet estimated -- and requires longer computational times.
 However, once the morphospace is defined its properties will not change.
-Post-ordination ancestral state estimation will not change the empirical
-ordination space and is faster to compute, but it will add elements in
-the space, whose estimated positions can be problematic for statistical
-tests and evolutionary inferences down the line
-\href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
+Post-ordination ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
 
-No ancestral state estimation method is without drawbacks and above all
-else are highly dependent on the data and method used. In general, using
-ancestral state estimation can help with recovering patterns of changes
-in disparity but should not be used simply to generate extra data points
-to increase statistical power. In fact, these extra points are not
-independent and can also have problematic side effects, especially when
-testing for the influence of mass extinctions on disparity.
+No ancestral state estimation method is without drawbacks
+% @@@ Maybe check in Pennel’s “fossils are needed” paper for ref.
+ and above all else are highly dependent on the data and method used.
+In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points to increase statistical power.
+In fact, these extra points are not independent and can also have problematic side effects, especially when testing for the influence of mass extinctions on disparity.
 
@@
 
-Morphological disparity analyses are widely employed in evolutionary
-palaeobiology, and they are based on a diversity of methods and data.
-There is no ``one-size-fits-all'' pipeline for morphological disparity
-analyses. As with any multidimensional analysis, there are many
-variables that have to be considered when deciding which data to use and
-how to analyse it, stemming from the explicit hypotheses being tested.
-Many of the problems in morphological disparity analysis arise from
-``blind'' application of established methodological pipelines without
-consideration of the biological question being addressed. We advocate
-the bespoke assembly of analytic approaches and, given the computational
-efficiency of these methods, an experimental approach that explores the
-impact of competing approaches, such as choice of distance measure,
-ordination method and ancestral state estimation method on disparity
-analysis results. Many of the methods employed in disparity analysis are
-used more widely in other fields, including genomics and ecology, which
-also encompass analyses of multidimensional datasets
-\href{https://paperpile.com/c/sTGYvp/krNU+60H0+cV3v}{(Donohue et al.
-2013; Canter et al. 2018; Saupe et al. 2015)}. Innovations in
-morphological disparity analyses likely await discovery in their
-respective literatures.
+Morphological disparity analyses are widely employed in evolutionary palaeobiology, and they are based on a diversity of methods and data.
+There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
+As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse it, stemming from the explicit hypotheses being tested.
+Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
+We advocate the bespoke
+%VW: I think there could be just a line or two that takes makes "bespoke design" less daunting.
+There are loads of well-documented R packages available in morphological and ecological studies that can be relatively straightforwardly adapted.
+assembly of analytic approaches and, given the computational efficiency of these methods, an experimental approach that explores the impact of competing approaches, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
+Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \href{https://paperpile.com/c/sTGYvp/krNU+60H0+cV3v}{(Donohue et al.
+2013; Canter et al.
+2018; Saupe et al.
+2015)}.
+% @@@ Add Mammola
+Innovations in morphological disparity analyses likely await discovery in their respective literatures.
+% AG: as above, i think you need to add in evodevo work here, it's really fundamental to disparity
 
-While studies of morphological disparity would benefit from advances in
-multidimensional analysis in other fields, the concept of a morphospace
-could reciprocally benefit other disciplines. For example, the
-multidimensional analysis of Diaz
-\href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, which
-analysed patterns of form and function in plants, is essentially an
-ecomorphospace; isotopic analyses of organisms
-\href{https://paperpile.com/c/sTGYvp/PwyQ}{(Jackson et al. 2011)} can be
-represented as an isotope-space; ecosystem functioning in Donohue et al.
-(2013) as an ecosystem-space, etc. These generalisations could also be
-exported for any set of traits (e.g. acousto-spaces for acoustic traits
-or glotto-spaces for linguistic traits). Cognate approaches have been
-adopted recently in the analysis of single cell comparative
-transcriptome data
-\href{https://paperpile.com/c/sTGYvp/856K}{(Sebé-Pedrós et al. 2018)}
-where interpretation of the resulting transcriptome-spaces would be
-improved by heeding the concerns we highlight concerning morphospaces.
+While studies of morphological disparity would benefit from advances in multidimensional analysis in other fields, the concept of a morphospace could reciprocally benefit other disciplines.
+For example, the multidimensional analysis of Diaz \href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al.
+2016)}, which analysed patterns of form and function in plants, is essentially an ecomorphospace; isotopic analyses of organisms \href{https://paperpile.com/c/sTGYvp/PwyQ}{(Jackson et al.
+2011)} can be represented as an isotope-space; ecosystem functioning in Donohue et al.
+(2013) as an ecosystem-space, etc.
+% AJ: this is a potential nice, although rather data hungry I believe, methods that I think could be worth dropping into the paper somewhere.
+Either here as a quick drop, or earlier up under overlap seems appropriate if you want to add it https://esajournals.onlinelibrary.wiley.com/doi/10.1890/14-0235.1
+These generalisations could also be exported for any set of traits (e.g. acousto-spaces for acoustic traits or glotto-spaces for linguistic traits).
+%SG: devising 
+Cognate approaches have been adopted recently in the analysis of single cell comparative transcriptome data \href{https://paperpile.com/c/sTGYvp/856K}{(Sebé-Pedrós et al.
+2018)} where interpretation of the resulting transcriptome-spaces would be improved by heeding the concerns we highlight concerning morphospaces.
+% ES: Suggested by Erin: reword this sentence
 
-Although disparity analyses are now simple to implement in freely
-available softwares
-\href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{(e.g.,
-Guillerme 2018;} Adams and Otárola-Castillo
-2013\href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{;
-Bouxin 2005; Harmon et al. 2008; Lloyd 2016; Navarro 2003; Dixon 2003)},
-it is crucial to remember that they are multidimensional analyses and
-multidimensional analyses are complex. We assert that future
-morphological analyses will benefit by emphasising the methodological
-decisions made, rather than simply using disparity analysis because
-\emph{we can}.
+Although disparity analyses are now simple to implement in freely available softwares \href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{(e.g., Guillerme 2018;} {; Bouxin 2005; Harmon et al.
+2008; Lloyd 2016; Navarro 2003; Oksanen 2007)},
+% @@@ add BAT
+it is crucial to remember that they are multidimensional analyses and multidimensional analyses are complex.
+We assert that future morphological analyses will benefit by emphasising the methodological decisions made, rather than simply using disparity analysis because \emph{we can}.
 
@@
 
-TG, NC and PD proposed this review; TG and NC led the writing supported
-by PD and GT. All authors edited drafts and approved the final version.
+TG, NC and PD proposed this review; TG and NC led the writing supported by PD and GT. All authors edited drafts and approved the final version.
 
@@
 
-This article results from discussion at the Royal Society International
-Science Seminar on `Reconciling disparate views on disparity' held at
-Chicheley Hall, 9-10th January, 2018.
+This article results from discussion at the Royal Society International Science Seminar on `Reconciling disparate views on disparity' held at Chicheley Hall, 9-10th January, 2018.
 
--AG was funded by European Research Council Starting grant 637171
-ADaPTiVE.
+-AG was funded by European Research Council Starting grant 637171 ADaPTiVE.
 
--ALJ was funded by an Irish Research Council Laureate Award
-IRCLA/2017/186. ES was funded by a University of Adelaide Research
-Fellowship;
+-ALJ was funded by an Irish Research Council Laureate Award IRCLA/2017/186. ES was funded by a University of Adelaide Research Fellowship;
 
@@
 
--PD was funded by NERC (NE/P013678/1; NE/N002067/1) and BBSRC
-(BB/N000919/1);
+-PD was funded by NERC (NE/P013678/1; NE/N002067/1) and BBSRC (BB/N000919/1);
 
--SB was funded by European Research Council (ERC) under the European
-Union's Horizon 2020 research and innovation program (grant agreement No
-756226, ERC Starting Grant: PalM) and a Leverhulme Trust Research
-Project Grant (RPG-2017-167);
+-SB was funded by European Research Council (ERC) under the European Union's Horizon 2020 research and innovation program (grant agreement No 756226, ERC Starting Grant: PalM) and a Leverhulme Trust Research Project Grant (RPG-2017-167);
 
@@
 
-Donoho, D. L. (2000). High-dimensional data analysis: The curses and
-blessings of dimensionality. \emph{AMS math challenges lecture},
-\emph{1}(2000), 32.
+\end{document}
 
-References
+% Donoho, D. L. (2000). High-dimensional data analysis: The curses and
+% blessings of dimensionality. \emph{AMS math challenges lecture},
+% \emph{1}(2000), 32.
 
-\href{http://paperpile.com/b/sTGYvp/ZnDd}{Adams, Dean C., and Michael L.
-Collyer. 2018. ``Multivariate Phylogenetic Comparative Methods:
-Evaluations, Comparisons, and Recommendations.'' \emph{Systematic
-Biology} 67 (1): 14--31.}
+% References
 
-\href{http://paperpile.com/b/sTGYvp/J2G1}{Adams, Dean C., and Erik
-Otárola-Castillo. 2013. ``Geomorph: Anrpackage for the Collection and
-Analysis of Geometric Morphometric Shape Data.'' \emph{Methods in
-Ecology and Evolution / British Ecological Society} 4 (4): 393--99.}
+% \href{http://paperpile.com/b/sTGYvp/ZnDd}{Adams, Dean C., and Michael L.
+% Collyer. 2018. ``Multivariate Phylogenetic Comparative Methods:
+% Evaluations, Comparisons, and Recommendations.'' \emph{Systematic
+% Biology} 67 (1): 14--31.}
 
-\href{http://paperpile.com/b/sTGYvp/SC6L}{Anderson, Marti J. 2001. ``A
-New Method for Non-Parametric Multivariate Analysis of Variance.''
-\emph{Austral Ecology} 26 (1): 32--46.}
+% \href{http://paperpile.com/b/sTGYvp/J2G1}{Adams, Dean C., and Erik
+% Otárola-Castillo. 2013. ``Geomorph: Anrpackage for the Collection and
+% Analysis of Geometric Morphometric Shape Data.'' \emph{Methods in
+% Ecology and Evolution / British Ecological Society} 4 (4): 393--99.}
 
-\href{http://paperpile.com/b/sTGYvp/3hy2}{Anderson, Marti J., and Daniel
-C. I. Walsh. 2013. ``PERMANOVA, ANOSIM, and the Mantel Test in the Face
-of Heterogeneous Dispersions: What Null Hypothesis Are You Testing?''
-\emph{Ecological Monographs} 83 (4): 557--74.}
+% \href{http://paperpile.com/b/sTGYvp/SC6L}{Anderson, Marti J. 2001. ``A
+% New Method for Non-Parametric Multivariate Analysis of Variance.''
+% \emph{Austral Ecology} 26 (1): 32--46.}
 
-\href{http://paperpile.com/b/sTGYvp/qjj9}{Anderson, Philip S. L., Matt
-Friedman, Martin D. Brazeau, and Emily J. Rayfield. 2011. ``Initial
-Radiation of Jaws Demonstrated Stability despite Faunal and
-Environmental Change.'' \emph{Nature} 476 (7359): 206--9.}
+% \href{http://paperpile.com/b/sTGYvp/3hy2}{Anderson, Marti J., and Daniel
+% C. I. Walsh. 2013. ``PERMANOVA, ANOSIM, and the Mantel Test in the Face
+% of Heterogeneous Dispersions: What Null Hypothesis Are You Testing?''
+% \emph{Ecological Monographs} 83 (4): 557--74.}
 
-\href{http://paperpile.com/b/sTGYvp/Qsl3}{Bellman, R. 1966. ``Dynamic
-Programming.'' \emph{Science} 153 (3731): 34--37.}
+% \href{http://paperpile.com/b/sTGYvp/qjj9}{Anderson, Philip S. L., Matt
+% Friedman, Martin D. Brazeau, and Emily J. Rayfield. 2011. ``Initial
+% Radiation of Jaws Demonstrated Stability despite Faunal and
+% Environmental Change.'' \emph{Nature} 476 (7359): 206--9.}
 
-\href{http://paperpile.com/b/sTGYvp/1SD2}{Bookstein, Fred L. 1997.
-\emph{Morphometric Tools for Landmark Data: Geometry and Biology}.
-Cambridge University Press.}
+% \href{http://paperpile.com/b/sTGYvp/Qsl3}{Bellman, R. 1966. ``Dynamic
+% Programming.'' \emph{Science} 153 (3731): 34--37.}
 
-\href{http://paperpile.com/b/sTGYvp/sN5d}{---------. 2015. ``The
-Relation between Geometric Morphometrics and Functional Morphology, as
-Explored by Procrustes Interpretation of Individual Shape Measures
-Pertinent to Function.'' \emph{Anatomical Record} 298 (1): 314--27.}
+% \href{http://paperpile.com/b/sTGYvp/1SD2}{Bookstein, Fred L. 1997.
+% \emph{Morphometric Tools for Landmark Data: Geometry and Biology}.
+% Cambridge University Press.}
 
-\href{http://paperpile.com/b/sTGYvp/o4w7}{---------. 2017a. ``A Newly
-Noticed Formula Enforces Fundamental Limits on Geometric Morphometric
-Analyses.'' \emph{Evolutionary Biology} 44 (4): 522--41.}
+% \href{http://paperpile.com/b/sTGYvp/sN5d}{---------. 2015. ``The
+% Relation between Geometric Morphometrics and Functional Morphology, as
+% Explored by Procrustes Interpretation of Individual Shape Measures
+% Pertinent to Function.'' \emph{Anatomical Record} 298 (1): 314--27.}
 
-\href{http://paperpile.com/b/sTGYvp/xaUx}{---------. 2017b. ``A Method
-of Factor Analysis for Shape Coordinates.'' \emph{American Journal of
-Physical Anthropology} 164 (2): 221--45.}
+% \href{http://paperpile.com/b/sTGYvp/o4w7}{---------. 2017a. ``A Newly
+% Noticed Formula Enforces Fundamental Limits on Geometric Morphometric
+% Analyses.'' \emph{Evolutionary Biology} 44 (4): 522--41.}
 
-\href{http://paperpile.com/b/sTGYvp/9JdS}{Bouxin, Guy. 2005. ``Ginkgo, a
-Multivariate Analysis Package.'' \emph{Journal of Vegetation Science:
-Official Organ of the International Association for Vegetation Science}
-16 (3): 355--59.}
+% \href{http://paperpile.com/b/sTGYvp/xaUx}{---------. 2017b. ``A Method
+% of Factor Analysis for Shape Coordinates.'' \emph{American Journal of
+% Physical Anthropology} 164 (2): 221--45.}
 
-\href{http://paperpile.com/b/sTGYvp/Yrbg}{Brazeau, Martin D., Thomas
-Guillerme, and Martin R. Smith. 2017. ``Morphological Phylogenetic
-Analysis with Inapplicable Data.''
-https://doi.org/}\href{http://dx.doi.org/10.1101/209775}{10.1101/209775}\href{http://paperpile.com/b/sTGYvp/Yrbg}{.}
+% \href{http://paperpile.com/b/sTGYvp/9JdS}{Bouxin, Guy. 2005. ``Ginkgo, a
+% Multivariate Analysis Package.'' \emph{Journal of Vegetation Science:
+% Official Organ of the International Association for Vegetation Science}
+% 16 (3): 355--59.}
 
-\href{http://paperpile.com/b/sTGYvp/tGyd}{Brusatte, S. L., M. J. Benton,
-M. Ruta, and G. T. Lloyd. 2008. ``The First 50 Myr of Dinosaur
-Evolution: Macroevolutionary Pattern and Morphological Disparity.''
-\emph{Biology Letters} 4 (6): 733--36.}
+% \href{http://paperpile.com/b/sTGYvp/Yrbg}{Brazeau, Martin D., Thomas
+% Guillerme, and Martin R. Smith. 2017. ``Morphological Phylogenetic
+% Analysis with Inapplicable Data.''
+% https://doi.org/}\href{http://dx.doi.org/10.1101/209775}{10.1101/209775}\href{http://paperpile.com/b/sTGYvp/Yrbg}{.}
 
-\href{http://paperpile.com/b/sTGYvp/EeC8}{Brusatte, Stephen L., Michael
-J. Benton, Marcello Ruta, and Graeme T. Lloyd. 2008. ``Superiority,
-Competition, and Opportunism in the Evolutionary Radiation of
-Dinosaurs.'' \emph{Science} 321 (5895): 1485--88.}
+% \href{http://paperpile.com/b/sTGYvp/tGyd}{Brusatte, S. L., M. J. Benton,
+% M. Ruta, and G. T. Lloyd. 2008. ``The First 50 Myr of Dinosaur
+% Evolution: Macroevolutionary Pattern and Morphological Disparity.''
+% \emph{Biology Letters} 4 (6): 733--36.}
 
-\href{http://paperpile.com/b/sTGYvp/aSSL}{Butler, Richard J., Stephen L.
-Brusatte, Brian Andres, and Roger B. J. Benson. 2012. ``How Do
-Geological Sampling Biases Affect Studies of Morphological Evolution in
-Deep Time? A Case Study of Pterosaur (Reptilia: Archosauria)
-Disparity.'' \emph{Evolution; International Journal of Organic
-Evolution} 66 (1): 147--62.}
+% \href{http://paperpile.com/b/sTGYvp/EeC8}{Brusatte, Stephen L., Michael
+% J. Benton, Marcello Ruta, and Graeme T. Lloyd. 2008. ``Superiority,
+% Competition, and Opportunism in the Evolutionary Radiation of
+% Dinosaurs.'' \emph{Science} 321 (5895): 1485--88.}
 
-\href{http://paperpile.com/b/sTGYvp/60H0}{Canter, Erin J., Catalina
-Cuellar-Gempeler, Abigail I. Pastore, Thomas E. Miller, and Olivia U.
-Mason. 2018. ``Predator Identity More than Predator Richness Structures
-Aquatic Microbial Assemblages in Sarracenia Purpurea Leaves.''
-\emph{Ecology} 99 (3): 652--60.}
+% \href{http://paperpile.com/b/sTGYvp/aSSL}{Butler, Richard J., Stephen L.
+% Brusatte, Brian Andres, and Roger B. J. Benson. 2012. ``How Do
+% Geological Sampling Biases Affect Studies of Morphological Evolution in
+% Deep Time? A Case Study of Pterosaur (Reptilia: Archosauria)
+% Disparity.'' \emph{Evolution; International Journal of Organic
+% Evolution} 66 (1): 147--62.}
 
-\href{http://paperpile.com/b/sTGYvp/ROH8}{Ciampaglio, Charles N.,
-Matthieu Kemp, and Daniel W. McShea. 2001. ``Detecting Changes in
-Morphospace Occupation Patterns in the Fossil Record: Characterization
-and Analysis of Measures of Disparity.'' \emph{Paleobiology} 27 (4):
-695--715.}
+% \href{http://paperpile.com/b/sTGYvp/60H0}{Canter, Erin J., Catalina
+% Cuellar-Gempeler, Abigail I. Pastore, Thomas E. Miller, and Olivia U.
+% Mason. 2018. ``Predator Identity More than Predator Richness Structures
+% Aquatic Microbial Assemblages in Sarracenia Purpurea Leaves.''
+% \emph{Ecology} 99 (3): 652--60.}
 
-\href{http://paperpile.com/b/sTGYvp/PbSx}{Close, Roger A., Matt
-Friedman, Graeme T. Lloyd, and Roger B. J. Benson. 2015. ``Evidence for
-a Mid-Jurassic Adaptive Radiation in Mammals.'' \emph{Current Biology:
-CB} 25 (16): 2137--42.}
+% \href{http://paperpile.com/b/sTGYvp/ROH8}{Ciampaglio, Charles N.,
+% Matthieu Kemp, and Daniel W. McShea. 2001. ``Detecting Changes in
+% Morphospace Occupation Patterns in the Fossil Record: Characterization
+% and Analysis of Measures of Disparity.'' \emph{Paleobiology} 27 (4):
+% 695--715.}
 
-\href{http://paperpile.com/b/sTGYvp/RjqE}{Cooney, Christopher R., Jen A.
-Bright, Elliot J. R. Capp, Angela M. Chira, Emma C. Hughes, Christopher
-J. A. Moody, Lara O. Nouri, Zoë K. Varley, and Gavin H. Thomas. 2017.
-``Mega-Evolutionary Dynamics of the Adaptive Radiation of Birds.''
-\emph{Nature} 542 (7641): 344--47.}
+% \href{http://paperpile.com/b/sTGYvp/PbSx}{Close, Roger A., Matt
+% Friedman, Graeme T. Lloyd, and Roger B. J. Benson. 2015. ``Evidence for
+% a Mid-Jurassic Adaptive Radiation in Mammals.'' \emph{Current Biology:
+% CB} 25 (16): 2137--42.}
 
-\href{http://paperpile.com/b/sTGYvp/0y4V}{Deline, Bradley, Jennifer M.
-Greenwood, James W. Clark, Mark N. Puttick, Kevin J. Peterson, and
-Philip C. J. Donoghue. 2018. ``Evolution of Metazoan Morphological
-Disparity.'' \emph{Proceedings of the National Academy of Sciences of
-the United States of America} 115 (38): E8909--18.}
+% \href{http://paperpile.com/b/sTGYvp/RjqE}{Cooney, Christopher R., Jen A.
+% Bright, Elliot J. R. Capp, Angela M. Chira, Emma C. Hughes, Christopher
+% J. A. Moody, Lara O. Nouri, Zoë K. Varley, and Gavin H. Thomas. 2017.
+% ``Mega-Evolutionary Dynamics of the Adaptive Radiation of Birds.''
+% \emph{Nature} 542 (7641): 344--47.}
 
-\href{http://paperpile.com/b/sTGYvp/47fI}{Díaz, Sandra, Jens Kattge,
-Johannes H. C. Cornelissen, Ian J. Wright, Sandra Lavorel, Stéphane
-Dray, Björn Reu, et al. 2016. ``The Global Spectrum of Plant Form and
-Function.'' \emph{Nature} 529 (7585): 167--71.}
+% \href{http://paperpile.com/b/sTGYvp/0y4V}{Deline, Bradley, Jennifer M.
+% Greenwood, James W. Clark, Mark N. Puttick, Kevin J. Peterson, and
+% Philip C. J. Donoghue. 2018. ``Evolution of Metazoan Morphological
+% Disparity.'' \emph{Proceedings of the National Academy of Sciences of
+% the United States of America} 115 (38): E8909--18.}
 
-\href{http://paperpile.com/b/sTGYvp/2KmX}{Dixon, Philip. 2003. ``VEGAN,
-a Package of R Functions for Community Ecology.'' \emph{Journal of
-Vegetation Science: Official Organ of the International Association for
-Vegetation Science} 14 (6): 927.}
+% \href{http://paperpile.com/b/sTGYvp/47fI}{Díaz, Sandra, Jens Kattge,
+% Johannes H. C. Cornelissen, Ian J. Wright, Sandra Lavorel, Stéphane
+% Dray, Björn Reu, et al. 2016. ``The Global Spectrum of Plant Form and
+% Function.'' \emph{Nature} 529 (7585): 167--71.}
 
-\href{http://paperpile.com/b/sTGYvp/krNU}{Donohue, Ian, Owen L. Petchey,
-José M. Montoya, Andrew L. Jackson, Luke McNally, Mafalda Viana, Kevin
-Healy, Miguel Lurgi, Nessa E. O'Connor, and Mark C. Emmerson. 2013. ``On
-the Dimensionality of Ecological Stability.'' \emph{Ecology Letters} 16
-(4): 421--29.}
+% \href{http://paperpile.com/b/sTGYvp/2KmX}{Dixon, Philip. 2003. ``VEGAN,
+% a Package of R Functions for Community Ecology.'' \emph{Journal of
+% Vegetation Science: Official Organ of the International Association for
+% Vegetation Science} 14 (6): 927.}
 
-\href{http://paperpile.com/b/sTGYvp/EPJ2}{Erwin, Douglas H. 2007.
-``DISPARITY: MORPHOLOGICAL PATTERN AND DEVELOPMENTAL CONTEXT.''
-\emph{Palaeontology}.
-https://doi.org/}\href{http://dx.doi.org/10.1111/j.1475-4983.2006.00614.x}{10.1111/j.1475-4983.2006.00614.x}\href{http://paperpile.com/b/sTGYvp/EPJ2}{.}
+% \href{http://paperpile.com/b/sTGYvp/krNU}{Donohue, Ian, Owen L. Petchey,
+% José M. Montoya, Andrew L. Jackson, Luke McNally, Mafalda Viana, Kevin
+% Healy, Miguel Lurgi, Nessa E. O'Connor, and Mark C. Emmerson. 2013. ``On
+% the Dimensionality of Ecological Stability.'' \emph{Ecology Letters} 16
+% (4): 421--29.}
 
-\href{http://paperpile.com/b/sTGYvp/Z6l6}{---------. 2011.
-``Evolutionary Uniformitarianism.'' \emph{Developmental Biology} 357
-(1): 27--34.}
+% \href{http://paperpile.com/b/sTGYvp/EPJ2}{Erwin, Douglas H. 2007.
+% ``DISPARITY: MORPHOLOGICAL PATTERN AND DEVELOPMENTAL CONTEXT.''
+% \emph{Palaeontology}.
+% https://doi.org/}\href{http://dx.doi.org/10.1111/j.1475-4983.2006.00614.x}{10.1111/j.1475-4983.2006.00614.x}\href{http://paperpile.com/b/sTGYvp/EPJ2}{.}
 
-\href{http://paperpile.com/b/sTGYvp/yyNa}{Finlay, Sive, and Natalie
-Cooper. 2015. ``Morphological Diversity in Tenrecs (Afrosoricida,
-Tenrecidae): Comparing Tenrec Skull Diversity to Their Closest
-Relatives.'' \emph{PeerJ} 3 (April): e927.}
+% \href{http://paperpile.com/b/sTGYvp/Z6l6}{---------. 2011.
+% ``Evolutionary Uniformitarianism.'' \emph{Developmental Biology} 357
+% (1): 27--34.}
 
-\href{http://paperpile.com/b/sTGYvp/2Neu}{Foote, Mike. 1989.
-``Perimeter-Based Fourier Analysis: A New Morphometric Method Applied to
-the Trilobite Cranidium.'' \emph{Journal of Paleontology}.
-https://doi.org/}\href{http://dx.doi.org/10.1017/s0022336000036556}{10.1017/s0022336000036556}\href{http://paperpile.com/b/sTGYvp/2Neu}{.}
+% \href{http://paperpile.com/b/sTGYvp/yyNa}{Finlay, Sive, and Natalie
+% Cooper. 2015. ``Morphological Diversity in Tenrecs (Afrosoricida,
+% Tenrecidae): Comparing Tenrec Skull Diversity to Their Closest
+% Relatives.'' \emph{PeerJ} 3 (April): e927.}
 
-\href{http://paperpile.com/b/sTGYvp/fTJ3}{---------. 1995.
-``Morphological Diversification of Paleozoic Crinoids.''
-\emph{Paleobiology} 21 (03): 273--99.}
+% \href{http://paperpile.com/b/sTGYvp/2Neu}{Foote, Mike. 1989.
+% ``Perimeter-Based Fourier Analysis: A New Morphometric Method Applied to
+% the Trilobite Cranidium.'' \emph{Journal of Paleontology}.
+% https://doi.org/}\href{http://dx.doi.org/10.1017/s0022336000036556}{10.1017/s0022336000036556}\href{http://paperpile.com/b/sTGYvp/2Neu}{.}
 
-\href{http://paperpile.com/b/sTGYvp/yqPw}{---------. 1997. ``THE
-EVOLUTION OF MORPHOLOGICAL DIVERSITY.'' \emph{Annual Review of Ecology
-and Systematics} 28 (1): 129--52.}
+% \href{http://paperpile.com/b/sTGYvp/fTJ3}{---------. 1995.
+% ``Morphological Diversification of Paleozoic Crinoids.''
+% \emph{Paleobiology} 21 (03): 273--99.}
 
-\href{http://paperpile.com/b/sTGYvp/2tbJ}{Fortey, R. A., D. E. G.
-Briggs, and M. A. Wills. 1996. ``The Cambrian Evolutionary `explosion':
-Decoupling Cladogenesis from Morphological Disparity.'' \emph{Biological
-Journal of the Linnean Society. Linnean Society of London} 57 (1):
-13--33.}
+% \href{http://paperpile.com/b/sTGYvp/yqPw}{---------. 1997. ``THE
+% EVOLUTION OF MORPHOLOGICAL DIVERSITY.'' \emph{Annual Review of Ecology
+% and Systematics} 28 (1): 129--52.}
 
-\href{http://paperpile.com/b/sTGYvp/EETc}{Friedman, Matt. 2010.
-``Explosive Morphological Diversification of Spiny-Finned Teleost Fishes
-in the Aftermath of the End-Cretaceous Extinction.'' \emph{Proceedings.
-Biological Sciences / The Royal Society} 277 (1688): 1675--83.}
+% \href{http://paperpile.com/b/sTGYvp/2tbJ}{Fortey, R. A., D. E. G.
+% Briggs, and M. A. Wills. 1996. ``The Cambrian Evolutionary `explosion':
+% Decoupling Cladogenesis from Morphological Disparity.'' \emph{Biological
+% Journal of the Linnean Society. Linnean Society of London} 57 (1):
+% 13--33.}
 
-\href{http://paperpile.com/b/sTGYvp/SJbC}{Gerber, Sylvain. 2014. ``Not
-All Roads Can Be Taken: Development Induces Anisotropic Accessibility in
-Morphospace.'' \emph{Evolution \& Development} 16 (6): 373--81.}
+% \href{http://paperpile.com/b/sTGYvp/EETc}{Friedman, Matt. 2010.
+% ``Explosive Morphological Diversification of Spiny-Finned Teleost Fishes
+% in the Aftermath of the End-Cretaceous Extinction.'' \emph{Proceedings.
+% Biological Sciences / The Royal Society} 277 (1688): 1675--83.}
 
-\href{http://paperpile.com/b/sTGYvp/QVvv}{---------. 2017. ``The
-Geometry of Morphospaces: Lessons from the Classic Raup Shell Coiling
-Model.'' \emph{Biological Reviews of the Cambridge Philosophical
-Society} 92 (2): 1142--55.}
+% \href{http://paperpile.com/b/sTGYvp/SJbC}{Gerber, Sylvain. 2014. ``Not
+% All Roads Can Be Taken: Development Induces Anisotropic Accessibility in
+% Morphospace.'' \emph{Evolution \& Development} 16 (6): 373--81.}
 
-\href{http://paperpile.com/b/sTGYvp/Uns3}{Gould, Stephen Jay. 2000.
-\emph{Wonderful Life: The Burgess Shale and the Nature of History}.
-Random House.}
+% \href{http://paperpile.com/b/sTGYvp/QVvv}{---------. 2017. ``The
+% Geometry of Morphospaces: Lessons from the Classic Raup Shell Coiling
+% Model.'' \emph{Biological Reviews of the Cambridge Philosophical
+% Society} 92 (2): 1142--55.}
 
-\href{http://paperpile.com/b/sTGYvp/xDqf}{Guillerme, Thomas. 2018.
-``dispRity : A Modular R Package for Measuring Disparity.''
-\emph{Methods in Ecology and Evolution / British Ecological Society} 9
-(7): 1755--63.}
+% \href{http://paperpile.com/b/sTGYvp/Uns3}{Gould, Stephen Jay. 2000.
+% \emph{Wonderful Life: The Burgess Shale and the Nature of History}.
+% Random House.}
 
-\href{http://paperpile.com/b/sTGYvp/ekU4}{Guillerme, Thomas, and Natalie
-Cooper. 2018. ``Time for a Rethink: Time Sub-Sampling Methods in
-Disparity-through-Time Analyses.'' \emph{Palaeontology} 61 (4):
-481--93.}
+% \href{http://paperpile.com/b/sTGYvp/xDqf}{Guillerme, Thomas. 2018.
+% ``dispRity : A Modular R Package for Measuring Disparity.''
+% \emph{Methods in Ecology and Evolution / British Ecological Society} 9
+% (7): 1755--63.}
 
-\href{http://paperpile.com/b/sTGYvp/u1KE}{Guillerme, Thomas, Mark N.
-Puttick, Ariel E. Marcy, and Vera Weisbecker. n.d. ``Shifting Spaces:
-Which Disparity or Dissimilarity Metrics Best Summarise Occupancy in
-Multidimensional Spaces?''
-https://doi.org/}\href{http://dx.doi.org/10.1101/801571}{10.1101/801571}\href{http://paperpile.com/b/sTGYvp/u1KE}{.}
+% \href{http://paperpile.com/b/sTGYvp/ekU4}{Guillerme, Thomas, and Natalie
+% Cooper. 2018. ``Time for a Rethink: Time Sub-Sampling Methods in
+% Disparity-through-Time Analyses.'' \emph{Palaeontology} 61 (4):
+% 481--93.}
 
-\href{http://paperpile.com/b/sTGYvp/9Zoi}{Harmon, Luke J., Jason T.
-Weir, Chad D. Brock, Richard E. Glor, and Wendell Challenger. 2008.
-``GEIGER: Investigating Evolutionary Radiations.'' \emph{Bioinformatics}
-24 (1): 129--31.}
+% \href{http://paperpile.com/b/sTGYvp/u1KE}{Guillerme, Thomas, Mark N.
+% Puttick, Ariel E. Marcy, and Vera Weisbecker. n.d. ``Shifting Spaces:
+% Which Disparity or Dissimilarity Metrics Best Summarise Occupancy in
+% Multidimensional Spaces?''
+% https://doi.org/}\href{http://dx.doi.org/10.1101/801571}{10.1101/801571}\href{http://paperpile.com/b/sTGYvp/u1KE}{.}
 
-\href{http://paperpile.com/b/sTGYvp/WXik}{Harvey, Paul H., and Mark D.
-Pagel. 1998. \emph{The Comparative Method in Evolutionary Biology}.
-Oxford University Press, USA.}
+% \href{http://paperpile.com/b/sTGYvp/9Zoi}{Harmon, Luke J., Jason T.
+% Weir, Chad D. Brock, Richard E. Glor, and Wendell Challenger. 2008.
+% ``GEIGER: Investigating Evolutionary Radiations.'' \emph{Bioinformatics}
+% 24 (1): 129--31.}
 
-\href{http://paperpile.com/b/sTGYvp/3JPy}{Hill, Jennifer J., Mark N.
-Puttick, Thomas L. Stubbs, Emily J. Rayfield, and Philip C. J. Donoghue.
-2018. ``Evolution of Jaw Disparity in Fishes.'' \emph{Palaeontology}.
-https://doi.org/}\href{http://dx.doi.org/10.1111/pala.12371}{10.1111/pala.12371}\href{http://paperpile.com/b/sTGYvp/3JPy}{.}
+% \href{http://paperpile.com/b/sTGYvp/WXik}{Harvey, Paul H., and Mark D.
+% Pagel. 1998. \emph{The Comparative Method in Evolutionary Biology}.
+% Oxford University Press, USA.}
 
-\href{http://paperpile.com/b/sTGYvp/xLdm}{Hopkins, Melanie J. 2017.
-``How Well Does a Part Represent the Whole? A Comparison of Cranidial
-Shape Evolution with Exoskeletal Character Evolution in the Trilobite
-Family Pterocephaliidae.'' \emph{Palaeontology} 60 (3): 309--18.}
+% \href{http://paperpile.com/b/sTGYvp/3JPy}{Hill, Jennifer J., Mark N.
+% Puttick, Thomas L. Stubbs, Emily J. Rayfield, and Philip C. J. Donoghue.
+% 2018. ``Evolution of Jaw Disparity in Fishes.'' \emph{Palaeontology}.
+% https://doi.org/}\href{http://dx.doi.org/10.1111/pala.12371}{10.1111/pala.12371}\href{http://paperpile.com/b/sTGYvp/3JPy}{.}
 
-\href{http://paperpile.com/b/sTGYvp/vTHS}{Hopkins, Melanie J., and
-Sylvain Gerber. 2017. ``Morphological Disparity.'' In \emph{Evolutionary
-Developmental Biology}, 1--12.}
+% \href{http://paperpile.com/b/sTGYvp/xLdm}{Hopkins, Melanie J. 2017.
+% ``How Well Does a Part Represent the Whole? A Comparison of Cranidial
+% Shape Evolution with Exoskeletal Character Evolution in the Trilobite
+% Family Pterocephaliidae.'' \emph{Palaeontology} 60 (3): 309--18.}
 
-\href{http://paperpile.com/b/sTGYvp/hea5}{Hopkins, M. J. 2013.
-``Decoupling of Taxonomic Diversity and Morphological Disparity during
-Decline of the Cambrian Trilobite Family Pterocephaliidae.''
-\emph{Journal of Evolutionary Biology} 26 (8): 1665--76.}
+% \href{http://paperpile.com/b/sTGYvp/vTHS}{Hopkins, Melanie J., and
+% Sylvain Gerber. 2017. ``Morphological Disparity.'' In \emph{Evolutionary
+% Developmental Biology}, 1--12.}
 
-\href{http://paperpile.com/b/sTGYvp/xxh5}{Hughes, Martin, Sylvain
-Gerber, and Matthew Albion Wills. 2013. ``Clades Reach Highest
-Morphological Disparity Early in Their Evolution.'' \emph{Proceedings of
-the National Academy of Sciences of the United States of America} 110
-(34): 13875--79.}
+% \href{http://paperpile.com/b/sTGYvp/hea5}{Hopkins, M. J. 2013.
+% ``Decoupling of Taxonomic Diversity and Morphological Disparity during
+% Decline of the Cambrian Trilobite Family Pterocephaliidae.''
+% \emph{Journal of Evolutionary Biology} 26 (8): 1665--76.}
 
-\href{http://paperpile.com/b/sTGYvp/PwyQ}{Jackson, Andrew L., Richard
-Inger, Andrew C. Parnell, and Stuart Bearhop. 2011. ``Comparing Isotopic
-Niche Widths among and within Communities: SIBER - Stable Isotope
-Bayesian Ellipses in R.'' \emph{The Journal of Animal Ecology} 80 (3):
-595--602.}
+% \href{http://paperpile.com/b/sTGYvp/xxh5}{Hughes, Martin, Sylvain
+% Gerber, and Matthew Albion Wills. 2013. ``Clades Reach Highest
+% Morphological Disparity Early in Their Evolution.'' \emph{Proceedings of
+% the National Academy of Sciences of the United States of America} 110
+% (34): 13875--79.}
 
-\href{http://paperpile.com/b/sTGYvp/oFiP}{Legendre, P., and Loic F. J.
-Legendre. 2012. \emph{Numerical Ecology}. Elsevier.}
+% \href{http://paperpile.com/b/sTGYvp/PwyQ}{Jackson, Andrew L., Richard
+% Inger, Andrew C. Parnell, and Stuart Bearhop. 2011. ``Comparing Isotopic
+% Niche Widths among and within Communities: SIBER - Stable Isotope
+% Bayesian Ellipses in R.'' \emph{The Journal of Animal Ecology} 80 (3):
+% 595--602.}
 
-\href{http://paperpile.com/b/sTGYvp/bCsU}{Lloyd, Graeme T. 2016.
-``Estimating Morphological Diversity and Tempo with Discrete
-Character-Taxon Matrices: Implementation, Challenges, Progress, and
-Future Directions.'' \emph{Biological Journal of the Linnean Society.
-Linnean Society of London} 118 (1): 131--51.}
+% \href{http://paperpile.com/b/sTGYvp/oFiP}{Legendre, P., and Loic F. J.
+% Legendre. 2012. \emph{Numerical Ecology}. Elsevier.}
 
-\href{http://paperpile.com/b/sTGYvp/53SJ}{---------. 2018. ``Journeys
-through Discrete-Character Morphospace: Synthesizing Phylogeny, Tempo,
-and Disparity.'' \emph{Palaeontology} 61 (5): 637--45.}
+% \href{http://paperpile.com/b/sTGYvp/bCsU}{Lloyd, Graeme T. 2016.
+% ``Estimating Morphological Diversity and Tempo with Discrete
+% Character-Taxon Matrices: Implementation, Challenges, Progress, and
+% Future Directions.'' \emph{Biological Journal of the Linnean Society.
+% Linnean Society of London} 118 (1): 131--51.}
 
-\href{http://paperpile.com/b/sTGYvp/dJHu}{Losos, Jonathan B. 2011.
-\emph{Lizards in an Evolutionary Tree: Ecology and Adaptive Radiation of
-Anoles}. Univ of California Press.}
+% \href{http://paperpile.com/b/sTGYvp/53SJ}{---------. 2018. ``Journeys
+% through Discrete-Character Morphospace: Synthesizing Phylogeny, Tempo,
+% and Disparity.'' \emph{Palaeontology} 61 (5): 637--45.}
 
-\href{http://paperpile.com/b/sTGYvp/aVVj}{Moyne, Sébastien, and Pascal
-Neige. 2007. ``The Space-Time Relationship of Taxonomic Diversity and
-Morphological Disparity in the Middle Jurassic Ammonite Radiation.''
-\emph{Palaeogeography, Palaeoclimatology, Palaeoecology} 248 (1-2):
-82--95.}
+% \href{http://paperpile.com/b/sTGYvp/dJHu}{Losos, Jonathan B. 2011.
+% \emph{Lizards in an Evolutionary Tree: Ecology and Adaptive Radiation of
+% Anoles}. Univ of California Press.}
 
-\href{http://paperpile.com/b/sTGYvp/EmTR}{Navarro, Nicolas. 2003. ``MDA:
-A MATLAB-Based Program for Morphospace-Disparity Analysis.''
-\emph{Computers \& Geosciences} 29 (5): 655--64.}
+% \href{http://paperpile.com/b/sTGYvp/aVVj}{Moyne, Sébastien, and Pascal
+% Neige. 2007. ``The Space-Time Relationship of Taxonomic Diversity and
+% Morphological Disparity in the Middle Jurassic Ammonite Radiation.''
+% \emph{Palaeogeography, Palaeoclimatology, Palaeoecology} 248 (1-2):
+% 82--95.}
 
-\href{http://paperpile.com/b/sTGYvp/yO2t}{Palci, Alessandro, and Michael
-S. Y. Lee. 2018. ``Geometric Morphometrics, Homology and Cladistics:
-Review and Recommendations.'' \emph{Cladistics: The International
-Journal of the Willi Hennig Society}.
-https://doi.org/}\href{http://dx.doi.org/10.1111/cla.12340}{10.1111/cla.12340}\href{http://paperpile.com/b/sTGYvp/yO2t}{.}
+% \href{http://paperpile.com/b/sTGYvp/EmTR}{Navarro, Nicolas. 2003. ``MDA:
+% A MATLAB-Based Program for Morphospace-Disparity Analysis.''
+% \emph{Computers \& Geosciences} 29 (5): 655--64.}
 
-\href{http://paperpile.com/b/sTGYvp/tSIy}{Pierce, Stephanie E., Kenneth
-D. Angielczyk, and Emily J. Rayfield. 2008. ``Patterns of Morphospace
-Occupation and Mechanical Performance in Extant Crocodilian Skulls: A
-Combined Geometric Morphometric and Finite Element Modeling Approach.''
-\emph{Journal of Morphology} 269 (7): 840--64.}
+% \href{http://paperpile.com/b/sTGYvp/yO2t}{Palci, Alessandro, and Michael
+% S. Y. Lee. 2018. ``Geometric Morphometrics, Homology and Cladistics:
+% Review and Recommendations.'' \emph{Cladistics: The International
+% Journal of the Willi Hennig Society}.
+% https://doi.org/}\href{http://dx.doi.org/10.1111/cla.12340}{10.1111/cla.12340}\href{http://paperpile.com/b/sTGYvp/yO2t}{.}
 
-\href{http://paperpile.com/b/sTGYvp/I0Ic}{Raup, D. M. 1961. ``THE
-GEOMETRY OF COILING IN GASTROPODS.'' \emph{Proceedings of the National
-Academy of Sciences of the United States of America} 47 (4): 602--9.}
+% \href{http://paperpile.com/b/sTGYvp/tSIy}{Pierce, Stephanie E., Kenneth
+% D. Angielczyk, and Emily J. Rayfield. 2008. ``Patterns of Morphospace
+% Occupation and Mechanical Performance in Extant Crocodilian Skulls: A
+% Combined Geometric Morphometric and Finite Element Modeling Approach.''
+% \emph{Journal of Morphology} 269 (7): 840--64.}
 
-\href{http://paperpile.com/b/sTGYvp/geAO}{Ruta, Marcello, Kenneth D.
-Angielczyk, Jörg Fröbisch, and Michael J. Benton. 2013. ``Decoupling of
-Morphological Disparity and Taxic Diversity during the Adaptive
-Radiation of Anomodont Therapsids.'' \emph{Proceedings. Biological
-Sciences / The Royal Society} 280 (1768): 20131071.}
+% \href{http://paperpile.com/b/sTGYvp/I0Ic}{Raup, D. M. 1961. ``THE
+% GEOMETRY OF COILING IN GASTROPODS.'' \emph{Proceedings of the National
+% Academy of Sciences of the United States of America} 47 (4): 602--9.}
 
-\href{http://paperpile.com/b/sTGYvp/cV3v}{Saupe, Erin E., Huijie Qiao,
-Jonathan R. Hendricks, Roger W. Portell, Stephen J. Hunter, Jorge
-Soberón, and Bruce S. Lieberman. 2015. ``Niche Breadth and Geographic
-Range Size as Determinants of Species Survival on Geological Time
-Scales.'' \emph{Global Ecology and Biogeography}.
-https://doi.org/}\href{http://dx.doi.org/10.1111/geb.12333}{10.1111/geb.12333}\href{http://paperpile.com/b/sTGYvp/cV3v}{.}
+% \href{http://paperpile.com/b/sTGYvp/geAO}{Ruta, Marcello, Kenneth D.
+% Angielczyk, Jörg Fröbisch, and Michael J. Benton. 2013. ``Decoupling of
+% Morphological Disparity and Taxic Diversity during the Adaptive
+% Radiation of Anomodont Therapsids.'' \emph{Proceedings. Biological
+% Sciences / The Royal Society} 280 (1768): 20131071.}
 
-\href{http://paperpile.com/b/sTGYvp/856K}{Sebé-Pedrós, Arnau, Elad
-Chomsky, Kevin Pang, David Lara-Astiaso, Federico Gaiti, Zohar Mukamel,
-Ido Amit, Andreas Hejnol, Bernard M. Degnan, and Amos Tanay. 2018.
-``Early Metazoan Cell Type Diversity and the Evolution of Multicellular
-Gene Regulation.'' \emph{Nature Ecology \& Evolution} 2 (7): 1176--88.}
+% \href{http://paperpile.com/b/sTGYvp/cV3v}{Saupe, Erin E., Huijie Qiao,
+% Jonathan R. Hendricks, Roger W. Portell, Stephen J. Hunter, Jorge
+% Soberón, and Bruce S. Lieberman. 2015. ``Niche Breadth and Geographic
+% Range Size as Determinants of Species Survival on Geological Time
+% Scales.'' \emph{Global Ecology and Biogeography}.
+% https://doi.org/}\href{http://dx.doi.org/10.1111/geb.12333}{10.1111/geb.12333}\href{http://paperpile.com/b/sTGYvp/cV3v}{.}
 
-\href{http://paperpile.com/b/sTGYvp/ZEDR}{Spriggs, Elizabeth L., Samuel
-B. Schmerler, Erika J. Edwards, and Michael J. Donoghue. 2018. ``Leaf
-Form Evolution in Viburnum Parallels Variation within Individual
-Plants.'' \emph{The American Naturalist} 191 (2): 235--49.}
+% \href{http://paperpile.com/b/sTGYvp/856K}{Sebé-Pedrós, Arnau, Elad
+% Chomsky, Kevin Pang, David Lara-Astiaso, Federico Gaiti, Zohar Mukamel,
+% Ido Amit, Andreas Hejnol, Bernard M. Degnan, and Amos Tanay. 2018.
+% ``Early Metazoan Cell Type Diversity and the Evolution of Multicellular
+% Gene Regulation.'' \emph{Nature Ecology \& Evolution} 2 (7): 1176--88.}
 
-\href{http://paperpile.com/b/sTGYvp/Ejzr}{Wainwright, Peter C., Michael
-E. Alfaro, Daniel I. Bolnick, and C. Darrin Hulsey. 2005. ``Many-to-One
-Mapping of Form to Function: A General Principle in Organismal Design?''
-\emph{Integrative and Comparative Biology} 45 (2): 256--62.}
+% \href{http://paperpile.com/b/sTGYvp/ZEDR}{Spriggs, Elizabeth L., Samuel
+% B. Schmerler, Erika J. Edwards, and Michael J. Donoghue. 2018. ``Leaf
+% Form Evolution in Viburnum Parallels Variation within Individual
+% Plants.'' \emph{The American Naturalist} 191 (2): 235--49.}
 
-\href{http://paperpile.com/b/sTGYvp/TZzO}{Weisbecker, Vera, Thomas
-Guillerme, Cruise Speck, Emma Sherratt, Hyab Mehari Abraha, Alana C.
-Sharp, Claire E. Terhune, Simon Collins, Stephen Johnston, and Olga
-Panagiotopoulou. 2019. ``Individual Variation of the Masticatory System
-Dominates 3D Skull Shape in the Herbivory-Adapted Marsupial Wombats.''
-\emph{Frontiers in Zoology} 16 (November): 41.}
+% \href{http://paperpile.com/b/sTGYvp/Ejzr}{Wainwright, Peter C., Michael
+% E. Alfaro, Daniel I. Bolnick, and C. Darrin Hulsey. 2005. ``Many-to-One
+% Mapping of Form to Function: A General Principle in Organismal Design?''
+% \emph{Integrative and Comparative Biology} 45 (2): 256--62.}
 
-\href{http://paperpile.com/b/sTGYvp/nFf7}{Wills, Matthew A. 2001.
-``Morphological Disparity: A Primer.'' In \emph{Topics in Geobiology},
-55--144.}
+% \href{http://paperpile.com/b/sTGYvp/TZzO}{Weisbecker, Vera, Thomas
+% Guillerme, Cruise Speck, Emma Sherratt, Hyab Mehari Abraha, Alana C.
+% Sharp, Claire E. Terhune, Simon Collins, Stephen Johnston, and Olga
+% Panagiotopoulou. 2019. ``Individual Variation of the Masticatory System
+% Dominates 3D Skull Shape in the Herbivory-Adapted Marsupial Wombats.''
+% \emph{Frontiers in Zoology} 16 (November): 41.}
 
-\href{http://paperpile.com/b/sTGYvp/eZ3F}{Wills, Matthew A., Derek E. G.
-Briggs, and Richard A. Fortey. 1994. ``Disparity as an Evolutionary
-Index: A Comparison of Cambrian and Recent Arthropods.''
-\emph{Paleobiology} 20 (02): 93--130.}
+% \href{http://paperpile.com/b/sTGYvp/nFf7}{Wills, Matthew A. 2001.
+% ``Morphological Disparity: A Primer.'' In \emph{Topics in Geobiology},
+% 55--144.}
 
-\href{http://paperpile.com/b/sTGYvp/s33b}{Wright, David F. 2017.
-``Phenotypic Innovation and Adaptive Constraints in the Evolutionary
-Radiation of Palaeozoic Crinoids.'' \emph{Scientific Reports} 7 (1):
-13745.}
+% \href{http://paperpile.com/b/sTGYvp/eZ3F}{Wills, Matthew A., Derek E. G.
+% Briggs, and Richard A. Fortey. 1994. ``Disparity as an Evolutionary
+% Index: A Comparison of Cambrian and Recent Arthropods.''
+% \emph{Paleobiology} 20 (02): 93--130.}
+
+% \href{http://paperpile.com/b/sTGYvp/s33b}{Wright, David F. 2017.
+% ``Phenotypic Innovation and Adaptive Constraints in the Evolutionary
+% Radiation of Palaeozoic Crinoids.'' \emph{Scientific Reports} 7 (1):
+% 13745.}

### 2020-02-20 Writing/disparity_review.tex
updated references

@@
 
-\hypertarget{introduction}{%
-\section{1. Introduction}\label{introduction}}
+\section{Introduction}
 
@@
 Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are a comparatively more recent phenomenon.
-Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (Runnegar 1987, Gould \href{https://paperpile.com/c/sTGYvp/Uns3}{@@@}),
-% Runnegar, Bruce "Rates and modes of evolution in the Mollusca." Rates of evolution. Allen and Unwin, London (1987): 39-60.
- this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
+Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
 However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
@@
 
-(1) \emph{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Foote 1995; Wills, Briggs, and Fortey 1994)}.
+\begin{enumerate}
+
+	\item \textbf{Descriptive disparity} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995-do, Briggs1992-pd}.
 	These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace?
-\href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Foote 1995; Raup 1961; Gerber 2017)}.
+\citep{Foote1995-do, Raup1961-vx, Gerber2017-xi}.
 
-(2) \emph{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
-This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \href{https://paperpile.com/c/sTGYvp/ekU4+s33b}{(Guillerme and Cooper 2018; Wright 2017)}, or how does disparity change leading up to and across mass extinction events \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}?
+	\item \textbf{Disparity-through-time} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
+	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{Guillerme2018-uj, Wright2017-jo}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010-ve}?
 
-(3) \emph{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e.
+	\item \textbf{Disparity and taxonomic diversity} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e.
 shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
 	This makes disparity an informative complement to diversity measures based on species richness alone.
-Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling \href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(e.g., Fortey, Briggs, and Wills 1996; Ruta et al.
-2013; M.
-J.
-Hopkins 2013; Moyne and Neige 2007)}.
-This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \href{https://paperpile.com/c/sTGYvp/dJHu}{(Losos 2011)}.
+Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling
+	\citep[][e.g.]{Fortey1996-kt, Moyne2007-jm, Ruta2013-iy, Hopkins2013-xt}.
+	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011-fq}.
 
-(4) \emph{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
-This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles \href{https://paperpile.com/c/sTGYvp/EETc+tSIy+qjj9}{(Friedman 2010; Pierce, Angielczyk, and Rayfield 2008; P.
-S.
-L.
-Anderson et al.
-2011)}.
+	\item \textbf{Disparity as a proxy for ecology} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
+	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles
+	\citep{Friedman2010-ve, Pierce2008-yr, Anderson2013-zt}.
 	The links between form and function, however, are not always clear cut.
-Traits can be linked to multiple functions and multiple functions can be linked to a single trait \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al. 2005)}.
-This approach has been used to investigate hypotheses of competitive replacement (Stephen L.
-Brusatte et al.
-2008) and changes in ecosystem function during and after mass extinctions \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}.
-It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.} 2005\href{https://paperpile.com/c/sTGYvp/Ejzr}{)}.
+	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005-or}.
+	This approach has been used to investigate hypotheses of competitive replacement \citep{Brusatte2008-vx} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010-ve}.
+	It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \citep{Wainwright2005-or}.
 
-(5) Disparity and evo-devo. Morphological disparity blalbalbal
+	\item \textbf{Disparity and evo-devo} Morphological disparity blalbalbal
 % AG: Although intrinsic constraints are mentioned earlier, it doesn’t enter into any of these topics explicitly, but it is certainly one of the most interesting things one can use disparity for – to bridge development and evolution, looking at developmental biases/constraints.
@@
 
+\end{enumerate}
+
 Fundamental insights into evolutionary biology have been elicited from these five types of analysis.
-One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \href{https://paperpile.com/c/sTGYvp/xxh5+yqPw}{(Hughes, Gerber, and Wills 2013; Foote 1997;} \href{https://paperpile.com/c/sTGYvp/EPJ2}{Erwin 2007)}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change \href{https://paperpile.com/c/sTGYvp/xxh5+Z6l6}{(Hughes, Gerber, and Wills 2013; Erwin 2011)}.
+One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997-nl, Erwin2007-jz, Hughes2013-td}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007-jz, Hughes2013-td}.
 However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?
@@
 
-\emph{Figure disparities {[}figure\_disparities.pdf{]}: The four main types of disparity analysis.
+\textit{Figure disparities {[}figure\_disparities.pdf{]}: The four main types of disparity analysis.
 \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
@@
 
-\hypertarget{data-and-disparity}{%
-\section{2. Data and disparity }\label{data-and-disparity}}
+\section{Data and disparity}
 
@@
 % SG: I agree. There's for instance Saunders et al. (2004) disparity study of Paleozoic ammonoids using Raup's shell coiling parameters
-(1) discrete ``cladistic'' characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)};
-(2) continuous measurements of features (e.g., lengths) \href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al. 2011)}; or
-(3) more mathematical descriptors from geometric morphometric landmark data (e.g.
-Procrustes coordinates) \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al.
-2017)}, and Fourier coefficients \href{https://paperpile.com/c/sTGYvp/2Neu}{(Foote 1989;} \href{https://paperpile.com/c/sTGYvp/ZEDR}{Spriggs et al.
-2018)}) (Fig.
+	1) discrete ``cladistic'' characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g]{Close2015-qi};
+	2) continuous measurements of features (e.g., lengths) \citep[][e.g]{Anderson2001-qb}; or
+	3) more mathematical descriptors from geometric morphometric landmark data (e.g.
+Procrustes coordinates) \citep[][e.g]{Cooney2017-ly}, and Fourier coefficients \citep[][e.g]{Foote1995-do, Spriggs2018-nu} (Fig.
 2).
-None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits (Hetherington et al 2015; Hopkins 2017).
+None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.
 
@@
 The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
-Discrete character data are commonly recycled from phylogenetic studies (e.g.Foote 1989\href{https://paperpile.com/c/sTGYvp/2Neu+0y4V}{; Deline et al. 2018}).
+Discrete character data are commonly recycled from phylogenetic studies \citep[][e.g]{Foote1989-fd, Deline2018-le}.
 % SG: Foote (1989) is a study of trilobite disparity from Fourier analysis of cranidia. I don't think it has anything to do with discrete data or phylogenetic studies

### 2020-02-20 Writing/disparity_review.tex
Update disparity_review.tex

@@
 I can look for some if needed.
-This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups \href{https://paperpile.com/c/sTGYvp/fTJ3}{(Foote 1995)}.
+This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups \citep{Foote1995-do}.
 This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
-Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J.
-Hopkins 2017)}.
-This effect of anatomical part on disparity patterns can be especially challenging when working on datasets where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline et al.
-2018)}.
+Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017-cf}.
+This effect of anatomical part on disparity patterns can be especially challenging when working on datasets where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018-le}.
 
-Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \href{https://paperpile.com/c/sTGYvp/Yrbg}{(Brazeau, Guillerme, and Smith 2017}; \href{https://paperpile.com/c/sTGYvp/yO2t}{Palci and Lee 2018)}.
+Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \citep{Brazeau2017-kg, Palci2018-ni}.
 Biological phenomena such as allometry and sexual dimorphism may also influence trait data.
@@
 
-\emph{Figure data {[}figure\_data.pdf{]}: Major routes to obtain morphological data for disparity analyses.
+\textit{Figure data {[}figure\_data.pdf{]}: Major routes to obtain morphological data for disparity analyses.
 Data can be collected as discrete trait observations (e.g.
@@
 
-\hypertarget{disparity-methods}{%
-\section{Disparity methods}\label{disparity-methods}}
+\section{Disparity methods}
 

### 2020-02-20 Writing/disparity_review.tex
Updated refs

@@
 This has clear advantages for interpreting the results.
-For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak in birds \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017}).
+For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak in birds \citep{Cooney2017-ly}.
 
@@
 %     -It is hard to interpret
-Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(}Bookstein \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1997, 2015, {[}b{]} 2017, {[}a{]} 2017)}{[}CITE Vera and Thomas paper{]}.
+Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \citep{Bookstein1997-dn, Bookstein2015-yy, Bookstein2017-qk, Bookstein2017-gu,Weisbecker2019-kp}.
  %TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
-Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \href{https://paperpile.com/c/sTGYvp/0y4V+QVvv}{(Deline et al.
-2018; Gerber 2017)}.
-Furthermore, interpreting biological variation along the axes is always a \emph{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
+Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018-le, Gerber2017-xi}.
+Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
 Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
 For example, when the analysed data is non-Euclidean (e.g.
-some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \href{https://paperpile.com/c/sTGYvp/SJbC}{(}Gerber 2014\href{https://paperpile.com/c/sTGYvp/SJbC}{)}.
-%TG: should be Gerber 2019
+some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \citep{Gerber2017-xi}.
+%@@@TG: should be Gerber 2019
 This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear.
-Finally, \emph{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible. %TG: Cite Gerber 2019 10.1111/pala.12407
+Finally, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible. %TG: Cite Gerber 2019 10.1111/pala.12407
 Although some gradients are easy to detect or interpret (e.g.
-the elongation and depth of mandibles in fishes on first and second PC axes, respectively; {[}\href{https://paperpile.com/c/sTGYvp/3JPy}{Hill et al.
-2018{]}}), some are not (e.g.
-\href{https://paperpile.com/c/sTGYvp/TZzO}{Weisbecker et al.
-2019)}.
+the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018-ye}), some are not \citep[][e.g.]{Weisbecker2019-kp}.
 For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0, has no biological meaning if these are binary alternate states.
@@
 contrary to a PCA, the first few axis do not encompass most variance of the dataset).
-Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric (\href{https://paperpile.com/c/sTGYvp/SJbC}{i.e.
-the distance between A and B is not equal to the distance between B and A; Gerber 2014)}.
+Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[][i.e.
+the distance between A and B is not equal to the distance between B and A]{Gerber2014-ol}.
 %     -It is often not necessary
 And last but not least reason, In many cases, ordination might not be necessary.
-For example, if a metric (e.g.
-sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below {[}metrics{]}), it is not necessary to calculate it on ordinated data {[}e.g.\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
-2015)}{]}.
+For example, if a metric (e.g. sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below metrics%indices
+), it is not necessary to calculate it on ordinated data \citep{Close2015-qi}.
 % -Disparity can be measured directly from the non-ordinated dataset (e.g. via distance matrices)
-For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination (Lloyd 2016, 2018).
+For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.
 
@@
 
-\emph{Figure methods: Morphospaces: different mathematical representations of a} \emph{morphospace.
+\textit{Figure methods: Morphospaces: different mathematical representations of a} \emph{morphospace.
 % SG: The morphospace is the original trait space within which lie the specimens.
@@
 
-\hypertarget{b-disparity-indices}{%
 % @@@ Change to indices throughout
-\subsection{(b) Measuring disparity}\label{c-disparity-indices}}
+\subsection{(b) Measuring disparity}
 
 Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
-This is usually achieved with a disparity metric or index \href{https://paperpile.com/c/sTGYvp/vTHS}{(Melanie J. Hopkins and Gerber 2017)}.
-As with any summary of multidimensional data, disparity metrics will reflect only some aspects of the morphological variation, never its whole complexity {[}metrics paper{]}.
+This is usually achieved with a disparity metric or index \citep{Hopkins2017-cf}.
+As with any summary of multidimensional data, disparity metrics will reflect only some aspects of the morphological variation, never its whole complexity %metrics paper. CITE moms
 It is therefore often beneficial to use more than one metric to summarise different aspects of variation, guided by the aim of the study.
@@
 does group A occupy as much space as group B?), metrics related to the spread of the group in the morphospace are most appropriate (e.g.
-volume \href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al.
-2016)}, distance from the centroid \href{https://paperpile.com/c/sTGYvp/vTHS+yyNa}{(Melanie J.
-Hopkins and Gerber 2017; Finlay and Cooper 2015)}, variance and range \href{https://paperpile.com/c/sTGYvp/tGyd}{(S.
-L.
-Brusatte et al.
-2008)}).
+volume
+\citealt{Diaz2016-mr}, distance from the centroid \citealt{Hopkins2017-cf, Finlay2015-ft}, variance and range \citealt{Brusatte2008-fx}).
 Conversely, if we wish to describe the ``position''
@@
  of a group in a morphospace (e.g.
-does group A occupy the same space as group B?), metrics related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate (e.g.\href{https://paperpile.com/c/sTGYvp/u1KE}{(Guillerme et al., n.d.)}).
+does group A occupy the same space as group B?), metrics related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate.
+%@@@ CITE moms
 Finally, if we aim to characterise the density of morphospace occupation (e.g.
@@
 nearest neighbour distance, pairwise distances, etc.
-\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
-2015)}).
+\citealt{Close2015-qi}).
 %ES: this para is probably most useful to refer to for the ordinate or not part
 
-In addition to considering what properties of disparity metrics should capture, it is important to also consider the mathematical properties of the metrics and their associated caveats \href{https://paperpile.com/c/sTGYvp/nFf7+ROH8}{(Wills 2001; Ciampaglio, Kemp, and McShea 2001)}.
+In addition to considering what properties of disparity metrics should capture, it is important to also consider the mathematical properties of the metrics and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
 For example, measuring the full sum of the variance in each dimension of the space does not require we add the covariance between the axes in an ordinated space using a PCA.
 % SG: Not very clear to me
-However, this is not true of other mathematical spaces or when not all dimensions or elements are considered, even in a PCA {[}Legendre \& Legendre 2012{]}.
+However, this is not true of other mathematical spaces or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012-va}.
 
-Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}.
+Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \citep{Bellman1966-mc}.
 In spaces with some axis of variance lower than 1, product-based metrics used as proxies of volumes (e.g.
-product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}; Donoho 2000).
-Some other types of metrics are also extremely sensitive to outliers and can be biased easily by sample size, for example range \href{https://paperpile.com/c/sTGYvp/aSSL}{(}Butler et al.
-2012\href{https://paperpile.com/c/sTGYvp/aSSL}{)}
+product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966-mc}.
+%Donoho 2000
+Some other types of metrics are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Butler2012-tr}
 %SG: Or even better, twenty years earlier, Foote (1992), who orginally described the problem in the context of disparity and suggested a rarefaction approach.
-or convex hull based metrics \href{https://paperpile.com/c/sTGYvp/aSSL+PwyQ}{(Butler et al. 2012; Jackson et al. 2011)}.
+or convex hull based metrics \citep{Butler2012-tr, Jackson2011-kq}.
 
-\hypertarget{c-testing-hypotheses-on-the-evolution-of-disparity}{%
 \subsection{(c) Testing hypotheses on the evolution of
-disparity}\label{c-testing-hypotheses-on-the-evolution-of-disparity}}
+disparity}
 
 No matter which disparity metrics have been calculated, the research question must be framed in an appropriate statistical context.
-The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer} 2018\href{https://paperpile.com/c/sTGYvp/ZnDd}{)}
+The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \citep{Adams2018-mg}
 %AG: there are others that should be mentioned - mvmorph for example, which is especially important given that there are extremely different approaches to multivariate data in these different packages
 , but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
-One commonly used test is the non-parametric permutation analysis of variance \href{https://paperpile.com/c/sTGYvp/3hy2+SC6L}{(M.
-J.
-Anderson and Walsh 2013; M.
-J.
-Anderson 2001)}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
+One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001-qb, Anderson2013-zt}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
 Although statistically valid, this test is not always directly related to the hypothesis under question.
@@
 %TG: Add section d) within c)
-\hypertarget{d-phylogenetic-autocorrelation}{%
-\subsection{(d) Phylogenetic
-autocorrelation}\label{d-phylogenetic-autocorrelation}}
+\subsection{(d) Phylogenetic autocorrelation}
 
-As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \href{https://paperpile.com/c/sTGYvp/WXik}{(Harvey and Pagel 1998)}.
+As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998-xg}.
 Thus, for many disparity analyses, phylogenetic relationships should be
@@
 % SG: In what way?
-if incorrect assumptions are made about the data (Uyeda et al). \href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)}
+if incorrect assumptions are made about the data \citep{Uyeda2015}.
 % @@@ref: https://academic.oup.com/sysbio/article/64/4/677/1649888
@@
 %TG: Add section e) within c)???
-\hypertarget{e-ancestral-state-estimation-in-disparity-through-time-analyses}{%
-\subsection{(e) Ancestral state estimation in disparity-through-time
-analyses}\label{e-ancestral-state-estimation-in-disparity-through-time-analyses}}
+\subsection{(e) Ancestral state estimation in disparity-through-time analyses}
 
@@
 the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g.
-the ordination scores) \href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
+the ordination scores) \citep{lloyd2018}.
 
@@
 However, once the morphospace is defined its properties will not change.
-Post-ordination ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.
+Post-ordination ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \citep{lloyd2018}.
 
@@
 
-\hypertarget{disparity-analyses-for-the-future}{%
-\section{4. Disparity analyses for the
-future}\label{disparity-analyses-for-the-future}}
+\section{4. Disparity analyses for the future}
 
@@
 assembly of analytic approaches and, given the computational efficiency of these methods, an experimental approach that explores the impact of competing approaches, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
-Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \href{https://paperpile.com/c/sTGYvp/krNU+60H0+cV3v}{(Donohue et al.
-2013; Canter et al.
-2018; Saupe et al.
-2015)}.
-% @@@ Add Mammola
+Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets
+\citep{Donohue2013-bg, Saupe2015-vm, Canter2018-hk, mammola2019}
+. % @@@ Add Mammola
 Innovations in morphological disparity analyses likely await discovery in their respective literatures.
@@
 While studies of morphological disparity would benefit from advances in multidimensional analysis in other fields, the concept of a morphospace could reciprocally benefit other disciplines.
-For example, the multidimensional analysis of Diaz \href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al.
-2016)}, which analysed patterns of form and function in plants, is essentially an ecomorphospace; isotopic analyses of organisms \href{https://paperpile.com/c/sTGYvp/PwyQ}{(Jackson et al.
-2011)} can be represented as an isotope-space; ecosystem functioning in Donohue et al.
-(2013) as an ecosystem-space, etc.
+For example, the multidimensional analysis of Diaz \citealt{Diaz2016-mr}, which analysed patterns of form and function in plants, is essentially an ecomorphospace; isotopic analyses of organisms \citep{Jackson2011-kq} can be represented as an isotope-space; ecosystem functioning in \citealt{Donohue2013-bg} as an ecosystem-space, etc.
 % AJ: this is a potential nice, although rather data hungry I believe, methods that I think could be worth dropping into the paper somewhere.

### 2020-02-20 Writing/disparity_review.tex
updated all references

@@
 %SG: devising 
-Cognate approaches have been adopted recently in the analysis of single cell comparative transcriptome data \href{https://paperpile.com/c/sTGYvp/856K}{(Sebé-Pedrós et al.
-2018)} where interpretation of the resulting transcriptome-spaces would be improved by heeding the concerns we highlight concerning morphospaces.
+Cognate approaches have been adopted recently in the analysis of single cell comparative transcriptome data \citep{Sebe-Pedros2018-sw} where interpretation of the resulting transcriptome-spaces would be improved by heeding the concerns we highlight concerning morphospaces.
 % ES: Suggested by Erin: reword this sentence
 
-Although disparity analyses are now simple to implement in freely available softwares \href{https://paperpile.com/c/sTGYvp/xDqf+J2G1+9JdS+9Zoi+bCsU+EmTR+2KmX}{(e.g., Guillerme 2018;} {; Bouxin 2005; Harmon et al.
-2008; Lloyd 2016; Navarro 2003; Oksanen 2007)},
-% @@@ add BAT
+Although disparity analyses are now simple to implement in freely available softwares \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc}.
 it is crucial to remember that they are multidimensional analyses and multidimensional analyses are complex.
-We assert that future morphological analyses will benefit by emphasising the methodological decisions made, rather than simply using disparity analysis because \emph{we can}.
+We assert that future morphological analyses will benefit by emphasising the methodological decisions made, rather than simply using disparity analysis because \textit{we can}.
 
-\hypertarget{author-contributions}{%
-\section{5. Author contributions}\label{author-contributions}}
+\subsection{Author contributions}
 
@@
 
-\hypertarget{acknowledgments}{%
-\section{6. Acknowledgments}\label{acknowledgments}}
+\subsection{Acknowledgments}
 

### 2020-02-21 Writing/disparity_review.tex
added figures to tex + compile OK

@@
 %Packages
-\usepackage{fixltx2e}
 \usepackage{textcomp}
@@
 
-Analyses of morphological disparity have been used to characterise and investigate the evolution of variation in the anatomy, function, and ecology of organisms since the 1980s.
+\noindent Analyses of morphological disparity have been used to characterise and investigate the evolution of variation in the anatomy, function, and ecology of organisms since the 1980s.
 While a diversity of methods have been employed, it is unclear whether they provide
@@
 
-Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
+\noindent Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
 At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g.
@@
 
-\textit{Figure disparities {[}figure\_disparities.pdf{]}: The four main types of disparity analysis.
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{Figures/figure_disparities.pdf}
+\caption{
+    The four main types of disparity analysis.
 \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
-These categories are not independent and many disparity studies will cover more than one.}
-
+    These categories are not independent and many disparity studies will cover more than one.
     %@@@ Add disparity and evo-devo
+}
+\label{Fig:disparity}
+\end{figure}
 
@@
 
-Disparity analyses are based on traits, but traits can be characterised in a number of ways:
+\noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
 % MH: We do not mention model-based descriptors at all, or the spaces defined by them (e.g.
@@
 
-%\includegraphics[width=5.63542in,height=3.875in]{media/image2.png}3.
-
-\textit{Figure data {[}figure\_data.pdf{]}: Major routes to obtain morphological data for disparity analyses.
-Data can be collected as discrete trait observations (e.g.
-presence or absence data) or as continuous data.
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{Figures/figure_data.pdf}
+\caption{
+    Major routes to obtain morphological data for disparity analyses. Data can be collected as discrete trait observations (e.g. presence or absence data) or as continuous data.
     Continuous data can be collected by various methods including linear measurements and landmark coordinates or contours (curves).
     These measurements can then be mathematically transformed (logarithm transforms, scaling, Procrustes superimposition, elliptic Fourier transforms, etc.).
-Regardless of the method, data collection produces a trait matrix where the observed traits constitute columns and the studied elements (generally taxa or OTUs) the rows.}
+    Regardless of the method, data collection produces a trait matrix where the observed traits constitute columns and the studied elements (generally taxa or OTUs) the rows.
+}
+\label{Fig:data}
+\end{figure}
 
-\section{Disparity methods}
+\section{Disparity analysis methods}
 
-Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
+\noindent Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
 Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the variety of ordination (dimension reduction) techniques available and their limitations; (c) the metrics used to summarise the relative disparity of groups; (d) the methods used for hypothesis testing within the disparity analysis framework; and (e) ancestral state estimations in disparity-through-time analyses.
@@
 
-%\includegraphics[width=6.5in,height=5.26389in]{media/image3.png}
-
-\textit{Figure methods: Morphospaces: different mathematical representations of a} \emph{morphospace.
+\begin{figure}[!htbp]
+\centering
+   \includegraphics[width=0.9\textwidth]{Figures/figure_space.pdf}
+\caption{
+    Morphospaces: different mathematical representations of a morphospace.
     % SG: The morphospace is the original trait space within which lie the specimens.
@@
 See also Mitteroecker and Huttegger (2009)
-A trait matrix can be transformed into a distance matrix (e.g. in Close et al., 2015) or an ordinated matrix (e.g. in Brusatte et al.} \emph{2008).
+    A trait matrix can be transformed into a distance matrix \citep[][e.g in]{Close2015-qi} or an ordinated matrix \citep[][e.g. in]{Brusatte2008-vx}
     In each case, all three matrices represent the morphospace for the observations at hand.
@@
     Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g.
-using multidimensional scaling} \emph{forcing all the variance to be contained in two dimensions)
+using multidimensional scaling forcing all the variance to be contained in two dimensions)
     % SG: No, the flattening will not force all the variance to be contained in 2D.
@@
 Some information (variance) is thus necessarily lost.
-; or ordination axis plots (directly from the ordinated matrix).}
+    ; or ordination axis plots (directly from the ordinated matrix).
     % ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
@@
     % @@@ Change references
+}
+\label{Fig:morphospace}
+\end{figure}
 
@@
 
-\section{4. Disparity analyses for the future}
+\section{Disparity analyses for the future}
 
-Morphological disparity analyses are widely employed in evolutionary palaeobiology, and they are based on a diversity of methods and data.
+\noindent Morphological disparity analyses are widely employed in evolutionary palaeobiology, and they are based on a diversity of methods and data.
 There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
@@
 
-\subsection{Acknowledgments}
+\subsection{Acknowledgements}
 
@@
 -TG was funded by ARC DP170103227 and FT180100634 awarded to VW;
+
+\bibliographystyle{sysbio}
+\bibliography{References}
 

### 2020-02-21 Writing/disparity_review.tex
rewritten ordination §

@@
 Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling
-	\citep[][e.g.]{Fortey1996-kt, Moyne2007-jm, Ruta2013-iy, Hopkins2013-xt}.
+	\citep[e.g.][]{Fortey1996-kt, Moyne2007-jm, Ruta2013-iy, Hopkins2013-xt}.
 	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011-fq}.
@@
 Throughout, it is important to remember these tools should always be used in the context of a specific scientific question, as this will drive data and methodological choices at every stage of the process.
-
-%\includegraphics[width=6.5in,height=4.94444in]{media/image1.png}
 
@@
 
-% -Disparity analysis have a history of using ordination techniques
 Disparity analyses often use ordination techniques for dimension reduction.
-Dimension reduction (or ordination) is a powerful tool for visualizing data distribution in disparity analyses, helping to elucidate the underlying structure within datasets.
-%     -Ordination techniques are blablabla
-These are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g.
-principal component analysis -- PCA; principal coordinates analysis -- PCO/PCoA; non-metric multidimensional scaling -- NMDS; etc.).
-Ordination techniques come in many flavours depending on the data and the desired morphospace properties.
-Quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCoA, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) (see Legendre and Legendre 2012,chapter 9, for a detailed explanation of ordination methods and properties).
+Ordinations are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g.
+principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS; etc.).
+They come in many flavours depending on the data and the desired morphospace properties.
+For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) \citep[see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012-va}.
+
 % -But mainly ordination is good for visualising.
-Ordination is advantageous for plotting and visualising data, and can reveal properties of the morphospace not intuitively captured by disparity metrics (see Metrics section below {[}metrics{]}).
-% -Ordination can be good for reducing dimensions if the data has a lot of measurements for a few number of observations (e.g. landmark data, discrete morphological characters, etc.)
-Using ordinations in disparity analysis makes it easier for researchers to comprehend patterns in just two or three spatial dimensions at a time - or, indeed, just the first few axes of a PCA or a PCO.
-Additionally, after ordinating the data, it is possible to reduce further the number of dimensions by focussing on just a subset of axes (e.g.
-selecting only those axes that describe the majority of the variation) or by fitting the variation to a prescribed number of axes.
-In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions (Legendre and Legendre 2012).
-%TG: cite also: Dryden and Mardia "Statistical Shape Analysis book"
-This has clear advantages for interpreting the results.
-For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak in birds \citep{Cooney2017-ly}.
+One of the reason why ordinations techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at the time which can be more intuitively captured than through disparity indices (see section below).
+%@@@ LINK
+Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e.
+selecting only those axes that describe the majority of the variation in the dataset - e.g.
+95\%).
+In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012-va,dryden2016statistical}.
+In practice, this can allow to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g.
+the elongation and flattening of birds beaks;][]{Cooney2017-ly}.
 
-% -However ordination can be problematic
 However, like most other aspects of disparity analyses, reducing dimensionality can be tricky.
-However, there are a number of potential problems with these transformations of the data.
-%TG: comment from Sylvain: difference between a morphospace and an ordination space
-%TG: comment from Graeme: the ordination space is decided by the data present (e.g. shoebill)
-%     -It should contain all axis
 In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results. 
-%TG: from Anjali: add that people should use 95% or 99% percent of the data (why not 100% though?)
-%     -It is hard to interpret
-Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \citep{Bookstein1997-dn, Bookstein2015-yy, Bookstein2017-qk, Bookstein2017-gu,Weisbecker2019-kp}.
- %TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
+Although a common technique is to consider the \textit{n} axis that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to miss-interpretation of the biological variation mapped on these axes \citep{Bookstein2015-yy, Bookstein2017-qk, Bookstein2017-gu,Weisbecker2019-kp}.
 Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018-le, Gerber2017-xi}.
-Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
+
+Interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
 Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
 For example, when the analysed data is non-Euclidean (e.g.
-some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \citep{Gerber2017-xi}.
-%@@@TG: should be Gerber 2019
+some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \citep{Gerber2019}.
 This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear.
-Finally, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible. %TG: Cite Gerber 2019 10.1111/pala.12407
+Furthermore, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible \citep{Gerber2019}.
 Although some gradients are easy to detect or interpret (e.g.
-the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018-ye}), some are not \citep[][e.g.]{Weisbecker2019-kp}.
+the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018-ye}), some are not \citep[e.g.][]{Weisbecker2019-kp}.
 For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0, has no biological meaning if these are binary alternate states.
-%     -It can make non-metric spaces
-Categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
+
+Finally, categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
 contrary to a PCA, the first few axis do not encompass most variance of the dataset).
-Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[][i.e.
-the distance between A and B is not equal to the distance between B and A]{Gerber2014-ol}.
-%     -It is often not necessary
+Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e.
+the distance between A and B is not equal to the distance between B and A][]{Gerber2014-ol}.
 And last but not least reason, In many cases, ordination might not be necessary.
-For example, if a metric (e.g. sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below metrics%indices
-), it is not necessary to calculate it on ordinated data \citep{Close2015-qi}.
-% -Disparity can be measured directly from the non-ordinated dataset (e.g. via distance matrices)
+For example, if a metric (e.g. sum or ranges of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015-qi}.
 For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.
-
-% \hypertarget{a-multidimensionality}{%
-% \subsection{(a) Multidimensionality}\label{a-multidimensionality}}
-
-% Disparity analyses often use ordination techniques for dimension reduction.
-% These are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g.
-principal component analysis -- PCA; principal coordinates analysis -- PCO/PCoA; non-metric multidimensional scaling -- NMDS; etc.).
-% Using ordinations in disparity analysis makes it easier for researchers to comprehend patterns in just two or three spatial dimensions at a time - or, indeed, just the first few axes of a PCA or a PCO.
-% Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(}Bookstein \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1997, 2015, {[}b{]} 2017, {[}a{]} 2017)}{[}CITE Vera and Thomas paper{]}.
-%TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
-% Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \href{https://paperpile.com/c/sTGYvp/0y4V+QVvv}{(Deline et al.
-2018; Gerber 2017)}.
-% Categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
-contrary to a PCA, the first few axis do not encompass most variance of the dataset).
-% Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric (\href{https://paperpile.com/c/sTGYvp/SJbC}{i.e.
-the distance between A and B is not equal to the distance between B and A; Gerber 2014)}.
-
-% \hypertarget{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}{%
-% \subsection{(b) To ordinate or not to ordinate - that is the
-% (multidimensional)
-% question}\label{b-to-ordinate-or-not-to-ordinate---that-is-the-multidimensional-question}}
-
-% Dimension reduction (or ordination) is a powerful tool for visualizing data distribution in disparity analyses, helping to elucidate the underlying structure within datasets.
-% However, like most other aspects of disparity analyses, reducing dimensionality can be tricky.
-% Ordination techniques come in many flavours depending on the data and the desired morphospace properties.
-% Quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCoA, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) (see Legendre and Legendre 2012,chapter 9, for a detailed explanation of ordination methods and properties).
-
-% Ordination is advantageous for plotting and visualising data, and can reveal properties of the morphospace not intuitively captured by disparity metrics (see Metrics section below {[}metrics{]}).
-% Additionally, after ordinating the data, it is possible to reduce further the number of dimensions by focussing on just a subset of axes (e.g.
-selecting only those axes that describe the majority of the variation) or by fitting the variation to a prescribed number of axes.
-% In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions (Legendre and Legendre 2012).
-%TG: cite also: Dryden and Mardia "Statistical Shape Analysis book"
-% This has clear advantages for interpreting the results.
-% For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak in birds \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017}).
-% However, there are a number of potential problems with these transformations of the data.
-% %TG: comment from Sylvain: difference between a morphospace and an ordination space
-% %TG: comment from Graeme: the ordination space is decided by the data present (e.g. shoebill)
-% In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results.
-%TG: from Anjali: add that people should use 95% or 99% percent of the data (why not 100% though?)
-% Furthermore, interpreting biological variation along the axes is always a \emph{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
-
-% However, In many cases, ordination might not be necessary.
-For example, if a metric (e.g.
-sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below {[}metrics{]}), it is not necessary to calculate it on ordinated data {[}e.g.\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al.
-2015)}{]}.
-% Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
-% For example, when the analysed data is non-Euclidean (e.g.
-some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \href{https://paperpile.com/c/sTGYvp/SJbC}{(}Gerber 2014\href{https://paperpile.com/c/sTGYvp/SJbC}{)}.
-%TG: should be Gerber 2019
-% This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear.
-% Finally, \emph{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible. %TG: Cite Gerber 2019 10.1111/pala.12407
-% Although some gradients are easy to detect or interpret (e.g.
-the elongation and depth of mandibles in fishes on first and second PC axes, respectively; {[}\href{https://paperpile.com/c/sTGYvp/3JPy}{Hill et al.
-2018{]}}), some are not (e.g.
-\href{https://paperpile.com/c/sTGYvp/TZzO}{Weisbecker et al.
-2019)}.
-% For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0, has no biological meaning if these are binary alternate states.
-% For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination (Lloyd 2016, 2018).
 
@@
 See also Mitteroecker and Huttegger (2009)
-    A trait matrix can be transformed into a distance matrix \citep[][e.g in]{Close2015-qi} or an ordinated matrix \citep[][e.g. in]{Brusatte2008-vx}
+    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}
     In each case, all three matrices represent the morphospace for the observations at hand.

### 2020-02-21 Writing/disparity_review.tex
added morphospace definition

@@
 
+%@@@ add links to figures
+
 \noindent Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
@@
 
+\subsection{From data to morphospace}
+
+The concept of what the morphospace \textit{is} can be different between authors and studies but generally used in the common (i.e.
+non-mathematical) sense of a space containing all possible morphologies.
+For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points \citep{mitteroecker2009concept}.
+In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combination of traits.
+We generalise this definition to any transformation of this space that retains the biological definition, i.e.
+the morphospace is the mathematical object that contains the observed combinations of traits or their transformation.
+Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see discussion below). %link
+If the morphospace is transformed mathematically (i.e.
+the dataset is transformed using logs, ordinations, distance matrices, etc...) it remains the morphospace as long as all the observation and traits remain in the datatset.
+
 \section{Disparity analysis methods}
@@
 \noindent Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
-Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the variety of ordination (dimension reduction) techniques available and their limitations; (c) the metrics used to summarise the relative disparity of groups; (d) the methods used for hypothesis testing within the disparity analysis framework; and (e) ancestral state estimations in disparity-through-time analyses.
-% @@@ Check the numbering here
+Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the indices used to summarise the relative disparity of groups; (c) the methods used for hypothesis testing within the disparity analysis framework; and (d) ancestral state estimations in disparity-through-time analyses.
 We consider these aspects in turn below.
 
-\subsection{a) To ordinate or not to ordinate - that is the (multidimensional) question}
+\subsection{(a) To ordinate or not to ordinate - that is the (multidimensional) question}
 
@@
 
-% -But mainly ordination is good for visualising.
 One of the reason why ordinations techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at the time which can be more intuitively captured than through disparity indices (see section below).
@@
     Morphospaces: different mathematical representations of a morphospace.
-    % SG: The morphospace is the original trait space within which lie the specimens.
-Then one performs an ordination to visualize the space.
-So I would distinguish the morphospace from the ordination space.
-See also Mitteroecker and Huttegger (2009)
     A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}

### 2020-02-21 Writing/disparity_review.tex
added precision
Added our morphospace definition. (I hope it's clear now!)

@@
 In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combination of traits.
-We generalise this definition to any transformation of this space that retains the biological definition, i.e.
+We generalise this definition to any linear transformation ($V \to W$) of this space that retains the biological definition, i.e.
 the morphospace is the mathematical object that contains the observed combinations of traits or their transformation.
@@
 the dataset is transformed using logs, ordinations, distance matrices, etc...) it remains the morphospace as long as all the observation and traits remain in the datatset.
+
+% This generalised definition that we will used throughout the manuscript considers to be a morphospace any mathematical object that contains all the traits and all the observation.
+% For example, the raw trait matrix is a morphospace, the distance matrix obtained from that one is the same morphospace, the ordination of that latter is the same morphospace, etc (again, properties of the space can change as long as all the traits and observations are used).
+% But if a space is comprised of 95\% of the axes of a PCA, this is not a morphospace anymore but a subset of it.
 
@@
     Morphospaces: different mathematical representations of a morphospace.
-    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}
-    In each case, all three matrices represent the morphospace for the observations at hand.
+    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}.
+    Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
     Visualisation: different  ways to represent the morphospace in 2D.
@@
 
+\subsection{Appendix: morphospace definition in this paper}
+
+\noindent \cite{mitteroecker2009concept} define a morphospace as:
+
+\setlength{\leftskip}{1cm}
+\setlength{\rightskip}{1cm}
+
+\noindent \textit{``[a] mathematical spaces describing and relating the phenotypic configuration of biological organisms [...].
+In a typical morphospace, the morphological configuration of an organism is represented by a single point, and the dimensionality of the space is determined by the number of measured variables (Q-space).''} \citealt{mitteroecker2009concept} p.55.
+
+\setlength{\leftskip}{0pt}
+\setlength{\rightskip}{0pt}
+
+\noindent Here, we generalise the definition of a morphospace following \cite{mitteroecker2009concept} as the Q-space or any of its linear transformation.
+The morphospace is any matrix $(a_{ij})$ with $n$ observations with $d$ trait values each or any linear transformation $f: V \to W$ of $(a_{ij})$.
+In other words the morphospace is the observed data matrix (e.g. the collected data) or any of it's linear transformation (e.g. a PCA, a distance matrix, a log, etc...).
+In \cite{mitteroecker2009concept}'s definition, the morphospace needs a conserved $n \times d$ dimensions whereas our definition propose only the conservation of $n$ dimensions (the number of observations) as long as the morphospace can be expressed as:
+
+\begin{equation}
+    M1 = f(a_{ij})\in \mathbb{R}^{{n}\times{d}}
+\end{equation}
+
+\noindent Where $f$ can be any linear transformation function (including no transformation: $f(x) = x$).
+This generalisation allows to consider linear transformations of $(a_{ij})$ as a different ``orientations'' or ``resolutions'' (not in the mathematical sense \textit{per se}) of the observed trait matrix $(a_{ij})$.
+For example, a distance matrix of $(a_{ij})$ can be represented through the following linear transformation:
+
+\begin{equation}
+    M2 = f_{a_{ij}} = D_{a_{ij}}
+\end{equation}
+
+\noindent Where $D$ is a distance function for any pair of $ij$ in the matrix (e.g. if the distance is Euclidean $D_{ij} = \sqrt{\sum(i_{d}-j_{d})^2}$).
+$M2$ is then just a different representation of the original morphospace $(a_{ij})$ where the properties between the $n$ observations are expressed as distances.
+Note that in our definition, $f_{a_{ij}}$ does not need to result in a metric (or metric Euclidean) space although this is often desired in disparity analysis.
+
+However our definition doesn't consider a matrix to be a morphospace of $(a_{ij})$ if it results from a non-linear transformation of $(a_{ij})$ matrix.
+For example if one performs an ordination (e.g. PCA) on the matrix $(a_{ij})$ such as:
+
+\begin{equation}
+    M3 = f_{a_{ij}} = P(a_{ij})
+\end{equation}
+
+\noindent Where $P$ is a principal components linear transformation.
+$M3$ is still a morphospace (i.e.
+a linear transformation of $(a_{ij})$).
+However if one only uses the two first axis of $M3$, this would result in a different space since it is no more a linear transformation of $(a_{ij})$.
+
+However, \textbf{\underline{\textit{DO NOT}}} declare that doing such a non-linear transformation is incorrect in a disparity analysis context.
+As pointed out throughout the main manuscript, different methods can be appropriate for different questions and different datasets.
+The point of this appendix is to develop \textit{our} definition of a morphospace but we do not argue that any other definition is incorrect.
+We do, however, suggest that a definition of ``morphospace'' should be at least briefly noted in papers using this mathematical or biological concept.
+
 \bibliographystyle{sysbio}
@@
 \end{document}
-
-% Donoho, D. L. (2000). High-dimensional data analysis: The curses and
-% blessings of dimensionality. \emph{AMS math challenges lecture},
-% \emph{1}(2000), 32.
-
-% References
-
-% \href{http://paperpile.com/b/sTGYvp/ZnDd}{Adams, Dean C., and Michael L.
-% Collyer. 2018. ``Multivariate Phylogenetic Comparative Methods:
-% Evaluations, Comparisons, and Recommendations.'' \emph{Systematic
-% Biology} 67 (1): 14--31.}
-
-% \href{http://paperpile.com/b/sTGYvp/J2G1}{Adams, Dean C., and Erik
-% Otárola-Castillo. 2013. ``Geomorph: Anrpackage for the Collection and
-% Analysis of Geometric Morphometric Shape Data.'' \emph{Methods in
-% Ecology and Evolution / British Ecological Society} 4 (4): 393--99.}
-
-% \href{http://paperpile.com/b/sTGYvp/SC6L}{Anderson, Marti J. 2001. ``A
-% New Method for Non-Parametric Multivariate Analysis of Variance.''
-% \emph{Austral Ecology} 26 (1): 32--46.}
-
-% \href{http://paperpile.com/b/sTGYvp/3hy2}{Anderson, Marti J., and Daniel
-% C. I. Walsh. 2013. ``PERMANOVA, ANOSIM, and the Mantel Test in the Face
-% of Heterogeneous Dispersions: What Null Hypothesis Are You Testing?''
-% \emph{Ecological Monographs} 83 (4): 557--74.}
-
-% \href{http://paperpile.com/b/sTGYvp/qjj9}{Anderson, Philip S. L., Matt
-% Friedman, Martin D. Brazeau, and Emily J. Rayfield. 2011. ``Initial
-% Radiation of Jaws Demonstrated Stability despite Faunal and
-% Environmental Change.'' \emph{Nature} 476 (7359): 206--9.}
-
-% \href{http://paperpile.com/b/sTGYvp/Qsl3}{Bellman, R. 1966. ``Dynamic
-% Programming.'' \emph{Science} 153 (3731): 34--37.}
-
-% \href{http://paperpile.com/b/sTGYvp/1SD2}{Bookstein, Fred L. 1997.
-% \emph{Morphometric Tools for Landmark Data: Geometry and Biology}.
-% Cambridge University Press.}
-
-% \href{http://paperpile.com/b/sTGYvp/sN5d}{---------. 2015. ``The
-% Relation between Geometric Morphometrics and Functional Morphology, as
-% Explored by Procrustes Interpretation of Individual Shape Measures
-% Pertinent to Function.'' \emph{Anatomical Record} 298 (1): 314--27.}
-
-% \href{http://paperpile.com/b/sTGYvp/o4w7}{---------. 2017a. ``A Newly
-% Noticed Formula Enforces Fundamental Limits on Geometric Morphometric
-% Analyses.'' \emph{Evolutionary Biology} 44 (4): 522--41.}
-
-% \href{http://paperpile.com/b/sTGYvp/xaUx}{---------. 2017b. ``A Method
-% of Factor Analysis for Shape Coordinates.'' \emph{American Journal of
-% Physical Anthropology} 164 (2): 221--45.}
-
-% \href{http://paperpile.com/b/sTGYvp/9JdS}{Bouxin, Guy. 2005. ``Ginkgo, a
-% Multivariate Analysis Package.'' \emph{Journal of Vegetation Science:
-% Official Organ of the International Association for Vegetation Science}
-% 16 (3): 355--59.}
-
-% \href{http://paperpile.com/b/sTGYvp/Yrbg}{Brazeau, Martin D., Thomas
-% Guillerme, and Martin R. Smith. 2017. ``Morphological Phylogenetic
-% Analysis with Inapplicable Data.''
-% https://doi.org/}\href{http://dx.doi.org/10.1101/209775}{10.1101/209775}\href{http://paperpile.com/b/sTGYvp/Yrbg}{.}
-
-% \href{http://paperpile.com/b/sTGYvp/tGyd}{Brusatte, S. L., M. J. Benton,
-% M. Ruta, and G. T. Lloyd. 2008. ``The First 50 Myr of Dinosaur
-% Evolution: Macroevolutionary Pattern and Morphological Disparity.''
-% \emph{Biology Letters} 4 (6): 733--36.}
-
-% \href{http://paperpile.com/b/sTGYvp/EeC8}{Brusatte, Stephen L., Michael
-% J. Benton, Marcello Ruta, and Graeme T. Lloyd. 2008. ``Superiority,
-% Competition, and Opportunism in the Evolutionary Radiation of
-% Dinosaurs.'' \emph{Science} 321 (5895): 1485--88.}
-
-% \href{http://paperpile.com/b/sTGYvp/aSSL}{Butler, Richard J., Stephen L.
-% Brusatte, Brian Andres, and Roger B. J. Benson. 2012. ``How Do
-% Geological Sampling Biases Affect Studies of Morphological Evolution in
-% Deep Time? A Case Study of Pterosaur (Reptilia: Archosauria)
-% Disparity.'' \emph{Evolution; International Journal of Organic
-% Evolution} 66 (1): 147--62.}
-
-% \href{http://paperpile.com/b/sTGYvp/60H0}{Canter, Erin J., Catalina
-% Cuellar-Gempeler, Abigail I. Pastore, Thomas E. Miller, and Olivia U.
-% Mason. 2018. ``Predator Identity More than Predator Richness Structures
-% Aquatic Microbial Assemblages in Sarracenia Purpurea Leaves.''
-% \emph{Ecology} 99 (3): 652--60.}
-
-% \href{http://paperpile.com/b/sTGYvp/ROH8}{Ciampaglio, Charles N.,
-% Matthieu Kemp, and Daniel W. McShea. 2001. ``Detecting Changes in
-% Morphospace Occupation Patterns in the Fossil Record: Characterization
-% and Analysis of Measures of Disparity.'' \emph{Paleobiology} 27 (4):
-% 695--715.}
-
-% \href{http://paperpile.com/b/sTGYvp/PbSx}{Close, Roger A., Matt
-% Friedman, Graeme T. Lloyd, and Roger B. J. Benson. 2015. ``Evidence for
-% a Mid-Jurassic Adaptive Radiation in Mammals.'' \emph{Current Biology:
-% CB} 25 (16): 2137--42.}
-
-% \href{http://paperpile.com/b/sTGYvp/RjqE}{Cooney, Christopher R., Jen A.
-% Bright, Elliot J. R. Capp, Angela M. Chira, Emma C. Hughes, Christopher
-% J. A. Moody, Lara O. Nouri, Zoë K. Varley, and Gavin H. Thomas. 2017.
-% ``Mega-Evolutionary Dynamics of the Adaptive Radiation of Birds.''
-% \emph{Nature} 542 (7641): 344--47.}
-
-% \href{http://paperpile.com/b/sTGYvp/0y4V}{Deline, Bradley, Jennifer M.
-% Greenwood, James W. Clark, Mark N. Puttick, Kevin J. Peterson, and
-% Philip C. J. Donoghue. 2018. ``Evolution of Metazoan Morphological
-% Disparity.'' \emph{Proceedings of the National Academy of Sciences of
-% the United States of America} 115 (38): E8909--18.}
-
-% \href{http://paperpile.com/b/sTGYvp/47fI}{Díaz, Sandra, Jens Kattge,
-% Johannes H. C. Cornelissen, Ian J. Wright, Sandra Lavorel, Stéphane
-% Dray, Björn Reu, et al. 2016. ``The Global Spectrum of Plant Form and
-% Function.'' \emph{Nature} 529 (7585): 167--71.}
-
-% \href{http://paperpile.com/b/sTGYvp/2KmX}{Dixon, Philip. 2003. ``VEGAN,
-% a Package of R Functions for Community Ecology.'' \emph{Journal of
-% Vegetation Science: Official Organ of the International Association for
-% Vegetation Science} 14 (6): 927.}
-
-% \href{http://paperpile.com/b/sTGYvp/krNU}{Donohue, Ian, Owen L. Petchey,
-% José M. Montoya, Andrew L. Jackson, Luke McNally, Mafalda Viana, Kevin
-% Healy, Miguel Lurgi, Nessa E. O'Connor, and Mark C. Emmerson. 2013. ``On
-% the Dimensionality of Ecological Stability.'' \emph{Ecology Letters} 16
-% (4): 421--29.}
-
-% \href{http://paperpile.com/b/sTGYvp/EPJ2}{Erwin, Douglas H. 2007.
-% ``DISPARITY: MORPHOLOGICAL PATTERN AND DEVELOPMENTAL CONTEXT.''
-% \emph{Palaeontology}.
-% https://doi.org/}\href{http://dx.doi.org/10.1111/j.1475-4983.2006.00614.x}{10.1111/j.1475-4983.2006.00614.x}\href{http://paperpile.com/b/sTGYvp/EPJ2}{.}
-
-% \href{http://paperpile.com/b/sTGYvp/Z6l6}{---------. 2011.
-% ``Evolutionary Uniformitarianism.'' \emph{Developmental Biology} 357
-% (1): 27--34.}
-
-% \href{http://paperpile.com/b/sTGYvp/yyNa}{Finlay, Sive, and Natalie
-% Cooper. 2015. ``Morphological Diversity in Tenrecs (Afrosoricida,
-% Tenrecidae): Comparing Tenrec Skull Diversity to Their Closest
-% Relatives.'' \emph{PeerJ} 3 (April): e927.}
-
-% \href{http://paperpile.com/b/sTGYvp/2Neu}{Foote, Mike. 1989.
-% ``Perimeter-Based Fourier Analysis: A New Morphometric Method Applied to
-% the Trilobite Cranidium.'' \emph{Journal of Paleontology}.
-% https://doi.org/}\href{http://dx.doi.org/10.1017/s0022336000036556}{10.1017/s0022336000036556}\href{http://paperpile.com/b/sTGYvp/2Neu}{.}
-
-% \href{http://paperpile.com/b/sTGYvp/fTJ3}{---------. 1995.
-% ``Morphological Diversification of Paleozoic Crinoids.''
-% \emph{Paleobiology} 21 (03): 273--99.}
-
-% \href{http://paperpile.com/b/sTGYvp/yqPw}{---------. 1997. ``THE
-% EVOLUTION OF MORPHOLOGICAL DIVERSITY.'' \emph{Annual Review of Ecology
-% and Systematics} 28 (1): 129--52.}
-
-% \href{http://paperpile.com/b/sTGYvp/2tbJ}{Fortey, R. A., D. E. G.
-% Briggs, and M. A. Wills. 1996. ``The Cambrian Evolutionary `explosion':
-% Decoupling Cladogenesis from Morphological Disparity.'' \emph{Biological
-% Journal of the Linnean Society. Linnean Society of London} 57 (1):
-% 13--33.}
-
-% \href{http://paperpile.com/b/sTGYvp/EETc}{Friedman, Matt. 2010.
-% ``Explosive Morphological Diversification of Spiny-Finned Teleost Fishes
-% in the Aftermath of the End-Cretaceous Extinction.'' \emph{Proceedings.
-% Biological Sciences / The Royal Society} 277 (1688): 1675--83.}
-
-% \href{http://paperpile.com/b/sTGYvp/SJbC}{Gerber, Sylvain. 2014. ``Not
-% All Roads Can Be Taken: Development Induces Anisotropic Accessibility in
-% Morphospace.'' \emph{Evolution \& Development} 16 (6): 373--81.}
-
-% \href{http://paperpile.com/b/sTGYvp/QVvv}{---------. 2017. ``The
-% Geometry of Morphospaces: Lessons from the Classic Raup Shell Coiling
-% Model.'' \emph{Biological Reviews of the Cambridge Philosophical
-% Society} 92 (2): 1142--55.}
-
-% \href{http://paperpile.com/b/sTGYvp/Uns3}{Gould, Stephen Jay. 2000.
-% \emph{Wonderful Life: The Burgess Shale and the Nature of History}.
-% Random House.}
-
-% \href{http://paperpile.com/b/sTGYvp/xDqf}{Guillerme, Thomas. 2018.
-% ``dispRity : A Modular R Package for Measuring Disparity.''
-% \emph{Methods in Ecology and Evolution / British Ecological Society} 9
-% (7): 1755--63.}
-
-% \href{http://paperpile.com/b/sTGYvp/ekU4}{Guillerme, Thomas, and Natalie
-% Cooper. 2018. ``Time for a Rethink: Time Sub-Sampling Methods in
-% Disparity-through-Time Analyses.'' \emph{Palaeontology} 61 (4):
-% 481--93.}
-
-% \href{http://paperpile.com/b/sTGYvp/u1KE}{Guillerme, Thomas, Mark N.
-% Puttick, Ariel E. Marcy, and Vera Weisbecker. n.d. ``Shifting Spaces:
-% Which Disparity or Dissimilarity Metrics Best Summarise Occupancy in
-% Multidimensional Spaces?''
-% https://doi.org/}\href{http://dx.doi.org/10.1101/801571}{10.1101/801571}\href{http://paperpile.com/b/sTGYvp/u1KE}{.}
-
-% \href{http://paperpile.com/b/sTGYvp/9Zoi}{Harmon, Luke J., Jason T.
-% Weir, Chad D. Brock, Richard E. Glor, and Wendell Challenger. 2008.
-% ``GEIGER: Investigating Evolutionary Radiations.'' \emph{Bioinformatics}
-% 24 (1): 129--31.}
-
-% \href{http://paperpile.com/b/sTGYvp/WXik}{Harvey, Paul H., and Mark D.
-% Pagel. 1998. \emph{The Comparative Method in Evolutionary Biology}.
-% Oxford University Press, USA.}
-
-% \href{http://paperpile.com/b/sTGYvp/3JPy}{Hill, Jennifer J., Mark N.
-% Puttick, Thomas L. Stubbs, Emily J. Rayfield, and Philip C. J. Donoghue.
-% 2018. ``Evolution of Jaw Disparity in Fishes.'' \emph{Palaeontology}.
-% https://doi.org/}\href{http://dx.doi.org/10.1111/pala.12371}{10.1111/pala.12371}\href{http://paperpile.com/b/sTGYvp/3JPy}{.}
-
-% \href{http://paperpile.com/b/sTGYvp/xLdm}{Hopkins, Melanie J. 2017.
-% ``How Well Does a Part Represent the Whole? A Comparison of Cranidial
-% Shape Evolution with Exoskeletal Character Evolution in the Trilobite
-% Family Pterocephaliidae.'' \emph{Palaeontology} 60 (3): 309--18.}
-
-% \href{http://paperpile.com/b/sTGYvp/vTHS}{Hopkins, Melanie J., and
-% Sylvain Gerber. 2017. ``Morphological Disparity.'' In \emph{Evolutionary
-% Developmental Biology}, 1--12.}
-
-% \href{http://paperpile.com/b/sTGYvp/hea5}{Hopkins, M. J. 2013.
-% ``Decoupling of Taxonomic Diversity and Morphological Disparity during
-% Decline of the Cambrian Trilobite Family Pterocephaliidae.''
-% \emph{Journal of Evolutionary Biology} 26 (8): 1665--76.}
-
-% \href{http://paperpile.com/b/sTGYvp/xxh5}{Hughes, Martin, Sylvain
-% Gerber, and Matthew Albion Wills. 2013. ``Clades Reach Highest
-% Morphological Disparity Early in Their Evolution.'' \emph{Proceedings of
-% the National Academy of Sciences of the United States of America} 110
-% (34): 13875--79.}
-
-% \href{http://paperpile.com/b/sTGYvp/PwyQ}{Jackson, Andrew L., Richard
-% Inger, Andrew C. Parnell, and Stuart Bearhop. 2011. ``Comparing Isotopic
-% Niche Widths among and within Communities: SIBER - Stable Isotope
-% Bayesian Ellipses in R.'' \emph{The Journal of Animal Ecology} 80 (3):
-% 595--602.}
-
-% \href{http://paperpile.com/b/sTGYvp/oFiP}{Legendre, P., and Loic F. J.
-% Legendre. 2012. \emph{Numerical Ecology}. Elsevier.}
-
-% \href{http://paperpile.com/b/sTGYvp/bCsU}{Lloyd, Graeme T. 2016.
-% ``Estimating Morphological Diversity and Tempo with Discrete
-% Character-Taxon Matrices: Implementation, Challenges, Progress, and
-% Future Directions.'' \emph{Biological Journal of the Linnean Society.
-% Linnean Society of London} 118 (1): 131--51.}
-
-% \href{http://paperpile.com/b/sTGYvp/53SJ}{---------. 2018. ``Journeys
-% through Discrete-Character Morphospace: Synthesizing Phylogeny, Tempo,
-% and Disparity.'' \emph{Palaeontology} 61 (5): 637--45.}
-
-% \href{http://paperpile.com/b/sTGYvp/dJHu}{Losos, Jonathan B. 2011.
-% \emph{Lizards in an Evolutionary Tree: Ecology and Adaptive Radiation of
-% Anoles}. Univ of California Press.}
-
-% \href{http://paperpile.com/b/sTGYvp/aVVj}{Moyne, Sébastien, and Pascal
-% Neige. 2007. ``The Space-Time Relationship of Taxonomic Diversity and
-% Morphological Disparity in the Middle Jurassic Ammonite Radiation.''
-% \emph{Palaeogeography, Palaeoclimatology, Palaeoecology} 248 (1-2):
-% 82--95.}
-
-% \href{http://paperpile.com/b/sTGYvp/EmTR}{Navarro, Nicolas. 2003. ``MDA:
-% A MATLAB-Based Program for Morphospace-Disparity Analysis.''
-% \emph{Computers \& Geosciences} 29 (5): 655--64.}
-
-% \href{http://paperpile.com/b/sTGYvp/yO2t}{Palci, Alessandro, and Michael
-% S. Y. Lee. 2018. ``Geometric Morphometrics, Homology and Cladistics:
-% Review and Recommendations.'' \emph{Cladistics: The International
-% Journal of the Willi Hennig Society}.
-% https://doi.org/}\href{http://dx.doi.org/10.1111/cla.12340}{10.1111/cla.12340}\href{http://paperpile.com/b/sTGYvp/yO2t}{.}
-
-% \href{http://paperpile.com/b/sTGYvp/tSIy}{Pierce, Stephanie E., Kenneth
-% D. Angielczyk, and Emily J. Rayfield. 2008. ``Patterns of Morphospace
-% Occupation and Mechanical Performance in Extant Crocodilian Skulls: A
-% Combined Geometric Morphometric and Finite Element Modeling Approach.''
-% \emph{Journal of Morphology} 269 (7): 840--64.}
-
-% \href{http://paperpile.com/b/sTGYvp/I0Ic}{Raup, D. M. 1961. ``THE
-% GEOMETRY OF COILING IN GASTROPODS.'' \emph{Proceedings of the National
-% Academy of Sciences of the United States of America} 47 (4): 602--9.}
-
-% \href{http://paperpile.com/b/sTGYvp/geAO}{Ruta, Marcello, Kenneth D.
-% Angielczyk, Jörg Fröbisch, and Michael J. Benton. 2013. ``Decoupling of
-% Morphological Disparity and Taxic Diversity during the Adaptive
-% Radiation of Anomodont Therapsids.'' \emph{Proceedings. Biological
-% Sciences / The Royal Society} 280 (1768): 20131071.}
-
-% \href{http://paperpile.com/b/sTGYvp/cV3v}{Saupe, Erin E., Huijie Qiao,
-% Jonathan R. Hendricks, Roger W. Portell, Stephen J. Hunter, Jorge
-% Soberón, and Bruce S. Lieberman. 2015. ``Niche Breadth and Geographic
-% Range Size as Determinants of Species Survival on Geological Time
-% Scales.'' \emph{Global Ecology and Biogeography}.
-% https://doi.org/}\href{http://dx.doi.org/10.1111/geb.12333}{10.1111/geb.12333}\href{http://paperpile.com/b/sTGYvp/cV3v}{.}
-
-% \href{http://paperpile.com/b/sTGYvp/856K}{Sebé-Pedrós, Arnau, Elad
-% Chomsky, Kevin Pang, David Lara-Astiaso, Federico Gaiti, Zohar Mukamel,
-% Ido Amit, Andreas Hejnol, Bernard M. Degnan, and Amos Tanay. 2018.
-% ``Early Metazoan Cell Type Diversity and the Evolution of Multicellular
-% Gene Regulation.'' \emph{Nature Ecology \& Evolution} 2 (7): 1176--88.}
-
-% \href{http://paperpile.com/b/sTGYvp/ZEDR}{Spriggs, Elizabeth L., Samuel
-% B. Schmerler, Erika J. Edwards, and Michael J. Donoghue. 2018. ``Leaf
-% Form Evolution in Viburnum Parallels Variation within Individual
-% Plants.'' \emph{The American Naturalist} 191 (2): 235--49.}
-
-% \href{http://paperpile.com/b/sTGYvp/Ejzr}{Wainwright, Peter C., Michael
-% E. Alfaro, Daniel I. Bolnick, and C. Darrin Hulsey. 2005. ``Many-to-One
-% Mapping of Form to Function: A General Principle in Organismal Design?''
-% \emph{Integrative and Comparative Biology} 45 (2): 256--62.}
-
-% \href{http://paperpile.com/b/sTGYvp/TZzO}{Weisbecker, Vera, Thomas
-% Guillerme, Cruise Speck, Emma Sherratt, Hyab Mehari Abraha, Alana C.
-% Sharp, Claire E. Terhune, Simon Collins, Stephen Johnston, and Olga
-% Panagiotopoulou. 2019. ``Individual Variation of the Masticatory System
-% Dominates 3D Skull Shape in the Herbivory-Adapted Marsupial Wombats.''
-% \emph{Frontiers in Zoology} 16 (November): 41.}
-
-% \href{http://paperpile.com/b/sTGYvp/nFf7}{Wills, Matthew A. 2001.
-% ``Morphological Disparity: A Primer.'' In \emph{Topics in Geobiology},
-% 55--144.}
-
-% \href{http://paperpile.com/b/sTGYvp/eZ3F}{Wills, Matthew A., Derek E. G.
-% Briggs, and Richard A. Fortey. 1994. ``Disparity as an Evolutionary
-% Index: A Comparison of Cambrian and Recent Arthropods.''
-% \emph{Paleobiology} 20 (02): 93--130.}
-
-% \href{http://paperpile.com/b/sTGYvp/s33b}{Wright, David F. 2017.
-% ``Phenotypic Innovation and Adaptive Constraints in the Evolutionary
-% Radiation of Palaeozoic Crinoids.'' \emph{Scientific Reports} 7 (1):
-% 13745.}

### 2020-02-21 Writing/disparity_review.tex
Updated until phylo methods

@@
 \section{Data and disparity}
+\label{section:data}
 
@@
 \subsection{From data to morphospace}
+\label{section:morphospace}
+
+% TG: The whole argument here is the same as for logging data in biology: we conserve the same properties of the biological observation (e.g.
+a distribution of body mass) but we modify it linearly to make it easier to analyse (e.g.
+the log(body mass)).
+In my opinion same goes with a data matrix: a distance matrix or a PCA of the observed trait matrix should no modify the biological properties as long as it it changed linearly (if you start arbitrarily removing axis of the PCA or points in the distance matrix, this results in something different).
+Therefore, morphospace = collected data matrix (any any linear transformation thereof).
 
@@
 We generalise this definition to any linear transformation ($V \to W$) of this space that retains the biological definition, i.e.
-the morphospace is the mathematical object that contains the observed combinations of traits or their transformation.
-Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see discussion below). %link
+the morphospace is the mathematical object that contains the observed combinations of traits or their transformation (see appendix \ref{section:appendixmorphospace}).
+Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see section \ref{section:ordination} below).
 If the morphospace is transformed mathematically (i.e.
@@
 \section{Disparity analysis methods}
+\label{section:methods}
 
@@
 \subsection{(a) To ordinate or not to ordinate - that is the (multidimensional) question}
+\label{section:ordination}
 
@@
 
-One of the reason why ordinations techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at the time which can be more intuitively captured than through disparity indices (see section below).
-%@@@ LINK
+One of the reason why ordinations techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at the time which can be more intuitively captured than through disparity indices (see section \ref{section:metrics} below).
 Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e.
@@
 And last but not least reason, In many cases, ordination might not be necessary.
-For example, if a metric (e.g. sum or ranges of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015-qi}.
+For example, if an index (e.g. sum or ranges of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015-qi}.
 For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.
@@
 \caption{
+    \tiny{
     Morphospaces: different mathematical representations of a morphospace.
@@
     Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g.
-using multidimensional scaling forcing all the variance to be contained in two dimensions)
-    % SG: No, the flattening will not force all the variance to be contained in 2D.
-First because NMDS is non metric and variance is therefore not preserved as such.
-Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation.
-Some information (variance) is thus necessarily lost.
-    ; or ordination axis plots (directly from the ordinated matrix).
-    % ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
-    % SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..).
-But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
+using non-metric multidimensional scaling forcing the dataset into two dimensions - note that variance is lost due to the distortion stress and non-metric properties of the resulting operation) ; or ordination axis plots (directly from the ordinated matrix).
+    Note that in 2D representations, we consider it as good practice to plot both axis with the same scale (so to not visually distort the importance of one axis).
+    }
     % @@@ Change references
@@
 
-% @@@ Change to indices throughout
 \subsection{(b) Measuring disparity}
+\label{section:metrics}
 
 Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
-This is usually achieved with a disparity metric or index \citep{Hopkins2017-cf}.
-As with any summary of multidimensional data, disparity metrics will reflect only some aspects of the morphological variation, never its whole complexity %metrics paper. CITE moms
-It is therefore often beneficial to use more than one metric to summarise different aspects of variation, guided by the aim of the study.
+This is usually achieved with a disparity measure or index \citep{Hopkins2017-cf}.
+As with any summary of multidimensional data, disparity indices will reflect only some aspects of the morphological variation, never its whole complexity \citep{GuillermeMOMS}.
+It is therefore often beneficial to use more than one index to summarise different aspects of variation, guided by the aim of the study.
 
-When considering only one dimension, disparity metrics can be used to reflect the spread of the distribution (e.g.
+When considering only one dimension, disparity indices can be used to reflect the spread of the distribution (e.g.
 the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
-Among these metrics, some will have more attractive properties than others, such as sensitivity to outliers.
+Among these indices, some will have more attractive properties than others, such as sensitivity to outliers.
 Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
 For example, if we want to characterise the extent of morphospace occupied by a group (e.g.
-does group A occupy as much space as group B?), metrics related to the spread of the group in the morphospace are most appropriate (e.g.
+does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g.
 volume
 \citealt{Diaz2016-mr}, distance from the centroid \citealt{Hopkins2017-cf, Finlay2015-ft}, variance and range \citealt{Brusatte2008-fx}).
-Conversely, if we wish to describe the ``position''
-% SG: Note that conceptually and technically,  disparity is about variation.
-So information on position is another aspect of morphospace occupation.
-but it is distinct from the meaning of  disparity as originally intended.
- of a group in a morphospace (e.g.
-does group A occupy the same space as group B?), metrics related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate.
-%@@@ CITE moms
+Furthermore, different aspects than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g.
+does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
 Finally, if we aim to characterise the density of morphospace occupation (e.g.
-is group A more closely packed than group B?) metrics related to the pairwise distances between elements will be most appropriate (e.g.
+is group A more closely packed than group B?) indices related to the pairwise distances between elements will be most appropriate (e.g.
 nearest neighbour distance, pairwise distances, etc.
-\citealt{Close2015-qi}).
-%ES: this para is probably most useful to refer to for the ordinate or not part
+\citealt{Close2015-qi} - see section \ref{section:ordination} above).
 
-In addition to considering what properties of disparity metrics should capture, it is important to also consider the mathematical properties of the metrics and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
-For example, measuring the full sum of the variance in each dimension of the space does not require we add the covariance between the axes in an ordinated space using a PCA.
-% SG: Not very clear to me
-However, this is not true of other mathematical spaces or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012-va}.
+In addition to considering what properties of disparity indices should capture, it is important to also consider the mathematical properties of the indices and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
+For example, measuring the sum of variance-covariance for each dimension of the space is equivalent of measuring the sum of variance in a PCA (since the co-variance is null).
+However, this is not true of other transformations of the space or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012-va}.
 
 Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \citep{Bellman1966-mc}.
-In spaces with some axis of variance lower than 1, product-based metrics used as proxies of volumes (e.g.
+In spaces with some axis of variance lower than 1, product-based indices used as proxies of volumes (e.g.
 product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966-mc}.
-%Donoho 2000
-Some other types of metrics are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Butler2012-tr}
-%SG: Or even better, twenty years earlier, Foote (1992), who orginally described the problem in the context of disparity and suggested a rarefaction approach.
-or convex hull based metrics \citep{Butler2012-tr, Jackson2011-kq}.
+Some other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992-zs} or convex hull based indices \citep{Jackson2011-kq}.
 
-\subsection{(c) Testing hypotheses on the evolution of
-disparity}
+\subsection{(c) Testing hypotheses on the evolution of disparity}
+\label{section:testing}
 
-No matter which disparity metrics have been calculated, the research question must be framed in an appropriate statistical context.
-The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \citep{Adams2018-mg}
-%AG: there are others that should be mentioned - mvmorph for example, which is especially important given that there are extremely different approaches to multivariate data in these different packages
-, but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
+No matter which disparity indices have been calculated, the research question must be framed in an appropriate statistical context.
+The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018-mg}, but some of these advances have yet to be implemented in disparity analyses.
+Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
 One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001-qb, Anderson2013-zt}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
 Although statistically valid, this test is not always directly related to the hypothesis under question.
-For example, PERMANOVA tests whether two groups share the same variance/covariance in a ``distance-space''.
-% SG: What is meant by distance-sapce and in what way is it distinguished from the spaces discussed above?
-This is not the same as testing whether the two groups overlap in morphospace.
-%SG: You're sure of this?
+For example, PERMANOVA tests whether two groups share the same variance/covariance based on a distance matrix between observations and therefore does not require the morphospace to be ordinated (see section \ref{section:ordination} above).
 Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.
@@
 It is also important to consider which data should be subjected to a statistical test.
-% AG: It would be good to also note here that data should be comparable in terms of having a common reference frame - e.g.
-comparing disparities across elements with different numbers of landmarks or different landmark configurations that haven't been/can't be superimposed together is problematic but is popping up in the disparity literature
 For example, in morphological disparity analysis, especially for palaeobiological questions, data are often bootstrapped.
-This has two advantages: (i) when the disparity metric is unidimensional (e.g.
-the sum of variances), bootstrapping the data generates a distribution of the metric that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
+This has two advantages: (i) when the disparity index is unidimensional (e.g.
+the sum of variances), bootstrapping the data generates a distribution of the index that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
 However, bootstrapped data are pseudoreplicates and thus non-independent.
@@
 Furthermore, the number of bootstrap pseudoreplicates will inevitably increase the false positive rate (Type I error).
-% @@@ check positive/negative type
 These factors are often ignored in disparity analyses.
@@
 % MH: I might be wrong about this, but my impression is that CIs generated from bootstrapping are compared visually (like to assess significance in changes in disparity over time), but that statistical tests are not commonly applied.
+Finally, it is important to understand the limitations of the dataset for performing statistical analysis.
+Mainly, disparity analysis should be restrained to groups within the same morphospace and are more difficult between different morphospaces.
+This can be the case when comparing elements with different numbers of landmarks or different landmark configurations for example: they will result in different morphospaces and comparing disparity indices between them is not trivial.
 
 %TG: Add section d) within c)
-\subsection{(d) Phylogenetic autocorrelation}
+\subsection{(d) Disparity and phylogeny}
+\label{section:phylo}
 
 As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998-xg}.
-Thus, for many disparity analyses, phylogenetic relationships should be
-taken into account.
-% SG: Why?
+Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common) the non-independence between observations should be taken into account to correctly measure phenomenons.
 However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate
@@
 % AG: as long as you are using most or ideally all of the variance in a dataset, the rotation of pPCA should have little to no impact on disparity analyses. polly et al 2013 i think
-provide a thorough review of multivariate phylogenetic comparative methods, and so we do not consider them further here.
-% AG: Given this is given quite short treatment, i think if you need room, you could cut it altogether or just have a sentence in another section stating this even more briefly
 
-%TG: Add section e) within c)???
-\subsection{(e) Ancestral state estimation in disparity-through-time analyses}
+        % provide a thorough review of multivariate phylogenetic comparative methods, and so we do not consider them further here.
+        % % AG: Given this is given quite short treatment, i think if you need room, you could cut it altogether or just have a sentence in another section stating this even more briefly
+
+        % %TG: Add section e) within c)???
+        % \subsection{(e) Ancestral state estimation in disparity-through-time analyses}
+        % \label{section:ACE}
 
@@
 \section{Disparity analyses for the future}
+\label{section:future}
 
@@
 \subsection{Appendix: morphospace definition in this paper}
+\label{section:appendixmorphospace}
 

### 2020-02-21 Writing/disparity_review.tex
Full updated up to the evo-devo part

@@
 \noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
-% MH: We do not mention model-based descriptors at all, or the spaces defined by them (e.g.
-Raup's shell coiling models).
-I'm not convinced that we must, but it may come across as an notable omission to reviewers
-% SG: I agree. There's for instance Saunders et al. (2004) disparity study of Paleozoic ammonoids using Raup's shell coiling parameters
-	1) discrete ``cladistic'' characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g]{Close2015-qi};
+	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g]{Foote1989-fd, Deline2018-le};
 	2) continuous measurements of features (e.g., lengths) \citep[][e.g]{Anderson2001-qb}; or
 	3) more mathematical descriptors from geometric morphometric landmark data (e.g.
-Procrustes coordinates) \citep[][e.g]{Cooney2017-ly}, and Fourier coefficients \citep[][e.g]{Foote1995-do, Spriggs2018-nu} (Fig.
+Procrustes coordinates) \citep[][e.g]{Cooney2017-ly}, Fourier coefficients \citep[][e.g]{Foote1995-do, Spriggs2018-nu} or model-based descriptors \citep{Raup1961-vx,Saunders2004} (Fig.
 2).
@@
 The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
-Discrete character data are commonly recycled from phylogenetic studies \citep[][e.g]{Foote1989-fd, Deline2018-le}.
-% SG: Foote (1989) is a study of trilobite disparity from Fourier analysis of cranidia. I don't think it has anything to do with discrete data or phylogenetic studies
-% MH: Also the Foote papers that used discrete data weren't taken from phylogenetic studies but collected for the specific purpose of the project -- I think this might be true for the Deline paper as well.
-But there are many studies out there that used matrices initially intended for phylogenetics.
-I can look for some if needed.
-This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups \citep{Foote1995-do}.
+Discrete character data can also be use
+
+are commonly recycled from phylogenetic studies \citep[][e.g]{Brusatte2008-fx,Close2015-qi,Wright2017-jo}.
+This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups.
 This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
@@
 
-% This generalised definition that we will used throughout the manuscript considers to be a morphospace any mathematical object that contains all the traits and all the observation.
-% For example, the raw trait matrix is a morphospace, the distance matrix obtained from that one is the same morphospace, the ordination of that latter is the same morphospace, etc (again, properties of the space can change as long as all the traits and observations are used).
-% But if a space is comprised of 95\% of the axes of a PCA, this is not a morphospace anymore but a subset of it.
-
 \section{Disparity analysis methods}
@@
 \noindent Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
-Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the indices used to summarise the relative disparity of groups; (c) the methods used for hypothesis testing within the disparity analysis framework; and (d) ancestral state estimations in disparity-through-time analyses.
+Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the indices used to summarise the relative disparity of groups; (c) the methods used for hypothesis testing within the disparity analysis framework; and (d) the influence of phylogeny on disparity analyses.
 We consider these aspects in turn below.
@@
 These factors are often ignored in disparity analyses.
-% VW: Surely there's a reference for this?
-% MH: I might be wrong about this, but my impression is that CIs generated from bootstrapping are compared visually (like to assess significance in changes in disparity over time), but that statistical tests are not commonly applied.
 Finally, it is important to understand the limitations of the dataset for performing statistical analysis.
@@
 
-%TG: Add section d) within c)
 \subsection{(d) Disparity and phylogeny}
@@
 Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common) the non-independence between observations should be taken into account to correctly measure phenomenons.
-However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate
-% SG: In what way?
-if incorrect assumptions are made about the data \citep{Uyeda2015}.
-% @@@ref: https://academic.oup.com/sysbio/article/64/4/677/1649888
-% SG: Which assumption? This part (e) seems a bit vague to me
-% AG: as long as you are using most or ideally all of the variance in a dataset, the rotation of pPCA should have little to no impact on disparity analyses. polly et al 2013 i think
+However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate, especially when using only the first $n$ axis of the ordination and can lead to artefactual interpretations of the data \citep[such as wrongly supported ``early burst'' type patterns;][]{Uyeda2015}.
 
-        % provide a thorough review of multivariate phylogenetic comparative methods, and so we do not consider them further here.
-        % % AG: Given this is given quite short treatment, i think if you need room, you could cut it altogether or just have a sentence in another section stating this even more briefly
-
-        % %TG: Add section e) within c)???
-        % \subsection{(e) Ancestral state estimation in disparity-through-time analyses}
-        % \label{section:ACE}
-
-Disparity-through-time analyses often use ancestral state estimation to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
-Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-ordination, i.e.
+One other common way to taken phylogeny into account in disparity analyses is using ancestral sates estimations in disparity-through-time analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
+Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e.
 the estimation is done before transformation of the data (e.g.
-ordination, or distance matrix construction) and is simply based on the original data; or (2) post-ordination, i.e.
+ordination, or distance matrix construction) and is simply based on the original data; or (2) post-transformation, i.e.
 the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g.
@@
 
-Pre-ordination ancestral state estimation will change the way the ordination space is defined -- i.e.
+Pre-transformation ancestral state estimation will change the way the ordination space is defined -- i.e.
 the relationship between the points are not yet estimated -- and requires longer computational times.
 However, once the morphospace is defined its properties will not change.
-Post-ordination ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \citep{lloyd2018}.
+Post-transformation ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \citep{lloyd2018}.
 
-No ancestral state estimation method is without drawbacks
-% @@@ Maybe check in Pennel’s “fossils are needed” paper for ref.
- and above all else are highly dependent on the data and method used.
+No ancestral state estimation method is without drawbacks and above all else are highly dependent on the data and method used \cite{Louca2019}.
 In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points to increase statistical power.
@@
 Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
-We advocate the bespoke
-%VW: I think there could be just a line or two that takes makes "bespoke design" less daunting.
-There are loads of well-documented R packages available in morphological and ecological studies that can be relatively straightforwardly adapted.
-assembly of analytic approaches and, given the computational efficiency of these methods, an experimental approach that explores the impact of competing approaches, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
-Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets
-\citep{Donohue2013-bg, Saupe2015-vm, Canter2018-hk, mammola2019}
-. % @@@ Add Mammola
+We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
+Thankfully this is now getting easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc}.
+Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013-bg, Saupe2015-vm, Canter2018-hk, mammola2019}.
 Innovations in morphological disparity analyses likely await discovery in their respective literatures.
-% AG: as above, i think you need to add in evodevo work here, it's really fundamental to disparity
 
 While studies of morphological disparity would benefit from advances in multidimensional analysis in other fields, the concept of a morphospace could reciprocally benefit other disciplines.
-For example, the multidimensional analysis of Diaz \citealt{Diaz2016-mr}, which analysed patterns of form and function in plants, is essentially an ecomorphospace; isotopic analyses of organisms \citep{Jackson2011-kq} can be represented as an isotope-space; ecosystem functioning in \citealt{Donohue2013-bg} as an ecosystem-space, etc.
-% AJ: this is a potential nice, although rather data hungry I believe, methods that I think could be worth dropping into the paper somewhere.
-Either here as a quick drop, or earlier up under overlap seems appropriate if you want to add it https://esajournals.onlinelibrary.wiley.com/doi/10.1890/14-0235.1
-These generalisations could also be exported for any set of traits (e.g. acousto-spaces for acoustic traits or glotto-spaces for linguistic traits).
-%SG: devising 
-Cognate approaches have been adopted recently in the analysis of single cell comparative transcriptome data \citep{Sebe-Pedros2018-sw} where interpretation of the resulting transcriptome-spaces would be improved by heeding the concerns we highlight concerning morphospaces.
-% ES: Suggested by Erin: reword this sentence
+For example, the multidimensional analysis of \citep{Diaz2016-mr}, which analysed patterns of form and function in plants, is essentially an eco-morphospace; isotopic analyses of organisms \citep{Jackson2011-kq,Swanson2015} can be represented as an isotope-space; ecosystem functioning in \citealt{Donohue2013-bg} as an ecosystem-space, etc.
+These generalisations could also be exported for any set of traits: cognate approaches have been adopted in the analysis of single cell comparative transcriptome data \citep{Sebe-Pedros2018-sw} where interpretation of the resulting transcriptome-spaces would be improved by giving careful attention to the concerns we highlight concerning morphospaces.
 
-Although disparity analyses are now simple to implement in freely available softwares \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc}.
-it is crucial to remember that they are multidimensional analyses and multidimensional analyses are complex.
+Although disparity analyses are now simple to implement in freely available softwares \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc} it is crucial to remember that they are multidimensional analyses and multidimensional analyses are complex.
 We assert that future morphological analyses will benefit by emphasising the methodological decisions made, rather than simply using disparity analysis because \textit{we can}.

### 2020-02-24 Writing/disparity_review.tex
Updated version

@@
 \setlength{\parindent}{0.5in}
-\setcounter{secnumdepth}{0} 
-\renewcommand{\section}[1]{%
-\bigskip
-\begin{center}
-\begin{Large}
-\normalfont\scshape #1
-\medskip
-\end{Large}
-\end{center}}
-\renewcommand{\subsection}[1]{%
-\bigskip
-\begin{center}
-\begin{large}
-\normalfont\itshape #1
-\end{large}
-\end{center}}
-\renewcommand{\subsubsection}[1]{%
-\vspace{2ex}
-\noindent
-\textit{#1.}---}
-\renewcommand{\tableofcontents}{}
+% \setcounter{secnumdepth}{0} 
+% \renewcommand{\section}[1]{%
+% \bigskip
+% \begin{center}
+% \begin{Large}
+% \normalfont\scshape #1
+% \medskip
+% \end{Large}
+% \end{center}}
+% \renewcommand{\subsection}[1]{%
+% \bigskip
+% \begin{center}
+% \begin{large}
+% \normalfont\itshape #1
+% \end{large}
+% \end{center}}
+% \renewcommand{\subsubsection}[1]{%
+% \vspace{2ex}
+% \noindent
+% \textit{#1.}---}
+% \renewcommand{\tableofcontents}{}
 %\bibpunct{(}{)}{;}{a}{}{,}
@@
 
+\modulolinenumbers[1]
+\linenumbers
+
 \textbf{Abstract (200 words max)}
@@
 
-\textbf{Keywords}: multidimensionality, palaeobiology, ecology,
-morphology, disparity, variance/variation
+\textbf{Keywords}: multidimensionality, palaeobiology, ecology, morphology, disparity, variance/variation
 
@@
 At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g.
-cichlids and molluscs), species-rich but phenotypically conservative (e.g.
-bacteria and nematodes), species-poor but phenotypically diverse (e.g.
+cichlids or molluscs), species-rich but phenotypically conservative (e.g.
+rodents or nematodes), species-poor but phenotypically diverse (e.g.
 Afrotheria), or depauperate in both species and phenotypic diversity (e.g.
@@
 However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
-Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (taxa) can be placed in this space based on their trait values.
+Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (e.g.
+taxa) can be placed in this space based on their trait values.
 Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):
@@
 \citep{Foote1995-do, Raup1961-vx, Gerber2017-xi}.
+    More recently, this approach as also been used to understand the relation between developmental processes and morphology in the evo-devo field.
+    For example, describing disparity patterns has been successfully used to compare modules of evolution in various groups \citep{goswami2010influence,bardua2019morphological} allowing to successfully link the variation in shape to a group's evolutionary or developmental constraints \citep{Hipsley2017}.
 
@@
 
-	\item \textbf{Disparity and evo-devo} Morphological disparity blalbalbal
-% AG: Although intrinsic constraints are mentioned earlier, it doesn’t enter into any of these topics explicitly, but it is certainly one of the most interesting things one can use disparity for – to bridge development and evolution, looking at developmental biases/constraints.
-This also includes integration and modularity, which are also central to the topic of disparity and evolvability.
-Although you touch on these in places, I think it's really important to incorporate it into one of these topics, as there probably isn't room to give it one of its own (although i would argue it warrants equal footing with ecology, given that these are the processes that generate variation in the first place).
-There is a massive literature on using disparity to test hypotheses of developmental constraints, and also a huge one on disparity and integration/modularity.
-Let me know if you want me to send you refs.
+	% \item \textbf{Disparity and evo-devo}
+ %    %TG: This section might be linked to "Descriptive disparity" above but Anjali might probably disagree.
+ %    The disparity of a group can also give information on a group's evolutionary or developmental constraints.
+ %    test hypotheses of developmental constraints,
+ %    disparity and integration/modularity
 
@@
 
-Fundamental insights into evolutionary biology have been elicited from these five types of analysis.
+Fundamental insights into evolutionary biology have been elicited from these four types of analysis.
 One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997-nl, Erwin2007-jz, Hughes2013-td}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007-jz, Hughes2013-td}.
@@
     The four main types of disparity analysis.
-\textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
+\textbf{Descriptive disparity} focuses on describing the features of morphospace occupation and can be linked to developmental constraints; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
     These categories are not independent and many disparity studies will cover more than one.
@@
 
-\section{Data and disparity}
-\label{section:data}
+\section{Data and disparity} \label{section:data}
 
@@
 For example, if investigating variation of bat wing shapes, both homologous landmarks and continuous measurements of bones may be appropriate to capture patterns of wing variation.
-However, if the question focuses on comparing wings between bats and birds, different measurements might be more appropriate depending on the specific question: for example if the focus is on wing function, i.e.
-whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
+However, if the question focuses on comparing wings between bats and birds, different measurements might be more appropriate depending on the specific question: if the focus is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
 wingspan, aspect ratio, etc.).
@@
 The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
-Discrete character data can also be use
-
-are commonly recycled from phylogenetic studies \citep[][e.g]{Brusatte2008-fx,Close2015-qi,Wright2017-jo}.
+Discrete character are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008-fx,Close2015-qi,Wright2017-jo}.
 This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups.
@@
 
-\subsection{From data to morphospace}
-\label{section:morphospace}
+\subsection{From data to morphospace} \label{section:morphospace}
 
@@
 
-The concept of what the morphospace \textit{is} can be different between authors and studies but generally used in the common (i.e.
+The concept of what the morphospace \textit{is} can differ between authors and studies but is generally used in the common (i.e.
 non-mathematical) sense of a space containing all possible morphologies.
-For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points \citep{mitteroecker2009concept}.
+For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points.
 In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combination of traits.
 We generalise this definition to any linear transformation ($V \to W$) of this space that retains the biological definition, i.e.
-the morphospace is the mathematical object that contains the observed combinations of traits or their transformation (see appendix \ref{section:appendixmorphospace}).
+the morphospace contains all the observed combinations of traits (see appendix \ref{section:appendixmorphospace}).
 Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see section \ref{section:ordination} below).
@@
 
-\section{Disparity analysis methods}
-\label{section:methods}
+\section{Disparity analysis methods} \label{section:methods}
 
 \noindent Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
-Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the indices used to summarise the relative disparity of groups; (c) the methods used for hypothesis testing within the disparity analysis framework; and (d) the influence of phylogeny on disparity analyses.
+Study design encompasses several key aspects including \ref{section:ordination} the difficulty of dealing with multidimensional data; \ref{section:metrics} the indices used to summarise the relative disparity of groups; \ref{section:testing} the methods used for hypothesis testing within the disparity analysis framework; and \ref{section:phylo} the influence of phylogeny on disparity analyses.
 We consider these aspects in turn below.
 
-\subsection{(a) To ordinate or not to ordinate - that is the (multidimensional) question}
-\label{section:ordination}
+\subsection{To ordinate or not to ordinate - that is the (multidimensional) question} \label{section:ordination}
 
@@
 For example, when the analysed data is non-Euclidean (e.g.
-some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \citep{Gerber2019}.
+some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
 This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear.
@@
 
-Finally, categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
+Categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
 contrary to a PCA, the first few axis do not encompass most variance of the dataset).
@@
 
-\subsection{(b) Measuring disparity}
-\label{section:metrics}
+\subsection{Measuring disparity} \label{section:metrics}
 
@@
 
-\subsection{(c) Testing hypotheses on the evolution of disparity}
-\label{section:testing}
+\subsection{Testing hypotheses on the evolution of disparity} \label{section:testing}
 
@@
 
-\subsection{(d) Disparity and phylogeny}
-\label{section:phylo}
+\subsection{Disparity and phylogeny} \label{section:phylo}
 
@@
 
-\section{Disparity analyses for the future}
-\label{section:future}
+\section{Disparity analyses for the future} \label{section:future}
 
@@
 This article results from discussion at the Royal Society International Science Seminar on `Reconciling disparate views on disparity' held at Chicheley Hall, 9-10th January, 2018.
+AG was funded by European Research Council Starting grant 637171 ADaPTiVE;
+ALJ was funded by an Irish Research Council Laureate Award IRCLA/2017/186;
+ES was funded by a University of Adelaide Research Fellowship;
+EES was funded by a Leverhulme Trust Research Project Grant (DGR01020);
+PD was funded by NERC (NE/P013678/1; NE/N002067/1) and BBSRC (BB/N000919/1);
+SB was funded by European Research Council (ERC) under the European Union's Horizon 2020 research and innovation program (grant agreement No 756226, ERC Starting Grant: PalM) and a Leverhulme Trust Research Project Grant (RPG-2017-167);
+TG was funded by ARC DP170103227 and FT180100634 awarded to VW;
 
--AG was funded by European Research Council Starting grant 637171 ADaPTiVE.
-
--ALJ was funded by an Irish Research Council Laureate Award IRCLA/2017/186. ES was funded by a University of Adelaide Research Fellowship;
-
--EES was funded by a Leverhulme Trust Research Project Grant (DGR01020).
-
--PD was funded by NERC (NE/P013678/1; NE/N002067/1) and BBSRC (BB/N000919/1);
-
--SB was funded by European Research Council (ERC) under the European Union's Horizon 2020 research and innovation program (grant agreement No 756226, ERC Starting Grant: PalM) and a Leverhulme Trust Research Project Grant (RPG-2017-167);
-
--TG was funded by ARC DP170103227 and FT180100634 awarded to VW;
-
-\subsection{Appendix: morphospace definition in this paper}
-\label{section:appendixmorphospace}
+\subsection{Appendix: morphospace definition in this paper} \label{section:appendixmorphospace}
 
@@
 
-However our definition doesn't consider a matrix to be a morphospace of $(a_{ij})$ if it results from a non-linear transformation of $(a_{ij})$ matrix.
+Our definition doesn't consider a matrix to be a morphospace of $(a_{ij})$ if it results from a non-linear transformation of $(a_{ij})$ matrix.
 For example if one performs an ordination (e.g. PCA) on the matrix $(a_{ij})$ such as:
@@
 
-However, \textbf{\underline{\textit{DO NOT}}} declare that doing such a non-linear transformation is incorrect in a disparity analysis context.
+However, we \textbf{\underline{\textit{DO NOT}}} declare that doing such a non-linear transformation is incorrect in a disparity analysis context.
 As pointed out throughout the main manuscript, different methods can be appropriate for different questions and different datasets.

### 2020-02-25 Writing/disparity_review.tex
Edits to intro

@@
 $^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
-$^2$IDepartment of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
+$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
 $^{3}$School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
@@
 rodents or nematodes), species-poor but phenotypically diverse (e.g.
-Afrotheria), or depauperate in both species and phenotypic diversity (e.g.
-lungfish).
-These phenomena suggest that taxonomic and phenotypic diversity are not inextricably linked, raising important questions about how phenotypic diversity arises, such as:
+Afrotheria), % NC: Can we expand Afrotheria a bit?
+All the other examples are using common names.
+Can we use common names of afrotheres?
+Or at least say afrotherian mammals?
+or depauperate in both species and phenotypic diversity (e.g. lungfish).
+These phenomena suggest that taxonomic and phenotypic diversity are not necessarily linked, raising important questions about how phenotypic diversity arises, such as:
 How does diversity evolve?
 Are some morphologies more common than others?
-Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve? (Mike Foote 1997, 1995)
+Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve?
+(Mike Foote 1997, 1995) % NC: Seems weird only having these two refs here and no refs in the other questions.
+Remove?
 What role does ecology play in structuring morphological diversity?
-Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are a comparatively more recent phenomenon.
-Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
+Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are comparatively more recent. 
+Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} the concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
 However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
@@
 
-	\item \textbf{Descriptive disparity} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995-do, Briggs1992-pd}.
+	\item \textbf{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995-do, Briggs1992-pd}.
 	These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace?
 \citep{Foote1995-do, Raup1961-vx, Gerber2017-xi}.
-    More recently, this approach as also been used to understand the relation between developmental processes and morphology in the evo-devo field.
-    For example, describing disparity patterns has been successfully used to compare modules of evolution in various groups \citep{goswami2010influence,bardua2019morphological} allowing to successfully link the variation in shape to a group's evolutionary or developmental constraints \citep{Hipsley2017}.
+    More recently, this approach as also been used to understand the relationship between developmental processes and morphology in the evo-devo field.
+    For example, describing disparity patterns has been successfully used to compare modules of evolution in various groups \citep{goswami2010influence,bardua2019morphological} allowing researchers to successfully link variation in shape to a group's evolutionary or developmental constraints \citep{Hipsley2017}.
 
-	\item \textbf{Disparity-through-time} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
+	\item \textbf{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
 	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{Guillerme2018-uj, Wright2017-jo}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010-ve}?
 
-	\item \textbf{Disparity and taxonomic diversity} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e.
+	\item \textbf{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e.
 shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
 	This makes disparity an informative complement to diversity measures based on species richness alone. 
-Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling
+	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two
 	\citep[e.g.][]{Fortey1996-kt, Moyne2007-jm, Ruta2013-iy, Hopkins2013-xt}.
@@
 
-	\item \textbf{Disparity as a proxy for ecology} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
+	\item \textbf{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
 	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles
@@
 
-Fundamental insights into evolutionary biology have been elicited from these four types of analysis.
+Fundamental insights into evolutionary biology have been elicited from these four types of disparity analysis.
 One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997-nl, Erwin2007-jz, Hughes2013-td}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007-jz, Hughes2013-td}.

### 2020-02-25 Writing/disparity_review.tex
Edits to part 2

@@
     The four main types of disparity analysis.
-\textbf{Descriptive disparity} focuses on describing the features of morphospace occupation and can be linked to developmental constraints; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
+\textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
     These categories are not independent and many disparity studies will cover more than one.
     %@@@ Add disparity and evo-devo
+    % NC: I removed the mention of constraints because I hate the idea it's all constraints - there are other things at play why are we only mentioning constraints?
+(I suspect Anjali or Vera put that in?) Also it didn't fit with the other examples which don't mention mechanisms...
 }
@@
 \noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
-	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g]{Foote1989-fd, Deline2018-le};
-	2) continuous measurements of features (e.g., lengths) \citep[][e.g]{Anderson2001-qb}; or
+	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g.][]{Foote1989-fd, Deline2018-le};
+	2) continuous measurements of features (e.g. lengths) \citep[e.g][]{Anderson2001-qb}; or
 	3) more mathematical descriptors from geometric morphometric landmark data (e.g.
-Procrustes coordinates) \citep[][e.g]{Cooney2017-ly}, Fourier coefficients \citep[][e.g]{Foote1995-do, Spriggs2018-nu} or model-based descriptors \citep{Raup1961-vx,Saunders2004} (Fig.
+Procrustes coordinates) \citep[e.g.][]{Cooney2017-ly}, Fourier coefficients \citep[e.g.][]{Foote1995-do, Spriggs2018-nu} or model-based descriptors \citep[e.g.][]{Raup1961-vx,Saunders2004} (Fig.
 2).
-None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.
+None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.
 
 For example, if investigating variation of bat wing shapes, both homologous landmarks and continuous measurements of bones may be appropriate to capture patterns of wing variation.
-However, if the question focuses on comparing wings between bats and birds, different measurements might be more appropriate depending on the specific question: if the focus is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
+If the question focuses on comparing wings between bats and birds, however, different measurements might be more appropriate depending on the specific question: if the focus is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
 wingspan, aspect ratio, etc.).
@@
 digit length, integumentary system, etc.).
-Where there is any doubt about the appropriate traits to choose, it may be preferable to use different kinds of data for the same feature to determine whether they capture the same pattern of disparity.
+Where there is any doubt about the appropriate traits to choose, it may be preferable to use several different kinds of data for the same feature to determine whether they capture the same pattern of disparity.
 
@@
 Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017-cf}.
-This effect of anatomical part on disparity patterns can be especially challenging when working on datasets where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018-le}.
+This effect of the anatomical feature used on disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018-le}.
 
@@
 Ultimately, disparity analyses characterise the data, and subsamples of the universe of possible data may not have the power to uncover holistic patterns of disparity.
-Therefore, trait data should be collected with the question in mind, or the question asked should be tailored to the limits of the data available.
+Therefore, trait data should be collected with the scientific question in mind, or the question asked should be tailored to the limits of the data available.
 
@@
 \caption{
-    Major routes to obtain morphological data for disparity analyses. Data can be collected as discrete trait observations (e.g. presence or absence data) or as continuous data.
+    Major routes to obtain morphological data for disparity analyses. Data can be collected as discrete trait observations (e.g. presence-absence data) or as continuous data.
     Continuous data can be collected by various methods including linear measurements and landmark coordinates or contours (curves).
-    These measurements can then be mathematically transformed (logarithm transforms, scaling, Procrustes superimposition, elliptic Fourier transforms, etc.).
+    These measurements can then be mathematically transformed (logarithmic transformations, scaling, Procrustes superimposition, elliptic Fourier transforms, etc.).
     Regardless of the method, data collection produces a trait matrix where the observed traits constitute columns and the studied elements (generally taxa or OTUs) the rows.
@@
 
-The concept of what the morphospace \textit{is} can differ between authors and studies but is generally used in the common (i.e.
-non-mathematical) sense of a space containing all possible morphologies.
+The concept of what the morphospace \textit{is} can differ among authors and studies but is generally used in the common (i.e. non-mathematical) sense of a space containing all possible morphologies.
 For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points.
-In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combination of traits.
+In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combinations of traits.
 We generalise this definition to any linear transformation ($V \to W$) of this space that retains the biological definition, i.e.
@@
 If the morphospace is transformed mathematically (i.e.
-the dataset is transformed using logs, ordinations, distance matrices, etc...) it remains the morphospace as long as all the observation and traits remain in the datatset.
+the dataset is transformed using logarithms, ordinations, distance matrices, etc.) it remains the morphospace provided all the observation and traits remain in the datatset.
 

### 2020-02-27 Writing/disparity_review.tex
Edits to the end of 3

@@
 
-One of the reason why ordinations techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at the time which can be more intuitively captured than through disparity indices (see section \ref{section:metrics} below).
+One of the reasons why ordination techniques are common in disparity analysis is that they make it easier for researchers to comprehend patterns in two or three spatial dimensions at a time, which can be more intuitive than through disparity indices (see section \ref{section:metrics} below).
 Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e.
@@
 In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012-va,dryden2016statistical}.
-In practice, this can allow to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g.
+In practice, this allows us to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g.
 the elongation and flattening of birds beaks;][]{Cooney2017-ly}.
 
-However, like most other aspects of disparity analyses, reducing dimensionality can be tricky.
+Like most other aspects of disparity analyses, however, reducing dimensionality can be tricky.
 In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results. 
-Although a common technique is to consider the \textit{n} axis that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to miss-interpretation of the biological variation mapped on these axes \citep{Bookstein2015-yy, Bookstein2017-qk, Bookstein2017-gu,Weisbecker2019-kp}.
-Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018-le, Gerber2017-xi}.
+Although a common technique is to consider the \textit{n} axes that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015-yy, Bookstein2017-qk, Bookstein2017-gu,Weisbecker2019-kp}.
+Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018-le, Gerber2017-xi}.
 
@@
 the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018-ye}), some are not \citep[e.g.][]{Weisbecker2019-kp}.
-For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and the ones that have more in state 0, has no biological meaning if these are binary alternate states.
+For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and those that have more in state 0, has no biological meaning if these are binary alternate states.
 
-Categorical data are a good deal more problematic, since the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean and the distribution of the variance is usually normal (i.e.
+Categorical data are a good deal more problematic, because the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean, and the distribution of the variance is usually more evenly distributed across axes (i.e.
 contrary to a PCA, the first few axis do not encompass most variance of the dataset).
-Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e.
+Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e.
 the distance between A and B is not equal to the distance between B and A][]{Gerber2014-ol}.
-And last but not least reason, In many cases, ordination might not be necessary.
-For example, if an index (e.g. sum or ranges of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015-qi}.
-For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.
+Last, but not least, in many cases, ordination might not be necessary.
+For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015-qi}.
+For all of these reasons, multidimensional data should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.
 
@@
     Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g.
-using non-metric multidimensional scaling forcing the dataset into two dimensions - note that variance is lost due to the distortion stress and non-metric properties of the resulting operation) ; or ordination axis plots (directly from the ordinated matrix).
-    Note that in 2D representations, we consider it as good practice to plot both axis with the same scale (so to not visually distort the importance of one axis).
+using non-metric multidimensional scaling forcing the dataset into two dimensions - note that variance is lost due to the distortion stress and non-metric properties of the resulting operation); or ordination axis plots (directly from the ordinated matrix).
+    Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).
     }
@@
 \citealt{Diaz2016-mr}, distance from the centroid \citealt{Hopkins2017-cf, Finlay2015-ft}, variance and range \citealt{Brusatte2008-fx}).
-Furthermore, different aspects than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g.
+Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g.
 does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
@@
 
-In addition to considering what properties of disparity indices should capture, it is important to also consider the mathematical properties of the indices and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
-For example, measuring the sum of variance-covariance for each dimension of the space is equivalent of measuring the sum of variance in a PCA (since the co-variance is null).
+In addition to considering what properties of disparity these indices should capture, it is also important to consider the mathematical properties of the indices and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
+For example, measuring the sum of variance-covariance for each dimension of the space is equivalent to measuring the sum of variance in a PCA (because the co-variance is null).
 However, this is not true of other transformations of the space or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012-va}.
 
-Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \citep{Bellman1966-mc}.
-In spaces with some axis of variance lower than 1, product-based indices used as proxies of volumes (e.g.
+Furthermore, multidimensional spaces have some counterintuitive properties that should be considered such as the ``curse of dimensionality'' \citep{Bellman1966-mc}.
+In spaces with some axes of variance lower than 1, product-based indices used as proxies of volumes (e.g.
 product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966-mc}.
-Some other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992-zs} or convex hull based indices \citep{Jackson2011-kq}.
+Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992-zs} or convex hull based indices \citep{Jackson2011-kq}.
 
-\subsection{Testing hypotheses on the evolution of disparity} \label{section:testing}
+\subsection{Testing hypotheses about the evolution of disparity} \label{section:testing}
 
 No matter which disparity indices have been calculated, the research question must be framed in an appropriate statistical context.
-The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018-mg}, but some of these advances have yet to be implemented in disparity analyses.
+The multidimensional statistical toolkit for ecology and evolution has greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018-mg}, but some of these advances have yet to be implemented in disparity analyses.
 Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
@@
 As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998-xg}.
-Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common) the non-independence between observations should be taken into account to correctly measure phenomenons.
-However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate, especially when using only the first $n$ axis of the ordination and can lead to artefactual interpretations of the data \citep[such as wrongly supported ``early burst'' type patterns;][]{Uyeda2015}.
+Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common) the non-independence between observations should be taken into account.
+It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first $n$ axis of the ordination, and can lead to incorrect interpretations of the data \citep[such as wrongly supporting ``early burst'' type patterns;][]{Uyeda2015}.
 
-One other common way to taken phylogeny into account in disparity analyses is using ancestral sates estimations in disparity-through-time analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
+One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in disparity-through-time analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
 Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e.
@@
 
-No ancestral state estimation method is without drawbacks and above all else are highly dependent on the data and method used \cite{Louca2019}.
+All ancestral state estimates are highly dependent on the data and method used \cite{Louca2019}.
 In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points to increase statistical power.
-In fact, these extra points are not independent and can also have problematic side effects, especially when testing for the influence of mass extinctions on disparity.
+In fact, these extra points are not independent and can also have problematic side effects, especially when testing for the influence of mass extinctions on disparity as they artificially increase taxon sampling.
 
@@
 a linear transformation of $(a_{ij})$).
-However if one only uses the two first axis of $M3$, this would result in a different space since it is no more a linear transformation of $(a_{ij})$.
+However if one only uses the two first axis of $M3$, this would result in a different space because it is no more a linear transformation of $(a_{ij})$.
 

### 2020-02-27 Writing/disparity_review.tex
Edits to the end

@@
 
-\noindent Morphological disparity analyses are widely employed in evolutionary palaeobiology, and they are based on a diversity of methods and data.
+\noindent Morphological disparity analyses are widely employed in evolutionary palaeobiology, and are based on a diversity of methods and data.
 There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
-As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse it, stemming from the explicit hypotheses being tested.
+As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse them, stemming from the explicit hypotheses being tested.
 Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
-We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
-Thankfully this is now getting easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc}.
+We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of indices, ordination method and ancestral state estimation method on disparity analysis results.
+Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc}.
 Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013-bg, Saupe2015-vm, Canter2018-hk, mammola2019}.
@@
 
-Although disparity analyses are now simple to implement in freely available softwares \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc} it is crucial to remember that they are multidimensional analyses and multidimensional analyses are complex.
-We assert that future morphological analyses will benefit by emphasising the methodological decisions made, rather than simply using disparity analysis because \textit{we can}.
+Although disparity analyses are now simple to implement in freely available software \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc} it is crucial to remember that they are multidimensional analyses and that multidimensional analyses are complex.
+We assert that future morphological analyses will benefit from emphasising the methodological decisions made, rather than simply using disparity analysis because it exists.
 
@@
 SB was funded by European Research Council (ERC) under the European Union's Horizon 2020 research and innovation program (grant agreement No 756226, ERC Starting Grant: PalM) and a Leverhulme Trust Research Project Grant (RPG-2017-167);
-TG was funded by ARC DP170103227 and FT180100634 awarded to VW;
+TG was funded by ARC DP170103227 and FT180100634 awarded to VW.
 

### 2020-02-27 Writing/disparity_review.tex
Updated Natalie's comments

@@
 rodents or nematodes), species-poor but phenotypically diverse (e.g.
-Afrotheria), % NC: Can we expand Afrotheria a bit?
-All the other examples are using common names.
-Can we use common names of afrotheres?
-Or at least say afrotherian mammals?
-or depauperate in both species and phenotypic diversity (e.g. lungfish).
+afrotherian mammals), or depauperate in both species and phenotypic diversity (e.g.
+lungfish).
 These phenomena suggest that taxonomic and phenotypic diversity are not necessarily linked, raising important questions about how phenotypic diversity arises, such as:
@@
 Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve?
-(Mike Foote 1997, 1995) % NC: Seems weird only having these two refs here and no refs in the other questions.
-Remove?
 What role does ecology play in structuring morphological diversity?
@@
     These categories are not independent and many disparity studies will cover more than one.
-    %@@@ Add disparity and evo-devo
-    % NC: I removed the mention of constraints because I hate the idea it's all constraints - there are other things at play why are we only mentioning constraints?
-(I suspect Anjali or Vera put that in?) Also it didn't fit with the other examples which don't mention mechanisms...
 }

### 2020-03-10 Writing/disparity_review.tex
Updated reference gender balance

@@
 Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are comparatively more recent. 
-Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} the concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
+Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991} the concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
 However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
@@
 
-	\item \textbf{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995-do, Briggs1992-pd}.
+	\item \textbf{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995, Briggs1992}.
 	These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace?
-\citep{Foote1995-do, Raup1961-vx, Gerber2017-xi}.
+\citep{Foote1995, Raup1961, Gerber2017}.
     More recently, this approach as also been used to understand the relationship between developmental processes and morphology in the evo-devo field.
@@
 	\item \textbf{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
-	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{Guillerme2018-uj, Wright2017-jo}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010-ve}?
+	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{prentice2011evolution, Guillerme2018}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010}?
 
@@
 	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two
-	\citep[e.g.][]{Fortey1996-kt, Moyne2007-jm, Ruta2013-iy, Hopkins2013-xt}.
-	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011-fq}.
+	\citep[e.g.][]{Fortey1996, Ruta2013, Hopkins2013}.
+	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011}.
 
@@
 	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles
-	\citep{Friedman2010-ve, Pierce2008-yr, Anderson2013-zt}.
+	\citep{Pierce2008, Friedman2010}.
 	The links between form and function, however, are not always clear cut.
-	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005-or}.
-	This approach has been used to investigate hypotheses of competitive replacement \citep{Brusatte2008-vx} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010-ve}.
-	It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \citep{Wainwright2005-or}.
+	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005}.
+	This approach has been used to investigate hypotheses of competitive replacement \citep{tyler2011detecting} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010}.
+	It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \citep{Wainwright2005}.
 
@@
 Fundamental insights into evolutionary biology have been elicited from these four types of disparity analysis.
-One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997-nl, Erwin2007-jz, Hughes2013-td}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007-jz, Hughes2013-td}.
+One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997, Erwin2007, Hughes2013}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007, Hughes2013}.
 However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?
@@
 \noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
-	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g.][]{Foote1989-fd, Deline2018-le};
-	2) continuous measurements of features (e.g. lengths) \citep[e.g][]{Anderson2001-qb}; or
+	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[e.g.][]{Foote1989, Deline2018};
+	2) continuous measurements of features (e.g. lengths) \citep[e.g][]{Friedman2010}; or
 	3) more mathematical descriptors from geometric morphometric landmark data (e.g.
-Procrustes coordinates) \citep[e.g.][]{Cooney2017-ly}, Fourier coefficients \citep[e.g.][]{Foote1995-do, Spriggs2018-nu} or model-based descriptors \citep[e.g.][]{Raup1961-vx,Saunders2004} (Fig.
+Procrustes coordinates) \citep[e.g.][]{Cooney2017}, Fourier coefficients \citep[e.g.][]{Foote1995, Spriggs2018} or model-based descriptors \citep[e.g.][]{Raup1961,Saunders2004} (Fig.
 2).
-None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.
+None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017}.
 
@@
 The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
-Discrete character are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008-fx,Close2015-qi,Wright2017-jo}.
+Discrete character are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008,Close2015}.
 This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups.
 This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
-Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017-cf}.
-This effect of the anatomical feature used on disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018-le}.
+Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017}.
+This effect of the anatomical feature used on disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018}.
 
-Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \citep{Brazeau2017-kg, Palci2018-ni}.
+Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \citep{Palci2018}.
 Biological phenomena such as allometry and sexual dimorphism may also influence trait data.
@@
 They come in many flavours depending on the data and the desired morphospace properties.
-For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) \citep[see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012-va}.
+For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) \citep[see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012}.
 
@@
 95\%).
-In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012-va,dryden2016statistical}.
+In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012,dryden2016statistical}.
 In practice, this allows us to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g.
-the elongation and flattening of birds beaks;][]{Cooney2017-ly}.
+the elongation and flattening of birds beaks;][]{Cooney2017}.
 
@@
 In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results. 
-Although a common technique is to consider the \textit{n} axes that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015-yy, Bookstein2017-qk, Bookstein2017-gu,Weisbecker2019-kp}.
-Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018-le, Gerber2017-xi}.
+Although a common technique is to consider the \textit{n} axes that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
+Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018, Gerber2017}.
 
@@
 Although some gradients are easy to detect or interpret (e.g.
-the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018-ye}), some are not \citep[e.g.][]{Weisbecker2019-kp}.
+the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018}), some are not \citep[e.g.][]{Weisbecker2019}.
 For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and those that have more in state 0, has no biological meaning if these are binary alternate states.
@@
 Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e.
-the distance between A and B is not equal to the distance between B and A][]{Gerber2014-ol}.
+the distance between A and B is not equal to the distance between B and A][]{Gerber2014}.
 Last, but not least, in many cases, ordination might not be necessary.
-For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015-qi}.
+For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015}.
 For all of these reasons, multidimensional data should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.
@@
     Morphospaces: different mathematical representations of a morphospace.
-    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015-qi} or an ordinated matrix \citep[e.g. in][]{Brusatte2008-vx}.
+    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015} or an ordinated matrix \citep[e.g. in][]{tyler2011detecting}.
     Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
@@
 Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
-This is usually achieved with a disparity measure or index \citep{Hopkins2017-cf}.
+This is usually achieved with a disparity measure or index \citep{Hopkins2017}.
 As with any summary of multidimensional data, disparity indices will reflect only some aspects of the morphological variation, never its whole complexity \citep{GuillermeMOMS}.
@@
 volume
-\citealt{Diaz2016-mr}, distance from the centroid \citealt{Hopkins2017-cf, Finlay2015-ft}, variance and range \citealt{Brusatte2008-fx}).
+\citealt{Diaz2016}, distance from the centroid \citealt{Hopkins2017, Finlay2015-ft}, variance and range \citealt{Brusatte2008}).
 Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g.
@@
 nearest neighbour distance, pairwise distances, etc.
-\citealt{Close2015-qi} - see section \ref{section:ordination} above).
+\citealt{Close2015} - see section \ref{section:ordination} above).
 
-In addition to considering what properties of disparity these indices should capture, it is also important to consider the mathematical properties of the indices and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
+In addition to considering what properties of disparity these indices should capture, it is also important to consider the mathematical properties of the indices and their associated caveats \citep{Wills2001, Ciampaglio2001}.
 For example, measuring the sum of variance-covariance for each dimension of the space is equivalent to measuring the sum of variance in a PCA (because the co-variance is null).
-However, this is not true of other transformations of the space or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012-va}.
+However, this is not true of other transformations of the space or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012}.
 
-Furthermore, multidimensional spaces have some counterintuitive properties that should be considered such as the ``curse of dimensionality'' \citep{Bellman1966-mc}.
+Furthermore, multidimensional spaces have some counterintuitive properties that should be considered such as the ``curse of dimensionality'' \citep{Bellman1966}.
 In spaces with some axes of variance lower than 1, product-based indices used as proxies of volumes (e.g.
-product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966-mc}.
-Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992-zs} or convex hull based indices \citep{Jackson2011-kq}.
+product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966}.
+Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992} or convex hull based indices \citep{Jackson2011}.
 
@@
 No matter which disparity indices have been calculated, the research question must be framed in an appropriate statistical context.
-The multidimensional statistical toolkit for ecology and evolution has greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018-mg}, but some of these advances have yet to be implemented in disparity analyses.
+The multidimensional statistical toolkit for ecology and evolution has greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018}, but some of these advances have yet to be implemented in disparity analyses.
 Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
-One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001-qb, Anderson2013-zt}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
+One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
 Although statistically valid, this test is not always directly related to the hypothesis under question.
@@
 
-As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998-xg}.
+As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998}.
 Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common) the non-independence between observations should be taken into account.
@@
 We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of indices, ordination method and ancestral state estimation method on disparity analysis results.
-Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc}.
-Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013-bg, Saupe2015-vm, Canter2018-hk, mammola2019}.
+Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Navarro2003, Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b}.
+Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013, Saupe2015, Canter2018, mammola2019}.
 Innovations in morphological disparity analyses likely await discovery in their respective literatures.
@@
 While studies of morphological disparity would benefit from advances in multidimensional analysis in other fields, the concept of a morphospace could reciprocally benefit other disciplines.
-For example, the multidimensional analysis of \citep{Diaz2016-mr}, which analysed patterns of form and function in plants, is essentially an eco-morphospace; isotopic analyses of organisms \citep{Jackson2011-kq,Swanson2015} can be represented as an isotope-space; ecosystem functioning in \citealt{Donohue2013-bg} as an ecosystem-space, etc.
-These generalisations could also be exported for any set of traits: cognate approaches have been adopted in the analysis of single cell comparative transcriptome data \citep{Sebe-Pedros2018-sw} where interpretation of the resulting transcriptome-spaces would be improved by giving careful attention to the concerns we highlight concerning morphospaces.
+For example, the multidimensional analysis of \citep{Diaz2016}, which analysed patterns of form and function in plants, is essentially an eco-morphospace; isotopic analyses of organisms \citep{Jackson2011,Swanson2015} can be represented as an isotope-space; ecosystem functioning in \citealt{Donohue2013} as an ecosystem-space, etc.
+These generalisations could also be exported for any set of traits: cognate approaches have been adopted in the analysis of single cell comparative transcriptome data \citep{Sebe-Pedros2018} where interpretation of the resulting transcriptome-spaces would be improved by giving careful attention to the concerns we highlight concerning morphospaces.
 
-Although disparity analyses are now simple to implement in freely available software \citep{Navarro2003-vz, Bouxin2005-wk, oksanen2007vegan, Harmon2008-gq, lloyd2016, Guillerme2018-uc} it is crucial to remember that they are multidimensional analyses and that multidimensional analyses are complex.
+Although disparity analyses are now simple to implement in freely available software \citep{Navarro2003, Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b} it is crucial to remember that they are multidimensional analyses and that multidimensional analyses are complex.
 We assert that future morphological analyses will benefit from emphasising the methodological decisions made, rather than simply using disparity analysis because it exists.

### 2020-03-11 Writing/disparity_review.tex
ref fix
Updated references and affiliations

@@
 
-\bigskip
-\noindent RH: disparate disparity analyses
-\bigskip
 \medskip
@@
 Kevin Healy$^{7}$,
-Melanie Hopkins$^{8}$,
-Marc EH Jones$^{9}$,
+Melanie J. Hopkins$^{8}$,
+Marc E. H. Jones$^{9}$,
 Graeme T. Lloyd$^{10}$,
@@
 Abi Pate$^{11}$,
-Mark N Puttick$^{12}$,
+Mark N. Puttick$^{12}$,
 Emiliy Rayfield$^{11}$,
@@
 Emma Sherratt$^{14}$,
-Graham Slater$^{15}$,
+Graham J. Slater$^{15}$,
 Vera Weisbecker$^{1}$,
-Gavin H Thomas$^{16}$
-and Philip Donoghue$^{11}$\\
+Gavin H. Thomas$^{16}$
+and Philip C. J. Donoghue$^{11}$\\
 
@@
 $^{7}$Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
-$^{8}$ @@@;
+$^{8}$ Division of Paleontology, American Museum of Natural History, New York, USA;
 $^{10}$School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
@@
 $^{14}$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
-$^{15}$ @@@;
-$^{16}$ @@@;\\
+$^{15}$ Department of the Geophysical Sciences, University of Chicago, Chicago, Illinois 60637;
+$^{16}$ Department of Animal and Plant Sciences, The University of Sheffield, Sheffield S10 2TN, UK;\\
 $^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding author: guillert@tcd.ie\\}
@@
 	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two
-	\citep[e.g.][]{Fortey1996, Ruta2013, Hopkins2013}.
-	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011}.
+	\citep[e.g.][]{Fortey1996, Hopkins2013}.
+	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{pierce2008patterns}.
 
@@
 	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[e.g.][]{Foote1989, Deline2018};
-	2) continuous measurements of features (e.g. lengths) \citep[e.g][]{Friedman2010}; or
+	2) continuous measurements of features \citep[e.g. lengths in][]{Friedman2010}; or
 	3) more mathematical descriptors from geometric morphometric landmark data (e.g.
-Procrustes coordinates) \citep[e.g.][]{Cooney2017}, Fourier coefficients \citep[e.g.][]{Foote1995, Spriggs2018} or model-based descriptors \citep[e.g.][]{Raup1961,Saunders2004} (Fig.
+Procrustes coordinates) \citep[e.g.][]{sherratt2017rates}, Fourier coefficients \citep[e.g.][]{Foote1995, Spriggs2018} or model-based descriptors \citep[e.g.][]{Raup1961,Saunders2004} (Fig.
 2).
@@
 They come in many flavours depending on the data and the desired morphospace properties.
-For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) \citep[see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012}.
+For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling \citep[NMDS; see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012}.
 
@@
 In practice, this allows us to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g.
-the elongation and flattening of birds beaks;][]{Cooney2017}.
+the elongation and flattening of birds beaks;][]{bright2016shapes}.
 
@@
 Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e.
-the distance between A and B is not equal to the distance between B and A][]{Gerber2014}.
+the distance between A and B is not equal to the distance between B and A;][]{Gerber2014}.
 Last, but not least, in many cases, ordination might not be necessary.
-For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep{Close2015}.
+For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep[e.g.][]{Close2015}.
 For all of these reasons, multidimensional data should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.
@@
 does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g.
-volume
-\citealt{Diaz2016}, distance from the centroid \citealt{Hopkins2017, Finlay2015-ft}, variance and range \citealt{Brusatte2008}).
+volume;
+\citealt{Diaz2016}, distance from the centroid; \citealt{Hopkins2017, Finlay2015}, variance and range; \citealt{Brusatte2008}).
 Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g.
@@
 product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966}.
-Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Foote1992} or convex hull based indices \citep{Jackson2011}.
+Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Wills2001} or convex hull based indices \citep{Jackson2011}.
 
@@
 ordination, or distance matrix construction) and is simply based on the original data; or (2) post-transformation, i.e.
-the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g.
-the ordination scores) \citep{lloyd2018}.
+the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix \citep[e.g.
+the ordination scores;][]{lloyd2018}.
 
@@
 We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of indices, ordination method and ancestral state estimation method on disparity analysis results.
-Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Navarro2003, Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b}.
+Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b}.
 Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013, Saupe2015, Canter2018, mammola2019}.
@@
 
-Although disparity analyses are now simple to implement in freely available software \citep{Navarro2003, Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b} it is crucial to remember that they are multidimensional analyses and that multidimensional analyses are complex.
+Although disparity analyses are now simple to implement in freely available software \citep{Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b} it is crucial to remember that they are multidimensional analyses and that multidimensional analyses are complex.
 We assert that future morphological analyses will benefit from emphasising the methodological decisions made, rather than simply using disparity analysis because it exists.

### 2020-03-12 Writing/disparity_review.tex
Updated refs + compiler

@@
 
-\noindent {\small \it 
-$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
-$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
-$^{3}$School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
-$^{4}$Department of Biology, University of York, YO10 5DD, UK;
-$^{5}$Department of Zoology, School of Natural Sciences, Trinity College, Ireland;
-$^{6}$Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
-$^{7}$Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
+\noindent { \it 
+$^1$ School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
+$^2$ Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
+$^{3}$ School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
+$^{4}$ Department of Biology, University of York, YO10 5DD, UK;
+$^{5}$ Department of Zoology, School of Natural Sciences, Trinity College, Ireland;
+$^{6}$ Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
+$^{7}$ Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
 $^{8}$ Division of Paleontology, American Museum of Natural History, New York, USA;
-$^{10}$School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
-$^{11}$School of Earth Sciences, University of Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
-$^{12}$Milner Centre for Evolution, University of Bath, BA2 7AY UK;
-$^{13}$Department of Earth Sciences, University of Oxford, S Parks Road, Oxford OX1 3AN, UK;
-$^{14}$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
+$^{9}$ Department of Cell and Developmental Biology, University College London, London, UK;
+$^{10}$ School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
+$^{11}$ School of Earth Sciences, University of Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
+$^{12}$ Milner Centre for Evolution, University of Bath, BA2 7AY UK;
+$^{13}$ Department of Earth Sciences, University of Oxford, S Parks Road, Oxford OX1 3AN, UK;
+$^{14}$ School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
 $^{15}$ Department of the Geophysical Sciences, University of Chicago, Chicago, Illinois 60637;
@@
     }
-    % @@@ Change references
 }

### 2020-03-12 Writing/pandoc.sh

@@ -1,0 +1,1 @@
+pandoc -o disparity_review.docx -i disparity_review.tex --bibliography References.bib --filter pandoc-citeproc --csl biologyletters.csl

### 2020-03-13 Writing/CoverLetter.tex
cover letter  INIT

\documentclass[11pt]{letter}
\usepackage[a4paper,left=2.5cm, right=2.5cm, top=1cm, bottom=1cm]{geometry}
\usepackage{hyperref}
\usepackage[osf]{mathpazo}
\signature{Thomas Guillerme, on behalf of my co-authors.}
\address{The University of Queensland \\School of Biological Sciences \\St Lucia QLD 4067, Australia \\guillert@tcd.ie}
\longindentation=0pt
\begin{document}

\begin{letter}{}
\opening{Dear Editors,}

Royal Society International Scientific Seminar: Reconciling disparate perspectives on the evolution of disparity on the 9th and 10th of January 2018 in Chicheley Hall.

Along with all 19 the meeting participants and 3 additional researchers, we wrote a review entitled: ``Disparities in the analysis of morphological disparity''.

This paper covers:
- The origin of disparity in litterature
- The types of analysis
- The types of data
- The types of statistical analysis (including phylogeny)
- The future of disparity methods

This paper especially aims to provide a common framework to disparity analysis that can be used as a base for macroevolution researchers to make choices in their analysis and go further.

It also aims to foster interdisciplinarity and disparity beyond morphology (e.g. in ecology or isotopes analysis).

Overall, this broad review proposes the state-of-the-art in disparity analyses and where the field has been going since its advent in the 1980s.

We look forward to hearing from you soon,

\closing{Yours sincerely,}

\end{letter}
\end{document}

%Editors:
%Sam Price, Daniele Silvestro

%Reviewers
%Graeme Lloyd, Dave Bapst, Emma Sherratt

### 2020-03-29 Writing/disparity_review.tex
Updated title and into

--- 2020-03-11 Writing/disparity_review.tex
@@
 
-Thomas Guillerme$^{1,16,*,+}$,
-Natalie Cooper$^{2,+}$,
+Thomas Guillerme$^{1,17,*, +}$,
+Natalie Cooper$^{2, +}$,
 Stephen L. Brusatte$^{3}$,
@@
 Graeme T. Lloyd$^{10}$,
-Joseph E. O'Reilly$^{11}$,
-Abi Pate$^{11}$,
-Mark N. Puttick$^{12}$,
-Emiliy Rayfield$^{11}$,
-Erin E. Saupe$^{13}$,
-Emma Sherratt$^{14}$,
-Graham J. Slater$^{15}$,
+Joseph E. O'Reilly$^{11}$, %was 17 - 11 forward + 1
+Abi Pate$^{12}$,
+Mark N. Puttick$^{13}$,
+Emily Rayfield$^{12}$,
+Erin E. Saupe$^{14}$,
+Emma Sherratt$^{15}$,
+Graham J. Slater$^{16}$,
 Vera Weisbecker$^{1}$,
-Gavin H. Thomas$^{16}$
-and Philip C. J. Donoghue$^{11}$\\
+Gavin H. Thomas$^{17}$
+and Philip C. J. Donoghue$^{12}$
 
-\noindent {\small \it 
-$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
-$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
-$^{3}$School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
-$^{4}$Department of Biology, University of York, YO10 5DD, UK;
-$^{5}$Department of Zoology, School of Natural Sciences, Trinity College, Ireland;
-$^{6}$Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
-$^{7}$Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
-$^{8}$ Division of Paleontology, American Museum of Natural History, New York, USA;
-$^{10}$School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
-$^{11}$School of Earth Sciences, University of Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
-$^{12}$Milner Centre for Evolution, University of Bath, BA2 7AY UK;
-$^{13}$Department of Earth Sciences, University of Oxford, S Parks Road, Oxford OX1 3AN, UK;
-$^{14}$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
-$^{15}$ Department of the Geophysical Sciences, University of Chicago, Chicago, Illinois 60637;
-$^{16}$ Department of Animal and Plant Sciences, The University of Sheffield, Sheffield S10 2TN, UK;\\
-$^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding author: guillert@tcd.ie\\}
+\noindent { \it 
+$^{1}$ School of Biological Sciences, University of Queensland, St. Lucia,
+Queensland, Australia;
+$^{2}$ Department of Life Sciences, Natural History Museum, London, Cromwell
+Road, London, SW7 5BD, UK;
+$^{3}$ School of GeoSciences, University of Edinburgh, Grant Institute,
+Edinburgh EH9 3FE, UK;
+$^{4}$ Leverhulme Centre for Anthropocene Biodiversity, Department of
+Biology, University of York, YO10 5DD, UK;
+$^{5}$ Department of Zoology, School of Natural Sciences, Trinity College
+Dublin, Ireland;
+$^{6}$ Institut de Systématique, Évolution, Biodiversité (ISYEB), Muséum
+national d'Histoire naturelle, CNRS, Sorbonne Université, EPHE,
+Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
+$^{7}$ Ryan Institute, School of Natural Sciences, National University of
+Ireland Galway, Ireland;
+$^{8}$ Division of Paleontology (Invertebrates), American Museum of Natural
+History, New York, USA;
+$^{9}$ Department of Cell and Developmental Biology, University College
+London, London, UK;
+$^{10}$ School of Earth and Environment, University of Leeds, Leeds LS2 9JT,
+UK;
+$^{11}$ MRC Institute of Genetic and Molecular Medicine, University of Edinburgh, UK;
+$^{12}$ School of Earth Sciences, University of Bristol, Life Sciences
+Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
+$^{13}$ Milner Centre for Evolution, University of Bath, BA2 7AY UK;
+$^{14}$ Department of Earth Sciences, University of Oxford, S Parks Road,
+Oxford OX1 3AN, UK;
+$^{15}$ School of Biological Sciences, The University of Adelaide, Adelaide,
+South Australia 5005, Australia.
+$^{16}$ Department of the Geophysical Sciences, University of Chicago,
+Chicago, Illinois 60637;
+$^{17}$ Department of Animal and Plant Sciences, The University of Sheffield,
+Sheffield S10 2TN, UK;\\
+$^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding
+author: guillert@tcd.ie\\}
 
@@
 \noindent Analyses of morphological disparity have been used to characterise and investigate the evolution of variation in the anatomy, function, and ecology of organisms since the 1980s.
-While a diversity of methods have been employed, it is unclear whether they provide
-equivalent insights.
+While a diversity of methods have been employed, it is unclear whether they provide equivalent insights.
 Here we review the most commonly used approaches for characterising and analysing morphological disparity, all of which have associated limitations that, if ignored, can lead to misinterpretation.
@@
 
-%@@@ add links to figures
-
 \noindent Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
-At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g.
-cichlids or molluscs), species-rich but phenotypically conservative (e.g.
-rodents or nematodes), species-poor but phenotypically diverse (e.g.
-afrotherian mammals), or depauperate in both species and phenotypic diversity (e.g.
+At the extremes, clades may be exceptionally rich in species and phenotypic diversity (hereafter \textit{disparity}) (e.g.
+cichlids or molluscs), species-rich but disparity-poor (e.g.
+rodents or nematodes), species-poor but rich in disparity (e.g.
+afrotherian mammals), or depauperate in both species diversity and disparity (e.g.
 lungfish).
-These phenomena suggest that taxonomic and phenotypic diversity are not necessarily linked, raising important questions about how phenotypic diversity arises, such as:
-How does diversity evolve?
+These phenomena suggest that taxonomic diversity and phenotypic disparity are not inextricably linked, raising important questions, such as: How does disparity evolve?
 Are some morphologies more common than others?
-Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve?
-What role does ecology play in structuring morphological diversity?
-Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are comparatively more recent. 
-Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991} the concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
-However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
+Is anatomical evolution unbounded or are some anatomies impossible to achieve?
+What role does ecology play in structuring disparity?
+Analyses of species diversity have a venerable history, but those of disparity are comparatively more recent.
+Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991}, the concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
+However, disparity analyses have since expanded into comparative biology as a means of capturing how intrinsic and extrinsic causal agents affect morphological evolution.
 Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (e.g.
 taxa) can be placed in this space based on their trait values.
-Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):
+Such multidimensional spaces (or morphospaces - defined broadly hereafter as a mathematical space relating morphological configurations generally based on some measure of similarity \citep{mitteroecker2009concept}) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig.
+1): %remove fig1?
 
@@
 
-	\item \textbf{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995, Briggs1992}.
-	These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace?
+	\item \textbf{Descriptive disparity.} Pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995, Briggs1992}.
+	These studies described multidimensional patterns in morphological trait diversity by addressing pertinent questions: why are some morphological trait combinations more common than others, and what are the biological (or mathematical) properties of the resulting morphospace?
 \citep{Foote1995, Raup1961, Gerber2017}.
-    More recently, this approach as also been used to understand the relationship between developmental processes and morphology in the evo-devo field.
-    For example, describing disparity patterns has been successfully used to compare modules of evolution in various groups \citep{goswami2010influence,bardua2019morphological} allowing researchers to successfully link variation in shape to a group's evolutionary or developmental constraints \citep{Hipsley2017}.
+	More recently, this approach has been used to understand the relationship between developmental processes and morphology in the field of evolutionary development (evo-devo).
+	For example, patterns of disparity have been used successfully to compare modules of evolution in various groups \citep{goswami2010influence,bardua2019morphological}, allowing researchers to link variation in shape to a group's evolutionary or developmental constraints \citep{Hipsley2017}.
 
-	\item \textbf{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
-	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{prentice2011evolution, Guillerme2018}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010}?
+	\item \textbf{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed over time, by focussing on the disparity of taxa in particular time intervals or slices.
+	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{prentice2011evolution, Guillerme2018}, or how does disparity change up to and across mass extinction events\citep{Friedman2010}?
 
 	\item \textbf{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e.
-shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
+	shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity \citep[but see][]{anderson2012}.
 	This makes disparity an informative complement to diversity measures based on species richness alone.
-	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two
-	\citep[e.g.][]{Fortey1996, Hopkins2013}.
-	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{pierce2008patterns}.
+	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two \citep[e.g.][]{Fortey1996, Hopkins2013}..
+	The approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{pierce2008patterns}.
 
 	\item \textbf{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
-	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles
-	\citep{Pierce2008, Friedman2010}.
-	The links between form and function, however, are not always clear cut.
+	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles \citep{Pierce2008, Friedman2010}.
+	The links between form and function, however, are not always clear.
 	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005}.
 	This approach has been used to investigate hypotheses of competitive replacement \citep{tyler2011detecting} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010}.
-	It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \citep{Wainwright2005}.
-
-	% \item \textbf{Disparity and evo-devo}
- %    %TG: This section might be linked to "Descriptive disparity" above but Anjali might probably disagree.
- %    The disparity of a group can also give information on a group's evolutionary or developmental constraints.
- %    test hypotheses of developmental constraints,
- %    disparity and integration/modularity
+	It is particularly common in palaeobiology where it is not possible to directly observe many ecological or functional characteristics of extinct species \citep{Wainwright2005}.
 
@@
 Fundamental insights into evolutionary biology have been elicited from these four types of disparity analysis.
-One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997, Erwin2007, Hughes2013}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007, Hughes2013}.
-However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?
+One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997, Erwin2007, Hughes2013}, indicating that capacity for evolutionary innovation wanes as clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change \citep{Erwin2007, Hughes2013}.
+However, this example also highlights one of the greatest challenges confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using a diversity of methods, approaches and data types be considered equivalent?
 
@@
 We first discuss the appropriate data required for characterising disparity, then review various challenging aspects of these approaches.
-Throughout, it is important to remember these tools should always be used in the context of a specific scientific question, as this will drive data and methodological choices at every stage of the process.
+Throughout, it is important to remember that these tools should always be used in the context of a specific scientific question, as this will drive data and methodological choices at every stage of the process.
 
-\begin{figure}[!htbp]
-\centering
-   \includegraphics[width=0.9\textwidth]{Figures/figure_disparities.pdf}
-\caption{
-    The four main types of disparity analysis.
+% \begin{figure}[!htbp]
+% \centering
+%    \includegraphics[width=0.9\textwidth]{Figures/figure_disparities.pdf}
+% \caption{
+%     The four main types of disparity analysis.
 \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
-    These categories are not independent and many disparity studies will cover more than one.
-}
-\label{Fig:disparity}
-\end{figure}
+%     These categories are not independent and many disparity studies will cover more than one.
+% }
+% \label{Fig:disparity}
+% \end{figure}
 

### 2020-03-29 Writing/disparity_review.tex
Updated up to "to ordinate"

@@
 1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[e.g.][]{Foote1989, Deline2018};
-	2) continuous measurements of features \citep[e.g. lengths in][]{Friedman2010}; or
-	3) more mathematical descriptors from geometric morphometric landmark data (e.g.
-Procrustes coordinates) \citep[e.g.][]{sherratt2017rates}, Fourier coefficients \citep[e.g.][]{Foote1995, Spriggs2018} or model-based descriptors \citep[e.g.][]{Raup1961,Saunders2004} (Fig.
-2).
-None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017}.
+2) continuous measurements of features \citep[e.g. lengths in][]{Friedman2010};
+or 3) more mathematical descriptors from geometric morphometric landmark data (e.g.
+Procrustes coordinates)  \citep[e.g.][]{sherratt2017rates}, Fourier coefficients \citep[e.g.][]{Foote1995, Spriggs2018} or model-based descriptors \citep[e.g.][]{Raup1961,Saunders2004} (Fig. 2).
+None of these approaches is superior, but they may be more or less well-suited to characterising the traits compared and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017}.
 
 For example, if investigating variation of bat wing shapes, both homologous landmarks and continuous measurements of bones may be appropriate to capture patterns of wing variation.
-If the question focuses on comparing wings between bats and birds, however, different measurements might be more appropriate depending on the specific question: if the focus is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
+If the question focuses on comparing wings between bats and birds, however, different measurements might be more appropriate depending on the specific question.
+That is, if the focus is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g.
 wingspan, aspect ratio, etc.).
@@
 digit length, integumentary system, etc.).
-Where there is any doubt about the appropriate traits to choose, it may be preferable to use several different kinds of data for the same feature to determine whether they capture the same pattern of disparity.
+Where there is any doubt about which traits to analyse, it may be preferable to use several different kinds of data for the same feature to determine whether they capture the same pattern of disparity.
 
-The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
-Discrete character are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008,Close2015}.
-This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups.
+The points above assume that researchers are collecting their own data for disparity analyses, but this is often not the case.
+Discrete characters are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008,Close2015}.
+This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups, since phylogenetic characters are often collected to discriminate among groups.
 This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
 Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017}.
-This effect of the anatomical feature used on disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018}.
+The influence of trait choice on resulting disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018}.
 
@@
 Biological phenomena such as allometry and sexual dimorphism may also influence trait data.
-More mundanely, data collection is constrained by the time and money available, making collating a ``perfect'' dataset difficult.
-Ultimately, disparity analyses characterise the data, and subsamples of the universe of possible data may not have the power to uncover holistic patterns of disparity.
+More mundanely, data collection is constrained by the time and money available, making collating a ``perfect'' dataset impossible.
+Ultimately, disparity analyses are characterised by the data, and subsamples of the universe of possible data may not have the power to uncover the full patterns of disparity.
 Therefore, trait data should be collected with the scientific question in mind, or the question asked should be tailored to the limits of the data available.
@@
 
-\subsection{From data to morphospace} \label{section:morphospace}
-
-% TG: The whole argument here is the same as for logging data in biology: we conserve the same properties of the biological observation (e.g.
-a distribution of body mass) but we modify it linearly to make it easier to analyse (e.g.
-the log(body mass)).
-In my opinion same goes with a data matrix: a distance matrix or a PCA of the observed trait matrix should no modify the biological properties as long as it it changed linearly (if you start arbitrarily removing axis of the PCA or points in the distance matrix, this results in something different).
-Therefore, morphospace = collected data matrix (any any linear transformation thereof).
-
-The concept of what the morphospace \textit{is} can differ among authors and studies but is generally used in the common (i.e. non-mathematical) sense of a space containing all possible morphologies.
-For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points.
-In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combinations of traits.
-We generalise this definition to any linear transformation ($V \to W$) of this space that retains the biological definition, i.e.
-the morphospace contains all the observed combinations of traits (see appendix \ref{section:appendixmorphospace}).
-Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see section \ref{section:ordination} below).
-If the morphospace is transformed mathematically (i.e.
-the dataset is transformed using logarithms, ordinations, distance matrices, etc.) it remains the morphospace provided all the observation and traits remain in the datatset.
-
 \section{Disparity analysis methods} \label{section:methods}
@@
 Study design encompasses several key aspects including \ref{section:ordination} the difficulty of dealing with multidimensional data; \ref{section:metrics} the indices used to summarise the relative disparity of groups; \ref{section:testing} the methods used for hypothesis testing within the disparity analysis framework; and \ref{section:phylo} the influence of phylogeny on disparity analyses.
-We consider these aspects in turn below.
+We consider these aspects in order below.
 

### 2020-03-29 Writing/disparity_review.tex
Updated metrics bit

@@
 Disparity analyses often use ordination techniques for dimension reduction.
-Ordinations are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g.
+Ordinations are statistical methods that map observed variables onto a new space of reduced dimension while maintaining the requirement that similar observations are closer together than dissimilar ones (e.g.
 principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS; etc.).
 They come in many flavours depending on the data and the desired morphospace properties.
-For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling \citep[NMDS; see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012}.
+For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling \citep[NMDS; see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012}.
 
@@
 Additionally, after ordinating the data, it is possible to focus on just a subset of axes of the morphospace (i.e.
-selecting only those axes that describe the majority of the variation in the dataset - e.g.
-95\%).
+selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
 In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012,dryden2016statistical}.
-In practice, this allows us to interpret only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g.
+In practice, this facilitates interpretation of only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g.
 the elongation and flattening of birds beaks;][]{bright2016shapes}.
 
-Like most other aspects of disparity analyses, however, reducing dimensionality can be tricky.
-In the case of ordination in a geometric morphometric context, subsampling axes from the ordination can lead to misinterpretation of the results. 
-Although a common technique is to consider the \textit{n} axes that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
+Like most other aspects of disparity analyses, however, reducing dimensionality can be fraught.
+In the case of ordination, subsampling axes from the ordination can lead to misinterpretation of the results.
+Although a common technique is to consider the \textit{d} axes that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
 Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018, Gerber2017}.
 
-Interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.
+Interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question (for example, if the first few ordination axes represent elongation of the beak in birds, but the question is about wing disparity).
 Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
-For example, when the analysed data is non-Euclidean (e.g.
+For example, when the analysed data are non-Euclidean (e.g.
 some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
@@
 the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018}), some are not \citep[e.g.][]{Weisbecker2019}.
-For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and those that have more in state 0, has no biological meaning if these are binary alternate states.
+For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and those that have more in state 0 has no biological meaning if these are binary alternate states.
 
-Categorical data are a good deal more problematic, because the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean, and the distribution of the variance is usually more evenly distributed across axes (i.e.
-contrary to a PCA, the first few axis do not encompass most variance of the dataset).
-Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[i.e.
+In general, categorical data are a good deal more problematic than continuous data, because the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean, and the distribution of the variance is usually more evenly distributed across axes (i.e.
+contrary to a PCA, the first few axes do not encompass most variance of the dataset).
+Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, character coding and missing data can make the pairwise dissimilarity matrix lose its metric properties \citep[i.e.
 the distance between A and B is not equal to the distance between B and A;][]{Gerber2014}.
 Last, but not least, in many cases, ordination might not be necessary.
-For example, if an index (e.g. sum or range of variance) used to characterise disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep[e.g.][]{Close2015}.
+For example, if an index characterising disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep[e.g.][]{Close2015}.
 For all of these reasons, multidimensional data should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.
@@
     \tiny{
-    Morphospaces: different mathematical representations of a morphospace.
-    A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015} or an ordinated matrix \citep[e.g. in][]{tyler2011detecting}.
+    \textbf{Morphospaces}: different mathematical representations of a morphospace.
+A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015} or an ordinated matrix \citep[e.g.
+in][]{tyler2011detecting}.
     Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
-    Visualisation: different  ways to represent the morphospace in 2D.
+    \textbf{Visualisation}: different  ways to represent the morphospace in 2D.
     Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g.
@@
     Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).
+Data depicted is from the classic ``iris'' dataset of \citep{edgar1935irises,fisher1936use}.
     }
@@
 
-Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
+Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e.
+interpretable) summary of disparity.
 This is usually achieved with a disparity measure or index \citep{Hopkins2017}.
@@
 Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
-For example, if we want to characterise the extent of morphospace occupied by a group (e.g.
+For example, if the goal is to characterise the extent of morphospace occupied by a group (e.g.
 does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g.
-volume;
-\citealt{Diaz2016}, distance from the centroid; \citealt{Hopkins2017, Finlay2015}, variance and range; \citealt{Brusatte2008}).
+volume; \citealt{Diaz2016}, distance from the centroid; \citealt{Hopkins2017, Finlay2015}, variance and range; \citealt{Brusatte2008}).
 Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g.
@@
 
-In addition to considering what properties of disparity these indices should capture, it is also important to consider the mathematical properties of the indices and their associated caveats \citep{Wills2001, Ciampaglio2001}.
-For example, measuring the sum of variance-covariance for each dimension of the space is equivalent to measuring the sum of variance in a PCA (because the co-variance is null).
-However, this is not true of other transformations of the space or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012}.
+In addition to considering which properties of disparity these indices capture, it is also important to consider the mathematical properties of the indices and their associated caveats \citep{Wills2001, Ciampaglio2001}.
+For example, measuring the sum of variance for each dimension of the space is equivalent to measuring the sum of variance in a PCA (because the co-variance is null).
+However, this is no longer true of other transformations of the space or when a subset of dimensions or elements are considered, such as is often done after PCA \citep{Legendre2012}.
 
-Furthermore, multidimensional spaces have some counterintuitive properties that should be considered such as the ``curse of dimensionality'' \citep{Bellman1966}.
-In spaces with some axes of variance lower than 1, product-based indices used as proxies of volumes (e.g.
-product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966}.
-Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Wills2001} or convex hull based indices \citep{Jackson2011}.
+Furthermore, multidimensional spaces have some counterintuitive properties that should be considered, such as the ``curse of dimensionality'' \citep{Bellman1966}.
+In spaces with some axes of variance lower than one, product-based indices used as proxies of volumes (e.g.
+product of ranges, hypervolume, hypercube, etc.) can tend quickly tend towards zero for spaces with even a modest number of dimensions \citep{Bellman1966}.
+Other types of indices are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Wills2001} or convex hull based \citep{Jackson2011} indices.
 

### 2020-03-29 Writing/disparity_review.tex
resting section done

@@
 The multidimensional statistical toolkit for ecology and evolution has greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018}, but some of these advances have yet to be implemented in disparity analyses.
-Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
-One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
-Although statistically valid, this test is not always directly related to the hypothesis under question.
-For example, PERMANOVA tests whether two groups share the same variance/covariance based on a distance matrix between observations and therefore does not require the morphospace to be ordinated (see section \ref{section:ordination} above).
+Instead, hypothesis testing has been mostly confined to a small set of well-established methods.
+One commonly used test is the non-parametric permutation analysis of variance (PERMANOVA) \citep{Anderson2001} that tests whether two groups share the same centroid and dispersion based on a distance matrix between observations.
+This test therefore does not require the morphospace to be ordinated (see section \ref{section:ordination} above).
 Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.
@@
 However, bootstrapped data are pseudoreplicates and thus non-independent.
-This violates the assumptions of most parametric statistical tests.
-Furthermore, the number of bootstrap pseudoreplicates will inevitably increase the false positive rate (Type I error).
+Furthermore, the number of bootstrap pseudoreplicates will inevitably increase the false positive rate (Type I error) \citep{strube1988bootstrap}.
 These factors are often ignored in disparity analyses.
+
 Finally, it is important to understand the limitations of the dataset for performing statistical analysis.
 Mainly, disparity analysis should be restrained to groups within the same morphospace and are more difficult between different morphospaces.
-This can be the case when comparing elements with different numbers of landmarks or different landmark configurations for example: they will result in different morphospaces and comparing disparity indices between them is not trivial.
+This can be the case when comparing elements with different numbers of landmarks or different landmark configurations which will result in different morphospaces;comparing disparity indices between these is not trivial.
 

### 2020-03-29 Writing/disparity_review.tex
Updated phylo

@@
 As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998}.
-Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common) the non-independence between observations should be taken into account.
-It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first $n$ axis of the ordination, and can lead to incorrect interpretations of the data \citep[such as wrongly supporting ``early burst'' type patterns;][]{Uyeda2015}.
+Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common), the non-independence between observations should be taken into account.
+It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first \textit{d} axis of the ordination, and can lead to incorrect interpretations of the data \citep[such as wrongly supporting ``early burst'' type patterns;][]{Uyeda2015}.
 
-One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in disparity-through-time analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
+One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in disparity-through-time analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny \citep{brusatte2011phylogenetic}.
 Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e.
@@
 Pre-transformation ancestral state estimation will change the way the ordination space is defined -- i.e.
-the relationship between the points are not yet estimated -- and requires longer computational times.
-However, once the morphospace is defined its properties will not change.
+the relationship between the points is not yet estimated -- and requires longer computational times.
+However, once the morphospace is defined, its properties will not change.
 Post-transformation ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \citep{lloyd2018}.
@@
 All ancestral state estimates are highly dependent on the data and method used \cite{Louca2019}.
-In general, using ancestral state estimation can help with recovering patterns of changes in disparity but should not be used simply to generate extra data points to increase statistical power.
-In fact, these extra points are not independent and can also have problematic side effects, especially when testing for the influence of mass extinctions on disparity as they artificially increase taxon sampling.
+In general, using ancestral state estimation can help with recovering patterns of change in disparity but should not be used simply to generate extra data points to increase statistical power.
+In fact, these extra points are not independent and can also have problematic side effects, especially when testing for the influence of mass extinctions on disparity as they artificially and asymmetrically increase taxon sampling.
 

### 2020-03-29 Writing/disparity_review.tex
Fully updated comments

@@
 Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
-We advocate that researchers should assemble their analytical approaches based on an experimental approach that explores the impact of competing methods, such as choice of indices, ordination method and ancestral state estimation method on disparity analysis results.
-Thankfully this is becoming easier through the availability of diverse well documented R packages for multidimensional analysis \citep{Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b}.
+We advocate that researchers should assemble their analytical protocol based on an experimental approach that explores the impact of competing methods, such as choice of indices, ordination method, and ancestral state estimation method on disparity analysis results.
+Thankfully, this is becoming easier through the availability of diverse, well documented R packages for multidimensional analysis \citep{Bouxin2005, oksanen2007vegan, Harmon2008, lloyd2016, Guillerme2018b}.
 Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \citep{Donohue2013, Saupe2015, Canter2018, mammola2019}.
@@
 While studies of morphological disparity would benefit from advances in multidimensional analysis in other fields, the concept of a morphospace could reciprocally benefit other disciplines.
-For example, the multidimensional analysis of \citep{Diaz2016}, which analysed patterns of form and function in plants, is essentially an eco-morphospace; isotopic analyses of organisms \citep{Jackson2011,Swanson2015} can be represented as an isotope-space; ecosystem functioning in \citealt{Donohue2013} as an ecosystem-space, etc.
+For example, the multidimensional analysis of \citep{Diaz2016}, which analysed patterns of form and function in plants, is essentially an eco-morphospace; isotopic analyses of organisms \citep{Jackson2011,Swanson2015} can be represented as an isotope-space; ecosystem functioning in \citealt{Donohue2013} as an ecosystem-space \citep{qiao2017}, etc.
 These generalisations could also be exported for any set of traits: cognate approaches have been adopted in the analysis of single cell comparative transcriptome data \citep{Sebe-Pedros2018} where interpretation of the resulting transcriptome-spaces would be improved by giving careful attention to the concerns we highlight concerning morphospaces.
@@
 
-TG, NC and PD proposed this review; TG and NC led the writing supported by PD and GT. All authors edited drafts and approved the final version.
+TG, NC and PD proposed this review; TG and NC led the writing supported by PD and GT.
+All authors edited drafts and approved the final version.
 
@@
 SB was funded by European Research Council (ERC) under the European Union's Horizon 2020 research and innovation program (grant agreement No 756226, ERC Starting Grant: PalM) and a Leverhulme Trust Research Project Grant (RPG-2017-167);
-TG was funded by ARC DP170103227 and FT180100634 awarded to VW.
-
-\subsection{Appendix: morphospace definition in this paper} \label{section:appendixmorphospace}
-
-\noindent \cite{mitteroecker2009concept} define a morphospace as:
-
-\setlength{\leftskip}{1cm}
-\setlength{\rightskip}{1cm}
-
-\noindent \textit{``[a] mathematical spaces describing and relating the phenotypic configuration of biological organisms [...].
-In a typical morphospace, the morphological configuration of an organism is represented by a single point, and the dimensionality of the space is determined by the number of measured variables (Q-space).''} \citealt{mitteroecker2009concept} p.55.
-
-\setlength{\leftskip}{0pt}
-\setlength{\rightskip}{0pt}
-
-\noindent Here, we generalise the definition of a morphospace following \cite{mitteroecker2009concept} as the Q-space or any of its linear transformation.
-The morphospace is any matrix $(a_{ij})$ with $n$ observations with $d$ trait values each or any linear transformation $f: V \to W$ of $(a_{ij})$.
-In other words the morphospace is the observed data matrix (e.g. the collected data) or any of it's linear transformation (e.g. a PCA, a distance matrix, a log, etc...).
-In \cite{mitteroecker2009concept}'s definition, the morphospace needs a conserved $n \times d$ dimensions whereas our definition propose only the conservation of $n$ dimensions (the number of observations) as long as the morphospace can be expressed as:
-
-\begin{equation}
-    M1 = f(a_{ij})\in \mathbb{R}^{{n}\times{d}}
-\end{equation}
-
-\noindent Where $f$ can be any linear transformation function (including no transformation: $f(x) = x$).
-This generalisation allows to consider linear transformations of $(a_{ij})$ as a different ``orientations'' or ``resolutions'' (not in the mathematical sense \textit{per se}) of the observed trait matrix $(a_{ij})$.
-For example, a distance matrix of $(a_{ij})$ can be represented through the following linear transformation:
-
-\begin{equation}
-    M2 = f_{a_{ij}} = D_{a_{ij}}
-\end{equation}
-
-\noindent Where $D$ is a distance function for any pair of $ij$ in the matrix (e.g. if the distance is Euclidean $D_{ij} = \sqrt{\sum(i_{d}-j_{d})^2}$).
-$M2$ is then just a different representation of the original morphospace $(a_{ij})$ where the properties between the $n$ observations are expressed as distances.
-Note that in our definition, $f_{a_{ij}}$ does not need to result in a metric (or metric Euclidean) space although this is often desired in disparity analysis.
-
-Our definition doesn't consider a matrix to be a morphospace of $(a_{ij})$ if it results from a non-linear transformation of $(a_{ij})$ matrix.
-For example if one performs an ordination (e.g. PCA) on the matrix $(a_{ij})$ such as:
-
-\begin{equation}
-    M3 = f_{a_{ij}} = P(a_{ij})
-\end{equation}
-
-\noindent Where $P$ is a principal components linear transformation.
-$M3$ is still a morphospace (i.e.
-a linear transformation of $(a_{ij})$).
-However if one only uses the two first axis of $M3$, this would result in a different space because it is no more a linear transformation of $(a_{ij})$.
-
-However, we \textbf{\underline{\textit{DO NOT}}} declare that doing such a non-linear transformation is incorrect in a disparity analysis context.
-As pointed out throughout the main manuscript, different methods can be appropriate for different questions and different datasets.
-The point of this appendix is to develop \textit{our} definition of a morphospace but we do not argue that any other definition is incorrect.
-We do, however, suggest that a definition of ``morphospace'' should be at least briefly noted in papers using this mathematical or biological concept.
+TG was funded by ARC DP170103227 and FT180100634 awarded to VW;
+GHT was funded by European Research Council Consolidator Grant 615709 ToLERates and Royal Society University Research Fellowship UF120016.
 

### 2020-03-29 Writing/disparity_review.tex
added paragraph to link back to the intro

@@
 As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse them, stemming from the explicit hypotheses being tested.
+Although this makes comparison between disparity analyses difficult and makes generalisation required to answer the general biological questions from the introduction still premature (e.g.
+how does phenotypic variation evolves?), this diversity of methodological approaches provides researchers with a great number of tools tailored to answer specific biological questions.
+
 Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.

### 2020-03-29 Writing/disparity_review.tex
Updated references for Biol.Letters export

@@
 	\item \textbf{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e.
-	shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity \citep[but see][]{anderson2012}.
+	shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (but see \citealt{anderson2012}).
 	This makes disparity an informative complement to diversity measures based on species richness alone.
-	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two \citep[e.g.][]{Fortey1996, Hopkins2013}..
+	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two (e.g. \citealt{Fortey1996, Hopkins2013}).
 	The approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{pierce2008patterns}.
@@
 \noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
-1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[e.g.][]{Foote1989, Deline2018};
-2) continuous measurements of features \citep[e.g. lengths in][]{Friedman2010};
+1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait (e.g. \citealt{Foote1989, Deline2018});
+2) continuous measurements of features (e.g. lengths in \citealt{Friedman2010});
 or 3) more mathematical descriptors from geometric morphometric landmark data (e.g.
-Procrustes coordinates)  \citep[e.g.][]{sherratt2017rates}, Fourier coefficients \citep[e.g.][]{Foote1995, Spriggs2018} or model-based descriptors \citep[e.g.][]{Raup1961,Saunders2004} (Fig. 2).
+Procrustes coordinates)  (e.g. \citealt{sherratt2017rates}), Fourier coefficients (e.g. \citealt{Foote1995, Spriggs2018}) or model-based descriptors (e.g. \citealt{Raup1961,Saunders2004} - Fig. 2).
 None of these approaches is superior, but they may be more or less well-suited to characterising the traits compared and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017}.
@@
 The points above assume that researchers are collecting their own data for disparity analyses, but this is often not the case.
-Discrete characters are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008,Close2015}.
+Discrete characters are commonly recycled from phylogenetic studies (e.g. \citealt{Brusatte2008,Close2015}).
 This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups, since phylogenetic characters are often collected to discriminate among groups.
@@
 They come in many flavours depending on the data and the desired morphospace properties.
-For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling \citep[NMDS; see][chapter 9 for a detailed overview of ordination methods and properties]{Legendre2012}.
+For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
 
@@
 In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012,dryden2016statistical}.
-In practice, this facilitates interpretation of only the major axis of a highly dimensional dataset as major gradients of biological variation \citep[e.g.
-the elongation and flattening of birds beaks;][]{bright2016shapes}.
+In practice, this facilitates interpretation of only the major axis of a highly dimensional dataset as major gradients of biological variation (e.g.
+the elongation and flattening of birds beaks; \citealt{bright2016shapes}).
 
@@
 Although some gradients are easy to detect or interpret (e.g.
-the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018}), some are not \citep[e.g.][]{Weisbecker2019}.
+the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018}), some are not (e.g. \citealt{Weisbecker2019}).
 For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and those that have more in state 0 has no biological meaning if these are binary alternate states.
@@
 contrary to a PCA, the first few axes do not encompass most variance of the dataset).
-Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, character coding and missing data can make the pairwise dissimilarity matrix lose its metric properties \citep[i.e.
-the distance between A and B is not equal to the distance between B and A;][]{Gerber2014}.
+Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, character coding and missing data can make the pairwise dissimilarity matrix lose its metric properties (i.e.
+the distance between A and B is not equal to the distance between B and A; \citealt{Gerber2014}).
 Last, but not least, in many cases, ordination might not be necessary.
-For example, if an index characterising disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset \citep[e.g.][]{Close2015}.
+For example, if an index characterising disparity can use all of the data, it is not necessary to calculate it on the ordinated dataset (e.g. \citealt{Close2015}).
 For all of these reasons, multidimensional data should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.
@@
     \textbf{Morphospaces}: different mathematical representations of a morphospace.
-A trait matrix can be transformed into a distance matrix \citep[e.g in][]{Close2015} or an ordinated matrix \citep[e.g.
-in][]{tyler2011detecting}.
+A trait matrix can be transformed into a distance matrix (e.g.
+in \citealt{Close2015}) or an ordinated matrix (e.g.
+in \citealt{tyler2011detecting}).
     Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
@@
 Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common), the non-independence between observations should be taken into account.
-It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first \textit{d} axis of the ordination, and can lead to incorrect interpretations of the data \citep[such as wrongly supporting ``early burst'' type patterns;][]{Uyeda2015}.
+It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first \textit{d} axis of the ordination, and can lead to incorrect interpretations of the data (such as wrongly supporting ``early burst'' type patterns; \citealt{Uyeda2015}).
 
@@
 ordination, or distance matrix construction) and is simply based on the original data; or (2) post-transformation, i.e.
-the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix \citep[e.g.
-the ordination scores;][]{lloyd2018}.
+the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g.
+the ordination scores; \citealt{lloyd2018}).
 

### 2020-03-29 Writing/disparity_review.tex
accents + minor stuff

@@
 Dublin, Ireland;
-$^{6}$ Institut de Systématique, Évolution, Biodiversité (ISYEB), Muséum
-national d'Histoire naturelle, CNRS, Sorbonne Université, EPHE,
-Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
+$^{6}$ Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Mus\'{e}um national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Universit\'{e} des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
 $^{7}$ Ryan Institute, School of Natural Sciences, National University of
@@
 
-\textbf{Abstract (200 words max)}
+\textbf{Abstract}
 

### 2020-03-29 Writing/disparity_review.tex
submission version?

@@
 Katie E. Davis$^{4}$,
-Andrew L Jackson$^{5}$,
+Andrew L. Jackson$^{5}$,
 Sylvain Gerber$^{6}$,
@@
 
-\noindent { \it 
+\noindent{
 $^{1}$ School of Biological Sciences, University of Queensland, St. Lucia,
@@
 $^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding
-author: guillert@tcd.ie\\}
+author: guillert@tcd.ie}
 
@@
 taxa) can be placed in this space based on their trait values.
-Such multidimensional spaces (or morphospaces - defined broadly hereafter as a mathematical space relating morphological configurations generally based on some measure of similarity \citep{mitteroecker2009concept}) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig.
-1): %remove fig1?
+Such multidimensional spaces (or morphospaces - defined broadly hereafter as a mathematical space relating morphological configurations generally based on some measure of similarity \citep{mitteroecker2009concept}) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes: %remove fig1?
 
@@
 or 3) more mathematical descriptors from geometric morphometric landmark data (e.g.
-Procrustes coordinates)  (e.g. \citealt{sherratt2017rates}), Fourier coefficients (e.g. \citealt{Foote1995, Spriggs2018}) or model-based descriptors (e.g. \citealt{Raup1961,Saunders2004} - Fig. 2).
+Procrustes coordinates)  (e.g.
+\citealt{sherratt2017rates}), Fourier coefficients (e.g.
+\citealt{Foote1995, Spriggs2018}) or model-based descriptors (e.g.
+\citealt{Raup1961,Saunders2004} - Fig.
+\ref{Fig:data}).
 None of these approaches is superior, but they may be more or less well-suited to characterising the traits compared and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017}.
@@
 \caption{
-    \tiny{
-    \textbf{Morphospaces}: different mathematical representations of a morphospace.
+    Morphospaces: different mathematical representations of a morphospace.
 A trait matrix can be transformed into a distance matrix (e.g.
@@
     Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
-    \textbf{Visualisation}: different  ways to represent the morphospace in 2D.
+    Visualisation: different  ways to represent the morphospace in 2D.
     Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g.
@@
 Data depicted is from the classic ``iris'' dataset of \citep{edgar1935irises,fisher1936use}.
-    }
 }

### 2020-03-30 Writing/disparity_review.tex
Updated vera's affiliation

@@
 
-Thomas Guillerme$^{1,17,*, +}$,
+Thomas Guillerme$^{1,18,*, +}$,
 Natalie Cooper$^{2, +}$,
@@
 Graham J. Slater$^{16}$,
-Vera Weisbecker$^{1}$,
-Gavin H. Thomas$^{17}$
+Vera Weisbecker$^{1,17}$,
+Gavin H. Thomas$^{18}$
 and Philip C. J. Donoghue$^{12}$
@@
 Chicago, Illinois 60637;
-$^{17}$ Department of Animal and Plant Sciences, The University of Sheffield,
+$^{18}$ College of Science and Engineering, Flinders University,
+Adelaide SA5042, Australia;\\
+$^{18}$ Department of Animal and Plant Sciences, The University of Sheffield,
 Sheffield S10 2TN, UK;\\

### 2020-04-30 Writing/disparity_review.tex
Added revisions to the paper

@@
 \usepackage{pdflscape}
+\usepackage{xcolor}
 
@@
 \begin{document}
+
+\section{Response to Referees}
+
+\subsection{Referee: 1}
+Concerning the section on disparity through time, perhaps hyphens should be removed; also, it may be misleading to use them, as sometimes the disparity-through-time phrase refers, in abbreviated format, to what is otherwise known (in full) as "mean relative subclade disparity through time", implemented in the DTT function in geiger.
+
+% TG: change naming of disparity through time
+
+At the beginning of page 7, I am not convinced that "synapomorphies will naturally lead an apparent shift or increase in disparity when new clades appear" in all cases.
+This very much depends on character-state distribution and, given a sufficient number of unique traits or combinations thereof across the sample, you may have comparable levels of disparity.
+
+% TG: tone this down
+
+On page 8, just to avoid confusion, PCO can also applied to quantitative data ONLY. End of page 8/beginning of page 9 could do with a small (few lines) exposition of protocols for axis selection.
+
+% TG: Add this precision
+
+Lines 211-213 are problematic. Do the authors mean that categorical data are ALWAYS more problematic because of the listed reasons? Some remedies at least exist for euclideanarity.
+
+% TG: Tone this down
+
+Line 269, about PERMANOVA, puzzles me. Yes, the test is used, but for geometric morphometric analyses at least, other tests have been devised.
+
+% TG: add more tests?
+
+\subsection{Referee: 2}
+
+That said, my biggest issue was the specific way the paper was subdivided.
+The "Measuring..", "Testing.." and "Phylogeny" subsections seemed a bit jumbled.
+For instance, a lot of the best advice on designing a custom test for your data is in the "Measuring.." section.
+Meanwhile, the "Testing.." section is very effective at convincing the reader not to take bootstrap values literally (a good thing to be effective at!) but doesn't contain much else in the way of advice or guidance.
+
+% TG: add guidance further than bootstraps.
+
+The "Phylogeny" section makes no explicit reference to the models of trait evolution that exist.
+Ancestral state estimates are mentioned, but the potential pitfalls of imposing a (possibly, if not likely) incorrect model of trait evolution is elided.
+Elsewhere in the paper the authors do an excellent job warning the reader away from just accepting default options, or doing something just because "that's the way it's done".
+Yet here, the reader is left without knowledge that accepting the default BM model uncritically could cause major issues with both Revell-style multivariate corrections and all ASRs.
+
+% TG: add a disclaimer about the methods
+
+In the first two sections, the authors reference back to the four main goals of disparity analyses they outlined.
+Using those four goals as a guide post for reworking the Testing and Phylogeny subsections may help improve the flow and usability of those parts.
+
+% TG: not sure what's meant here
+
+Last bit: I think Figure 2 should be redesigned, personally.
+I'm not sure to whom it is targeted.
+The top part's clear, but I suspect the bottom part would be opaque to those who would most benefit from reading this paper.
+Right now, it shows every step of every option all at once.
+But I think focusing in on a single complex and general concept (eg, a clear visual explanation of ordination alone, or a clear representation of distance generation, or...) would serve the paper better.
+
+% TG: redesign the figure 2 to only be PCA?/PCO?
+
+\subsubsection{Specific Comments:}
+
+Line 98: "[Disparity as a proxy for ecology] is particularly common in palaeobiology" I actually don't think this is true (at least, not in the way it is most likely to be read; see last paragraph).
+There are entire journals (eg, Functional Ecology) dedicated to neontological studies using variation in easily measured traits across space/time (disparity) as a proxy for changes in ecosystem functions \& species interactions (ecology).
+Plant physiologists in particular routinely use various measurable morphological features (stem diameters, root depth, etc) as proxies for various ecological features (e.g., aridity tolerance) and compare locations.
+Functional ecologists use different language, and rarely use the literal word "disparity", but the ideas are basically the same.
+The big differences I perceive are in terms of (1) how easy it is to test the reliability of the proxy and (2) how often workers even attempt to rigorously test that proxy.
+I think, especially given the specific citation here \& preceding sentence, that the authors mean something along the lines of "disparity-as-ecology is one of the primary ways paleobiologist have for investigating ecosystem function".
+That is, the "particularly common" is in reference to the large proportion of studies that use this approach in palaeobiology.
+It's just that, as-written, I think it makes this common approach sound like a "weird" palaeobiologist think.
+
+% TG: change the terminology here and reference moms for parallels with ecology
+
+Line 150 - 158: Nothing wrong or incorrect here, but the flow of this paragraph is odd.
+I find it very choppy.
+The first sentence is a long list of problems.
+The second is a set of two more problems.
+The third is a statement that everything has problems.
+The above three sentences, I think, could be reworked to flow more directly attached to the main point, which is that questions need to fit their data and vice versa.
+
+% TG: change the paragraph flow
+
+Line 171-172: Don't need e.g. and etc. in same parenthetical
+
+% TG: change the paragraph flow
+
+Lines 175 - 177: The discussion of dimension reduction using PCO here makes me think that more discussion in this paper of the importance of choosing an appropriate distance metric might be needed.
+That is, it'd be a whole separate paper trying to review or describe what is or is not appropriate!
+But I do think it's worth further emphasizing that it's a critical methodological choice that shouldn't be taken lightly for those intending to use MDS/NMDS
+
+% TG: insit that the distance metric is important (cite Lloyd)
+
+Line 195: I think you mean explicitly "geometric morphometric data". If that's true, being explicit is, I think, better.
+
+% TG: fix this
+
+Lines 234 - 239: This implies that a central tendency is, on its own, a measure of disparity.
+Below, in the discussion of the position within morphospace of a clade, I think this point is made clearer.
+But I think the first few sentences here should be clarified a bit.
+Something like "or differences in the central tendencies (mean/median/mode) of clades [or subsets to be more general]"
+
+% TG: add this
+
+Lines 263 - 289: The "Testing hypotheses..." section.
+I feel as though this section needs to be fleshed out.
+A lot of very good points are made, especially with respect to bootstrapping, but a reader will walk away from this section (1) knowing that PERMANOVA exists, (2) being a bit skeptical of bootstrap support, and (3) not much else.
+The last paragraph is, I think, the most important in this section, and I would recommend cutting from the other two to expand this.
+
+% TG: change these paragraphs
+
+The general point about tailoring analyses to your data is sound, and that does make general advice hard, but not impossible.
+A few clear principles or even just an explanation of "same" morphospace could be useful here.
+That is, up in the Measuring subsection in lines 239 - 249 the authors provide a lot of good sound advice that seems more fitting here.
+
+% TG: not sure what's meant here
+
+I'm just imagining a student who reads this paper early in their project.
+Then later, once they have data, tries to go back to look up some best practices and misses some good tips because they were in the "Measuring..." subsection instead of the "Testing..." subsection.
+It may be that the subsections should simply be reworked.
+
+% TG: move things around?
+
+Disparity \& Phylogeny section: This section is good, but perhaps overlong.
+Although, it doesn't discuss (what I perceive to be) the biggest problem with phylogenetic corrections to disparity analyses: the choice of trait evolution model.
+That is, "correcting" an ordination by imposing a specific model (e.g., Brownian) might cause more trouble than it solves if that model is a poor fit to the data.
+
+% TG: add something about trait evolution model for pPCA
+
+This is especially relevant as the discussion of Ancestral state estimation, which is entirely reliant on the model of trait reconstruction (as is briefly noted in line 315).
+
+% TG: also add something about the model in ACE
+
 %Running head
@@
 This article results from discussion at the Royal Society International Science Seminar on `Reconciling disparate views on disparity' held at Chicheley Hall, 9-10th January, 2018.
+\textcolor{blue}{We thank the two anonymous reviewers for their helpful and positive comments.}
 AG was funded by European Research Council Starting grant 637171 ADaPTiVE;

### 2020-04-30 Writing/disparity_review.tex
Reviewer 1 sorted

@@
 
-% TG: change naming of disparity through time
+\textcolor{blue}{We have removed hyphens from ``disparity-through-time''}
 
@@
 
-% TG: tone this down
+\textcolor{blue}{We changed the sentence to:}
+
+\textit{especially as synapomorphies can lead to apparent shift or increase in disparity when new clade appear (especially if the character-state distribution is skewed towards a particular clade).} l.@@@
 
@@
 
-% TG: Add this precision
+\textcolor{blue}{We've added the precision for the PCO with quantitative data:}
+
+\textit{or quantitative, or both} l.@@@
+
+\textcolor{blue}{And added a brief mention of protocols for axis selection:}
+
+\textit{(either by manually selecting the d axes that encompasses the desired variance or using methods such as the broken stick model; Legendre \& Legendre 2012 p.410)} l.@@@ + change cite to number
 
@@
 
-% TG: Tone this down
+\textcolor{blue}{We've now added a reference to the Cailliez (1983) correction to remedy for euclideanatrity}:
+
+\textit{although this can sometimes be corrected, Cailliez 1983} l.@@@
 
@@
 
-% TG: add more tests?
+\textcolor{blue}{We have added mentions to other tests common in morphological disparity analyses:}
+
+\textit{The last decade has also seen a series of development based on this test (e.g.
+the linear regression for multidimensional data Collyer et al 2015; or the phylogenetic ANOVA Adams 2014; but see Adams \& Collyer 2018; Lloyd 2016 for more).
+It is worth noting that most of these test do not require the morphospace to be ordinated (see section 1 above).
+Regardless of the statistical test used, they should be employed that are tailored to the question at hand, rather than simply following common practices.}
 
@@
 
-	\item \textbf{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed over time, by focussing on the disparity of taxa in particular time intervals or slices.
+	\item \textbf{\textcolor{blue}{Disparity through time}.} This approach investigates how the morphologies of organisms have changed over time, by focussing on the disparity of taxa in particular time intervals or slices.
 	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{prentice2011evolution, Guillerme2018}, or how does disparity change up to and across mass extinction events\citep{Friedman2010}?
@@
 %     The four main types of disparity analysis.
-\textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
+\textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{\textcolor{blue}{disparity through time}} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
 %     These categories are not independent and many disparity studies will cover more than one.
@@
 This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups, since phylogenetic characters are often collected to discriminate among groups.
-This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
+This needs to be considered when interpreting results, especially \textcolor{blue}{as synapomorphies can lead to apparent shift or increase in disparity when new clade appear (especially if the character-state distribution is skewed towards a particular clade).}
 Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017}.
@@
 They come in many flavours depending on the data and the desired morphospace properties.
-For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
+For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (\textcolor{blue}{or quantitative, or both}) can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
 
@@
 In the case of ordination, subsampling axes from the ordination can lead to misinterpretation of the results.
-Although a common technique is to consider the \textit{d} axes that encompasses 95\% or 99\% of the variance in the dataset, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
+Although a common technique is to consider the \textit{d} axes that encompasses 95\% or 99\% of the variance in the dataset \textcolor{blue}{(either by manually selecting the \textit{d} axes that encompasses the desired variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410)}, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
 Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018, Gerber2017}.
@@
 some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
-This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear.
+This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear \textcolor{blue}{(although this can sometimes be corrected \citealt{cailliez1983})}.
 Furthermore, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible \citep{Gerber2019}.
@@
 One commonly used test is the non-parametric permutation analysis of variance (PERMANOVA) \citep{Anderson2001} that tests whether two groups share the same centroid and dispersion based on a distance matrix between observations.
-This test therefore does not require the morphospace to be ordinated (see section \ref{section:ordination} above).
-Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.
+\textcolor{blue}{The last decade has also seen a series of development based on this test (e.g.
+the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{Adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
+It is worth noting that most of these test do not require the morphospace to be ordinated (see section \ref{section:ordination} above).
+Regardless of the statistical test used, they should be employed that are tailored to the question at hand, rather than simply following common practices.}
 
@@
 
-One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in disparity-through-time analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny \citep{brusatte2011phylogenetic}.
+One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in \textcolor{blue}{disparity through time} analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny \citep{brusatte2011phylogenetic}.
 Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e.

### 2020-04-30 Writing/disparity_review.tex
Added reviewers comments (in the main text)

@@
 \subsection{Referee: 1}
-Concerning the section on disparity through time, perhaps hyphens should be removed; also, it may be misleading to use them, as sometimes the disparity-through-time phrase refers, in abbreviated format, to what is otherwise known (in full) as "mean relative subclade disparity through time", implemented in the DTT function in geiger.
+\noindent Concerning the section on disparity through time, perhaps hyphens should be removed; also, it may be misleading to use them, as sometimes the disparity-through-time phrase refers, in abbreviated format, to what is otherwise known (in full) as "mean relative subclade disparity through time", implemented in the DTT function in geiger.
 
@@
 
-At the beginning of page 7, I am not convinced that "synapomorphies will naturally lead an apparent shift or increase in disparity when new clades appear" in all cases.
+\noindent At the beginning of page 7, I am not convinced that "synapomorphies will naturally lead an apparent shift or increase in disparity when new clades appear" in all cases.
 This very much depends on character-state distribution and, given a sufficient number of unique traits or combinations thereof across the sample, you may have comparable levels of disparity.
@@
 
-On page 8, just to avoid confusion, PCO can also applied to quantitative data ONLY. End of page 8/beginning of page 9 could do with a small (few lines) exposition of protocols for axis selection.
+\noindent On page 8, just to avoid confusion, PCO can also applied to quantitative data ONLY.
+End of page 8/beginning of page 9 could do with a small (few lines) exposition of protocols for axis selection.
 
@@
 
-Lines 211-213 are problematic. Do the authors mean that categorical data are ALWAYS more problematic because of the listed reasons? Some remedies at least exist for euclideanarity.
+\noindent Lines 211-213 are problematic. Do the authors mean that categorical data are ALWAYS more problematic because of the listed reasons? Some remedies at least exist for euclideanarity.
 
@@
 
-Line 269, about PERMANOVA, puzzles me. Yes, the test is used, but for geometric morphometric analyses at least, other tests have been devised.
+\noindent Line 269, about PERMANOVA, puzzles me. Yes, the test is used, but for geometric morphometric analyses at least, other tests have been devised.
 
@@
 It is worth noting that most of these test do not require the morphospace to be ordinated (see section 1 above).
-Regardless of the statistical test used, they should be employed that are tailored to the question at hand, rather than simply following common practices.}
+Regardless of the statistical test used, they should be employed that are tailored to the question at hand, rather than simply following common practices.} l.@@@
 
@@
 
-That said, my biggest issue was the specific way the paper was subdivided.
+\noindent That said, my biggest issue was the specific way the paper was subdivided.
 The "Measuring..", "Testing.." and "Phylogeny" subsections seemed a bit jumbled.
@@
 
-% TG: add guidance further than bootstraps.
+\textcolor{blue}{We've added some references to specific tests and insisted throughout the section that it is important to chose the right statistical test according to the data and question at hand (and pointed out to some general reviews detailing some of the options - see response to reviewer 1 above and response to specific comments below).}
+% TG: Not sure if that's enough but lacking inspiration for now
 
-The "Phylogeny" section makes no explicit reference to the models of trait evolution that exist.
+\noindent The "Phylogeny" section makes no explicit reference to the models of trait evolution that exist.
 Ancestral state estimates are mentioned, but the potential pitfalls of imposing a (possibly, if not likely) incorrect model of trait evolution is elided.
@@
 
-% TG: add a disclaimer about the methods
+\textcolor{blue}{We've added the following caveat to the section about ASR:}
 
-In the first two sections, the authors reference back to the four main goals of disparity analyses they outlined.
+\textit{Furthermore, phylogenetic PCA are also really dependent on the evolutionary model which can be sometimes difficult to choose (e.g.
+the default Brownian motion model is not always appropriate; \citealt{blomberg2020})} .
+l.@@@
+
+\noindent In the first two sections, the authors reference back to the four main goals of disparity analyses they outlined.
 Using those four goals as a guide post for reworking the Testing and Phylogeny subsections may help improve the flow and usability of those parts.
@@
 
-Last bit: I think Figure 2 should be redesigned, personally.
+\noindent Last bit: I think Figure 2 should be redesigned, personally.
 I'm not sure to whom it is targeted.
@@
 
-% TG: redesign the figure 2 to only be PCA?/PCO?
+\textit{We have now updated the figure to be just illustrating the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:}
+
+\textit{Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of \citealt{edgar1935irises,fisher1936use}).
+Morphospaces: different mathematical representations of a morphospace.
+A trait matrix can be ordinated matrix (e.g.
+in \citealt{tyler2011detecting} - or transformed into a distance matrix \citealt{Close2015}, not represented here).
+Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
+Visualisation: different  ways to represent the morphospace in 2D.
+Visualisations can use either trait plots (directly from the trait matrix); or ordination axis plots (directly from the ordinated matrix).
+Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).}
 
@@
 
-Line 98: "[Disparity as a proxy for ecology] is particularly common in palaeobiology" I actually don't think this is true (at least, not in the way it is most likely to be read; see last paragraph).
+\noindent Line 98: "[Disparity as a proxy for ecology] is particularly common in palaeobiology" I actually don't think this is true (at least, not in the way it is most likely to be read; see last paragraph).
 There are entire journals (eg, Functional Ecology) dedicated to neontological studies using variation in easily measured traits across space/time (disparity) as a proxy for changes in ecosystem functions \& species interactions (ecology).
@@
 
-% TG: change the terminology here and reference moms for parallels with ecology
+\textcolor{blue}{We've changed the sentence mentioning disparity and ecology and palaeobiology as follows:}
 
-Line 150 - 158: Nothing wrong or incorrect here, but the flow of this paragraph is odd.
+\textit{It is one of the primary ways to investigate ecosystem functioning in palaeobiology when the study species (and their functional characteristics) are extinct \citep{Wainwright2005}} l.@@@
+
+\noindent Line 150 - 158: Nothing wrong or incorrect here, but the flow of this paragraph is odd.
 I find it very choppy.
@@
 
-Line 171-172: Don't need e.g. and etc. in same parenthetical
+\noindent Line 171-172: Don't need e.g. and etc. in same parenthetical
 
-% TG: change the paragraph flow
+\textcolor{blue}{We removed the  ``etc.''}
 
-Lines 175 - 177: The discussion of dimension reduction using PCO here makes me think that more discussion in this paper of the importance of choosing an appropriate distance metric might be needed.
+\noindent Lines 175 - 177: The discussion of dimension reduction using PCO here makes me think that more discussion in this paper of the importance of choosing an appropriate distance metric might be needed.
 That is, it'd be a whole separate paper trying to review or describe what is or is not appropriate!
@@
 
-% TG: insit that the distance metric is important (cite Lloyd)
+\textcolor{blue}{We've added the following sentence highlighting the effect of the chosen distance metric and referring to a great paper looking at it in more details:}
 
-Line 195: I think you mean explicitly "geometric morphometric data". If that's true, being explicit is, I think, better.
+\textit{Note that when using a PCO, the distance metric on which to calculate the PCO can have some significant impact on the result morphospace \citep{lehmann2019}}. l.@@@
 
-% TG: fix this
+\noindent Line 195: I think you mean explicitly "geometric morphometric data". If that's true, being explicit is, I think, better.
 
-Lines 234 - 239: This implies that a central tendency is, on its own, a measure of disparity.
+\textcolor{blue}{We've added the term ``geometric'' to be more explicit.}
+
+\noindent Lines 234 - 239: This implies that a central tendency is, on its own, a measure of disparity.
 Below, in the discussion of the position within morphospace of a clade, I think this point is made clearer.
@@
 
-% TG: add this
+\textcolor{blue}{We've changed the sentence to:}
 
-Lines 263 - 289: The "Testing hypotheses..." section.
+\textit{disparity indices can be used compare the spread of the distributions (e.g.
+the range, quantiles or variance) or the differences in the central tendencies (i.e.
+mean, median or mode) of groups in the morphospace.} l.@@@
+
+%TG: point 1
+\noindent Lines 263 - 289: The "Testing hypotheses..." section.
 I feel as though this section needs to be fleshed out.
@@
 The last paragraph is, I think, the most important in this section, and I would recommend cutting from the other two to expand this.
-
-% TG: change these paragraphs
-
+%TG: point 2
 The general point about tailoring analyses to your data is sound, and that does make general advice hard, but not impossible.
@@
 
-% TG: not sure what's meant here
+%TG: I think these two points where meant to be merged no? See my response
 
-I'm just imagining a student who reads this paper early in their project.
+\textcolor{blue}{We've now added more references to explicit test (see response to comment of reviewer 1).
+However, to respond to this comment, we have shortened the second paragraph and highlighted the importance to thing twice about which test to apply to which dataset at the end of each paragraph to really insist on this point.}
+
+\noindent I'm just imagining a student who reads this paper early in their project.
 Then later, once they have data, tries to go back to look up some best practices and misses some good tips because they were in the "Measuring..." subsection instead of the "Testing..." subsection.
@@
 
-% TG: move things around?
+% TG: note sure how to handle this without rewriting and writing a "pro tips" section? Hence my non-response below:
+\textcolor{blue}{We understand the reviewer's point of view and acknowledge that some parts could be missed out by inattentive readers.
+However, we wrote this paper with the idea in mind to make it the least prescriptive as possible (our message after the meeting at the orgin of the paper was: ``do whatever disparity analysis you want but to it for the right reasons'') so we already had to tone down some prescriptive aspects in previous draft versions of this paper.}
 
-Disparity \& Phylogeny section: This section is good, but perhaps overlong.
+\noindent Disparity \& Phylogeny section: This section is good, but perhaps overlong.
 Although, it doesn't discuss (what I perceive to be) the biggest problem with phylogenetic corrections to disparity analyses: the choice of trait evolution model.
@@
 
-% TG: add something about trait evolution model for pPCA
+\textcolor{blue}{We've added a caveat to this section (see main comments above).}
 
-This is especially relevant as the discussion of Ancestral state estimation, which is entirely reliant on the model of trait reconstruction (as is briefly noted in line 315).
+\noindent This is especially relevant as the discussion of Ancestral state estimation, which is entirely reliant on the model of trait reconstruction (as is briefly noted in line 315).
 
-% TG: also add something about the model in ACE
+\textcolor{blue}{We've added the following to insist on the model part:}
+
+\textit{All ancestral state estimates are highly dependent on the data and method used (especially which evolutionary model) [60].} l.@@@
 
@@
 	This approach has been used to investigate hypotheses of competitive replacement \citep{tyler2011detecting} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010}.
+    \textcolor{blue}{It is one of the primary ways to investigate ecosystem functioning in palaeobiology when the study species (and their functional characteristics) are extinct} \citep{Wainwright2005}.
 	It is particularly common in palaeobiology where it is not possible to directly observe many ecological or functional characteristics of extinct species \citep{Wainwright2005}.
@@
 Ordinations are statistical methods that map observed variables onto a new space of reduced dimension while maintaining the requirement that similar observations are closer together than dissimilar ones (e.g.
-principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS; etc.).
+principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS).
 They come in many flavours depending on the data and the desired morphospace properties.
 For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (\textcolor{blue}{or quantitative, or both}) can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
+\textcolor{blue}{Note that when using a PCO, the distance metric on which to calculate the PCO can have some significant impact on the result morphospace} \citep{lehmann2019}.
 
@@
 Although a common technique is to consider the \textit{d} axes that encompasses 95\% or 99\% of the variance in the dataset \textcolor{blue}{(either by manually selecting the \textit{d} axes that encompasses the desired variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410)}, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
-Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018, Gerber2017}.
+Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing \textcolor{blue}{geometric} morphometric data \citep{Deline2018, Gerber2017}.
 
@@
 \centering
-   \includegraphics[width=0.9\textwidth]{Figures/figure_space.pdf}
+   \includegraphics[width=0.9\textwidth]{Figures/figure_space2.pdf}
 \caption{
-    Morphospaces: different mathematical representations of a morphospace.
-A trait matrix can be transformed into a distance matrix (e.g.
-in \citealt{Close2015}) or an ordinated matrix (e.g.
-in \citealt{tyler2011detecting}).
-    Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
+    \textcolor{blue}{Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of} \citealt{edgar1935irises,fisher1936use}).
+    \textcolor{blue}{Morphospaces: different mathematical representations of a morphospace.
+A trait matrix can be ordinated matrix (e.g.
+in }\citealt{tyler2011detecting}\textcolor{blue}{ - or transformed into a distance matrix }\citealt{Close2015}\textcolor{blue}{, not represented here).}
+    \textcolor{blue}{Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
     Visualisation: different  ways to represent the morphospace in 2D.
-    Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g.
-using non-metric multidimensional scaling forcing the dataset into two dimensions - note that variance is lost due to the distortion stress and non-metric properties of the resulting operation); or ordination axis plots (directly from the ordinated matrix).
-    Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).
-Data depicted is from the classic ``iris'' dataset of \citep{edgar1935irises,fisher1936use}.
+    Visualisations can use either trait plots (directly from the trait matrix); or ordination axis plots (directly from the ordinated matrix).
+    Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).}
 }
@@
 
-When considering only one dimension, disparity indices can be used to reflect the spread of the distribution (e.g.
-the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
+When considering only one dimension, \textcolor{blue}{disparity indices can be used compare the spread of the distributions (e.g.
+the range, quantiles or variance) or the differences in the central tendencies (i.e.
+mean, median or mode) of groups in the morphospace.}
 Among these indices, some will have more attractive properties than others, such as sensitivity to outliers.
@@
 the sum of variances), bootstrapping the data generates a distribution of the index that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
-However, bootstrapped data are pseudoreplicates and thus non-independent.
-Furthermore, the number of bootstrap pseudoreplicates will inevitably increase the false positive rate (Type I error) \citep{strube1988bootstrap}.
-These factors are often ignored in disparity analyses.
+However, bootstrapped data are pseudoreplicates and thus non-independent \textcolor{blue}{and can increase the false positive rate (Type I error)} \citep{strube1988bootstrap}.
+\textcolor{blue}{This, again, highlights the importance of tailoring the statistical test to the data and question at hand.}
 
@@
 It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first \textit{d} axis of the ordination, and can lead to incorrect interpretations of the data (such as wrongly supporting ``early burst'' type patterns; \citealt{Uyeda2015}).
+\textcolor{blue}{Furthermore, phylogenetic PCA are also really dependent on the evolutionary model which can be sometimes difficult to choose (e.g.
+the default Brownian motion model is not always appropriate;} \citealt{blomberg2020}).
 
-One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in \textcolor{blue}{disparity through time} analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny \citep{brusatte2011phylogenetic}.
+One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in \textcolor{blue}{disparity through time} analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny \citep{brusatte2011phylogenetic,Guillerme2018}.
 Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e.
@@
 
-All ancestral state estimates are highly dependent on the data and method used \cite{Louca2019}.
+All ancestral state estimates are highly dependent on the data and method used \textcolor{blue}{(especially which evolutionary model)} \cite{louca2020}.
 In general, using ancestral state estimation can help with recovering patterns of change in disparity but should not be used simply to generate extra data points to increase statistical power.

### 2020-05-01 Writing/disparity_review.tex
typo!
Edits to R1 responses

@@
 
-\textit{especially as synapomorphies can lead to apparent shift or increase in disparity when new clade appear (especially if the character-state distribution is skewed towards a particular clade).} l.@@@
+\textit{especially as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).} l.@@@
 
@@
 
-\textcolor{blue}{We've added the precision for the PCO with quantitative data:}
+\textcolor{blue}{We've added that PCO can be used on dissimilarity matrices built using qualitative, or quantitative data, or mixed data types:}
 
-\textit{or quantitative, or both} l.@@@
+\textit{dissimilarity matrices based on qualitative, quantitative or mixed data types can be reduced using PCO} l.@@@
 
@@
 
-\textit{(either by manually selecting the d axes that encompasses the desired variance or using methods such as the broken stick model; Legendre \& Legendre 2012 p.410)} l.@@@ + change cite to number
+\textit{(either by manually selecting the d axes that encompasses the desired cumulative variance or using methods such as the broken stick model; Legendre \& Legendre 2012 p.410)} l.@@@ + change cite to number
 
@@
 
-\textcolor{blue}{We've now added a reference to the Cailliez (1983) correction to remedy for euclideanatrity}:
+% NC: Not sure if this really answers the issue they had. Do we mean that categorical are always more problematic???
+
+\textcolor{blue}{We've now added a reference to the Cailliez (1983) correction to remedy for euclideanarity}:
 
@@
 
-\textcolor{blue}{We have added mentions to other tests common in morphological disparity analyses:}
+\textcolor{blue}{We have added mentions of other tests common in morphological disparity analyses:}
 
-\textit{The last decade has also seen a series of development based on this test (e.g.
+\textit{The last decade has also seen a series of developments based on this test (e.g.
 the linear regression for multidimensional data Collyer et al 2015; or the phylogenetic ANOVA Adams 2014; but see Adams \& Collyer 2018; Lloyd 2016 for more).
 It is worth noting that most of these test do not require the morphospace to be ordinated (see section 1 above).
-Regardless of the statistical test used, they should be employed that are tailored to the question at hand, rather than simply following common practices.} l.@@@
+Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.} l.@@@
 
@@
 
-\textit{We have now updated the figure to be just illustrating the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:}
+\textcolor{blue}{We have now updated the figure to be just illustrating the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:}
 
@@
 This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups, since phylogenetic characters are often collected to discriminate among groups.
-This needs to be considered when interpreting results, especially \textcolor{blue}{as synapomorphies can lead to apparent shift or increase in disparity when new clade appear (especially if the character-state distribution is skewed towards a particular clade).}
+This needs to be considered when interpreting results, especially \textcolor{blue}{as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).}
 Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017}.
@@
 They come in many flavours depending on the data and the desired morphospace properties.
-For example, quantitative (continuous) data can be reduced using PCA, and qualitative data (\textcolor{blue}{or quantitative, or both}) can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
+For example, quantitative (continuous) data can be reduced using PCA, and \textcolor{blue}{dissimilarity matrices based on qualitative, quantitative or mixed data types} can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
 \textcolor{blue}{Note that when using a PCO, the distance metric on which to calculate the PCO can have some significant impact on the result morphospace} \citep{lehmann2019}.
@@
 In the case of ordination, subsampling axes from the ordination can lead to misinterpretation of the results.
-Although a common technique is to consider the \textit{d} axes that encompasses 95\% or 99\% of the variance in the dataset \textcolor{blue}{(either by manually selecting the \textit{d} axes that encompasses the desired variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410)}, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
+Although a common technique is to consider the \textit{d} axes that encompass 95\% or 99\% of the variance in the dataset \textcolor{blue}{(either by manually selecting the \textit{d} axes that encompasses the desired cumulative variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410)}, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
 Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing \textcolor{blue}{geometric} morphometric data \citep{Deline2018, Gerber2017}.
@@
 One commonly used test is the non-parametric permutation analysis of variance (PERMANOVA) \citep{Anderson2001} that tests whether two groups share the same centroid and dispersion based on a distance matrix between observations.
-\textcolor{blue}{The last decade has also seen a series of development based on this test (e.g.
+\textcolor{blue}{The last decade has also seen a series of developments based on this test (e.g.
 the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{Adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
 It is worth noting that most of these test do not require the morphospace to be ordinated (see section \ref{section:ordination} above).
-Regardless of the statistical test used, they should be employed that are tailored to the question at hand, rather than simply following common practices.}
+Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.}
 

### 2020-05-01 Writing/disparity_review.tex
Edits to R2 responses

@@
 
+We thank the reviewers for their useful and insightful comments. We respond to them below in blue.
+
 \subsection{Referee: 1}
@@
 
-\textcolor{blue}{We've added some references to specific tests and insisted throughout the section that it is important to chose the right statistical test according to the data and question at hand (and pointed out to some general reviews detailing some of the options - see response to reviewer 1 above and response to specific comments below).}
+\textcolor{blue}{We suspect this is probably due to us doing a poor job at explaining how these sections differ.
+In a standard disparity workflow, we tend to first extract disparity values for our chosen subsets of taxa which we refer to as "Measuring disparity".
+Following from this, some (though not all) studies go on to determine whether differences among subsets are statistically different.
+We refer to this step as "Testing hypotheses".
+To reduce confusion we have renamed these sections as 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity.
+We have retained the Disparity and phylogeny subsection as this is somewhat of a side issue that is important for these studies, but depending on the clade in question might be dealt with well or might be ignored due to lack of phylogenies for the group that could actually be used for these analyses.}
+
+\textcolor{blue}{For the Testing section, we've added some references to specific tests and insisted throughout the section that it is important to chose the right statistical test according to the data and question at hand (and pointed to some general reviews detailing some of the options - see response to reviewer 1 above and response to specific comments below).}
 % TG: Not sure if that's enough but lacking inspiration for now
+% NC: I think the error the reviewer made is what I've written above, so changing the names should deal with it.
 
@@
 
-\textcolor{blue}{We've added the following caveat to the section about ASR:}
+\textcolor{blue}{We've now added the following caveats to this section:}
 
-\textit{Furthermore, phylogenetic PCA are also really dependent on the evolutionary model which can be sometimes difficult to choose (e.g.
-the default Brownian motion model is not always appropriate; \citealt{blomberg2020})} .
+\textit{Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution.
+Standard methods assume a model of Brownian motion, i.e.
+a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution.
+In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{blomberg2020}.
+If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.)} .
 l.@@@
+
+\textit{All ancestral state estimates are highly dependent on the data and method used (especially on the underlying model of trait evolution)}
 
@@
 
+\textcolor{blue}{We appreciate the reviewer's suggestion here and tried hard to implement it.
+Unfortunately it led to the sections being even more confused as many of our points apply across the four types of disparity analysis.
+We also didn't feel we had stuck to this structure in the earlier sections, so the structure here didn't make as much sense as the reviewer assumed it might.
+We therefore reverted back to the original structure.}
+
 % TG: not sure what's meant here
+% NC: What they mean is can we use the 4 types of disparity analysis from the intro to say: For this type of analysis you should do this test, for this type do this test. But I don't think we can...
 
@@
 
-\textcolor{blue}{We have now updated the figure to be just illustrating the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:}
+\textcolor{blue}{We have now updated the figure to be just illustrate the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:}
 
@@
 Morphospaces: different mathematical representations of a morphospace.
-A trait matrix can be ordinated matrix (e.g.
+A trait matrix can be an ordinated matrix (e.g.
 in \citealt{tyler2011detecting} - or transformed into a distance matrix \citealt{Close2015}, not represented here).
+% NC: Not sure the brackets are in the right places here?
+As is the sentence reads "A trait matrix can be an ordinated matrix (....)."
 Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
@@
 
-\textcolor{blue}{We've changed the sentence mentioning disparity and ecology and palaeobiology as follows:}
+\textcolor{blue}{This is a very fair observation. We've changed the sentence mentioning disparity and ecology and palaeobiology as follows:}
 
@@
 
-% TG: change the paragraph flow
+\textcolor{blue}{We phrased the paragraph as shown below. Hopefully this has improved the flow to the main point.}
+
+\textit{Ultimately, disparity analyses are characterised by the data they use.
+Unfortunately, trait data suffer from the same shortcomings as most biological datasets.
+The data within them can be non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, and/or correlated \citep{Palci2018}.
+There are also issues of missing data, both where a particular character cannot be measured for a given taxon, or where a given taxon cannot be sampled at all.
+Trait data may also be influenced by biological phenomena such as allometry and sexual dimorphism.
+More practically, data collection is constrained by the time and money available, making collating a ``perfect'' dataset impossible.
+Even when care is taken, subsamples of the universe of possible data may not have the power to uncover the full patterns of disparity.
+These issues should be considered when collecting data.
+It is particularly important to collect trait data with the scientific question in mind, or, where there are limits on the data available, to tailor the question being asked to match the data.
+}
 
@@
 
-\textcolor{blue}{We've added the following sentence highlighting the effect of the chosen distance metric and referring to a great paper looking at it in more details:}
+\textcolor{blue}{We've added the following sentence highlighting the effect of the chosen distance metric and referring to a great paper looking at it in more detail:}
 
-\textit{Note that when using a PCO, the distance metric on which to calculate the PCO can have some significant impact on the result morphospace \citep{lehmann2019}}. l.@@@
+\textit{Note that for PCO, the distance metric used can have significant impacts on the resulting morphospace \citep{lehmann2019}.
+Choice of distance metric is therefore crucial, and should not be overlooked when using PCO.}.
+l.@@@
 
@@
 
-\textit{disparity indices can be used compare the spread of the distributions (e.g.
+\textit{disparity indices can be used to compare the spread of distributions (e.g.
 the range, quantiles or variance) or the differences in the central tendencies (i.e.
@@
 
-%TG: point 1
+\textcolor{blue}{The following three points by the reviewer are closely linked so we deal with them together}
+
 \noindent Lines 263 - 289: The "Testing hypotheses..." section.
@@
 The last paragraph is, I think, the most important in this section, and I would recommend cutting from the other two to expand this.
-%TG: point 2
-The general point about tailoring analyses to your data is sound, and that does make general advice hard, but not impossible.
+
+\noindent The general point about tailoring analyses to your data is sound, and that does make general advice hard, but not impossible.
 A few clear principles or even just an explanation of "same" morphospace could be useful here.
 That is, up in the Measuring subsection in lines 239 - 249 the authors provide a lot of good sound advice that seems more fitting here.
-
-%TG: I think these two points where meant to be merged no? See my response
-
-\textcolor{blue}{We've now added more references to explicit test (see response to comment of reviewer 1).
-However, to respond to this comment, we have shortened the second paragraph and highlighted the importance to thing twice about which test to apply to which dataset at the end of each paragraph to really insist on this point.}
 
@@
 
-% TG: note sure how to handle this without rewriting and writing a "pro tips" section? Hence my non-response below:
-\textcolor{blue}{We understand the reviewer's point of view and acknowledge that some parts could be missed out by inattentive readers.
-However, we wrote this paper with the idea in mind to make it the least prescriptive as possible (our message after the meeting at the orgin of the paper was: ``do whatever disparity analysis you want but to it for the right reasons'') so we already had to tone down some prescriptive aspects in previous draft versions of this paper.}
+\textcolor{blue}{In response to this, and comments by reviewer 1, we have fleshed out the "Testing" section to mention other existing tests.
+We have also, as suggested by the reviewer, shortened the other paragraphs and expanded this section.
+This is a difficult section to write because the possible statistical tests heavily rely on the kind of data available and the question being asked.
+In many studies only visual observations of differences are made without statistical testing.
+The edited text is as follows:}
+
+% add the editted text here
+
+\textcolor{blue}{We appreciate the reviewer's point about inattentive readers missing key sections.
+We hope that we have solved this in part by renaming these sections to more accurately represent their content (changed to 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity).
+As mentioned above, we tried reshuffling some of this material, but the current structure best suits the way most disparity analyses are planned and carried out (in our experience).}
+
+\textcolor{blue}{We avoid giving general advice because the idea with this paper was to inform but to be as non-prescriptive as possible (our take home message after the meeting at the origin of the paper was: ``do whatever disparity analysis you want but do it for the right reasons'') so we already had to tone down some prescriptive aspects in previous draft versions of this paper.
+We have however, highlighted the importance of applying the appropriate test to for the question and data twice in this section to really emphasize the point.}
+
+% add edited text here
 
@@
 
-\textcolor{blue}{We've added a caveat to this section (see main comments above).}
+\textcolor{blue}{We agree this is a really important point that we have not expnaded on previously du to worries about the word count.
+We've now added a caveat to this section, as mentioned in the main comments:}
+
+\textit{Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution.
+Standard methods assume a model of Brownian motion, i.e.
+a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution.
+In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{blomberg2020}.
+If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.)} .
+l.@@@
 
@@
 
-\textcolor{blue}{We've added the following to insist on the model part:}
+\textcolor{blue}{We've also added the following to emphasize this point:}
 
-\textit{All ancestral state estimates are highly dependent on the data and method used (especially which evolutionary model) [60].} l.@@@
+\textit{All ancestral state estimates are highly dependent on the data and method used (especially on the underlying model of trait evolution)}
 
@@
     \textcolor{blue}{It is one of the primary ways to investigate ecosystem functioning in palaeobiology when the study species (and their functional characteristics) are extinct} \citep{Wainwright2005}.
-	It is particularly common in palaeobiology where it is not possible to directly observe many ecological or functional characteristics of extinct species \citep{Wainwright2005}.
 
@@
 
-Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \citep{Palci2018}.
-Biological phenomena such as allometry and sexual dimorphism may also influence trait data.
-More mundanely, data collection is constrained by the time and money available, making collating a ``perfect'' dataset impossible.
-Ultimately, disparity analyses are characterised by the data, and subsamples of the universe of possible data may not have the power to uncover the full patterns of disparity.
-Therefore, trait data should be collected with the scientific question in mind, or the question asked should be tailored to the limits of the data available.
+Ultimately, disparity analyses are characterised by the data they use.
+Unfortunately, trait data suffer from the same shortcomings as most biological datasets.
+The data within them can be non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, and/or correlated \citep{Palci2018}.
+There are also issues of missing data, both where a particular character cannot be measured for a given taxon, or where a given taxon cannot be sampled at all.
+Trait data may also be influenced by biological phenomena such as allometry and sexual dimorphism.
+More practically, data collection is constrained by the time and money available, making collating a ``perfect'' dataset impossible.
+Even when care is taken, subsamples of the universe of possible data may not have the power to uncover the full patterns of disparity.
+These issues should be considered when collecting data.
+It is particularly important to collect trait data with the scientific question in mind, or, where there are limits on the data available, to tailor the question being asked to match the data.
 
@@
 For example, quantitative (continuous) data can be reduced using PCA, and \textcolor{blue}{dissimilarity matrices based on qualitative, quantitative or mixed data types} can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
-\textcolor{blue}{Note that when using a PCO, the distance metric on which to calculate the PCO can have some significant impact on the result morphospace} \citep{lehmann2019}.
+\textcolor{blue}{Note that for PCO, the distance metric used can have significant impacts on the resulting morphospace} \citep{lehmann2019}. 
+\textcolor{blue}{Choice of distance metric is therefore crucial, and should not be overlooked when using PCO.}
 
@@
 
-\subsection{Measuring disparity} \label{section:metrics}
+\textcolor{blue}{\subsection{Summarising disparity using disparity indices} \label{section:metrics}}
 
@@
 
-When considering only one dimension, \textcolor{blue}{disparity indices can be used compare the spread of the distributions (e.g.
+When considering only one dimension, \textcolor{blue}{disparity indices can be used to compare the spread of distributions (e.g.
 the range, quantiles or variance) or the differences in the central tendencies (i.e.
@@
 
-\subsection{Testing hypotheses about the evolution of disparity} \label{section:testing}
+\textcolor{blue}{\subsection{Testing for differences in disparity} \label{section:testing}}
 
@@
 Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common), the non-independence between observations should be taken into account.
-It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first \textit{d} axis of the ordination, and can lead to incorrect interpretations of the data (such as wrongly supporting ``early burst'' type patterns; \citealt{Uyeda2015}).
-\textcolor{blue}{Furthermore, phylogenetic PCA are also really dependent on the evolutionary model which can be sometimes difficult to choose (e.g.
-the default Brownian motion model is not always appropriate;} \citealt{blomberg2020}).
+It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first \textit{d} axes of the ordination, and can lead to incorrect interpretations of the data (such as wrongly supporting ``early burst'' type patterns; \citealt{Uyeda2015}).
+\textcolor{blue}{Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution.
+Standard methods assume a model of Brownian motion, i.e.
+a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution.
+In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{blomberg2020}.
+If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.}
 
@@
 
-All ancestral state estimates are highly dependent on the data and method used \textcolor{blue}{(especially which evolutionary model)} \cite{louca2020}.
+All ancestral state estimates are highly dependent on the data and method used \textcolor{blue}{(especially on the underlying model of trait evolution)} \cite{louca2020}.
 In general, using ancestral state estimation can help with recovering patterns of change in disparity but should not be used simply to generate extra data points to increase statistical power.

### 2020-05-01 Writing/disparity_review.tex
added color tags for conversion

@@
 
-\textcolor{blue}{We have removed hyphens from ``disparity-through-time''}
+\textcolor{blue}{COLORTAGIN[We have removed hyphens from ``disparity-through-time'']COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We changed the sentence to:}
+\textcolor{blue}{COLORTAGIN[We changed the sentence to:]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We've added that PCO can be used on dissimilarity matrices built using qualitative, or quantitative data, or mixed data types:}
+\textcolor{blue}{COLORTAGIN[We've added that PCO can be used on dissimilarity matrices built using qualitative, or quantitative data, or mixed data types:]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{And added a brief mention of protocols for axis selection:}
+\textcolor{blue}{COLORTAGIN[And added a brief mention of protocols for axis selection:]COLORTAGOUT}
 
@@
 
-% NC: Not sure if this really answers the issue they had. Do we mean that categorical are always more problematic???
+\textcolor{blue}{COLORTAGIN[We've now stated that it is not always problematic and added a reference to the Cailliez (1983) correction to remedy for euclideanarity]COLORTAGOUT}:
 
-\textcolor{blue}{We've now added a reference to the Cailliez (1983) correction to remedy for euclideanarity}:
-
-\textit{although this can sometimes be corrected, Cailliez 1983} l.@@@
+\textit{This can be sometimes problematic when comparing the position of groups in multidimensional space, as the distances might not be linear (although this can sometimes be corrected; Cailliez 1983} l.@@@
 
@@
 
-\textcolor{blue}{We have added mentions of other tests common in morphological disparity analyses:}
+\textcolor{blue}{COLORTAGIN[We have added mentions of other tests common in morphological disparity analyses:]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We suspect this is probably due to us doing a poor job at explaining how these sections differ.
+\textcolor{blue}{COLORTAGIN[We suspect this is probably due to us doing a poor job at explaining how these sections differ.
 In a standard disparity workflow, we tend to first extract disparity values for our chosen subsets of taxa which we refer to as "Measuring disparity".
@@
 To reduce confusion we have renamed these sections as 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity.
-We have retained the Disparity and phylogeny subsection as this is somewhat of a side issue that is important for these studies, but depending on the clade in question might be dealt with well or might be ignored due to lack of phylogenies for the group that could actually be used for these analyses.}
+We have retained the Disparity and phylogeny subsection as this is somewhat of a side issue that is important for these studies, but depending on the clade in question might be dealt with well or might be ignored due to lack of phylogenies for the group that could actually be used for these analyses.]COLORTAGOUT}
 
-\textcolor{blue}{For the Testing section, we've added some references to specific tests and insisted throughout the section that it is important to chose the right statistical test according to the data and question at hand (and pointed to some general reviews detailing some of the options - see response to reviewer 1 above and response to specific comments below).}
-% TG: Not sure if that's enough but lacking inspiration for now
-% NC: I think the error the reviewer made is what I've written above, so changing the names should deal with it.
+\textcolor{blue}{COLORTAGIN[For the Testing section, we've added some references to specific tests and insisted throughout the section that it is important to chose the right statistical test according to the data and question at hand (and pointed to some general reviews detailing some of the options - see response to reviewer 1 above and response to specific comments below).]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We've now added the following caveats to this section:}
+\textcolor{blue}{COLORTAGIN[We've now added the following caveats to this section:]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We appreciate the reviewer's suggestion here and tried hard to implement it.
+\textcolor{blue}{COLORTAGIN[We appreciate the reviewer's suggestion here and tried hard to implement it.
 Unfortunately it led to the sections being even more confused as many of our points apply across the four types of disparity analysis.
 We also didn't feel we had stuck to this structure in the earlier sections, so the structure here didn't make as much sense as the reviewer assumed it might.
-We therefore reverted back to the original structure.}
+We therefore reverted back to the original structure.]COLORTAGOUT}
 
-% TG: not sure what's meant here
-% NC: What they mean is can we use the 4 types of disparity analysis from the intro to say: For this type of analysis you should do this test, for this type do this test. But I don't think we can...
+\textcolor{blue}{COLORTAGIN[If we understand this reviewer well, they suggest we partition the methodological section per type of disparity analysis (e.g.
+for descriptive disparity analysis, one could apply such and such tests, etc...).
+If this is the case, we unfortunately think that this will make the paper to prescriptive and probably more complex to read since methods can be applied to different types of analysis.]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We have now updated the figure to be just illustrate the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:}
+\textcolor{blue}{COLORTAGIN[We have now updated the figure to be just illustrate the path from raw data to PCA (illustrating that both can be morphospaces).
+We updated the caption accordingly:]COLORTAGOUT}
 
@@
 A trait matrix can be an ordinated matrix (e.g.
-in \citealt{tyler2011detecting} - or transformed into a distance matrix \citealt{Close2015}, not represented here).
-% NC: Not sure the brackets are in the right places here?
-As is the sentence reads "A trait matrix can be an ordinated matrix (....)."
+in \citealt{tyler2011detecting}) or transformed into a distance matrix (e.g.
+in \citealt{Close2015}, not represented here).
 Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
@@
 
-\textcolor{blue}{This is a very fair observation. We've changed the sentence mentioning disparity and ecology and palaeobiology as follows:}
+\textcolor{blue}{COLORTAGIN[This is a very fair observation. We've changed the sentence mentioning disparity and ecology and palaeobiology as follows:]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We phrased the paragraph as shown below. Hopefully this has improved the flow to the main point.}
+\textcolor{blue}{COLORTAGIN[We phrased the paragraph as shown below. Hopefully this has improved the flow to the main point.]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We removed the  ``etc.''}
+\textcolor{blue}{COLORTAGIN[We removed the  ``etc.'']COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We've added the following sentence highlighting the effect of the chosen distance metric and referring to a great paper looking at it in more detail:}
+\textcolor{blue}{COLORTAGIN[We've added the following sentence highlighting the effect of the chosen distance metric and referring to a great paper looking at it in more detail:]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We've added the term ``geometric'' to be more explicit.}
+\textcolor{blue}{COLORTAGIN[We've added the term ``geometric'' to be more explicit.]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We've changed the sentence to:}
+\textcolor{blue}{COLORTAGIN[We've changed the sentence to:]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{The following three points by the reviewer are closely linked so we deal with them together}
+\textcolor{blue}{COLORTAGIN[The following three points by the reviewer are closely linked so we deal with them together]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{In response to this, and comments by reviewer 1, we have fleshed out the "Testing" section to mention other existing tests.
+\textcolor{blue}{COLORTAGIN[In response to this, and comments by reviewer 1, we have fleshed out the "Testing" section to mention other existing tests.
 We have also, as suggested by the reviewer, shortened the other paragraphs and expanded this section.
@@
 In many studies only visual observations of differences are made without statistical testing.
-The edited text is as follows:}
+The edited text is as follows:]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We appreciate the reviewer's point about inattentive readers missing key sections.
+\textcolor{blue}{COLORTAGIN[We appreciate the reviewer's point about inattentive readers missing key sections.
 We hope that we have solved this in part by renaming these sections to more accurately represent their content (changed to 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity).
-As mentioned above, we tried reshuffling some of this material, but the current structure best suits the way most disparity analyses are planned and carried out (in our experience).}
+As mentioned above, we tried reshuffling some of this material, but the current structure best suits the way most disparity analyses are planned and carried out (in our experience).]COLORTAGOUT}
 
-\textcolor{blue}{We avoid giving general advice because the idea with this paper was to inform but to be as non-prescriptive as possible (our take home message after the meeting at the origin of the paper was: ``do whatever disparity analysis you want but do it for the right reasons'') so we already had to tone down some prescriptive aspects in previous draft versions of this paper.
-We have however, highlighted the importance of applying the appropriate test to for the question and data twice in this section to really emphasize the point.}
+\textcolor{blue}{COLORTAGIN[We avoid giving general advice because the idea with this paper was to inform but to be as non-prescriptive as possible (our take home message after the meeting at the origin of the paper was: ``do whatever disparity analysis you want but do it for the right reasons'') so we already had to tone down some prescriptive aspects in previous draft versions of this paper.
+We have however, highlighted the importance of applying the appropriate test to for the question and data twice in this section to really emphasize the point.]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We agree this is a really important point that we have not expnaded on previously du to worries about the word count.
-We've now added a caveat to this section, as mentioned in the main comments:}
+\textcolor{blue}{COLORTAGIN[We agree this is a really important point that we have not expnaded on previously du to worries about the word count.
+We've now added a caveat to this section, as mentioned in the main comments:]COLORTAGOUT}
 
@@
 
-\textcolor{blue}{We've also added the following to emphasize this point:}
+\textcolor{blue}{COLORTAGIN[We've also added the following to emphasize this point:]COLORTAGOUT}
 
@@
 
-	\item \textbf{\textcolor{blue}{Disparity through time}.} This approach investigates how the morphologies of organisms have changed over time, by focussing on the disparity of taxa in particular time intervals or slices.
+	\item \textbf{\textcolor{blue}{COLORTAGIN[Disparity through time}.]COLORTAGOUT} This approach investigates how the morphologies of organisms have changed over time, by focussing on the disparity of taxa in particular time intervals or slices.
 	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{prentice2011evolution, Guillerme2018}, or how does disparity change up to and across mass extinction events\citep{Friedman2010}?
@@
 	This approach has been used to investigate hypotheses of competitive replacement \citep{tyler2011detecting} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010}.
-    \textcolor{blue}{It is one of the primary ways to investigate ecosystem functioning in palaeobiology when the study species (and their functional characteristics) are extinct} \citep{Wainwright2005}.
+    \textcolor{blue}{COLORTAGIN[It is one of the primary ways to investigate ecosystem functioning in palaeobiology when the study species (and their functional characteristics) are extinct]COLORTAGOUT} \citep{Wainwright2005}.
 
@@
 %     The four main types of disparity analysis.
-\textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{\textcolor{blue}{disparity through time}} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
+\textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{\textcolor{blue}{COLORTAGIN[disparity through time}} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
 %     These categories are not independent and many disparity studies will cover more than one.
@@
 This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups, since phylogenetic characters are often collected to discriminate among groups.
-This needs to be considered when interpreting results, especially \textcolor{blue}{as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).}
+This needs to be considered when interpreting results, especially \textcolor{blue}{COLORTAGIN[as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).]COLORTAGOUT}
 Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017}.
@@
 They come in many flavours depending on the data and the desired morphospace properties.
-For example, quantitative (continuous) data can be reduced using PCA, and \textcolor{blue}{dissimilarity matrices based on qualitative, quantitative or mixed data types} can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
-\textcolor{blue}{Note that for PCO, the distance metric used can have significant impacts on the resulting morphospace} \citep{lehmann2019}. 
-\textcolor{blue}{Choice of distance metric is therefore crucial, and should not be overlooked when using PCO.}
+For example, quantitative (continuous) data can be reduced using PCA, and \textcolor{blue}{COLORTAGIN[dissimilarity matrices based on qualitative, quantitative or mixed data types]COLORTAGOUT} can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
+\textcolor{blue}{COLORTAGIN[Note that for PCO, the distance metric used can have significant impacts on the resulting morphospace]COLORTAGOUT} \citep{lehmann2019}. 
+\textcolor{blue}{COLORTAGIN[Choice of distance metric is therefore crucial, and should not be overlooked when using PCO.]COLORTAGOUT}
 
@@
 In the case of ordination, subsampling axes from the ordination can lead to misinterpretation of the results.
-Although a common technique is to consider the \textit{d} axes that encompass 95\% or 99\% of the variance in the dataset \textcolor{blue}{(either by manually selecting the \textit{d} axes that encompasses the desired cumulative variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410)}, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
-Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing \textcolor{blue}{geometric} morphometric data \citep{Deline2018, Gerber2017}.
+Although a common technique is to consider the \textit{d} axes that encompass 95\% or 99\% of the variance in the dataset \textcolor{blue}{COLORTAGIN[(either by manually selecting the \textit{d} axes that encompasses the desired cumulative variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410]COLORTAGOUT)}, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
+Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing \textcolor{blue}{COLORTAGIN[geometric]COLORTAGOUT} morphometric data \citep{Deline2018, Gerber2017}.
 
@@
 some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
-This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear \textcolor{blue}{(although this can sometimes be corrected \citealt{cailliez1983})}.
+\textcolor{blue}{COLORTAGIN[This can be sometimes problematic when comparing the position of groups in multidimensional space, as the distances might not be linear (although this can sometimes be corrected ]COLORTAGOUT \citealt{cailliez1983})}.
 Furthermore, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible \citep{Gerber2019}.
@@
 \caption{
-    \textcolor{blue}{Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of} \citealt{edgar1935irises,fisher1936use}).
-    \textcolor{blue}{Morphospaces: different mathematical representations of a morphospace.
-A trait matrix can be ordinated matrix (e.g.
-in }\citealt{tyler2011detecting}\textcolor{blue}{ - or transformed into a distance matrix }\citealt{Close2015}\textcolor{blue}{, not represented here).}
-    \textcolor{blue}{Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
+    \textcolor{blue}{COLORTAGIN[Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of]COLORTAGOUT} \citealt{edgar1935irises,fisher1936use}).
+    \textcolor{blue}{COLORTAGIN[Morphospaces: different mathematical representations of a morphospace.
+A trait matrix can be an ordinated matrix (e.g.
+in }\citealt{tyler2011detecting}\textcolor{blue}{COLORTAGIN[ ) or transformed into a distance matrix (e.g.
+in }\citealt{Close2015}\textcolor{blue}{COLORTAGIN[, not represented here).}
+    \textcolor{blue}{COLORTAGIN[Here we consider all these matrices as being \textit{morphospaces}, i.e.
+objects containing all the combinations of traits and observations (albeit transformed differently).
     Visualisation: different  ways to represent the morphospace in 2D.
     Visualisations can use either trait plots (directly from the trait matrix); or ordination axis plots (directly from the ordinated matrix).
-    Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).}
+    Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).]COLORTAGOUT}
 }
@@
 
-\textcolor{blue}{\subsection{Summarising disparity using disparity indices} \label{section:metrics}}
+\textcolor{blue}{COLORTAGIN[\subsection{Summarising disparity using disparity indices}]COLORTAGOUT \label{section:metrics}}
 
@@
 
-When considering only one dimension, \textcolor{blue}{disparity indices can be used to compare the spread of distributions (e.g.
+When considering only one dimension, \textcolor{blue}{COLORTAGIN[disparity indices can be used to compare the spread of distributions (e.g.
 the range, quantiles or variance) or the differences in the central tendencies (i.e.
-mean, median or mode) of groups in the morphospace.}
+mean, median or mode) of groups in the morphospace.]COLORTAGOUT}
 Among these indices, some will have more attractive properties than others, such as sensitivity to outliers.
@@
 
-\textcolor{blue}{\subsection{Testing for differences in disparity} \label{section:testing}}
+\textcolor{blue}{COLORTAGIN[\subsection{Testing for differences in disparity} ]COLORTAGOUT\label{section:testing}}
 
@@
 One commonly used test is the non-parametric permutation analysis of variance (PERMANOVA) \citep{Anderson2001} that tests whether two groups share the same centroid and dispersion based on a distance matrix between observations.
-\textcolor{blue}{The last decade has also seen a series of developments based on this test (e.g.
+\textcolor{blue}{COLORTAGIN[The last decade has also seen a series of developments based on this test (e.g.
 the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{Adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
 It is worth noting that most of these test do not require the morphospace to be ordinated (see section \ref{section:ordination} above).
-Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.}
+Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.]COLORTAGOUT}
 
@@
 the sum of variances), bootstrapping the data generates a distribution of the index that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
-However, bootstrapped data are pseudoreplicates and thus non-independent \textcolor{blue}{and can increase the false positive rate (Type I error)} \citep{strube1988bootstrap}.
-\textcolor{blue}{This, again, highlights the importance of tailoring the statistical test to the data and question at hand.}
+However, bootstrapped data are pseudoreplicates and thus non-independent \textcolor{blue}{COLORTAGIN[and can increase the false positive rate (Type I error)]COLORTAGOUT} \citep{strube1988bootstrap}.
+\textcolor{blue}{COLORTAGIN[This, again, highlights the importance of tailoring the statistical test to the data and question at hand.]COLORTAGOUT}
 
@@
 It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first \textit{d} axes of the ordination, and can lead to incorrect interpretations of the data (such as wrongly supporting ``early burst'' type patterns; \citealt{Uyeda2015}).
-\textcolor{blue}{Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution.
+\textcolor{blue}{COLORTAGIN[Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution.
 Standard methods assume a model of Brownian motion, i.e.
@@
 In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{blomberg2020}.
-If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.}
+If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.]COLORTAGOUT}
 
-One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in \textcolor{blue}{disparity through time} analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny \citep{brusatte2011phylogenetic,Guillerme2018}.
+One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in \textcolor{blue}{COLORTAGIN[disparity through time]COLORTAGOUT} analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny \citep{brusatte2011phylogenetic,Guillerme2018}.
 Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e.
@@
 
-All ancestral state estimates are highly dependent on the data and method used \textcolor{blue}{(especially on the underlying model of trait evolution)} \cite{louca2020}.
+All ancestral state estimates are highly dependent on the data and method used \textcolor{blue}{COLORTAGIN[(especially on the underlying model of trait evolution)]COLORTAGOUT} \cite{louca2020}.
 In general, using ancestral state estimation can help with recovering patterns of change in disparity but should not be used simply to generate extra data points to increase statistical power.
@@
 This article results from discussion at the Royal Society International Science Seminar on `Reconciling disparate views on disparity' held at Chicheley Hall, 9-10th January, 2018.
-\textcolor{blue}{We thank the two anonymous reviewers for their helpful and positive comments.}
+\textcolor{blue}{COLORTAGIN[We thank the two anonymous reviewers for their helpful and positive comments.]COLORTAGOUT}
 AG was funded by European Research Council Starting grant 637171 ADaPTiVE;

### 2020-05-01 Writing/disparity_review.tex
finalised version

@@
 \textcolor{blue}{COLORTAGIN[The last decade has also seen a series of developments based on this test (e.g.
-the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{Adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
+the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
 It is worth noting that most of these test do not require the morphospace to be ordinated (see section \ref{section:ordination} above).

### 2020-05-05 Writing/pandoc.sh
compiler with color tags

@@ -1,1 +1,5 @@
-pandoc -o disparity_review.docx -i disparity_review.tex --bibliography References.bib --filter pandoc-citeproc --csl biologyletters.csl
+sed 's/textcolor{blue}{/textcolor{blue}{COLORTAGIN[/g' disparity_review.tex > tmp
+sed 's/}/]COLORTAGOUT}/g' tmp > compile.tex
+pandoc -o disparity_review.docx -i compile.tex --bibliography References.bib --filter pandoc-citeproc --csl biologyletters.csl
+rm compile.tex
+rm tmp

### 2020-05-05 Writing/response_to_reviewers.tex
Updated manuscript and response

\documentclass[12pt,letterpaper]{article}
\usepackage{natbib}

%Packages
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{pdflscape}
\usepackage{xcolor}

\usepackage{color,soul}

\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
\newcolumntype{d}[1]{D{.}{.}{#1}}

\pagenumbering{arabic}

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
% \setcounter{secnumdepth}{0} 
% \renewcommand{\section}[1]{%
% \bigskip
% \begin{center}
% \begin{Large}
% \normalfont\scshape #1
% \medskip
% \end{Large}
% \end{center}}
% \renewcommand{\subsection}[1]{%
% \bigskip
% \begin{center}
% \begin{large}
% \normalfont\itshape #1
% \end{large}
% \end{center}}
% \renewcommand{\subsubsection}[1]{%
% \vspace{2ex}
% \noindent
% \textit{#1.}---}
% \renewcommand{\tableofcontents}{}
%\bibpunct{(}{)}{;}{a}{}{,}

%---------------------------------------------
%
%       START
%
%---------------------------------------------

\begin{document}

\section{Response to Referees}

We thank the reviewers for their useful and insightful comments. We respond to them below in blue.

\subsection{Referee: 1}
\noindent Concerning the section on disparity through time, perhaps hyphens should be removed; also, it may be misleading to use them, as sometimes the disparity-through-time phrase refers, in abbreviated format, to what is otherwise known (in full) as "mean relative subclade disparity through time", implemented in the DTT function in geiger.

\textcolor{blue}{We have removed hyphens from ``disparity-through-time''}

\noindent At the beginning of page 7, I am not convinced that "synapomorphies will naturally lead an apparent shift or increase in disparity when new clades appear" in all cases. This very much depends on character-state distribution and, given a sufficient number of unique traits or combinations thereof across the sample, you may have comparable levels of disparity.

\textcolor{blue}{We changed the sentence to:}

\textcolor{blue}{\textit{especially as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).}} l.129-131 %@@@ lines

\noindent On page 8, just to avoid confusion, PCO can also applied to quantitative data ONLY. End of page 8/beginning of page 9 could do with a small (few lines) exposition of protocols for axis selection.

\textcolor{blue}{We've added that PCO can be used on dissimilarity matrices built using qualitative, or quantitative data, or mixed data types:}

\textcolor{blue}{\textit{dissimilarity matrices based on qualitative, quantitative or mixed data types can be reduced using PCO}} l.166-167 %@@@ lines

\textcolor{blue}{And added a brief mention of protocols for axis selection:}

\textcolor{blue}{\textit{(either by manually selecting the d axes that encompasses the desired cumulative variance or using methods such as the broken stick model; Legendre and Legendre 2012)}} l.188-190 %@@@ lines + author

\noindent Lines 211-213 are problematic. Do the authors mean that categorical data are ALWAYS more problematic because of the listed reasons? Some remedies at least exist for euclideanarity.

\textcolor{blue}{We've now stated that it is not always problematic and added a reference to the Cailliez (1983) correction to remedy for euclideanarity}:

\textcolor{blue}{\textit{This can be sometimes problematic when comparing the position of groups in multidimensional space, as the distances might not be linear (although this can sometimes be corrected; Cailliez 1983}} l.203-205 %@@@ lines + author

\noindent Line 269, about PERMANOVA, puzzles me. Yes, the test is used, but for geometric morphometric analyses at least, other tests have been devised.

\textcolor{blue}{We have added mention of other tests common in morphological disparity analyses:}

\textcolor{blue}{\textit{The last decade has also seen a series of developments based on this test (e.g. the linear regression for multidimensional data  Collyer et al. 2015 or the phylogenetic ANOVA Adams 2014; but see Adams and Collyer 2018; Lloyd 2016 for more.
It is worth noting that most of these test do not require the morphospace to be ordinated (see section 1 above).
Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.}} l.281-287 %@@@ lines + authors (3)

\subsection{Referee: 2}

\noindent That said, my biggest issue was the specific way the paper was subdivided. The "Measuring..", "Testing.." and "Phylogeny" subsections seemed a bit jumbled. For instance, a lot of the best advice on designing a custom test for your data is in the "Measuring.." section. Meanwhile, the "Testing.." section is very effective at convincing the reader not to take bootstrap values literally (a good thing to be effective at!) but doesn't contain much else in the way of advice or guidance.

\textcolor{blue}{We suspect this is probably due to us doing a poor job at explaining how these sections differ. In a standard disparity workflow, we tend to first extract disparity values for our chosen subsets of taxa which we refer to as "Measuring disparity". Following from this, some (though not all) studies go on to determine whether differences among subsets are statistically different. We refer to this step as "Testing hypotheses". To reduce confusion we have renamed these sections as 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity. We have retained the Disparity and phylogeny subsection as this is somewhat of a side issue that is important for these studies, but depending on the clade in question might be dealt with well or might be ignored due to lack of phylogenies for the group that could actually be used for these analyses.}

\textcolor{blue}{For the Testing section, we've added some references to specific tests and insisted throughout the section that it is important to choose the right statistical test according to the data and question at hand (and pointed to some general reviews detailing some of the options - see response to reviewer 1 above and response to specific comments below).}

\noindent The "Phylogeny" section makes no explicit reference to the models of trait evolution that exist. Ancestral state estimates are mentioned, but the potential pitfalls of imposing a (possibly, if not likely) incorrect model of trait evolution is elided. Elsewhere in the paper the authors do an excellent job warning the reader away from just accepting default options, or doing something just because "that's the way it's done". Yet here, the reader is left without knowledge that accepting the default BM model uncritically could cause major issues with both Revell-style multivariate corrections and all ASRs.

\textcolor{blue}{We've now added the following caveats to this section:}

\textcolor{blue}{\textit{Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution. Standard methods assume a model of Brownian motion, i.e. a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution. In many biological situations this model is not realistic, and different models of evolution should be considered Blomberg et al. 2020. If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.)}} l.313-321 %@@@ lines + author

\textcolor{blue}{\textit{All ancestral state estimates are highly dependent on the data and method used (especially on the underlying model of trait evolution).}} l.338-339 %@@@ lines

\noindent In the first two sections, the authors reference back to the four main goals of disparity analyses they outlined. Using those four goals as a guide post for reworking the Testing and Phylogeny subsections may help improve the flow and usability of those parts.

\textcolor{blue}{We appreciate the reviewer's suggestion here and tried hard to implement it. Unfortunately it led to the sections being even more confused as many of our points apply across the four types of disparity analysis. We also didn't feel we had stuck to this structure in the earlier sections, so the structure here didn't make as much sense as the reviewer assumed it might. We therefore reverted back to the original structure.}

\textcolor{blue}{If we understand this reviewer well, they suggest we partition the methodological section per type of disparity analysis (e.g. for descriptive disparity analysis, one could apply such and such tests, etc...). If this is the case, we unfortunately think that this will make the paper to prescriptive and probably more complex to read since methods can be applied to different types of analysis.}

\noindent Last bit: I think Figure 2 should be redesigned, personally. I'm not sure to whom it is targeted. The top part's clear, but I suspect the bottom part would be opaque to those who would most benefit from reading this paper. Right now, it shows every step of every option all at once. But I think focusing in on a single complex and general concept (eg, a clear visual explanation of ordination alone, or a clear representation of distance generation, or...) would serve the paper better.

\textcolor{blue}{We have now updated the figure to be just illustrate the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:}

\textcolor{blue}{\textit{Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of Edgar 1935; Fisher 1936).
Morphospaces: different mathematical representations of a morphospace. A trait matrix can be an ordinated matrix (e.g. in Tyler and Leighton 2011) or transformed into a distance matrix (e.g. in Close et al. 2015, not represented here).
Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
Visualisation: different  ways to represent the morphospace in 2D.
Visualisations can use either trait plots (directly from the trait matrix); or ordination axis plots (directly from the ordinated matrix).
Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).}} Figure 2. %@@@ authors (4)

\subsubsection{Specific Comments:}

\noindent Line 98: "[Disparity as a proxy for ecology] is particularly common in palaeobiology" I actually don't think this is true (at least, not in the way it is most likely to be read; see last paragraph). There are entire journals (eg, Functional Ecology) dedicated to neontological studies using variation in easily measured traits across space/time (disparity) as a proxy for changes in ecosystem functions \& species interactions (ecology). Plant physiologists in particular routinely use various measurable morphological features (stem diameters, root depth, etc) as proxies for various ecological features (e.g., aridity tolerance) and compare locations. 
Functional ecologists use different language, and rarely use the literal word "disparity", but the ideas are basically the same. The big differences I perceive are in terms of (1) how easy it is to test the reliability of the proxy and (2) how often workers even attempt to rigorously test that proxy. 
I think, especially given the specific citation here \& preceding sentence, that the authors mean something along the lines of "disparity-as-ecology is one of the primary ways paleobiologist have for investigating ecosystem function". That is, the "particularly common" is in reference to the large proportion of studies that use this approach in palaeobiology. It's just that, as-written, I think it makes this common approach sound like a "weird" palaeobiologist think.

\textcolor{blue}{This is a very fair observation. We've changed the sentence mentioning disparity and ecology and palaeobiology as follows:}

\textcolor{blue}{\textit{It is one of the primary ways to investigate ecosystem functioning in palaeobiology when the study species (and their functional characteristics) are extinct (Wainwright et al., 2005).}} l.80-82 %@@@ lines + author

\noindent Line 150 - 158: Nothing wrong or incorrect here, but the flow of this paragraph is odd. I find it very choppy. The first sentence is a long list of problems. The second is a set of two more problems. The third is a statement that everything has problems. The above three sentences, I think, could be reworked to flow more directly attached to the main point, which is that questions need to fit their data and vice versa.

\textcolor{blue}{We phrased the paragraph as shown below. Hopefully this has improved the flow to the main point.}

\textcolor{blue}{\textit{Ultimately, disparity analyses are characterised by the data they use. Unfortunately, trait data suffer from the same shortcomings as most biological datasets. The data within them can be non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, and/or correlated  (Palci and Lee, 2018). There are also issues of missing data, both where a particular character cannot be measured for a given taxon, or where a given taxon cannot be sampled at all. Trait data may also be influenced by biological phenomena such as allometry and sexual dimorphism. More practically, data collection is constrained by the time and money available, making collating a ``perfect'' dataset impossible. Even when care is taken, subsamples of the universe of possible data may not have the power to uncover the full patterns of disparity. These issues should be considered when collecting data. It is particularly important to collect trait data with the scientific question in mind, or, where there are limits on the data available, to tailor the question being asked to match the data.}} l.138-150 %@@@ lines + author

\noindent Line 171-172: Don't need e.g. and etc. in same parenthetical

\textcolor{blue}{We removed the  ``etc.''}

\noindent Lines 175 - 177: The discussion of dimension reduction using PCO here makes me think that more discussion in this paper of the importance of choosing an appropriate distance metric might be needed. That is, it'd be a whole separate paper trying to review or describe what is or is not appropriate! But I do think it's worth further emphasizing that it's a critical methodological choice that shouldn't be taken lightly for those intending to use MDS/NMDS

\textcolor{blue}{We've added the following sentence highlighting the effect of the chosen distance metric and referring to a great paper looking at it in more detail:}

\textcolor{blue}{\textit{Note that for PCO, the distance metric used can have significant impacts on the resulting morphospace (Lehmann et al., 2019). Choice of distance metric is therefore crucial, and should not be overlooked when using PCO.}} l.170-173 %@@@lines + author

\noindent Line 195: I think you mean explicitly "geometric morphometric data". If that's true, being explicit is, I think, better.

\textcolor{blue}{We've added the term ``geometric'' to be more explicit.}

\noindent Lines 234 - 239: This implies that a central tendency is, on its own, a measure of disparity. Below, in the discussion of the position within morphospace of a clade, I think this point is made clearer. But I think the first few sentences here should be clarified a bit. Something like "or differences in the central tendencies (mean/median/mode) of clades [or subsets to be more general]"

\textcolor{blue}{We've changed the sentence to:}

\textcolor{blue}{\textit{disparity indices can be used to compare the spread of distributions (e.g. the range, quantiles or variance) or the differences in the central tendencies (i.e. mean, median or mode) of groups in the morphospace.}} l.237-239 %@@@ lines

\textcolor{blue}{The following three points by the reviewer are closely linked so we deal with them together}

\noindent Lines 263 - 289: The "Testing hypotheses..." section. I feel as though this section needs to be fleshed out. A lot of very good points are made, especially with respect to bootstrapping, but a reader will walk away from this section (1) knowing that PERMANOVA exists, (2) being a bit skeptical of bootstrap support, and (3) not much else. The last paragraph is, I think, the most important in this section, and I would recommend cutting from the other two to expand this.

\noindent The general point about tailoring analyses to your data is sound, and that does make general advice hard, but not impossible. A few clear principles or even just an explanation of "same" morphospace could be useful here. That is, up in the Measuring subsection in lines 239 - 249 the authors provide a lot of good sound advice that seems more fitting here.

\noindent I'm just imagining a student who reads this paper early in their project. Then later, once they have data, tries to go back to look up some best practices and misses some good tips because they were in the "Measuring..." subsection instead of the "Testing..." subsection. It may be that the subsections should simply be reworked.

\textcolor{blue}{In response to this, and comments by reviewer 1, we have fleshed out the "Testing" section to mention other existing tests. We have also, as suggested by the reviewer, shortened the other paragraphs and expanded this section. This is a difficult section to write because the possible statistical tests heavily rely on the kind of data available and the question being asked. In many studies only visual observations of differences are made without statistical testing.}

\textcolor{blue}{We appreciate the reviewer's point about inattentive readers missing key sections. We hope that we have solved this in part by renaming these sections to more accurately represent their content (changed to 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity). As mentioned above, we tried reshuffling some of this material, but the current structure best suits the way most disparity analyses are planned and carried out (in our experience).}

\textcolor{blue}{We avoid giving general advice because the idea with this paper was to inform but to be as non-prescriptive as possible (our take home message after the meeting at the origin of the paper was: ``do whatever disparity analysis you want but do it for the right reasons'') so we already had to tone down some prescriptive aspects in previous draft versions of this paper. We have however, highlighted the importance of applying the appropriate test to for the question and data twice in this section to really emphasize the point.}

\noindent Disparity \& Phylogeny section: This section is good, but perhaps overlong. Although, it doesn't discuss (what I perceive to be) the biggest problem with phylogenetic corrections to disparity analyses: the choice of trait evolution model.
That is, "correcting" an ordination by imposing a specific model (e.g., Brownian) might cause more trouble than it solves if that model is a poor fit to the data.

\textcolor{blue}{We agree this is a really important point that we have not expanded on previously du to worries about the word count. We've now added a caveat to this section, as mentioned in the main comments:}

\textcolor{blue}{\textit{Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution. Standard methods assume a model of Brownian motion, i.e. a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution. In many biological situations this model is not realistic, and different models of evolution should be considered Blomberg et al. 2020. If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.)}} l.313-321 %@@@ lines + author

\noindent This is especially relevant as the discussion of Ancestral state estimation, which is entirely reliant on the model of trait reconstruction (as is briefly noted in line 315).

\textcolor{blue}{We've also added the following to emphasize this point:}

\textcolor{blue}{\textit{All ancestral state estimates are highly dependent on the data and method used (especially on the underlying model of trait evolution).}} l.338-339 %@@@ lines

\end{document}

### 2020-05-05 Writing/disparity_review.tex
Updated manuscript and response

--- 2020-05-01 Writing/disparity_review.tex
@@
 
-\section{Response to Referees}
-
-We thank the reviewers for their useful and insightful comments. We respond to them below in blue.
-
-\subsection{Referee: 1}
-\noindent Concerning the section on disparity through time, perhaps hyphens should be removed; also, it may be misleading to use them, as sometimes the disparity-through-time phrase refers, in abbreviated format, to what is otherwise known (in full) as "mean relative subclade disparity through time", implemented in the DTT function in geiger.
-
-\textcolor{blue}{We have removed hyphens from ``disparity-through-time''}
-
-\noindent At the beginning of page 7, I am not convinced that "synapomorphies will naturally lead an apparent shift or increase in disparity when new clades appear" in all cases.
-This very much depends on character-state distribution and, given a sufficient number of unique traits or combinations thereof across the sample, you may have comparable levels of disparity.
-
-\textcolor{blue}{We changed the sentence to:}
-
-\textit{especially as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).} l.@@@
-
-\noindent On page 8, just to avoid confusion, PCO can also applied to quantitative data ONLY.
-End of page 8/beginning of page 9 could do with a small (few lines) exposition of protocols for axis selection.
-
-\textcolor{blue}{We've added that PCO can be used on dissimilarity matrices built using qualitative, or quantitative data, or mixed data types:}
-
-\textit{dissimilarity matrices based on qualitative, quantitative or mixed data types can be reduced using PCO} l.@@@
-
-\textcolor{blue}{And added a brief mention of protocols for axis selection:}
-
-\textit{(either by manually selecting the d axes that encompasses the desired cumulative variance or using methods such as the broken stick model; Legendre \& Legendre 2012 p.410)} l.@@@ + change cite to number
-
-\noindent Lines 211-213 are problematic. Do the authors mean that categorical data are ALWAYS more problematic because of the listed reasons? Some remedies at least exist for euclideanarity.
-
-% NC: Not sure if this really answers the issue they had. Do we mean that categorical are always more problematic???
-
-\textcolor{blue}{We've now added a reference to the Cailliez (1983) correction to remedy for euclideanarity}:
-
-\textit{although this can sometimes be corrected, Cailliez 1983} l.@@@
-
-\noindent Line 269, about PERMANOVA, puzzles me. Yes, the test is used, but for geometric morphometric analyses at least, other tests have been devised.
-
-\textcolor{blue}{We have added mentions of other tests common in morphological disparity analyses:}
-
-\textit{The last decade has also seen a series of developments based on this test (e.g.
-the linear regression for multidimensional data Collyer et al 2015; or the phylogenetic ANOVA Adams 2014; but see Adams \& Collyer 2018; Lloyd 2016 for more).
-It is worth noting that most of these test do not require the morphospace to be ordinated (see section 1 above).
-Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.} l.@@@
-
-\subsection{Referee: 2}
-
-\noindent That said, my biggest issue was the specific way the paper was subdivided.
-The "Measuring..", "Testing.." and "Phylogeny" subsections seemed a bit jumbled.
-For instance, a lot of the best advice on designing a custom test for your data is in the "Measuring.." section.
-Meanwhile, the "Testing.." section is very effective at convincing the reader not to take bootstrap values literally (a good thing to be effective at!) but doesn't contain much else in the way of advice or guidance.
-
-\textcolor{blue}{We suspect this is probably due to us doing a poor job at explaining how these sections differ.
-In a standard disparity workflow, we tend to first extract disparity values for our chosen subsets of taxa which we refer to as "Measuring disparity".
-Following from this, some (though not all) studies go on to determine whether differences among subsets are statistically different.
-We refer to this step as "Testing hypotheses".
-To reduce confusion we have renamed these sections as 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity.
-We have retained the Disparity and phylogeny subsection as this is somewhat of a side issue that is important for these studies, but depending on the clade in question might be dealt with well or might be ignored due to lack of phylogenies for the group that could actually be used for these analyses.}
-
-\textcolor{blue}{For the Testing section, we've added some references to specific tests and insisted throughout the section that it is important to chose the right statistical test according to the data and question at hand (and pointed to some general reviews detailing some of the options - see response to reviewer 1 above and response to specific comments below).}
-% TG: Not sure if that's enough but lacking inspiration for now
-% NC: I think the error the reviewer made is what I've written above, so changing the names should deal with it.
-
-\noindent The "Phylogeny" section makes no explicit reference to the models of trait evolution that exist.
-Ancestral state estimates are mentioned, but the potential pitfalls of imposing a (possibly, if not likely) incorrect model of trait evolution is elided.
-Elsewhere in the paper the authors do an excellent job warning the reader away from just accepting default options, or doing something just because "that's the way it's done".
-Yet here, the reader is left without knowledge that accepting the default BM model uncritically could cause major issues with both Revell-style multivariate corrections and all ASRs.
-
-\textcolor{blue}{We've now added the following caveats to this section:}
-
-\textit{Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution.
-Standard methods assume a model of Brownian motion, i.e.
-a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution.
-In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{blomberg2020}.
-If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.)} .
-l.@@@
-
-\textit{All ancestral state estimates are highly dependent on the data and method used (especially on the underlying model of trait evolution)}
-
-\noindent In the first two sections, the authors reference back to the four main goals of disparity analyses they outlined.
-Using those four goals as a guide post for reworking the Testing and Phylogeny subsections may help improve the flow and usability of those parts.
-
-\textcolor{blue}{We appreciate the reviewer's suggestion here and tried hard to implement it.
-Unfortunately it led to the sections being even more confused as many of our points apply across the four types of disparity analysis.
-We also didn't feel we had stuck to this structure in the earlier sections, so the structure here didn't make as much sense as the reviewer assumed it might.
-We therefore reverted back to the original structure.}
-
-% TG: not sure what's meant here
-% NC: What they mean is can we use the 4 types of disparity analysis from the intro to say: For this type of analysis you should do this test, for this type do this test. But I don't think we can...
-
-\noindent Last bit: I think Figure 2 should be redesigned, personally.
-I'm not sure to whom it is targeted.
-The top part's clear, but I suspect the bottom part would be opaque to those who would most benefit from reading this paper.
-Right now, it shows every step of every option all at once.
-But I think focusing in on a single complex and general concept (eg, a clear visual explanation of ordination alone, or a clear representation of distance generation, or...) would serve the paper better.
-
-\textcolor{blue}{We have now updated the figure to be just illustrate the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:}
-
-\textit{Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of \citealt{edgar1935irises,fisher1936use}).
-Morphospaces: different mathematical representations of a morphospace.
-A trait matrix can be an ordinated matrix (e.g.
-in \citealt{tyler2011detecting} - or transformed into a distance matrix \citealt{Close2015}, not represented here).
-% NC: Not sure the brackets are in the right places here?
-As is the sentence reads "A trait matrix can be an ordinated matrix (....)."
-Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
-Visualisation: different  ways to represent the morphospace in 2D.
-Visualisations can use either trait plots (directly from the trait matrix); or ordination axis plots (directly from the ordinated matrix).
-Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).}
-
-\subsubsection{Specific Comments:}
-
-\noindent Line 98: "[Disparity as a proxy for ecology] is particularly common in palaeobiology" I actually don't think this is true (at least, not in the way it is most likely to be read; see last paragraph).
-There are entire journals (eg, Functional Ecology) dedicated to neontological studies using variation in easily measured traits across space/time (disparity) as a proxy for changes in ecosystem functions \& species interactions (ecology).
-Plant physiologists in particular routinely use various measurable morphological features (stem diameters, root depth, etc) as proxies for various ecological features (e.g., aridity tolerance) and compare locations.
-Functional ecologists use different language, and rarely use the literal word "disparity", but the ideas are basically the same.
-The big differences I perceive are in terms of (1) how easy it is to test the reliability of the proxy and (2) how often workers even attempt to rigorously test that proxy.
-I think, especially given the specific citation here \& preceding sentence, that the authors mean something along the lines of "disparity-as-ecology is one of the primary ways paleobiologist have for investigating ecosystem function".
-That is, the "particularly common" is in reference to the large proportion of studies that use this approach in palaeobiology.
-It's just that, as-written, I think it makes this common approach sound like a "weird" palaeobiologist think.
-
-\textcolor{blue}{This is a very fair observation. We've changed the sentence mentioning disparity and ecology and palaeobiology as follows:}
-
-\textit{It is one of the primary ways to investigate ecosystem functioning in palaeobiology when the study species (and their functional characteristics) are extinct \citep{Wainwright2005}} l.@@@
-
-\noindent Line 150 - 158: Nothing wrong or incorrect here, but the flow of this paragraph is odd.
-I find it very choppy.
-The first sentence is a long list of problems.
-The second is a set of two more problems.
-The third is a statement that everything has problems.
-The above three sentences, I think, could be reworked to flow more directly attached to the main point, which is that questions need to fit their data and vice versa.
-
-\textcolor{blue}{We phrased the paragraph as shown below. Hopefully this has improved the flow to the main point.}
-
-\textit{Ultimately, disparity analyses are characterised by the data they use.
-Unfortunately, trait data suffer from the same shortcomings as most biological datasets.
-The data within them can be non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, and/or correlated \citep{Palci2018}.
-There are also issues of missing data, both where a particular character cannot be measured for a given taxon, or where a given taxon cannot be sampled at all.
-Trait data may also be influenced by biological phenomena such as allometry and sexual dimorphism.
-More practically, data collection is constrained by the time and money available, making collating a ``perfect'' dataset impossible.
-Even when care is taken, subsamples of the universe of possible data may not have the power to uncover the full patterns of disparity.
-These issues should be considered when collecting data.
-It is particularly important to collect trait data with the scientific question in mind, or, where there are limits on the data available, to tailor the question being asked to match the data.
-}
-
-\noindent Line 171-172: Don't need e.g. and etc. in same parenthetical
-
-\textcolor{blue}{We removed the  ``etc.''}
-
-\noindent Lines 175 - 177: The discussion of dimension reduction using PCO here makes me think that more discussion in this paper of the importance of choosing an appropriate distance metric might be needed.
-That is, it'd be a whole separate paper trying to review or describe what is or is not appropriate!
-But I do think it's worth further emphasizing that it's a critical methodological choice that shouldn't be taken lightly for those intending to use MDS/NMDS
-
-\textcolor{blue}{We've added the following sentence highlighting the effect of the chosen distance metric and referring to a great paper looking at it in more detail:}
-
-\textit{Note that for PCO, the distance metric used can have significant impacts on the resulting morphospace \citep{lehmann2019}.
-Choice of distance metric is therefore crucial, and should not be overlooked when using PCO.}.
-l.@@@
-
-\noindent Line 195: I think you mean explicitly "geometric morphometric data". If that's true, being explicit is, I think, better.
-
-\textcolor{blue}{We've added the term ``geometric'' to be more explicit.}
-
-\noindent Lines 234 - 239: This implies that a central tendency is, on its own, a measure of disparity.
-Below, in the discussion of the position within morphospace of a clade, I think this point is made clearer.
-But I think the first few sentences here should be clarified a bit.
-Something like "or differences in the central tendencies (mean/median/mode) of clades [or subsets to be more general]"
-
-\textcolor{blue}{We've changed the sentence to:}
-
-\textit{disparity indices can be used to compare the spread of distributions (e.g.
-the range, quantiles or variance) or the differences in the central tendencies (i.e.
-mean, median or mode) of groups in the morphospace.} l.@@@
-
-\textcolor{blue}{The following three points by the reviewer are closely linked so we deal with them together}
-
-\noindent Lines 263 - 289: The "Testing hypotheses..." section.
-I feel as though this section needs to be fleshed out.
-A lot of very good points are made, especially with respect to bootstrapping, but a reader will walk away from this section (1) knowing that PERMANOVA exists, (2) being a bit skeptical of bootstrap support, and (3) not much else.
-The last paragraph is, I think, the most important in this section, and I would recommend cutting from the other two to expand this.
-
-\noindent The general point about tailoring analyses to your data is sound, and that does make general advice hard, but not impossible.
-A few clear principles or even just an explanation of "same" morphospace could be useful here.
-That is, up in the Measuring subsection in lines 239 - 249 the authors provide a lot of good sound advice that seems more fitting here.
-
-\noindent I'm just imagining a student who reads this paper early in their project.
-Then later, once they have data, tries to go back to look up some best practices and misses some good tips because they were in the "Measuring..." subsection instead of the "Testing..." subsection.
-It may be that the subsections should simply be reworked.
-
-\textcolor{blue}{In response to this, and comments by reviewer 1, we have fleshed out the "Testing" section to mention other existing tests.
-We have also, as suggested by the reviewer, shortened the other paragraphs and expanded this section.
-This is a difficult section to write because the possible statistical tests heavily rely on the kind of data available and the question being asked.
-In many studies only visual observations of differences are made without statistical testing.
-The edited text is as follows:}
-
-% add the editted text here
-
-\textcolor{blue}{We appreciate the reviewer's point about inattentive readers missing key sections.
-We hope that we have solved this in part by renaming these sections to more accurately represent their content (changed to 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity).
-As mentioned above, we tried reshuffling some of this material, but the current structure best suits the way most disparity analyses are planned and carried out (in our experience).}
-
-\textcolor{blue}{We avoid giving general advice because the idea with this paper was to inform but to be as non-prescriptive as possible (our take home message after the meeting at the origin of the paper was: ``do whatever disparity analysis you want but do it for the right reasons'') so we already had to tone down some prescriptive aspects in previous draft versions of this paper.
-We have however, highlighted the importance of applying the appropriate test to for the question and data twice in this section to really emphasize the point.}
-
-% add edited text here
-
-\noindent Disparity \& Phylogeny section: This section is good, but perhaps overlong.
-Although, it doesn't discuss (what I perceive to be) the biggest problem with phylogenetic corrections to disparity analyses: the choice of trait evolution model.
-That is, "correcting" an ordination by imposing a specific model (e.g., Brownian) might cause more trouble than it solves if that model is a poor fit to the data.
-
-\textcolor{blue}{We agree this is a really important point that we have not expnaded on previously du to worries about the word count.
-We've now added a caveat to this section, as mentioned in the main comments:}
-
-\textit{Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution.
-Standard methods assume a model of Brownian motion, i.e.
-a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution.
-In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{blomberg2020}.
-If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.)} .
-l.@@@
-
-\noindent This is especially relevant as the discussion of Ancestral state estimation, which is entirely reliant on the model of trait reconstruction (as is briefly noted in line 315).
-
-\textcolor{blue}{We've also added the following to emphasize this point:}
-
-\textit{All ancestral state estimates are highly dependent on the data and method used (especially on the underlying model of trait evolution)}
-
 %Running head
@@
 Graeme T. Lloyd$^{10}$,
-Joseph E. O'Reilly$^{11}$, %was 17 - 11 forward + 1
+Joseph E. O'Reilly$^{11,12}$,
 Abi Pate$^{12}$,
-Mark N. Puttick$^{13}$,
+Mark N. Puttick$^{12,13}$,
 Emily Rayfield$^{12}$,
@@
 \noindent{
+\scriptsize{
 $^{1}$ School of Biological Sciences, University of Queensland, St. Lucia,
@@
 $^{18}$ College of Science and Engineering, Flinders University,
-Adelaide SA5042, Australia;\\
+Adelaide SA5042, Australia;
 $^{18}$ Department of Animal and Plant Sciences, The University of Sheffield,
-Sheffield S10 2TN, UK;\\
+Sheffield S10 2TN, UK;}
+
 $^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding
-author: guillert@tcd.ie}
-
+author: guillert@tcd.ie
+}
 \end{center}
+\newpage
 
@@
 Here we review the most commonly used approaches for characterising and analysing morphological disparity, all of which have associated limitations that, if ignored, can lead to misinterpretation.
-We provide best practice guidelines for disparity analyses, while noting that there can be no ``one-size-fits-all'' approach.
+We propose best practice guidelines for disparity analyses, while noting that there can be no ``one-size-fits-all'' approach.
 The available tools should always be used in the context of a specific biological question that will determine data and method selection at every stage of the analysis.
@@
 
-Ultimately, disparity analyses are characterised by the data they use.
+\textcolor{blue}{Ultimately, disparity analyses are characterised by the data they use.
 Unfortunately, trait data suffer from the same shortcomings as most biological datasets.
@@
 These issues should be considered when collecting data.
-It is particularly important to collect trait data with the scientific question in mind, or, where there are limits on the data available, to tailor the question being asked to match the data.
+It is particularly important to collect trait data with the scientific question in mind, or, where there are limits on the data available, to tailor the question being asked to match the data.}
 
@@
 Ordinations are statistical methods that map observed variables onto a new space of reduced dimension while maintaining the requirement that similar observations are closer together than dissimilar ones (e.g.
-principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS).
+principal component analysis - PCA; principal coordinates analysis - PCO; non-metric multidimensional scaling - NMDS).
 They come in many flavours depending on the data and the desired morphospace properties.
@@
 some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
-This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear \textcolor{blue}{(although this can sometimes be corrected \citealt{cailliez1983})}.
+\textcolor{blue}{This can be sometimes problematic when comparing the position of groups in multidimensional space, as the distances might not be linear (although this can sometimes be corrected  \citealt{cailliez1983})}.
 Furthermore, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible \citep{Gerber2019}.
@@
 \centering
-   \includegraphics[width=0.9\textwidth]{Figures/figure_space2.pdf}
+   \includegraphics[width=0.7\textwidth]{Figures/figure_space2.pdf}
 \caption{
@@
     \textcolor{blue}{Morphospaces: different mathematical representations of a morphospace.
-A trait matrix can be ordinated matrix (e.g.
-in }\citealt{tyler2011detecting}\textcolor{blue}{ - or transformed into a distance matrix }\citealt{Close2015}\textcolor{blue}{, not represented here).}
+A trait matrix can be an ordinated matrix (e.g.
+in }\citealt{tyler2011detecting}\textcolor{blue}{ ) or transformed into a distance matrix (e.g.
+in }\citealt{Close2015}\textcolor{blue}{, not represented here).}
     \textcolor{blue}{Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
@@
 \textcolor{blue}{The last decade has also seen a series of developments based on this test (e.g.
-the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{Adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
+the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
 It is worth noting that most of these test do not require the morphospace to be ordinated (see section \ref{section:ordination} above).
@@
 
-All ancestral state estimates are highly dependent on the data and method used \textcolor{blue}{(especially on the underlying model of trait evolution)} \cite{louca2020}.
+\textcolor{blue}{All ancestral state estimates are highly dependent on the data and method used (especially on the underlying model of trait evolution)} \cite{louca2020}.
 In general, using ancestral state estimation can help with recovering patterns of change in disparity but should not be used simply to generate extra data points to increase statistical power.
@@
 As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse them, stemming from the explicit hypotheses being tested.
-Although this makes comparison between disparity analyses difficult and makes generalisation required to answer the general biological questions from the introduction still premature (e.g.
-how does phenotypic variation evolves?), this diversity of methodological approaches provides researchers with a great number of tools tailored to answer specific biological questions.
+\textcolor{blue}{Although this makes comparison between disparity analyses difficult and renders premature attempts to achieve the generalisation required to answer the broad biological questions (e.g.
+how does phenotypic variation evolves?), this diversity of methodological approaches provides researchers with a great number of tools tailored to answer specific biological questions.}
 

### 2020-05-12 Writing/response_to_reviewers.tex
final revised version!

@@
 
-We thank the reviewers for their useful and insightful comments. We respond to them below in blue.
+We thank the reviewers for their useful and insightful comments. We respond to them below in \textcolor{blue}{blue}.
 
@@
 
-\textcolor{blue}{\textit{especially as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).}} l.129-131 %@@@ lines
+\textcolor{blue}{\textit{especially as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (particularly if the character-state distribution is skewed towards a particular clade).}} l.129-131 %@@@ lines
 
@@
 
-\textcolor{blue}{\textit{This can be sometimes problematic when comparing the position of groups in multidimensional space, as the distances might not be linear (although this can sometimes be corrected; Cailliez 1983}} l.203-205 %@@@ lines + author
+\textcolor{blue}{\textit{
+This can sometimes problematic be when comparing the position of groups in multidimensional space, as true dissimilarities might not be reliably conveyed (although this can sometimes be improved; Cailliez 1983)}} l.203-205 %@@@ lines + author
 
@@
 the linear regression for multidimensional data  Collyer et al.
-2015 or the phylogenetic ANOVA Adams 2014; but see Adams and Collyer 2018; Lloyd 2016 for more.
-It is worth noting that most of these test do not require the morphospace to be ordinated (see section 1 above).
+2015 or the phylogenetic ANOVA Adams 2014; but see Adams and Collyer 2018; Lloyd 2016 for more.)
+It is worth noting that most of these tests do not require the morphospace to be ordinated (see section 1 above).
 Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.}} l.281-287 %@@@ lines + authors (3)
@@
 To reduce confusion we have renamed these sections as 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity.
-We have retained the Disparity and phylogeny subsection as this is somewhat of a side issue that is important for these studies, but depending on the clade in question might be dealt with well or might be ignored due to lack of phylogenies for the group that could actually be used for these analyses.}
+We have retained the Disparity and phylogeny section as this is a separate issue that, while important for particular studies, is not always included in disparity analysis of groups without appropriate phylogentic data.}
 
@@
 a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution.
-In many biological situations this model is not realistic, and different models of evolution should be considered Blomberg et al.
+In many biological situations this model is not realistic, and different models of evolution should be considered; Gerber 2014, Blomberg et al.
 2020.
@@
 
-\textcolor{blue}{If we understand this reviewer well, they suggest we partition the methodological section per type of disparity analysis (e.g.
+\textcolor{blue}{If we understand the reviewer well, they suggest we partition the methodological section per type of disparity analysis (e.g.
 for descriptive disparity analysis, one could apply such and such tests, etc...).
-If this is the case, we unfortunately think that this will make the paper to prescriptive and probably more complex to read since methods can be applied to different types of analysis.}
+If this is the case, we unfortunately think that this will make the paper too prescriptive and probably more complex to read since methods can be applied to different types of analysis.}
 
@@
 
-\textcolor{blue}{We phrased the paragraph as shown below. Hopefully this has improved the flow to the main point.}
+\textcolor{blue}{We phrased the paragraph as shown below. We feel this has improved the flow to the main point.}
 
@@
 
-\textcolor{blue}{We've added the term ``geometric'' to be more explicit.}
+\textcolor{blue}{We've changed the sentence to be more general to:}
+
+\textcolor{blue}{\textit{Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces might not possess the Euclidean properties one often intuitively assumes}} l.170-173 %@@@lines + author
 
@@
 
-\textcolor{blue}{We agree this is a really important point that we have not expanded on previously du to worries about the word count.
+\textcolor{blue}{We agree this is a really important point that we have not expanded on previously due to worries about the word count.
 We've now added a caveat to this section, as mentioned in the main comments:}

### 2020-05-12 Writing/disparity_review.tex
final revised version!

@@
 	\item \textbf{\textcolor{blue}{Disparity through time}.} This approach investigates how the morphologies of organisms have changed over time, by focussing on the disparity of taxa in particular time intervals or slices.
-	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{prentice2011evolution, Guillerme2018}, or how does disparity change up to and across mass extinction events\citep{Friedman2010}?
+	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{Brusatte2008, prentice2011evolution, Guillerme2018}, or how does disparity change up to and across mass extinction events\citep{Friedman2010}?
 
@@
 \citealt{sherratt2017rates}), Fourier coefficients (e.g.
-\citealt{Foote1995, Spriggs2018}) or model-based descriptors (e.g.
+\citealt{Foote1989, Spriggs2018}) or model-based descriptors (e.g.
 \citealt{Raup1961,Saunders2004} - Fig.
@@
 Discrete characters are commonly recycled from phylogenetic studies (e.g. \citealt{Brusatte2008,Close2015}).
-This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups, since phylogenetic characters are often collected to discriminate among groups.
-This needs to be considered when interpreting results, especially \textcolor{blue}{as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).}
+This approach may artifactually increase disparity between phylogenetically distinct groups, since phylogenetic characters are often collected to discriminate among groups.
+This needs to be considered when interpreting results, especially \textcolor{blue}{as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (particularly if the character-state distribution is skewed towards a particular clade).}
 Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017}.
@@
 
-\subsection{To ordinate or not to ordinate - that is the (multidimensional) question} \label{section:ordination}
+\subsection{To ordinate or not to ordinate? That is the (multidimensional) question} \label{section:ordination}
 
-Disparity analyses often use ordination techniques for dimension reduction.
+Disparity analyses often use ordination techniques for dimensionality reduction.
 Ordinations are statistical methods that map observed variables onto a new space of reduced dimension while maintaining the requirement that similar observations are closer together than dissimilar ones (e.g.
@@
 selecting only those axes that describe the majority of the variation in the dataset - e.g. 95\%).
-In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{Legendre2012,dryden2016statistical}.
+In the case of geometric morphometric data, some ordination techniques (e.g.
+PCA) can be particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions \citep{dryden2016statistical}.
 In practice, this facilitates interpretation of only the major axis of a highly dimensional dataset as major gradients of biological variation (e.g.
@@
 In the case of ordination, subsampling axes from the ordination can lead to misinterpretation of the results.
-Although a common technique is to consider the \textit{d} axes that encompass 95\% or 99\% of the variance in the dataset \textcolor{blue}{(either by manually selecting the \textit{d} axes that encompasses the desired cumulative variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410)}, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
-Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing \textcolor{blue}{geometric} morphometric data \citep{Deline2018, Gerber2017}.
+Although a common technique is to consider the \textit{d} axes that encompass 95\% or 99\% of the variance in the dataset \textcolor{blue}{(either by manually selecting the \textit{d} axes that encompasses the desired cumulative variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410)}, the interpretation of these principal axes can \textcolor{blue}{miss some aspects of the structure of the data} and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
+Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces 
+\textcolor{blue}{might not possess the Euclidean properties one often intuitively assumes} \citep{Deline2018, Gerber2017}.
 
@@
 Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
-For example, when the analysed data are non-Euclidean (e.g.
-some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
-\textcolor{blue}{This can be sometimes problematic when comparing the position of groups in multidimensional space, as the distances might not be linear (although this can sometimes be corrected  \citealt{cailliez1983})}.
+For example, when the analysed \textcolor{blue}{dissimilarity} data are non-Euclidean (e.g.
+\textcolor{blue}{as induced for instance with inapplicable characters in discrete character schemes}), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
+\textcolor{blue}{This can sometimes problematic be when comparing the position of groups in multidimensional space, as true dissimilarities might not be reliably conveyed (although this can sometimes be improved \citealt{cailliez1983})}.
 Furthermore, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible \citep{Gerber2019}.
@@
 
-In general, categorical data are a good deal more problematic than continuous data, because the characters themselves are invariably non-equivalent, non-independent, and mostly non-Euclidean, and the distribution of the variance is usually more evenly distributed across axes (i.e.
+In general, categorical data are a good deal more problematic than continuous data, because the characters themselves are invariably non-equivalent, non-independent, and the distribution of the variance is usually more evenly distributed across axes (i.e.
 contrary to a PCA, the first few axes do not encompass most variance of the dataset).
 Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some dimensions are not actually straight and, character coding and missing data can make the pairwise dissimilarity matrix lose its metric properties (i.e.
-the distance between A and B is not equal to the distance between B and A; \citealt{Gerber2014}).
+the distance between A and B is not equal to the distance between B and A; \citealt{Gerber2019}).
 Last, but not least, in many cases, ordination might not be necessary.
@@
 Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g.
-does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
+does group A occupy the same region of morphospace as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
 Finally, if we aim to characterise the density of morphospace occupation (e.g.
@@
 In addition to considering which properties of disparity these indices capture, it is also important to consider the mathematical properties of the indices and their associated caveats \citep{Wills2001, Ciampaglio2001}.
-For example, measuring the sum of variance for each dimension of the space is equivalent to measuring the sum of variance in a PCA (because the co-variance is null).
+For example, measuring the sum of variance for each dimension of the space before or after ordination via PCA is equivalent.
 However, this is no longer true of other transformations of the space or when a subset of dimensions or elements are considered, such as is often done after PCA \citep{Legendre2012}.
@@
 the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
-It is worth noting that most of these test do not require the morphospace to be ordinated (see section \ref{section:ordination} above).
+It is worth noting that most of these tests do not require the morphospace to be ordinated (see section \ref{section:ordination} above).
 Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.}
@@
 a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution.
-In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{blomberg2020}.
+In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{Gerber2014, blomberg2020}.
 If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.}
@@
 However, once the morphospace is defined, its properties will not change.
-Post-transformation ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \citep{lloyd2018}.
+Post-transformation ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \citep{lloyd2018,Gerber2019}.
 

### 2020-06-12 Analysis/Figures_in_R.Rmd
added penguin dataset

--- 2018-11-19 Analysis/Figures_in_R.Rmd
@@
 ## Creating a 5 trait space
-iris_rand <- cbind(iris[seq(from = 1, to = 150, by = 5), 1:4], rnorm(30))
+penguins_sub_rand <- cbind(penguins_sub[seq(from = 1, to = 150, by = 5), 1:4], rnorm(30))
 
@@
 pdf(file = "../Writing/Figures/figure_methods_biplot.pdf", width = 5, height = 5)
-plot(iris_rand, pch = 19, xaxt = "n", yaxt = "n")
+plot(penguins_sub_rand, pch = 19, xaxt = "n", yaxt = "n")
 dev.off()
@@
 ## Making a distance matrix
-iris_dist <- dist(iris_rand)
+penguins_sub_dist <- dist(penguins_sub_rand)
 
@@
 pdf(file = "../Writing/Figures/figure_methods_2dPCO.pdf", width = 5, height = 5)
-plot(cmdscale(iris_dist, k = 2), pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
+plot(cmdscale(penguins_sub_dist, k = 2), pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
 dev.off()
@@
 ## Making a PCA
-pca <- princomp(iris_rand)$scores
+pca <- princomp(penguins_sub_rand)$scores
 pdf(file = "../Writing/Figures/figure_methods_PCA1.pdf", width = 5, height = 5)
@@
 ```
+
+## Iris example
+
+```{r}
+set.seed(42)
+## Preparing the dataset
+library(palmerpenguins)
+
+## Sorting the dataset into the correct format (to look like penguins_sub: 150*4)
+penguins_sub <- penguins
+class(penguins_sub) <- "data.frame"
+penguins_sub <- penguins_sub[,c(1, 3:6)]
+
+## Removing NAs
+nas <- apply(penguins_sub[,-1], 1, function(x) any(is.na(x)))
+penguins_sub <- penguins_sub[!nas, ]
+
+## Selecting only 50 specimens per species
+selected <- numeric()
+for(species in as.character(unique(penguins_sub[,1]))) {
+    specimens <- which(penguins_sub[,1] == species)
+    selected <- c(selected, sample(specimens, 50))
+}
+penguins_sub <- penguins_sub[selected, ]
+
+## Z transform the data
+# penguins_sub[,c(2:5)] <- apply(penguins_sub[, c(2:5)], 1, scale)
+
+## Rownames
+rownames(penguins_sub) <- c(paste0(unique(penguins_sub[,1])[[1]], 1:50),
+                            paste0(unique(penguins_sub[,1])[[2]], 1:50),
+                            paste0(unique(penguins_sub[,1])[[3]], 1:50))
+penguins_sp <- penguins_sub[, 1]
+penguins_sub <- penguins_sub[,-1]
+
+## Colnames 
+colnames(penguins_sub) <- c("Culmen\nlength", "Culmen\ndepth", "Flipper\nlength", "Body\nmass")
+
+## Morphospaces:
+## No transformations
+write.csv(head(penguins_sub), file = "../Writing/Figures/trait.space.no_trans.csv")
+
+## Distance matrix
+penguins_sub_dist <- dist(penguins_sub)
+write.csv(round(head(as.matrix(penguins_sub_dist)[1:5, 1:5]), digits = 2), file = "../Writing/Figures/trait.space.distance.csv")
+
+## Ordinations
+penguins_sub_pca <- princomp(apply(penguins_sub, 2, scale))$scores
+write.csv(round(head(penguins_sub_pca), digits = 2), file = "../Writing/Figures/trait.space.pca.csv")
+penguins_sub_pco <-cmdscale(penguins_sub_dist, k = 2)
+
+## Visualisations
+## biplots
+col_var <- c("orange", "blue", "green")[as.factor(penguins_sp)]
+pdf(file = "../Writing/Figures/figure_methods_biplots.pdf", width = 5, height = 5)
+plot(penguins_sub,
+    pch = 19,
+    col = col_var)
+dev.off()
+
+## PCO
+pdf(file = "../Writing/Figures/figure_methods_PCO.pdf", width = 5, height = 5)
+
+vars <- round(apply(penguins_sub_pco, 2, var)/sum(apply(penguins_sub_pco, 2, var))*100, 2)
+plot(penguins_sub_pco,
+    pch = 19,
+    col = col_var,
+    xlab = paste0("PCO 1 (", vars[1], "%)"),
+    ylab = paste0("PCO 21 (", vars[2], "%)"),
+    xlim = range(penguins_sub_pco),
+    ylim = range(penguins_sub_pco))
+dev.off()
+
+## PCA
+pdf(file = "../Writing/Figures/figure_methods_PCA.pdf", width = 5, height = 5)
+vars <- round(apply(penguins_sub_pca, 2, var)/sum(apply(penguins_sub_pca, 2, var))*100, 2) 
+ranges <- apply(penguins_sub_pca, 2, range)
+
+par(mfrow = c(4,4), mar = c(1,1,1,1), oma = c(2,2,2,2))
+## First row
+# 1
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PC 1\n(", vars[1], "%)"), cex = 1.9)
+# 2
+plot(penguins_sub_pca[, c(1,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(3)
+# 3
+plot(penguins_sub_pca[, c(1,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 4
+plot(penguins_sub_pca[, c(1,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(3)
+axis(4)
+
+## Second row
+# 1
+plot(penguins_sub_pca[, c(2,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(2)
+# 2
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PC 2\n(", vars[2], "%)"), cex = 1.9)
+# 3
+plot(penguins_sub_pca[, c(2,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 4
+plot(penguins_sub_pca[, c(2,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(4)
+
+## Third row
+# 1
+plot(penguins_sub_pca[, c(3,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(2)
+# 3
+plot(penguins_sub_pca[, c(3,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 3
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PC 3\n(", vars[3], "%)"), cex = 1.9)
+# 4
+plot(penguins_sub_pca[, c(3,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(4)
+
+## Fourth row
+# 1
+plot(penguins_sub_pca[, c(4,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(1)
+axis(2)
+# 3
+plot(penguins_sub_pca[, c(4,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+# 3
+plot(penguins_sub_pca[, c(4,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
+    pch = 19,
+    col = col_var)
+axis(1)
+# 4
+plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
+text(0,0, paste0("PC 4\n(", vars[4], "%)"), cex = 1.9)
+dev.off()
+
+## PCA with rotation
+pdf(file = "../Writing/Figures/figure_methods_PCA_rotation.pdf", width = 5, height = 5)
+
+peguins_ordination <- princomp(apply(penguins_sub, 2, scale))
+
+xaxis_loading <- peguins_ordination$loadings[2,1]
+yaxis_loading <- peguins_ordination$loadings[1,2]
+
+plot(apply(penguins_sub, 2, scale)[,c(1,2)],
+    pch = 19,
+    col = col_var,
+    xlab = "Scaled culmen length",
+    ylab = "Scaled culmen depth")
+
+arrows(x0 = 0, y0 = 0, x1 = 1, y1 = xaxis_loading, lwd = 2.5)
+arrows(x0 = 0, y0 = 0, x1 = yaxis_loading, y1 = 1, lwd = 2.5)
+text(x = 1.35, y = xaxis_loading, "PC1", cex = 2)
+text(x = yaxis_loading+0.05, y = 1.15, "PC2", cex = 2)
+
+dev.off()
